### 2020-05-14 why-negative-selection.Rmd
Initial commit

---
title: "Negative Selection in the 20th Century"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "April 2020"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    keep_tex: true
---

```md
# TODO
* Improve imputation of parents' YOB.
* Selection bias. 
  - You could weight subjects by education level; sex; birth region?

## Data to gather
* f.2139 - age first had sex (includes "never had sex" which may explain
  some of the many NAs for f.2141, num sex partners)

```

```{r setup, include = FALSE}

library(drake)
library(magrittr)
library(dplyr)
library(forcats)
library(ggplot2)
library(tidyr)
library(santoku) # keep this after tidyr as it masks chop
library(purrr)
library(glue)
library(broom)
library(huxtable)

drake::loadd(famhist)
drake::loadd(rgs)
drake::loadd(score_names)
drake::loadd(reverse_code)

options(digits = 2)
knitr::opts_chunk$set(
        echo = FALSE, 
        warning = FALSE, 
        cache = FALSE, 
        error = TRUE
      )

theme_set(theme_minimal())
pval_palette <- c("FALSE" = "grey50", "TRUE" = "orange2")
my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dashed")
order_abs <- function (x) x[1] # for use when reordering factors

huxtable::set_default_properties(latex_float = "h!")

pretty_names <- function (names) {
  reverse <- names %in% reverse_code
  names <- gsub("_", " ", names)
  names %<>% 
        strsplit(" ") %>% 
        map_chr(~paste0(
          toupper(substring(.x, 1, 1)), 
          substring(.x, 2),
          collapse = " "
        ))
  names[reverse] %<>% paste0("*")

  return(names)
}

```

# Data

Data is taken from UK Biobank. Polygenic scores were  normalized to mean 0,
variance 1. Scores for ADHD, substance use, alcohol, Alzheimer's disease,
autism, bipolarity, BMI, body fat, caffeine, cannabis use, coronary artery
disease, eating disorders, waist-hip ratio, major depressive disorder,
neuroticism, and schizophrenia were reverse coded. High scores are therefore
"positive" in the loose sense that their target phenotypes would be considered
desirable by many people.

# Results

We run regressions on two dependent variables:

* *total siblings*, the number of full siblings of the respondent (including
  himself or herself).
* The number of *children* ever born to/fathered by the respondent.

We run regressions both with and without controls for the 40 top principal components
of the genetic data.

```{r regs-data}

drake::loadd(res_all)

res_all_wide <- res_all %>% 
  select(reg.type, dep.var, term, estimate) %>% 
  pivot_wider(names_from = c(reg.type, dep.var), values_from = estimate) %>% 
  mutate(
    consistent = sign(`raw_N siblings`) == sign(`controlled_N siblings`) &
      sign(`raw_N siblings`) == sign(`raw_N children`) &
      sign(`raw_N siblings`) == sign(`controlled_N children`)
  )

n_pgs  <- as.double(nrow(res_all_wide))
prop_consistent <- mean(res_all_wide$consistent)
controlled_smaller_sibs <- abs(res_all_wide[["controlled_N siblings"]]) < 
  abs(res_all_wide[["raw_N siblings"]])
controlled_smaller_children <- abs(res_all_wide[["controlled_N children"]]) < 
  abs(res_all_wide[["raw_N children"]])
median_prop_sibs <- median(abs(res_all_wide[["controlled_N siblings"]]) /
    abs(res_all_wide[["raw_N siblings"]]))
median_prop_children <- median(abs(res_all_wide[["controlled_N children"]]) /
    abs(res_all_wide[["raw_N children"]]))

```

Figure \@ref(fig:plot-basic-regs) shows effect sizes of a one-standard
deviation shift in each polygenic score.

Estimates are broadly consistent across generations. For 
`r sum(res_all_wide$consistent)` out of `r n_pgs` polygenic scores, all 4 estimates have the same sign.

Within siblings regressions, effect sizes are smaller when controlling for
principal components -- sometimes much smaller, as in the case of height. 
`r sum(controlled_smaller_sibs)` out of `r n_pgs` "controlled" effect sizes have 
a smaller absolute value than the corresponding "raw" effect size. The median
proportion between raw and controlled effect sizes is `r median_prop_sibs`.
Among the children regressions, this no longer holds. Effect sizes are barely
affected by controlling for principal components.

```{r plot-basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children. Scores marked * are reverse coded.", fig.height = 7}

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate, order_abs),
        dep.var = fct_inorder(dep.var) 
      ) %>% 
      ggplot(aes(x = term, y = estimate, shape = reg.type, 
        colour = p.value < 0.05/{{n_regs}})) +
      my_hline + 
      geom_point(size = 2.5, alpha = 0.8) + 
      labs(y = "Effect size", x = "") +
      facet_wrap(vars(dep.var), ncol = 2, scales = "free") +
      coord_flip() +
      scale_colour_manual(values = pval_palette) +
      scale_shape_manual(values = c("circle", "triangle"), 
        guide = guide_legend(title = "")) +
      theme(
        legend.position = "bottom", 
        legend.margin = margin(2,2,2,2)
      )

```

# Selection over time

```{r period-regs-data}

drake::loadd(res_period)
res_period %<>% 
      tidyr::extract(term, c("score_name", "period"),
        regex="(.*):year_split(.*)")
n_tests <- nrow(res_period)/2

summary_period <- res_period %>% 
      select(-std.error, -statistic) %>% 
      pivot_wider(
        names_from  = period, 
        values_from = c(estimate, conf.low, conf.high, p.value)
      ) %>% 
      group_by(score_name, children) %>% 
      mutate(
        significant = diff.p.value < 0.05/n_tests,
        increasing  = estimate_early < estimate_late,
        size_inc    = abs(estimate_early) < abs(estimate_late),
        change_sign = sign(estimate_early) != sign(estimate_late),
        Change      = case_when(
                        ! significant ~ "Insignificant",
                        change_sign  & increasing   ~ "Change sign to +",
                        change_sign  & ! increasing ~ "Change sign to -",
                        increasing   & size_inc   ~ "Increasing +",
                        ! increasing & size_inc   ~ "Decreasing -",
                        increasing   & ! size_inc ~ "Increasing towards 0",
                        ! increasing & ! size_inc ~ "Decreasing towards 0"
                      )
      ) %>% 
      select(score_name, children, Change, increasing, significant) %>% 
      mutate(children = ifelse(children, "Respondents", "Parents")) %>% 
      pivot_wider(names_from = children, 
        values_from = c(Change, increasing, significant)) %>% 
      mutate(overall = case_when(
              increasing_Respondents   & increasing_Parents   ~ 
                "Consistent increase",
              ! increasing_Respondents & ! increasing_Parents ~ 
                "Consistent decrease",
              increasing_Respondents & ! increasing_Parents
                 & (significant_Parents | significant_Respondents) ~ 
                "Changes direction",
              TRUE ~ 
              "Other"
            ))

```

Negative selection seems to decrease over time. Figure \@ref(fig:plot-by-period) shows
effect sizes for *number of siblings* and *number of children*, median-split by
parents' year of birth and own year of birth respectively.  Parents' year of
birth is imputed, which is likely to produce some bias.

```{r plot-by-period, fig.cap = "Effect sizes of PGS on number of children/siblings by own/parents' year of birth. PGS with the largest mean effect sizes are shown.", fig.height = 7, fig.width = 6}

max_abs <- function (x) max(abs(x))
res_period %>% 
      mutate(
        dep.var       = ifelse(children, "Children", "Siblings"),
        period        = (period == "late") + children,
        score_name    = pretty_names(score_name),
        score_name    = fct_reorder(score_name, estimate, max_abs, 
                          .desc = TRUE),
        p             = diff.p.value
      ) %>% 
      arrange(score_name) %>% 
      slice(1:48) %>% 
ggplot(aes(period, estimate, colour = dep.var, group = dep.var)) + 
      geom_hline(yintercept = 0, colour = "grey30") + 
      geom_point(size = 1.5) +
      geom_line(aes(linetype = p < 0.05/{{n_tests}})) +
      facet_wrap(vars(score_name), nrow = 4) + 
      scale_fill_manual(values = pval_palette) +
      scale_colour_manual(values = c("darkgreen", "navy")) +
      scale_linetype_manual(values = c("FALSE" = "dashed", "TRUE" = "solid"))

```

By definition, the sibling regressions exclude members of the parents'
generation who had no children. This is likely to bias results towards zero, since
much of the effect in children regressions is due to respondents with high scores
being more likely to have no children. So, we cannot directly compare 
effect sizes for the two sets of regressions. Within the sibling regressions,
the most common pattern is that negative effects shrink in absolute size 
(Table \@ref(tab:siblings-by-period)).

```{r siblings-by-period}

period_table <- function (tbl, caption) {
  tbl %>% 
      as_hux() %>%
      set_bold(1, everywhere) %>% 
      set_bottom_border(1, everywhere) %>%

      add_footnote(glue("Significance is measured at p < 0.05/{n_tests}"), 
        border = 0.4) %>% 
      set_contents(1, everywhere, 
        c("Change", "Number of scores")) %>% 
      set_right_border(final(1), 1, 0) %>% 
      set_top_padding(1) %>% 
      set_bottom_padding(1) %>%
      set_left_padding(everywhere, 1, 0) %>% 
      set_right_padding(everywhere, 2, 0) %>% 
      set_caption(caption) %>% 
      set_caption_width(1)
}

summary_period %>% 
      group_by(Change_Parents) %>% 
      count() %>% 
      period_table("Change in effect sizes between early and late born parents, 'sibling' regressions")

```

In children regressions, no clear pattern is visible 
(Table \@ref(tab:children-by-period)).

```{r children-by-period}
summary_period %>% 
      group_by(Change_Respondents) %>% 
      count() %>% 
      period_table("Change in effect sizes between early and late born respondents, 'children' regressions")
```

Despite this, effect sizes tend to be smaller for children regressions. This
could be caused by ascertainment bias in the UK Biobank sample -- e.g.,
if respondents themselves are a more selected sample than the respondents' 
parents.

# Causality

```{r pgs-corrs}
cor_tbl <- cor(famhist[score_names], use = "pair") %>% 
      as_tibble() %>% 
      mutate(score_name = score_names, .before = 1) %>% 
      pivot_longer(cols = -score_name, values_to = "Correlation") %>% 
      filter(score_name < name) %>% 
      arrange(desc(Correlation)) %>% 
      slice(1:10) %>% 
      mutate(across(c(score_name, name), pretty_names))

as_hux(cor_tbl, add_colnames = TRUE) %>% 
      set_contents(1, 1:2, c("PGS", "PGS")) %>% 
      theme_basic(header_col = FALSE) %>% 
      set_bottom_border(1, everywhere, 0.4) %>% 
      set_bottom_border(final(1), everywhere, 0.4) %>% 
      set_caption("Top 10 correlations between polygenic scores")

```

Different polygenic scores are correlated. Table \@ref(tab:pgs-corrs) shows the top
correlations in the sample. Because of this, bivariate correlations between
PGS and number of children might be driven by other genetic scores. To explore which polygenic scores are driving negative selection, we run a single
omnibus regression of *number of children* on all the PGS. We exclude EA2, 
waist-hip ratio, WC (XXX what is it?) and "Hip combined" since they are 
highly correlated with other scores, which could make our estimates unstable. Figure \@ref(fig:plot-together) shows the results. Interestingly, several PGS remain independently significant, although
effect sizes are reduced.

```{r plot-together, fig.cap = "Partial correlations with number of children"}

loadd(res_together)
n_regs <- as.double(nrow(res_together))
res_together %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate),
        dep.var = recode(dep.var, 
          "n_children" = "N Children", "n_sibs" = "N Siblings")
      ) %>% 
      ggplot(aes(term, estimate, color = p.value < 0.05/{{n_regs}}, shape  = PCs)) + 
        geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
        my_hline +
        facet_wrap(vars(dep.var), scales = "free_x") +
        coord_flip() +
        scale_color_manual(values = pval_palette) +
        theme(legend.position = "bottom")
```

# Subgroups

We next examine how different subgroups contribute to natural selection.

## Males and females

```{r}

drake::loadd(res_sex)

female_lower <- res_sex %>% 
      pivot_wider(id_cols = c(term), names_from = sex, 
        values_from = c(estimate, conf.low, conf.high)) %>% 
      mutate(female_lower = estimate_Female < estimate_Male) %>% 
      pull(female_lower)
```

Figure \@ref(fig:plot-sexes) shows effect sizes of PGS on number of children
separately for males and females. For `r sum(female_lower)` out of
`r length(female_lower)` PGS, selection is more negative for women
than for men. Differences are particularly large for educational attainment and height PGS.

```{r plot-sexes, fig.cap = "Effect sizes on number of children by sex"}

res_sex %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, estimate, order_abs)
      ) %>% 
      ggplot(aes(term, estimate, group = sex, colour = sex)) +
        my_hline +
        geom_pointrange(aes(ymin = conf.low, ymax = conf.high), alpha = 0.7) +
        coord_flip() +
        scale_colour_brewer(palette = 2, type = "qual")

```

## Number of sexual partners

Figure \@ref(fig:plot-n-partners) splits males and females up by lifetime number
of sexual partners. Remarkably, across both sexes, negative selection is
strongly reversed for respondents who had 3 or fewer sexual partners in their
lifetime.

```{r plot-n-partners, fig.cap = "Effect sizes on number of children by number of sexual partners", fig.height = 7, fig.width = 6}

drake::loadd(res_partners)
res_partners %<>%       
      filter(term != "(Intercept)", term != "lo_partnersTRUE") %>% 
      mutate(`N partners` = ifelse(grepl(":", term), "<= 3", "> 3"))

n_regs <- nrow(res_partners)
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name),
        score_name = fct_reorder(score_name, estimate, order_abs)
      ) %>% 
      ggplot(aes(score_name, estimate, colour = `N partners`)) + 
        my_hline +
        geom_pointrange(aes(ymin = conf.low, ymax = conf.high), alpha = .8) +
        facet_wrap(vars(sex)) +
        coord_flip() +
        scale_colour_manual(values = c("darkred", "steelblue")) +
        theme(legend.position = "bottom")

```

## Education levels

Figure \@ref(fig:plot-educ-levels) splits respondents up by education levels.
Both negative and positive selection are typically larger and more significant
for those who left school before 16. Table \@ref(tab:educ-levels) summarizes
the results.

```{r educ-levels}
loadd(res_edu)
res_edu %>% 
      mutate(
        age_fte_cat = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18")
      ) %>% 
      group_by(age_fte_cat) %>% 
      summarize(
        prop_sig = 100 * mean(p.value < 0.05/nrow(res_edu)), 
        .groups  = "drop"
      ) %>% 
      as_huxtable(add_colnames = TRUE) %>% 
      set_contents(1, everywhere, 
        c("Age left FTE", "% PGS significant")) %>%
      set_caption("Negative selection by education level") %>% 
      set_bold(1, everywhere, TRUE) %>% 
      set_bottom_border(1, everywhere, 0.4) %>% 
      set_left_padding(everywhere, 1, 0) %>% 
      set_right_padding(everywhere, final(1), 0) %>% 
      set_top_padding(14)

```

```{r plot-educ-levels, fig.cap = "Effect sizes on number of children by age left full-time education", fig.height = 7}

loadd(res_edu)
res_edu %>% 
      mutate(
        term = pretty_names(term), 
        term = fct_reorder(term, estimate, order_abs),
        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18")
      ) %>% 
      ggplot(aes(term, estimate, colour = `Age left FTE`)) +
        my_hline +
        geom_pointrange(aes(ymin=conf.low, ymax=conf.high), alpha = 0.8,
          position = position_dodge(width = 0.4)) + 
        scale_colour_viridis_d(begin = 0.1, end = 0.85) +
        coord_flip()
```

# Number of children

Figure \@ref(fig:plot-n-children-bar) shows the full distribution of number 
of children born for different ventiles of the EA3 polygenic score. The 
strongest relationship seems to be for having 0 children versus 1 or more.

```{r plot-n-children-bar, fig.cap = "Number of children by ventiles of EA3 PGS"}
famhist %>% 
      select(n_children, EA3_excl_23andMe_UK, sex) %>% 
      mutate(
        "EA3 Ventile" = chop_equally(EA3_excl_23andMe_UK, 20, labels = 1:20),
        "N Children"  = santoku::chop(n_children, brk_left(0:4, FALSE), 
                          labels = c(0:3, "4+")),
        Sex           = factor(sex, levels = 0:1, labels = c("Female", "Male"))
        ) %>% 
      filter(! is.na(`EA3 Ventile`), ! is.na(`N Children`)) %>% 
      ggplot(aes(`EA3 Ventile`, fill = `N Children`, group = `N Children`)) + 
        geom_bar(position = position_fill(reverse = TRUE), alpha = 0.7) + 
        facet_wrap(vars(Sex)) +
        scale_fill_viridis_d(option = "A") +
        scale_y_continuous(n.breaks = 10, labels = scales::percent) +
        theme(legend.position = "bottom")

```

```{r plot-sd-years, eval = FALSE}
# inequality over time. Nothing to see, almost everything is flat.
grid <- expand_grid(score_name = score_names, YOB = 1940:1965)  
grid$sd <- pmap_dbl(grid, ~sd(famhist[famhist$YOB==.y, ][[.x]], na.rm = TRUE))
ggplot(grid, aes(YOB, sd)) + 
      geom_line() + 
      facet_wrap(~score_name, scales = "free_y", ncol = 4)
```

```{r rgs-by-effect-size, eval = FALSE, fig.cap = "Genetic correlations with EA3 by effect size on n siblings."}
rgs <- left_join(rgs, res %>% filter(reg.type == "residualized"), by = c("p2" = "term"))

rgs %>% 
      filter(p2 != "EA3_excl_23andMe_UK", p2 != "EA2_noUKB") %>% 
      ggplot(aes(rg, estimate)) + 
        geom_smooth(method = lm) +
        geom_point(colour = "red2") + 
        geom_label(aes(label = p2, hjust = "left", vjust = "middle"), 
          size = 3, nudge_x = 0.02, alpha = 0.5, 
          label.padding = unit(0.1, "lines")) +
        labs(x = "rG", y = "Effect size", 
          title = "rG with EA3 vs. effect size on # siblings")
```

```{r correlate-effect-sizes, eval = FALSE}
combined_results <- res %>% 
      mutate(PCs = ifelse(reg.type=="raw", "no", "yes")) %>% 
      dplyr::select(-reg.type) %>% 
      {bind_rows(bivariate = ., multivariate = all_res, .id = "reg.type")} %>% 
      arrange(term, reg.type, PCs) %>% 
      dplyr::select(term, reg.type, PCs, estimate) %>% 
      pivot_wider(names_from = c(reg.type, PCs), values_from = estimate)

cor(combined_results[-1], use = "complete")

```

### 2020-05-14 why-selection-against-EA-parents.Rmd
Initial commit

---
title: "Negative Selection in the 20th Century"
author: "David Hugh-Jones"
date: "April 2020"
output: 
  bookdown::pdf_document2:
    toc: false
---

```{r setup, include = FALSE}

draft <- FALSE # set to FALSE to reset the cache

pgs_dir  <- "~/UKBB data 2019/polygenic_scores/"
ph_file  <- "~/UKBB data 2019/UKB.EA_pheno.coordinates.QC.david.csv"
pcs_file <- "~/UKBB data 2019/ukb30545.40PCs.csv"
famhist_file <- "~/Dropbox/assortative mating/biobank-analysis/david.family_history.traits.out.csv"
famhist2_file <- "~/UKBB data 2019/david.family_history.traits.20042020.out.csv"
famhist3_file <- "~/UKBB data 2019/david.family_history.traits.05052020.out.csv"
rgs_file <- "EA3_rgs.10052019.rgs.csv"
mf_pairs_file <- "~/UKBB data 2019/spouse_pair_info/UKB_out.mf_pairs_rebadged.csv"

library(readr)
library(multcomp) # keep before dplyr, it masks select
library(magrittr)
library(dplyr)
library(forcats)
library(ggplot2)
library(tidyr)
library(santoku) # keep this after tidyr as it masks chop
library(purrr)
library(glue)
library(broom)
library(huxtable)

knitr::opts_chunk$set(
        echo = FALSE, 
        warning = FALSE, 
        cache = draft, 
        error = draft
      )
theme_set(theme_minimal())

order_abs <- function (x) x[1] # for use when reordering factors

ph_spec <- spec_csv(ph_file)
ph <- read_csv(ph_file, col_types = cols(
        geno_measurement_plate = col_skip(),
        geno_measurement_well = col_skip(),
        .default = col_double()
      ))
pcs <- read_csv(pcs_file)
rgs <- read_csv(rgs_file)
famhist <- read_csv(famhist_file, col_types = strrep("d", 40))
famhist2 <- read_csv(famhist2_file, col_types = strrep("d", 17))
famhist3 <- read_csv(famhist3_file)
mf_pairs <- read_csv(mf_pairs_file, col_types = "dddccccc")

names(famhist) <- paste0("f.", names(famhist))
names(famhist) <- gsub("\\-", ".", names(famhist))
names(famhist2) <- paste0("f.", names(famhist2))
names(famhist2) <- gsub("\\-", ".", names(famhist2))
names(famhist3) <- paste0("f.", names(famhist3))
names(famhist3) <- gsub("\\-", ".", names(famhist3))

famhist %<>% left_join(pcs, by = c("f.eid" = "eid"))
famhist %<>% left_join(ph, by = c("f.eid" = "eid"))
famhist %<>% left_join(famhist2, by = "f.eid")
famhist %<>% left_join(famhist3, by = "f.eid")

# only "genetic" whites
famhist %<>% filter(! is.na(genetic_ethnic_grouping))

for (pgs_file in list.files(pgs_dir, pattern = "csv$", full.names = TRUE)) {
  score_name <- sub(".*UKB\\.AMC\\.(.*?)\\..*", "\\1", pgs_file, perl = TRUE)

  pgs <- read_delim(pgs_file, delim = " ", col_types = "dd")

  pgs %<>% filter(FID > 0) 
  names(pgs)[2] <- score_name # instead of "SCORE"
  pgs[[score_name]] <- c(scale(pgs[[score_name]]))

  famhist %<>% left_join(pgs, by = c("f.eid" = "FID"))

  resid_fml <- paste(score_name, "~", paste0("PC", 1:40, collapse = " + "))
  resid_score <- resid(lm(as.formula(resid_fml), famhist, na.action = na.exclude))
  famhist[[paste0(score_name, "_resid")]] <- c(scale(resid_score))
}

score_names <- sub(".*UKB\\.AMC\\.(.*?)\\..*", "\\1", 
      list.files(pgs_dir, pattern = "csv$"), perl = TRUE)

# we get very few extra cases from adding f.2946.1.0 etc, and it makes calculating
# father's year of birth more complex
# roughly speaking, these are ages in 2007-10
# TODO: get f.53 from Abdel for these EIDs
famhist$fath_age <- famhist$f.2946.0.0
famhist$fath_age[famhist$fath_age < 0] <- NA
famhist$moth_age <- famhist$f.1845.0.0
famhist$moth_age[famhist$moth_age < 0] <- NA

# full brothers and sisters
famhist$nbro <- pmax(famhist$f.1873.0.0, famhist$f.1873.1.0, 
      famhist$f.1873.2.0, na.rm = TRUE)
famhist$nbro[famhist$nbro < 0] <- NA
famhist$nsis <- pmax(famhist$f.1883.0.0, famhist$f.1883.1.0, 
      famhist$f.1883.2.0, na.rm = TRUE)
famhist$nsis[famhist$nsis < 0] <- NA
famhist$n_sibs <- famhist$nbro + famhist$nsis

famhist$n_partners <- pmax(famhist$f.2149.0.0, famhist$f.2149.1.0, 
     famhist$f.2149.2.0, na.rm = TRUE)
famhist$n_partners[famhist$n_partners < 0] <- NA

famhist$n_children <- pmax(famhist$f.2405.0.0, famhist$f.2405.1.0,
        famhist$f.2405.2.0, famhist$f.2734.0.0, famhist$f.2734.1.0, 
        famhist$f.2734.2.0, 
        na.rm = TRUE
      )
famhist$n_children[famhist$n_children < 0] <- NA

child_vars <- quos(f.2754.0.0, f.2754.1.0, f.2754.2.0)
famhist %<>% mutate_at(vars(!!!child_vars), ~ ifelse(.x < 0, NA, .x))
famhist$age_flb <- rowMeans(famhist %>% select(!!!child_vars), na.rm = TRUE)
famhist$year_flb <- famhist$YOB + famhist$age_flb

rgs$p2 <- rgs$p2 %>% recode(
        EA3_excl_23andMe_and_allUK = "EA3_excl_23andMe_UK",
        cognitve_ability.noUKB     = "cognitive_ability",
        autism_2017.ipsych.pgc     = "autism_2017",
      ) %>% {sub("\\.GPC\\.23andme$", "", .)}

rm(famhist2, famhist3, pgs, pcs, ph)
```

```{r make-pairs, eval = FALSE}
famhist_tmp <- famhist %>% dplyr::select(! (starts_with("f.") | 
      starts_with("PC") | starts_with("home") |
      starts_with("assessment") | starts_with("birth_")), f.eid)
mf_pairs %<>% 
      left_join(famhist_tmp, by = c("ID.m" = "f.eid")) %>% 
      left_join(famhist_tmp, by = c("ID.f" = "f.eid"), suffix = c(".m", ".f"))
rm(famhist_tmp)

```

# Results

We run regressions on two dependent variables:

* log(total siblings + 1), where *total siblings* includes the
  respondent. 
* log(children + 1) where *children* is the number of 
  children ever born to/fathered by the respondent.

We run regressions both with and without controls for the 40 top principal components
of the genetic data.

```{r run-regressions}
# for each score, calculate the correlation with number of siblings
# 
run_regs <- function (dep_var, score_name) {
  fml <- as.formula(glue("{dep_var} ~ {score_name}"))
  reg_bv <- tidy(lm(fml, famhist), conf.int = TRUE)

  pcs <- paste0("PC", 1:40, collapse = " + ")
  fml_resid <- as.formula(glue("{dep_var} ~ {score_name} + {pcs}"))
  reg_resid <- tidy(lm(fml_resid, famhist), conf.int = TRUE)

  regs <- bind_rows(raw = reg_bv, controlled = reg_resid, .id = "reg.type")
  regs %<>% filter(term != "(Intercept)", ! grepl("^PC", term))

  return(regs)
}

# we use n_sibs + 2 to be directly comparable to log(n_children + 1)
res_sibs <- map_dfr(score_names, run_regs, dep_var = "log(n_sibs + 2)")
res_chn <- map_dfr(score_names, run_regs, dep_var = "log(n_children + 1)")

res_all <- bind_rows("N siblings" = res_sibs, "N children" = res_chn, 
      .id = "dep.var")

res_all_wide <- res_all %>% 
      select(reg.type, dep.var, term, estimate) %>% 
      pivot_wider(names_from = c(reg.type, dep.var), values_from = estimate) %>% 
      mutate(
        consistent = sign(`raw_N siblings`) == sign(`controlled_N siblings`) &
                     sign(`raw_N siblings`) == sign(`raw_N children`) &
                     sign(`raw_N siblings`) == sign(`controlled_N children`)
      )

n_regs <- as.double(nrow(res_all))
n_pgs  <- as.double(nrow(res_all_wide))
prop_consistent <- mean(res_all_wide$consistent)
controlled_smaller_sibs <- abs(res_all_wide[["controlled_N siblings"]]) < 
               abs(res_all_wide[["raw_N siblings"]])
controlled_smaller_children <- abs(res_all_wide[["controlled_N children"]]) < 
               abs(res_all_wide[["raw_N children"]])
median_prop_sibs <- median(abs(res_all_wide[["controlled_N siblings"]]) /
               abs(res_all_wide[["raw_N siblings"]]))
median_prop_children <- median(abs(res_all_wide[["controlled_N children"]]) /
               abs(res_all_wide[["raw_N children"]]))
```

Figure \@ref(fig:basic-regs) shows effect sizes of a one-standard
deviation shift in each polygenic score.

Estimates are broadly consistent
across generations. For `r sum(res_all_wide$consistent)` out of 
`r n_pgs` polygenic scores, all 4 estimates have the same sign.

Within siblings regressions, effect sizes are smaller when controlling for 
principal components -- sometimes much smaller, as in the case of height. 
`r sum(controlled_smaller_sibs)` out of `r n_pgs`
"controlled" effect sizes have a smaller absolute value than the corresponding 
"raw" effect size. The median proportion between raw and controlled effect sizes
is just `r median_prop_sibs`.

Among the children regressions, this no longer holds. Only 
`r sum(controlled_smaller_children)` residualized effect sizes out of `r n_pgs`
are smaller. The median proportion between effect sizes is `r median_prop_children`.

```{r basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children", fig.height = 7}

shape21 <- guide_legend(override.aes = list(shape = 21))
res_all %>% 
      mutate(term = fct_reorder(term, estimate, order_abs)) %>% 
      ggplot(aes(x = term, y = estimate, fill = reg.type, shape = dep.var,
        colour = p.value < 0.05/{{n_regs}})) +
      geom_hline(yintercept = 0, colour = "grey40", linetype = "dashed") + 
      geom_point(stroke = 0.75, size = 2) + 
      labs(y = "Effect size", x = "") +
      coord_flip() +
      scale_shape_manual(values = c("N siblings" = 21, "N children" = 23)) +
      scale_fill_brewer(guide = shape21, type = "qual", palette = 3) +
      scale_colour_manual(
        values = c("FALSE" = "grey80", "TRUE" = "orange2"), 
        guide = shape21
      )

```

# Selection over time

```{r}
run_regs_time <- function(children, score_name, period) {
  dep_var  <- if (children) "log(n_children + 1)" else "log(n_sibs + 2)"
  year_var <- if (children) "year_flb" else "YOB"
  famhist$in_period <- ! is.na(famhist[[year_var]]) & famhist[[year_var]] %in% period:(period+4)

  dep_var_col <- if (children) "n_children" else "n_sibs"
  fh_subset <- famhist[
          famhist$in_period & 
          ! is.na(famhist[[dep_var_col]]) & 
          ! is.na(famhist[[score_name]])
        , ]
  # for strict comparability to the siblings regression, we include only
  # those with 1 child or more
  if (children) fh_subset <- fh_subset[fh_subset$n_children > 0, ] 
  if (nrow(fh_subset) < 1000) return(NULL)

  fml <- as.formula(glue("{dep_var} ~ {score_name}"))
  mod <- lm(fml, fh_subset)
  res <- tidy(mod, conf.int = TRUE)
  res <- filter(res, term == score_name)
  res$period <- period
  res$children <- children
  res
}

pars <- expand_grid(
        children = c(TRUE, FALSE),
        score_name = score_names, 
        period = seq(1940, 2000, 5)
      )
res_yr <- pmap_dfr(pars, run_regs_time)

```

```{r over-time}

ggplot(res_yr, aes(period, estimate, colour = children)) + 
      geom_linerange(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
      geom_line() + 
      facet_wrap(vars(term))
```

Selection over time appears to be decreasing, if we focus on raw PGS. Figure
\@ref(fig:effect-sizes-by-period) shows effect sizes for  survey respondents
born in the 1930s and 1940s, versus respondents born in the 1950s to 1970s.

```{r effect-sizes-by-period, fig.cap = "Effect sizes of PGS on n children by period. Significance tests compare early to late effect sizes", fig.width = 6, fig.height = 7}

famhist$period <- chop(famhist$YOB, c(1930, 1950, 1974), 
      labels = c("30s_40s", "50s_70s"))

run_time_regs <- function(score_name) {

  fml <- as.formula(glue("log(n_sibs+1) ~ {score_name}:period"))
  mod <- lm(fml, famhist)
  res <- tidy(mod)
  res <- filter(res, ! term %in% c("(Intercept)"))
  res$term <- gsub(".*:period", "", res$term)

  glhts <- list(
    glht(mod, glue("{score_name}:period30s_40s - {score_name}:period50s_70s = 0"))
  )
  glhts <- glhts %>% map(summary) %>% map(tidy) %>% map_dbl("p.value")
  res$glht.p.value <- c(NA, glhts)

  res
}

names(score_names) <- score_names
time_regs <- map_dfr(score_names, run_time_regs, .id = "reg")

time_regs %<>% 
      rename(period = term) %>% 
      mutate(reg = fct_reorder(reg, estimate, first))

time_regs_wide <- time_regs %>% 
      select(reg, period, estimate, glht.p.value) %>% 
      pivot_wider(reg, names_from = period, 
        values_from = c(estimate, glht.p.value)) %>% 
      mutate(Significant = glht.p.value_50s_70s < 0.05/(nrow(time_regs) * 3/4))

time_regs %>% 
      group_by(reg) %>% 
      mutate(
        Significant = na.omit(glht.p.value) < 0.05/(nrow(time_regs) * 3/4)
      ) %>% 
      ggplot(aes(x = reg)) +
        geom_hline(yintercept = 0, colour = "darkgrey") +
        geom_linerange(aes(
          ymin = estimate_30s_40s, 
          ymax = estimate_50s_70s, colour = Significant), 
          data = time_regs_wide) + 
        geom_point(aes(y = estimate, fill = period, colour = Significant), 
          shape = 21, size = 2) + 
        scale_colour_manual(values = c("grey", "orange3")) +
        scale_fill_brewer(type = "qual", palette = 3) +
        coord_flip()
```

# Causality

To be selected against, an allele does not need to cause fewer children, only to
be correlated with fewer children. Nevertheless, it is interesting to ask
whether individual PGS caused sample members' parents to have fewer children. As
a first cut, we regress log(number of siblings _+ 1) on all PGS together. We run
two regressions: one controlling for genetic principal components, and one
without these controls. Figure \@ref(fig:multivariate-reg-plots) shows the results.

```{r multivariate-regs}

# EA3 and EA2 have a 0.9 raw correlation. These are the only pair of PGS with
# a raw correlation of more than 0.8
reg_score_names <- setdiff(score_names, "EA2_noUKB")

without_pcs <- paste0("log(n_sibs+1) ~ ", paste(reg_score_names, collapse = " + "))
fml <- as.formula(without_pcs)
fml_with_pcs <- as.formula(paste(without_pcs, "+", paste0("PC", 1:40, 
      collapse = " + ")))

all_mod <- lm(fml, famhist)
all_res <- tidy(all_mod, conf.int = TRUE)

all_mod_pcs <- lm(fml_with_pcs, famhist)
all_res_pcs <- tidy(all_mod_pcs, conf.int = TRUE)

all_res <- bind_rows(no = all_res, yes = all_res_pcs, .id = "PCs")
all_res <- filter(all_res, term != "(Intercept)", ! grepl("^PC", term))

```

```{r multivariate-reg-plots, fig.cap = "Multivariate regressions. Significance tests compare effect sizes to zero."}

all_res$term <- fct_reorder(all_res$term, all_res$estimate, order_abs)

n_regs <- as.double(nrow(all_res)) # avoids an ugly "L"
ggplot(all_res, aes(term)) + 
      geom_point(aes(y = estimate, fill = PCs, 
        colour = p.value <= 0.05/{{n_regs}}), shape = 21) + 
      scale_colour_manual(values = c("grey", "black")) +
      geom_hline(yintercept = 0, colour = "darkgrey") +
      labs(x = "PGS", y = "Effect size") +
      coord_flip() 
```

```{r rgs-by-effect-size, eval = FALSE, fig.cap = "Genetic correlations with EA3 by effect size on n siblings."}
rgs <- left_join(rgs, res %>% filter(reg.type == "residualized"), by = c("p2" = "term"))

rgs %>% 
      filter(p2 != "EA3_excl_23andMe_UK", p2 != "EA2_noUKB") %>% 
      ggplot(aes(rg, estimate)) + 
        geom_smooth(method = lm) +
        geom_point(colour = "red2") + 
        geom_label(aes(label = p2, hjust = "left", vjust = "middle"), 
          size = 3, nudge_x = 0.02, alpha = 0.5, 
          label.padding = unit(0.1, "lines")) +
        labs(x = "rG", y = "Effect size", 
          title = "rG with EA3 vs. effect size on # siblings")
```

```{r correlate-effect-sizes, eval = FALSE}
combined_results <- res %>% 
      mutate(PCs = ifelse(reg.type=="raw", "no", "yes")) %>% 
      dplyr::select(-reg.type) %>% 
      {bind_rows(bivariate = ., multivariate = all_res, .id = "reg.type")} %>% 
      arrange(term, reg.type, PCs) %>% 
      dplyr::select(term, reg.type, PCs, estimate) %>% 
      pivot_wider(names_from = c(reg.type, PCs), values_from = estimate)

cor(combined_results[-1], use = "complete")

```

### 2020-05-14 why-selection-against-EA.Rmd
Initial commit

--- 2020-05-14 why-selection-against-EA-parents.Rmd
@@
 
+options(digits = 2)
 knitr::opts_chunk$set(
@@
 order_abs <- function (x) x[1] # for use when reordering factors
+pretty_names <- function (name) {
+  name <- gsub("_", " ", name)
+  name <- strsplit(name, " ")
+  map_chr(name, 
+        ~paste0(
+          toupper(substring(.x, 1, 1)), 
+          substring(.x, 2),
+          collapse = " "
+        ))
+}
 
@@
 
+reverse_code <- c("agreeableness", "age_at_menarche", "age_at_menopauze",
+  "cognitive_ability", "conscientiousness", "EA2_noUKB",
+  "EA3_excl_23andMe_UK", "extraversion", "height_combined",
+  "openness")
+famhist[reverse_code] <- famhist[reverse_code] * -1
+
 rgs$p2 <- rgs$p2 %>% recode(
@@
 
+# Data
+
+Data is taken from UK Biobank. Polygenic scores were  normalized to mean 0,
+variance 1. Scores for educational attainment, cognitive ability, height,
+extraversion, agreeableness, conscientiousness, age at menarche, and age at
+menopause were reverse coded. High scores are therefore "negative" in the loose
+sense that their target phenotypes would be considered undesirable by many
+people.
+
 # Results
@@
 
-```{r run-regressions}
+```{r run-regs}
 # for each score, calculate the correlation with number of siblings
@@
 
-```{r}
-run_regs_time <- function(children, score_name, period) {
-  dep_var  <- if (children) "log(n_children + 1)" else "log(n_sibs + 2)"
+```{r run-regs-period}
+run_regs_period <- function(children, score_name, period) {
+  dep_var  <- if (children) "log(n_children)" else "log(n_sibs + 1)"
   year_var <- if (children) "year_flb" else "YOB"
-  famhist$in_period <- ! is.na(famhist[[year_var]]) & famhist[[year_var]] %in% period:(period+4)
+  famhist$in_period <- ! is.na(famhist[[year_var]]) & 
+                       famhist[[year_var]] %in% period:(period+9)
 
@@
         score_name = score_names, 
-        period = seq(1940, 2000, 5)
+        period = seq(1940, 2000, 10)
       )
-res_yr <- pmap_dfr(pars, run_regs_time)
+res_yr <- pmap_dfr(pars, run_regs_period)
 
@@
 
-```{r over-time}
-
-ggplot(res_yr, aes(period, estimate, colour = children)) + 
-      geom_linerange(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2) +
-      geom_line() + 
-      facet_wrap(vars(term))
+```{r plot-by-period, fig.cap = "Effect sizes of PGS on number of children/siblings by ten year periods", fig.height = 7, fig.width = 6}
+res_yr %>% 
+      mutate(
+        dep.var = ifelse(children, "Children", "Siblings")
+      ) %>% 
+ggplot(aes(period, estimate, group = dep.var)) + 
+      geom_hline(yintercept = 0, colour = "grey50", size = 0.5) +
+      geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.3, fill = "lightblue") +
+      geom_line(aes(linetype = dep.var), colour = "navy") + 
+      facet_wrap(
+        vars(term), 
+        scales = "free_y", 
+        ncol = 4, 
+        strip.position = "bottom"
+      ) +
+      theme(
+        panel.grid.minor.y = element_blank(), 
+        legend.title = element_blank()
+      )
 ```
 
-Selection over time appears to be decreasing, if we focus on raw PGS. Figure
-\@ref(fig:effect-sizes-by-period) shows effect sizes for  survey respondents
-born in the 1930s and 1940s, versus respondents born in the 1950s to 1970s.
+Selection over time appears to be decreasing. Figure \@ref(fig:plot-by-period) shows
+effect sizes estimated within 10 year periods. Sibling regressions use the
+respondent's own birth year. Children regressions use the year of the
+respondent's first live birth. Since this is not available for men, only female
+respondents are included. By definition, the sibling regressions exclude members
+of the parents' generation who had no children. To make the effect size
+estimates as comparable as possible, children regressions include only
+respondents with at least one child ever born; the dependent variable is
+*log(number of children)* or *log(total number of siblings)*.
 
-```{r effect-sizes-by-period, fig.cap = "Effect sizes of PGS on n children by period. Significance tests compare early to late effect sizes", fig.width = 6, fig.height = 7}
+Despite this, effect sizes tend to be smaller for children regressions. This
+could be caused by ascertainment bias in the UK Biobank sample -- e.g.,
+if respondents themselves are a more selected sample than the respondents' 
+parents.
 
-famhist$period <- chop(famhist$YOB, c(1930, 1950, 1974), 
-      labels = c("30s_40s", "50s_70s"))
+## Subgroups
 
-run_time_regs <- function(score_name) {
+We next examine the correlates of natural selection by running regressions on
+*log(number of children + 1)* within various subgroups of the sample.
 
-  fml <- as.formula(glue("log(n_sibs+1) ~ {score_name}:period"))
-  mod <- lm(fml, famhist)
-  res <- tidy(mod)
-  res <- filter(res, ! term %in% c("(Intercept)"))
-  res$term <- gsub(".*:period", "", res$term)
+```{r run-regs-subset}
 
-  glhts <- list(
-    glht(mod, glue("{score_name}:period30s_40s - {score_name}:period50s_70s = 0"))
-  )
-  glhts <- glhts %>% map(summary) %>% map(tidy) %>% map_dbl("p.value")
-  res$glht.p.value <- c(NA, glhts)
-
+run_regs_subset <- function(score_name, subset) {
+  fml <- as.formula(glue("log(n_children + 1) ~ {score_name}"))
+  mod <- lm(fml, famhist, subset = eval(subset))
+  res <- tidy(mod, conf.int = TRUE)
+  res <- filter(res, term == {{score_name}})
+  res$subset <- list(subset)
   res
@@
 
-names(score_names) <- score_names
-time_regs <- map_dfr(score_names, run_time_regs, .id = "reg")
-
-time_regs %<>% 
-      rename(period = term) %>% 
-      mutate(reg = fct_reorder(reg, estimate, first))
-
-time_regs_wide <- time_regs %>% 
-      select(reg, period, estimate, glht.p.value) %>% 
-      pivot_wider(reg, names_from = period, 
-        values_from = c(estimate, glht.p.value)) %>% 
-      mutate(Significant = glht.p.value_50s_70s < 0.05/(nrow(time_regs) * 3/4))
-
-time_regs %>% 
-      group_by(reg) %>% 
-      mutate(
-        Significant = na.omit(glht.p.value) < 0.05/(nrow(time_regs) * 3/4)
-      ) %>% 
-      ggplot(aes(x = reg)) +
-        geom_hline(yintercept = 0, colour = "darkgrey") +
-        geom_linerange(aes(
-          ymin = estimate_30s_40s, 
-          ymax = estimate_50s_70s, colour = Significant), 
-          data = time_regs_wide) + 
-        geom_point(aes(y = estimate, fill = period, colour = Significant), 
-          shape = 21, size = 2) + 
-        scale_colour_manual(values = c("grey", "orange3")) +
-        scale_fill_brewer(type = "qual", palette = 3) +
-        coord_flip()
+subsets <- rlang::exprs(sex == 0, sex == 1)
+par <- expand_grid(score_name = score_names, subset = subsets)
+res_subsets <- pmap_dfr(par, run_regs_subset)
 ```
 
-# Causality
-
-To be selected against, an allele does not need to cause fewer children, only to
-be correlated with fewer children. Nevertheless, it is interesting to ask
-whether individual PGS caused sample members' parents to have fewer children. As
-a first cut, we regress log(number of siblings _+ 1) on all PGS together. We run
-two regressions: one controlling for genetic principal components, and one
-without these controls. Figure \@ref(fig:multivariate-reg-plots) shows the results.
-
-```{r multivariate-regs}
-
-# EA3 and EA2 have a 0.9 raw correlation. These are the only pair of PGS with
-# a raw correlation of more than 0.8
-reg_score_names <- setdiff(score_names, "EA2_noUKB")
-
-without_pcs <- paste0("log(n_sibs+1) ~ ", paste(reg_score_names, collapse = " + "))
-fml <- as.formula(without_pcs)
-fml_with_pcs <- as.formula(paste(without_pcs, "+", paste0("PC", 1:40, 
-      collapse = " + ")))
-
-all_mod <- lm(fml, famhist)
-all_res <- tidy(all_mod, conf.int = TRUE)
-
-all_mod_pcs <- lm(fml_with_pcs, famhist)
-all_res_pcs <- tidy(all_mod_pcs, conf.int = TRUE)
-
-all_res <- bind_rows(no = all_res, yes = all_res_pcs, .id = "PCs")
-all_res <- filter(all_res, term != "(Intercept)", ! grepl("^PC", term))
-
+```{r fn-plot-subsets}
+plot_subsets <- function(ss_name, ss_expr, ...) {
+  ss_name <- rlang::enquo(ss_name)
+  subset_expr <- rlang::enexprs(...)
+  res_subsets %>% 
+        filter(subset %in% !!subset_expr) %>% 
+        mutate(
+          subset = as.character(subset), 
+          !!ss_name := {{ss_expr}},
+          term = fct_reorder(term, estimate, order_abs)
+        ) %>% 
+        ungroup() %>% 
+        ggplot(aes(term, estimate, group = !!ss_name, colour = !!ss_name)) +
+          geom_hline(yintercept = 0, colour = "grey20") +
+          geom_pointrange(aes(ymin = conf.low, ymax = conf.high), alpha = 0.7) +
+          coord_flip() +
+          scale_colour_brewer(palette = 2, type = "qual")
+}
 ```
 
-```{r multivariate-reg-plots, fig.cap = "Multivariate regressions. Significance tests compare effect sizes to zero."}
+Figure \@ref(fig:plot-sexes) shows effect sizes of PGS on number of children
+separately for males and females. Education and height PGS are strongly
+negatively selected among women.
 
-all_res$term <- fct_reorder(all_res$term, all_res$estimate, order_abs)
+```{r plot-sexes, fig.cap = "Effect sizes on number of children by sex"}
+plot_subsets(
+        Sex, 
+        recode(subset, "sex == 0" = "Female", "sex == 1" = "Male"),
+        sex == 0, sex == 1
+      )
+```
 
-n_regs <- as.double(nrow(all_res)) # avoids an ugly "L"
-ggplot(all_res, aes(term)) + 
-      geom_point(aes(y = estimate, fill = PCs, 
-        colour = p.value <= 0.05/{{n_regs}}), shape = 21) + 
-      scale_colour_manual(values = c("grey", "black")) +
-      geom_hline(yintercept = 0, colour = "darkgrey") +
-      labs(x = "PGS", y = "Effect size") +
-      coord_flip() 
+```{r plot-sd-years, eval = FALSE}
+# inequality over time. Nothing to see, almost everything is flat.
+grid <- expand_grid(score_name = score_names, YOB = 1940:1965)  
+grid$sd <- pmap_dbl(grid, ~sd(famhist[famhist$YOB==.y, ][[.x]], na.rm = TRUE))
+ggplot(grid, aes(YOB, sd)) + 
+      geom_line() + 
+      facet_wrap(~score_name, scales = "free_y", ncol = 4)
 ```

### 2020-05-14 _drake.R

@@ -1,0 +1,191 @@
+
+suppressPackageStartupMessages({
+  library(drake)
+  library(rlang)
+  library(purrr)
+  library(tidyr)
+})
+
+source("make-data.R")
+source("regressions.R")
+
+pgs_dir       <- "~/UKBB data 2019/polygenic_scores/"
+ph_file       <- "~/UKBB data 2019/UKB.EA_pheno.coordinates.QC.david.csv"
+pcs_file      <- "~/UKBB data 2019/ukb30545.40PCs.csv"
+famhist_file  <- "~/Dropbox/assortative mating/biobank-analysis/david.family_history.traits.out.csv"
+famhist2_file <- "~/UKBB data 2019/david.family_history.traits.20042020.out.csv"
+famhist3_file <- "~/UKBB data 2019/david.family_history.traits.05052020.out.csv"
+rgs_file      <- "EA3_rgs.10052019.rgs.csv"
+mf_pairs_file <- "~/UKBB data 2019/spouse_pair_info/UKB_out.mf_pairs_rebadged.csv"
+
+
+plan <- drake_plan(
+  report       = rmarkdown::render(
+                   input       = knitr_in("why-negative-selection.Rmd"), 
+                   output_file = file_out("why-negative-selection.pdf"), 
+                   quiet       = TRUE
+                 ),
+  
+  
+  famhist      = edit_famhist(famhist_raw, reverse_code),
+  
+  
+  famhist_raw  = target(
+                   make_famhist( 
+                      file_in(!! ph_file), 
+                      file_in(!! pcs_file), 
+                      file_in(!! famhist_file),
+                      file_in(!! famhist2_file),
+                      file_in(!! famhist3_file),
+                      file_in(!! pgs_dir)
+                    ), 
+                    format = "fst"
+                  ),
+  
+  
+  mf_pairs     = target(
+                   make_mf_pairs(file_in(!! mf_pairs_file), famhist), 
+                   format = "fst"
+                 ), 
+  
+  
+  rgs          = make_rgs(file_in(!! rgs_file)),
+   
+  
+  score_names  = {
+                   sub(
+                           ".*UKB\\.AMC\\.(.*?)\\..*", 
+                           "\\1", 
+                           list.files(file_in(!! pgs_dir), pattern = "csv$"), 
+                           perl = TRUE
+                         )
+                 },
+  
+  
+  reverse_code =  {
+                    setdiff(score_names, c("agreeableness", "age_at_menarche",
[…]

### 2020-05-14 regressions.R

@@ -1,0 +1,66 @@
+
+suppressPackageStartupMessages({
+  library(glue)
+  library(broom)
+  library(dplyr)
+  library(lmtest)
+})
+
+
+run_regs <- function (dep_var, score_name, famhist) {
+  fml <- as.formula(glue("{dep_var} ~ {score_name}"))
+  reg_bv <- tidy(lm(fml, famhist), conf.int = TRUE) 
+  
+  pcs <- paste0("PC", 1:40, collapse = " + ")
+  fml_resid <- as.formula(glue("{dep_var} ~ {score_name} + {pcs}"))
+  reg_resid <- tidy(lm(fml_resid, famhist), conf.int = TRUE)
+  
+  regs <- bind_rows(raw = reg_bv, controlled = reg_resid, .id = "reg.type")
+  regs %<>% filter(term != "(Intercept)", ! grepl("^PC", term))
+  
+  return(regs)
+}
+
+
+run_regs_period <- function (children, score_name, famhist) {
+  dep_var  <- if (children) "n_children" else "n_sibs"
+  year_var <- if (children) "YOB" else "parents_imp_YOB"
+  famhist$year_split <- famhist[[year_var]] >= median(famhist[[year_var]], 
+        na.rm = TRUE)
+  famhist$year_split <- factor(famhist$year_split, labels = c("early", "late"))
+  fml <- as.formula(glue("{dep_var} ~ {score_name}:year_split"))
+  mod <- lm(fml, famhist)
+  res <- tidy(mod, conf.int = TRUE)
+  res <- filter(res, grepl(score_name, term))
+  
+  test_spec <- glue("{score_name}:year_splitearly - {score_name}:year_splitlate = 0")
+  test <- car::lht(mod, test_spec)
+  pval <- test[["Pr(>F)"]]
+  res$diff.p.value <- pval[2]
+  
+  res$children <- children
+  
+  res
+}
+
+
+run_regs_subset <- function(score_name, subset, famhist) {
+  fml <- as.formula(glue::glue("n_children ~ {score_name}"))
+  mod <- lm(fml, famhist, subset = eval(subset))
+  res <- tidy(mod, conf.int = TRUE)
+  res <- filter(res, term == {{score_name}})
+  res$subset <- list(subset)
+
+  res
+}
+
+
+run_regs_fml <- function(fml, ..., subset = NULL, famhist) {
+  glue_args <- list(...)
[…]

### 2020-05-14 why-negative-selection.Rmd
Add regressions on PCs

@@
 Data is taken from UK Biobank. Polygenic scores were  normalized to mean 0,
-variance 1. Scores for ADHD, substance use, alcohol, Alzheimer's disease,
-autism, bipolarity, BMI, body fat, caffeine, cannabis use, coronary artery
-disease, eating disorders, waist-hip ratio, major depressive disorder,
-neuroticism, and schizophrenia were reverse coded. High scores are therefore
-"positive" in the loose sense that their target phenotypes would be considered
-desirable by many people.
+variance 1.
 
@@
 
-* *total siblings*, the number of full siblings of the respondent (including
+* *siblings*, the number of full siblings in the respondent's sib (including
   himself or herself).
@@
 
+prop_gens_raw <- res_all_wide[["raw_N children"]]/res_all_wide[["raw_N siblings"]]
+prop_gens_controlled <- res_all_wide[["controlled_N children"]]/res_all_wide[["controlled_N siblings"]]
+median_prop_gens_raw <- median(prop_gens_raw[prop_gens_raw > 0])
+median_prop_gens_controlled <- median(prop_gens_controlled[prop_gens_controlled > 0])
 ```
@@
 Estimates are broadly consistent across generations. For 
-`r sum(res_all_wide$consistent)` out of `r n_pgs` polygenic scores, all 4 estimates have the same sign.
+`r sum(res_all_wide$consistent)` out of `r n_pgs` polygenic scores, all 4 
+estimates have the same sign.
 
-Within siblings regressions, effect sizes are smaller when controlling for
+However, effect sizes are much smaller for *children* than *siblings*
+regressions. Among consistently-signed estimates, the median effect size for
+children as a proportion of the effect size for  siblings is 
+`r median_prop_gens_raw`, or `r median_prop_gens_controlled` with controls.
+
+In siblings regressions, effect sizes are smaller when controlling for
 principal components -- sometimes much smaller, as in the case of height. 
@@
 
-```{r plot-basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children. Scores marked * are reverse coded.", fig.height = 7}
+```{r plot-basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children. Scores are plotted on different scales.", fig.height = 7}
 
@@
       labs(y = "Effect size", x = "") +
-      facet_wrap(vars(dep.var), ncol = 2, scales = "free") +
+      facet_wrap(vars(dep.var), ncol = 2, scales = "free_x") +
       coord_flip() +
@@
       )
+
+```
+
+To get a further insight into this we regress *siblings* and *children* on
+individual principal components themselves. As Figure \@ref(fig:plot-pc-regs) shows,
+there are many more large and significant effects  in siblings regressions.
+
+```{r plot-pc-regs, fig.cap = "Effect of principal components on number of siblings/children. Each dot represents one principal component. Points are jittered on the Y axis.", fig.height = 2}
+
+drake::loadd(res_pcs)
+n_regs <- as.double(nrow(res_pcs))
+
+res_pcs %>% 
+      mutate(
+        estimate = abs(estimate),
+        dep.var  = fct_relevel(dep.var, "N children")
+      ) %>% 
+      ggplot(aes(estimate, dep.var, colour = p.value < 0.05/{{n_regs}})) + 
+        geom_jitter(height = 0.15, width = 0) + 
+        scale_x_log10(labels = scales::label_number(accuracy = 0.0001)) +
+        labs(x = "Effect size (log scale)", y = "") + 
+        scale_colour_manual(values = pval_palette)
 

### 2020-05-14 _drake.R

@@ -20,11 +20,6 @@ mf_pairs_file <- "~/UKBB data 2019/spouse_pair_info/UKB_out.mf_pairs_rebadged.cs
 
 
 plan <- drake_plan(
-  report       = rmarkdown::render(
-                   input       = knitr_in("why-negative-selection.Rmd"), 
-                   output_file = file_out("why-negative-selection.pdf"), 
-                   quiet       = TRUE
-                 ),
   
   
   famhist      = edit_famhist(famhist_raw, reverse_code),
@@ -53,63 +48,83 @@ plan <- drake_plan(
    
   
   score_names  = {
-                   sub(
-                           ".*UKB\\.AMC\\.(.*?)\\..*", 
-                           "\\1", 
-                           list.files(file_in(!! pgs_dir), pattern = "csv$"), 
-                           perl = TRUE
-                         )
-                 },
+    sub(
+            ".*UKB\\.AMC\\.(.*?)\\..*", 
+            "\\1", 
+            list.files(file_in(!! pgs_dir), pattern = "csv$"), 
+            perl = TRUE
+          )
+  },
   
   
-  reverse_code =  {
-                    setdiff(score_names, c("agreeableness", "age_at_menarche",
-                      "age_at_menopauze", "cognitive_ability", 
-                      "conscientiousness", "EA2_noUKB", "EA3_excl_23andMe_UK", 
-                      "extraversion", "height_combined", "openness"))
-                  },
+  reverse_code = c(), # don't reverse code anything
+  # {
+  #   setdiff(score_names, c("agreeableness", "age_at_menarche",
+  #           "age_at_menopauze", "cognitive_ability", 
+  #           "conscientiousness", "EA2_noUKB", "EA3_excl_23andMe_UK", 
+  #           "extraversion", "height_combined", "openness")
+  #         )
+  # },
   
   
   res_all      = {
-                    res_sibs <- map_dfr(score_names, run_regs, 
-                      dep_var = "n_sibs", 
-                      famhist = famhist
-                    )
-                    res_chn <- map_dfr(score_names, run_regs, 
-                      dep_var = "n_children", 
-                      famhist = famhist
-                    )
-                    dplyr::bind_rows(
-                      "N siblings" = res_sibs, 
-                      "N children" = res_chn, 
-                      .id = "dep.var"
-                    )
-                  },
[…]

### 2020-05-14 regressions.R

@@ -22,6 +22,15 @@ run_regs <- function (dep_var, score_name, famhist) {
 }
 
 
+run_regs_pcs <- function (dep_var, pc_name, famhist) {
+  fml <- as.formula(glue("{dep_var} ~ {pc_name}"))
+  regs <- tidy(lm(fml, famhist), conf.int = TRUE) 
+  regs %<>% filter(term != "(Intercept)")
+  
+  return(regs)
+}
+
+
 run_regs_period <- function (children, score_name, famhist) {
   dep_var  <- if (children) "n_children" else "n_sibs"
   year_var <- if (children) "YOB" else "parents_imp_YOB"

### 2020-05-14 why-negative-selection.Rmd
Add income regressions

@@
 
-```{r plot-pc-regs, fig.cap = "Effect of principal components on number of siblings/children. Each dot represents one principal component. Points are jittered on the Y axis.", fig.height = 2}
+```{r plot-pc-regs, fig.cap = "Effect of principal components on number of siblings/children.
+Absolute effect sizes are plotted.
+Each dot represents one bivariate regression.
+Points are jittered on the Y axis.", fig.height = 2}
 
@@
 
-## Education levels
+## Education and income
 
@@
 
-```{r plot-educ-levels, fig.cap = "Effect sizes on number of children by age left full-time education", fig.height = 7}
+```{r plot-educ-levels, fig.cap = "Effect sizes on number of children by age left full-time education"}
 
-loadd(res_edu)
+drake::loadd(res_edu)
 res_edu %>% 
@@
 ```
+
+Figure \@ref(fig:plot-income) splits respondents by household income category.
+A very similar pattern holds, with selection effects being larger for those
+in the poorest income category.
+
+```{r plot-income, fig.cap = "Effect sizes on number of children by household income"}
+
+drake::loadd(res_income)
+
+n_regs <- as.double(nrow(res_income))
+res_income %>% 
+      mutate(
+        term = pretty_names(term), 
+        term = fct_reorder(term, estimate, order_abs),
+        Income = factor(income_cat, 
+          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
+        )
+      ) %>% 
+      ggplot(aes(term, estimate, fill = Income, colour = p.value < 0.05/{{n_regs}})) +
+        my_hline +
+        geom_point(alpha = 0.8, shape = 21, size = 1.5, stroke = 0.9) + 
+        scale_fill_brewer(direction = -1) +
+        scale_colour_manual(values = pval_palette) +
+        coord_flip()
+```
+
+```{r income-edu-controlled}
+drake::loadd(res_income_controlled, res_edu_controlled)
+n_regs_ic <- nrow(res_income_controlled)
+n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
+
+n_regs_educ <- nrow(res_edu_controlled)
+n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
+```
+
+These results could be driven by age, if older respondents are poorer and less
+educated, and also more subject to selection on polygenic scores. However, if we
+rerun the regressions, interacting the polygenic score with income category and
+also with a quadratic in age, the interaction with income remains significant at
+0.05/`r n_regs_ic` for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
+if we interact the PGS with age of leaving full time education and a quadratic
+in age, the interaction with age leaving FTE remains significant at 
+0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.
 

### 2020-05-14 _drake.R

@@ -149,9 +149,6 @@ plan <- drake_plan(
   
   
   res_edu = {
-    famhist$age_fte_cat <- santoku::chop(famhist$age_fulltime_edu, 
-          c(16, 18), 
-          c("< 16", "16-18", "> 18"))
     subsets <- rlang::exprs(
             age_fte_cat == "< 16",
             age_fte_cat == "16-18",
@@ -174,6 +171,58 @@ plan <- drake_plan(
             age_fte_cat = sub("age_fte_cat == \"(.*)\"", "\\1", subset)
           ) %>% 
           select(-row_number, -subset)
+  },
+  
+  
+  res_income = {
+    subsets = rlang::exprs(
+            income_cat == 1,
+            income_cat == 2,
+            income_cat == 3,
+            income_cat == 4,
+            income_cat == 5
+          ) 
+    pars <- expand_grid(score_name = score_names, subset = subsets) 
+    res_income <- pmap_dfr(pars,
+            ~ run_regs_fml(
+              "n_children ~ {score_name}", 
+              score_name = .x, 
+              subset     = .y,
+              famhist    = famhist
+            ),
+            .id = "row_number"
+          ) 
+    pars$row_number <- as.character(seq_len(nrow(pars)))
+    res_income %>% left_join(pars, by = "row_number") %>% 
+          filter(term != "(Intercept)") %>% 
+          mutate(
+            income_cat = sub("income_cat == (.*)", "\\1", subset)
+          ) %>% 
+          select(-row_number, -subset)
+  },
+  
+  
+  res_income_controlled = {
+    res <- map_dfr(score_names, 
+            ~run_regs_fml(
+              "n_children ~ {score_name}*(income_cat + YOB +I(YOB^2))", 
+              score_name = .x, famhist = famhist
+            ),
+            .id = "score_name"
+          )
+    res %>% filter(grepl(":income_cat", term))
+  },
+  
+  
+  res_edu_controlled = {
+    res <- map_dfr(score_names, 
+      ~run_regs_fml(
+        "n_children ~ {score_name}*(as.numeric(age_fte_cat) + YOB + I(YOB^2))", 
+        score_name = .x, famhist = famhist
+      ),
+      .id = "score_name"
+    )
+    res %>% filter(grepl(":.*age_fte_cat", term))
   },
   
   

### 2020-05-14 make-data.R

@@ -4,6 +4,7 @@ suppressPackageStartupMessages({
   library(magrittr)
   library(dplyr)
   library(rlang)
+  library(santoku)
 })
 
 
@@ -106,11 +107,16 @@ edit_famhist <- function (famhist, reverse_code) {
   famhist$n_children[famhist$n_children < 0] <- NA
   
   famhist$age_fulltime_edu[famhist$age_fulltime_edu < 0] <- NA
-  
+  famhist$age_fte_cat <- santoku::chop(famhist$age_fulltime_edu, 
+    c(16, 18), 
+    c("< 16", "16-18", "> 18"))
   child_vars <- rlang::quos(f.2754.0.0, f.2754.1.0, f.2754.2.0)
   famhist %<>% mutate_at(vars(!!!child_vars), ~ ifelse(.x < 0, NA, .x))
   famhist$age_flb <- rowMeans(famhist %>% select(!!!child_vars), na.rm = TRUE)
   famhist$year_flb <- famhist$YOB + famhist$age_flb
+  
+  famhist$income_cat <- famhist$f.738.0.0
+  famhist$income_cat[famhist$income_cat < 0] <- NA
   
   famhist[reverse_code] <- famhist[reverse_code] * -1
   

### 2020-05-14 why-negative-selection.Rmd
weight data using GHS 2006.

@@
 ```{r plot-basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children. Scores are plotted on different scales.", fig.height = 7}
+
+drake::loadd(res_all)
 
@@
 
-Despite this, effect sizes tend to be smaller for children regressions. This
+```{r res-weighted}
+
+drake::loadd(res_weighted)
+res_weighted_all <- res_weighted %>% 
+      left_join(
+        res_all %>% 
+              filter(dep.var == "N children", reg.type == "raw") %>% 
+              select(term, estimate),
+        by = "term",
+        suffix = c("_weighted", "_unweighted")
+      ) %>% 
+      mutate(
+        prop_wtd_unwtd = estimate_weighted/estimate_unweighted
+      )
+
+mean_prop_wtd_unwtd <- res_weighted_all %>% 
+      filter(sign(estimate_weighted) == sign(estimate_unweighted)) %>% 
+      summarize(mean = mean(prop_wtd_unwtd)) %>% 
+      pull(mean)
+
+```
+
+# Accounting for ascertainment bias
+
+Effect sizes tend to be smaller for children regressions. This
 could be caused by ascertainment bias in the UK Biobank sample -- e.g.,
 if respondents themselves are a more selected sample than the respondents' 
-parents.
+parents. To check this, we weighted UK Biobank participants by age of leaving
+full time education. Our sample weights are based on the 2006 General Household
+Survey, calculating proportions within age and sex cells among White British
+respondents. We then rerun the basic *children* regressions.
+
+Figure \@ref(fig:plot-weighted) shows the results, with unweighted estimates
+plotted for comparison. Weighting increases effect sizes on average by 
+`r 100 * (mean_prop_wtd_unwtd - 1)` per cent. These weights are only basic 
+attempts to correct for ascertainment bias. Effect sizes might be increased 
+further by more precise weighting.
+
+```{r plot-weighted, fig.cap = "Effects of polygenic scores on number of children, regressions weighted by education levels within age categories"}
+
+n_regs <- as.double(nrow(res_weighted_all))
+
+res_weighted_all %>% 
+      mutate(
+        term    = pretty_names(term),
+        term    = fct_reorder(term, estimate_weighted, order_abs)
+      ) %>% 
+      pivot_longer(cols = c(estimate_weighted, estimate_unweighted), 
+        names_to = "Regression", values_to = "estimate") %>% 
+      mutate(
+        Regression = sub("estimate_", "", Regression)
+      ) %>% 
+      ggplot(aes(x = term, y = estimate, fill = Regression,
+        colour = p.value < 0.05/{{n_regs}})) +
+      my_hline + 
+      geom_point(size = 1.5, stroke = .8, shape = 21) + 
+      labs(y = "Effect size", x = "") +
+      coord_flip() +
+      scale_fill_manual(values = 
+        c("unweighted" = "transparent", "weighted" = "darkgreen")) +
+      scale_colour_manual(values = pval_palette)
+
+```
 
@@
 omnibus regression of *number of children* on all the PGS. We exclude EA2, 
-waist-hip ratio, WC (XXX what is it?) and "Hip combined" since they are 
+waist-hip ratio, waist-circumference, and "Hip combined" since they are 
 highly correlated with other scores, which could make our estimates unstable. Figure \@ref(fig:plot-together) shows the results. Interestingly, several PGS remain independently significant, although

### 2020-05-14 _drake.R

@@ -6,8 +6,10 @@ suppressPackageStartupMessages({
   library(tidyr)
 })
 
+
 source("make-data.R")
 source("regressions.R")
+
 
 pgs_dir       <- "~/UKBB data 2019/polygenic_scores/"
 ph_file       <- "~/UKBB data 2019/UKB.EA_pheno.coordinates.QC.david.csv"
@@ -17,13 +19,11 @@ famhist2_file <- "~/UKBB data 2019/david.family_history.traits.20042020.out.csv"
 famhist3_file <- "~/UKBB data 2019/david.family_history.traits.05052020.out.csv"
 rgs_file      <- "EA3_rgs.10052019.rgs.csv"
 mf_pairs_file <- "~/UKBB data 2019/spouse_pair_info/UKB_out.mf_pairs_rebadged.csv"
-
+nomis_file    <- "nomis-2011-statistics.csv"
+ghs_file      <- "UKDA-5804-stata8/stata8/Ghs06client.dta"
 
 plan <- drake_plan(
-  
-  
-  famhist      = edit_famhist(famhist_raw, reverse_code),
-  
+
   
   famhist_raw  = target(
                    make_famhist( 
@@ -36,6 +36,15 @@ plan <- drake_plan(
                     ), 
                     format = "fst"
                   ),
+  
+  
+  famhist      = target(
+                  edit_famhist(famhist_raw, reverse_code), 
+                  format = "fst"
+                 ),
+
+    
+  ghs_weights  = weight_by_ghs(file_in(!! ghs_file), famhist),
   
   
   mf_pairs     = target(
@@ -100,6 +109,13 @@ plan <- drake_plan(
       .id = "dep.var"
     ) 
   },
+  
+  
+  res_weighted =  map_dfr(score_names, 
+                    run_regs_weighted, 
+                    famhist = famhist, 
+                    weights = ghs_weights
+                  ),
   
   
   res_period   = {

### 2020-05-14 make-data.R

@@ -148,3 +148,79 @@ make_mf_pairs <- function (mf_pairs_file, famhist) {
   
   mf_pairs
 }
+
+
+weight_by_ghs <- function (ghs_file, famhist) {
+  library(haven)
+  
+  ghs <- haven::read_dta(ghs_file)
+  # variables of interest: 
+  # sex (1 male)
+  # age
+  # ethnic = 1 = white nbritish
+  # EdAge - age left FTE
+  # hiqual - highest qualification 
+  # edlev00 - education level
+  # edlev10 - ditto, fewer categories
+  # weight06 - "weight you should use to weight the data"
+  # 
+  # grfam1h - gross weekly income of family (harmonized)
+  # grhhold - gross weekly household income
+  # for comparison, the UKBB question was "what is the average total
+  # income before tax received by your household" (p.a.)
+  # 
+  # ten1 - tenure
+  # llord - landlord
+  
+  ghs <- ghs %>% 
+    zap_labels() %>% 
+    filter(ethnic == 1, sex %in% 1:2, edage > 0, edage < age) %>% 
+    select(sex, age, edage, weight06, ten1, llord) 
+  
+  ghs_props <- ghs %>% 
+    group_by(sex, age, edage) %>% 
+    summarize(
+      weighted_n = sum(weight06)
+    )
+  ghs_prop_totals <- ghs %>% 
+    group_by(sex, age) %>% 
+    summarize(
+      weighted_n_sex_age = sum(weight06)
+    )
+  ghs_props <- ghs_props %>% 
+    left_join(ghs_prop_totals, by = c("sex", "age")) %>% 
+    mutate(
+      prop_age_fte = weighted_n/weighted_n_sex_age,
+      sex = recode(sex, "1" = "male", "2" = "female")
+    ) %>% 
+    select(-weighted_n, -weighted_n_sex_age)
+  
+  ukbb_props <- famhist %>% 
+    group_by(sex, age_at_reqruitment, age_fulltime_edu) %>% 
+    summarize(n = n())
+  ukbb_prop_totals <- famhist %>% 
+    group_by(sex, age_at_reqruitment) %>% 
+    summarize(n_sex_age = n())
+  ukbb_props <- ukbb_props %>% 
+    left_join(ukbb_prop_totals, by = c("sex", "age_at_reqruitment")) %>% 
+    mutate(
+      prop_age_fte = n/n_sex_age,
[…]

### 2020-05-14 why-negative-selection.Rmd
trivia

@@
     latex_engine: xelatex
-    keep_tex: true
 ---

### 2020-05-15 why-negative-selection.Rmd
Use 100 PCs from Abdel. Split out some data files in the drake plan.

@@
 # TODO
-* Improve imputation of parents' YOB.
-* Selection bias. 
-  - You could weight subjects by education level; sex; birth region?
+
+* Use Abdel's 100 PCs, not UKBB's 40. DONE
+
+* Look at geography, esp. in siblings regressions.
+
+* Divide siblings regressions by own YOB, into early and late.
+  - Use this to confirm the basic "selection decreasing" story
+
+* Plot average scores over time, as an intro graph
+  - Can we work out how to plot selection effects on the "same scale"?
+  - Let X_i be i's polygenic score.
+  - If E(X_parents) = 1/n \sum X_i in the parent's generation, then
+    E(X_kids) = 1/n \sum X_i K_i, in the children's gernation, where K_i is 
+    the number of kids i has. 
+  - Up to a change of means, this is the same as the sample correlation between
+    X and K. 
+  - Do we calculate correlations per year, then plot? Probably too low N....
+    Maybe calculate the correlation over the whole sample but then use
+    it with each year's X_i profile
+
+* How much does age at first live birth "account for" selection?
+
+* Work more on weighting the data?
+
+* control for age in n_partners regressions
+
+* check for new 100 PCs
 
@@
 drake::loadd(score_names)
-drake::loadd(reverse_code)
 
@@
 pretty_names <- function (names) {
-  reverse <- names %in% reverse_code
   names <- gsub("_", " ", names)
@@
         ))
-  names[reverse] %<>% paste0("*")
 
@@
 
+```{r}
+pgs_over_time <- famhist %>% 
+      group_by(
+        YOB = santoku::chop(YOB, 
+          brk_left(seq(1940, 1970, 5), close_end = FALSE), 
+          seq(40, 65, 5),
+          extend = FALSE
+        ) 
+      ) %>% 
+      filter(! is.na(YOB)) %>% 
+      summarize(across(all_of(score_names), mean, na.rm = TRUE)) %>% 
+      pivot_longer(-YOB, names_to = "score_name", values_to = "score")
+
+```
+
+```{r plot-means-over-time, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Lines are means for 5-year intervals"}
+pgs_over_time %>% 
+      mutate(score_name = pretty_names(score_name),) %>% 
+      ggplot(aes(YOB, score, group = score_name)) + 
+      geom_line() +
+      facet_wrap(~score_name) +
+      theme(strip.text = element_text(size = 7)) +
+      labs(y = "PGS", x = "Birth year (5 year intervals)")
+```
+
 We run regressions on two dependent variables:
@@
 
-We run regressions both with and without controls for the 40 top principal components
-of the genetic data.
+We run regressions both with and without controls for the 100 top principal
+components of the genetic data.
 
@@
 
+```{r}
+drake::loadd(res_pcs)
+n_regs <- as.double(nrow(res_pcs))
+prop_sig <- res_pcs %>% 
+      group_by(dep.var) %>% 
+      summarize(sig = sum(p.value <= 0.05/nrow(res_pcs))) %>% 
+      pull(sig, name = dep.var)
+```
+
 To get a further insight into this we regress *siblings* and *children* on
-individual principal components themselves. As Figure \@ref(fig:plot-pc-regs) shows,
-there are many more large and significant effects  in siblings regressions.
+individual principal components. As Figure \@ref(fig:plot-pc-regs)
+shows, effects are larger and more significant in siblings regressions. 
+`r prop_sig[["N siblings"]]` principal components significantly predicted number of
+siblings, while only `r prop_sig[["N children"]]` significantly predicted number
+of children.
 
-```{r plot-pc-regs, fig.cap = "Effect of principal components on number of siblings/children.
+```{r plot-pc-regs, fig.cap = "Effect of principal components of genetic data on number of siblings/children.
 Absolute effect sizes are plotted.
@@
 Points are jittered on the Y axis.", fig.height = 2}
-
-drake::loadd(res_pcs)
-n_regs <- as.double(nrow(res_pcs))
 

### 2020-05-15 _drake.R

@@ -4,6 +4,7 @@ suppressPackageStartupMessages({
   library(rlang)
   library(purrr)
   library(tidyr)
+  library(readr)
 })
 
 
@@ -13,7 +14,7 @@ source("regressions.R")
 
 pgs_dir       <- "~/UKBB data 2019/polygenic_scores/"
 ph_file       <- "~/UKBB data 2019/UKB.EA_pheno.coordinates.QC.david.csv"
-pcs_file      <- "~/UKBB data 2019/ukb30545.40PCs.csv"
+pcs_file      <- "~/UKBB data 2019/UKB.HM3.100PCs.40310.txt"
 famhist_file  <- "~/Dropbox/assortative mating/biobank-analysis/david.family_history.traits.out.csv"
 famhist2_file <- "~/UKBB data 2019/david.family_history.traits.20042020.out.csv"
 famhist3_file <- "~/UKBB data 2019/david.family_history.traits.05052020.out.csv"
@@ -23,12 +24,21 @@ nomis_file    <- "nomis-2011-statistics.csv"
 ghs_file      <- "UKDA-5804-stata8/stata8/Ghs06client.dta"
 
 plan <- drake_plan(
-
+  
+  
+  score_names  = {
+    sub(
+      ".*UKB\\.AMC\\.(.*?)\\..*", 
+      "\\1", 
+      list.files(file_in(!! pgs_dir), pattern = "csv$"), 
+      perl = TRUE
+    )
+  },
+  
   
   famhist_raw  = target(
                    make_famhist( 
                       file_in(!! ph_file), 
-                      file_in(!! pcs_file), 
                       file_in(!! famhist_file),
                       file_in(!! famhist2_file),
                       file_in(!! famhist3_file),
@@ -38,9 +48,18 @@ plan <- drake_plan(
                   ),
   
   
+  pcs          = read_table2(file_in(!! pcs_file)),
+  
+  
   famhist      = target(
-                  edit_famhist(famhist_raw, reverse_code), 
+                  edit_famhist(famhist_raw, score_names), 
                   format = "fst"
+                 ),
+  
+  
+  resid_scores = target(
+                   make_resid_scores(famhist, pcs, score_names), 
+                   format = "fst"
                  ),
 
     
@@ -54,36 +73,20 @@ plan <- drake_plan(
   
   
   rgs          = make_rgs(file_in(!! rgs_file)),
-   
-  
-  score_names  = {
-    sub(
[…]

### 2020-05-15 make-data.R

@@ -8,21 +8,53 @@ suppressPackageStartupMessages({
 })
 
 
+make_pcs <- function (pcs_file) {
+  pcs <- read_table(pcs_file)
+  pc_names <- grep("PC", names(pcs), value = TRUE)
+  pcs[pc_names] <- scale(pcs[pc_names])
+  pcs
+}
+
+
+join_famhist_pcs <- function (famhist, pcs) {
+  left_join(famhist, pcs, by = c("f.eid" = "IID"))
+}
+
+
+join_famhist_resid_scores <- function (famhist, resid_scores) {
+  cbind(famhist, resid_scores)
+}
+
+
+make_resid_scores <- function (famhist, pcs, score_names) {
+  famhist <- join_famhist_pcs(famhist, pcs)
+  resid_scores <- data.frame(dummy = numeric(nrow(famhist))) 
+  
+  for (score_name in score_names) {
+    resid_fml <- paste(score_name, "~", paste0("PC", 1:40, collapse = " + "))
+    resid_score <- resid(lm(as.formula(resid_fml), famhist, 
+          na.action = na.exclude))
+    resid_scores[[paste0(score_name, "_resid")]] <- resid_score
+  }
+  
+  resid_score$dummy <- NULL
+  resid_scores
+}
+
+
 make_famhist <- function (
         ph_file, 
-        pcs_file, 
         famhist_file,
         famhist2_file,
         famhist3_file,
         pgs_dir
       ) {
-  ph_spec <- spec_csv(ph_file)
   ph <- read_csv(ph_file, col_types = cols(
     geno_measurement_plate = col_skip(),
     geno_measurement_well = col_skip(),
     .default = col_double()
   ))
-  pcs <- read_csv(pcs_file)
+  
   famhist <- read_csv(famhist_file, col_types = strrep("d", 40))
   famhist2 <- read_csv(famhist2_file, col_types = strrep("d", 17))
   famhist3 <- read_csv(famhist3_file)
@@ -34,7 +66,6 @@ make_famhist <- function (
   names(famhist3) <- paste0("f.", names(famhist3))
   names(famhist3) <- gsub("\\-", ".", names(famhist3))
   
-  famhist %<>% left_join(pcs, by = c("f.eid" = "eid"))
   famhist %<>% left_join(ph, by = c("f.eid" = "eid"))
   famhist %<>% left_join(famhist2, by = "f.eid")
   famhist %<>% left_join(famhist3, by = "f.eid")
[…]

### 2020-05-15 regressions.R

@@ -7,27 +7,41 @@ suppressPackageStartupMessages({
 })
 
 
-run_regs <- function (dep_var, score_name, famhist) {
-  fml <- as.formula(glue("{dep_var} ~ {score_name}"))
-  reg_bv <- tidy(lm(fml, famhist), conf.int = TRUE) 
+run_regs <- function (dep_var, score_names, famhist, pcs) {
   
-  pcs <- paste0("PC", 1:40, collapse = " + ")
-  fml_resid <- as.formula(glue("{dep_var} ~ {score_name} + {pcs}"))
-  reg_resid <- tidy(lm(fml_resid, famhist), conf.int = TRUE)
+  run_reg <- function (score_name) {
+    fml <- as.formula(glue("{dep_var} ~ {score_name}"))
+    reg_bv <- tidy(lm(fml, famhist), conf.int = TRUE)  
+  }
+  basic_res <- map_dfr(score_names, run_reg, .id = "score_name")
   
-  regs <- bind_rows(raw = reg_bv, controlled = reg_resid, .id = "reg.type")
-  regs %<>% filter(term != "(Intercept)", ! grepl("^PC", term))
+  famhist <- join_famhist_pcs(famhist, pcs)
+  pc_names <- grep("PC", names(pcs), value = TRUE)
+  run_resid_reg <- function (score_name) {
+    fml_resid <- as.formula(glue("{dep_var} ~ {score_name} + {pc_names}"))
+    reg_resid <- tidy(lm(fml_resid, famhist), conf.int = TRUE)  
+  }
+  resid_res <- map_dfr(score_names, run_resid_reg, .id = "score_name")
   
-  return(regs)
+  res <- bind_rows(raw = basic_res, controlled = resid_res, .id = "reg.type")
+  res %<>% filter(term != "(Intercept)", ! grepl("^PC", term))
+  
+  return(res)
 }
 
 
-run_regs_pcs <- function (dep_var, pc_name, famhist) {
-  fml <- as.formula(glue("{dep_var} ~ {pc_name}"))
-  regs <- tidy(lm(fml, famhist), conf.int = TRUE) 
-  regs %<>% filter(term != "(Intercept)")
+run_regs_pcs <- function (dep_var, famhist, pcs) {
+  pc_names <- grep("PC", names(pcs), value = TRUE)
   
-  return(regs)
[…]

### 2020-05-15 why-negative-selection.Rmd
Fix paths in _drake.R; split time regressions by own YOB always. Put data files in gitignore. Stop storing drake history.

@@
   - Use this to confirm the basic "selection decreasing" story
+  - Actually almost no change.
 
-* Plot average scores over time, as an intro graph
+* Plot average scores over time, as an intro graph. DONE
   - Can we work out how to plot selection effects on the "same scale"?
-  - Let X_i be i's polygenic score.
-  - If E(X_parents) = 1/n \sum X_i in the parent's generation, then
-    E(X_kids) = 1/n \sum X_i K_i, in the children's gernation, where K_i is 
-    the number of kids i has. 
-  - Up to a change of means, this is the same as the sample correlation between
-    X and K. 
-  - Do we calculate correlations per year, then plot? Probably too low N....
-    Maybe calculate the correlation over the whole sample but then use
-    it with each year's X_i profile
+  - Simplest thing would be to use birth dates
 
 * How much does age at first live birth "account for" selection?
+  - Need f.3872 age at flb for women with one child
 
@@
         significant = diff.p.value < 0.05/n_tests,
-        increasing  = estimate_early < estimate_late,
+        sign_early  = sign(estimate_early),
+        sign_late   = sign(estimate_late),
         size_inc    = abs(estimate_early) < abs(estimate_late),
-        change_sign = sign(estimate_early) != sign(estimate_late),
+        change_sign = sign_early != sign_late,
         Change      = case_when(
                         ! significant ~ "Insignificant",
-                        change_sign  & increasing   ~ "Change sign to +",
-                        change_sign  & ! increasing ~ "Change sign to -",
-                        increasing   & size_inc   ~ "Increasing +",
-                        ! increasing & size_inc   ~ "Decreasing -",
-                        increasing   & ! size_inc ~ "Increasing towards 0",
-                        ! increasing & ! size_inc ~ "Decreasing towards 0"
+                        change_sign   ~ "Change sign",
+                        size_inc      ~ "Size increasing",
+                        ! size_inc    ~ "Size decreasing"
                       )
       ) %>% 
-      select(score_name, children, Change, increasing, significant) %>% 
+      select(score_name, children, Change, significant, change_sign, 
+        sign_early, sign_late) %>% 
       mutate(children = ifelse(children, "Respondents", "Parents")) %>% 
       pivot_wider(names_from = children, 
-        values_from = c(Change, increasing, significant)) %>% 
+        values_from = c(Change, significant, change_sign, sign_early, 
+        sign_late)) %>% 
       mutate(overall = case_when(
-              increasing_Respondents   & increasing_Parents   ~ 
-                "Consistent increase",
-              ! increasing_Respondents & ! increasing_Parents ~ 
-                "Consistent decrease",
-              increasing_Respondents & ! increasing_Parents
-                 & (significant_Parents | significant_Respondents) ~ 
-                "Changes direction",
-              TRUE ~ 
-              "Other"
+              ! change_sign_Parents & ! change_sign_Respondents & 
+                sign_late_Parents == sign_early_Parents ~ 
+                "Consistent",
+              TRUE ~ "Changes direction"
             ))

### 2020-05-15 _drake.R

@@ -12,14 +12,14 @@ source("make-data.R")
 source("regressions.R")
 
 
-pgs_dir       <- "~/UKBB data 2019/polygenic_scores/"
-ph_file       <- "~/UKBB data 2019/UKB.EA_pheno.coordinates.QC.david.csv"
-pcs_file      <- "~/UKBB data 2019/UKB.HM3.100PCs.40310.txt"
-famhist_file  <- "~/Dropbox/assortative mating/biobank-analysis/david.family_history.traits.out.csv"
-famhist2_file <- "~/UKBB data 2019/david.family_history.traits.20042020.out.csv"
-famhist3_file <- "~/UKBB data 2019/david.family_history.traits.05052020.out.csv"
+pgs_dir       <- "polygenic_scores/"
+ph_file       <- "UKB.EA_pheno.coordinates.QC.david.csv"
+pcs_file      <- "UKB.HM3.100PCs.40310.txt"
+famhist_file  <- "david.family_history.traits.out.csv"
+famhist2_file <- "david.family_history.traits.20042020.out.csv"
+famhist3_file <- "david.family_history.traits.05052020.out.csv"
 rgs_file      <- "EA3_rgs.10052019.rgs.csv"
-mf_pairs_file <- "~/UKBB data 2019/spouse_pair_info/UKB_out.mf_pairs_rebadged.csv"
+mf_pairs_file <- "spouse_pair_info/UKB_out.mf_pairs_rebadged.csv"
 nomis_file    <- "nomis-2011-statistics.csv"
 ghs_file      <- "UKDA-5804-stata8/stata8/Ghs06client.dta"
 
@@ -133,18 +133,18 @@ plan <- drake_plan(
   
   
   res_sex      = {
-   sexes <- rlang::exprs(
-           sex == 0, 
-           sex == 1
-         )
-   res_sex <- expand_grid(
-           score_name = score_names, 
-           subset     = sexes
-         ) %>% 
-           pmap_dfr(run_regs_subset, famhist = famhist)
-         res_sex$sex <- ifelse(res_sex$subset == "sex == 0", 
-           "Female", "Male") 
-   res_sex
+    sexes <- rlang::exprs(
+            sex == 0, 
+            sex == 1
+          )
+    res_sex <- expand_grid(
+            score_name = score_names, 
+            subset     = sexes
+          ) %>% 
+            pmap_dfr(run_regs_subset, famhist = famhist)
[…]

### 2020-05-15 make-data.R

@@ -95,19 +95,6 @@ edit_famhist <- function (famhist, score_names) {
   famhist$moth_age <- famhist$f.1845.0.0
   famhist$moth_age[famhist$moth_age < 0] <- NA
   
-  # TODO: get f.53 (year attended assessment) from Abdel
-  # the below is a quick hack
-  famhist$fath_YOB <- 2008 - famhist$fath_age
-  famhist$moth_YOB <- 2008 - famhist$moth_age
-  
-  famhist %<>% group_by(YOB) %>% mutate(
-          parents_imp_YOB  = rowMeans(cbind(fath_YOB, moth_YOB), na.rm = TRUE),
-          mean_parents_YOB = mean(parents_imp_YOB, na.rm = TRUE),
-          parents_imp_YOB  = ifelse(is.na(parents_imp_YOB),
-                               mean_parents_YOB, parents_imp_YOB),
-          mean_parents_YOB = NULL
-        ) %>% ungroup()
-  
   # full brothers and sisters
   famhist$nbro <- pmax(famhist$f.1873.0.0, famhist$f.1873.1.0, 
     famhist$f.1873.2.0, na.rm = TRUE)

### 2020-05-15 regressions.R

@@ -59,9 +59,7 @@ run_regs_weighted <- function (score_name, famhist, weights) {
 
 run_regs_period <- function (children, score_name, famhist) {
   dep_var  <- if (children) "n_children" else "n_sibs"
-  year_var <- if (children) "YOB" else "parents_imp_YOB"
-  famhist$year_split <- famhist[[year_var]] >= median(famhist[[year_var]], 
-        na.rm = TRUE)
+  famhist$year_split <- famhist$YOB >= median(famhist$YOB, na.rm = TRUE)
   famhist$year_split <- factor(famhist$year_split, labels = c("early", "late"))
   fml <- as.formula(glue("{dep_var} ~ {score_name}:year_split"))
   mod <- lm(fml, famhist)

### 2020-05-16 why-negative-selection.Rmd
trivia

@@
   - Actually almost no change.
+  - Overall, diffs between siblings and children regressions
+    now look more like ascertainment bias than change over time.
 
@@
   - Can we work out how to plot selection effects on the "same scale"?
-  - Simplest thing would be to use birth dates
+  - Simplest thing would be to use birth dates of children.
 
@@
         term = pretty_names(term),
-        term = fct_reorder(term, estimate),
+        term = fct_reorder(term, estimate, order_abs),
         dep.var = recode(dep.var, 

### 2020-05-18 why-negative-selection.Rmd
trivia
Add regressions controlling for age at first live birth.

@@
 
-* Use Abdel's 100 PCs, not UKBB's 40. DONE
+* Why so many missings for age_fulltime_edu?
+  - ah, need f.6138 for those who went to college
+  - but note that YearsEdu.ISCED has many fewer missings
+  - ask Abdel how it was calculated
+  - use this for survey weighting?
 
@@
     now look more like ascertainment bias than change over time.
+  - (I checked my code: I hadn't been controlling for one main term when
+    running the interaction. Doh.)
 
@@
   - Can we work out how to plot selection effects on the "same scale"?
-  - Simplest thing would be to use birth dates of children.
+  - Simplest thing would be to use birth dates of children. (Maybe average of
+    first and last live birth)
+  - Also weight?
 
@@
 * control for age in n_partners regressions
-
-* check for new 100 PCs
 
@@
 
-drake::loadd(famhist)
 drake::loadd(rgs)
@@
 ```{r}
+
+drake::loadd(famhist)
 pgs_over_time <- famhist %>% 
@@
       ggplot(aes(estimate, dep.var, colour = p.value < 0.05/{{n_regs}})) + 
-        geom_jitter(height = 0.15, width = 0) + 
+        geom_jitter(height = 0.15, width = 0, alpha = 0.9) + 
         scale_x_log10(labels = scales::label_number(accuracy = 0.0001)) +
@@
 res_period %<>% 
-      tidyr::extract(term, c("score_name", "period"),
-        regex="(.*):year_split(.*)")
+      tidyr::extract(term, c("period", "score_name"),
+        regex="year_split(.*):(.*)")
 n_tests <- nrow(res_period)/2
@@
         term = fct_reorder(term, estimate, order_abs),
-        dep.var = recode(dep.var, 
+        dep.var = dplyr::recode(dep.var, 
           "n_children" = "N Children", "n_sibs" = "N Siblings")
@@
 
+```
+
+## Age at first live birth
+
+```{r corr-age-flb}
+
+drake::loadd(res_age_flb)
+drake::loadd(res_sex)
+
+res_sex_comparison <- res_sex %>% 
+      filter(sex == "Female") %>% 
+      mutate(score_name = term) %>% 
+      select(score_name, term, estimate:conf.high)
+
+res_age_flb %<>% filter(term != "age_flb")
+
+res_combined <- bind_rows(
+        raw      = res_sex_comparison, 
+        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
+        .id      = "reg.type"
+      ) %>% 
+      arrange(score_name)
+
+raw_flb_comparison <- res_combined %>% 
+      pivot_wider(
+        id_cols     = score_name, 
+        names_from  = reg.type, 
+        values_from = estimate
+      )
+
+cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
+opp_signed <- sum(
+        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
+      )
+
+```
+
+Figure \@ref(fig:plot-age-flb) shows the results of *children* regressions for 
+women only,
+controlling for age at first live birth. Effect sizes are greatly reduced.
+In `r opp_signed` out of `r length(score_names)` cases, they are of the 
+opposite sign. The correlation between effect sizes
+controlling for age at first live birth, and raw effect sizes, is `r cor_raw_flb`.
+
+```{r plot-age-flb, fig.cap = "Effect sizes on number of children, controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}
+
+n_regs <- as.double(nrow(res_age_flb))
+
+res_combined %>% 
+      mutate(
+        score_name = fct_reorder(score_name, estimate, order_abs),
+        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for age FLB")
+      ) %>% 
+      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
+        color = p.value < 0.05/{{n_regs}})) + 
+        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
+        geom_point(shape = 21, size = 1.8, stroke = 0.8) + 
+        scale_color_manual(values = pval_palette) +
+        scale_fill_brewer(type = "qual", palette = 3)
 ```

### 2020-05-18 make-data.R

@@ -5,6 +5,8 @@ suppressPackageStartupMessages({
   library(dplyr)
   library(rlang)
   library(santoku)
+  library(haven)
+  loadNamespace("car") # very annoying if it overwrites recode
 })
 
 
@@ -47,6 +49,7 @@ make_famhist <- function (
         famhist_file,
         famhist2_file,
         famhist3_file,
+        famhist4_file,
         pgs_dir
       ) {
   ph <- read_csv(ph_file, col_types = cols(
@@ -58,17 +61,24 @@ make_famhist <- function (
   famhist <- read_csv(famhist_file, col_types = strrep("d", 40))
   famhist2 <- read_csv(famhist2_file, col_types = strrep("d", 17))
   famhist3 <- read_csv(famhist3_file)
-  
+  famhist4 <- read_csv(famhist4_file, col_types = strrep("d", 33))
+  famhist5 <- read_csv(famhist5_file, col_types = strrep("d", 4))
   names(famhist) <- paste0("f.", names(famhist))
   names(famhist) <- gsub("\\-", ".", names(famhist))
   names(famhist2) <- paste0("f.", names(famhist2))
   names(famhist2) <- gsub("\\-", ".", names(famhist2))
   names(famhist3) <- paste0("f.", names(famhist3))
   names(famhist3) <- gsub("\\-", ".", names(famhist3))
+  names(famhist4) <- paste0("f.", names(famhist4))
+  names(famhist4) <- gsub("\\-", ".", names(famhist4))
+  names(famhist5) <- paste0("f.", names(famhist5))
+  names(famhist5) <- gsub("\\-", ".", names(famhist5))
   
   famhist %<>% left_join(ph, by = c("f.eid" = "eid"))
   famhist %<>% left_join(famhist2, by = "f.eid")
   famhist %<>% left_join(famhist3, by = "f.eid")
+  famhist %<>% left_join(famhist4, by = "f.eid")
+  famhist %<>% left_join(famhist5, by = "f.eid")
   
   # only "genetic" whites
   famhist %<>% filter(! is.na(genetic_ethnic_grouping))
@@ -89,24 +99,36 @@ edit_famhist <- function (famhist, score_names) {
   # we get very few extra cases from adding f.2946.1.0 etc, and it makes calculating
[…]

### 2020-05-19 why-negative-selection.Rmd
clean up graph styling. Create n_older_siblings and fath_age_birth/moth_age_birth. Run regressions with controls for these.

@@
 * Divide siblings regressions by own YOB, into early and late.
-  - Use this to confirm the basic "selection decreasing" story
+  - Use this to confirm the basic "selection decreasing" story. DONE.
   - Actually almost no change.
@@
 
-* Plot average scores over time, as an intro graph. DONE
-  - Can we work out how to plot selection effects on the "same scale"?
-  - Simplest thing would be to use birth dates of children. (Maybe average of
-    first and last live birth)
-  - Also weight?
+* Work more on weighting the data?
+  - weight by age at first live birth
+  - education using ISCED
 
-* How much does age at first live birth "account for" selection?
-  - Need f.3872 age at flb for women with one child
-
-* Work more on weighting the data?
+* direct regressions of scores on age_flb
 
@@
 
+huxtable::set_default_properties(latex_float = "h!")
+
 theme_set(theme_minimal())
+my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dotted")
+my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
 pval_palette <- c("FALSE" = "grey50", "TRUE" = "orange2")
-my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dashed")
-order_abs <- function (x) x[1] # for use when reordering factors
 
-huxtable::set_default_properties(latex_float = "h!")
+pval_scale <- function (...) scale_color_manual(values = pval_palette, ...)
+
+pval_geom_point <- function (size = 1.5, stroke = 0.7, ...) {
+  geom_point(shape = 21, size = size, stroke = stroke, ...)
+}
+
+my_fill_scale <- function (n, aesthetics = "fill", ...) {
+  if (n == 2) {
+    scale_color_brewer(type = "qual", palette = 3, aesthetics = aesthetics, ...)
+  } else {
+    scale_color_brewer(aesthetics = aesthetics, ...)
+  }
+}
+
+# for use when reordering factors. Order by the nth score in a group
+order_abs <- function (n = 1) {
+  function (x) x[n] 
+}
 
@@
         ))
+
+  acronyms <- c("WHR", "BMI", "DPW", "CPD", "SI", "AI", "WC")
+  for (acronym in acronyms) {
+    regex <- paste0("\\b", acronym, "\\b")
+    names <- gsub(regex, toupper(acronym), names, ignore.case = TRUE)
+  }
 
@@
 
-```{r}
+```{r calc-means-over-time}
 
@@
       summarize(across(all_of(score_names), mean, na.rm = TRUE)) %>% 
+      pivot_longer(
+        -YOB, 
+        names_to = "score_name",
+        values_to = "score"
+      )
+
+```
+
+```{r calc-means-over-time-children, eval = FALSE}
+pgs_over_time_children <- famhist %>% 
+      mutate(
+        YOB_kids = (year_flb + year_llb)/2,
+        YOB      = santoku::chop(
+                     YOB_kids, 
+                     brk_left(seq(1960, 1995, 5), close_end = FALSE), 
+                     seq(60, 90, 5),
+                     extend = FALSE
+                   )
+      ) %>% 
+      filter(! is.na(YOB)) %>% 
+      group_by(YOB) %>% 
+      summarize(across(all_of(score_names), 
+        ~weighted.mean(.x, n_children, na.rm = TRUE)
+      )) %>% 
       pivot_longer(-YOB, names_to = "score_name", values_to = "score")
+
+pgs_over_time <- bind_rows(
+        Respondents = pgs_over_time,
+        Children    = pgs_over_time_children,
+        .id         = "Generation"
+      )
 
@@
 pgs_over_time %>% 
-      mutate(score_name = pretty_names(score_name),) %>% 
+      mutate(score_name = pretty_names(score_name)) %>% 
       ggplot(aes(YOB, score, group = score_name)) + 
@@
       facet_wrap(~score_name) +
-      theme(strip.text = element_text(size = 7)) +
+      theme(
+        strip.text  = element_text(size = 7), 
+        axis.text.x = element_text(size = 7)
+      ) +
       labs(y = "PGS", x = "Birth year (5 year intervals)")
@@
 
-```{r plot-basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children. Scores are plotted on different scales.", fig.height = 7}
+```{r plot-basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children.", fig.height = 7}
 
@@
         term    = pretty_names(term),
-        term    = fct_reorder(term, estimate, order_abs),
+        term    = fct_reorder(term, estimate, order_abs(1)),
         dep.var = fct_inorder(dep.var) 
       ) %>% 
-      ggplot(aes(x = term, y = estimate, shape = reg.type, 
+      ggplot(aes(y = term, x = estimate, fill = reg.type, 
         colour = p.value < 0.05/{{n_regs}})) +
-      my_hline + 
-      geom_point(size = 2.5, alpha = 0.8) + 
-      labs(y = "Effect size", x = "") +
-      facet_wrap(vars(dep.var), ncol = 2, scales = "free_x") +
-      coord_flip() +
-      scale_colour_manual(values = pval_palette) +
-      scale_shape_manual(values = c("circle", "triangle"), 
-        guide = guide_legend(title = "")) +
+      my_vline + 
+      pval_geom_point() +
+      labs(x = "Effect size", y = "") +
+      facet_wrap(vars(dep.var), ncol = 2) +
+      pval_scale() +
+      my_fill_scale(n = 2, guide = guide_legend(title = "")) +
       theme(
         legend.position = "bottom", 
-        legend.margin = margin(2,2,2,2)
+        legend.margin   = margin(2, 2, 2, 2)
       )
@@
         labs(x = "Effect size (log scale)", y = "") + 
-        scale_colour_manual(values = pval_palette)
+        pval_scale(aesthetics = "colour")
 
@@
       facet_wrap(vars(score_name), nrow = 4) + 
-      scale_fill_manual(values = pval_palette) +
-      scale_colour_manual(values = c("darkgreen", "navy")) +
+      pval_scale() +
+      my_fill_scale(n = 2, aesthetics = "colour") +
       scale_linetype_manual(values = c("FALSE" = "dashed", "TRUE" = "solid"))
@@
         term    = pretty_names(term),
-        term    = fct_reorder(term, estimate_weighted, order_abs)
+        term    = fct_reorder(term, estimate_weighted, order_abs(1))
       ) %>% 
@@
       ) %>% 
-      ggplot(aes(x = term, y = estimate, fill = Regression,
+      ggplot(aes(y = term, x = estimate, fill = Regression,
         colour = p.value < 0.05/{{n_regs}})) +
-      my_hline + 
-      geom_point(size = 1.5, stroke = .8, shape = 21) + 
-      labs(y = "Effect size", x = "") +
-      coord_flip() +
-      scale_fill_manual(values = 
-        c("unweighted" = "transparent", "weighted" = "darkgreen")) +
-      scale_colour_manual(values = pval_palette)
+      my_vline + 
+      pval_geom_point() +
+      labs(x = "Effect size", y = "") +
+      my_fill_scale(n = 2) +
+      pval_scale(aesthetics = "colour")
 
@@
         term = pretty_names(term),
-        term = fct_reorder(term, estimate, order_abs),
+        term = fct_reorder(term, estimate, order_abs(2)),
         dep.var = dplyr::recode(dep.var, 
-          "n_children" = "N Children", "n_sibs" = "N Siblings")
+          "n_children" = "N Children", "n_sibs" = "N Siblings"),
+        dep.var = fct_relevel(dep.var, "N Siblings")
       ) %>% 
-      ggplot(aes(term, estimate, color = p.value < 0.05/{{n_regs}}, shape  = PCs)) + 
-        geom_pointrange(aes(ymin = conf.low, ymax = conf.high)) +
-        my_hline +
-        facet_wrap(vars(dep.var), scales = "free_x") +
-        coord_flip() +
-        scale_color_manual(values = pval_palette) +
+      ggplot(aes(y = term, x = estimate, color = p.value < 0.05/{{n_regs}}, 
+        fill  = PCs)) + 
+        pval_geom_point(alpha = 0.8) +
+        my_vline +
+        facet_wrap(vars(dep.var)) +
+        pval_scale() +
+        my_fill_scale(n = 2, direction = -1) +
         theme(legend.position = "bottom")
@@
 
+n_regs <- as.double(nrow(res_sex))
 res_sex %>% 
@@
         term = pretty_names(term),
-        term = fct_reorder(term, estimate, order_abs)
+        term = fct_reorder(term, estimate, order_abs(1))
       ) %>% 
-      ggplot(aes(term, estimate, group = sex, colour = sex)) +
-        my_hline +
-        geom_pointrange(aes(ymin = conf.low, ymax = conf.high), alpha = 0.7) +
-        coord_flip() +
-        scale_colour_brewer(palette = 2, type = "qual")
+      ggplot(aes(y = term, x = estimate, fill = sex, 
+            colour = p.value < 0.05/{{n_regs}})) +
+        my_vline +
+        pval_geom_point() +
+        pval_scale() +
+        my_fill_scale(n = 2, direction = -1)
 
@@
       mutate(
-        score_name = fct_reorder(score_name, estimate, order_abs),
+        score_name = fct_reorder(score_name, estimate, order_abs(2)),
         Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for age FLB")
@@
         geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
-        geom_point(shape = 21, size = 1.8, stroke = 0.8) + 
-        scale_color_manual(values = pval_palette) +
+        pval_geom_point() +
+        pval_scale() +
         scale_fill_brewer(type = "qual", palette = 3)
+```
+
+```{r calc-age-birth-parents}
+
+drake::loadd(res_age_birth_parents)
+drake::loadd(res_all)
+
+res_sibs_raw <- filter(res_all, dep.var == "N siblings", reg.type == "raw")
+compare_abp_raw <- res_age_birth_parents %>% 
+      filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
+      left_join(res_sibs_raw, by = "term")
+
+cor_raw_fath <- with(
+  compare_abp_raw %>% filter(control == "fath_age_birth"), 
+  cor(estimate.x, estimate.y))
+cor_raw_moth <- with(
+  compare_abp_raw %>% filter(control == "moth_age_birth"), 
+  cor(estimate.x, estimate.y))
+```
+
+The dataset has no information on fathers' age at birth of their first child.
+However, we can calculate this information for the parents' generation,
+for the subsets of respondents who reported their mother's or father's age
+and who had no elder siblings. We run *sibling* regressions on these subsets,
+controlling for either parent's age at their birth.
+Figure \@ref(fig:plot-age-birth-parents) shows
+the results. Effect sizes are very similar, whether controlling for
+father's or mother's age at respondent's birth or mother's age at respondent's birth. Unlike for the respondent's own generation, effect sizes are positively correlated with
+the effect sizes from bivariate regressions (father's age at birth: 
+$\rho$ `r cor_raw_fath`; mother's age at birth: $\rho$ `r cor_raw_moth`).
+
+```{r plot-age-birth-parents}
+
+n_regs <- as.double(nrow(res_age_birth_parents)/2)
+
+res_age_birth_parents %>% 
+    filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
+    mutate(
+      term    = pretty_names(term),
+      term    = fct_reorder(term, estimate, order_abs(2)),
+      Control = ifelse(control == "fath_age_birth", 
+                  "Father's age at birth", 
+                  "Mother's age at birth"
+                )
+    ) %>% 
+    ggplot(aes(y = term, x = estimate, fill = Control, 
+      colour = p.value < 0.05/{{n_regs}})) + 
+      pval_geom_point(alpha = 0.8) +  
+      my_vline +
+      my_fill_scale(2) +
+      pval_scale()
+
 ```
@@
 
-n_regs <- nrow(res_partners)
+n_regs <- as.double(nrow(res_partners))
 res_partners %>% 
@@
         score_name = pretty_names(score_name),
-        score_name = fct_reorder(score_name, estimate, order_abs)
+        score_name = fct_reorder(score_name, estimate, order_abs(1))
       ) %>% 
-      ggplot(aes(score_name, estimate, colour = `N partners`)) + 
-        my_hline +
-        geom_pointrange(aes(ymin = conf.low, ymax = conf.high), alpha = .8) +
+      ggplot(aes(y = score_name, x = estimate, fill = `N partners`, 
+        colour = p.value < 0.05/{{n_regs}})) + 
+        my_vline +
+        pval_geom_point() +
         facet_wrap(vars(sex)) +
-        coord_flip() +
-        scale_colour_manual(values = c("darkred", "steelblue")) +
+        my_fill_scale(n = 2) +
+        pval_scale() +
         theme(legend.position = "bottom")
@@
 drake::loadd(res_edu)
+n_regs <- as.double(nrow(res_edu))
 res_edu %>% 
@@
         term = pretty_names(term), 
-        term = fct_reorder(term, estimate, order_abs),
+        term = fct_reorder(term, estimate, order_abs(1)),
         "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18")
       ) %>% 
-      ggplot(aes(term, estimate, colour = `Age left FTE`)) +
-        my_hline +
-        geom_pointrange(aes(ymin=conf.low, ymax=conf.high), alpha = 0.8,
-          position = position_dodge(width = 0.4)) + 
-        scale_colour_viridis_d(begin = 0.1, end = 0.85) +
-        coord_flip()
+      ggplot(aes(y = term, x = estimate, colour = p.value < 0.05/{{n_regs}}, 
+        fill = `Age left FTE`)) +
+        my_vline +
+        pval_geom_point() + 
+        pval_scale() +
+        my_fill_scale(n = 3, direction = -1)
+
 ```
@@
         term = pretty_names(term), 
-        term = fct_reorder(term, estimate, order_abs),
+        term = fct_reorder(term, estimate, order_abs(1)),
         Income = factor(income_cat, 
@@
       ) %>% 
-      ggplot(aes(term, estimate, fill = Income, colour = p.value < 0.05/{{n_regs}})) +
-        my_hline +
-        geom_point(alpha = 0.8, shape = 21, size = 1.5, stroke = 0.9) + 
-        scale_fill_brewer(direction = -1) +
-        scale_colour_manual(values = pval_palette) +
-        coord_flip()
+      ggplot(aes(y = term, x = estimate, fill = Income, 
+        colour = p.value < 0.05/{{n_regs}})) +
+        my_vline +
+        pval_geom_point(alpha = 0.8) + 
+        my_fill_scale(n = 5, direction = -1) +
+        pval_scale()
 ```

### 2020-05-19 make-data.R

@@ -7,6 +7,7 @@ suppressPackageStartupMessages({
   library(santoku)
   library(haven)
   loadNamespace("car") # very annoying if it overwrites recode
+  loadNamespace("matrixStats")
 })
 
 
@@ -99,16 +100,19 @@ edit_famhist <- function (famhist, score_names) {
   # we get very few extra cases from adding f.2946.1.0 etc, and it makes calculating
   # father's year of birth more complex
   
+  names(famhist) <- sub("age_at_reqruitment", "age_at_recruitment", 
+        names(famhist))
+  
   # remove negatives
   negative_to_na <- function (x) {
     x[x < 0] <- NA
     x
   }
-  
   famhist %<>% mutate(across(
       c(age_fulltime_edu, starts_with(c(
         "f.2946", "f.1845", "f.2754", "f.738",  "f.2764", "f.2405", "f.2734",
-        "f.2149", "f.1873", "f.1883", "f.2784", "f.2794", "f.709",  "f.3872"
+        "f.2149", "f.1873", "f.1883", "f.2784", "f.2794", "f.709",  "f.3872",
+        "f.5057"
       ))), 
       negative_to_na
     )
@@ -119,6 +123,8 @@ edit_famhist <- function (famhist, score_names) {
   # roughly speaking, these are ages in 2007-10
   famhist$fath_age <- famhist$f.2946.0.0
   famhist$moth_age <- famhist$f.1845.0.0
+  famhist$fath_age_birth <- famhist$fath_age - famhist$age_at_recruitment
+  famhist$moth_age_birth <- famhist$moth_age - famhist$age_at_recruitment
   
   # full brothers and sisters
   famhist$nbro <- pmax(famhist$f.1873.0.0, famhist$f.1873.1.0, 
@@ -126,6 +132,12 @@ edit_famhist <- function (famhist, score_names) {
   famhist$nsis <- pmax(famhist$f.1883.0.0, famhist$f.1883.1.0, 
         famhist$f.1883.2.0, na.rm = TRUE)
   famhist$n_sibs <- famhist$nbro + famhist$nsis + 1
+  # a few people give varying answers, we assume median is fine.
+  # including later answers picks up c. 10K extra people
+  famhist$n_older_sibs <- matrixStats::rowMedians(
+          as.matrix(famhist[, c("f.5057.0.0", "f.5057.1.0", "f.5057.2.0")]),
+          na.rm = TRUE
+        )
   
[…]

### 2020-05-19 why-negative-selection.Rmd
regressions on age of parents at own birth

@@
 
-* Divide siblings regressions by own YOB, into early and late.
-  - Use this to confirm the basic "selection decreasing" story. DONE.
-  - Actually almost no change.
-  - Overall, diffs between siblings and children regressions
-    now look more like ascertainment bias than change over time.
-  - (I checked my code: I hadn't been controlling for one main term when
-    running the interaction. Doh.)
-
 * Work more on weighting the data?
   - weight by age at first live birth
-  - education using ISCED
-
-* direct regressions of scores on age_flb
+  - education using ISCED?
+  - geography?
+  - ask for existing work by others?
 
@@
 }
+
+my_labs <- function (x = "Effect size", y = "", ...) labs(x = x, y = y, ...)
 
@@
         my_fill_scale(n = 2, direction = -1) +
+        my_labs() +
         theme(legend.position = "bottom")
@@
         pval_scale() +
-        my_fill_scale(n = 2, direction = -1)
+        my_fill_scale(n = 2, direction = -1) +
+        my_labs()
 
@@
         pval_scale() +
-        scale_fill_brewer(type = "qual", palette = 3)
+        scale_fill_brewer(type = "qual", palette = 3) +
+        my_labs()
 ```
@@
 
-The dataset has no information on fathers' age at birth of their first child.
-However, we can calculate this information for the parents' generation,
-for the subsets of respondents who reported their mother's or father's age
+We can run similar regressions for the parents' generation, using the subsets of respondents who reported their mother's or father's age
 and who had no elder siblings. We run *sibling* regressions on these subsets,
@@
 the results. Effect sizes are very similar, whether controlling for
-father's or mother's age at respondent's birth or mother's age at respondent's birth. Unlike for the respondent's own generation, effect sizes are positively correlated with
+father's or mother's age at respondent's birth or mother's age at respondent's birth. Unlike for the respondents' generation, effect sizes are positively correlated with
 the effect sizes from bivariate regressions (father's age at birth: 
@@
 
-```{r plot-age-birth-parents}
+```{r plot-age-birth-parents, fig.cap = "Effect sizes on number of siblings controlling for parents' age at birth, eldest siblings"}
 
@@
       my_fill_scale(2) +
-      pval_scale()
+      pval_scale() +
+      my_labs()
+
+```
+
+```{r calc-age-flb-dv}
+
+drake::loadd(res_age_flb_dv)
+drake::loadd(res_age_birth_parents_dv)
+
+compare_flb_dv <- left_join(
+        res_age_flb_dv           %>% select(term, estimate),
+        res_age_birth_parents_dv %>% select(term, estimate, dep.var),
+        by = "term"
+      )
+
+cor_flb_fath <- with(compare_flb_dv %>% filter(dep.var == "fath_age_birth"),
+        cor(estimate.x, estimate.y)
+      )
+cor_flb_moth <- with(compare_flb_dv %>% filter(dep.var == "moth_age_birth"),
+        cor(estimate.x, estimate.y)
+      )
+```
+
+These results suggest that polygenic scores may directly correlate with
+age at first live birth. Figure \@ref(fig:plot-age-flb-dv) plots
+estimated effect sizes from bivariate regressions for respondents, and
+Figure \@ref(fig:plot-age-birth-parents-dv) does the same for their parents. 
+Effect sizes are reasonably large. They are also very highly correlated across 
+generations. Effect sizes of PGS on father's age at own birth, and on own age
+at first live birth, have a correlation of `r cor_flb_fath`; for mother's age
+and own age it is `r cor_flb_moth`
+
+```{r plot-age-flb-dv, fig.cap = "Effects of polygenic scores on age at first live birth"}
+
+n_regs <- as.double(nrow(res_age_flb_dv))
+
+res_age_flb_dv %>% 
+      mutate(
+        term = pretty_names(term),
+        term = fct_reorder(term, estimate, order_abs(1))
+      ) %>% 
+      ggplot(aes(estimate, term, colour = p.value < 0.05 / {{n_regs}})) + 
+      my_vline +
+      pval_geom_point(fill = "steelblue4") +
+      pval_scale() + 
+      my_labs()
+
+```
+
+```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at own birth, eldest siblings"}
+
+n_regs <- as.double(nrow(res_age_birth_parents_dv))
+
+res_age_birth_parents_dv %>% 
+      mutate(
+        term = pretty_names(term),
+        term = fct_reorder(term, estimate, order_abs(1)),
+        "Dependent variable" = dplyr::recode(dep.var, 
+          "fath_age_birth" = "Father's age", 
+          "moth_age_birth" = "Mother's age"
+        )
+      ) %>% 
+      ggplot(aes(estimate, term, fill = `Dependent variable`, 
+        colour = p.value < 0.05 / {{n_regs}})) + 
+        my_vline +
+        pval_geom_point() +
+        pval_scale() +
+        my_fill_scale(n = 2, direction = -1) +
+        my_labs()
 
@@
         pval_scale() +
-        theme(legend.position = "bottom")
+        theme(legend.position = "bottom") +
+        my_labs(x = "", y = "")
 
@@
         pval_scale() +
-        my_fill_scale(n = 3, direction = -1)
+        my_fill_scale(n = 3, direction = -1) +
+        my_labs()
 
@@
         my_fill_scale(n = 5, direction = -1) +
-        pval_scale()
+        pval_scale() +
+        my_labs()
 ```

### 2020-05-19 why-negative-selection.Rmd
Improve GHS weights. Weight by age/sex/educ and by age at first live birth (women).

@@
 
-* Work more on weighting the data?
-  - weight by age at first live birth
-  - education using ISCED?
-  - geography?
+* Work more on weighting the data
   - ask for existing work by others?
+  - deal with the fact that GHS 06 has its own weights and use
+    subset.svydesign to reweight?
 
@@
   some of the many NAs for f.2141, num sex partners)
+
+## Paper plan
+
+* Intro: changes in PGS over time
+* We investigate using n siblings and n children. Effect sizes are much larger
+  for n siblings.
+  - This isn't because of change over time. (Details in appendix.)
+  - Instead we hypoth that it is due to selection bias.
+* To support this argument, we show that the strength of selection 
+  varies dramatically across social groups.
+  - Women more than men.
+  - Effect sizes are larger among less educated and poorer respondents. (Appendix.)
+  - Effect sizes are reversed in people with less than median sexual partners.
+  - Effect sizes are reversed if we control for age at first live birth.
+  - There are strong effects of PGS on age at flb, both among respondents and
+    their parents
+* The UK Biobank sample is highly selected with respect to education, income
+  and age at FLB. (Show using GHS.)
+  - GHS variable for age at CHBNBM1; raw data babdata
+* We reweight the data to get more accurate estimates.
+* The data can be explained by a story where...
+  -
 
@@
 
+# Accounting for ascertainment bias
+
+Effect sizes tend to be smaller for children regressions. This could be caused
+by ascertainment bias in the UK Biobank sample -- e.g., if respondents
+themselves are a more selected sample than the respondents' parents. To check
+this, we weighted UK Biobank participants by sex, age (within the 40-71 age
+group), and education. Our sample weights are calibrated based on the 2006
+General Household Survey. We then rerun the basic *children* regressions.
+
 ```{r res-weighted}
@@
 
-# Accounting for ascertainment bias
-
-Effect sizes tend to be smaller for children regressions. This
-could be caused by ascertainment bias in the UK Biobank sample -- e.g.,
-if respondents themselves are a more selected sample than the respondents' 
-parents. To check this, we weighted UK Biobank participants by age of leaving
-full time education. Our sample weights are based on the 2006 General Household
-Survey, calculating proportions within age and sex cells among White British
-respondents. We then rerun the basic *children* regressions.
-
 Figure \@ref(fig:plot-weighted) shows the results, with unweighted estimates
 plotted for comparison. Weighting increases effect sizes on average by 
-`r 100 * (mean_prop_wtd_unwtd - 1)` per cent. These weights are only basic 
-attempts to correct for ascertainment bias. Effect sizes might be increased 
-further by more precise weighting.
+`r 100 * (mean_prop_wtd_unwtd - 1)` per cent.
 
@@
 
+```
+
+We also weighted data by age at first live birth and age, among women only.
+These weights are only basic attempts to correct for ascertainment bias. Effect
+sizes might be increased further by more precise weighting.
+
+```{r plot-flb-weights, fig.cap = "Effect sizes on number of children, female respondents weighted by age at first live birth"}
+
+drake::loadd(res_flb_weights)
+drake::loadd(res_sex)
+
+bind_rows(
+        Weighted   = res_flb_weights, 
+        Unweighted = res_sex %>% filter(sex == "Female") %>% select(-subset, - sex),
+        .id        = "Regression"
+      ) %>% 
+      mutate(
+        term = pretty_names(term),
+        term = fct_reorder(term, estimate, order_abs(1))
+      ) %>% 
+      ggplot(aes(estimate, term, fill = Regression, 
+        color = p.value < 0.05/66)) +
+        my_vline +
+        pval_geom_point() +
+        pval_scale() +
+        my_fill_scale(n = 2) +
+        my_labs()
 ```

### 2020-05-19 make-data.R

@@ -10,6 +10,11 @@ suppressPackageStartupMessages({
   loadNamespace("matrixStats")
 })
 
+# utility function:
+negative_to_na <- function (x) {
+  x[x < 0] <- NA
+  x
+}
 
 make_pcs <- function (pcs_file) {
   pcs <- read_table(pcs_file)
@@ -104,10 +109,6 @@ edit_famhist <- function (famhist, score_names) {
         names(famhist))
   
   # remove negatives
-  negative_to_na <- function (x) {
-    x[x < 0] <- NA
-    x
-  }
   famhist %<>% mutate(across(
       c(age_fulltime_edu, starts_with(c(
         "f.2946", "f.1845", "f.2754", "f.738",  "f.2764", "f.2405", "f.2734",
@@ -154,6 +155,9 @@ edit_famhist <- function (famhist, score_names) {
   famhist$age_fte_cat <- santoku::chop(famhist$age_fulltime_edu, 
     c(16, 18), 
     c("< 16", "16-18", "> 18"))
+  # -7 means never went to school. We recode to 0 for simpliciy
+  famhist$edu_qual[famhist$edu_qual == -7] <- 0
+  famhist$edu_qual[famhist$edu_qual == -3] <- NA
   
   # we use pmax, assuming that people *can* have given birth for the first
   # time in between surveys.
@@ -210,7 +214,9 @@ make_mf_pairs <- function (mf_pairs_file, famhist) {
 }
 
 
-weight_by_ghs <- function (ghs_file, famhist) {
+make_ghs_subset <- function(ghs_file) {
+  # calculate population totals from ghs file for variables of interest
+  # using only 40-70 year olds I guess!
   
   ghs <- haven::read_dta(ghs_file)
   # variables of interest: 
@@ -233,55 +239,70 @@ weight_by_ghs <- function (ghs_file, famhist) {
   # ten1 - tenure
   # llord - landlord
   
-  ghs <- ghs %>% 
+  ghs_subset <- ghs %>% 
     zap_labels() %>% 
-    filter(ethnic == 1, sex %in% 1:2, edage > 0, edage < age) %>% 
-    select(sex, age, edage, weight06, ten1, llord) 
-  
-  ghs_props <- ghs %>% 
-    group_by(sex, age, edage) %>% 
-    summarize(
-      weighted_n = sum(weight06)
[…]

### 2020-05-19 regressions.R

@@ -45,12 +45,12 @@ run_regs_pcs <- function (dep_var, famhist, pcs) {
 }
 
 
-run_regs_weighted <- function (score_name, famhist, weights) {
-  famhist <- left_join(famhist, weights, by = "f.eid")
-  
+run_regs_weighted <- function (score_name, famhist, weight_data) {
+  famhist <- inner_join(famhist, weight_data, by = "f.eid")
+
   fml <- as.formula(glue("n_children ~ {score_name}"))
-  regs <- tidy(lm(fml, famhist, weights = educ_weight,
-        na.action = na.exclude), conf.int = TRUE) 
+  mod <- lm(fml, data = famhist, weights = weights, na.action = na.exclude)
+  regs <- tidy(mod, conf.int = TRUE)
   regs %<>% filter(term != "(Intercept)")
   
   return(regs)

### 2020-05-19 weight-data.R

@@ -1,0 +1,35 @@
+
+suppressPackageStartupMessages({
+  library(survey)
+  library(dplyr)
+  library(haven)
+})
+
+
+#' @param fml A formula to be passed in to `survey::calibrate()`
+#'   Possible variables: sex, age_at_recruitment, edu_qual, age_flb...
+#'   These could be specified as factors, and should be for e.g. educ_level
+weight_by_ghs <- function (fml, ghs_subset, famhist) {
+
+  fml_vars <- all.vars(fml)
+  famhist %<>%  filter(across(all_of(fml_vars), Negate(is.na)))
+  ghs_subset %<>% filter(across(all_of(fml_vars), Negate(is.na)))
+  
+  pop_mx <- model.matrix(fml, data = ghs_subset)
+  pop_mx <- pop_mx * ghs_subset$weight06
+  pop_mx_totals <- colSums(pop_mx)
+  
+  design <- survey::svydesign(~1, data = famhist)
+  # `rake` works on each margin individually. `postStratify` works on multiple 
+  # margins, which might be better if we have enough data! `calibrate` uses a 
+  # linear model (presumably to predict the population frequencies? or the
+  # response probability?)
+  calibrate_design <- survey::calibrate(design,
+          formula = fml,
+          population = pop_mx_totals,
+          bounds = c(0, Inf)
+        )
+  calibrate_weights <- weights(calibrate_design)
+  
+  data.frame(f.eid = famhist$f.eid, weights = calibrate_weights)
+}

### 2020-05-20 why-negative-selection.Rmd
Include all women in res_flb_weights. Fix bug which only used one PC in basic regressions; rewrite to use residualized variables. Interact scores with age_flb terciles.

@@
 
-mean_prop_wtd_unwtd <- res_weighted_all %>% 
-      filter(sign(estimate_weighted) == sign(estimate_unweighted)) %>% 
-      summarize(mean = mean(prop_wtd_unwtd)) %>% 
-      pull(mean)
+mean_prop_wtd_unwtd <- mean(res_weighted_all$prop_wtd_unwtd)
 
@@
 
-We also weighted data by age at first live birth and age, among women only.
+```{r calc-res-flb-weights}
+
+drake::loadd(res_flb_weights)
+drake::loadd(res_sex)
+res_flb_all <- left_join(
+        res_flb_weights, 
+        res_sex %>% filter(sex=="Female") %>% select(-subset, -sex), 
+        by = "term", 
+        suffix = c(".wt", ".nowt")
+      ) %>% 
+      mutate(prop_estimate = estimate.wt/estimate.nowt)
+
+prop_wt_nowt <- mean(res_flb_all$prop_estimate)
+pct_change_score <- function (score) {
+  100 * (res_flb_all$prop_estimate[res_flb_all$term == score] - 1)
+}
+```
+
+We also weighted data for women with children only, by age, education and age 
+at first live birth. Weighting increased effect sizes on average by 
+`r 100 * (prop_wt_nowt - 1)`%. Figure \@ref(fig:plot-flb-weights) shows the results. There 
+are large changes for some variables, including EA3 
+(`r pct_change_score("EA3_excl_23andMe_UK")`%), cognitive
+ability (`r pct_change_score("cognitive_ability")`%) and
+waist-hip ratio (WHR; `r pct_change_score("whr_combined")`%).
+
 These weights are only basic attempts to correct for ascertainment bias. Effect
@@
 
-drake::loadd(res_flb_weights)
-drake::loadd(res_sex)
-
 bind_rows(
         Weighted   = res_flb_weights, 
-        Unweighted = res_sex %>% filter(sex == "Female") %>% select(-subset, - sex),
+        Unweighted = res_flb_nowt %>% select(-subset),
         .id        = "Regression"
@@
 women only,
-controlling for age at first live birth. Effect sizes are greatly reduced.
-In `r opp_signed` out of `r length(score_names)` cases, they are of the 
-opposite sign. The correlation between effect sizes
+controlling for age at first live birth. Effect sizes are greatly reduced. Partly,
+this is because the regressions exclude childless women. However,
+in `r opp_signed` out of `r length(score_names)` cases, effect sizes actually
+change sign with the controls. The correlation between effect sizes
 controlling for age at first live birth, and raw effect sizes, is `r cor_raw_flb`.
+
+Figure \@ref(fig:plot-age-flb-cross) shows *children* regressions estimated
+separately for each tercile of age at first live birth. Several effects
+are strikingly different. ADHD is selected for amongst the youngest
+third of mothers, but selected against among the oldest two-thirds. Educational
+attainment is selected for among the oldest two-thirds of mothers, but is not
+significantly selected among the youngest third. Similarly, several PGS of
+body measurements are selected against only among older mothers.
 
@@
         my_labs()
+```
+
+```{r plot-age-flb-cross, fig.cap = "Effect sizes on number of children, by age at first live birth terciles"}
+
+drake::loadd(res_age_flb_cross)
+
+# this counts each cross term as a separate test:
+n_regs <- as.double(nrow(res_age_flb_cross))
+
+res_age_flb_cross %>% 
+      mutate(
+        score_name = pretty_names(score_name),
+        score_name = fct_reorder(score_name, estimate, order_abs(1)),
+        term       = gsub("age_flb_cat(.*):.*", "\\1", term)
+      ) %>% 
+      ggplot(aes(estimate, score_name, fill = term, 
+        colour = p.value < 0.05/{{n_regs}})) +
+        my_vline +
+        pval_geom_point() +
+        pval_scale() +
+        my_fill_scale(n = 3, direction = -1, 
+          guide = guide_legend(title = "Age at FLB")) +
+        my_labs()
+
 ```

### 2020-05-20 _drake.R

@@ -76,11 +76,13 @@ plan <- drake_plan(
                     famhist
                   ),
   
-  flb_weights = weight_by_ghs(
-                  ~ age_at_recruitment + age_flb, 
-                  ghs_subset, 
-                  famhist
-                ),
+  flb_weights = {
+    weight_by_ghs(
+            ~ age_at_recruitment + factor(flb_cat) + factor(edu_qual), 
+            ghs_subset, 
+            famhist
+          )
+  },
   
   mf_pairs     = target(
                    make_mf_pairs(file_in(!! mf_pairs_file), famhist), 
@@ -92,17 +94,16 @@ plan <- drake_plan(
   
   
   res_all      = {
-    res_sibs <- run_regs(
+    famhist <- join_famhist_resid_scores(famhist, resid_scores)
+    res_sibs <- run_regs_basic(
             dep_var     = "n_sibs", 
             score_names = score_names,
-            famhist     = famhist, 
-            pcs         = pcs
+            famhist     = famhist
           )
-    res_chn <- run_regs(
+    res_chn <- run_regs_basic(
             dep_var     = "n_children", 
             score_names = score_names, 
-            famhist     = famhist, 
-            pcs         = pcs
+            famhist     = famhist
           )
     dplyr::bind_rows(
             "N siblings" = res_sibs, 
@@ -144,7 +145,7 @@ plan <- drake_plan(
                       famhist = famhist, 
                       weight_data = flb_weights
                     ),
-  
+
   
   res_period   = {
     expand_grid(
@@ -174,15 +175,31 @@ plan <- drake_plan(
   res_age_flb = {
     res <- map_dfr(setNames(score_names, score_names), 
             ~run_regs_fml(
-              fml = "n_children ~ {score_name} + age_flb",
+              fml        = "n_children ~ {score_name} + age_flb",
               score_name = .x, 
-              famhist = famhist
+              famhist    = famhist
             ), 
             .id = "score_name"
           )
     res %<>% filter(term != "(Intercept)")
     
     res
+  },
+  
+  
+  res_age_flb_cross = {
+    famhist$age_flb_cat <- santoku::chop_equally(famhist$age_flb, 3, 
+          lbl_integer("-"))
+    res <- map_dfr(setNames(score_names, score_names), 
+            ~run_regs_fml(
+              fml        = "n_children ~ age_flb_cat + {score_name}:age_flb_cat",
[…]

### 2020-05-20 make-data.R

@@ -174,6 +174,10 @@ edit_famhist <- function (famhist, score_names) {
   famhist$year_flb <- famhist$YOB + famhist$age_flb
   famhist$year_llb <- famhist$YOB + famhist$age_llb
   
+  famhist$flb_cat <- santoku::fillet(famhist$age_flb, c(13, 20, 23, 26, 30, 47))
+  famhist$flb_cat %<>% forcats::fct_expand("No children")
+  famhist$flb_cat[famhist$sex == 0 &  famhist$n_children == 0] <- "No children"
+  
   famhist$urbrur <- car::recode(famhist$f.20118.0.0,
           "c(1, 5, 11, 12) = 'urban';
           c(2, 6, 13, 14, 15)  = 'town';
@@ -245,13 +249,13 @@ make_ghs_subset <- function(ghs_file) {
       ethnic == 1, sex %in% 1:2, edage < age, age %in% c(40:71),
       chbnbm1 < age
     ) %>% 
-    select(sex, age, edage, weight06, ten1, llord, chbnbm1, edlev00) 
+    select(sex, age, edage, weight06, ten1, llord, chbnbm1, edlev00, chnbrnt) 
   
   ghs_subset$edlev00[ghs_subset$edlev00 == -9] <- 0 # "never went to school"
-  ghs_subset %<>% mutate(
-    across(c(edage, chbnbm1, ten1, llord, chbnbm1, edlev00), negative_to_na)
-  )
-  ghs_subset$chbnbm1[ghs_subset$chbnbm1 == 0] <- NA # "none", presumably no children
+  ghs_subset %<>% mutate(across(
+    c(edage, chbnbm1, ten1, llord, chbnbm1, edlev00, chnbrnt), 
+    negative_to_na
+  ))
   
   # create variables with the same meaning as in famhist
   # ghs: 1 male, 2 female; famhist: 0 female, 1 male
@@ -259,6 +263,7 @@ make_ghs_subset <- function(ghs_file) {
   ghs_subset$age_at_recruitment <- ghs_subset$age
   ghs_subset$age_fulltime_edu   <- ghs_subset$edage
   ghs_subset$age_flb            <- ghs_subset$chbnbm1
+
   # YearsEdu is 7, 10, 13, 15, 19 or 20. It maps from
   # edu_qual, which is:
   # 1	College or University degree
@@ -302,6 +307,14 @@ make_ghs_subset <- function(ghs_file) {
[…]

### 2020-05-20 regressions.R

@@ -7,7 +7,7 @@ suppressPackageStartupMessages({
 })
 
 
-run_regs <- function (dep_var, score_names, famhist, pcs) {
+run_regs_basic <- function (dep_var, score_names, famhist) {
   
   run_reg <- function (score_name) {
     fml <- as.formula(glue("{dep_var} ~ {score_name}"))
@@ -15,16 +15,17 @@ run_regs <- function (dep_var, score_names, famhist, pcs) {
   }
   basic_res <- map_dfr(score_names, run_reg, .id = "score_name")
   
-  famhist <- join_famhist_pcs(famhist, pcs)
-  pc_names <- grep("PC", names(pcs), value = TRUE)
   run_resid_reg <- function (score_name) {
-    fml_resid <- as.formula(glue("{dep_var} ~ {score_name} + {pc_names}"))
-    reg_resid <- tidy(lm(fml_resid, famhist), conf.int = TRUE)  
+    score_resid <- paste0(score_name, "_resid")
+    fml_resid <- as.formula(glue("{dep_var} ~ {score_resid}"))
+    reg_resid <- tidy(lm(fml_resid, famhist), conf.int = TRUE) 
+    reg_resid <- mutate(reg_resid, term = gsub("_resid", "", term))
+    reg_resid
   }
   resid_res <- map_dfr(score_names, run_resid_reg, .id = "score_name")
   
   res <- bind_rows(raw = basic_res, controlled = resid_res, .id = "reg.type")
-  res %<>% filter(term != "(Intercept)", ! grepl("^PC", term))
+  res %<>% filter(term != "(Intercept)")
   
   return(res)
 }

### 2020-05-21 why-negative-selection.Rmd
Reorganize paper. Add intro and appendix. Weighted regressions for parents. Correlations beween effect sizes and correlations with income/education.

--- 2020-05-19 why-negative-selection.Rmd
@@
 ---
-
-```md
-# TODO
-
-* Why so many missings for age_fulltime_edu?
-  - ah, need f.6138 for those who went to college
-  - but note that YearsEdu.ISCED has many fewer missings
-  - ask Abdel how it was calculated
-  - use this for survey weighting?
-
-* Look at geography, esp. in siblings regressions.
-
-* Work more on weighting the data?
-  - weight by age at first live birth
-  - education using ISCED?
-  - geography?
-  - ask for existing work by others?
-
-* control for age in n_partners regressions
-
-## Data to gather
-* f.2139 - age first had sex (includes "never had sex" which may explain
-  some of the many NAs for f.2141, num sex partners)
-
-```
 
@@
 options(digits = 2)
+options(dplyr.summarise.inform = FALSE) # shut up about summarize
 knitr::opts_chunk$set(
@@
         cache = FALSE, 
-        error = TRUE
+        error = TRUE,
+        fig.height = 3.5
       )
@@
 pval_palette <- c("FALSE" = "grey50", "TRUE" = "orange2")
+
+my_theme <- function (...) theme(axis.text.y = element_text(size = 7), ...)
 
@@
 
+# Introduction
+
+Existing work suggests that natural selection is taking place in modern populations,
+but that effect sizes are small. In this paper we investigate natural selection
+in the UK over the twentieth century, using data from UK Biobank. We use
+questions on number of children and number of siblings to reconstruct two
+generations of family size per respondent. We then relate these to polygenic
+scores for several different characteristics.
+
+Natural selection is taking place on most polygenic scores, and is broadly
+"negative" in that characteristics that most people would consider undesirable
+are being selected for. Effect sizes are larger for number of siblings than for
+number of children. This is not because the strength of natural selection has
+changed over time: effect sizes do not differ between early- and late-born
+respondents, within either children or siblings. Instead, the difference is
+probably due to ascertainment bias in UK Biobank. Indeed, natural selection is
+concentrated among population subgroups that are under-represented in the
+sample, including less-educated people and poorer households. When we weight 
+the sample to correct for ascertainment bias, many effect sizes are considerably 
+larger.
+
+Among mothers who were older at first birth, and among parents with fewer sexual
+partners, effect signs are reversed so that natural selection is broadly
+"positive". The effects of PGS can be decomposed into two channels: an effect on
+age at first live birth, and an opposite-signed effect on number of children
+conditional on age at first live birth. This pattern may be explained by
+economic theories of fertility. In these, higher potential earnings have two
+opposite effects on fertility: a fertility-increasing income effect (higher
+income makes children more affordable), and a fertility-lowering substitution
+effect (time spent on childrearing has a higher cost in foregone earnings).
+Among couples who can take advantage of coordination in childcare activities,
+the income effect dominates; among single parents or those in unstable
+relationships, the substitution effect dominates.
+
+# Open questions/TODO
+
+* If this theory is true then the pattern of differences is explained by
+  (potential) earnings. So the strength of the relationship of a PGS with 
+  number of children should be explained by the strength of its relationship
+  with earnings and education.
+* Even after weighting, effect sizes for siblings are still higher than for
+  children. Why?
+  - Remaining ascertainment bias?
+  - Siblings don't include those with 0 children?
+  - Something else?
+* If we weight with the GHS, and plot the means of figure 1 using this w
+  weighting, most things stay the same - but the change in EA is reversed! EA
+  increases over time. Could this be because the sample includes many
+  relatively educated older people? If so, then mean effects are *overestimated*
+  by figure 1.
+* Could be useful: f.2139 - age first had sex (includes "never had sex" 
+  which may explain some of the many NAs for f.2141, num sex partners)
+*  Need f.6138 to correctly calculate `age_fulltime_edu` including
+  those who went to college.
+* Look at geography, esp. in siblings regressions.
+* Work more on weighting the data
+  - check for existing work by others?
+  - deal with the fact that GHS 06 has its own weights and use
+    subset.svydesign to reweight?
+* control for age in n_partners regressions
+
 # Data
@@
 
+Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores by 5-year
+birth intervals, for the entire sample. Several scores show consistent increases
+or declines over this 30-year period, of the order of 5% of a standard deviation.
+
 ```{r calc-means-over-time}
@@
 drake::loadd(famhist)
+drake::loadd(ghs_weights)
+famhist <- left_join(famhist, ghs_weights, by = "f.eid")
 pgs_over_time <- famhist %>% 
@@
       summarize(across(all_of(score_names), mean, na.rm = TRUE)) %>% 
-      pivot_longer(
-        -YOB, 
-        names_to = "score_name",
-        values_to = "score"
-      )
-
-```
-
-```{r calc-means-over-time-children, eval = FALSE}
-pgs_over_time_children <- famhist %>% 
-      mutate(
-        YOB_kids = (year_flb + year_llb)/2,
-        YOB      = santoku::chop(
-                     YOB_kids, 
-                     brk_left(seq(1960, 1995, 5), close_end = FALSE), 
-                     seq(60, 90, 5),
-                     extend = FALSE
-                   )
-      ) %>% 
-      filter(! is.na(YOB)) %>% 
-      group_by(YOB) %>% 
-      summarize(across(all_of(score_names), 
-        ~weighted.mean(.x, n_children, na.rm = TRUE)
-      )) %>% 
       pivot_longer(-YOB, names_to = "score_name", values_to = "score")
-
-pgs_over_time <- bind_rows(
-        Respondents = pgs_over_time,
-        Children    = pgs_over_time_children,
-        .id         = "Generation"
-      )
 
@@
 
-We run regressions on two dependent variables:
+We run bivariate regressions on two dependent variables:
 
@@
 
-We run regressions both with and without controls for the 100 top principal
-components of the genetic data.
-
-```{r regs-data}
+```{r calc-regs-all}
 
@@
 res_all_wide <- res_all %>% 
+  filter(reg.type == "raw") %>% 
+  select(dep.var, term, estimate) %>% 
+  pivot_wider(names_from = dep.var, values_from = estimate)
+
+n_pgs  <- as.double(nrow(res_all_wide))
+
+prop_gens <- res_all_wide[["N siblings"]]/res_all_wide[["N children"]]
+median_prop_gens <- median(prop_gens[prop_gens > 0])
+```
+
+Figure \@ref(fig:plot-basic-regs) shows effect sizes of a one-standard
+deviation shift in each polygenic score.
+
+Estimates are broadly consistent across generations. However, effect sizes are 
+much larger for *siblings* than for *children* 
+regressions. Among consistently-signed estimates, the effect size for
+siblings is `r 100*median_prop_gens` per cent at the median.
+
+```{r plot-basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children."}
+
+drake::loadd(res_all)
+
+n_regs <- as.double(nrow(res_all))
+res_all %>% 
+      filter(reg.type == "raw") %>% 
+      mutate(
+        term    = pretty_names(term),
+        term    = fct_reorder(term, estimate, order_abs(1)),
+        dep.var = fct_inorder(dep.var) 
+      ) %>% 
+      ggplot(aes(y = term, x = estimate, fill = dep.var, 
+        colour = p.value < 0.05/{{n_regs}})) +
+      my_vline + 
+      pval_geom_point() +
+      my_labs() +
+      pval_scale() +
+      my_fill_scale(n = 2, direction = -1, guide = guide_legend(title = "")) +
+      my_theme()
+
+```
+
+# Subgroups
+
+We next examine how different subgroups of the population contribute to natural 
+selection. These analyses have two goals: to get a sense of how ascertainment bias
+could affect our basic results, and to learn more about the causes of natural 
+selection in the UK population. Because we have more data on respondents than on
+their parents, we focus on *children* regressions.
+
+## Sex, income and education
+
+Figure \@ref(fig:plot-sexes) shows effect sizes of PGS on number of children
+separately for males and females. Differences are particularly large for
+educational attainment, height and MDD. There is no overall pattern in
+differences of effect size, however.
+
+```{r plot-sexes, fig.cap = "Effect sizes on number of children by sex"}
+
+drake::loadd(res_sex)
+n_regs <- as.double(nrow(res_sex))
+
+res_sex %>% 
+      mutate(
+        term = pretty_names(term),
+        term = fct_reorder(term, estimate, order_abs(1))
+      ) %>% 
+      ggplot(aes(y = term, x = estimate, fill = sex, 
+            colour = p.value < 0.05/{{n_regs}})) +
+        my_vline +
+        pval_geom_point() +
+        pval_scale() +
+        my_fill_scale(n = 2, direction = -1) +
+        my_labs() +
+        my_theme()
+
+```
+
+Figure \@ref(fig:plot-educ-level) splits respondents up by education levels.
+Selection effects are typically larger and more significant for those who left
+school before 16.  Figure \@ref(fig:plot-income) splits respondents by household
+income category. A very similar pattern holds, with selection effects being
+larger for those in the poorest income category.
+
+```{r plot-educ-level, fig.cap = "Effect sizes on number of children by age left full-time education"}
+
+drake::loadd(res_edu)
+n_regs <- as.double(nrow(res_edu))
+res_edu %>% 
+      mutate(
+        term = pretty_names(term), 
+        term = fct_reorder(term, estimate, order_abs(1)),
+        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18")
+      ) %>% 
+      ggplot(aes(y = term, x = estimate, colour = p.value < 0.05/{{n_regs}}, 
+        fill = `Age left FTE`)) +
+        my_vline +
+        pval_geom_point() + 
+        pval_scale() +
+        my_fill_scale(n = 3, direction = -1) +
+        my_labs() +
+        my_theme()
+
+```
+
+```{r plot-income, fig.cap = "Effect sizes on number of children by household income"}
+
+drake::loadd(res_income)
+
+n_regs <- as.double(nrow(res_income))
+res_income %>% 
+      mutate(
+        term = pretty_names(term), 
+        term = fct_reorder(term, estimate, order_abs(1)),
+        Income = factor(income_cat, 
+          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
+        )
+      ) %>% 
+      ggplot(aes(y = term, x = estimate, fill = Income, 
+        colour = p.value < 0.05/{{n_regs}})) +
+        my_vline +
+        pval_geom_point(alpha = 0.8) + 
+        my_fill_scale(n = 5, direction = -1) +
+        pval_scale() +
+        my_labs() +
+        my_theme()
+```
+
+```{r calc-income-edu-controlled}
+drake::loadd(res_income_controlled, res_edu_controlled)
+n_regs_ic <- nrow(res_income_controlled)
+n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
+
+n_regs_educ <- nrow(res_edu_controlled)
+n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
+```
+
+These results could be driven by age, if older respondents are poorer and less
+educated, and also more subject to selection on polygenic scores. However, if we
+rerun the regressions, interacting the polygenic score with income category and
+also with a quadratic in age, the interaction with income remains significant at
+0.05/`r n_regs_ic` for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
+if we interact the PGS with age of leaving full time education and a quadratic
+in age, the interaction with age leaving FTE remains significant at 
+0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.
+
+The UK Biobank sample overrepresents highly educated people and high-income
+households. This suggests that statistics from the sample may underestimate
+the level of natural selection in the population.
+
+## Age at first live birth
+
+```{r calc-age-flb-cross}
+
+drake::loadd(res_age_flb_cross)
+
+cor_age_flb <- res_age_flb_cross %>% 
+      mutate(category = gsub(":.*", "", term)) %>% 
+      pivot_wider(score_name, names_from = category, values_from = estimate) %>%              {cor(.$`age_flb_cat10-22`, .$`age_flb_cat28-52`)}
+
+```
+
+Age at first live birth correlates with number of children ever born. Figure
+\@ref(fig:plot-age-flb-cross) shows *children* regressions estimated separately
+for each tercile of age at first live birth. Several effects are strikingly
+different across terciles. ADHD and MDD are selected for amongst the youngest
+third of mothers, but selected against among the oldest two-thirds. Educational
+attainment is selected for among the oldest two-thirds of mothers, but is not
+significantly selected among the youngest third. Similarly, several PGS of body
+measurements are selected against only among older mothers. The correlation
+between effect sizes for the youngest and oldest terciles is `r cor_age_flb`.
+This is the first sign that selection may work in different directions across
+different subgroups.
+
+Figure \@ref(fig:plot-n-partners) splits males and females by lifetime number of
+sexual partners, at the median value of 3. Again, selection effects are reversed
+across the groups. Among men and women with more than 3 sexual partners, for
+example, EA is selected against; among those with 3 or fewer sexual partners, it
+is selected for. Selection is broadly negative (positive) among those 
+with more (less) partners.
+
+```{r plot-age-flb-cross, fig.cap = "Effect sizes on number of children, by age at first live birth terciles"}
+
+drake::loadd(res_age_flb_cross)
+
+# this counts each cross term as a separate test:
+n_regs <- as.double(nrow(res_age_flb_cross))
+
+res_age_flb_cross %>% 
+      mutate(
+        score_name = pretty_names(score_name),
+        score_name = fct_reorder(score_name, estimate, order_abs(1)),
+        term       = gsub("age_flb_cat(.*):.*", "\\1", term)
+      ) %>% 
+      ggplot(aes(estimate, score_name, fill = term, 
+        colour = p.value < 0.05/{{n_regs}})) +
+        my_vline +
+        pval_geom_point() +
+        pval_scale() +
+        my_fill_scale(n = 3, direction = -1, 
+          guide = guide_legend(title = "Age at FLB")) +
+        my_labs() +
+        my_theme()
+
+```
+
+```{r plot-n-partners, fig.cap = "Effect sizes on number of children by lnumber of sexual partners", fig.height = 4}
+
+drake::loadd(res_partners)
+res_partners %<>%       
+      filter(term != "(Intercept)", term != "lo_partnersTRUE") %>% 
+      mutate(`N partners` = ifelse(grepl(":", term), "<= 3", "> 3"))
+
+n_regs <- as.double(nrow(res_partners))
+res_partners %>% 
+      mutate(
+        score_name = pretty_names(score_name),
+        score_name = fct_reorder(score_name, estimate, order_abs(1))
+      ) %>% 
+      ggplot(aes(y = score_name, x = estimate, fill = `N partners`, 
+        colour = p.value < 0.05/{{n_regs}})) + 
+        my_vline +
+        pval_geom_point() +
+        facet_wrap(vars(sex)) +
+        my_fill_scale(n = 2) +
+        pval_scale() +
+        my_theme(legend.position = "bottom") +
+        my_labs(x = "", y = "")
+
+```
+
+# Accounting for ascertainment bias
+
+```{r calc-res-weighted}
+
+drake::loadd(res_weighted)
+res_weighted_all <- res_weighted %>% 
+      left_join(
+        res_all %>% 
+              filter(dep.var == "N children", reg.type == "raw") %>% 
+              select(term, estimate),
+        by = "term",
+        suffix = c("_weighted", "_unweighted")
+      ) %>% 
+      mutate(
+        prop_wtd_unwtd = estimate_weighted/estimate_unweighted
+      )
+
+mean_prop_wtd_unwtd <- mean(res_weighted_all$prop_wtd_unwtd)
+
+```
+
+As a fix for ascertainment bias in the UK Biobank sample, we weight participants
+by sex, age (within the 40-71 age group), and education. Our sample weights are
+calibrated based on the 2006 General Household Survey. We then rerun the basic
+*children* regressions. Figure \@ref(fig:plot-weighted) shows the results, with
+unweighted estimates plotted for comparison. Weighting increases effect sizes on
+average by `r 100 * (mean_prop_wtd_unwtd - 1)` per cent.
+
+```{r plot-weighted, fig.cap = "Effects of polygenic scores on number of children, regressions weighted by education levels within age categories"}
+
+n_regs <- as.double(nrow(res_weighted_all))
+
+res_weighted_all %>% 
+      mutate(
+        term    = pretty_names(term),
+        term    = fct_reorder(term, estimate_weighted, order_abs(1))
+      ) %>% 
+      pivot_longer(cols = c(estimate_weighted, estimate_unweighted), 
+        names_to = "Regression", values_to = "estimate") %>% 
+      mutate(
+        Regression = sub("estimate_", "", Regression)
+      ) %>% 
+      ggplot(aes(y = term, x = estimate, shape = Regression,
+        colour = p.value < 0.05/{{n_regs}})) +
+      my_vline + 
+      geom_point(size = 1.5, stroke = 0.7, fill = "steelblue") +
+      labs(x = "Effect size", y = "") +
+      scale_shape_manual(values = c("weighted" = 22, "unweighted" = 21)) +
+      pval_scale(aesthetics = "colour") +
+      my_theme()
+
+```
+
+```{r plot-sibs-weighted, eval = FALSE}
+
+# what this shows is that even after weighting, sibling effect sizes are
+# still stronger. In fact weighting hasn't changed things much.
+drake::loadd(res_weighted, res_sibs_weighted)
+n_regs <- as.double(nrow(res_weighted))
+# bind_rows(children = res_weighted, sibs = res_sibs_weighted, .id = "Dep. Var.") %>% 
+res_all %>% 
+      filter(reg.type == "raw") %>% 
+      mutate(
+        term    = pretty_names(term),
+        term    = fct_reorder(term, estimate, order_abs(1))
+      ) %>% 
+      ggplot(aes(y = term, x = estimate, fill = `dep.var`,
+        colour = p.value < 0.05/{{n_regs}})) +
+      my_vline + 
+      pval_geom_point() +
+      labs(x = "Effect size", y = "") +
+      my_fill_scale(n = 2) +
+      pval_scale(aesthetics = "colour")
+
+```
+
+```{r calc-flb-weights}
+
+drake::loadd(res_flb_weights)
+drake::loadd(res_sex)
+res_flb_all <- left_join(
+        res_flb_weights, 
+        res_sex %>% filter(sex=="Female") %>% select(-subset, -sex), 
+        by = "term", 
+        suffix = c(".wt", ".nowt")
+      ) %>% 
+      mutate(prop_estimate = estimate.wt/estimate.nowt)
+
+prop_wt_nowt <- mean(res_flb_all$prop_estimate)
+pct_change_score <- function (score) {
+  100 * (res_flb_all$prop_estimate[res_flb_all$term == score] - 1)
+}
+```
+
+We also weighted data for women with children only, by age, education and age 
+at first live birth.  Figure \@ref(fig:plot-flb-weights) 
+shows the results. Weighting increased effect sizes on average by 
+`r 100 * (prop_wt_nowt - 1)` per cent. There 
+are large changes for some variables, including EA3 
+(`r pct_change_score("EA3_excl_23andMe_UK")` per cent), cognitive
+ability (`r pct_change_score("cognitive_ability")` per cent) and
+waist-hip ratio (WHR; `r pct_change_score("whr_combined")` percent).
+
+```{r plot-flb-weights, fig.cap = "Effect sizes on number of children, female respondents weighted by age at first live birth"}
+
+bind_rows(
+        Weighted   = res_flb_weights, 
+        Unweighted = res_sex %>% filter(sex == "Female") %>% select(-subset, -sex),
+        .id        = "Regression"
+      ) %>% 
+      mutate(
+        term = pretty_names(term),
+        term = fct_reorder(term, estimate, order_abs(1))
+      ) %>% 
+      ggplot(aes(estimate, term, shape = Regression, 
+        color = p.value < 0.05/66)) +
+        my_vline +
+        geom_point(size = 1.5, stroke = 0.7, fill = "steelblue") +
+        scale_shape_manual(values = c("Weighted" = 22, "Unweighted" = 21)) +  
+        pval_scale() +
+        my_labs() + 
+        my_theme()
+```
+
+These results suggest that previous work may have underestimated the level of
+selection occurring in the population. [XXX who?]
+
+```{r calc-corr-age-flb}
+
+drake::loadd(res_age_flb)
+drake::loadd(res_sex)
+
+res_sex_comparison <- res_sex %>% 
+      filter(sex == "Female") %>% 
+      mutate(score_name = term) %>% 
+      select(score_name, term, estimate:conf.high)
+
+res_age_flb %<>% filter(term != "age_flb")
+
+res_combined <- bind_rows(
+        raw      = res_sex_comparison, 
+        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
+        .id      = "reg.type"
+      ) %>% 
+      arrange(score_name)
+
+raw_flb_comparison <- res_combined %>% 
+      pivot_wider(
+        id_cols     = score_name, 
+        names_from  = reg.type, 
+        values_from = estimate
+      )
+
+cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
+opp_signed <- sum(
+        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
+      )
+
+```
+
+## Mechanisms behind natural selection
+
+In this section, we dig into the mechanisms causing natural selection in the
+population. In particular, why are selection effects *oppositely* signed in
+certain subgroups -- for example, in older and younger mothers (by age at first
+live birth) or in those with more and fewer lifetime sexual partners?
+
+A possible answer is given by the economic theory of fertility [XXX cite Becker 1960].
+According to this, increases in potential earned income affect fertility via two opposing channels. There is an "income effect" by which children become more
+affordable, like any other good. Since childrearing has a cost in time, there is
+also a substitution effect: the opportunity cost of childrearing increases if
+one's market wage is higher. The income effect would lead higher earners to
+have more children. The substitution effect would lead higher earners to have
+fewer children.
+
+Traditionally, it was assumed that since women
+spent more of their time on childcare, the substitution effect would dominate
+for women's earnings, while the income effect would dominate for men's earnings. 
+However, a more relevant split may be between those raising children alone, and
+those raising children in a two-parent household. The substitution effect is likely
+to be particularly strong for lone parents, since they have less opportunity
+to share childcare responsibilities.
+
+If so, then genetic characteristics which affect one's earnings potential in the
+labour market may lead to opposing effects on fertility. Genetic
+variants which improve one's earnings may increase fertility among couples, but
+decrease fertility among single parents.
+
+This explanation predicts that the strength of polygenic scores' effects on
+fertility will correlate with the strength of their relationship with earnings.
+We measure the correlation of each PGS with household income, as well as with
+education level, a phenotypic variable which is likely to predict earnings. We
+then correlate these measurements with effect sizes on number of children.
+Across scores, there is a clear negative correlation between effect on
+fertility, and link to both income and education.
+
+```{r plot-cors-income-educ, fig.cap="Effect sizes on number of children and correlations with income and education"}
+
+drake::loadd(res_all)
+cors_income <- cor(famhist[c(score_names)], famhist$income_cat, use = "pairwise")
+# TODO: use age_left_edu when this includes college
+cors_educ <-   cor(famhist[c(score_names)], famhist$edu_qual, use = "pairwise")
+
+effect_size <- res_all %>% 
+      filter(dep.var == "N children", reg.type == "raw") %>% 
+      pull(estimate)
+
+dfr <- data.frame(
+      type = rep(c("Cor. with income", "Cor. with educ"), each = 33),
+      cor = c(cors_income[,1], cors_educ[,1]),
+      effect_size = rep(effect_size, 2)
+)
+
+ggplot(dfr, aes(effect_size, cor, colour = type)) + 
+      geom_point() +
+      facet_wrap(vars(type)) + 
+      my_vline + 
+      my_hline +
+      labs(x = "Effect on n children", y = "Correlation") +
+      scale_colour_manual(values = c("darkblue", "darkgreen")) +
+      theme(legend.position = "none")
+
+```
+
+# Bits to integrate
+
+Figure \@ref(fig:plot-age-flb) shows the results of *children* regressions for 
+women only,
+controlling for age at first live birth. Effect sizes are greatly reduced. Partly,
+this is because the regressions exclude childless women. However,
+in `r opp_signed` out of `r length(score_names)` cases, effect sizes actually
+change sign with the controls. The correlation between effect sizes
+controlling for age at first live birth, and raw effect sizes, is `r cor_raw_flb`.
+```{r plot-age-flb, fig.cap = "Effect sizes on number of children, controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}
+
+n_regs <- as.double(nrow(res_age_flb))
+
+res_combined %>% 
+      mutate(
+        score_name = fct_reorder(score_name, estimate, order_abs(2)),
+        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for age FLB")
+      ) %>% 
+      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
+        color = p.value < 0.05/{{n_regs}})) + 
+        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
+        pval_geom_point() +
+        pval_scale() +
+        scale_fill_brewer(type = "qual", palette = 3) +
+        my_labs() +
+        my_theme()
+```
+
+```{r calc-age-birth-parents}
+
+drake::loadd(res_age_birth_parents)
+drake::loadd(res_all)
+
+res_sibs_raw <- filter(res_all, dep.var == "N siblings", reg.type == "raw")
+compare_abp_raw <- res_age_birth_parents %>% 
+      filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
+      left_join(res_sibs_raw, by = "term")
+
+cor_raw_fath <- with(
+  compare_abp_raw %>% filter(control == "fath_age_birth"), 
+  cor(estimate.x, estimate.y))
+cor_raw_moth <- with(
+  compare_abp_raw %>% filter(control == "moth_age_birth"), 
+  cor(estimate.x, estimate.y))
+```
+
+We can run similar regressions for the parents' generation, using the subsets of respondents who reported their mother's or father's age
+and who had no elder siblings. We run *sibling* regressions on these subsets,
+controlling for either parent's age at their birth.
+Figure \@ref(fig:plot-age-birth-parents) shows
+the results. Effect sizes are very similar, whether controlling for
+father's or mother's age at respondent's birth or mother's age at respondent's birth. Unlike for the respondents' generation, effect sizes are positively correlated with
+the effect sizes from bivariate regressions (father's age at birth: 
+$\rho$ `r cor_raw_fath`; mother's age at birth: $\rho$ `r cor_raw_moth`).
+
+```{r plot-age-birth-parents, fig.cap = "Effect sizes on number of siblings controlling for parents' age at birth, eldest siblings"}
+
+n_regs <- as.double(nrow(res_age_birth_parents)/2)
+
+res_age_birth_parents %>% 
+    filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
+    mutate(
+      term    = pretty_names(term),
+      term    = fct_reorder(term, estimate, order_abs(2)),
+      Control = ifelse(control == "fath_age_birth", 
+                  "Father's age at birth", 
+                  "Mother's age at birth"
+                )
+    ) %>% 
+    ggplot(aes(y = term, x = estimate, fill = Control, 
+      colour = p.value < 0.05/{{n_regs}})) + 
+      pval_geom_point(alpha = 0.8) +  
+      my_vline +
+      my_fill_scale(2) +
+      pval_scale() +
+      my_labs()
+
+```
+
+```{r calc-age-flb-dv}
+
+drake::loadd(res_age_flb_dv)
+drake::loadd(res_age_birth_parents_dv)
+
+compare_flb_dv <- left_join(
+        res_age_flb_dv           %>% select(term, estimate),
+        res_age_birth_parents_dv %>% select(term, estimate, dep.var),
+        by = "term"
+      )
+
+cor_flb_fath <- with(compare_flb_dv %>% filter(dep.var == "fath_age_birth"),
+        cor(estimate.x, estimate.y)
+      )
+cor_flb_moth <- with(compare_flb_dv %>% filter(dep.var == "moth_age_birth"),
+        cor(estimate.x, estimate.y)
+      )
+```
+
+These results suggest that polygenic scores may directly correlate with
+age at first live birth. Figure \@ref(fig:plot-age-flb-dv) plots
+estimated effect sizes from bivariate regressions for respondents, and
+Figure \@ref(fig:plot-age-birth-parents-dv) does the same for their parents. 
+Effect sizes are reasonably large. They are also very highly correlated across 
+generations. Effect sizes of PGS on father's age at own birth, and on own age
+at first live birth, have a correlation of `r cor_flb_fath`; for mother's age
+and own age it is `r cor_flb_moth`
+
+```{r plot-age-flb-dv, fig.cap = "Effects of polygenic scores on age at first live birth"}
+
+n_regs <- as.double(nrow(res_age_flb_dv))
+
+res_age_flb_dv %>% 
+      mutate(
+        term = pretty_names(term),
+        term = fct_reorder(term, estimate, order_abs(1))
+      ) %>% 
+      ggplot(aes(estimate, term, colour = p.value < 0.05 / {{n_regs}})) + 
+      my_vline +
+      pval_geom_point(fill = "steelblue4") +
+      pval_scale() + 
+      my_labs() +
+      my_theme()
+
+```
+
+```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at own birth, eldest siblings"}
+
+n_regs <- as.double(nrow(res_age_birth_parents_dv))
+
+res_age_birth_parents_dv %>% 
+      mutate(
+        term = pretty_names(term),
+        term = fct_reorder(term, estimate, order_abs(1)),
+        "Dependent variable" = dplyr::recode(dep.var, 
+          "fath_age_birth" = "Father's age", 
+          "moth_age_birth" = "Mother's age"
+        )
+      ) %>% 
+      ggplot(aes(estimate, term, fill = `Dependent variable`, 
+        colour = p.value < 0.05 / {{n_regs}})) + 
+        my_vline +
+        pval_geom_point() +
+        pval_scale() +
+        my_fill_scale(n = 2, direction = -1) +
+        my_labs()
+
+```
+
+# Appendix
+
+## Controlling for principal components
+
+Polygenic scores could capture effects that are really due to population 
+stratification, although would not change our results for natural selection of
+the scores. In \@ref(fig:plot-regs-controlled) we show results for selection on
+polygenic scores residualized for the top 100 principal components of the
+genetic data, calculated within the UK Biobank population. (XXX Abdel details.)
+
+```{r calc-regs-controlled}
+
+drake::loadd(res_all)
+
+res_all_controlled <- res_all %>% 
   select(reg.type, dep.var, term, estimate) %>% 
@@
   mutate(
-    consistent = sign(`raw_N siblings`) == sign(`controlled_N siblings`) &
-      sign(`raw_N siblings`) == sign(`raw_N children`) &
+    consistent = sign(`raw_N siblings`) == sign(`raw_N children`) &
+                 sign(`raw_N siblings`) == sign(`controlled_N siblings`) &
                  sign(`raw_N siblings`) == sign(`controlled_N children`)
@@
 
-n_pgs  <- as.double(nrow(res_all_wide))
-prop_consistent <- mean(res_all_wide$consistent)
-controlled_smaller_sibs <- abs(res_all_wide[["controlled_N siblings"]]) < 
-  abs(res_all_wide[["raw_N siblings"]])
-controlled_smaller_children <- abs(res_all_wide[["controlled_N children"]]) < 
-  abs(res_all_wide[["raw_N children"]])
-median_prop_sibs <- median(abs(res_all_wide[["controlled_N siblings"]]) /
-    abs(res_all_wide[["raw_N siblings"]]))
-median_prop_children <- median(abs(res_all_wide[["controlled_N children"]]) /
-    abs(res_all_wide[["raw_N children"]]))
-
-prop_gens_raw <- res_all_wide[["raw_N children"]]/res_all_wide[["raw_N siblings"]]
-prop_gens_controlled <- res_all_wide[["controlled_N children"]]/res_all_wide[["controlled_N siblings"]]
-median_prop_gens_raw <- median(prop_gens_raw[prop_gens_raw > 0])
-median_prop_gens_controlled <- median(prop_gens_controlled[prop_gens_controlled > 0])
-```
-
-Figure \@ref(fig:plot-basic-regs) shows effect sizes of a one-standard
-deviation shift in each polygenic score.
-
-Estimates are broadly consistent across generations. For 
-`r sum(res_all_wide$consistent)` out of `r n_pgs` polygenic scores, all 4 
-estimates have the same sign.
-
-However, effect sizes are much smaller for *children* than *siblings*
-regressions. Among consistently-signed estimates, the median effect size for
-children as a proportion of the effect size for  siblings is 
-`r median_prop_gens_raw`, or `r median_prop_gens_controlled` with controls.
-
-In siblings regressions, effect sizes are smaller when controlling for
+n_pgs  <- as.double(nrow(res_all_controlled))
+prop_consistent_controlled <- mean(res_all_controlled$consistent)
+controlled_smaller_sibs <- abs(res_all_controlled[["controlled_N siblings"]]) < 
+  abs(res_all_controlled[["raw_N siblings"]])
+controlled_smaller_children <- abs(res_all_controlled[["controlled_N children"]]) < 
+  abs(res_all_controlled[["raw_N children"]])
+median_prop_sibs <- median(abs(res_all_controlled[["controlled_N siblings"]]) /
+    abs(res_all_controlled[["raw_N siblings"]]))
+median_prop_children <- median(abs(res_all_controlled[["controlled_N children"]]) /
+    abs(res_all_controlled[["raw_N children"]]))
+
+```
+
+In siblings regressions, effect sizes are smaller when residualizing for
 principal components -- sometimes much smaller, as in the case of height. 
@@
 
-```{r plot-basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children.", fig.height = 7}
+Overall, `r prop_consistent_controlled * 100` per cent of effect sizes are
+consistently signed across all four regressions (on children and siblings, and
+with and without residualization).
+
+```{r plot-regs-controlled, fig.cap = "Effects of residualized polygenic scores on number of siblings/children.", fig.height = 7}
 
@@
       my_vline + 
-      pval_geom_point() +
-      labs(x = "Effect size", y = "") +
-      facet_wrap(vars(dep.var), ncol = 2) +
-      pval_scale() +
-      my_fill_scale(n = 2, guide = guide_legend(title = "")) +
-      theme(
-        legend.position = "bottom", 
-        legend.margin   = margin(2, 2, 2, 2)
-      )
+      facet_wrap(vars(dep.var)) +
+      pval_geom_point(alpha = .7) +
+      my_labs() +
+      pval_scale() +
+      my_fill_scale(n = 2, guide = guide_legend(title = ""))
 
@@
 
-# Selection over time
-
-```{r period-regs-data}
+## Selection over time
+
+```{r calc-period-regs}
 
@@
 
-Negative selection seems to decrease over time. Figure \@ref(fig:plot-by-period) shows
-effect sizes for *number of siblings* and *number of children*, median-split by
-parents' year of birth and own year of birth respectively.  Parents' year of
-birth is imputed, which is likely to produce some bias.
-
-```{r plot-by-period, fig.cap = "Effect sizes of PGS on number of children/siblings by own/parents' year of birth. PGS with the largest mean effect sizes are shown.", fig.height = 7, fig.width = 6}
-
-max_abs <- function (x) max(abs(x))
-res_period %>% 
-      mutate(
-        dep.var       = ifelse(children, "Children", "Siblings"),
-        period        = (period == "late") + children,
-        score_name    = pretty_names(score_name),
-        score_name    = fct_reorder(score_name, estimate, max_abs, 
-                          .desc = TRUE),
-        p             = diff.p.value
-      ) %>% 
-      arrange(score_name) %>% 
-      slice(1:48) %>% 
-ggplot(aes(period, estimate, colour = dep.var, group = dep.var)) + 
-      geom_hline(yintercept = 0, colour = "grey30") + 
-      geom_point(size = 1.5) +
-      geom_line(aes(linetype = p < 0.05/{{n_tests}})) +
-      facet_wrap(vars(score_name), nrow = 4) + 
-      pval_scale() +
-      my_fill_scale(n = 2, aesthetics = "colour") +
-      scale_linetype_manual(values = c("FALSE" = "dashed", "TRUE" = "solid"))
-
-```
-
-By definition, the sibling regressions exclude members of the parents'
-generation who had no children. This is likely to bias results towards zero, since
-much of the effect in children regressions is due to respondents with high scores
-being more likely to have no children. So, we cannot directly compare 
-effect sizes for the two sets of regressions. Within the sibling regressions,
-the most common pattern is that negative effects shrink in absolute size 
-(Table \@ref(tab:siblings-by-period)).
-
-```{r siblings-by-period}
+To check whether effect sizes were changing over time, we ran regressions 
+interacting PGS with birth year, median split at 1950. 
+Tables \@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)
+summarize the results. Very few scores are significantly different. Only
+one score, EA3, changes significantly across time in both generations: the
+absolute size of the (negative) effect significantly increases in sibling 
+regressions, and significantly decreases in children regressions.
+
+```{r tbl-siblings-by-period}
 
@@
       add_footnote(glue("Significance is measured at p < 0.05/{n_tests}"), 
-        border = 0.4) %>% 
+        border = 0.4, font_size = 8) %>% 
       set_contents(1, everywhere, 
@@
 
-In children regressions, no clear pattern is visible 
-(Table \@ref(tab:children-by-period)).
-
-```{r children-by-period}
+```{r tbl-children-by-period}
 summary_period %>% 
@@
 
-```{r res-weighted}
-
-drake::loadd(res_weighted)
-res_weighted_all <- res_weighted %>% 
-      left_join(
-        res_all %>% 
-              filter(dep.var == "N children", reg.type == "raw") %>% 
-              select(term, estimate),
-        by = "term",
-        suffix = c("_weighted", "_unweighted")
-      ) %>% 
-      mutate(
-        prop_wtd_unwtd = estimate_weighted/estimate_unweighted
-      )
-
-mean_prop_wtd_unwtd <- res_weighted_all %>% 
-      filter(sign(estimate_weighted) == sign(estimate_unweighted)) %>% 
-      summarize(mean = mean(prop_wtd_unwtd)) %>% 
-      pull(mean)
-
-```
-
-# Accounting for ascertainment bias
-
-Effect sizes tend to be smaller for children regressions. This
-could be caused by ascertainment bias in the UK Biobank sample -- e.g.,
-if respondents themselves are a more selected sample than the respondents' 
-parents. To check this, we weighted UK Biobank participants by age of leaving
-full time education. Our sample weights are based on the 2006 General Household
-Survey, calculating proportions within age and sex cells among White British
-respondents. We then rerun the basic *children* regressions.
-
-Figure \@ref(fig:plot-weighted) shows the results, with unweighted estimates
-plotted for comparison. Weighting increases effect sizes on average by 
-`r 100 * (mean_prop_wtd_unwtd - 1)` per cent. These weights are only basic 
-attempts to correct for ascertainment bias. Effect sizes might be increased 
-further by more precise weighting.
-
-```{r plot-weighted, fig.cap = "Effects of polygenic scores on number of children, regressions weighted by education levels within age categories"}
-
-n_regs <- as.double(nrow(res_weighted_all))
-
-res_weighted_all %>% 
-      mutate(
-        term    = pretty_names(term),
-        term    = fct_reorder(term, estimate_weighted, order_abs(1))
-      ) %>% 
-      pivot_longer(cols = c(estimate_weighted, estimate_unweighted), 
-        names_to = "Regression", values_to = "estimate") %>% 
-      mutate(
-        Regression = sub("estimate_", "", Regression)
-      ) %>% 
-      ggplot(aes(y = term, x = estimate, fill = Regression,
-        colour = p.value < 0.05/{{n_regs}})) +
-      my_vline + 
-      pval_geom_point() +
-      labs(x = "Effect size", y = "") +
-      my_fill_scale(n = 2) +
-      pval_scale(aesthetics = "colour")
-
-```
-
-# Causality
-
-```{r pgs-corrs}
-cor_tbl <- cor(famhist[score_names], use = "pair") %>% 
-      as_tibble() %>% 
-      mutate(score_name = score_names, .before = 1) %>% 
-      pivot_longer(cols = -score_name, values_to = "Correlation") %>% 
-      filter(score_name < name) %>% 
-      arrange(desc(Correlation)) %>% 
-      slice(1:10) %>% 
-      mutate(across(c(score_name, name), pretty_names))
-
-as_hux(cor_tbl, add_colnames = TRUE) %>% 
-      set_contents(1, 1:2, c("PGS", "PGS")) %>% 
-      theme_basic(header_col = FALSE) %>% 
-      set_bottom_border(1, everywhere, 0.4) %>% 
-      set_bottom_border(final(1), everywhere, 0.4) %>% 
-      set_caption("Top 10 correlations between polygenic scores")
-
-```
-
-Different polygenic scores are correlated. Table \@ref(tab:pgs-corrs) shows the top
-correlations in the sample. Because of this, bivariate correlations between
-PGS and number of children might be driven by other genetic scores. To explore which polygenic scores are driving negative selection, we run a single
-omnibus regression of *number of children* on all the PGS. We exclude EA2, 
-waist-hip ratio, waist-circumference, and "Hip combined" since they are 
-highly correlated with other scores, which could make our estimates unstable. Figure \@ref(fig:plot-together) shows the results. Interestingly, several PGS remain independently significant, although
-effect sizes are reduced.
-
-```{r plot-together, fig.cap = "Partial correlations with number of children"}
-
-loadd(res_together)
-n_regs <- as.double(nrow(res_together))
-res_together %>% 
-      mutate(
-        term = pretty_names(term),
-        term = fct_reorder(term, estimate, order_abs(2)),
-        dep.var = dplyr::recode(dep.var, 
-          "n_children" = "N Children", "n_sibs" = "N Siblings"),
-        dep.var = fct_relevel(dep.var, "N Siblings")
-      ) %>% 
-      ggplot(aes(y = term, x = estimate, color = p.value < 0.05/{{n_regs}}, 
-        fill  = PCs)) + 
-        pval_geom_point(alpha = 0.8) +
-        my_vline +
-        facet_wrap(vars(dep.var)) +
-        pval_scale() +
-        my_fill_scale(n = 2, direction = -1) +
-        my_labs() +
-        theme(legend.position = "bottom")
-```
-
-# Subgroups
-
-We next examine how different subgroups contribute to natural selection.
-
-## Males and females
-
-```{r}
-
-drake::loadd(res_sex)
-
-female_lower <- res_sex %>% 
-      pivot_wider(id_cols = c(term), names_from = sex, 
-        values_from = c(estimate, conf.low, conf.high)) %>% 
-      mutate(female_lower = estimate_Female < estimate_Male) %>% 
-      pull(female_lower)
-```
-
-Figure \@ref(fig:plot-sexes) shows effect sizes of PGS on number of children
-separately for males and females. For `r sum(female_lower)` out of
-`r length(female_lower)` PGS, selection is more negative for women
-than for men. Differences are particularly large for educational attainment and height PGS.
-
-```{r plot-sexes, fig.cap = "Effect sizes on number of children by sex"}
-
-n_regs <- as.double(nrow(res_sex))
-res_sex %>% 
-      mutate(
-        term = pretty_names(term),
-        term = fct_reorder(term, estimate, order_abs(1))
-      ) %>% 
-      ggplot(aes(y = term, x = estimate, fill = sex, 
-            colour = p.value < 0.05/{{n_regs}})) +
-        my_vline +
-        pval_geom_point() +
-        pval_scale() +
-        my_fill_scale(n = 2, direction = -1) +
-        my_labs()
-
-```
-
-## Age at first live birth
-
-```{r corr-age-flb}
-
-drake::loadd(res_age_flb)
-drake::loadd(res_sex)
-
-res_sex_comparison <- res_sex %>% 
-      filter(sex == "Female") %>% 
-      mutate(score_name = term) %>% 
-      select(score_name, term, estimate:conf.high)
-
-res_age_flb %<>% filter(term != "age_flb")
-
-res_combined <- bind_rows(
-        raw      = res_sex_comparison, 
-        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
-        .id      = "reg.type"
-      ) %>% 
-      arrange(score_name)
-
-raw_flb_comparison <- res_combined %>% 
-      pivot_wider(
-        id_cols     = score_name, 
-        names_from  = reg.type, 
-        values_from = estimate
-      )
-
-cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
-opp_signed <- sum(
-        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
-      )
-
-```
-
-Figure \@ref(fig:plot-age-flb) shows the results of *children* regressions for 
-women only,
-controlling for age at first live birth. Effect sizes are greatly reduced.
-In `r opp_signed` out of `r length(score_names)` cases, they are of the 
-opposite sign. The correlation between effect sizes
-controlling for age at first live birth, and raw effect sizes, is `r cor_raw_flb`.
-
-```{r plot-age-flb, fig.cap = "Effect sizes on number of children, controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}
-
-n_regs <- as.double(nrow(res_age_flb))
-
-res_combined %>% 
-      mutate(
-        score_name = fct_reorder(score_name, estimate, order_abs(2)),
-        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for age FLB")
-      ) %>% 
-      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
-        color = p.value < 0.05/{{n_regs}})) + 
-        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
-        pval_geom_point() +
-        pval_scale() +
-        scale_fill_brewer(type = "qual", palette = 3) +
-        my_labs()
-```
-
-```{r calc-age-birth-parents}
-
-drake::loadd(res_age_birth_parents)
-drake::loadd(res_all)
-
-res_sibs_raw <- filter(res_all, dep.var == "N siblings", reg.type == "raw")
-compare_abp_raw <- res_age_birth_parents %>% 
-      filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
-      left_join(res_sibs_raw, by = "term")
-
-cor_raw_fath <- with(
-  compare_abp_raw %>% filter(control == "fath_age_birth"), 
-  cor(estimate.x, estimate.y))
-cor_raw_moth <- with(
-  compare_abp_raw %>% filter(control == "moth_age_birth"), 
-  cor(estimate.x, estimate.y))
-```
-
-We can run similar regressions for the parents' generation, using the subsets of respondents who reported their mother's or father's age
-and who had no elder siblings. We run *sibling* regressions on these subsets,
-controlling for either parent's age at their birth.
-Figure \@ref(fig:plot-age-birth-parents) shows
-the results. Effect sizes are very similar, whether controlling for
-father's or mother's age at respondent's birth or mother's age at respondent's birth. Unlike for the respondents' generation, effect sizes are positively correlated with
-the effect sizes from bivariate regressions (father's age at birth: 
-$\rho$ `r cor_raw_fath`; mother's age at birth: $\rho$ `r cor_raw_moth`).
-
-```{r plot-age-birth-parents, fig.cap = "Effect sizes on number of siblings controlling for parents' age at birth, eldest siblings"}
-
-n_regs <- as.double(nrow(res_age_birth_parents)/2)
-
-res_age_birth_parents %>% 
-    filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
-    mutate(
-      term    = pretty_names(term),
-      term    = fct_reorder(term, estimate, order_abs(2)),
-      Control = ifelse(control == "fath_age_birth", 
-                  "Father's age at birth", 
-                  "Mother's age at birth"
-                )
-    ) %>% 
-    ggplot(aes(y = term, x = estimate, fill = Control, 
-      colour = p.value < 0.05/{{n_regs}})) + 
-      pval_geom_point(alpha = 0.8) +  
-      my_vline +
-      my_fill_scale(2) +
-      pval_scale() +
-      my_labs()
-
-```
-
-```{r calc-age-flb-dv}
-
-drake::loadd(res_age_flb_dv)
-drake::loadd(res_age_birth_parents_dv)
-
-compare_flb_dv <- left_join(
-        res_age_flb_dv           %>% select(term, estimate),
-        res_age_birth_parents_dv %>% select(term, estimate, dep.var),
-        by = "term"
-      )
-
-cor_flb_fath <- with(compare_flb_dv %>% filter(dep.var == "fath_age_birth"),
-        cor(estimate.x, estimate.y)
-      )
-cor_flb_moth <- with(compare_flb_dv %>% filter(dep.var == "moth_age_birth"),
-        cor(estimate.x, estimate.y)
-      )
-```
-
-These results suggest that polygenic scores may directly correlate with
-age at first live birth. Figure \@ref(fig:plot-age-flb-dv) plots
-estimated effect sizes from bivariate regressions for respondents, and
-Figure \@ref(fig:plot-age-birth-parents-dv) does the same for their parents. 
-Effect sizes are reasonably large. They are also very highly correlated across 
-generations. Effect sizes of PGS on father's age at own birth, and on own age
-at first live birth, have a correlation of `r cor_flb_fath`; for mother's age
-and own age it is `r cor_flb_moth`
-
-```{r plot-age-flb-dv, fig.cap = "Effects of polygenic scores on age at first live birth"}
-
-n_regs <- as.double(nrow(res_age_flb_dv))
-
-res_age_flb_dv %>% 
-      mutate(
-        term = pretty_names(term),
-        term = fct_reorder(term, estimate, order_abs(1))
-      ) %>% 
-      ggplot(aes(estimate, term, colour = p.value < 0.05 / {{n_regs}})) + 
-      my_vline +
-      pval_geom_point(fill = "steelblue4") +
-      pval_scale() + 
-      my_labs()
-
-```
-
-```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at own birth, eldest siblings"}
-
-n_regs <- as.double(nrow(res_age_birth_parents_dv))
-
-res_age_birth_parents_dv %>% 
-      mutate(
-        term = pretty_names(term),
-        term = fct_reorder(term, estimate, order_abs(1)),
-        "Dependent variable" = dplyr::recode(dep.var, 
-          "fath_age_birth" = "Father's age", 
-          "moth_age_birth" = "Mother's age"
-        )
-      ) %>% 
-      ggplot(aes(estimate, term, fill = `Dependent variable`, 
-        colour = p.value < 0.05 / {{n_regs}})) + 
-        my_vline +
-        pval_geom_point() +
-        pval_scale() +
-        my_fill_scale(n = 2, direction = -1) +
-        my_labs()
-
-```
-
-## Number of sexual partners
-
-Figure \@ref(fig:plot-n-partners) splits males and females up by lifetime number
-of sexual partners. Remarkably, across both sexes, negative selection is
-strongly reversed for respondents who had 3 or fewer sexual partners in their
-lifetime.
-
-```{r plot-n-partners, fig.cap = "Effect sizes on number of children by number of sexual partners", fig.height = 7, fig.width = 6}
-
-drake::loadd(res_partners)
-res_partners %<>%       
-      filter(term != "(Intercept)", term != "lo_partnersTRUE") %>% 
-      mutate(`N partners` = ifelse(grepl(":", term), "<= 3", "> 3"))
-
-n_regs <- as.double(nrow(res_partners))
-res_partners %>% 
-      mutate(
-        score_name = pretty_names(score_name),
-        score_name = fct_reorder(score_name, estimate, order_abs(1))
-      ) %>% 
-      ggplot(aes(y = score_name, x = estimate, fill = `N partners`, 
-        colour = p.value < 0.05/{{n_regs}})) + 
-        my_vline +
-        pval_geom_point() +
-        facet_wrap(vars(sex)) +
-        my_fill_scale(n = 2) +
-        pval_scale() +
-        theme(legend.position = "bottom") +
-        my_labs(x = "", y = "")
-
-```
-
-## Education and income
-
-Figure \@ref(fig:plot-educ-levels) splits respondents up by education levels.
-Both negative and positive selection are typically larger and more significant
-for those who left school before 16. Table \@ref(tab:educ-levels) summarizes
-the results.
-
-```{r educ-levels}
-loadd(res_edu)
-res_edu %>% 
-      mutate(
-        age_fte_cat = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18")
-      ) %>% 
-      group_by(age_fte_cat) %>% 
-      summarize(
-        prop_sig = 100 * mean(p.value < 0.05/nrow(res_edu)), 
-        .groups  = "drop"
-      ) %>% 
-      as_huxtable(add_colnames = TRUE) %>% 
-      set_contents(1, everywhere, 
-        c("Age left FTE", "% PGS significant")) %>%
-      set_caption("Negative selection by education level") %>% 
-      set_bold(1, everywhere, TRUE) %>% 
-      set_bottom_border(1, everywhere, 0.4) %>% 
-      set_left_padding(everywhere, 1, 0) %>% 
-      set_right_padding(everywhere, final(1), 0) %>% 
-      set_top_padding(14)
-
-```
-
-```{r plot-educ-levels, fig.cap = "Effect sizes on number of children by age left full-time education"}
-
-drake::loadd(res_edu)
-n_regs <- as.double(nrow(res_edu))
-res_edu %>% 
-      mutate(
-        term = pretty_names(term), 
-        term = fct_reorder(term, estimate, order_abs(1)),
-        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18")
-      ) %>% 
-      ggplot(aes(y = term, x = estimate, colour = p.value < 0.05/{{n_regs}}, 
-        fill = `Age left FTE`)) +
-        my_vline +
-        pval_geom_point() + 
-        pval_scale() +
-        my_fill_scale(n = 3, direction = -1) +
-        my_labs()
-
-```
-
-Figure \@ref(fig:plot-income) splits respondents by household income category.
-A very similar pattern holds, with selection effects being larger for those
-in the poorest income category.
-
-```{r plot-income, fig.cap = "Effect sizes on number of children by household income"}
-
-drake::loadd(res_income)
-
-n_regs <- as.double(nrow(res_income))
-res_income %>% 
-      mutate(
-        term = pretty_names(term), 
-        term = fct_reorder(term, estimate, order_abs(1)),
-        Income = factor(income_cat, 
-          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
-        )
-      ) %>% 
-      ggplot(aes(y = term, x = estimate, fill = Income, 
-        colour = p.value < 0.05/{{n_regs}})) +
-        my_vline +
-        pval_geom_point(alpha = 0.8) + 
-        my_fill_scale(n = 5, direction = -1) +
-        pval_scale() +
-        my_labs()
-```
-
-```{r income-edu-controlled}
-drake::loadd(res_income_controlled, res_edu_controlled)
-n_regs_ic <- nrow(res_income_controlled)
-n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
-
-n_regs_educ <- nrow(res_edu_controlled)
-n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
-```
-
-These results could be driven by age, if older respondents are poorer and less
-educated, and also more subject to selection on polygenic scores. However, if we
-rerun the regressions, interacting the polygenic score with income category and
-also with a quadratic in age, the interaction with income remains significant at
-0.05/`r n_regs_ic` for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
-if we interact the PGS with age of leaving full time education and a quadratic
-in age, the interaction with age leaving FTE remains significant at 
-0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.
-
-# Number of children
+## Causality
+
+## Number of children
 
@@
 ```
+
+```{r tbl-pgs-corrs}
+cor_tbl <- cor(famhist[score_names], use = "pair") %>% 
+      as_tibble() %>% 
+      mutate(score_name = score_names, .before = 1) %>% 
+      pivot_longer(cols = -score_name, values_to = "Correlation") %>% 
+      filter(score_name < name) %>% 
+      arrange(desc(Correlation)) %>% 
+      slice(1:10) %>% 
+      mutate(across(c(score_name, name), pretty_names))
+
+as_hux(cor_tbl, add_colnames = TRUE) %>% 
+      set_contents(1, 1:2, c("PGS", "PGS")) %>% 
+      theme_basic(header_col = FALSE) %>% 
+      set_bottom_border(1, everywhere, 0.4) %>% 
+      set_bottom_border(final(1), everywhere, 0.4) %>% 
+      set_tb_padding(4) %>% 
+      set_caption("Top 10 correlations between polygenic scores")
+
+```
+
+Different polygenic scores are correlated. Table \@ref(tab:tbl-pgs-corrs) shows the top
+correlations in the sample. Because of this, bivariate correlations between
+PGS and number of children might be driven by other genetic scores. To explore which polygenic scores are driving negative selection, we run a single
+omnibus regression of *number of children* on all the PGS. We exclude EA2, 
+waist-hip ratio, waist-circumference, and "Hip combined" since they are 
+highly correlated with other scores, which could make our estimates unstable.
+Figure \@ref(fig:plot-together) shows the results.
+Interestingly, several PGS remain independently significant, although effect sizes are reduced.
+
+```{r plot-together, fig.cap = "Partial correlations with number of children"}
+
+loadd(res_together)
+n_regs <- as.double(nrow(res_together))
+res_together %>% 
+      mutate(
+        term = pretty_names(term),
+        term = fct_reorder(term, estimate, order_abs(2)),
+        dep.var = dplyr::recode(dep.var, 
+          "n_children" = "N Children", "n_sibs" = "N Siblings"),
+        dep.var = fct_relevel(dep.var, "N Siblings")
+      ) %>% 
+      ggplot(aes(y = term, x = estimate, color = p.value < 0.05/{{n_regs}}, 
+        fill  = PCs)) + 
+        pval_geom_point(alpha = 0.8) +
+        my_vline +
+        facet_wrap(vars(dep.var)) +
+        pval_scale() +
+        my_fill_scale(n = 2, direction = -1) +
+        my_labs() +
+        theme(legend.position = "bottom")
+```

### 2020-05-21 _drake.R

@@ -135,15 +135,25 @@ plan <- drake_plan(
   
   res_weighted =  map_dfr(score_names, 
                     run_regs_weighted, 
-                    famhist = famhist, 
-                    weight_data = ghs_weights
+                    famhist     = famhist, 
+                    weight_data = ghs_weights,
+                    dep.var     = "n_children"
                   ),
+  
+  
+  res_sibs_weighted = map_dfr(score_names, 
+                        run_regs_weighted, 
+                        famhist     = famhist, 
+                        weight_data = ghs_weights,
+                        dep.var     = "n_sibs"
+                      ),
   
   
   res_flb_weights = map_dfr(score_names, 
                       run_regs_weighted, 
-                      famhist = famhist, 
-                      weight_data = flb_weights
+                      famhist     = famhist, 
+                      weight_data = flb_weights,
+                      dep.var     = "n_children"
                     ),
 
   

### 2020-05-21 regressions.R

@@ -46,10 +46,10 @@ run_regs_pcs <- function (dep_var, famhist, pcs) {
 }
 
 
-run_regs_weighted <- function (score_name, famhist, weight_data) {
+run_regs_weighted <- function (score_name, famhist, weight_data, dep.var) {
   famhist <- inner_join(famhist, weight_data, by = "f.eid")
 
-  fml <- as.formula(glue("n_children ~ {score_name}"))
+  fml <- as.formula(glue("{dep.var} ~ {score_name}"))
   mod <- lm(fml, data = famhist, weights = weights, na.action = na.exclude)
   regs <- tidy(mod, conf.int = TRUE)
   regs %<>% filter(term != "(Intercept)")
@@ -60,7 +60,9 @@ run_regs_weighted <- function (score_name, famhist, weight_data) {
 
 run_regs_period <- function (children, score_name, famhist) {
   dep_var  <- if (children) "n_children" else "n_sibs"
-  famhist$year_split <- famhist$YOB >= median(famhist$YOB, na.rm = TRUE)
+  famhist$year_split <- famhist$YOB >= 1950
+  # evaluates to 1950:
+  # median(famhist$YOB, na.rm = TRUE)
   famhist$year_split <- factor(famhist$year_split, labels = c("early", "late"))
   fml <- as.formula(glue("{dep_var} ~ year_split + {score_name}:year_split"))
   mod <- lm(fml, famhist)

### 2020-05-22 why-negative-selection.Rmd
Weight data for parents. standard_ggplot for plots. Attempt (but fail) to run multinomial logits.

@@
 my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
-pval_palette <- c("FALSE" = "grey50", "TRUE" = "orange2")
+pval_palette <- c("FALSE" = "grey65", "TRUE" = "orange2")
+
+standard_ggplot <- function (
+        dfr, 
+        fill_col, 
+        n_regs, 
+        ...,
+        score_col = quo(term), 
+        order_idx = 1, 
+        n_cats    = NULL
+      ) {
+  if (! missing(score_col)) score_col <- enquo(score_col)
+  fill_col <- if (! missing(fill_col)) enquo(fill_col) else quo(NULL)
+  if (missing(n_cats)) {
+    n_cats <- if (missing(fill_col)) 1 else length(unique(pull(dfr, 
+                {{fill_col}})))
+  }
+
+  n_regs <- as.double(n_regs)
+
+  dfr %>% mutate(
+          !! score_col := pretty_names(!! score_col),
+          !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
+        ) %>% 
+        ggplot(aes(estimate, {{score_col}}, fill = {{fill_col}}, ...,
+          colour = p.value < 0.05/{{n_regs}})) +
+          pval_geom_point() +
+          my_vline + 
+          pval_scale() +
+          my_fill_scale(n = n_cats) +
+          my_labs() +
+          my_theme()
+}
 
@@
   if (n == 2) {
-    scale_color_brewer(type = "qual", palette = 3, aesthetics = aesthetics, ...)
+    scale_colour_manual(values = c("steelblue4", "indianred2"), 
+          aesthetics = aesthetics, ...)
+    # scale_color_brewer(type = "qual", palette = 3, aesthetics = aesthetics, ...)
   } else {
@@
   return(names)
+}
+
+# from https://github.com/dgrtwo/drlib/blob/master/R/reorder_within.R
+# for ordering within facets
+reorder_within <- function(x, by, within, fun = mean, sep = "___", ...) {
+  new_x <- paste(x, within, sep = sep)
+  stats::reorder(new_x, by, FUN = fun)
+}
+
+scale_x_reordered <- function(..., sep = "___") {
+  reg <- paste0(sep, ".+$")
+  ggplot2::scale_x_discrete(labels = function(x) gsub(reg, "", x), ...)
+}
+
+scale_y_reordered <- function(..., sep = "___") {
+  reg <- paste0(sep, ".+$")
+  ggplot2::scale_y_discrete(labels = function(x) gsub(reg, "", x), ...)
 }
@@
 
-```{r plot-means-over-time, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Lines are means for 5-year intervals"}
+```{r plot-means-over-time, fig.height = 5, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Lines are means for 5-year intervals"}
 pgs_over_time %>% 
@@
       ggplot(aes(YOB, score, group = score_name)) + 
-      geom_line() +
+      geom_line(colour = "steelblue4") +
       facet_wrap(~score_name) +
+      my_hline +
       theme(
@@
       pval_scale() +
-      my_fill_scale(n = 2, direction = -1, guide = guide_legend(title = "")) +
+      my_fill_scale(n = 2, guide = guide_legend(title = "")) +
       my_theme()
@@
         pval_scale() +
-        my_fill_scale(n = 2, direction = -1) +
+        my_fill_scale(n = 2) +
         my_labs() +
@@
 
-```{r plot-n-partners, fig.cap = "Effect sizes on number of children by lnumber of sexual partners", fig.height = 4}
+```{r plot-n-partners, fig.cap = "Effect sizes on number of children by number of sexual partners", fig.height = 4}
 
@@
       mutate(
-        score_name = pretty_names(score_name),
-        score_name = fct_reorder(score_name, estimate, order_abs(1))
-      ) %>% 
-      ggplot(aes(y = score_name, x = estimate, fill = `N partners`, 
+        score_name = pretty_names(score_name)
+      ) %>% 
+      ggplot(aes(y = reorder_within(score_name, estimate, sex, fun = order_abs(1)), x = estimate, fill = `N partners`, 
         colour = p.value < 0.05/{{n_regs}})) + 
@@
         pval_geom_point() +
-        facet_wrap(vars(sex)) +
+        facet_wrap(vars(sex), scales = "free_y") +
         my_fill_scale(n = 2) +
@@
         my_theme(legend.position = "bottom") +
-        my_labs(x = "", y = "")
+        my_labs(x = "", y = "") + 
+        scale_y_reordered()
 
@@
 
+Our results are subject to two kinds of ascertainment bias. First, Biobank is
+not a random sample of the UK population: for example, it overrepresents
+educated people and high income households. Second, sampling parents of
+respondents inherently overweights parents who have many children. For instance,
+parents of three children will have three times more children as respondents in
+UK Biobank than parents of one child, on average. Parents of no children will,
+by definition, not be represented at all.
+
 ```{r calc-res-weighted}
@@
 drake::loadd(res_weighted)
-res_weighted_all <- res_weighted %>% 
-      left_join(
-        res_all %>% 
-              filter(dep.var == "N children", reg.type == "raw") %>% 
-              select(term, estimate),
+drake::loadd(res_sibs_parent_weights)
+drake::loadd(res_all)
+res_weighted_all <- bind_rows(
+        "N children" = res_weighted, 
+        "N siblings" = res_sibs_parent_weights, 
+      .id = "dep.var") 
+res_weighted_all$reg.type <- "weighted"
+res_weighted_all <- bind_rows(
+        res_weighted_all,
+        res_all %>% filter(reg.type == "raw") %>% select(-score_name)
+      )
+
+res_w_wide <- full_join(
+        res_weighted,
+        res_sibs_parent_weights,
         by = "term",
-        suffix = c("_weighted", "_unweighted")
-      ) %>% 
-      mutate(
-        prop_wtd_unwtd = estimate_weighted/estimate_unweighted
-      )
-
-mean_prop_wtd_unwtd <- mean(res_weighted_all$prop_wtd_unwtd)
-
-```
-
-As a fix for ascertainment bias in the UK Biobank sample, we weight participants
+        suffix = c(".children", ".sibs")
+      ) 
+# barbaric but it works:
+res_uw_wide <- left_join(
+  res_all %>% filter(dep.var == "N children", reg.type == "raw"),
+  res_all %>% filter(dep.var == "N siblings", reg.type == "raw"),
+  by = "term",
+  suffix = c(".children", ".sibs")
+)
+
+res_wide <- left_join(res_w_wide, res_uw_wide, by = "term", 
+      suffix = c(".wtd", ".uwtd"))
+
+res_wide %<>% mutate(
+        prop_children = estimate.children.wtd/estimate.children.uwtd,
+        prop_sibs     = estimate.sibs.wtd/estimate.sibs.uwtd
+      )
+
+```
+
+To compensate for ascertainment bias in the UK Biobank sample, we weight participants
 by sex, age (within the 40-71 age group), and education. Our sample weights are
-calibrated based on the 2006 General Household Survey. We then rerun the basic
-*children* regressions. Figure \@ref(fig:plot-weighted) shows the results, with
-unweighted estimates plotted for comparison. Weighting increases effect sizes on
-average by `r 100 * (mean_prop_wtd_unwtd - 1)` per cent.
-
-```{r plot-weighted, fig.cap = "Effects of polygenic scores on number of children, regressions weighted by education levels within age categories"}
-
-n_regs <- as.double(nrow(res_weighted_all))
+calibrated based on the 2006 General Household Survey. We then calculate the 
+proportions of respondents within each sibling group size, based on the respondent
+weights, and use this to create weights for the sibling regressions.
+
+We rerun the basic regressions for both children and siblings, using the new
+weights. Figure \@ref(fig:plot-res-weighted) shows the results. The weightings
+are fairly successful at reconciling the overall size of effects from
+siblings and children regressions. Tables \@ref(tab:tbl-sibs-weights) and
+\@ref(tab:tbl-children-weights) show the largest and smallest changes due to 
+weighting in siblings and children regressions.
+
+```{r tbl-sibs-weights}
+
+stars <- function (x) ifelse(x < 0.05/33, "*", "")
+
+top_bottom_5 <- function (x, caption) {
+  x %>% 
+      slice(c(1:5, nrow(.) - 0:4)) %>% 
+      as_hux() %>% 
+      insert_row(rep("...", 4), after = 6) %>% 
+      set_number_format(-1, -1, 3) %>% 
+      set_number_format(-1, 4, 0) %>% 
+      set_align(-1, -1, ".") %>% 
+      theme_basic() %>% 
+      set_font_size(8) %>% 
+      set_caption(caption) %>% 
+      set_caption_pos("bottom") %>% 
+      set_caption_width(1) %>% 
+      add_footnote("Insignificant effects not shown", font_size = 7) %>% 
+      set_all_padding(0) 
+}
+
+res_wide %>% 
+      arrange(desc(prop_sibs)) %>% 
+      select(term, matches("estimate.sibs"), matches("p.value.sibs"), 
+        prop_sibs) %>% 
+      filter(p.value.sibs.uwtd < 0.05/33) %>% 
+      transmute(
+        Score = pretty_names(term),
+        `Weighted estimate` = estimate.sibs.wtd,
+        `Unweighted estimate` = estimate.sibs.uwtd,
+        `Proportion (%)` = prop_sibs * 100
+      ) %>% 
+      top_bottom_5(caption = "Effects of weighting on sibling regressions: top and bottom 5 changes")
+```
+
+```{r tbl-children-weights}
+
+res_wide %>% 
+      arrange(desc(prop_children)) %>% 
+      select(term, matches("estimate.children"), matches("p.value.children"), 
+        prop_children) %>% 
+      filter(p.value.children.uwtd < 0.05/33) %>% 
+      transmute(
+        Score = pretty_names(term),
+        `Weighted estimate` = estimate.children.wtd,
+        `Unweighted estimate` = estimate.children.uwtd,
+        `Proportion (%)` = prop_children * 100
+      ) %>% 
+      top_bottom_5(caption = "Effects of weighting on children regressions: top and bottom 5 changes")
+```
+
+```{r plot-res-weighted, fig.cap = "Effects of polygenic scores on number of children/siblings. Weighted regressions."}
 
 res_weighted_all %>% 
-      mutate(
-        term    = pretty_names(term),
-        term    = fct_reorder(term, estimate_weighted, order_abs(1))
-      ) %>% 
-      pivot_longer(cols = c(estimate_weighted, estimate_unweighted), 
-        names_to = "Regression", values_to = "estimate") %>% 
-      mutate(
-        Regression = sub("estimate_", "", Regression)
-      ) %>% 
-      ggplot(aes(y = term, x = estimate, shape = Regression,
-        colour = p.value < 0.05/{{n_regs}})) +
-      my_vline + 
-      geom_point(size = 1.5, stroke = 0.7, fill = "steelblue") +
-      labs(x = "Effect size", y = "") +
-      scale_shape_manual(values = c("weighted" = 22, "unweighted" = 21)) +
-      pval_scale(aesthetics = "colour") +
-      my_theme()
-
-```
-
-```{r plot-sibs-weighted, eval = FALSE}
-
-# what this shows is that even after weighting, sibling effect sizes are
-# still stronger. In fact weighting hasn't changed things much.
-drake::loadd(res_weighted, res_sibs_weighted)
-n_regs <- as.double(nrow(res_weighted))
-# bind_rows(children = res_weighted, sibs = res_sibs_weighted, .id = "Dep. Var.") %>% 
-res_all %>% 
-      filter(reg.type == "raw") %>% 
-      mutate(
-        term    = pretty_names(term),
-        term    = fct_reorder(term, estimate, order_abs(1))
-      ) %>% 
-      ggplot(aes(y = term, x = estimate, fill = `dep.var`,
-        colour = p.value < 0.05/{{n_regs}})) +
-      my_vline + 
-      pval_geom_point() +
-      labs(x = "Effect size", y = "") +
-      my_fill_scale(n = 2) +
-      pval_scale(aesthetics = "colour")
+      rename(
+        "Dependent variable" = dep.var, 
+      ) %>% 
+      filter(reg.type == "weighted") %>% 
+      standard_ggplot(fill = `Dependent variable`, n_regs = 33, order_idx = 1, 
+        shape = reg.type) + 
+        scale_shape_manual(values = c(21, 4))
 
@@
 These results suggest that previous work may have underestimated the level of
-selection occurring in the population. [XXX who?]
+selection occurring in the population. [TODO: who?]
 
@@
 
-A possible answer is given by the economic theory of fertility [XXX cite Becker 1960].
+A possible answer is given by the economic theory of fertility [TODO cite Becker 1960].
 According to this, increases in potential earned income affect fertility via two opposing channels. There is an "income effect" by which children become more
@@
         pval_scale() +
-        my_fill_scale(n = 2, direction = -1) +
+        my_fill_scale(n = 2) +
         my_labs()
@@
 polygenic scores residualized for the top 100 principal components of the
-genetic data, calculated within the UK Biobank population. (XXX Abdel details.)
+genetic data, calculated within the UK Biobank population. (TODO Abdel details.)
 
@@
 
-```{r plot-regs-controlled, fig.cap = "Effects of residualized polygenic scores on number of siblings/children.", fig.height = 7}
+```{r plot-regs-controlled, fig.cap = "Effects of residualized polygenic scores on number of siblings/children."}
 
@@
         term    = pretty_names(term),
-        term    = fct_reorder(term, estimate, order_abs(1)),
+        term    = reorder_within(term, estimate, dep.var, order_abs(1)),
         dep.var = fct_inorder(dep.var) 
@@
         my_vline + 
-      facet_wrap(vars(dep.var)) +
-      pval_geom_point(alpha = .7) +
+        facet_wrap(vars(dep.var), scales = "free_y") +
+        pval_geom_point(position = position_dodge(0.4)) +
         my_labs() +
         pval_scale() +
-      my_fill_scale(n = 2, guide = guide_legend(title = ""))
+        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
+        my_theme() +
+        scale_y_reordered()
 
@@
 
-## Causality
-
-## Number of children
-
-Figure \@ref(fig:plot-n-children-bar) shows the full distribution of number 
-of children born for different ventiles of the EA3 polygenic score. The 
-strongest relationship seems to be for having 0 children versus 1 or more.
-
-```{r plot-n-children-bar, fig.cap = "Number of children by ventiles of EA3 PGS"}
-famhist %>% 
-      select(n_children, EA3_excl_23andMe_UK, sex) %>% 
-      mutate(
-        "EA3 Ventile" = chop_equally(EA3_excl_23andMe_UK, 20, labels = 1:20),
-        "N Children"  = santoku::chop(n_children, brk_left(0:4, FALSE), 
-                          labels = c(0:3, "4+")),
-        Sex           = factor(sex, levels = 0:1, labels = c("Female", "Male"))
-        ) %>% 
-      filter(! is.na(`EA3 Ventile`), ! is.na(`N Children`)) %>% 
-      ggplot(aes(`EA3 Ventile`, fill = `N Children`, group = `N Children`)) + 
-        geom_bar(position = position_fill(reverse = TRUE), alpha = 0.7) + 
-        facet_wrap(vars(Sex)) +
-        scale_fill_viridis_d(option = "A") +
-        scale_y_continuous(n.breaks = 10, labels = scales::percent) +
-        theme(legend.position = "bottom")
-
-```
-
-```{r plot-sd-years, eval = FALSE}
-# inequality over time. Nothing to see, almost everything is flat.
-grid <- expand_grid(score_name = score_names, YOB = 1940:1965)  
-grid$sd <- pmap_dbl(grid, ~sd(famhist[famhist$YOB==.y, ][[.x]], na.rm = TRUE))
-ggplot(grid, aes(YOB, sd)) + 
-      geom_line() + 
-      facet_wrap(~score_name, scales = "free_y", ncol = 4)
-```
-
-```{r rgs-by-effect-size, eval = FALSE, fig.cap = "Genetic correlations with EA3 by effect size on n siblings."}
-rgs <- left_join(rgs, res %>% filter(reg.type == "residualized"), by = c("p2" = "term"))
+## Genetic correlations with EA3
+
+Another way to examine the "earnings" theory of natural selection is to compare
+effect sizes of PGS with their genetic correlation with educational attainment
+(EA3). Since EA3 strongly predicts earnings, if earnings drives differences
+in fertility, we'd expect a correlation between the two sets of results. Figure
+\@ref(fig:plot-rgs-by-effect-size) shows this is so.
+
+TODO: ref source of data.
+
+```{r plot-rgs-by-effect-size, fig.cap = "Effect size on n siblings by genetic correlation with EA3"}
+
+drake::loadd(rgs)
+drake::loadd(res_all)
+
+rgs <- left_join(rgs, 
+        res_all %>% filter(reg.type == "raw", dep.var == "N children"), 
+        by = c("p2" = "term"))
 
@@
       ggplot(aes(rg, estimate)) + 
-        geom_smooth(method = lm) +
-        geom_point(colour = "red2") + 
-        geom_label(aes(label = p2, hjust = "left", vjust = "middle"), 
-          size = 3, nudge_x = 0.02, alpha = 0.5, 
-          label.padding = unit(0.1, "lines")) +
-        labs(x = "rG", y = "Effect size", 
-          title = "rG with EA3 vs. effect size on # siblings")
+        geom_point() +
+        my_vline +
+        my_hline +
+        labs(x = "rG with EA3", y = "Effect on number of children")
+
 ```
@@
 ```
+
+## Number of children
+
+Figure \@ref(fig:plot-n-children-bar) shows the full distribution of number 
+of children born for different ventiles of the EA3 polygenic score. The 
+strongest relationship seems to be for having 0 children versus 1 or more.
+
+```{r plot-n-children-bar, fig.cap = "Number of children by ventiles of EA3 PGS"}
+famhist %>% 
+      select(n_children, EA3_excl_23andMe_UK, sex) %>% 
+      mutate(
+        "EA3 Ventile" = chop_equally(EA3_excl_23andMe_UK, 20, labels = 1:20),
+        "N Children"  = santoku::chop(n_children, brk_left(0:4, FALSE), 
+                          labels = c(0:3, "4+")),
+        Sex           = factor(sex, levels = 0:1, labels = c("Female", "Male"))
+        ) %>% 
+      filter(! is.na(`EA3 Ventile`), ! is.na(`N Children`)) %>% 
+      ggplot(aes(`EA3 Ventile`, fill = `N Children`, group = `N Children`)) + 
+        geom_bar(position = position_fill(reverse = TRUE), alpha = 0.7) + 
+        facet_wrap(vars(Sex)) +
+        scale_fill_viridis_d(option = "A") +
+        scale_y_continuous(n.breaks = 10, labels = scales::percent) +
+        theme(legend.position = "bottom")
+
+```
+
+## Causality
+
+Different polygenic scores are correlated. Table \@ref(tab:tbl-pgs-corrs) shows the top
+correlations in the sample. Because of this, bivariate correlations between
+PGS and number of children might be driven by other genetic scores. To explore which polygenic scores are driving negative selection, we run a single
+omnibus regression of *number of children* on all the PGS. We exclude EA2, 
+waist-hip ratio, waist-circumference, and "Hip combined" since they are 
+highly correlated with other scores, which could make our estimates unstable.
+Figure \@ref(fig:plot-together) shows the results.
+Interestingly, several PGS remain independently significant, although effect sizes are reduced.
 
@@
 
-Different polygenic scores are correlated. Table \@ref(tab:tbl-pgs-corrs) shows the top
-correlations in the sample. Because of this, bivariate correlations between
-PGS and number of children might be driven by other genetic scores. To explore which polygenic scores are driving negative selection, we run a single
-omnibus regression of *number of children* on all the PGS. We exclude EA2, 
-waist-hip ratio, waist-circumference, and "Hip combined" since they are 
-highly correlated with other scores, which could make our estimates unstable.
-Figure \@ref(fig:plot-together) shows the results.
-Interestingly, several PGS remain independently significant, although effect sizes are reduced.
-
 ```{r plot-together, fig.cap = "Partial correlations with number of children"}

### 2020-05-22 _drake.R

@@ -28,7 +28,6 @@ ghs_file      <- file.path(data_dir, "UKDA-5804-stata8/stata8/Ghs06client.dta")
 
 plan <- drake_plan(
   
-  
   score_names  = {
     sub(
       ".*UKB\\.AMC\\.(.*?)\\..*", 
@@ -37,7 +36,6 @@ plan <- drake_plan(
       perl = TRUE
     )
   },
-  
   
   famhist_raw  = target(
                    make_famhist( 
@@ -51,24 +49,21 @@ plan <- drake_plan(
                     format = "fst"
                   ),
   
-  
   pcs          = read_table2(file_in(!! pcs_file)),
   
+  famhist      =  target(
+                   edit_famhist(famhist_raw, score_names), 
+                   format = "fst"
+                  ),
   
-  famhist      = target(
-                  edit_famhist(famhist_raw, score_names), 
-                  format = "fst"
-                 ),
-  
+  fhl_mlogit   =  make_famhist_long_mlogit(famhist, score_names),
   
   resid_scores = target(
                    make_resid_scores(famhist, pcs, score_names), 
                    format = "fst"
                  ),
 
-  
   ghs_subset   = target(make_ghs_subset(file_in(!! ghs_file)), format = "fst"),
-  
   
   ghs_weights  =  weight_by_ghs(
                     ~ factor(sex) + age_at_recruitment + factor(edu_qual), 
@@ -84,14 +79,14 @@ plan <- drake_plan(
           )
   },
   
+  parent_weights = weight_parents(famhist, ghs_weights),
+  
   mf_pairs     = target(
                    make_mf_pairs(file_in(!! mf_pairs_file), famhist), 
                    format = "fst"
                  ), 
   
-  
   rgs          = make_rgs(file_in(!! rgs_file)),
-  
   
   res_all      = {
     famhist <- join_famhist_resid_scores(famhist, resid_scores)
@@ -112,7 +107,6 @@ plan <- drake_plan(
           )
   },
   
-  
   res_pcs     = {
     pc_names <- grep("PC", names(pcs), value = TRUE)
     res_sibs_pcs <- run_regs_pcs( 
@@ -132,7 +126,6 @@ plan <- drake_plan(
     ) 
   },
   
-  
   res_weighted =  map_dfr(score_names, 
                     run_regs_weighted, 
                     famhist     = famhist, 
@@ -140,24 +133,21 @@ plan <- drake_plan(
                     dep.var     = "n_children"
[…]

### 2020-05-22 make-data.R

@@ -8,6 +8,7 @@ suppressPackageStartupMessages({
   library(haven)
   loadNamespace("car") # very annoying if it overwrites recode
   loadNamespace("matrixStats")
+  loadNamespace("mlogit")
 })
 
 # utility function:
@@ -189,6 +190,15 @@ edit_famhist <- function (famhist, score_names) {
   famhist[score_names] <- scale(famhist[score_names])
   
   return(famhist)
+}
+
+
+make_famhist_long_mlogit <- function (famhist, score_names) {
+  famhist$n_ch_fac <- santoku::chop(famhist$n_children, 
+        brk_left(0:5, close_end = FALSE), lbl_integer())
+  fh_subset <- famhist %>% select(n_ch_fac, all_of(score_names))
+  mlogit::mlogit.data(fh_subset, choice = "n_ch_fac", shape = "wide", 
+        alt.levels = levels(fh_subset$n_ch_fac))
 }
 
 

### 2020-05-22 regressions.R

@@ -100,3 +100,16 @@ run_regs_fml <- function(fml, ..., subset = NULL, famhist) {
   res
 }
 
+
+run_regs_mnlogit <- function(score_names, fhl_mlogit) {
+  loadNamespace("mnlogit")
+  run_1_mnlogit <- function(score_name) {
+    fml <- as.formula(glue::glue("n_ch_fac ~ 1 | {score_name}"))
+    res <- try(mod <- mnlogit::mnlogit(fml, fhl_mlogit, ncores = 2))
+    if (inherits(res, "try-error")) return(res)
+    mod$model <- NULL # otherwise it stores a copy of the entire data 8-(
+    return(mod)
+  }
+  lapply(score_names, run_1_mnlogit)
+}
+

### 2020-05-22 why-negative-selection.Rmd
Work on reporting results of weighting

@@
 
-* If this theory is true then the pattern of differences is explained by
-  (potential) earnings. So the strength of the relationship of a PGS with 
-  number of children should be explained by the strength of its relationship
-  with earnings and education.
-* Even after weighting, effect sizes for siblings are still higher than for
-  children. Why?
-  - Remaining ascertainment bias?
-  - Siblings don't include those with 0 children?
-  - Something else?
 * If we weight with the GHS, and plot the means of figure 1 using this w
@@
   by figure 1. 
+* Need to make the weighting very watertight. 
+  - Include income? 
+  - Interact age and other variables?
+  - Investigate "MRP" (Gelman 2005)
+  - Understand reweighting on subsets of survey data.
+  - Check for existing work
+* Reweight other regressions (e.g. the omnibus regression in the appendix)
 * Could be useful: f.2139 - age first had sex (includes "never had sex" 
@@
 * Look at geography, esp. in siblings regressions.
-* Work more on weighting the data
-  - check for existing work by others?
-  - deal with the fact that GHS 06 has its own weights and use
-    subset.svydesign to reweight?
-* control for age in n_partners regressions
+* control for age in n_partners regressions.
 
@@
 
-res_w_wide <- full_join(
+res_wide <- full_join(
         res_weighted,
@@
         suffix = c(".children", ".sibs")
-      ) 
-# barbaric but it works:
-res_uw_wide <- left_join(
-  res_all %>% filter(dep.var == "N children", reg.type == "raw"),
-  res_all %>% filter(dep.var == "N siblings", reg.type == "raw"),
-  by = "term",
-  suffix = c(".children", ".sibs")
-)
-
-res_wide <- left_join(res_w_wide, res_uw_wide, by = "term", 
-      suffix = c(".wtd", ".uwtd"))
-
-res_wide %<>% mutate(
-        prop_children = estimate.children.wtd/estimate.children.uwtd,
-        prop_sibs     = estimate.sibs.wtd/estimate.sibs.uwtd
-      )
+      ) %>% 
+      filter(p.value.children < 0.05/33 | p.value.sibs < 0.05/33)
+
+chn_larger <- with(res_wide, estimate.children > estimate.sibs)
 
@@
 weights. Figure \@ref(fig:plot-res-weighted) shows the results. The weightings
-are fairly successful at reconciling the overall size of effects from
-siblings and children regressions. Tables \@ref(tab:tbl-sibs-weights) and
-\@ref(tab:tbl-children-weights) show the largest and smallest changes due to 
-weighting in siblings and children regressions.
-
-```{r tbl-sibs-weights}
-
-stars <- function (x) ifelse(x < 0.05/33, "*", "")
-
-top_bottom_5 <- function (x, caption) {
-  x %>% 
-      slice(c(1:5, nrow(.) - 0:4)) %>% 
-      as_hux() %>% 
-      insert_row(rep("...", 4), after = 6) %>% 
-      set_number_format(-1, -1, 3) %>% 
-      set_number_format(-1, 4, 0) %>% 
-      set_align(-1, -1, ".") %>% 
-      theme_basic() %>% 
-      set_font_size(8) %>% 
-      set_caption(caption) %>% 
-      set_caption_pos("bottom") %>% 
-      set_caption_width(1) %>% 
-      add_footnote("Insignificant effects not shown", font_size = 7) %>% 
-      set_all_padding(0) 
-}
-
-res_wide %>% 
-      arrange(desc(prop_sibs)) %>% 
-      select(term, matches("estimate.sibs"), matches("p.value.sibs"), 
-        prop_sibs) %>% 
-      filter(p.value.sibs.uwtd < 0.05/33) %>% 
-      transmute(
-        Score = pretty_names(term),
-        `Weighted estimate` = estimate.sibs.wtd,
-        `Unweighted estimate` = estimate.sibs.uwtd,
-        `Proportion (%)` = prop_sibs * 100
-      ) %>% 
-      top_bottom_5(caption = "Effects of weighting on sibling regressions: top and bottom 5 changes")
-```
-
-```{r tbl-children-weights}
-
-res_wide %>% 
-      arrange(desc(prop_children)) %>% 
-      select(term, matches("estimate.children"), matches("p.value.children"), 
-        prop_children) %>% 
-      filter(p.value.children.uwtd < 0.05/33) %>% 
-      transmute(
-        Score = pretty_names(term),
-        `Weighted estimate` = estimate.children.wtd,
-        `Unweighted estimate` = estimate.children.uwtd,
-        `Proportion (%)` = prop_children * 100
-      ) %>% 
-      top_bottom_5(caption = "Effects of weighting on children regressions: top and bottom 5 changes")
-```
-
-```{r plot-res-weighted, fig.cap = "Effects of polygenic scores on number of children/siblings. Weighted regressions."}
+are fairly successful at reconciling the overall size of effects from siblings
+and children regressions. Among weighted scores that are significant (at p <
+0.05/33) in either regression, `r sum(chn_larger)` estimates are
+higher for children and `r length(chn_larger) - sum(chn_larger)` are
+lower. However, some estimates differ widely or change
+sign across the generations.
+
+```{r plot-res-weighted, fig.cap = "Effects of polygenic scores on number of children/siblings. Regressions weighted by sex, age and education."}
 
@@
 at first live birth.  Figure \@ref(fig:plot-flb-weights) 
-shows the results. Weighting increased effect sizes on average by 
-`r 100 * (prop_wt_nowt - 1)` per cent. There 
+shows the results. There 
 are large changes for some variables, including EA3 
@@
 
+Overall, these results suggest that previous work may have underestimated the level of
+selection occurring in the population. [TODO: who?]
+
 ```{r plot-flb-weights, fig.cap = "Effect sizes on number of children, female respondents weighted by age at first live birth"}
@@
       ) %>% 
-      mutate(
-        term = pretty_names(term),
-        term = fct_reorder(term, estimate, order_abs(1))
-      ) %>% 
-      ggplot(aes(estimate, term, shape = Regression, 
-        color = p.value < 0.05/66)) +
-        my_vline +
-        geom_point(size = 1.5, stroke = 0.7, fill = "steelblue") +
-        scale_shape_manual(values = c("Weighted" = 22, "Unweighted" = 21)) +  
-        pval_scale() +
-        my_labs() + 
-        my_theme()
-```
-
-These results suggest that previous work may have underestimated the level of
-selection occurring in the population. [TODO: who?]
+      standard_ggplot(Regression, n_regs = 66)
+```
+
+## Mechanisms behind natural selection
+
+Why are selection effects *oppositely* signed in certain subgroups -- for
+example, in older and younger mothers (by age at first live birth) or in those
+with more and fewer lifetime sexual partners? A possible answer is given by the
+economic theory of fertility [TODO cite Becker 1960]. According to this,
+increases in potential earned income affect fertility via two opposing channels.
+There is an "income effect" by which children become more affordable, like any
+other good. Since childrearing has a cost in time, there is also a substitution
+effect: the opportunity cost of childrearing increases if one's market wage is
+higher. The income effect would lead higher earners to have more children. The
+substitution effect would lead higher earners to have fewer children. It is
+often assumed that the substitution effect will dominate for lone parents, since
+they have less opportunity to share childcare responsibilities, while the income
+effect will dominate for couples who are able to reap the gains from
+specialization.
+
+If so, then genetic characteristics which affect one's earnings potential in the
+labour market may lead to opposing effects on fertility. Genetic
+variants which improve one's earnings may increase fertility among couples, but
+decrease fertility among single parents.
+
+This explanation predicts that the strength of polygenic scores' effects on
+fertility will correlate with the strength of their relationship with earnings.
+We measure the correlation of each PGS with household income, as well as with
+education level, a phenotypic variable which is likely to predict earnings. We
+then relate these measurements to effect sizes on number of children.
+Across scores, there is a clear negative relationship between effect on
+fertility, and the correlation with income and/or education (Figure \@ref(fig:plot-cors-income-educ).
+
+```{r plot-cors-income-educ, fig.cap="Effect sizes on number of children and correlations with income and education"}
+
+drake::loadd(res_all)
+cors_income <- cor(famhist[c(score_names)], famhist$income_cat, use = "pairwise")
+# TODO: use age_left_edu when this includes college
+cors_educ <-   cor(famhist[c(score_names)], famhist$edu_qual, use = "pairwise")
+
+effect_size <- res_all %>% 
+      filter(dep.var == "N children", reg.type == "raw") %>% 
+      pull(estimate)
+
+dfr <- data.frame(
+      type = rep(c("Cor. with income", "Cor. with educ"), each = 33),
+      cor = c(cors_income[,1], cors_educ[,1]),
+      effect_size = rep(effect_size, 2)
+)
+
+ggplot(dfr, aes(effect_size, cor, colour = type)) + 
+      geom_point() +
+      facet_wrap(vars(type)) + 
+      my_vline + 
+      my_hline +
+      labs(x = "Effect on n children", y = "Correlation") +
+      scale_colour_manual(values = c("darkblue", "darkgreen")) +
+      theme(legend.position = "none")
+
+```
+
+# Bits to integrate
 
@@
 ```
-
-## Mechanisms behind natural selection
-
-In this section, we dig into the mechanisms causing natural selection in the
-population. In particular, why are selection effects *oppositely* signed in
-certain subgroups -- for example, in older and younger mothers (by age at first
-live birth) or in those with more and fewer lifetime sexual partners?
-
-A possible answer is given by the economic theory of fertility [TODO cite Becker 1960].
-According to this, increases in potential earned income affect fertility via two opposing channels. There is an "income effect" by which children become more
-affordable, like any other good. Since childrearing has a cost in time, there is
-also a substitution effect: the opportunity cost of childrearing increases if
-one's market wage is higher. The income effect would lead higher earners to
-have more children. The substitution effect would lead higher earners to have
-fewer children.
-
-Traditionally, it was assumed that since women
-spent more of their time on childcare, the substitution effect would dominate
-for women's earnings, while the income effect would dominate for men's earnings. 
-However, a more relevant split may be between those raising children alone, and
-those raising children in a two-parent household. The substitution effect is likely
-to be particularly strong for lone parents, since they have less opportunity
-to share childcare responsibilities.
-
-If so, then genetic characteristics which affect one's earnings potential in the
-labour market may lead to opposing effects on fertility. Genetic
-variants which improve one's earnings may increase fertility among couples, but
-decrease fertility among single parents.
-
-This explanation predicts that the strength of polygenic scores' effects on
-fertility will correlate with the strength of their relationship with earnings.
-We measure the correlation of each PGS with household income, as well as with
-education level, a phenotypic variable which is likely to predict earnings. We
-then correlate these measurements with effect sizes on number of children.
-Across scores, there is a clear negative correlation between effect on
-fertility, and link to both income and education.
-
-```{r plot-cors-income-educ, fig.cap="Effect sizes on number of children and correlations with income and education"}
-
-drake::loadd(res_all)
-cors_income <- cor(famhist[c(score_names)], famhist$income_cat, use = "pairwise")
-# TODO: use age_left_edu when this includes college
-cors_educ <-   cor(famhist[c(score_names)], famhist$edu_qual, use = "pairwise")
-
-effect_size <- res_all %>% 
-      filter(dep.var == "N children", reg.type == "raw") %>% 
-      pull(estimate)
-
-dfr <- data.frame(
-      type = rep(c("Cor. with income", "Cor. with educ"), each = 33),
-      cor = c(cors_income[,1], cors_educ[,1]),
-      effect_size = rep(effect_size, 2)
-)
-
-ggplot(dfr, aes(effect_size, cor, colour = type)) + 
-      geom_point() +
-      facet_wrap(vars(type)) + 
-      my_vline + 
-      my_hline +
-      labs(x = "Effect on n children", y = "Correlation") +
-      scale_colour_manual(values = c("darkblue", "darkgreen")) +
-      theme(legend.position = "none")
-
-```
-
-# Bits to integrate
 
@@
 controlling for age at first live birth, and raw effect sizes, is `r cor_raw_flb`.
+
 ```{r plot-age-flb, fig.cap = "Effect sizes on number of children, controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}
@@
         term = pretty_names(term),
-        term = fct_reorder(term, estimate, order_abs(2)),
         dep.var = dplyr::recode(dep.var, 
@@
       ) %>% 
-      ggplot(aes(y = term, x = estimate, color = p.value < 0.05/{{n_regs}}, 
-        fill  = PCs)) + 
+      ggplot(aes(y = reorder_within(term, estimate, dep.var, order_abs(1)), 
+        x = estimate, color = p.value < 0.05/{{n_regs}}, fill  = PCs)) + 
         pval_geom_point(alpha = 0.8) +
         my_vline +
-        facet_wrap(vars(dep.var)) +
+        facet_wrap(vars(dep.var), scales = "free_y") +
         pval_scale() +
@@
         my_labs() + 
+        scale_y_reordered() +
         theme(legend.position = "bottom")

### 2020-05-26 why-negative-selection.Rmd
Add regression split by household type

@@
 
+Lastly Figure \@ref(fig:plot-with-partner) splits respondents up by whether they
+were living with a spouse or partner at the time of interview. Effects are 
+generally the same sign in both groups, but are much larger among those not
+living with a spouse or partner.
+
 ```{r plot-age-flb-cross, fig.cap = "Effect sizes on number of children, by age at first live birth terciles"}
@@
 
+```
+
+```{r plot-with-partner, fig.cap = "Effect sizes on number of children by household type"}
+
+drake::loadd(res_with_partner)
+res_with_partner %<>% filter(grepl(":", term))
+
+n_regs <- nrow(res_with_partner)
+
+res_with_partner %>% 
+      mutate(
+        Household = ifelse(grepl("TRUE", term), "With partner", "Without partner")
+      ) %>% 
+      standard_ggplot(fill = Household, n_regs = n_regs, score_col = score_name)
 ```

### 2020-05-26 _drake.R

@@ -272,7 +272,20 @@ plan <- drake_plan(
     left_join(res, pars, by = "combo") %>%
           mutate(sex = ifelse(subset == "sex == 1", "Male", "Female")) %>%
           select(-combo, -subset)
-  }, 
+  },
+  
+  res_with_partner = {
+    res <- map_dfr(setNames(score_names, score_names),
+            ~run_regs_fml(
+              "n_children ~ with_partner + {score_name}:with_partner",
+              score_name = .x,
+              famhist    = famhist
+            ),
+            .id = "score_name"
+          )
+    res %<>% filter(term != "(Intercept)")
+    res
+  },
   
   res_edu = {
     subsets <- rlang::exprs(

### 2020-05-26 make-data.R

@@ -46,7 +46,7 @@ make_resid_scores <- function (famhist, pcs, score_names) {
     resid_scores[[paste0(score_name, "_resid")]] <- resid_score
   }
   
-  resid_score$dummy <- NULL
+  resid_scores$dummy <- NULL
   resid_scores
 }
 
@@ -151,7 +151,13 @@ edit_famhist <- function (famhist, score_names) {
     na.rm = TRUE
   )
   
+  
   famhist$n_in_household <- famhist$f.709.0.0
+  
+  famhist$with_partner   <- famhist$f.6141.0.0 == 1
+  # Many NAs, almost all from people living alone i.e. f.709 == 1
+  famhist$with_partner[famhist$n_in_household == 1] <- FALSE
+  famhist$with_partner[famhist$f.6141.0.0 == -3] <- NA
   
   famhist$age_fte_cat <- santoku::chop(famhist$age_fulltime_edu, 
     c(16, 18), 

### 2020-06-12 why-negative-selection.Rmd
Updates to graphs in response to Abdel's comments

@@
 drake::loadd(score_names)
+drake::loadd(famhist)
 
@@
 my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
-pval_palette <- c("FALSE" = "grey65", "TRUE" = "orange2")
+pval_palette <- c("FALSE" = "grey65", "TRUE" = "#FAC32D")
 
@@
   if (n == 2) {
-    scale_colour_manual(values = c("steelblue4", "indianred2"), 
+    scale_colour_manual(values = c("steelblue4", "#D63C3C"), 
           aesthetics = aesthetics, ...)
@@
 pretty_names <- function (names) {
-  names <- gsub("_", " ", names)
-  names %<>% 
-        strsplit(" ") %>% 
-        map_chr(~paste0(
-          toupper(substring(.x, 1, 1)), 
-          substring(.x, 2),
-          collapse = " "
-        ))
-
-  acronyms <- c("WHR", "BMI", "DPW", "CPD", "SI", "AI", "WC")
-  for (acronym in acronyms) {
-    regex <- paste0("\\b", acronym, "\\b")
-    names <- gsub(regex, toupper(acronym), names, ignore.case = TRUE)
-  }
-
-  return(names)
+  pretty <- c(
+    ADHD_2017               = "ADHD",
+    age_at_menarche         = "Age at menarche",
+    age_at_menopauze        = "Age at menopause",
+    agreeableness           = "Agreeableness",
+    ai_substance_use        = "Age at smoking initiation",
+    alcohol_schumann        = "Alcohol use",
+    alzheimer               = "Alzheimer",
+    autism_2017             = "Autism",
+    bipolar                 = "Bipolar",
+    bmi_combined            = "BMI",
+    body_fat                = "Body Fat",
+    caffeine                = "Caffeine",
+    cannabis                = "Cannabis (ever vs. never)",
+    cognitive_ability       = "Cognitive Ability",
+    conscientiousness       = "Conscientiousness",
+    coronary_artery_disease = "Coronary Artery Disease",
+    cpd_substance_use       = "Cigarettes per day",
+    diagram_T2D             = "Type 2 Diabetes",
+    dpw_substance_use       = "Drinks per week",
+    EA2_noUKB               = "Educ. attainment 2 (no UK)",
+    EA3_excl_23andMe_UK     = "Educ. attainment 3 (no UK)",
+    eating_disorder         = "Eating disorder",
+    extraversion            = "Extraversion",
+    height_combined         = "Height",
+    hip_combined            = "Hip circumference",
+    MDD_PGC2_noUKB          = "Major Depressive Disorder",
+    neuroticism             = "Neuroticism",
+    openness                = "Openness",
+    sc_substance_use        = "Smoking cessation",
+    SCZ2                    = "Schizophrenia",
+    si_substance_use        = "Smoking initiation",
+    wc_combined             = "Waist circumference",
+    whr_combined            = "Waist-hip ratio"
+  )
+
+  pretty[names]
+}
+
+n_in_regs <- function (var, dv = "n_children") {
+  # all PGS have the same NA pattern, so we just use one
+  tbl <- table(famhist[
+                  ! is.na(famhist[dv]) & ! is.na(famhist$whr_combined), 
+                  var
+                ])
+  c(tbl)
+}
+
+add_n <- function (var, famhist_var = var, dv = "n_children") {
+  var <- as.factor(var)
+  n <- n_in_regs(famhist_var, dv)
+  levels(var) <- sprintf("%s (N = %s)", levels(var), 
+                         prettyNum(n[levels(var)], big.mark = ","))
+  # var_n <- sprintf("%s (N = %s)", var, prettyNum(n[var], big.mark = ","))
+  # var_n[is.na(var)] <- NA
+  var
 }
@@
         YOB = santoku::chop(YOB, 
-          brk_left(seq(1940, 1970, 5), close_end = FALSE), 
-          seq(40, 65, 5),
+          seq(1940, 1970, 5), 
+          labels    = seq(40, 65, 5),
           extend    = FALSE
@@
       filter(! is.na(YOB)) %>% 
-      summarize(across(all_of(score_names), mean, na.rm = TRUE)) %>% 
-      pivot_longer(-YOB, names_to = "score_name", values_to = "score")
-
-```
-
-```{r plot-means-over-time, fig.height = 5, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Lines are means for 5-year intervals"}
+      summarize(across(all_of(score_names), ggplot2::mean_cl_normal, na.rm = TRUE)) %>% 
+      pivot_longer(-YOB, names_to = "score_name", values_to = "score") %>% 
+      mutate(
+        score_lo = score$ymin,
+        score_hi = score$ymax,
+        score    = score$y
+      )
+
+```
+
+```{r plot-means-over-time, fig.height = 5, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Points are means for 5-year intervals. Lines are 95% confidence intervals."}
+
 pgs_over_time %>% 
       mutate(score_name = pretty_names(score_name)) %>% 
-      ggplot(aes(YOB, score, group = score_name)) + 
-      geom_line(colour = "steelblue4") +
+      ggplot(aes(YOB, group = score_name)) + 
+      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
+        colour = "steelblue4", fatten = 2) +
       facet_wrap(~score_name) +
@@
       theme(
-        strip.text  = element_text(size = 7), 
+        strip.text  = element_text(size = 6), 
         axis.text.x = element_text(size = 7)
@@
       labs(y = "PGS", x = "Birth year (5 year intervals)")
+
 ```
@@
 much larger for *siblings* than for *children* 
-regressions. Among consistently-signed estimates, the effect size for
-siblings is `r 100*median_prop_gens` per cent at the median.
+regressions. Among consistently-signed estimates, at the median, the proportion between
+effect sizes for siblings and for children is `r 100 * median_prop_gens` per cent.
 
@@
 
-```{r plot-sexes, fig.cap = "Effect sizes on number of children by sex"}
+```{r plot-sexes, fig.cap = "Effect sizes on number of children by sex. Lines show significant differences"}
 
@@
 
+res_sex_wide <- res_sex %>% 
+      select(term, estimate, sex, diff.p.value) %>% 
+      pivot_wider(names_from = "sex", values_from = "estimate") %>% 
+      mutate(
+        min_est = pmin(Female, Male),
+        max_est = pmax(Female, Male),
+        term = pretty_names(term),
+        term = fct_reorder(term, Female)
+      )
+
 res_sex %>% 
@@
         term = pretty_names(term),
-        term = fct_reorder(term, estimate, order_abs(1))
       ) %>% 
@@
         my_vline +
+        geom_linerange(data = res_sex_wide, 
+          aes(
+            y      = term,
+            x      = NULL,
+            fill   = NULL,
+            xmin   = min_est, 
+            xmax   = max_est, 
+            colour = diff.p.value < 0.05/{{n_regs}}
+          ), 
+        group = "") +
         pval_geom_point() +
@@
 n_regs <- as.double(nrow(res_edu))
+
 res_edu %>% 
@@
         term = fct_reorder(term, estimate, order_abs(1)),
-        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18")
+        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
+        "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
       ) %>% 
@@
           labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
-        )
+        ),
+        Income = add_n(Income, "income_cat")
       ) %>% 
@@
         score_name = fct_reorder(score_name, estimate, order_abs(1)),
-        term       = gsub("age_flb_cat(.*):.*", "\\1", term)
+        term       = gsub("age_flb_cat(.*):.*", "\\1", term),
+        term       = add_n(term, "age_flb_cat")
       ) %>% 
@@
         my_fill_scale(n = 3, direction = -1, 
-          guide = guide_legend(title = "Age at FLB")) +
+          guide = guide_legend(title = "Age at first live birth")) +
         my_labs() +
@@
       ) %>% 
-      ggplot(aes(y = reorder_within(score_name, estimate, sex, fun = order_abs(1)), x = estimate, fill = `N partners`, 
-        colour = p.value < 0.05/{{n_regs}})) + 
+      ggplot(aes(
+        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        x      = estimate, 
+        fill   = `N partners`, 
+        colour = p.value < 0.05/{{n_regs}}
+      )) + 
         my_vline +
@@
 n_regs <- nrow(res_with_partner)
+hhtype_n <- n_in_regs("with_partner")
 
@@
       mutate(
-        Household = ifelse(grepl("TRUE", term), "With partner", "Without partner")
+        Household = ifelse(grepl("TRUE", term), "With partner", 
+                      "Without partner"),
+        with_p    = as.character(grepl("TRUE", term)),
+        Household = sprintf("%s (N = %s)", Household, 
+                      prettyNum(hhtype_n[with_p], big.mark = ","))
       ) %>% 
@@
         "EA3 Ventile" = chop_equally(EA3_excl_23andMe_UK, 20, labels = 1:20),
-        "N Children"  = santoku::chop(n_children, brk_left(0:4, FALSE), 
-                          labels = c(0:3, "4+")),
+        "N Children"  = santoku::chop(n_children, 0:4, labels = c(0:3, "4+")),
         Sex           = factor(sex, levels = 0:1, labels = c("Female", "Male"))

### 2020-06-12 make-data.R

@@ -119,7 +119,8 @@ edit_famhist <- function (famhist, score_names) {
       negative_to_na
     )
   )
-  
+# TEMPORARY hack. TODO: use f.6138
+famhist$age_fulltime_edu[is.na(famhist$age_fulltime_edu)]  <- 21 
   famhist$income_cat <- famhist$f.738.0.0
   
   # roughly speaking, these are ages in 2007-10
@@ -173,6 +174,8 @@ edit_famhist <- function (famhist, score_names) {
           famhist$f.2754.0.0, famhist$f.2754.1.0, famhist$f.2754.2.0,
           na.rm = TRUE
         )
+  famhist$age_flb_cat <- santoku::chop_equally(famhist$age_flb, 3, 
+                                               labels = lbl_discrete("-"))
   famhist$age_llb <- pmax(
           famhist$f.2764.0.0, famhist$f.2764.1.0, famhist$f.2764.2.0,
           na.rm = TRUE
@@ -200,11 +203,11 @@ edit_famhist <- function (famhist, score_names) {
 
 
 make_famhist_long_mlogit <- function (famhist, score_names) {
-  famhist$n_ch_fac <- santoku::chop(famhist$n_children, 
-        brk_left(0:5, close_end = FALSE), lbl_integer())
+  famhist$n_ch_fac <- santoku::chop(famhist$n_children, 0:5, 
+                                      labels = santoku::lbl_discrete())
   fh_subset <- famhist %>% select(n_ch_fac, all_of(score_names))
   mlogit::mlogit.data(fh_subset, choice = "n_ch_fac", shape = "wide", 
-        alt.levels = levels(fh_subset$n_ch_fac))
+                        alt.levels = levels(fh_subset$n_ch_fac))
 }
 
 

### 2020-06-16 why-negative-selection.Rmd
work and notes on census weighting. Some simplifications in the text.

@@
 number of children. This is not because the strength of natural selection has
-changed over time: effect sizes do not differ between early- and late-born
-respondents, within either children or siblings. Instead, the difference is
-probably due to ascertainment bias in UK Biobank. Indeed, natural selection is
-concentrated among population subgroups that are under-represented in the
-sample, including less-educated people and poorer households. When we weight 
-the sample to correct for ascertainment bias, many effect sizes are considerably 
-larger.
+changed over time. Instead, the difference is probably due to ascertainment bias
+in UK Biobank. Indeed, natural selection is concentrated among population
+subgroups that are under-represented in the sample, including less-educated
+people and poorer households. When we weight the sample to correct for
+ascertainment bias, many effect sizes are considerably larger.
 
 Among mothers who were older at first birth, and among parents with fewer sexual
-partners, effect signs are reversed so that natural selection is broadly
-"positive". The effects of PGS can be decomposed into two channels: an effect on
-age at first live birth, and an opposite-signed effect on number of children
-conditional on age at first live birth. This pattern may be explained by
-economic theories of fertility. In these, higher potential earnings have two
-opposite effects on fertility: a fertility-increasing income effect (higher
-income makes children more affordable), and a fertility-lowering substitution
-effect (time spent on childrearing has a higher cost in foregone earnings).
-Among couples who can take advantage of coordination in childcare activities,
-the income effect dominates; among single parents or those in unstable
-relationships, the substitution effect dominates.
+partners, effect signs are reversed. The effects of PGS can be decomposed into
+two channels: an effect on age at first live birth, and an opposite-signed
+effect on number of children conditional on age at first live birth. This
+pattern may be explained by economic theories of fertility. In these, higher
+potential earnings have two opposite effects on fertility: a
+fertility-increasing income effect (higher income makes children more
+affordable), and a fertility-lowering substitution effect (time spent on
+childrearing has a higher cost in foregone earnings). Among couples who can take
+advantage of coordination in childcare activities, the income effect dominates;
+among single parents, or those in unstable relationships, the substitution effect
+dominates.
 
@@
   - Check for existing work
+  - Chris Hanretty suggested using 2011 census. Household size/type would also
+    be good.
 * Reweight other regressions (e.g. the omnibus regression in the appendix)
@@
       mutate(
+        score_name = pretty_names(score_name),
         score_name = fct_reorder(score_name, estimate, order_abs(2)),

### 2020-06-16 _drake.R

@@ -25,6 +25,8 @@ rgs_file      <- file.path(data_dir, "EA3_rgs.10052019.rgs.csv")
 mf_pairs_file <- file.path(data_dir, "spouse_pair_info/UKB_out.mf_pairs_rebadged.csv")
 nomis_file    <- file.path(data_dir, "nomis-2011-statistics.csv")
 ghs_file      <- file.path(data_dir, "UKDA-5804-stata8/stata8/Ghs06client.dta")
+DC1108EW_file <- file.path(data_dir, "DC1108EW.csv")
+DC5202SC_file <- file.path(data_dir, "DC5202SC.csv")
 
 plan <- drake_plan(
   
@@ -72,6 +74,14 @@ plan <- drake_plan(
                     famhist
                   ),
   
+  census_age_qual = target(
+                      make_census_age_qual(
+                        file_in(!! DC1108EW_file),
+                        file_in(!! DC5202SC_file)
+                      ), 
+                      format = "fst"
+                    ),
+  
   flb_weights = {
     weight_by_ghs(
             ~ age_at_recruitment + factor(flb_cat) + factor(edu_qual), 
@@ -82,14 +92,14 @@ plan <- drake_plan(
   
   parent_weights = weight_parents(famhist, ghs_weights),
   
-  mf_pairs     = target(
-                   make_mf_pairs(file_in(!! mf_pairs_file), famhist), 
-                   format = "fst"
-                 ), 
+  mf_pairs = target(
+                     make_mf_pairs(file_in(!! mf_pairs_file), famhist), 
+                     format = "fst"
+                   ), 
   
-  rgs          = make_rgs(file_in(!! rgs_file)),
+  rgs = make_rgs(file_in(!! rgs_file)),
   
-  res_all      = {
+  res_all = {
     famhist <- join_famhist_resid_scores(famhist, resid_scores)
     res_sibs <- run_regs_basic(
             dep_var     = "n_sibs", 
@@ -108,7 +118,7 @@ plan <- drake_plan(
           )
   },
   
-  res_pcs     = {
+  res_pcs = {
     pc_names <- grep("PC", names(pcs), value = TRUE)
     res_sibs_pcs <- run_regs_pcs( 
       dep_var     = "n_sibs", 

### 2020-06-16 make-data.R

@@ -338,3 +338,87 @@ make_ghs_subset <- function(ghs_file) {
   ghs_subset
 }
 
+
+make_census_age_qual <- function (DC1108EW_file, DC5202SC_file) {
+  ew_age_qual <- readr::read_csv(DC1108EW_file, col_names = FALSE)
+  # cols 1-3 are date, geography and geography code. We only want 
+  # row 1: "England and Wales"
+  ew_age_qual <- t(ew_age_qual)
+  ew_age_qual %<>% 
+        as.data.frame(stringsAsFactors = FALSE) %>% 
+        select(name = 1, value = 2) %>% 
+        slice(-(1:3)) %>% 
+        filter(
+          ! grepl("All categories", name, fixed = TRUE)
+        ) %>% 
+        tidyr::separate(name, 
+          into = c("age_cat", "qual", "ethnic", "measure"), 
+          sep  = ";"
+        ) %>% 
+        select(-measure) %>% 
+        mutate(
+          age_cat = sub("Age: Age ", "", age_cat),
+          qual    = sub(".*: ", "", qual),
+          qual    = sub("(Level.*) qualifications", "\\1", qual),
+          ethnic  = sub("Ethnic Group: *", "", ethnic)
+        ) %>% 
+        filter(ethnic == " White: Total") %>% 
+        select(-ethnic) %>% 
+        mutate(
+          value = as.numeric(value)
+        )
+  
+  # alternatively, "Ethnic Group: White: English" would
+  # capture English/Welsh/Scottish/Northern Irish/British only
+  
+  # England splits up age categories to 65-74, 75+. Scotland doesn't.
+  ew_age_qual %<>% 
+        tidyr::pivot_wider(
+          id_cols     = qual,
+          names_from  = age_cat, 
+          values_from = value
+        ) %>% 
+        mutate(
+          `65 and over` = `65 to 74` + `75 and over`
+        ) %>% 
+        select(- `65 to 74`, - `75 and over`) %>% 
+        tidyr::pivot_longer(
+          -qual,
+          names_to  = "age_cat",
+          values_to = "value"
+        ) %>% 
+        mutate(
+          place = "England and Wales"
+        )
+  
+  sc_age_qual <- readr::read_csv(DC5202SC_file, skip = 12)
+  sc_age_qual %<>% 
+        select(
+          age_cat = `X1`,              # new names reflect values in the columns
+          qual    = `Ethnicity (Flat)`, 
+          value   = `White: Total`
+        ) %>% 
[…]

### 2020-06-18 why-negative-selection.Rmd
Add census weighting and use in paper. Rework _drake to use "static branching".

@@
   if (! missing(score_col)) score_col <- enquo(score_col)
-  fill_col <- if (! missing(fill_col)) enquo(fill_col) else quo(NULL)
+  mfc <- missing(fill_col)
+  fill_col <- if (mfc) quo(NULL) else enquo(fill_col)
   if (missing(n_cats)) {
-    n_cats <- if (missing(fill_col)) 1 else length(unique(pull(dfr, 
-                {{fill_col}})))
+    n_cats <- if (mfc) 1 else length(unique(pull(dfr, {{fill_col}})))
   }
@@
   by figure 1. 
-* Need to make the weighting very watertight. 
-  - Include income? 
-  - Interact age and other variables?
-  - Investigate "MRP" (Gelman 2005)
-  - Understand reweighting on subsets of survey data.
-  - Check for existing work
-  - Chris Hanretty suggested using 2011 census. Household size/type would also
-    be good.
 * Reweight other regressions (e.g. the omnibus regression in the appendix)
@@
   which may explain some of the many NAs for f.2141, num sex partners)
-*  Need f.6138 to correctly calculate `age_fulltime_edu` including
-  those who went to college.
 * Look at geography, esp. in siblings regressions.
@@
 
-drake::loadd(res_weighted)
+drake::loadd(res_wt_ghs_weights)
+drake::loadd(res_wt_flb_weights)
+drake::loadd(res_wt_msoa_weights)
+drake::loadd(res_wt_age_qual_weights)
 drake::loadd(res_sibs_parent_weights)
 drake::loadd(res_all)
+
+res_wt_combined <- bind_rows(
+  None                     = res_all %>% 
+                               filter(reg.type == "raw", dep.var == "N children") %>% 
+                               select(-reg.type, -dep.var, -score_name),
+  Geographical             = res_wt_msoa_weights,
+  `Age/Qualification`      = res_wt_age_qual_weights,
+  `Age/Qual/FLB`  = res_wt_flb_weights,
+  .id = "Weights"
+)
+
+res_wt_cb_wide <- res_wt_combined %>% 
+                    tidyr::pivot_wider(
+                      id_cols     = term,
+                      names_from  = Weights,
+                      values_from = estimate
+                    ) %>% 
+                    rowwise() %>% 
+                    mutate(
+                      consistent = all(c_across(-term) > 0) ||
+                        all(c_across(-term) < 0)
+                    ) %>% 
+                    ungroup() %>% 
+                    left_join(
+                      res_wt_combined %>% 
+                        filter(Weights == "None") %>% 
+                        select(term, p.value)
+                    )
+
+rel_sizes <- res_wt_cb_wide %>% 
+                filter(
+                  consistent, 
+                  p.value < 0.05/length(score_names)
+                ) %>% 
+                mutate(
+                  across(Geographical:`Age/Qual/FLB`, ~ .x/None)
+                ) %>% 
+                select(- consistent, - None, - p.value) %>% 
+                mutate(
+                  term = pretty_names(term)
+                )
+
+rel_sizes_summary <- rel_sizes %>% 
+                summarize(
+                  across(Geographical:`Age/Qual/FLB`, ~ c(mean(.x), median(.x)))
+                ) %>% 
+                mutate(
+                  term = c("Mean", "Median")
+                )
+
+# TODO: move later
 res_weighted_all <- bind_rows(
-        "N children" = res_weighted, 
+        "N children" = res_wt_ghs_weights, 
         "N siblings" = res_sibs_parent_weights, 
@@
 res_wide <- full_join(
-        res_weighted,
+        res_wt_ghs_weights,
         res_sibs_parent_weights,
@@
 
-To compensate for ascertainment bias in the UK Biobank sample, we weight participants
-by sex, age (within the 40-71 age group), and education. Our sample weights are
-calibrated based on the 2006 General Household Survey. We then calculate the 
+To compensate for ascertainment bias in the UK Biobank sample, we weight 
+participants using data from the 2011 Census and from the 2006 General Household
+Survey (GHS). We try three alternative weighting schemes:
+
+* Geographical weighting by census MSOA, age and presence/absence of a partner;
+* Weighting by age and highest educational qualification;
+* Weighting by age, highest qualification, and age at first live birth, using
+  GHS data.
+
+Figure \@ref(fig:plot-res-wt) shows the results. Effect sizes are significantly
+larger in absolute value. Table \@ref(tab:tbl-res-wt) gives effect sizes as a 
+proportion of unweighted effect sizes.
+
+```{r tbl-res-wt}
+rel_sizes %>% 
+      arrange(desc(`Age/Qual/FLB`)) %>% 
+      bind_rows(rel_sizes_summary) %>% 
+      rename(PGS = term) %>% 
+      as_hux() %>% 
+      set_caption(
+        "Weighted effect sizes as a proportion of unweighted"
+      ) %>% 
+      insert_row("", "Weighting", fill = "") %>% 
+      merge_cells(1, 2:4) %>% 
+      set_align(1, 2, "centre") %>% 
+      set_header_rows(1, TRUE) %>% 
+      theme_compact() %>% 
+      style_headers(bold = TRUE) %>% 
+      set_bottom_border(1, 2:4) %>% 
+      set_bottom_border(2, everywhere) %>% 
+      set_background_color(final(2), everywhere, "grey80") %>% 
+      add_footnote(
+        "Only consistent and significant (when unweighted) estimates are shown",
+        font_size   = 8, 
+        top_padding = 2
+      )
+
+```
+
+```{r plot-res-wt, fig.cap = "Effects of polygenic scores on fertility. Weighted regressions."}
+
+res_wt_combined %>% 
+  mutate(
+    Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
+                "Age/Qual/FLB", "None"),
+    term     = pretty_names(term),
+    term     = fct_reorder(term, estimate, order_abs(4))
+  ) %>% 
+  ggplot(aes(x = estimate, y = term, shape = Weights, colour = Weights)) +
+  geom_point(size = 2) +
+  my_vline +
+  scale_colour_manual(values = c(
+    "None"              = "grey25", 
+    "Geographical"      = "darkgreen",
+    "Age/Qualification" = "chocolate4",
+    "Age/Qual/FLB"      = "midnightblue"
+  )) +
+  scale_shape_manual(values = c("None" = "square", "Geographical" = "circle",
+        "Age/Qualification" = "circle", "Age/Qual/FLB" = "circle"))
+
+```
+
+```{r plot-res-weighted, fig.cap = "Effects of polygenic scores on number of children/siblings. Regressions weighted by sex, age and education."}
+
+res_weighted_all %>% 
+      rename(
+        "Dependent variable" = dep.var, 
+      ) %>% 
+      filter(reg.type == "weighted") %>% 
+      standard_ggplot(fill = `Dependent variable`, n_regs = 33, order_idx = 1, 
+        shape = reg.type) + 
+        scale_shape_manual(values = c(21, 4))
+
+```
+
+We then calculate the 
 proportions of respondents within each sibling group size, based on the respondent
@@
 
-```{r plot-res-weighted, fig.cap = "Effects of polygenic scores on number of children/siblings. Regressions weighted by sex, age and education."}
-
-res_weighted_all %>% 
-      rename(
-        "Dependent variable" = dep.var, 
-      ) %>% 
-      filter(reg.type == "weighted") %>% 
-      standard_ggplot(fill = `Dependent variable`, n_regs = 33, order_idx = 1, 
-        shape = reg.type) + 
-        scale_shape_manual(values = c(21, 4))
-
-```
-
 ```{r calc-flb-weights}
 
-drake::loadd(res_flb_weights)
+drake::loadd(res_wt_flb_weights)
 drake::loadd(res_sex)
 res_flb_all <- left_join(
-        res_flb_weights, 
+        res_wt_flb_weights, 
         res_sex %>% filter(sex=="Female") %>% select(-subset, -sex), 
@@
 bind_rows(
-        Weighted   = res_flb_weights, 
+        Weighted   = res_wt_flb_weights, 
         Unweighted = res_sex %>% filter(sex == "Female") %>% select(-subset, -sex),

### 2020-06-18 _drake.R

@@ -9,27 +9,37 @@ suppressPackageStartupMessages({
 
 
 source("make-data.R")
+source("make-census-weights.R")
 source("regressions.R")
 source("weight-data.R")
 
-data_dir      <- "../negative-selection-data"
-pgs_dir       <- file.path(data_dir, "polygenic_scores/")
-ph_file       <- file.path(data_dir, "UKB.EA_pheno.coordinates.QC.david.csv")
-pcs_file      <- file.path(data_dir, "UKB.HM3.100PCs.40310.txt")
-famhist_file  <- file.path(data_dir, "david.family_history.traits.out.csv")
-famhist2_file <- file.path(data_dir, "david.family_history.traits.20042020.out.csv")
-famhist3_file <- file.path(data_dir, "david.family_history.traits.05052020.out.csv")
-famhist4_file <- file.path(data_dir, "david.family_history.traits.16052020.out.csv")
-famhist5_file <- file.path(data_dir, "david.family_history.traits.18052020.out.csv")
-rgs_file      <- file.path(data_dir, "EA3_rgs.10052019.rgs.csv")
-mf_pairs_file <- file.path(data_dir, "spouse_pair_info/UKB_out.mf_pairs_rebadged.csv")
-nomis_file    <- file.path(data_dir, "nomis-2011-statistics.csv")
-ghs_file      <- file.path(data_dir, "UKDA-5804-stata8/stata8/Ghs06client.dta")
-DC1108EW_file <- file.path(data_dir, "DC1108EW.csv")
-DC5202SC_file <- file.path(data_dir, "DC5202SC.csv")
+data_dir           <- "../negative-selection-data"
+pgs_dir            <- file.path(data_dir, "polygenic_scores/")
+ph_file            <- file.path(data_dir, "UKB.EA_pheno.coordinates.QC.david.csv")
+pcs_file           <- file.path(data_dir, "UKB.HM3.100PCs.40310.txt")
+famhist_file       <- file.path(data_dir, "david.family_history.traits.out.csv")
+famhist2_file      <- file.path(data_dir, "david.family_history.traits.20042020.out.csv")
+famhist3_file      <- file.path(data_dir, "david.family_history.traits.05052020.out.csv")
+famhist4_file      <- file.path(data_dir, "david.family_history.traits.16052020.out.csv")
+famhist5_file      <- file.path(data_dir, "david.family_history.traits.18052020.out.csv")
[…]

### 2020-06-18 make-census-weights.R

@@ -1,0 +1,159 @@
+
+suppressPackageStartupMessages({
+  loadNamespace("readr")
+  loadNamespace("tidyr")
+  library(dplyr)
+  library(sf)
+})
+
+make_census_age_qual <- function (DC1108EW_file, DC5202SC_file) {
+  ew_age_qual <- readr::read_csv(DC1108EW_file, col_names = FALSE)
+  # cols 1-3 are date, geography and geography code. We only want 
+  # row 1: "England and Wales"
+  ew_age_qual <- t(ew_age_qual)
+  ew_age_qual %<>% 
+    as.data.frame(stringsAsFactors = FALSE) %>% 
+    select(name = 1, value = 2) %>% 
+    slice(-(1:3)) %>% 
+    filter(
+      ! grepl("All categories", name, fixed = TRUE)
+    ) %>% 
+    tidyr::separate(name, 
+                    into = c("age_cat", "qual", "ethnic", "measure"), 
+                    sep  = ";"
+    ) %>% 
+    select(-measure) %>% 
+    mutate(
+      age_cat = sub("Age: Age ", "", age_cat),
+      qual    = sub(".*: ", "", qual),
+      qual    = sub("(Level.*) qualifications", "\\1", qual),
+      ethnic  = sub("Ethnic Group: *", "", ethnic)
+    ) %>% 
+    filter(ethnic == " White: Total") %>% 
+    select(-ethnic) %>% 
+    mutate(
+      value = as.numeric(value)
+    )
+  
+  # alternatively, "Ethnic Group: White: English" would
+  # capture English/Welsh/Scottish/Northern Irish/British only
+  
+  # England splits up age categories to 65-74, 75+. Scotland doesn't.
+  ew_age_qual %<>% 
+    tidyr::pivot_wider(
+      id_cols     = qual,
+      names_from  = age_cat, 
+      values_from = value
+    ) %>% 
+    mutate(
+      `65 and over` = `65 to 74` + `75 and over`
+    ) %>% 
+    select(- `65 to 74`, - `75 and over`) %>% 
+    tidyr::pivot_longer(
+      -qual,
+      names_to  = "age_cat",
+      values_to = "value"
+    ) %>% 
+    mutate(
+      place = "England and Wales"
+    )
+  
+  sc_age_qual <- readr::read_csv(DC5202SC_file, skip = 12)
+  sc_age_qual %<>% 
+    select(
+      age_cat = `X1`,              # new names reflect values in the columns
+      qual    = `Ethnicity (Flat)`, 
[…]

### 2020-06-18 make-data.R

@@ -6,6 +6,7 @@ suppressPackageStartupMessages({
   library(rlang)
   library(santoku)
   library(haven)
+  library(sf)
   loadNamespace("car") # very annoying if it overwrites recode
   loadNamespace("matrixStats")
   loadNamespace("mlogit")
@@ -57,6 +58,8 @@ make_famhist <- function (
         famhist2_file,
         famhist3_file,
         famhist4_file,
+        famhist5_file,
+        famhist6_file,
         pgs_dir
       ) {
   ph <- read_csv(ph_file, col_types = cols(
@@ -70,6 +73,7 @@ make_famhist <- function (
   famhist3 <- read_csv(famhist3_file)
   famhist4 <- read_csv(famhist4_file, col_types = strrep("d", 33))
   famhist5 <- read_csv(famhist5_file, col_types = strrep("d", 4))
+  famhist6 <- read_csv(famhist6_file, col_types = strrep("d", 22))
   names(famhist) <- paste0("f.", names(famhist))
   names(famhist) <- gsub("\\-", ".", names(famhist))
   names(famhist2) <- paste0("f.", names(famhist2))
@@ -80,12 +84,15 @@ make_famhist <- function (
   names(famhist4) <- gsub("\\-", ".", names(famhist4))
   names(famhist5) <- paste0("f.", names(famhist5))
   names(famhist5) <- gsub("\\-", ".", names(famhist5))
+  names(famhist6) <- paste0("f.", names(famhist6))
+  names(famhist6) <- gsub("\\-", ".", names(famhist6))
   
   famhist %<>% left_join(ph, by = c("f.eid" = "eid"))
   famhist %<>% left_join(famhist2, by = "f.eid")
   famhist %<>% left_join(famhist3, by = "f.eid")
   famhist %<>% left_join(famhist4, by = "f.eid")
   famhist %<>% left_join(famhist5, by = "f.eid")
+  famhist %<>% left_join(famhist6, by = "f.eid")
   
   # only "genetic" whites
   famhist %<>% filter(! is.na(genetic_ethnic_grouping))
@@ -114,13 +121,17 @@ edit_famhist <- function (famhist, score_names) {
       c(age_fulltime_edu, starts_with(c(
         "f.2946", "f.1845", "f.2754", "f.738",  "f.2764", "f.2405", "f.2734",
         "f.2149", "f.1873", "f.1883", "f.2784", "f.2794", "f.709",  "f.3872",
[…]

### 2020-06-18 weight-data.R

@@ -3,6 +3,9 @@ suppressPackageStartupMessages({
   library(survey)
   library(dplyr)
   library(haven)
+  library(forcats)
+  library(santoku)
+  library(magrittr)
 })
 
 
@@ -35,8 +38,8 @@ weight_by_ghs <- function (fml, ghs_subset, famhist) {
 }
 
 
-weight_parents <- function (famhist, ghs_weights) {
-  famhist <- inner_join(famhist, ghs_weights, by = "f.eid")
+weight_parents <- function (famhist, input_weights) {
+  famhist <- inner_join(famhist, input_weights, by = "f.eid")
   n_sibs_are_up_to_8 <- ! is.na(famhist$n_sibs) & famhist$n_sibs <= 8
   n_sibs_up_to_8 <- famhist$n_sibs[n_sibs_are_up_to_8]
   wts <- famhist$weights[n_sibs_are_up_to_8]
@@ -54,6 +57,95 @@ weight_parents <- function (famhist, ghs_weights) {
 }
 
 
-weight_census_age_qual <- function (famhist, census_age_qual) {
+weight_by_census_age_qual <- function (famhist, census_age_qual) {
+  famhist$age_cat <-  santoku::chop(famhist$age_at_recruitment, 
+                        c(16, 25, 35, 50, 65), 
+                        lbl_discrete(" to ")) %>% 
+                      fct_recode("65 and over" = "65 to 73")
+
+  # edu_qual is:
+  # 1	College or University degree
+  # 2	A levels/AS levels or equivalent
+  # 3	O levels/GCSEs or equivalent
+  # 4	CSEs or equivalent
+  # 5	NVQ or HND or HNC or equivalent
+  # 6	Other professional qualifications eg: nursing, teaching
+  # -7 None (but I recoded this to 0)
+  # 
+  # QUAL in the census data is
+  # Apprenticeship: only in EW, what it says
+  # Level 1: 1-4 O levels/CSE/GCSEs, NVQ level 1, foundation GNVQ; et al.
+  # Level 2: 5+ O levels/CSE/GCSEs; 1 A level; or 2-3 AS levels; NVQ level 2; et al
+  # Level 3: 2+ A levels, 4+ AS levels; advanced GNVQ; et al
+  # Level 4 and above: degree or higher; professional qualification e.g. teaching
+  # nursing accountancy; et al.
+  # Other: only in EW.
+  # 
[…]

### 2020-06-19 why-negative-selection.Rmd
More work on article

@@
 
+```
 # Open questions/TODO
@@
 * Reweight other regressions (e.g. the omnibus regression in the appendix)
-* Could be useful: f.2139 - age first had sex (includes "never had sex" 
-  which may explain some of the many NAs for f.2141, num sex partners)
 * Look at geography, esp. in siblings regressions.
 * control for age in n_partners regressions.
+* can we weight men by age at first child, using GHS? If not, may need
+  to plot the age_flb weighted regressions separately, against unweighted
+  women-only regressions.
+* Can we find representative data for weights by n_partners?
+```
 
@@
 # Results
-
-Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores by 5-year
-birth intervals, for the entire sample. Several scores show consistent increases
-or declines over this 30-year period, of the order of 5% of a standard deviation.
 
@@
 ```
+
+Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores by 5-year
+birth intervals, for the entire sample. Several scores show consistent increases
+or declines over this 30-year period, of the order of 5% of a standard deviation.
 
@@
 
-We run bivariate regressions on two dependent variables:
-
-* *siblings*, the number of full siblings in the respondent's sib (including
-  himself or herself).
-* The number of *children* ever born to/fathered by the respondent.
-
-```{r calc-regs-all}
-
-drake::loadd(res_all)
-
-res_all_wide <- res_all %>% 
-  filter(reg.type == "raw") %>% 
-  select(dep.var, term, estimate) %>% 
-  pivot_wider(names_from = dep.var, values_from = estimate)
-
-n_pgs  <- as.double(nrow(res_all_wide))
-
-prop_gens <- res_all_wide[["N siblings"]]/res_all_wide[["N children"]]
-median_prop_gens <- median(prop_gens[prop_gens > 0])
-```
-
-Figure \@ref(fig:plot-basic-regs) shows effect sizes of a one-standard
-deviation shift in each polygenic score.
-
-Estimates are broadly consistent across generations. However, effect sizes are 
-much larger for *siblings* than for *children* 
-regressions. Among consistently-signed estimates, at the median, the proportion between
-effect sizes for siblings and for children is `r 100 * median_prop_gens` per cent.
-
-```{r plot-basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children."}
-
-drake::loadd(res_all)
-
-n_regs <- as.double(nrow(res_all))
-res_all %>% 
-      filter(reg.type == "raw") %>% 
+These estimates may suffer from ascertainment bias within the UK Biobank sample.
+Respondents are higher income and better educated than the UK population, and
+they may also differ on other unobserved characteristics.
+
+To examine the effects of sampling, we regress respondents' polygenic
+scores on their number of children:
+
+\[
+y_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
+\]
+
+The effect size $\beta$ measures the strength of natural selection within the
+sample (or a given subgroup). In fact, since polygenic scores are normalized to
+mean 0, $\beta$ is the expected polygenic score among children of the sample.
+
+Figures \@ref(fig:plot-educ-level) and \@ref(fig:plot-income) plot effect sizes
+for each polygenic score among different subgroups of the sample. Figure
+\@ref(fig:plot-educ-level) groups respondents by age of completing full-time
+education. Figure \@ref(fig:plot-income) groups respondents by household income.
+Effect sizes are larger and more significant for the lowest income category, and
+for the lowest education category. Note that the overall effect is not a simple
+average of the effect among the different subgroups, because polygenic scores 
+may also shift respondents between the subgroups. For example, a high 
+PGS for educational attainment will increase the age at which a respondent 
+leaves school.
+
+```{r plot-educ-level, fig.cap = "Effect sizes on number of children by age left full-time education"}
+
+drake::loadd(res_edu)
+n_regs <- as.double(nrow(res_edu))
+
+res_edu %>% 
       mutate(
@@
         term = fct_reorder(term, estimate, order_abs(1)),
-        dep.var = fct_inorder(dep.var) 
-      ) %>% 
-      ggplot(aes(y = term, x = estimate, fill = dep.var, 
+        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
+        "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
+      ) %>% 
+      ggplot(aes(y = term, x = estimate, colour = p.value < 0.05/{{n_regs}}, 
+        fill = `Age left FTE`)) +
+        my_vline +
+        pval_geom_point() + 
+        pval_scale() +
+        my_fill_scale(n = 3, direction = -1) +
+        my_labs() +
+        my_theme()
+
+```
+
+```{r plot-income, fig.cap = "Effect sizes on number of children by household income"}
+
+drake::loadd(res_income)
+
+n_regs <- as.double(nrow(res_income))
+res_income %>% 
+      mutate(
+        term = pretty_names(term), 
+        term = fct_reorder(term, estimate, order_abs(1)),
+        Income = factor(income_cat, 
+          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
+        ),
+        Income = add_n(Income, "income_cat")
+      ) %>% 
+      ggplot(aes(y = term, x = estimate, fill = Income, 
         colour = p.value < 0.05/{{n_regs}})) +
         my_vline +
-      pval_geom_point() +
+        pval_geom_point(alpha = 0.8) + 
+        my_fill_scale(n = 5, direction = -1) +
+        pval_scale() +
         my_labs() +
-      pval_scale() +
-      my_fill_scale(n = 2, guide = guide_legend(title = "")) +
         my_theme()
-
-```
-
-# Subgroups
-
-We next examine how different subgroups of the population contribute to natural 
-selection. These analyses have two goals: to get a sense of how ascertainment bias
-could affect our basic results, and to learn more about the causes of natural 
-selection in the UK population. Because we have more data on respondents than on
-their parents, we focus on *children* regressions.
-
-## Sex, income and education
+```
+
+```{r calc-income-edu-controlled}
+drake::loadd(res_income_controlled, res_edu_controlled)
+n_regs_ic <- nrow(res_income_controlled)
+n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
+
+n_regs_educ <- nrow(res_edu_controlled)
+n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
+```
+
+These results could be driven by age, if older respondents are poorer and less
+educated, and also more subject to selection on polygenic scores. However, if we
+rerun the regressions, interacting the polygenic score with income category and
+also with a quadratic in age, the interaction with income remains significant at
+0.05/`r n_regs_ic` for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
+if we interact the PGS with age of leaving full time education and a quadratic
+in age, the interaction with age leaving FTE remains significant at 
+0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.
 
@@
 
-Figure \@ref(fig:plot-educ-level) splits respondents up by education levels.
-Selection effects are typically larger and more significant for those who left
-school before 16.  Figure \@ref(fig:plot-income) splits respondents by household
-income category. A very similar pattern holds, with selection effects being
-larger for those in the poorest income category.
-
-```{r plot-educ-level, fig.cap = "Effect sizes on number of children by age left full-time education"}
-
-drake::loadd(res_edu)
-n_regs <- as.double(nrow(res_edu))
-
-res_edu %>% 
-      mutate(
-        term = pretty_names(term), 
-        term = fct_reorder(term, estimate, order_abs(1)),
-        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
-        "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
-      ) %>% 
-      ggplot(aes(y = term, x = estimate, colour = p.value < 0.05/{{n_regs}}, 
-        fill = `Age left FTE`)) +
-        my_vline +
-        pval_geom_point() + 
-        pval_scale() +
-        my_fill_scale(n = 3, direction = -1) +
-        my_labs() +
-        my_theme()
-
-```
-
-```{r plot-income, fig.cap = "Effect sizes on number of children by household income"}
-
-drake::loadd(res_income)
-
-n_regs <- as.double(nrow(res_income))
-res_income %>% 
-      mutate(
-        term = pretty_names(term), 
-        term = fct_reorder(term, estimate, order_abs(1)),
-        Income = factor(income_cat, 
-          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
-        ),
-        Income = add_n(Income, "income_cat")
-      ) %>% 
-      ggplot(aes(y = term, x = estimate, fill = Income, 
-        colour = p.value < 0.05/{{n_regs}})) +
-        my_vline +
-        pval_geom_point(alpha = 0.8) + 
-        my_fill_scale(n = 5, direction = -1) +
-        pval_scale() +
-        my_labs() +
-        my_theme()
-```
-
-```{r calc-income-edu-controlled}
-drake::loadd(res_income_controlled, res_edu_controlled)
-n_regs_ic <- nrow(res_income_controlled)
-n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
-
-n_regs_educ <- nrow(res_edu_controlled)
-n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
-```
-
-These results could be driven by age, if older respondents are poorer and less
-educated, and also more subject to selection on polygenic scores. However, if we
-rerun the regressions, interacting the polygenic score with income category and
-also with a quadratic in age, the interaction with income remains significant at
-0.05/`r n_regs_ic` for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
-if we interact the PGS with age of leaving full time education and a quadratic
-in age, the interaction with age leaving FTE remains significant at 
-0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.
-
-The UK Biobank sample overrepresents highly educated people and high-income
-households. This suggests that statistics from the sample may underestimate
-the level of natural selection in the population.
-
-## Age at first live birth
+When we split respondents up by other demographic variables, we get a more
+surprising result: the direction of natural selection is different among
+different subgroups.
 
@@
 
-Age at first live birth correlates with number of children ever born. Figure
-\@ref(fig:plot-age-flb-cross) shows *children* regressions estimated separately
-for each tercile of age at first live birth. Several effects are strikingly
+Figure \@ref(fig:plot-age-flb-cross) shows effect sizes estimated separately for
+each tercile of age at first live birth. Several effects are strikingly
 different across terciles. ADHD and MDD are selected for amongst the youngest
@@
 between effect sizes for the youngest and oldest terciles is `r cor_age_flb`.
-This is the first sign that selection may work in different directions across
-different subgroups.
 
@@
 across the groups. Among men and women with more than 3 sexual partners, for
-example, EA is selected against; among those with 3 or fewer sexual partners, it
-is selected for. Selection is broadly negative (positive) among those 
-with more (less) partners.
-
-Lastly Figure \@ref(fig:plot-with-partner) splits respondents up by whether they
+example, educational attainment is selected against; among those with 3 or fewer
+sexual partners, it is selected for.
+
+Figure \@ref(fig:plot-with-partner) splits respondents up by whether they
 were living with a spouse or partner at the time of interview. Effects are 
@@
 
-# Accounting for ascertainment bias
-
-Our results are subject to two kinds of ascertainment bias. First, Biobank is
-not a random sample of the UK population: for example, it overrepresents
-educated people and high income households. Second, sampling parents of
-respondents inherently overweights parents who have many children. For instance,
-parents of three children will have three times more children as respondents in
-UK Biobank than parents of one child, on average. Parents of no children will,
-by definition, not be represented at all.
+```{r calc-corr-age-flb}
+
+drake::loadd(res_age_flb)
+drake::loadd(res_sex)
+
+res_sex_comparison <- res_sex %>% 
+      filter(sex == "Female") %>% 
+      mutate(score_name = term) %>% 
+      select(score_name, term, estimate:conf.high)
+
+res_age_flb %<>% filter(term != "age_flb")
+
+res_combined <- bind_rows(
+        raw      = res_sex_comparison, 
+        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
+        .id      = "reg.type"
+      ) %>% 
+      arrange(score_name)
+
+raw_flb_comparison <- res_combined %>% 
+      pivot_wider(
+        id_cols     = score_name, 
+        names_from  = reg.type, 
+        values_from = estimate
+      )
+
+cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
+opp_signed <- sum(
+        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
+      )
+
+```
+
+Figure \@ref(fig:plot-age-flb) shows the results of regressions for women only,
+controlling for age at first live birth. Effect sizes are greatly reduced.
+This could be because the regressions exclude childless women. However, in 
+`r opp_signed` out of `r length(score_names)` cases, effects actually change
+sign when controls are added. The correlation between effect sizes controlling for age
+at first live birth, and raw effect sizes, is `r cor_raw_flb`.
+
+```{r plot-age-flb, fig.cap = "Effect sizes on number of children, controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}
+
+n_regs <- as.double(nrow(res_age_flb))
+
+res_combined %>% 
+      mutate(
+        score_name = pretty_names(score_name),
+        score_name = fct_reorder(score_name, estimate, order_abs(2)),
+        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for age FLB")
+      ) %>% 
+      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
+        color = p.value < 0.05/{{n_regs}})) + 
+        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
+        pval_geom_point() +
+        pval_scale() +
+        scale_fill_brewer(type = "qual", palette = 3) +
+        my_labs() +
+        my_theme()
+```
+
+## Weighted regressions
 
@@
 drake::loadd(res_all)
+drake::loadd(res_sex)
 
@@
                         filter(Weights == "None") %>% 
-                        select(term, p.value)
+                        select(term, p.value),
+                      by = "term"
+                    ) %>% 
+                    left_join(
+                      res_sex %>% 
+                        filter(sex == "Female") %>% 
+                        select(term, estimate_females = estimate),
+                      by = "term"
                     )
@@
                 mutate(
-                  across(Geographical:`Age/Qual/FLB`, ~ .x/None)
-                ) %>% 
-                select(- consistent, - None, - p.value) %>% 
+                  Geographical = Geographical/None,
+                  `Age/Qualification` = `Age/Qualification`/None,
+                  `Age/Qual/FLB` = `Age/Qual/FLB`/estimate_females
+                ) %>% 
+                select(-consistent, -None, -p.value, -estimate_females) %>% 
                 mutate(
@@
                 summarize(
-                  across(Geographical:`Age/Qual/FLB`, ~ c(mean(.x), median(.x)))
+                  across(-term, ~ c(mean(.x), median(.x)))
                 ) %>% 
@@
 * Weighting by age, highest qualification, and age at first live birth, using
-  GHS data.
-
-Figure \@ref(fig:plot-res-wt) shows the results. Effect sizes are significantly
-larger in absolute value. Table \@ref(tab:tbl-res-wt) gives effect sizes as a 
-proportion of unweighted effect sizes.
+  GHS data (women only).
+
+Figure \@ref(fig:plot-res-wt) plots effect sizes among the entire sample, using
+these weighting schemes. We include unweighted regressions for comparison.
+Weighted effect sizes are significantly larger in absolute value. Table
+\@ref(tab:tbl-res-wt) gives effect sizes as a proportion of unweighted effect
+sizes for the census-based weightings.
+
+Effect sizes are reasonably consistent across the geographical and
+age/qualification weights. But they are increased again when we weight by age at
+first live birth.This suggests that estimates might be increased still more by
+weighting on other relevant demographic variables. Overall, these results
+suggest that previous work may have underestimated the level of selection
+occurring in the population. [TODO: who?]
 
 ```{r tbl-res-wt}
+
 rel_sizes %>% 
-      arrange(desc(`Age/Qual/FLB`)) %>% 
+      arrange(desc(Geographical)) %>% 
       bind_rows(rel_sizes_summary) %>% 
@@
       as_hux() %>% 
+      insert_row("", "Weighting", fill = "") %>% 
       set_caption(
@@
       ) %>% 
-      insert_row("", "Weighting", fill = "") %>% 
       merge_cells(1, 2:4) %>% 
@@
       set_bottom_border(2, everywhere) %>% 
-      set_background_color(final(2), everywhere, "grey80") %>% 
-      add_footnote(
-        "Only consistent and significant (when unweighted) estimates are shown",
-        font_size   = 8, 
-        top_padding = 2
+      set_italic(final(2), 1) %>% 
+      set_font_size(8) %>% 
+      set_width(0.7) %>% # ensures line wrapping below
+      set_col_width(c(.34, .22, .22, .22)) %>% 
+      add_footnote(paste0(
+        "Only consistent and significant (when unweighted) estimates ",
+        "are shown.\n",
+        "Age/Qual/FLB as a proportion of unweighted regressions ",
+        "for females only."),
+        font_size    = 6, 
+        top_padding  = 0,
+        left_padding = 0
       )
@@
 
+n_regs <- length(score_names)
 res_wt_combined %>% 
@@
     term     = pretty_names(term),
-    term     = fct_reorder(term, estimate, order_abs(4))
-  ) %>% 
-  ggplot(aes(x = estimate, y = term, shape = Weights, colour = Weights)) +
-  geom_point(size = 2) +
+    term     = fct_reorder(term, estimate, order_abs(1))
+  ) %>% 
+  ggplot(aes(x = estimate, y = term, shape = Weights, fill = Weights, colour = p.value < 0.05/{{n_regs}})) +
+  geom_point(size = 1.5, stroke = 0.5) +
   my_vline +
-  scale_colour_manual(values = c(
-    "None"              = "grey25", 
-    "Geographical"      = "darkgreen",
-    "Age/Qualification" = "chocolate4",
-    "Age/Qual/FLB"      = "midnightblue"
+  pval_scale() +
+  scale_fill_manual(values = c(
+    "None"              = "grey15", 
+    "Geographical"      = "dodgerblue4",
+    "Age/Qualification" = "steelblue2",
+    "Age/Qual/FLB"      = "darkseagreen4"
   )) +
-  scale_shape_manual(values = c("None" = "square", "Geographical" = "circle",
-        "Age/Qualification" = "circle", "Age/Qual/FLB" = "circle"))
-
-```
-
-```{r plot-res-weighted, fig.cap = "Effects of polygenic scores on number of children/siblings. Regressions weighted by sex, age and education."}
-
-res_weighted_all %>% 
-      rename(
-        "Dependent variable" = dep.var, 
-      ) %>% 
-      filter(reg.type == "weighted") %>% 
-      standard_ggplot(fill = `Dependent variable`, n_regs = 33, order_idx = 1, 
-        shape = reg.type) + 
-        scale_shape_manual(values = c(21, 4))
-
-```
-
-We then calculate the 
-proportions of respondents within each sibling group size, based on the respondent
-weights, and use this to create weights for the sibling regressions.
+  scale_shape_manual(values = c(
+    "None" = "square filled", 
+    "Geographical" = "circle filled",
+    "Age/Qualification" = "circle filled", 
+    "Age/Qual/FLB" = "circle filled"
+  ))
+
+```
+
+## Change over time
+
+The UK Biobank data contains information on respondents' number of siblings.
+Since respondents' polygenic scores are equal in expectation to their parents' 
+mean scores, we can use this to look at natural selection in the parents'
+generation. The dependent variable is now *number of siblings* (including
+the respondent him- or herself).
+
+The parents' generation has a second source of ascertainment bias, beyond the
+sampling bias of the survey. Sampling parents of respondents inherently
+overweights parents who have many children. For instance, parents of three
+children will have, on average, three times more children as respondents in UK
+Biobank than parents of one child. Parents of no children will
+not be represented at all, by definition.
+
+To compensate for this, we reweight our preferred weightings by the inverse of
+*number of siblings*.
+
+TODO: 
+- weight regressions using "best effort" weights
+- look at change over time to confirm your "no change" hypothesis
+- look at siblings, again using "best" weights
 
@@
 
-```{r calc-flb-weights}
-
-drake::loadd(res_wt_flb_weights)
-drake::loadd(res_sex)
-res_flb_all <- left_join(
-        res_wt_flb_weights, 
-        res_sex %>% filter(sex=="Female") %>% select(-subset, -sex), 
-        by = "term", 
-        suffix = c(".wt", ".nowt")
-      ) %>% 
-      mutate(prop_estimate = estimate.wt/estimate.nowt)
-
-prop_wt_nowt <- mean(res_flb_all$prop_estimate)
-pct_change_score <- function (score) {
-  100 * (res_flb_all$prop_estimate[res_flb_all$term == score] - 1)
-}
-```
-
-We also weighted data for women with children only, by age, education and age 
-at first live birth.  Figure \@ref(fig:plot-flb-weights) 
-shows the results. There 
-are large changes for some variables, including EA3 
-(`r pct_change_score("EA3_excl_23andMe_UK")` per cent), cognitive
-ability (`r pct_change_score("cognitive_ability")` per cent) and
-waist-hip ratio (WHR; `r pct_change_score("whr_combined")` percent).
-
-Overall, these results suggest that previous work may have underestimated the level of
-selection occurring in the population. [TODO: who?]
-
-```{r plot-flb-weights, fig.cap = "Effect sizes on number of children, female respondents weighted by age at first live birth"}
-
-bind_rows(
-        Weighted   = res_wt_flb_weights, 
-        Unweighted = res_sex %>% filter(sex == "Female") %>% select(-subset, -sex),
-        .id        = "Regression"
-      ) %>% 
-      standard_ggplot(Regression, n_regs = 66)
-```
-
-## Mechanisms behind natural selection
-
-Why are selection effects *oppositely* signed in certain subgroups -- for
-example, in older and younger mothers (by age at first live birth) or in those
+```{r plot-res-weighted, fig.cap = "Effects of polygenic scores on number of children/siblings. Regressions weighted by sex, age and education."}
+
+res_weighted_all %>% 
+      rename(
+        "Dependent variable" = dep.var, 
+      ) %>% 
+      filter(reg.type == "weighted") %>% 
+      standard_ggplot(fill = `Dependent variable`, n_regs = 33, order_idx = 1, 
+        shape = reg.type) + 
+        scale_shape_manual(values = c(21, 4))
+
+```
+
+## Mechanisms behind modern natural selection
+
+Why do selection effects have *opposite* signs across certain subgroups -- for
+example, across older and younger mothers by age at first live birth, or across those
 with more and fewer lifetime sexual partners? A possible answer is given by the
@@
 variants which improve one's earnings may increase fertility among couples, but
-decrease fertility among single parents.
+decrease fertility among single parents. They may also decrease the likelihood
+of becoming a young mother, if this is correlated with single motherhood. Lastly,
+if lifetime number of sexual partners predicts single parent versus 
+two-parent status, the theory could also explain the differently-signed
+effects across people with more and fewer sexual partners.
 
@@
 We measure the correlation of each PGS with household income, as well as with
-education level, a phenotypic variable which is likely to predict earnings. We
+education level, a phenotype which predicts earnings. We
 then relate these measurements to effect sizes on number of children.
-Across scores, there is a clear negative relationship between effect on
-fertility, and the correlation with income and/or education (Figure \@ref(fig:plot-cors-income-educ).
-
-```{r plot-cors-income-educ, fig.cap="Effect sizes on number of children and correlations with income and education"}
+There is a clear negative relationship between a score's effect on
+fertility, and its correlation with income or education (Figure \@ref(fig:plot-cors-income-educ)). If we control for age at first live birth,
+the relationship changes sign and becomes positive.
+
+```{r plot-cors-income-educ, fig.cap="Effect sizes on number of children, with and without controls for age at first live birth, by correlations with income and education"}
 
 drake::loadd(res_all)
-cors_income <- cor(famhist[c(score_names)], famhist$income_cat, use = "pairwise")
+drake::loadd(res_age_flb)
+
+cors_income <- cor(famhist[score_names], famhist$income_cat, use = "pairwise")
 # TODO: use age_left_edu when this includes college
-cors_educ <-   cor(famhist[c(score_names)], famhist$edu_qual, use = "pairwise")
+cors_educ <-   cor(famhist[score_names], famhist$edu_qual, use = "pairwise")
 
@@
 
+effect_size_flb <-  res_age_flb %>% 
+                      filter(term != "age_flb") %>% 
+                      pull(estimate)
+
 dfr <- data.frame(
-      type = rep(c("Cor. with income", "Cor. with educ"), each = 33),
-      cor = c(cors_income[,1], cors_educ[,1]),
-      effect_size = rep(effect_size, 2)
-)
-
-ggplot(dfr, aes(effect_size, cor, colour = type)) + 
+        Correlation  = rep(c("Income", "Education"), each = 33, 2),
+        cor          = rep(c(cors_income[,1], cors_educ[,1]), 2),
+        effect_size  = c(rep(effect_size, 2), rep(effect_size_flb, 2)),
+        Controls     = rep(c("None", 
+                        "Age at first live birth"), each = 66)
+      )
+
+dfr %>% 
+      mutate(
+        Controls = fct_relevel(Controls, "None")
+      ) %>% 
+      ggplot(aes(effect_size, cor, colour = Correlation)) + 
         geom_point() +
-      facet_wrap(vars(type)) + 
+        facet_grid(Correlation ~ Controls, labeller = label_both) + 
         my_vline + 
         my_hline +
-      labs(x = "Effect on n children", y = "Correlation") +
-      scale_colour_manual(values = c("darkblue", "darkgreen")) +
+        labs(x = "Effect on number of children", y = "Correlation") +
+        scale_colour_brewer(type = "qual", palette = 2) +
         theme(legend.position = "none")
@@
 # Bits to integrate
-
-```{r calc-corr-age-flb}
-
-drake::loadd(res_age_flb)
-drake::loadd(res_sex)
-
-res_sex_comparison <- res_sex %>% 
-      filter(sex == "Female") %>% 
-      mutate(score_name = term) %>% 
-      select(score_name, term, estimate:conf.high)
-
-res_age_flb %<>% filter(term != "age_flb")
-
-res_combined <- bind_rows(
-        raw      = res_sex_comparison, 
-        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
-        .id      = "reg.type"
-      ) %>% 
-      arrange(score_name)
-
-raw_flb_comparison <- res_combined %>% 
-      pivot_wider(
-        id_cols     = score_name, 
-        names_from  = reg.type, 
-        values_from = estimate
-      )
-
-cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
-opp_signed <- sum(
-        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
-      )
-
-```
-
-Figure \@ref(fig:plot-age-flb) shows the results of *children* regressions for 
-women only,
-controlling for age at first live birth. Effect sizes are greatly reduced. Partly,
-this is because the regressions exclude childless women. However,
-in `r opp_signed` out of `r length(score_names)` cases, effect sizes actually
-change sign with the controls. The correlation between effect sizes
-controlling for age at first live birth, and raw effect sizes, is `r cor_raw_flb`.
-
-```{r plot-age-flb, fig.cap = "Effect sizes on number of children, controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}
-
-n_regs <- as.double(nrow(res_age_flb))
-
-res_combined %>% 
-      mutate(
-        score_name = pretty_names(score_name),
-        score_name = fct_reorder(score_name, estimate, order_abs(2)),
-        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for age FLB")
-      ) %>% 
-      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
-        color = p.value < 0.05/{{n_regs}})) + 
-        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
-        pval_geom_point() +
-        pval_scale() +
-        scale_fill_brewer(type = "qual", palette = 3) +
-        my_labs() +
-        my_theme()
-```
 

### 2020-07-15 why-negative-selection.Rmd
Pushing update before reinstalling hard drive

@@
 date: "April 2020"
+bibliography: "negative-selection.bib"
 output: 
@@
 
-pval_geom_point <- function (size = 1.5, stroke = 0.7, ...) {
+pval_geom_point <- function (size = 1.5, stroke = 0.5, ...) {
   geom_point(shape = 21, size = size, stroke = stroke, ...)
@@
   levels(var) <- sprintf("%s (N = %s)", levels(var), 
-                         prettyNum(n[levels(var)], big.mark = ","))
+                         prettyNum(n[seq_len(nlevels(var))], big.mark = ","))
   # var_n <- sprintf("%s (N = %s)", var, prettyNum(n[var], big.mark = ","))
@@
 questions on number of children and number of siblings to reconstruct two
-generations of family size per respondent. We then relate these to polygenic
-scores for several different characteristics.
-
-Natural selection is taking place on most polygenic scores, and is broadly
-"negative" in that characteristics that most people would consider undesirable
-are being selected for. Effect sizes are larger for number of siblings than for
-number of children. This is not because the strength of natural selection has
-changed over time. Instead, the difference is probably due to ascertainment bias
-in UK Biobank. Indeed, natural selection is concentrated among population
-subgroups that are under-represented in the sample, including less-educated
-people and poorer households. When we weight the sample to correct for
-ascertainment bias, many effect sizes are considerably larger.
+generations of family size per respondent. We then relate these to 33 different
+polygenic scores for a variety of  characteristics.
+
+Natural selection is taking place on most polygenic scores, and is concentrated
+among population subgroups that are under-represented in the sample, including
+less-educated people and poorer households. When we weight the sample to correct
+for ascertainment bias, many effect sizes are considerably larger.
 
@@
 effect on number of children conditional on age at first live birth. This
-pattern may be explained by economic theories of fertility. In these, higher
+pattern can be explained by economic theories of fertility. In these, higher
 potential earnings have two opposite effects on fertility: a
@@
 
-```
+```{r, eval = FALSE}
 # Open questions/TODO
@@
 * control for age in n_partners regressions.
-* can we weight men by age at first child, using GHS? If not, may need
-  to plot the age_flb weighted regressions separately, against unweighted
-  women-only regressions.
-* Can we find representative data for weights by n_partners?
+* Beauchamp 2016 used the HRS. He reckons it's representative. Is it? How do
+  his effect sizes compare?
+  - he also worries about the quality/quantity tradeoff, which might mean that
+    people with fewer children have more grandchildren (than you'd expect). 
+    As we have the two generations, we can test this. (Actually, there's a 
+    positive correlation of about 0.1)
+
 ```
@@
 drake::loadd(famhist)
-drake::loadd(ghs_weights)
-famhist <- left_join(famhist, ghs_weights, by = "f.eid")
+
 pgs_over_time <- famhist %>% 
@@
 may also shift respondents between the subgroups. For example, a high 
-PGS for educational attainment will increase the age at which a respondent 
+PGS for educational attainment may increase the age at which a respondent 
 leaves school.
@@
 separately for males and females. Differences are particularly large for
-educational attainment, height and MDD. There is no overall pattern in
-differences of effect size, however.
+educational attainment, height and MDD. Several PGS for mental illness and
+personality traits are more selected for (or less against) among women, including 
+major depressive disorder (MDD), schizophrenia and neuroticism, while 
+extraversion is more selected for among men. Waist circumference and waist-hip 
+ratio are less selected for among women, and educational attainment is more
+selected against, though the difference is only significant for EA3.
 
@@
 
-drake::loadd(res_wt_ghs_weights)
 drake::loadd(res_wt_flb_weights)
@@
 
-# TODO: move later
-res_weighted_all <- bind_rows(
-        "N children" = res_wt_ghs_weights, 
-        "N siblings" = res_sibs_parent_weights, 
-      .id = "dep.var") 
-res_weighted_all$reg.type <- "weighted"
-res_weighted_all <- bind_rows(
-        res_weighted_all,
-        res_all %>% filter(reg.type == "raw") %>% select(-score_name)
-      )
-
-res_wide <- full_join(
-        res_wt_ghs_weights,
-        res_sibs_parent_weights,
-        by = "term",
-        suffix = c(".children", ".sibs")
-      ) %>% 
-      filter(p.value.children < 0.05/33 | p.value.sibs < 0.05/33)
-
-chn_larger <- with(res_wide, estimate.children > estimate.sibs)
-
 ```
@@
 Effect sizes are reasonably consistent across the geographical and
-age/qualification weights. But they are increased again when we weight by age at
-first live birth.This suggests that estimates might be increased still more by
-weighting on other relevant demographic variables. Overall, these results
-suggest that previous work may have underestimated the level of selection
-occurring in the population. [TODO: who?]
+age/qualification weights, but they are increased again when we weight by age at
+first live birth. This suggests that estimates might be further affected by
+weighting on other demographic variables. Given that the Biobank sample
+seems to be ascertained in ways that reduce estimates of natural selection, 
+we view our estimates as likely to be conservative.
 
@@
 
-To compensate for this, we reweight our preferred weightings by the inverse of
-*number of siblings*.
-
-TODO: 
-- weight regressions using "best effort" weights
-- look at change over time to confirm your "no change" hypothesis
-- look at siblings, again using "best" weights
-
-We rerun the basic regressions for both children and siblings, using the new
-weights. Figure \@ref(fig:plot-res-weighted) shows the results. The weightings
-are fairly successful at reconciling the overall size of effects from siblings
-and children regressions. Among weighted scores that are significant (at p <
-0.05/33) in either regression, `r sum(chn_larger)` estimates are
-higher for children and `r length(chn_larger) - sum(chn_larger)` are
-lower. However, some estimates differ widely or change
-sign across the generations.
-
-```{r plot-res-weighted, fig.cap = "Effects of polygenic scores on number of children/siblings. Regressions weighted by sex, age and education."}
-
-res_weighted_all %>% 
-      rename(
-        "Dependent variable" = dep.var, 
-      ) %>% 
-      filter(reg.type == "weighted") %>% 
-      standard_ggplot(fill = `Dependent variable`, n_regs = 33, order_idx = 1, 
-        shape = reg.type) + 
-        scale_shape_manual(values = c(21, 4))
-
+To compensate for this, we reweight our preferred weightings (Age/Qualification)
+by the inverse of *number of siblings*. We then regress this on PGS.
+Figure \@ref(fig:plot-res-sibs) shows the results. For comparison, we show
+regressions on *number of children*, where we have excluded respondents with no 
+children.
+
+Effect sizes appear larger in the parents' generation, i.e.
+for *number of siblings*. However, we treat this result cautiously, because
+when we split respondents up by year of birth, we find few differences in
+effect sizes between early- and late-born respondents, whether our 
+dependent variable is *number of siblings* or *number of children*. In other words,
+since estimated effect sizes change when we change the dependent variable,
+but do not change over time within either dependent variable, the explanation
+may be due to remaining ascertainment bias within our sample.
+
+```{r plot-res-sibs, fig.cap = "Effects of polygenic scores on fertility, respondents' parents vs. respondents"}
+drake::loadd(res_sibs_parent_weights)
+drake::loadd(res_children_comparison)
+
+res_sibs_kids <-  bind_rows(
+                    Siblings = res_sibs_parent_weights, 
+                    Children = res_children_comparison,
+                    .id = "Dep. var."
+                  )
+
+standard_ggplot(res_sibs_kids, `Dep. var.`, n_regs = length(score_names))
 ```
@@
 with more and fewer lifetime sexual partners? A possible answer is given by the
-economic theory of fertility [TODO cite Becker 1960]. According to this,
+economic theory of fertility [@becker1960economic]. According to this,
 increases in potential earned income affect fertility via two opposing channels.
@@
 
-# Bits to integrate
-
-```{r calc-age-birth-parents}
-
-drake::loadd(res_age_birth_parents)
-drake::loadd(res_all)
-
-res_sibs_raw <- filter(res_all, dep.var == "N siblings", reg.type == "raw")
-compare_abp_raw <- res_age_birth_parents %>% 
-      filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
-      left_join(res_sibs_raw, by = "term")
-
-cor_raw_fath <- with(
-  compare_abp_raw %>% filter(control == "fath_age_birth"), 
-  cor(estimate.x, estimate.y))
-cor_raw_moth <- with(
-  compare_abp_raw %>% filter(control == "moth_age_birth"), 
-  cor(estimate.x, estimate.y))
-```
-
-We can run similar regressions for the parents' generation, using the subsets of respondents who reported their mother's or father's age
-and who had no elder siblings. We run *sibling* regressions on these subsets,
-controlling for either parent's age at their birth.
-Figure \@ref(fig:plot-age-birth-parents) shows
-the results. Effect sizes are very similar, whether controlling for
-father's or mother's age at respondent's birth or mother's age at respondent's birth. Unlike for the respondents' generation, effect sizes are positively correlated with
-the effect sizes from bivariate regressions (father's age at birth: 
-$\rho$ `r cor_raw_fath`; mother's age at birth: $\rho$ `r cor_raw_moth`).
-
-```{r plot-age-birth-parents, fig.cap = "Effect sizes on number of siblings controlling for parents' age at birth, eldest siblings"}
-
-n_regs <- as.double(nrow(res_age_birth_parents)/2)
-
-res_age_birth_parents %>% 
-    filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
-    mutate(
-      term    = pretty_names(term),
-      term    = fct_reorder(term, estimate, order_abs(2)),
-      Control = ifelse(control == "fath_age_birth", 
-                  "Father's age at birth", 
-                  "Mother's age at birth"
-                )
-    ) %>% 
-    ggplot(aes(y = term, x = estimate, fill = Control, 
-      colour = p.value < 0.05/{{n_regs}})) + 
-      pval_geom_point(alpha = 0.8) +  
-      my_vline +
-      my_fill_scale(2) +
-      pval_scale() +
-      my_labs()
-
-```
-
-```{r calc-age-flb-dv}
-
-drake::loadd(res_age_flb_dv)
-drake::loadd(res_age_birth_parents_dv)
-
-compare_flb_dv <- left_join(
-        res_age_flb_dv           %>% select(term, estimate),
-        res_age_birth_parents_dv %>% select(term, estimate, dep.var),
-        by = "term"
-      )
-
-cor_flb_fath <- with(compare_flb_dv %>% filter(dep.var == "fath_age_birth"),
-        cor(estimate.x, estimate.y)
-      )
-cor_flb_moth <- with(compare_flb_dv %>% filter(dep.var == "moth_age_birth"),
-        cor(estimate.x, estimate.y)
-      )
-```
-
-These results suggest that polygenic scores may directly correlate with
-age at first live birth. Figure \@ref(fig:plot-age-flb-dv) plots
-estimated effect sizes from bivariate regressions for respondents, and
-Figure \@ref(fig:plot-age-birth-parents-dv) does the same for their parents. 
-Effect sizes are reasonably large. They are also very highly correlated across 
-generations. Effect sizes of PGS on father's age at own birth, and on own age
-at first live birth, have a correlation of `r cor_flb_fath`; for mother's age
-and own age it is `r cor_flb_moth`
-
-```{r plot-age-flb-dv, fig.cap = "Effects of polygenic scores on age at first live birth"}
-
-n_regs <- as.double(nrow(res_age_flb_dv))
-
-res_age_flb_dv %>% 
-      mutate(
-        term = pretty_names(term),
-        term = fct_reorder(term, estimate, order_abs(1))
-      ) %>% 
-      ggplot(aes(estimate, term, colour = p.value < 0.05 / {{n_regs}})) + 
-      my_vline +
-      pval_geom_point(fill = "steelblue4") +
-      pval_scale() + 
-      my_labs() +
-      my_theme()
-
-```
-
-```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at own birth, eldest siblings"}
-
-n_regs <- as.double(nrow(res_age_birth_parents_dv))
-
-res_age_birth_parents_dv %>% 
-      mutate(
-        term = pretty_names(term),
-        term = fct_reorder(term, estimate, order_abs(1)),
-        "Dependent variable" = dplyr::recode(dep.var, 
-          "fath_age_birth" = "Father's age", 
-          "moth_age_birth" = "Mother's age"
-        )
-      ) %>% 
-      ggplot(aes(estimate, term, fill = `Dependent variable`, 
-        colour = p.value < 0.05 / {{n_regs}})) + 
-        my_vline +
-        pval_geom_point() +
-        pval_scale() +
-        my_fill_scale(n = 2) +
-        my_labs()
-
-```
+# Conclusion
+
+Natural selection is taking place in modern populations on variants underlying
+many polygenic scores. In a loose sense, phenotypes relating to scores which are
+selected for (against) would be seen as undesirable (desirable) by many people.
+For example, most people would probably prefer to have high educational
+attainment, a low risk of major depressive disorder, and a low risk of coronary
+artery disease, but PGS underlying these traits are being selected for.
+Potentially, natural selection could increase the health burden on modern
+societies, but that depends on effect sizes. Existing work argues that these are
+small. However, as we show, estimates can be crucially affected by ascertainment
+bias in population samples. In addition, we do not know whether estimated effect
+sizes of natural selection will increase or decrease as more accurate polygenic
+scores are produced. We are also unsure about whether polygenic scores for new
+phenotypes will show a similar pattern of natural selection as existing ones. In
+short, it is probably too early to tell whether modern natural selection is
+substantively important or not.
+
+Previous work has documented natural selection in modern populations
+[@beauchamp2016genetic; @kong2017selection; @sanjak2018evidence] but has not
+explained its underlying causes. The economic theory of fertility provides a
+parsimonious explanation for which traits are selected, and for why the
+direction of selection is reversed among certain subgroups. An alternative
+theory is that traits selected for are linked to impulsivity and low time
+discount rates, via the channel of early sexual behaviour. The data here provide
+some support for this: scores which might plausibly be linked to
+impulsivity,like ADHD and age at smoking initiation, are selected for. However,
+the "impulsivity" theory is less good at explaining variation in selection
+across the full range of scores, including e.g. waist-hip ratio and BMI.
+Impulsivity also does not explain the reversed sign of selection among e.g.
+older parents. Overall, then, the economic theory is probably the leading
+explanation at present.
 
@@
 
-drake::loadd(res_period)
+drake::loadd(res_period_parents)
+drake::loadd(res_period_children)
+
+res_period <- bind_rows(res_period_parents, res_period_children)
+
 res_period %<>% 
@@
 To check whether effect sizes were changing over time, we ran regressions 
-interacting PGS with birth year, median split at 1950. 
+interacting PGS with birth year, median split at 1950 ("early born" versus "late born"). We use both 
+*number of children* and *number of siblings* as a dependent variable. We weight 
+using age/qualification cells, and further adjust for selection in the parents'
+generation (see above).
+
 Tables \@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)
-summarize the results. Very few scores are significantly different. Only
-one score, EA3, changes significantly across time in both generations: the
-absolute size of the (negative) effect significantly increases in sibling 
-regressions, and significantly decreases in children regressions.
+summarize the results. There is no change in natural selection in the parents'
+generation. In the respondents' generation, effect sizes were larger in
+absolute size among the late born for five PGS: autism, cognitive ability,
+EA2, EA3 and extraversion. Three PGS showed significant changes in sign: 
+alcohol use (positive to negative); conscientiousness (negative to positive);
+and type II diabetes (negative to positive).
 
@@
       count() %>% 
-      period_table("Change in effect sizes between early and late born parents, 'sibling' regressions")
+      period_table("Change in effect sizes between parents of early and late born respondents (regressions on number of siblings)")
 
@@
       count() %>% 
-      period_table("Change in effect sizes between early and late born respondents, 'children' regressions")
+      period_table("Change in effect sizes between early and late born respondents (regressions on number of children)")
+```
+
+## Selection controlling for age at first live birth: respondents' parents
+
+```{r calc-age-birth-parents}
+
+drake::loadd(res_age_birth_parents)
+drake::loadd(res_all)
+
+res_sibs_raw <- filter(res_all, dep.var == "N siblings", reg.type == "raw")
+compare_abp_raw <- res_age_birth_parents %>% 
+      filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
+      left_join(res_sibs_raw, by = "term")
+
+cor_raw_fath <- with(
+  compare_abp_raw %>% filter(control == "fath_age_birth"), 
+  cor(estimate.x, estimate.y))
+cor_raw_moth <- with(
+  compare_abp_raw %>% filter(control == "moth_age_birth"), 
+  cor(estimate.x, estimate.y))
+```
+
+Among the parents' generation, we can control for age at first live birth using
+the subsets of respondents who reported their mother's or father's age, and who
+had no elder siblings. We run regressions on *number of siblings* on these
+subsets, controlling for either parent's age at their birth. Figure
+\@ref(fig:plot-age-birth-parents) shows the results. Effect sizes are very
+similar, whether controlling for father's or mother's age. Unlike for the
+respondents' generation, effect sizes are positively correlated with the effect
+sizes from bivariate regressions without the age at birth control (father's age
+at birth: $\rho$ `r cor_raw_fath`; mother's age at birth: $\rho$ `r
+cor_raw_moth`).
+
+```{r plot-age-birth-parents, fig.cap = "Effect sizes on number of siblings controlling for parents' age at birth, eldest siblings"}
+
+n_regs <- as.double(nrow(res_age_birth_parents)/2)
+
+res_age_birth_parents %>% 
+    filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
+    mutate(
+      term    = pretty_names(term),
+      term    = fct_reorder(term, estimate, order_abs(2)),
+      Control = ifelse(control == "fath_age_birth", 
+                  "Father's age at birth", 
+                  "Mother's age at birth"
+                )
+    ) %>% 
+    ggplot(aes(y = term, x = estimate, fill = Control, 
+      colour = p.value < 0.05/{{n_regs}})) + 
+      pval_geom_point(alpha = 0.8) +  
+      my_vline +
+      my_fill_scale(2) +
+      pval_scale() +
+      my_labs()
+
+```
+
+## Effects of PGS on age at first live birth
+
+```{r calc-age-flb-dv}
+
+drake::loadd(res_age_flb_dv)
+drake::loadd(res_age_birth_parents_dv)
+
+compare_flb_dv <- left_join(
+        res_age_flb_dv           %>% select(term, estimate),
+        res_age_birth_parents_dv %>% select(term, estimate, dep.var),
+        by = "term"
+      )
+
+cor_flb_fath <- with(compare_flb_dv %>% filter(dep.var == "fath_age_birth"),
+        cor(estimate.x, estimate.y)
+      )
+cor_flb_moth <- with(compare_flb_dv %>% filter(dep.var == "moth_age_birth"),
+        cor(estimate.x, estimate.y)
+      )
+```
+
+Our results suggest that polygenic scores may directly correlate with
+age at first live birth. Figure \@ref(fig:plot-age-flb-dv) plots
+estimated effect sizes from bivariate regressions for respondents, and
+Figure \@ref(fig:plot-age-birth-parents-dv) does the same for their parents. 
+Effect sizes are reasonably large. They are also very highly correlated across 
+generations. Effect sizes of PGS on father's age at own birth, and on own age
+at first live birth, have a correlation of `r cor_flb_fath`; for mother's age
+and own age it is `r cor_flb_moth`
+
+```{r plot-age-flb-dv, fig.cap = "Effects of polygenic scores on age at first live birth"}
+
+n_regs <- as.double(nrow(res_age_flb_dv))
+
+res_age_flb_dv %>% 
+      mutate(
+        term = pretty_names(term),
+        term = fct_reorder(term, estimate, order_abs(1))
+      ) %>% 
+      ggplot(aes(estimate, term, colour = p.value < 0.05 / {{n_regs}})) + 
+      my_vline +
+      pval_geom_point(fill = "steelblue4") +
+      pval_scale() + 
+      my_labs() +
+      my_theme()
+
+```
+
+```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at own birth, eldest siblings"}
+
+n_regs <- as.double(nrow(res_age_birth_parents_dv))
+
+res_age_birth_parents_dv %>% 
+      mutate(
+        term = pretty_names(term),
+        term = fct_reorder(term, estimate, order_abs(1)),
+        "Dependent variable" = dplyr::recode(dep.var, 
+          "fath_age_birth" = "Father's age", 
+          "moth_age_birth" = "Mother's age"
+        )
+      ) %>% 
+      ggplot(aes(estimate, term, fill = `Dependent variable`, 
+        colour = p.value < 0.05 / {{n_regs}})) + 
+        my_vline +
+        pval_geom_point() +
+        pval_scale() +
+        my_fill_scale(n = 2) +
+        my_labs()
+
 ```
@@
 ```
+
+# References

### 2020-07-15 _drake.R

@@ -38,6 +38,7 @@ msoa_shapefile     <- file.path(data_dir, "infuse_msoa_lyr_2011_clipped",
 
 weighting_schemes <- rlang::syms(c("flb_weights", "age_qual_weights", 
                         "msoa_weights"))
+period_weight_syms <- rlang::syms((c("age_qual_weights", "parent_weights")))
 
 plan <- drake_plan(
   score_names  = {
@@ -98,14 +99,6 @@ plan <- drake_plan(
                   format = "fst"
                 ),
   
-  # ghs_weights  =  {
-  #   weight_by_ghs(
-  #     ~ factor(sex) + age_at_recruitment + factor(edu_qual), 
-  #     ghs_subset, 
-  #     famhist
-  #   )
-  # },
-  
   flb_weights = {
     weight_by_ghs(
             ~ age_at_recruitment + factor(flb_cat) + factor(edu_qual), 
@@ -120,10 +113,10 @@ plan <- drake_plan(
   
   msoa_weights = weight_by_census_msoa(famhist, census_msoa, famhist_msoa),
   
-  mf_pairs = target(
-                     make_mf_pairs(file_in(!! mf_pairs_file), famhist), 
-                     format = "fst"
-                   ), 
+  mf_pairs =  target(
+                make_mf_pairs(file_in(!! mf_pairs_file), famhist), 
+                format = "fst"
+              ), 
   
   rgs = make_rgs(file_in(!! rgs_file)),
   
@@ -184,14 +177,27 @@ plan <- drake_plan(
                               dep.var     = "n_sibs"
                             ),
   
-  res_period = {
-    expand_grid(
-            children = c(TRUE, FALSE),
-            score_name = score_names
-          ) %>% 
-          pmap_dfr(run_regs_period, famhist = famhist)
-  },
+  res_children_comparison = map_dfr(score_names,
+                              run_regs_weighted,
+                              famhist     = famhist %>% filter(n_children > 0),
+                              weight_data = age_qual_weights,
+                              dep.var     = "n_children"
+                            ),
   
+  res_period =  target(
+                  map_dfr(score_names, 
+                    run_regs_period,
+                    famhist     = famhist,
+                    weight_data = weight_data,
+                    children    = children
+                  ),
+                  transform = map(
+                    children    = !! (c(TRUE, FALSE)),
+                    weight_data = !! period_weight_syms,
+                    .names      = c("res_period_children", "res_period_parents")
+                  )
[…]

### 2020-10-07 why-negative-selection.Rmd
Changes to get drake working again after reinstating files

@@
 author: "David Hugh-Jones, Abdel Abdellaoui"
-date: "April 2020"
+date: "July 2020"
 bibliography: "negative-selection.bib"

### 2020-10-07 make-data.R

@@ -60,6 +60,7 @@ make_famhist <- function (
         famhist4_file,
         famhist5_file,
         famhist6_file,
+        famhist7_file,
         pgs_dir
       ) {
   ph <- read_csv(ph_file, col_types = cols(
@@ -68,31 +69,24 @@ make_famhist <- function (
     .default = col_double()
   ))
   
-  famhist <- read_csv(famhist_file, col_types = strrep("d", 40))
-  famhist2 <- read_csv(famhist2_file, col_types = strrep("d", 17))
-  famhist3 <- read_csv(famhist3_file)
-  famhist4 <- read_csv(famhist4_file, col_types = strrep("d", 33))
-  famhist5 <- read_csv(famhist5_file, col_types = strrep("d", 4))
-  famhist6 <- read_csv(famhist6_file, col_types = strrep("d", 22))
-  names(famhist) <- paste0("f.", names(famhist))
-  names(famhist) <- gsub("\\-", ".", names(famhist))
-  names(famhist2) <- paste0("f.", names(famhist2))
-  names(famhist2) <- gsub("\\-", ".", names(famhist2))
-  names(famhist3) <- paste0("f.", names(famhist3))
-  names(famhist3) <- gsub("\\-", ".", names(famhist3))
-  names(famhist4) <- paste0("f.", names(famhist4))
-  names(famhist4) <- gsub("\\-", ".", names(famhist4))
-  names(famhist5) <- paste0("f.", names(famhist5))
-  names(famhist5) <- gsub("\\-", ".", names(famhist5))
-  names(famhist6) <- paste0("f.", names(famhist6))
-  names(famhist6) <- gsub("\\-", ".", names(famhist6))
+  famhist <- list()
+  famhist[[1]] <- read_csv(famhist_file, col_types = strrep("d", 40))
+  famhist[[2]] <- read_csv(famhist2_file, col_types = strrep("d", 17))
+  famhist[[3]] <- read_csv(famhist3_file)
+  famhist[[4]] <- read_csv(famhist4_file, col_types = strrep("d", 33))
+  famhist[[5]] <- read_csv(famhist5_file, col_types = strrep("d", 4))
+  famhist[[6]] <- read_csv(famhist6_file, col_types = strrep("d", 22))
+  famhist[[7]] <- read_csv(famhist7_file, col_types = strrep("d", 3)) 
[…]

### 2020-10-13 why-negative-selection.Rmd
New intro; age-controlled n_partners regression

@@
 ---
-title: "Negative Selection in the 20th Century"
+title: "Natural Selection in the 20th Century"
 author: "David Hugh-Jones, Abdel Abdellaoui"
@@
 
-Existing work suggests that natural selection is taking place in modern populations,
-but that effect sizes are small. In this paper we investigate natural selection
-in the UK over the twentieth century, using data from UK Biobank. We use
-questions on number of children and number of siblings to reconstruct two
-generations of family size per respondent. We then relate these to 33 different
-polygenic scores for a variety of  characteristics.
-
-Natural selection is taking place on most polygenic scores, and is concentrated
-among population subgroups that are under-represented in the sample, including
-less-educated people and poorer households. When we weight the sample to correct
-for ascertainment bias, many effect sizes are considerably larger.
-
-Among mothers who were older at first birth, and among parents with fewer sexual
-partners, effect signs are reversed. The effects of PGS can be decomposed into
-two channels: an effect on age at first live birth, and an opposite-signed
-effect on number of children conditional on age at first live birth. This
-pattern can be explained by economic theories of fertility. In these, higher
-potential earnings have two opposite effects on fertility: a
-fertility-increasing income effect (higher income makes children more
-affordable), and a fertility-lowering substitution effect (time spent on
-childrearing has a higher cost in foregone earnings). Among couples who can take
-advantage of coordination in childcare activities, the income effect dominates;
-among single parents, or those in unstable relationships, the substitution effect
-dominates.
+Geneticists have long hypothesized that natural selection is taking place in
+modern human populations. Recent work confirms this using genome-wide analysis
+[@Barban_2016; @beauchamp2016genetic; @kong2017selection; @sanjak2018evidence].
+In particular, genetic variants associated with higher educational attainment
+are being selected against, although effect sizes appear small.
+
+Several gaps in our knowledge remain. First, while existing work examines
+contemporary populations, we would like to extend the analysis over time, since
+natural selection in the same direction over multiple generations
+could lead to significant effects in the long run. Second, there are concerns
+about ascertainment bias in the samples used. This could potentially bias
+estimates of the magnitude of natural selection. Lastly, we know little about
+the *mechanisms* behind natural selection in contemporary populations.
+
+Here we attempt to answer these questions, using data from UK Biobank. We use
+questions on respondents' number of children, and their number of siblings, to 
+reconstruct two generations of family size. We then relate these to 33 different
+polygenic scores for a variety of characteristics. This analysis provides a 
+broader picture of natural selection in the UK than has previously been available.
+
+Our key results are as follows. First, comparing natural selection across the
+two generations, there are few signs of change in size or direction over time.
+Most polygenic scores are selected in the same direction in both generations. We
+also see little evidence of changes across birth cohorts within either
+generation.
+
+Second, natural selection is concentrated among population subgroups which are
+under-represented in the sample, including less-educated people and poorer
+households. When we weight the sample to correct for ascertainment bias, many
+effect sizes are larger, sometimes much larger. We emphasize that our weights
+are only a partial solution to ascertainment bias in the sample. As a result,
+much uncertainty remains about the strength of natural selection in the UK
+population, but we can be reasonably sure that it is larger than naïve estimates
+suggest.
+
+Lastly, patterns in natural selection give us clues to the underlying
+mechanism. In particular, among mothers who were older at their first birth, and
+among parents with fewer sexual partners, the direction of natural selection is
+*reversed*: for example, among these subgroups, higher polygenic scores (PGS)
+for educational attainment are associated with having more offspring, rather
+than fewer. The effects of PGS can be decomposed into two channels: an effect on
+age at first live birth, and an opposite-signed effect on number of children
+conditional on age at first live birth.
+
+This pattern can be explained by economic theories of fertility
+[@becker1960economic]. In these, higher potential earnings have two opposite
+effects on fertility: a fertility-increasing *income effect* (higher income
+makes children more affordable), and a fertility-lowering *substitution effect*
+(time spent on childrearing has a higher cost in foregone earnings). Among
+couples who can take advantage of coordination in childcare activities, the
+income effect dominates; among single parents, or those in unstable
+relationships, the substitution effect dominates. Supporting this explanation,
+polygenic scores' raw effects on fertility are negatively correlated with their
+effects on income; on the other hand, controlling for age at first live birth,
+scores' effects on fertility and on household income are *positively*
+correlated. These two results are compatible with the substitution effect and the
+income effect respectively.
 
@@
 Respondents are higher income and better educated than the UK population, and
-they may also differ on other unobserved characteristics.
-
-To examine the effects of sampling, we regress respondents' polygenic
-scores on their number of children:
+they may also differ on other unobserved characteristics [@10.1093/aje/kwx246].
+
+To examine the effects of sampling, we start by measuring the strength of
+natural selection within different subgroups. To do this we regress respondents' 
+polygenic scores on their number of children:
 
@@
 sample (or a given subgroup). In fact, since polygenic scores are normalized to
-mean 0, $\beta$ is the expected polygenic score among children of the sample.
+mean 0, $\beta$ is the expected polygenic score among children of the sample or
+subgroup.
 
 Figures \@ref(fig:plot-educ-level) and \@ref(fig:plot-income) plot effect sizes
-for each polygenic score among different subgroups of the sample. Figure
-\@ref(fig:plot-educ-level) groups respondents by age of completing full-time
-education. Figure \@ref(fig:plot-income) groups respondents by household income.
-Effect sizes are larger and more significant for the lowest income category, and
-for the lowest education category. Note that the overall effect is not a simple
-average of the effect among the different subgroups, because polygenic scores 
-may also shift respondents between the subgroups. For example, a high 
-PGS for educational attainment may increase the age at which a respondent 
-leaves school.
+for each polygenic score. Figure \@ref(fig:plot-educ-level) groups respondents
+by age of completing full-time education. Figure \@ref(fig:plot-income) groups
+respondents by household income. Effect sizes are larger and more significant
+for the lowest income category, and for the lowest education category. Note that
+the overall effect is not a simple average of the effect among the different
+subgroups, because polygenic scores may also shift respondents between the
+subgroups. For example, a high PGS for educational attainment may increase the
+age at which a respondent leaves school.
 

### 2020-10-13 _drake.R

@@ -419,6 +419,27 @@ plan <- drake_plan(
     res %>% filter(grepl(":.*age_fte_cat", term))
   },
   
+  res_partners_controlled = {
+    sexes <- rlang::exprs(sex == 0, sex == 1)
+    pars <- expand_grid(subset = sexes, score_name = score_names)
+    res <- pmap_dfr(pars,
+            ~ run_regs_fml(
+              fml        =
+                "n_children ~ {score_name}*(lo_partners + YOB + I(YOB^2))",
+              score_name = .y,
+              famhist    = famhist,
+              subset     = .x
+            ),
+              .id = "row_number"
+          )
+    pars$row_number <- as.character(seq_len(nrow(pars)))
+    
+    left_join(res, pars, by = "row_number") %>%
+          mutate(sex = ifelse(subset == "sex == 1", "Male", "Female")) %>%
+          select(-row_number, -subset) %>% 
+          filter(! grepl("YOB", term))
+  },
+  
   res_together = {
     most_score_names <- setdiff(score_names, 
           c("EA2_noUKB", "whr_combined", "wc_combined", "hip_combined"))

### 2020-10-15 why-negative-selection.Rmd
Colour changes over time in raw PGS. More work on text.

@@
 * Look at geography, esp. in siblings regressions.
-* control for age in n_partners regressions.
 * Beauchamp 2016 used the HRS. He reckons it's representative. Is it? How do
@@
 
-```{r calc-means-over-time}
-
-drake::loadd(famhist)
-
-pgs_over_time <- famhist %>% 
-      group_by(
-        YOB = santoku::chop(YOB, 
-          seq(1940, 1970, 5), 
-          labels    = seq(40, 65, 5),
-          extend    = FALSE
-        ) 
-      ) %>% 
-      filter(! is.na(YOB)) %>% 
-      summarize(across(all_of(score_names), ggplot2::mean_cl_normal, na.rm = TRUE)) %>% 
-      pivot_longer(-YOB, names_to = "score_name", values_to = "score") %>% 
-      mutate(
-        score_lo = score$ymin,
-        score_hi = score$ymax,
-        score    = score$y
-      )
-
-```
-
-Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores by 5-year
-birth intervals, for the entire sample. Several scores show consistent increases
-or declines over this 30-year period, of the order of 5% of a standard deviation.
-
-```{r plot-means-over-time, fig.height = 5, fig.cap = "Mean polygenic scores by birth year in UK Biobank. Points are means for 5-year intervals. Lines are 95% confidence intervals."}
+We first show raw changes in polygenic scores over birth years, within the entire UK Biobank
+sample. Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores by 5-year
+birth intervals. Several scores show consistent increases or declines over this 
+30-year period, of the order of 5% of a standard deviation.
+
+```{r plot-means-over-time, fig.height = 5, fig.cap = "Mean polygenic scores by birth year in UK Biobank.
+Points are means for 5-year intervals.
+Lines are 95% confidence intervals.
+Green lines show a significant increase over time (p < 0.05).
+Red lines show a significant decrease."}
+
+drake::loadd(pgs_over_time)
 
@@
       mutate(score_name = pretty_names(score_name)) %>% 
-      ggplot(aes(YOB, group = score_name)) + 
+      ggplot(aes(YOB, group = score_name, colour = Change)) + 
       geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
-        colour = "steelblue4", fatten = 2) +
-      facet_wrap(~score_name) +
+        fatten = 2) +
+      facet_wrap(~score_name, ncol = 5) +
       my_hline +
+      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
       theme(
         strip.text  = element_text(size = 6), 
-        axis.text.x = element_text(size = 7)
+        axis.text.x = element_text(size = 7),
+        legend.position = "none"
       ) +
@@
 
-These estimates may suffer from ascertainment bias within the UK Biobank sample.
+These changes could reflect either natural selection within the UK population,
+or ascertainment bias within the sample. 
 Respondents are higher income and better educated than the UK population, and
 they may also differ on other unobserved characteristics [@10.1093/aje/kwx246].
+Since richer and educated people also live longer, this bias might also increase
+with age.
 
@@
 natural selection within different subgroups. To do this we regress respondents' 
-polygenic scores on their number of children:
+polygenic scores on their number of children ($y_i$):
 
@@
 
+```{r calc-sex-ratio}
+female_ratio <- mean(famhist$sex == 0)
+```
+
 The effect size $\beta$ measures the strength of natural selection within the
-sample (or a given subgroup). In fact, since polygenic scores are normalized to
+sample or a given subgroup. In fact, since polygenic scores are normalized to
 mean 0, $\beta$ is the expected polygenic score among children of the sample or
-subgroup.
+subgroup [@beauchamp2016genetic]. We begin with basic demographic variables
+including education, income and sex. These are all potential sources of ascertainment
+bias: as well as the ascertainment for income and education, mentioned above, the
+sample sex ratio  skews `r female_ratio * 100`% female.
 
@@
 subgroups, because polygenic scores may also shift respondents between the
-subgroups. For example, a high PGS for educational attainment may increase the
+subgroups. For example, a high PGS for educational attainment may correlate
+with fewer children among early school leavers, but may also increase the
 age at which a respondent leaves school.
@@
 
-When we split respondents up by other demographic variables, we get a more
-surprising result: the direction of natural selection is different among
-different subgroups.
-
-```{r calc-age-flb-cross}
-
-drake::loadd(res_age_flb_cross)
-
-cor_age_flb <- res_age_flb_cross %>% 
-      mutate(category = gsub(":.*", "", term)) %>% 
-      pivot_wider(score_name, names_from = category, values_from = estimate) %>%              {cor(.$`age_flb_cat10-22`, .$`age_flb_cat28-52`)}
-
-```
-
-Figure \@ref(fig:plot-age-flb-cross) shows effect sizes estimated separately for
-each tercile of age at first live birth. Several effects are strikingly
-different across terciles. ADHD and MDD are selected for amongst the youngest
-third of mothers, but selected against among the oldest two-thirds. Educational
-attainment is selected for among the oldest two-thirds of mothers, but is not
-significantly selected among the youngest third. Similarly, several PGS of body
-measurements are selected against only among older mothers. The correlation
-between effect sizes for the youngest and oldest terciles is `r cor_age_flb`.
+These results show that ascertainment within UK Biobank could bias estimates
+of the size of natural selection towards zero. We now focus in on variables 
+specifically related to marriage markets and raising children. Whatever processes 
+give rise to natural selection, these variables are likely to be involved; at
+the same time, they may also be ascertained within the sample.
 
 Figure \@ref(fig:plot-n-partners) splits males and females by lifetime number of
-sexual partners, at the median value of 3. Again, selection effects are reversed
-across the groups. Among men and women with more than 3 sexual partners, for
-example, educational attainment is selected against; among those with 3 or fewer
-sexual partners, it is selected for.
-
-Figure \@ref(fig:plot-with-partner) splits respondents up by whether they
-were living with a spouse or partner at the time of interview. Effects are 
-generally the same sign in both groups, but are much larger among those not
-living with a spouse or partner.
-
-```{r plot-age-flb-cross, fig.cap = "Effect sizes on number of children, by age at first live birth terciles"}
-
-drake::loadd(res_age_flb_cross)
-
-# this counts each cross term as a separate test:
-n_regs <- as.double(nrow(res_age_flb_cross))
-
-res_age_flb_cross %>% 
-      mutate(
-        score_name = pretty_names(score_name),
-        score_name = fct_reorder(score_name, estimate, order_abs(1)),
-        term       = gsub("age_flb_cat(.*):.*", "\\1", term),
-        term       = add_n(term, "age_flb_cat")
-      ) %>% 
-      ggplot(aes(estimate, score_name, fill = term, 
-        colour = p.value < 0.05/{{n_regs}})) +
-        my_vline +
-        pval_geom_point() +
-        pval_scale() +
-        my_fill_scale(n = 3, direction = -1, 
-          guide = guide_legend(title = "Age at first live birth")) +
-        my_labs() +
-        my_theme()
-
-```
+sexual partners, at the median value of 3. For both sexes, effects are larger
+and more significant among those with more than 3 partners. Results are
+unchanged if we control for respondents' age (Figure
+\@ref(fig:plot-n-partners-controlled) in the appendix). Figure
+\@ref(fig:plot-with-partner) splits respondents up by whether they were living
+with a spouse or partner at the time of interview. Again, effects are larger
+among those not living with a spouse or partner.
 
@@
 res_partners %<>%       
-      filter(term != "(Intercept)", term != "lo_partnersTRUE") %>% 
-      mutate(`N partners` = ifelse(grepl(":", term), "<= 3", "> 3"))
+      filter(grepl(":", term)) %>% 
+      mutate(`N partners` = ifelse(grepl("TRUE:", term), "<= 3", "> 3"))
 
@@
 
+```{r calc-age-flb-cross}
+
+drake::loadd(res_age_flb_cross)
+
+cor_age_flb <- res_age_flb_cross %>% 
+      mutate(category = gsub(":.*", "", term)) %>% 
+      pivot_wider(score_name, names_from = category, values_from = estimate) %>%              {cor(.$`age_flb_cat10-22`, .$`age_flb_cat28-52`)}
+
+```
+
+Our final variable is age at first live birth (AFLB), for females only. There is evidence
+for genetic effects on AFLB [@Barban_2016]. There is also 
+a close link between this variable and number of children born.
+
+Figure \@ref(fig:plot-age-flb-cross) shows effect sizes estimated separately for
+each tercile of AFLB. Several effects are strikingly
+different across terciles. ADHD and MDD are selected for amongst the youngest
+third of mothers, but selected against among the oldest two-thirds. Educational
+attainment is selected for among the oldest two-thirds of mothers, but is not
+significantly selected among the youngest third. Similarly, several PGS of body
+measurements are selected against only among older mothers. The correlation
+between effect sizes for the youngest and oldest terciles is `r cor_age_flb`.
+As before, recall that PGS may have two effects: they may change the number of
+children within categories, but may also shift respondents between categories.
+
+```{r plot-age-flb-cross, fig.cap = "Effect sizes on number of children, by age at first live birth terciles"}
+
+drake::loadd(res_age_flb_cross)
+
+# this counts each cross term as a separate test:
+n_regs <- as.double(nrow(res_age_flb_cross))
+
+res_age_flb_cross %>% 
+      mutate(
+        score_name = pretty_names(score_name),
+        score_name = fct_reorder(score_name, estimate, order_abs(1)),
+        term       = gsub("age_flb_cat(.*):.*", "\\1", term),
+        term       = add_n(term, "age_flb_cat")
+      ) %>% 
+      ggplot(aes(estimate, score_name, fill = term, 
+        colour = p.value < 0.05/{{n_regs}})) +
+        my_vline +
+        pval_geom_point() +
+        pval_scale() +
+        my_fill_scale(n = 3, direction = -1, 
+          guide = guide_legend(title = "Age at first live birth")) +
+        my_labs() +
+        my_theme()
+
+```
+
 ```{r calc-corr-age-flb}
@@
 
-Figure \@ref(fig:plot-age-flb) shows the results of regressions for women only,
-controlling for age at first live birth. Effect sizes are greatly reduced.
-This could be because the regressions exclude childless women. However, in 
-`r opp_signed` out of `r length(score_names)` cases, effects actually change
-sign when controls are added. The correlation between effect sizes controlling for age
-at first live birth, and raw effect sizes, is `r cor_raw_flb`.
+To investigate this further, we regress number of children on PGS *controlling*
+for AFLB. Figure \@ref(fig:plot-age-flb) shows the results, along with raw
+effect sizes for comparison. Controlling for AFLB greatly reduces effect sizes.
+This could be because the controlled regressions exclude childless women.
+However, in `r opp_signed` out of `r length(score_names)` cases, effects
+actually change sign when controls are added. The correlation between effect
+sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`. So,
+natural selection "effects" seem to come through two channels: an effect on
+AFLB, and an effect on number of children controlling for AFLB. Moreover, these
+channels often push in opposite directions.
 
@@
   `Age/Qualification`      = res_wt_age_qual_weights,
-  `Age/Qual/FLB`  = res_wt_flb_weights,
+  `Age/Qual/AFLB`  = res_wt_flb_weights,
   .id = "Weights"
@@
                   `Age/Qualification` = `Age/Qualification`/None,
-                  `Age/Qual/FLB` = `Age/Qual/FLB`/estimate_females
+                  `Age/Qual/AFLB` = `Age/Qual/AFLB`/estimate_females
                 ) %>% 
@@
 
-To compensate for ascertainment bias in the UK Biobank sample, we weight 
+These results show that there may be ascertainment bias in estimates of natural
+selection derived from UK Biobank. One way to compensate for this is by survey
+weighting. Weighting is not guaranteed to lead to unbiased estimates, because as
+well as ascertainment on characteristics which we measure and weight for, there
+may also be ascertainment on other, unmeasured characteristics. But the exercise
+will give us a rough sense of the direction and magnitude of bias. We weight
 participants using data from the 2011 Census and from the 2006 General Household
@@
 Figure \@ref(fig:plot-res-wt) plots effect sizes among the entire sample, using
-these weighting schemes. We include unweighted regressions for comparison.
-Weighted effect sizes are significantly larger in absolute value. Table
-\@ref(tab:tbl-res-wt) gives effect sizes as a proportion of unweighted effect
-sizes for the census-based weightings.
-
-Effect sizes are reasonably consistent across the geographical and
-age/qualification weights, but they are increased again when we weight by age at
-first live birth. This suggests that estimates might be further affected by
-weighting on other demographic variables. Given that the Biobank sample
-seems to be ascertained in ways that reduce estimates of natural selection, 
-we view our estimates as likely to be conservative.
+these weighting schemes. We include unweighted regressions for comparison. Table
+\@ref(tab:tbl-res-wt) gives effect sizes as a proportion of the unweighted
+effect size. As expected, weighting generally increases estimated effect sizes.
+Sizes are reasonably similar across the geographical and age/qualification
+weights, but they are increased further when we weight by AFLB. This suggests
+that estimates might be further affected by weighting on other demographic
+variables. Given that the Biobank sample seems to be ascertained in ways that
+reduce estimates of natural selection, we suspect that these weighted estimates
+are still conservative.
 
@@
         "are shown.\n",
-        "Age/Qual/FLB as a proportion of unweighted regressions ",
+        "Age/Qual/AFLB as a proportion of unweighted regressions ",
         "for females only."),
@@
 
-Why do selection effects have *opposite* signs across certain subgroups -- for
-example, across older and younger mothers by age at first live birth, or across those
-with more and fewer lifetime sexual partners? A possible answer is given by the
+Why do selection effects have *opposite* signs across older and younger mothers at first live birth, and opposite signs when AFLB is controlled for? A possible answer is given by the
 economic theory of fertility [@becker1960economic]. According to this,
@@
 if lifetime number of sexual partners predicts single parent versus 
-two-parent status, the theory could also explain the differently-signed
-effects across people with more and fewer sexual partners.
+two-parent status, the theory could also explain the different
+effect sizes across people with more and fewer sexual partners.
 
@@
 fertility will correlate with the strength of their relationship with earnings.
-We measure the correlation of each PGS with household income, as well as with
-education level, a phenotype which predicts earnings. We
+To test this, we measure the correlation of each PGS with household income, 
+as well as with education level, a phenotype which predicts earnings. We
 then relate these measurements to effect sizes on number of children.
@@
 attainment, a low risk of major depressive disorder, and a low risk of coronary
-artery disease, but PGS underlying these traits are being selected for.
-Potentially, natural selection could increase the health burden on modern
-societies, but that depends on effect sizes. Existing work argues that these are
-small. However, as we show, estimates can be crucially affected by ascertainment
-bias in population samples. In addition, we do not know whether estimated effect
-sizes of natural selection will increase or decrease as more accurate polygenic
-scores are produced. We are also unsure about whether polygenic scores for new
-phenotypes will show a similar pattern of natural selection as existing ones. In
-short, it is probably too early to tell whether modern natural selection is
-substantively important or not.
+artery disease, but PGS underlying these traits are being selected for. In addition,
+the size of selection on PGS is negatively correlated with the score's correlation 
+with household income. Potentially, natural selection could increase the health 
+burden on modern societies, but that depends on effect sizes. Existing work 
+argues that these are small [@beauchamp2016genetic].
+
+However, as we show, estimates can be crucially affected by ascertainment
+bias in population samples. This problem may be less serious in other surveys,
+such as the US Health and Retirement Study, which explicitly aims to be
+representative (as the UK Biobank does not). Nevertheless, ascertainment bias 
+can still be a problem, since not all respondents consent to the collection of
+genetic data: for instance, completion rates were around 80-85% for HRS genotype
+data [@hrs2020].
+
+In addition, we do not know whether estimated effect sizes of natural selection
+will increase or decrease as more accurate polygenic scores are produced. We are
+also unsure about whether polygenic scores for new phenotypes will show a
+similar pattern of natural selection as existing ones. In short, it is probably
+too early to tell whether modern natural selection is substantively important or
+not.
 
@@
 direction of selection is reversed among certain subgroups. An alternative
-theory is that traits selected for are linked to impulsivity and low time
-discount rates, via the channel of early sexual behaviour. The data here provide
-some support for this: scores which might plausibly be linked to
-impulsivity,like ADHD and age at smoking initiation, are selected for. However,
-the "impulsivity" theory is less good at explaining variation in selection
-across the full range of scores, including e.g. waist-hip ratio and BMI.
-Impulsivity also does not explain the reversed sign of selection among e.g.
-older parents. Overall, then, the economic theory is probably the leading
-explanation at present.
+theory is that traits selected for are linked to externalizing behaviour and low
+time discount rates, via the channel of early sexual behaviour
+[@Mills2020.05.06.081273]. The data here provide some support for this: scores
+which might plausibly be linked to externalizing behaviour, like ADHD and age at
+smoking initiation, are selected for. However, this theory is less
+good at explaining variation in selection across the full range of scores,
+including for example waist-hip ratio and BMI. Externalizing behaviour also does 
+not explain the reversed sign of selection among older parents. Overall, 
+we believe that the economic theory is the leading explanation at present.
 
@@
 
+## Selection by number of partners, controlling for age
+
+Figure \@ref(fig:plot-n-partners-controlled) shows selection effects by
+lifetime number of sexual partners. Age, age squared, and interactions of 
+age and age squared with PGS are controlled for.
+
+```{r plot-n-partners-controlled, fig.cap = "Effect sizes on number of siblings interacted with number of sexual partners (median split at 3), respondent's age and age squared."}
+drake::loadd(res_partners_controlled)
+res_partners_controlled %<>%       
+      filter(term != "(Intercept)", term != "lo_partnersTRUE") %>% 
+      mutate(`N partners` = ifelse(grepl("TRUE:", term), "<= 3", "> 3"))
+
+n_regs <- as.double(nrow(res_partners_controlled))
+res_partners_controlled %>% 
+      mutate(
+        score_name = pretty_names(score_name)
+      ) %>% 
+      ggplot(aes(
+        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        x      = estimate, 
+        fill   = `N partners`, 
+        colour = p.value < 0.05/{{n_regs}}
+      )) + 
+        my_vline +
+        pval_geom_point() +
+        facet_wrap(vars(sex), scales = "free_y") +
+        my_fill_scale(n = 2) +
+        pval_scale() +
+        my_theme(legend.position = "bottom") +
+        my_labs(x = "", y = "") + 
+        scale_y_reordered()
+
+```
+
 ## Selection controlling for age at first live birth: respondents' parents

### 2020-10-15 _drake.R

@@ -90,6 +90,8 @@ plan <- drake_plan(
 
   ghs_subset   = target(make_ghs_subset(file_in(!! ghs_file)), format = "fst"),
   
+  pgs_over_time = calc_pgs_over_time(famhist, score_names),
+  
   census_age_qual = target(
                       make_census_age_qual(
                         file_in(!! DC1108EW_file),
@@ -316,7 +318,7 @@ plan <- drake_plan(
     res <- pmap_dfr(pars,
             ~ run_regs_fml(
               fml        =
-                "n_children ~ {score_name}*lo_partners",
+                "n_children ~ lo_partners + {score_name}:lo_partners",
               score_name = .y,
               famhist    = famhist,
               subset     = .x
@@ -425,7 +427,7 @@ plan <- drake_plan(
     res <- pmap_dfr(pars,
             ~ run_regs_fml(
               fml        =
-                "n_children ~ {score_name}*(lo_partners + YOB + I(YOB^2))",
+                "n_children ~ lo_partners + lo_partners:({score_name} + YOB + I(YOB^2))",
               score_name = .y,
               famhist    = famhist,
               subset     = .x

### 2020-10-15 make-data.R

@@ -7,6 +7,7 @@ suppressPackageStartupMessages({
   library(santoku)
   library(haven)
   library(sf)
+  loadNamespace("ggplot2")
   loadNamespace("car") # very annoying if it overwrites recode
   loadNamespace("matrixStats")
   loadNamespace("mlogit")
@@ -346,5 +347,3 @@ make_ghs_subset <- function(ghs_file) {
   
   ghs_subset
 }
-
-

### 2020-10-15 regressions.R

@@ -2,16 +2,56 @@
 suppressPackageStartupMessages({
   library(glue)
   library(broom)
+  library(magrittr)
+  library(santoku)
+  library(tidyr)
   library(dplyr)
   library(lmtest)
+  loadNamespace("ggplot2")
 })
+
+calc_pgs_over_time <- function (famhist, score_names) {
+  pgs_over_time <- famhist %>% 
+        group_by(
+          YOB = santoku::chop(YOB, 
+            seq(1940, 1970, 5), 
+            labels    = seq(40, 65, 5),
+            extend    = FALSE
+          ) 
+        ) %>% 
+        filter(! is.na(YOB)) %>% 
+        summarize(across(all_of(score_names), ggplot2::mean_cl_normal, na.rm = TRUE)) %>% 
+        pivot_longer(-YOB, names_to = "score_name", values_to = "score") %>% 
+        mutate(
+          score_lo = score$ymin,
+          score_hi = score$ymax,
+          score    = score$y
+        )
+  
+  run_reg <- function (score_name) {
+    fml <- as.formula(glue("{score_name} ~ YOB"))
+    tidy(lm(fml, famhist))  
+  }
+  time_regs <- map_dfr(score_names, run_reg, .id = "score_name")
+  time_regs %<>% 
+        filter(term != "(Intercept)") %>% 
+        select(score_name, p.value, statistic) %>% 
+        mutate(
+          Change = santoku::chop(statistic, c(-1.96, 1.96), labels = c("-", "o", "+"))
+        )
+  
+  pgs_over_time %<>% left_join(time_regs, by = "score_name")
+  
+  pgs_over_time
+}
 
 
 run_regs_basic <- function (dep_var, score_names, famhist) {
   
   run_reg <- function (score_name) {
     fml <- as.formula(glue("{dep_var} ~ {score_name}"))
-    reg_bv <- tidy(lm(fml, famhist), conf.int = TRUE)  
+    reg_bv <- tidy(lm(fml, famhist), conf.int = TRUE) 
+    reg_bv
   }
   basic_res <- map_dfr(score_names, run_reg, .id = "score_name")
   

### 2020-11-09 why-negative-selection.Rmd
Use earnings not household income.

@@
     latex_engine: xelatex
+header-includes:
+  - \usepackage{subfig}
+  - \usepackage{setspace}\doublespacing
+  - \usepackage{placeins}
 ---
@@
 
+\normalem
+
 # Introduction
@@
 
-```{r, eval = FALSE}
+```{r TODO, eval = FALSE}
 # Open questions/TODO
@@
   by figure 1. 
-* Reweight other regressions (e.g. the omnibus regression in the appendix)
+* Reweight other regressions (e.g. the omnibus regression in the appendix)?
 * Look at geography, esp. in siblings regressions.
+* We say we don't see change over time. But, we do see the effect of 
+  controlling for AFLB changing over generations.
 * Beauchamp 2016 used the HRS. He reckons it's representative. Is it? How do
@@
     positive correlation of about 0.1)
-
+* Another point is that negative selection only among poor will increase 
+    genetic inequality.
 ```
@@
 Lines are 95% confidence intervals.
-Green lines show a significant increase over time (p < 0.05).
+Green lines show a significant linear increase over time (p < 0.05/33).
 Red lines show a significant decrease."}
@@
     Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
-                "Age/Qual/FLB", "None"),
+                "Age/Qual/AFLB", "None"),
     term     = pretty_names(term),
@@
     "Age/Qualification" = "steelblue2",
-    "Age/Qual/FLB"      = "darkseagreen4"
+    "Age/Qual/AFLB"      = "darkseagreen4"
   )) +
@@
     "Age/Qualification" = "circle filled", 
-    "Age/Qual/FLB" = "circle filled"
+    "Age/Qual/AFLB" = "circle filled"
   ))
@@
 fertility will correlate with the strength of their relationship with earnings.
-To test this, we measure the correlation of each PGS with household income, 
-as well as with education level, a phenotype which predicts earnings. We
+To test this, we calculate the correlation of each PGS with estimated
+earnings from respondent's first job[^est-income], and with
+ education level, a phenotype which predicts earnings. We
 then relate these measurements to effect sizes on number of children.
 There is a clear negative relationship between a score's effect on
-fertility, and its correlation with income or education (Figure \@ref(fig:plot-cors-income-educ)). If we control for age at first live birth,
+fertility, and its correlation with earnings or education (Figure \@ref(fig:plot-cors-earnings-educ)). If we control for age at first live birth,
 the relationship changes sign and becomes positive.
 
-```{r plot-cors-income-educ, fig.cap="Effect sizes on number of children, with and without controls for age at first live birth, by correlations with income and education"}
+[^est-income]: Earnings are estimated from the 2007 Annual Survey of Hours and
+Earnings, using the SOC 2000 job code.
+
+```{r plot-cors-earnings-educ, fig.cap="Effect sizes on number of children, with and without controls for age at first live birth, by correlations with earnings and education.
+Each point represents one PGS."}
 
@@
 
-cors_income <- cor(famhist[score_names], famhist$income_cat, use = "pairwise")
+cors_income <- cor(famhist[score_names], famhist$first_job_pay, use = "pairwise")
 # TODO: use age_left_edu when this includes college
@@
 dfr <- data.frame(
-        Correlation  = rep(c("Income", "Education"), each = 33, 2),
+        Correlation  = rep(c("Earnings", "Education"), each = 33, 2),
         cor          = rep(c(cors_income[,1], cors_educ[,1]), 2),
@@
 sizes from bivariate regressions without the age at birth control (father's age
-at birth: $\rho$ `r cor_raw_fath`; mother's age at birth: $\rho$ `r
-cor_raw_moth`).
+at birth: $\rho$ `r cor_raw_fath`; mother's age at birth: $\rho$ `r cor_raw_moth`).
 

### 2020-11-09 _drake.R

@@ -24,6 +24,7 @@ famhist4_file      <- file.path(data_dir, "david.family_history.traits.16052020.
 famhist5_file      <- file.path(data_dir, "david.family_history.traits.18052020.out.csv")
 famhist6_file      <- file.path(data_dir, "david.family_history.traits.17062020.out.csv")
 famhist7_file      <- file.path(data_dir, "david.birthinfo.traits.14072020.out.csv")
+famhist8_file      <- file.path(data_dir, "david.traits.03112020.out.csv")
 rgs_file           <- file.path(data_dir, "EA3_rgs.10052019.rgs.csv")
 mf_pairs_file      <- file.path(data_dir, "spouse_pair_info", 
                         "UKB_out.mf_pairs_rebadged.csv")
@@ -35,6 +36,9 @@ DC1108EW_msoa_file <- file.path(data_dir, "DC1108EW-msoas.csv")
 DC5202SC_file      <- file.path(data_dir, "DC5202SC.csv")
 msoa_shapefile     <- file.path(data_dir, "infuse_msoa_lyr_2011_clipped", 
                         "infuse_msoa_lyr_2011_clipped.shp")
+ashe_income_file   <- file.path(data_dir, 
+                        "SOC-income", 
+                        "Occupation (4) Table 14.7a   Annual pay - Gross 2007.xls") 
 
 
 weighting_schemes <- rlang::syms(c("flb_weights", "age_qual_weights", 
@@ -62,6 +66,7 @@ plan <- drake_plan(
                       file_in(!! famhist5_file),
                       file_in(!! famhist6_file),
                       file_in(!! famhist7_file),
+                      file_in(!! famhist8_file),
                       file_in(!! pgs_dir)
                     ), 
                     format = "fst"
@@ -69,8 +74,10 @@ plan <- drake_plan(
   
   pcs          = read_table2(file_in(!! pcs_file)),
   
+  ashe_income = make_ashe_income(file_in(!! ashe_income_file)),
+  
   famhist      =  target(
-                   edit_famhist(famhist_raw, score_names), 
+                   edit_famhist(famhist_raw, score_names, ashe_income), 
                    format = "fst"
                   ),
   

### 2020-11-09 regressions.R

@@ -37,8 +37,11 @@ calc_pgs_over_time <- function (famhist, score_names) {
         filter(term != "(Intercept)") %>% 
         select(score_name, p.value, statistic) %>% 
         mutate(
-          Change = santoku::chop(statistic, c(-1.96, 1.96), labels = c("-", "o", "+"))
-        )
+          Change = case_when(
+            p.value > 0.05/33 ~ "o", 
+            statistic > 0     ~ "+",
+            statistic <= 0    ~ "-"
+        ))
   
   pgs_over_time %<>% left_join(time_regs, by = "score_name")
   

### 2020-11-09 why-negative-selection.Rmd
Add abstract. Rewrite intro. Rewrite conclusion. Reorganize and shorten. Move material to appendix.

@@
 author: "David Hugh-Jones, Abdel Abdellaoui"
-date: "July 2020"
+date: "`r Sys.Date()`"
+abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it.
+We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank.
+Selection effects appear consistent over time.
+The strength of natural selection on a polygenic score can be predicted by its correlation with earnings and education.
+Effects are larger among low-income and less educated individuals, younger parents, and people not living with a partner.
+Effect sizes are larger when we correct for ascertainment bias.
+The direction of natural selection is reversed among older parents, or  controlling for age at first live birth.
+These patterns can be explained by economic theories of fertility, in which higher earnings can either increase or decrease fertility."
 bibliography: "negative-selection.bib"
@@
   - \usepackage{placeins}
+editor_options: 
+  markdown: 
+    wrap: 72
 ---
@@
 
-Geneticists have long hypothesized that natural selection is taking place in
-modern human populations. Recent work confirms this using genome-wide analysis
-[@Barban_2016; @beauchamp2016genetic; @kong2017selection; @sanjak2018evidence].
-In particular, genetic variants associated with higher educational attainment
-are being selected against, although effect sizes appear small.
-
-Several gaps in our knowledge remain. First, while existing work examines
-contemporary populations, we would like to extend the analysis over time, since
-natural selection in the same direction over multiple generations
-could lead to significant effects in the long run. Second, there are concerns
-about ascertainment bias in the samples used. This could potentially bias
-estimates of the magnitude of natural selection. Lastly, we know little about
-the *mechanisms* behind natural selection in contemporary populations.
-
-Here we attempt to answer these questions, using data from UK Biobank. We use
-questions on respondents' number of children, and their number of siblings, to 
-reconstruct two generations of family size. We then relate these to 33 different
-polygenic scores for a variety of characteristics. This analysis provides a 
-broader picture of natural selection in the UK than has previously been available.
-
-Our key results are as follows. First, comparing natural selection across the
-two generations, there are few signs of change in size or direction over time.
-Most polygenic scores are selected in the same direction in both generations. We
-also see little evidence of changes across birth cohorts within either
-generation.
-
-Second, natural selection is concentrated among population subgroups which are
-under-represented in the sample, including less-educated people and poorer
-households. When we weight the sample to correct for ascertainment bias, many
-effect sizes are larger, sometimes much larger. We emphasize that our weights
-are only a partial solution to ascertainment bias in the sample. As a result,
-much uncertainty remains about the strength of natural selection in the UK
-population, but we can be reasonably sure that it is larger than naïve estimates
-suggest.
-
-Lastly, patterns in natural selection give us clues to the underlying
-mechanism. In particular, among mothers who were older at their first birth, and
-among parents with fewer sexual partners, the direction of natural selection is
-*reversed*: for example, among these subgroups, higher polygenic scores (PGS)
-for educational attainment are associated with having more offspring, rather
-than fewer. The effects of PGS can be decomposed into two channels: an effect on
-age at first live birth, and an opposite-signed effect on number of children
-conditional on age at first live birth.
-
-This pattern can be explained by economic theories of fertility
-[@becker1960economic]. In these, higher potential earnings have two opposite
-effects on fertility: a fertility-increasing *income effect* (higher income
-makes children more affordable), and a fertility-lowering *substitution effect*
-(time spent on childrearing has a higher cost in foregone earnings). Among
-couples who can take advantage of coordination in childcare activities, the
-income effect dominates; among single parents, or those in unstable
-relationships, the substitution effect dominates. Supporting this explanation,
-polygenic scores' raw effects on fertility are negatively correlated with their
-effects on income; on the other hand, controlling for age at first live birth,
-scores' effects on fertility and on household income are *positively*
-correlated. These two results are compatible with the substitution effect and the
-income effect respectively.
+Geneticists have long hypothesized that natural selection is taking
+place in modern human populations. Recent work confirms this using
+genome-wide analysis [@Barban_2016; @beauchamp2016genetic;
+@kong2017selection; @sanjak2018evidence]. In particular, genetic
+variants associated with higher educational attainment are being
+selected against, although effect sizes appear small.
+
+As yet we know little about the *mechanisms* behind natural selection in
+contemporary populations. This paper uses data from UK Biobank to learn
+more. We test for natural selection on 33 different polygenic scores. We
+extend the analysis over two generations, using data on respondents'
+number of siblings as well as their number of children to reconstruct
+two generations of family size. This is interesting because consistent
+natural selection over multiple generations could lead to substantive
+effects in the long run. Most importantly, we examine the strength of
+natural selection in different subgroups of the population, in order to
+uncover patterns that can help illuminate the mechanisms involved.
 
@@
 
-# Data
-
-Data is taken from UK Biobank. Polygenic scores were normalized to mean 0,
-variance 1.
-
 # Results
 
-We first show raw changes in polygenic scores over birth years, within the entire UK Biobank
-sample. Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores by 5-year
-birth intervals. Several scores show consistent increases or declines over this 
-30-year period, of the order of 5% of a standard deviation.
+We use data from UK Biobank. We compute 33 polygenic scores, normalized
+to mean 0, variance 1. Figure \@ref(fig:plot-means-over-time) plots mean
+polygenic scores by 5-year birth intervals. Several scores show
+consistent increases or declines over this 30-year period, of the order
+of 5% of a standard deviation.
 
@@
 
-These changes could reflect either natural selection within the UK population,
-or ascertainment bias within the sample. 
-Respondents are higher income and better educated than the UK population, and
-they may also differ on other unobserved characteristics [@10.1093/aje/kwx246].
-Since richer and educated people also live longer, this bias might also increase
-with age.
-
-To examine the effects of sampling, we start by measuring the strength of
-natural selection within different subgroups. To do this we regress respondents' 
-polygenic scores on their number of children ($y_i$):
-
-\[
+These changes could reflect either natural selection within the UK
+population, or ascertainment bias within the sample. Respondents are
+higher income and better educated than the UK population, and they may
+also differ on other unobserved characteristics [@10.1093/aje/kwx246].
+Since richer and educated people also live longer, this bias might also
+increase with age.
+
+We next regress respondents'polygenic scores on their number of children
+($y_i$):
+
+```{=tex}
+\begin{equation}
 y_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
-\]
+\label{eq:regression}
+\end{equation}
+```
 
@@
 
-The effect size $\beta$ measures the strength of natural selection within the
-sample or a given subgroup. In fact, since polygenic scores are normalized to
-mean 0, $\beta$ is the expected polygenic score among children of the sample or
-subgroup [@beauchamp2016genetic]. We begin with basic demographic variables
-including education, income and sex. These are all potential sources of ascertainment
-bias: as well as the ascertainment for income and education, mentioned above, the
-sample sex ratio  skews `r female_ratio * 100`% female.
-
-Figures \@ref(fig:plot-educ-level) and \@ref(fig:plot-income) plot effect sizes
-for each polygenic score. Figure \@ref(fig:plot-educ-level) groups respondents
-by age of completing full-time education. Figure \@ref(fig:plot-income) groups
-respondents by household income. Effect sizes are larger and more significant
-for the lowest income category, and for the lowest education category. Note that
-the overall effect is not a simple average of the effect among the different
-subgroups, because polygenic scores may also shift respondents between the
-subgroups. For example, a high PGS for educational attainment may correlate
-with fewer children among early school leavers, but may also increase the
-age at which a respondent leaves school.
-
-```{r plot-educ-level, fig.cap = "Effect sizes on number of children by age left full-time education"}
+The "selection effect", $\beta$, measures the strength of natural
+selection within the sample. In fact, since polygenic scores are
+normalized to mean 0, $\beta$ is the expected polygenic score among
+children of the sample [@beauchamp2016genetic]. To learn more about the
+underlying mechanisms, we split the sample, starting with basic
+demographic variables including education, income and sex. These are all
+potential sources of ascertainment bias: as well as the ascertainment
+for income and education, mentioned above, the sample sex ratio skews
+`r female_ratio * 100`% female.
+
+Figure \@ref(fig:plot-income-educ-level) plots selection effects for
+each polygenic score, grouping respondents by age of completing
+full-time education, and by household income. Effects are larger and
+more significant for the lowest income category, and for the lowest
+education category. Note that the overall effect is not a simple average
+of the effect among the different subgroups, because polygenic scores
+may also shift respondents between the subgroups. For example, a high
+PGS for educational attainment may correlate with fewer children among
+early school leavers, but may also increase the age at which a
+respondent leaves school.
+
+```{r plot-income-educ-level, fig.cap = "Selection effects by education and income", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}
 
@@
 
-```
-
-```{r plot-income, fig.cap = "Effect sizes on number of children by household income"}
-
 drake::loadd(res_income)
@@
 
-These results could be driven by age, if older respondents are poorer and less
-educated, and also more subject to selection on polygenic scores. However, if we
-rerun the regressions, interacting the polygenic score with income category and
-also with a quadratic in age, the interaction with income remains significant at
-0.05/`r n_regs_ic` for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
-if we interact the PGS with age of leaving full time education and a quadratic
-in age, the interaction with age leaving FTE remains significant at 
-0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.
-
-Figure \@ref(fig:plot-sexes) shows effect sizes of PGS on number of children
-separately for males and females. Differences are particularly large for
-educational attainment, height and MDD. Several PGS for mental illness and
-personality traits are more selected for (or less against) among women, including 
-major depressive disorder (MDD), schizophrenia and neuroticism, while 
-extraversion is more selected for among men. Waist circumference and waist-hip 
-ratio are less selected for among women, and educational attainment is more
-selected against, though the difference is only significant for EA3.
-
-```{r plot-sexes, fig.cap = "Effect sizes on number of children by sex. Lines show significant differences"}
+These results could be driven by age, if older respondents are poorer
+and less educated, and also more subject to selection on polygenic
+scores. However, if we rerun the regressions, interacting the polygenic
+score with income category and also with a quadratic in age, the
+interaction with income remains significant at 0.05/`r n_regs_ic` for
+`r n_sig_ic` out of `r n_regs_ic` regressions. Similarly if we interact
+the PGS with age of leaving full time education and a quadratic in age,
+the interaction with age leaving FTE remains significant at
+0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ`
+regressions.
+
+Selection effects are also different between men and women (Appendix
+Figure \@ref(fig:plot-sexes)). Differences are particularly large for
+educational attainment, height and MDD. Several PGS for mental illness
+and personality traits are more selected for (or less against) among
+women, including major depressive disorder (MDD), schizophrenia and
+neuroticism, while extraversion is more selected for among men. Waist
+circumference and waist-hip ratio are less selected for among women, and
+educational attainment is more selected against, though the difference
+is only significant for EA3.
+
+```{r calc-age-flb-cross}
+
+drake::loadd(res_age_flb_cross)
+
+cor_age_flb <- res_age_flb_cross %>% 
+      mutate(category = gsub(":.*", "", term)) %>% 
+      pivot_wider(score_name, names_from = category, values_from = estimate) %>%              {cor(.$`age_flb_cat10-22`, .$`age_flb_cat28-52`)}
+
+```
+
+We next focus on variables related to mate search and household
+formation. We split males and females by lifetime number of sexual
+partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
+For both sexes, selection effects are larger and more significant among
+those with more than 3 partners. Results are unchanged if we control for
+respondents' age (appendix Figure
+\@ref(fig:plot-n-partners-controlled)). Splitting respondents by whether
+they were living with a spouse or partner at the time of interview,
+effects are larger among those not living with a spouse or partner
+(Figure \@ref(fig:plot-n-partners)b). Lastly, we split female
+respondents by age at first live birth (AFLB).[^1] There is evidence for
+genetic effects on AFLB [@Barban_2016], and there is a close link
+between this variable and number of children born. Figure
+\@ref(fig:plot-n-partners)c shows effect sizes estimated separately for
+each tercile of AFLB. Several effects are strikingly different across
+terciles. ADHD and MDD are selected for amongst the youngest third of
+mothers, but selected against among the oldest two-thirds. Educational
+attainment is selected for among the oldest two-thirds of mothers, but
+is not significantly selected among the youngest third. Similarly,
+several PGS of body measurements are selected against only among older
+mothers. The correlation between effect sizes for the youngest and
+oldest terciles is `r cor_age_flb`. As before, recall that PGS may have
+two effects: changing the number of children within categories, and
+shifting respondents between categories.
+
+[^1]: This information is unavailable for men.
+
+```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth", fig.subcap = c("Number of partners", "Household type", "Age at first live birth"), fig.ncol = 1, out.height = c("35%", "27%", "27%"), fig.align = "center"}
+
+drake::loadd(res_partners)
+res_partners %<>%       
+      filter(grepl(":", term)) %>% 
+      mutate(`N partners` = ifelse(grepl("TRUE:", term), "<= 3", "> 3"))
+
+n_regs <- as.double(nrow(res_partners))
+res_partners %>% 
+      mutate(
+        score_name = pretty_names(score_name)
+      ) %>% 
+      ggplot(aes(
+        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        x      = estimate, 
+        fill   = `N partners`, 
+        colour = p.value < 0.05/{{n_regs}}
+      )) + 
+        my_vline +
+        pval_geom_point() +
+        facet_wrap(vars(sex), scales = "free_y") +
+        my_fill_scale(n = 2) +
+        pval_scale() +
+        my_theme(legend.position = "bottom") +
+        my_labs(x = "", y = "") + 
+        scale_y_reordered()
+
+drake::loadd(res_with_partner)
+res_with_partner %<>% filter(grepl(":", term))
+
+n_regs <- nrow(res_with_partner)
+hhtype_n <- n_in_regs("with_partner")
+
+res_with_partner %>% 
+      mutate(
+        Household = ifelse(grepl("TRUE", term), "With partner", 
+                      "Without partner"),
+        with_p    = as.character(grepl("TRUE", term)),
+        Household = sprintf("%s (N = %s)", Household, 
+                      prettyNum(hhtype_n[with_p], big.mark = ","))
+      ) %>% 
+      standard_ggplot(fill = Household, n_regs = n_regs, score_col = score_name)
+
+drake::loadd(res_age_flb_cross)
+
+# this counts each cross term as a separate test:
+n_regs <- as.double(nrow(res_age_flb_cross))
+
+res_age_flb_cross %>% 
+      mutate(
+        score_name = pretty_names(score_name),
+        score_name = fct_reorder(score_name, estimate, order_abs(1)),
+        term       = gsub("age_flb_cat(.*):.*", "\\1", term),
+        term       = add_n(term, "age_flb_cat")
+      ) %>% 
+      ggplot(aes(estimate, score_name, fill = term, 
+        colour = p.value < 0.05/{{n_regs}})) +
+        my_vline +
+        pval_geom_point() +
+        pval_scale() +
+        my_fill_scale(n = 3, direction = -1, 
+          guide = guide_legend(title = "Age at first live birth")) +
+        my_labs() +
+        my_theme()
+```
+
+```{r calc-corr-age-flb}
+
+drake::loadd(res_age_flb)
+drake::loadd(res_sex)
+
+res_sex_comparison <- res_sex %>% 
+      filter(sex == "Female") %>% 
+      mutate(score_name = term) %>% 
+      select(score_name, term, estimate:conf.high)
+
+res_age_flb %<>% filter(term != "age_flb")
+
+res_combined <- bind_rows(
+        raw      = res_sex_comparison, 
+        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
+        .id      = "reg.type"
+      ) %>% 
+      arrange(score_name)
+
+raw_flb_comparison <- res_combined %>% 
+      pivot_wider(
+        id_cols     = score_name, 
+        names_from  = reg.type, 
+        values_from = estimate
+      )
+
+cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
+opp_signed <- sum(
+        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
+      )
+
+```
+
+To investigate this further, we regress number of children on PGS
+*controlling* for AFLB (appendix Figure \@ref(fig:plot-age-flb)). In
+`r opp_signed` out of `r length(score_names)` cases, effects
+change sign when controls are added. The correlation between effect
+sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`.
+Selection effects appear to come through two opposing channels: an
+effect on AFLB, and an opposite-signed effect on number of children 
+controlling for AFLB.
+
+```{r calc-res-weighted}
+
+drake::loadd(res_wt_flb_weights)
+drake::loadd(res_wt_msoa_weights)
+drake::loadd(res_wt_age_qual_weights)
+drake::loadd(res_sibs_parent_weights)
+drake::loadd(res_all)
+drake::loadd(res_sex)
+
+res_wt_combined <- bind_rows(
+  None                     = res_all %>% 
+                               filter(reg.type == "raw", dep.var == "N children") %>% 
+                               select(-reg.type, -dep.var, -score_name),
+  Geographical             = res_wt_msoa_weights,
+  `Age/Qualification`      = res_wt_age_qual_weights,
+  `Age/Qual/AFLB`  = res_wt_flb_weights,
+  .id = "Weights"
+)
+
+res_wt_cb_wide <- res_wt_combined %>% 
+                    tidyr::pivot_wider(
+                      id_cols     = term,
+                      names_from  = Weights,
+                      values_from = estimate
+                    ) %>% 
+                    rowwise() %>% 
+                    mutate(
+                      consistent = all(c_across(-term) > 0) ||
+                        all(c_across(-term) < 0)
+                    ) %>% 
+                    ungroup() %>% 
+                    left_join(
+                      res_wt_combined %>% 
+                        filter(Weights == "None") %>% 
+                        select(term, p.value),
+                      by = "term"
+                    ) %>% 
+                    left_join(
+                      res_sex %>% 
+                        filter(sex == "Female") %>% 
+                        select(term, estimate_females = estimate),
+                      by = "term"
+                    )
+
+rel_sizes <- res_wt_cb_wide %>% 
+                filter(
+                  consistent, 
+                  p.value < 0.05/length(score_names)
+                ) %>% 
+                mutate(
+                  Geographical = Geographical/None,
+                  `Age/Qualification` = `Age/Qualification`/None,
+                  `Age/Qual/AFLB` = `Age/Qual/AFLB`/estimate_females
+                ) %>% 
+                select(-consistent, -None, -p.value, -estimate_females) %>% 
+                mutate(
+                  term = pretty_names(term)
+                )
+
+rel_sizes_summary <- rel_sizes %>% 
+                summarize(
+                  across(-term, ~ c(mean(.x), median(.x)))
+                ) %>% 
+                mutate(
+                  term = c("Mean", "Median")
+                )
+
+mean_rel <- c(rel_sizes_summary[1, 1:3])
+```
+
+Since UK Biobank subjects are not representative of the wider population, our
+results suggest that naïve estimates of natural selection are at risk of
+ascertainment bias. To correct for this, we weight participants using data from
+the 2011 Census and from the 2006 General Household Survey (GHS). We try three
+alternative weighting schemes: geographical weighting by census MSOA, age and
+presence/absence of a partner; weighting by age and highest educational
+qualification; and age, highest qualification, and age at first live birth, for
+women only. Mean effect sizes across all polygenic scores are increased by 
+`r mean_rel$Geographical` (geographical weighting), 
+`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
+`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
+further affected by weighting on other demographic variables. Since the Biobank
+sample seems to be ascertained in ways that reduce estimates of selection
+effects, we suspect that our weighted estimates remain conservative.
+
+The UK Biobank data contains information on respondents' number of
+siblings. Since respondents' polygenic scores are equal in expectation
+to the mean scores of their parents, we can use this to look at selection 
+effects in the parents' generation. We estimate equation (\ref{eq:regression})
+using *number of siblings* (including the respondent) as the dependent variable.
+
+The parents' generation has a further source of ascertainment bias: sampling
+parents of respondents overweights parents who have many children. For instance,
+parents of three children will have, on average, three times more children
+represented in UK Biobank than parents of one child. Parents of no children will
+by definition not be represented. To compensate, we reweight our preferred
+weightings (Age/Qualification) by the inverse of *number of siblings*. We then
+regress this on our polygenic scores. Selection effects are highly correlated across
+the two generations and most share the same sign (appendix Figure \@ref(fig:plot-res-sibs)). Effect sizes appear larger in the parents' generation. 
+We treat this result cautiously, because when we
+split respondents up by year of birth, we find few differences in effect
+sizes between early- and late-born respondents, for either generation. In other
+words, since estimated effect sizes change when we change the dependent
+variable, but do not change over time within either dependent variable,
+the explanation may be due to remaining ascertainment bias within our
+sample.
+
+Why do selection effects have *opposite* signs across older and younger
+mothers at first live birth, and opposite signs when AFLB is controlled
+for? A possible answer is given by the economic theory of fertility
+[@becker1960economic]. According to this, increases in potential earned
+income affect fertility via two opposing channels. There is an "income
+effect" by which children become more affordable, like any other good.
+Since childrearing has a cost in time, there is also a "substitution
+effect": the opportunity cost of childrearing increases if one's market wage is
+higher. The income effect would lead higher earners to have more children. The
+substitution effect would lead higher earners to have fewer children. It is
+often assumed that the substitution effect will dominate for lone parents, or
+those in unstable relationships, since they have less opportunity to share
+childcare responsibilities, while the income effect will dominate for couples
+who are able to reap the gains from specialization.
+
+If so, then genetic characteristics which affect one's earnings
+potential in the labour market may lead to opposing effects on
+fertility. Genetic variants which improve one's earnings may increase
+fertility among couples, but decrease fertility among single parents.
+They may also decrease the likelihood of becoming a young mother, if
+this is correlated with single motherhood. Lastly, if lifetime number of
+sexual partners predicts single parent versus two-parent status, the
+theory could also explain the different effect sizes across people with
+more and fewer sexual partners.
+
+This explanation predicts that the strength of polygenic scores' effects
+on fertility will correlate with the strength of their relationship with
+earnings. To test this, we calculate the correlation of each PGS with
+estimated earnings from respondent's first job[^earnings-est], and with education
+level, a phenotype which predicts earnings. We then relate these
+measurements to effect sizes on number of children. There is a clear
+negative relationship between a score's effect on fertility, and its
+correlation with earnings or education. If we control for age at first live
+birth, the relationship changes sign and becomes positive  (Figure
+\@ref(fig:plot-cors-earnings-educ)).
+
+[^earnings-est]: Earnings are estimated from the 2007 Annual Survey of Hours and
+    Earnings, using the SOC 2000 job code.
+
+```{r plot-cors-earnings-educ, fig.cap="Effect sizes on number of children, with and without controls for age at first live birth, by correlations with earnings and education.
+Each point represents one PGS."}
+
+drake::loadd(res_all)
+drake::loadd(res_age_flb)
+
+cors_income <- cor(famhist[score_names], famhist$first_job_pay, use = "pairwise")
+# TODO: use age_left_edu when this includes college
+cors_educ <-   cor(famhist[score_names], famhist$edu_qual, use = "pairwise")
+
+effect_size <-  res_all %>% 
+                  filter(dep.var == "N children", reg.type == "raw") %>% 
+                  pull(estimate)
+
+effect_size_flb <-  res_age_flb %>% 
+                      filter(term != "age_flb") %>% 
+                      pull(estimate)
+
+dfr <- data.frame(
+        Correlation  = rep(c("Earnings", "Education"), each = 33, 2),
+        cor          = rep(c(cors_income[,1], cors_educ[,1]), 2),
+        effect_size  = c(rep(effect_size, 2), rep(effect_size_flb, 2)),
+        Controls     = rep(c("None", 
+                        "Age at first live birth"), each = 66)
+      )
+
+dfr %>% 
+      mutate(
+        Controls = fct_relevel(Controls, "None")
+      ) %>% 
+      ggplot(aes(effect_size, cor, colour = Correlation)) + 
+        geom_point() +
+        facet_grid(Correlation ~ Controls, labeller = label_both) + 
+        my_vline + 
+        my_hline +
+        labs(x = "Effect on number of children", y = "Correlation") +
+        scale_colour_brewer(type = "qual", palette = 2) +
+        theme(legend.position = "none")
+
+```
+
+# Conclusion
+
+Previous work has documented natural selection in modern populations on variants
+underlying many polygenic scores [@beauchamp2016genetic; @kong2017selection;
+@sanjak2018evidence]. We show that natural selection effects are highly
+concentrated among specific subgroups of the population, including poorer
+people, less educated people, younger parents and those with more lifetime
+sexual partners. Indeed, among older parents selection effects are actually
+reversed. Furthermore, the size of selection effects correlates with the
+association between a PGS and labour market earnings. The economic theory of
+fertility provides a parsimonious explanation for this.
+
+An alternative theory is that traits selected for are linked to
+externalizing behaviour and low time discount rates, via the channel of
+early sexual behaviour [@Mills2020.05.06.081273]. The data here provide
+some support for this: scores which might plausibly be linked to
+externalizing behaviour, like ADHD and age at smoking initiation, are
+selected for. However, this theory is less good at explaining variation
+in selection across the full range of scores, including physical measures, e.g. waist-hip ratio and BMI. Externalizing behaviour also does not explain
+that selection effects change sign among older parents. Overall, we believe
+that the economic theory is the leading explanation at present.
+
+PGS which correlate with high (low) earnings and more (less) education are being
+selected against (selected for). In addition, many of the phenotypes under
+positive selection are linked to poor health, or would be seen as undesirable to
+have by many people. For example, most people would probably prefer to have high
+educational attainment, a low risk of major depressive disorder, and a low risk
+of coronary artery disease, but natural selection is pushing against these
+traits. Potentially, this could increase the health burden on modern societies,
+but that depends on effect sizes.
+
+Our results suggest that naïve estimates can be affected by sample ascertainment
+bias. This problem may be less serious in surveys which aim to be representative
+(as the UK Biobank does not). Nevertheless, not all respondents consent to the
+collection of genetic data: for instance, completion rates for genotype data in
+the US Health and Retirement Study  were around 80-85% [@hrs2020]. Researchers
+should be aware of the risks of ascertainment when studying contemporary 
+natural selection.
+
+In addition, we do not know whether estimated effect sizes of natural
+selection will increase or decrease as more accurate polygenic scores
+are produced. We are also unsure whether genetic variants underlying other
+phenotypes will show a similar pattern of natural selection as these ones. 
+In short, it is probably too early to tell whether modern natural
+selection is substantively important or not.
+
+Research on natural selection is controversial. It brings to mind eugenics,
+which has been used to justify mass murder and human rights violations, in the
+past and possibly the present [@mao1998chinese; @sleeboom2011genetic;
+@zenz2020]. Whilst we claim scientific legitimacy for our empirical results, we
+do not claim to speak with any special authority on how society should react to
+them. This can only be resolved by open debate, in which scientists take part as
+equals among many. We therefore choose not to discuss policy implications here.
+
+\FloatBarrier
+# Appendix
+
+## Natural selection by sex
+
+```{r plot-sexes, fig.cap = "Selection effects by sex. Lines show significant differences."}
 
@@
 
-These results show that ascertainment within UK Biobank could bias estimates
-of the size of natural selection towards zero. We now focus in on variables 
-specifically related to marriage markets and raising children. Whatever processes 
-give rise to natural selection, these variables are likely to be involved; at
-the same time, they may also be ascertained within the sample.
-
-Figure \@ref(fig:plot-n-partners) splits males and females by lifetime number of
-sexual partners, at the median value of 3. For both sexes, effects are larger
-and more significant among those with more than 3 partners. Results are
-unchanged if we control for respondents' age (Figure
-\@ref(fig:plot-n-partners-controlled) in the appendix). Figure
-\@ref(fig:plot-with-partner) splits respondents up by whether they were living
-with a spouse or partner at the time of interview. Again, effects are larger
-among those not living with a spouse or partner.
-
-```{r plot-n-partners, fig.cap = "Effect sizes on number of children by number of sexual partners", fig.height = 4}
-
-drake::loadd(res_partners)
-res_partners %<>%       
-      filter(grepl(":", term)) %>% 
-      mutate(`N partners` = ifelse(grepl("TRUE:", term), "<= 3", "> 3"))
-
-n_regs <- as.double(nrow(res_partners))
-res_partners %>% 
-      mutate(
-        score_name = pretty_names(score_name)
-      ) %>% 
-      ggplot(aes(
-        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        x      = estimate, 
-        fill   = `N partners`, 
-        colour = p.value < 0.05/{{n_regs}}
-      )) + 
-        my_vline +
-        pval_geom_point() +
-        facet_wrap(vars(sex), scales = "free_y") +
-        my_fill_scale(n = 2) +
-        pval_scale() +
-        my_theme(legend.position = "bottom") +
-        my_labs(x = "", y = "") + 
-        scale_y_reordered()
-
-```
-
-```{r plot-with-partner, fig.cap = "Effect sizes on number of children by household type"}
-
-drake::loadd(res_with_partner)
-res_with_partner %<>% filter(grepl(":", term))
-
-n_regs <- nrow(res_with_partner)
-hhtype_n <- n_in_regs("with_partner")
-
-res_with_partner %>% 
-      mutate(
-        Household = ifelse(grepl("TRUE", term), "With partner", 
-                      "Without partner"),
-        with_p    = as.character(grepl("TRUE", term)),
-        Household = sprintf("%s (N = %s)", Household, 
-                      prettyNum(hhtype_n[with_p], big.mark = ","))
-      ) %>% 
-      standard_ggplot(fill = Household, n_regs = n_regs, score_col = score_name)
-```
-
-```{r calc-age-flb-cross}
-
-drake::loadd(res_age_flb_cross)
-
-cor_age_flb <- res_age_flb_cross %>% 
-      mutate(category = gsub(":.*", "", term)) %>% 
-      pivot_wider(score_name, names_from = category, values_from = estimate) %>%              {cor(.$`age_flb_cat10-22`, .$`age_flb_cat28-52`)}
-
-```
-
-Our final variable is age at first live birth (AFLB), for females only. There is evidence
-for genetic effects on AFLB [@Barban_2016]. There is also 
-a close link between this variable and number of children born.
-
-Figure \@ref(fig:plot-age-flb-cross) shows effect sizes estimated separately for
-each tercile of AFLB. Several effects are strikingly
-different across terciles. ADHD and MDD are selected for amongst the youngest
-third of mothers, but selected against among the oldest two-thirds. Educational
-attainment is selected for among the oldest two-thirds of mothers, but is not
-significantly selected among the youngest third. Similarly, several PGS of body
-measurements are selected against only among older mothers. The correlation
-between effect sizes for the youngest and oldest terciles is `r cor_age_flb`.
-As before, recall that PGS may have two effects: they may change the number of
-children within categories, but may also shift respondents between categories.
-
-```{r plot-age-flb-cross, fig.cap = "Effect sizes on number of children, by age at first live birth terciles"}
-
-drake::loadd(res_age_flb_cross)
-
-# this counts each cross term as a separate test:
-n_regs <- as.double(nrow(res_age_flb_cross))
-
-res_age_flb_cross %>% 
-      mutate(
-        score_name = pretty_names(score_name),
-        score_name = fct_reorder(score_name, estimate, order_abs(1)),
-        term       = gsub("age_flb_cat(.*):.*", "\\1", term),
-        term       = add_n(term, "age_flb_cat")
-      ) %>% 
-      ggplot(aes(estimate, score_name, fill = term, 
-        colour = p.value < 0.05/{{n_regs}})) +
-        my_vline +
-        pval_geom_point() +
-        pval_scale() +
-        my_fill_scale(n = 3, direction = -1, 
-          guide = guide_legend(title = "Age at first live birth")) +
-        my_labs() +
-        my_theme()
-
-```
-
-```{r calc-corr-age-flb}
-
-drake::loadd(res_age_flb)
-drake::loadd(res_sex)
-
-res_sex_comparison <- res_sex %>% 
-      filter(sex == "Female") %>% 
-      mutate(score_name = term) %>% 
-      select(score_name, term, estimate:conf.high)
-
-res_age_flb %<>% filter(term != "age_flb")
-
-res_combined <- bind_rows(
-        raw      = res_sex_comparison, 
-        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
-        .id      = "reg.type"
-      ) %>% 
-      arrange(score_name)
-
-raw_flb_comparison <- res_combined %>% 
-      pivot_wider(
-        id_cols     = score_name, 
-        names_from  = reg.type, 
-        values_from = estimate
-      )
-
-cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
-opp_signed <- sum(
-        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
-      )
-
-```
-
-To investigate this further, we regress number of children on PGS *controlling*
-for AFLB. Figure \@ref(fig:plot-age-flb) shows the results, along with raw
-effect sizes for comparison. Controlling for AFLB greatly reduces effect sizes.
-This could be because the controlled regressions exclude childless women.
-However, in `r opp_signed` out of `r length(score_names)` cases, effects
-actually change sign when controls are added. The correlation between effect
-sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`. So,
-natural selection "effects" seem to come through two channels: an effect on
-AFLB, and an effect on number of children controlling for AFLB. Moreover, these
-channels often push in opposite directions.
-
-```{r plot-age-flb, fig.cap = "Effect sizes on number of children, controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}
+\FloatBarrier
+## Controlling for AFLB
+
+```{r plot-age-flb, fig.cap = "Selection effects controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}
 
@@
         score_name = fct_reorder(score_name, estimate, order_abs(2)),
-        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for age FLB")
+        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for AFLB")
       ) %>% 
@@
 
+\FloatBarrier
 ## Weighted regressions
 
-```{r calc-res-weighted}
-
-drake::loadd(res_wt_flb_weights)
-drake::loadd(res_wt_msoa_weights)
-drake::loadd(res_wt_age_qual_weights)
-drake::loadd(res_sibs_parent_weights)
-drake::loadd(res_all)
-drake::loadd(res_sex)
-
-res_wt_combined <- bind_rows(
-  None                     = res_all %>% 
-                               filter(reg.type == "raw", dep.var == "N children") %>% 
-                               select(-reg.type, -dep.var, -score_name),
-  Geographical             = res_wt_msoa_weights,
-  `Age/Qualification`      = res_wt_age_qual_weights,
-  `Age/Qual/AFLB`  = res_wt_flb_weights,
-  .id = "Weights"
-)
-
-res_wt_cb_wide <- res_wt_combined %>% 
-                    tidyr::pivot_wider(
-                      id_cols     = term,
-                      names_from  = Weights,
-                      values_from = estimate
-                    ) %>% 
-                    rowwise() %>% 
-                    mutate(
-                      consistent = all(c_across(-term) > 0) ||
-                        all(c_across(-term) < 0)
-                    ) %>% 
-                    ungroup() %>% 
-                    left_join(
-                      res_wt_combined %>% 
-                        filter(Weights == "None") %>% 
-                        select(term, p.value),
-                      by = "term"
-                    ) %>% 
-                    left_join(
-                      res_sex %>% 
-                        filter(sex == "Female") %>% 
-                        select(term, estimate_females = estimate),
-                      by = "term"
-                    )
-
-rel_sizes <- res_wt_cb_wide %>% 
-                filter(
-                  consistent, 
-                  p.value < 0.05/length(score_names)
-                ) %>% 
-                mutate(
-                  Geographical = Geographical/None,
-                  `Age/Qualification` = `Age/Qualification`/None,
-                  `Age/Qual/AFLB` = `Age/Qual/AFLB`/estimate_females
-                ) %>% 
-                select(-consistent, -None, -p.value, -estimate_females) %>% 
-                mutate(
-                  term = pretty_names(term)
-                )
-
-rel_sizes_summary <- rel_sizes %>% 
-                summarize(
-                  across(-term, ~ c(mean(.x), median(.x)))
-                ) %>% 
-                mutate(
-                  term = c("Mean", "Median")
-                )
-
-```
-
-These results show that there may be ascertainment bias in estimates of natural
-selection derived from UK Biobank. One way to compensate for this is by survey
-weighting. Weighting is not guaranteed to lead to unbiased estimates, because as
-well as ascertainment on characteristics which we measure and weight for, there
-may also be ascertainment on other, unmeasured characteristics. But the exercise
-will give us a rough sense of the direction and magnitude of bias. We weight
-participants using data from the 2011 Census and from the 2006 General Household
-Survey (GHS). We try three alternative weighting schemes:
-
-* Geographical weighting by census MSOA, age and presence/absence of a partner;
-* Weighting by age and highest educational qualification;
-* Weighting by age, highest qualification, and age at first live birth, using
-  GHS data (women only).
-
-Figure \@ref(fig:plot-res-wt) plots effect sizes among the entire sample, using
-these weighting schemes. We include unweighted regressions for comparison. Table
-\@ref(tab:tbl-res-wt) gives effect sizes as a proportion of the unweighted
-effect size. As expected, weighting generally increases estimated effect sizes.
-Sizes are reasonably similar across the geographical and age/qualification
-weights, but they are increased further when we weight by AFLB. This suggests
-that estimates might be further affected by weighting on other demographic
-variables. Given that the Biobank sample seems to be ascertained in ways that
-reduce estimates of natural selection, we suspect that these weighted estimates
-are still conservative.
+Figure \@ref(fig:plot-res-wt) plots effect sizes among the entire
+sample, using our three weighting schemes. We include unweighted regressions
+for comparison. Table \@ref(tab:tbl-res-wt) gives effect sizes as a
+proportion of the unweighted effect size.
 
@@
 
-## Change over time
-
-The UK Biobank data contains information on respondents' number of siblings.
-Since respondents' polygenic scores are equal in expectation to their parents' 
-mean scores, we can use this to look at natural selection in the parents'
-generation. The dependent variable is now *number of siblings* (including
-the respondent him- or herself).
-
-The parents' generation has a second source of ascertainment bias, beyond the
-sampling bias of the survey. Sampling parents of respondents inherently
-overweights parents who have many children. For instance, parents of three
-children will have, on average, three times more children as respondents in UK
-Biobank than parents of one child. Parents of no children will
-not be represented at all, by definition.
-
-To compensate for this, we reweight our preferred weightings (Age/Qualification)
-by the inverse of *number of siblings*. We then regress this on PGS.
-Figure \@ref(fig:plot-res-sibs) shows the results. For comparison, we show
-regressions on *number of children*, where we have excluded respondents with no 
+\FloatBarrier
+## Parents' generation
+
+Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings* on
+polygenic scores. For comparison, we show regressions on the respondents'
+generation (*number of children*), excluding respondents with no
 children.
 
-Effect sizes appear larger in the parents' generation, i.e.
-for *number of siblings*. However, we treat this result cautiously, because
-when we split respondents up by year of birth, we find few differences in
-effect sizes between early- and late-born respondents, whether our 
-dependent variable is *number of siblings* or *number of children*. In other words,
-since estimated effect sizes change when we change the dependent variable,
-but do not change over time within either dependent variable, the explanation
-may be due to remaining ascertainment bias within our sample.
-
-```{r plot-res-sibs, fig.cap = "Effects of polygenic scores on fertility, respondents' parents vs. respondents"}
+```{r plot-res-sibs, fig.cap = "Selection effects, respondents' parents vs. respondents"}
 drake::loadd(res_sibs_parent_weights)
@@
 
-## Mechanisms behind modern natural selection
-
-Why do selection effects have *opposite* signs across older and younger mothers at first live birth, and opposite signs when AFLB is controlled for? A possible answer is given by the
-economic theory of fertility [@becker1960economic]. According to this,
-increases in potential earned income affect fertility via two opposing channels.
-There is an "income effect" by which children become more affordable, like any
-other good. Since childrearing has a cost in time, there is also a substitution
-effect: the opportunity cost of childrearing increases if one's market wage is
-higher. The income effect would lead higher earners to have more children. The
-substitution effect would lead higher earners to have fewer children. It is
-often assumed that the substitution effect will dominate for lone parents, since
-they have less opportunity to share childcare responsibilities, while the income
-effect will dominate for couples who are able to reap the gains from
-specialization.
-
-If so, then genetic characteristics which affect one's earnings potential in the
-labour market may lead to opposing effects on fertility. Genetic
-variants which improve one's earnings may increase fertility among couples, but
-decrease fertility among single parents. They may also decrease the likelihood
-of becoming a young mother, if this is correlated with single motherhood. Lastly,
-if lifetime number of sexual partners predicts single parent versus 
-two-parent status, the theory could also explain the different
-effect sizes across people with more and fewer sexual partners.
-
-This explanation predicts that the strength of polygenic scores' effects on
-fertility will correlate with the strength of their relationship with earnings.
-To test this, we calculate the correlation of each PGS with estimated
-earnings from respondent's first job[^est-income], and with
- education level, a phenotype which predicts earnings. We
-then relate these measurements to effect sizes on number of children.
-There is a clear negative relationship between a score's effect on
-fertility, and its correlation with earnings or education (Figure \@ref(fig:plot-cors-earnings-educ)). If we control for age at first live birth,
-the relationship changes sign and becomes positive.
-
-[^est-income]: Earnings are estimated from the 2007 Annual Survey of Hours and
-Earnings, using the SOC 2000 job code.
-
-```{r plot-cors-earnings-educ, fig.cap="Effect sizes on number of children, with and without controls for age at first live birth, by correlations with earnings and education.
-Each point represents one PGS."}
-
-drake::loadd(res_all)
-drake::loadd(res_age_flb)
-
-cors_income <- cor(famhist[score_names], famhist$first_job_pay, use = "pairwise")
-# TODO: use age_left_edu when this includes college
-cors_educ <-   cor(famhist[score_names], famhist$edu_qual, use = "pairwise")
-
-effect_size <-  res_all %>% 
-                  filter(dep.var == "N children", reg.type == "raw") %>% 
-                  pull(estimate)
-
-effect_size_flb <-  res_age_flb %>% 
-                      filter(term != "age_flb") %>% 
-                      pull(estimate)
-
-dfr <- data.frame(
-        Correlation  = rep(c("Earnings", "Education"), each = 33, 2),
-        cor          = rep(c(cors_income[,1], cors_educ[,1]), 2),
-        effect_size  = c(rep(effect_size, 2), rep(effect_size_flb, 2)),
-        Controls     = rep(c("None", 
-                        "Age at first live birth"), each = 66)
-      )
-
-dfr %>% 
-      mutate(
-        Controls = fct_relevel(Controls, "None")
-      ) %>% 
-      ggplot(aes(effect_size, cor, colour = Correlation)) + 
-        geom_point() +
-        facet_grid(Correlation ~ Controls, labeller = label_both) + 
-        my_vline + 
-        my_hline +
-        labs(x = "Effect on number of children", y = "Correlation") +
-        scale_colour_brewer(type = "qual", palette = 2) +
-        theme(legend.position = "none")
-
-```
-
-# Conclusion
-
-Natural selection is taking place in modern populations on variants underlying
-many polygenic scores. In a loose sense, phenotypes relating to scores which are
-selected for (against) would be seen as undesirable (desirable) by many people.
-For example, most people would probably prefer to have high educational
-attainment, a low risk of major depressive disorder, and a low risk of coronary
-artery disease, but PGS underlying these traits are being selected for. In addition,
-the size of selection on PGS is negatively correlated with the score's correlation 
-with household income. Potentially, natural selection could increase the health 
-burden on modern societies, but that depends on effect sizes. Existing work 
-argues that these are small [@beauchamp2016genetic].
-
-However, as we show, estimates can be crucially affected by ascertainment
-bias in population samples. This problem may be less serious in other surveys,
-such as the US Health and Retirement Study, which explicitly aims to be
-representative (as the UK Biobank does not). Nevertheless, ascertainment bias 
-can still be a problem, since not all respondents consent to the collection of
-genetic data: for instance, completion rates were around 80-85% for HRS genotype
-data [@hrs2020].
-
-In addition, we do not know whether estimated effect sizes of natural selection
-will increase or decrease as more accurate polygenic scores are produced. We are
-also unsure about whether polygenic scores for new phenotypes will show a
-similar pattern of natural selection as existing ones. In short, it is probably
-too early to tell whether modern natural selection is substantively important or
-not.
-
-Previous work has documented natural selection in modern populations
-[@beauchamp2016genetic; @kong2017selection; @sanjak2018evidence] but has not
-explained its underlying causes. The economic theory of fertility provides a
-parsimonious explanation for which traits are selected, and for why the
-direction of selection is reversed among certain subgroups. An alternative
-theory is that traits selected for are linked to externalizing behaviour and low
-time discount rates, via the channel of early sexual behaviour
-[@Mills2020.05.06.081273]. The data here provide some support for this: scores
-which might plausibly be linked to externalizing behaviour, like ADHD and age at
-smoking initiation, are selected for. However, this theory is less
-good at explaining variation in selection across the full range of scores,
-including for example waist-hip ratio and BMI. Externalizing behaviour also does 
-not explain the reversed sign of selection among older parents. Overall, 
-we believe that the economic theory is the leading explanation at present.
-
-# Appendix
-
+\FloatBarrier
 ## Controlling for principal components
@@
 Polygenic scores could capture effects that are really due to population
-stratification, although would not change our results for natural selection of
-the scores. In \@ref(fig:plot-regs-controlled) we show results for selection on
-polygenic scores residualized for the top 100 principal components of the
-genetic data, calculated within the UK Biobank population. (TODO Abdel details.)
+stratification, although would not change our results for natural
+selection of the scores. In \@ref(fig:plot-regs-controlled) we show
+results for selection on polygenic scores residualized for the top 100
+principal components of the genetic data, calculated within the UK
+Biobank population. (TODO Abdel details.)
 
@@
 In siblings regressions, effect sizes are smaller when residualizing for
-principal components -- sometimes much smaller, as in the case of height. 
-`r sum(controlled_smaller_sibs)` out of `r n_pgs` "controlled" effect sizes have 
-a smaller absolute value than the corresponding "raw" effect size. The median
-proportion between raw and controlled effect sizes is `r median_prop_sibs`.
-Among the children regressions, this no longer holds. Effect sizes are barely
-affected by controlling for principal components.
-
-Overall, `r prop_consistent_controlled * 100` per cent of effect sizes are
-consistently signed across all four regressions (on children and siblings, and
-with and without residualization).
+principal components -- sometimes much smaller, as in the case of
+height. `r sum(controlled_smaller_sibs)` out of `r n_pgs` "controlled"
+effect sizes have a smaller absolute value than the corresponding "raw"
+effect size. The median proportion between raw and controlled effect
+sizes is `r median_prop_sibs`. Among the children regressions, this no
+longer holds. Effect sizes are barely affected by controlling for
+principal components.
+
+Overall, `r prop_consistent_controlled * 100` per cent of effect sizes
+are consistently signed across all four regressions (on children and
+siblings, and with and without residualization).
 
@@
 
-To get a further insight into this we regress *siblings* and *children* on
-individual principal components. As Figure \@ref(fig:plot-pc-regs)
+To get a further insight into this we regress *siblings* and *children*
+on individual principal components. As Figure \@ref(fig:plot-pc-regs)
 shows, effects are larger and more significant in siblings regressions.
-`r prop_sig[["N siblings"]]` principal components significantly predicted number of
-siblings, while only `r prop_sig[["N children"]]` significantly predicted number
-of children.
+`r prop_sig[["N siblings"]]` principal components significantly
+predicted number of siblings, while only `r prop_sig[["N children"]]`
+significantly predicted number of children.
 
@@
 
+\FloatBarrier
 ## Selection over time
@@
 
-To check whether effect sizes were changing over time, we ran regressions 
-interacting PGS with birth year, median split at 1950 ("early born" versus "late born"). We use both 
-*number of children* and *number of siblings* as a dependent variable. We weight 
-using age/qualification cells, and further adjust for selection in the parents'
-generation (see above).
-
-Tables \@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)
-summarize the results. There is no change in natural selection in the parents'
-generation. In the respondents' generation, effect sizes were larger in
-absolute size among the late born for five PGS: autism, cognitive ability,
-EA2, EA3 and extraversion. Three PGS showed significant changes in sign: 
-alcohol use (positive to negative); conscientiousness (negative to positive);
-and type II diabetes (negative to positive).
+To check whether effect sizes were changing over time, we ran
+regressions interacting PGS with birth year, median split at 1950
+("early born" versus "late born"). We use both *number of children* and
+*number of siblings* as a dependent variable. We weight using
+age/qualification cells, and further adjust for selection in the
+parents' generation (see above).
+
+Tables \@ref(tab:tbl-siblings-by-period) and
+\@ref(tab:tbl-children-by-period) summarize the results. There is no
+change in natural selection in the parents' generation. In the
+respondents' generation, effect sizes were larger in absolute size among
+the late born for five PGS: autism, cognitive ability, EA2, EA3 and
+extraversion. Three PGS showed significant changes in sign: alcohol use
+(positive to negative); conscientiousness (negative to positive); and
+type II diabetes (negative to positive).
 
@@
 Figure \@ref(fig:plot-n-partners-controlled) shows selection effects by
-lifetime number of sexual partners. Age, age squared, and interactions of 
-age and age squared with PGS are controlled for.
+lifetime number of sexual partners. Age, age squared, and interactions
+of age and age squared with PGS are controlled for.
 
@@
 
-Among the parents' generation, we can control for age at first live birth using
-the subsets of respondents who reported their mother's or father's age, and who
-had no elder siblings. We run regressions on *number of siblings* on these
-subsets, controlling for either parent's age at their birth. Figure
-\@ref(fig:plot-age-birth-parents) shows the results. Effect sizes are very
-similar, whether controlling for father's or mother's age. Unlike for the
-respondents' generation, effect sizes are positively correlated with the effect
-sizes from bivariate regressions without the age at birth control (father's age
-at birth: $\rho$ `r cor_raw_fath`; mother's age at birth: $\rho$ `r cor_raw_moth`).
+Among the parents' generation, we can control for age at first live
+birth using the subsets of respondents who reported their mother's or
+father's age, and who had no elder siblings. We run regressions on
+*number of siblings* on these subsets, controlling for either parent's
+age at their birth. Figure \@ref(fig:plot-age-birth-parents) shows the
+results. Effect sizes are very similar, whether controlling for father's
+or mother's age. Unlike for the respondents' generation, effect sizes
+are positively correlated with the effect sizes from bivariate
+regressions without the age at birth control (father's age at birth:
+$\rho$ `r cor_raw_fath`; mother's age at birth: $\rho$
+`r cor_raw_moth`).
 
@@
 estimated effect sizes from bivariate regressions for respondents, and
-Figure \@ref(fig:plot-age-birth-parents-dv) does the same for their parents. 
-Effect sizes are reasonably large. They are also very highly correlated across 
-generations. Effect sizes of PGS on father's age at own birth, and on own age
-at first live birth, have a correlation of `r cor_flb_fath`; for mother's age
-and own age it is `r cor_flb_moth`
+Figure \@ref(fig:plot-age-birth-parents-dv) does the same for their
+parents. Effect sizes are reasonably large. They are also very highly
+correlated across generations. Effect sizes of PGS on father's age at
+own birth, and on own age at first live birth, have a correlation of
+`r cor_flb_fath`; for mother's age and own age it is `r cor_flb_moth`
 
@@
 
-Another way to examine the "earnings" theory of natural selection is to compare
-effect sizes of PGS with their genetic correlation with educational attainment
-(EA3). Since EA3 strongly predicts earnings, if earnings drives differences
-in fertility, we'd expect a correlation between the two sets of results. Figure
+Another way to examine the "earnings" theory of natural selection is to
+compare effect sizes of PGS with their genetic correlation with
+educational attainment (EA3). Since EA3 strongly predicts earnings, if
+earnings drives differences in fertility, we'd expect a correlation
+between the two sets of results. Figure
 \@ref(fig:plot-rgs-by-effect-size) shows this is so.
@@
 
-Figure \@ref(fig:plot-n-children-bar) shows the full distribution of number 
-of children born for different ventiles of the EA3 polygenic score. The 
-strongest relationship seems to be for having 0 children versus 1 or more.
+Figure \@ref(fig:plot-n-children-bar) shows the full distribution of
+number of children born for different ventiles of the EA3 polygenic
+score. The strongest relationship seems to be for having 0 children
+versus 1 or more.
 
@@
 
-Different polygenic scores are correlated. Table \@ref(tab:tbl-pgs-corrs) shows the top
-correlations in the sample. Because of this, bivariate correlations between
-PGS and number of children might be driven by other genetic scores. To explore which polygenic scores are driving negative selection, we run a single
-omnibus regression of *number of children* on all the PGS. We exclude EA2, 
+Different polygenic scores are correlated. Table
+\@ref(tab:tbl-pgs-corrs) shows the top correlations in the sample.
+Because of this, bivariate correlations between PGS and number of
+children might be driven by other genetic scores. To explore which
+polygenic scores are driving negative selection, we run a single omnibus
+regression of *number of children* on all the PGS. We exclude EA2,
 waist-hip ratio, waist-circumference, and "Hip combined" since they are
-highly correlated with other scores, which could make our estimates unstable.
-Figure \@ref(fig:plot-together) shows the results.
-Interestingly, several PGS remain independently significant, although effect sizes are reduced.
+highly correlated with other scores, which could make our estimates
+unstable. Figure \@ref(fig:plot-together) shows the results.
+Interestingly, several PGS remain independently significant, although
+effect sizes are reduced.
 
@@
 
+\FloatBarrier
+
 # References

### 2020-11-12 why-negative-selection.Rmd
papers; notes; exploration of sex, pay, married couples

@@
     latex_engine: xelatex
+    number_sections: false
 header-includes:
@@
 
-# Introduction
-
 Geneticists have long hypothesized that natural selection is taking
@@
 
-```{r TODO, eval = FALSE}
+```{r NOTES-TODO, eval=FALSE}
+"
 # Open questions/TODO
@@
     genetic inequality.
-```
-
-# Results
+* When we have better definition of "partner", bring it across from the other
+  project to define with_partner
+* If you take "year of last job started" and see if it is after/before
+  the first child, you can figure out whether mothers keep working.
+  - better still, look at all jobs started and see if any are within e.g. 5 years of
+  first child's birth
+  - this could then be another way to get at income vs substitution effects. You'd
+  predict negative selection effects among mothers who worked. Of course the
+  choice to work is endogenous.
+* Put mediating effects of pay and education in there (see explore-sex-vs-pay.R)
+  - in a multivariate regression, most effects vanish once we control for these, including EA3
+
+## LITERATURE NOTES
+
+* Lundberg and Pollak JEP 2007:
+    - several models where nonmarital fertility is more likely for poorer women
+    - Burdett and Ermisch 2002, Willis 1999, Neal 2004
+    - Unmarried parents may not be able to credibly commit to each other
+
+* Happel et al. 1994
+   - childbirth is chosen to minimize loss of earnings from interrupted work.
+   - early childbirth if your skills decay a lot  while childrearing. (No point
+     gaining them then losing them.) Late childbirth otherwise.
+   - late childbirth if you have acquired skills before marriage.
+   - Capital-market imperfections make it wise to delay childbirth (till man has
+     high income to cope while wife is at home). More so for men's wages which
+     rise faster.
+   - empirically: (among married couples) 
+     - high-skill occupation women have children later. Higher-earning men have 
+     them earlier.
+
+- Caucutt et al. 2001
+  - high waged and high educated women (of recent cohorts) more likely to
+    delay childbirth
+
+- Cohen et al. 2013
+  - analyse effect of Israeli subsidy
+  - along the way, find that income affects fertility negatively at low income,
+    positively at high income, supporting Becker and Tomes 76
+
+- Becker and Tomes 1976
+  - investment in children quality and quantity when quality must be equal
+    across children; and some quality comes from an innate endowment.
+  - "observed" quantity-of-children income elasticity rises as income rises
+  - while "observed" quality-of-children elasticity falls as income rises, 
+    b/c more of total quality must be paid for by parents themselves. 
+  - the need to pay more for quality results in lower quantity, since total cost is 
+    quantity X quality
+  - something fishy about argument somehow...
+
+- Lindo 2008
+  - men who lose a job have children sooner, but fewer children in total
+  - controls for all skills effects
+  - so overall effect of income is positive, but timing changes
+
+- Monstad et al. 2008
+  - using an education reform to vary education
+  - no evidence of a causal effect on fertility, but births start later
+  * maybe the usual correlation reflects underlying skills. Your data could
+    test this.
+
+- Autor et al. 2017
+  - the China shock (exogenous shock to income) 
+  - they use Chinese import penetration by industry (instrumented using
+    other countries) times industry employment shares; and separate it for
+    males and females
+  - shocks to male employment decreased fertility and marriage; 
+    increased unmarried mothers
+  - shocks to female employment *increase* marriage and motherhood
+  * this roughly seems to be the consensus
+  * Whereas, you find that male and female *skills* (as 'instrumented' by PGS)
+    have the same effect on fertility - i.e. less skills = higher fertility
+
+# WHAT MODEL FITS YOUR DATA
+## ASSUMPTIONS
+* Genetics produce human capital h. 
+* h and education are compliments in producing earnings
+* There are income and substitution effects on fertility:
+  - income effects are stronger at high income levels (Becker and Tomes 1976)
+    and among couples
+* People seek partners when in the labour market. Women match a person, observe
+  their quality, and decide whether to have a first child.
+* Then they (may) have more children, either with the partner or not 
+  (high quality partners stay around more)
+* High quality partners increase household income
+
+## RESULTS
+* People with low h are more likely to leave education early and enter the labour market
+  - because e and h are complements
+  - this is part of the substitution effect, as they expect to spend some time
+    raising children rather than being in work
+* At low values of h, increases in h reduce fertility (substitution effect dominates)
+  at higher values, smaller reductions or increases (income effect stronger)
+  - because h predicts income
+* In 1-parent households, increases in h reduce fertility (sub. effect), in 
+  2-parent households, smaller reductions or increases (income effect stronger)
+  - because h predicts income and because stronger income effect among couples
+* People with low h start having children earlier (b/c leave educ earlier)
+  - so part of the substitution effect comes via this
+  - controlling for this the income effect dominates
+  - hence increases in h reduce fertility less once we control for AFLB
+
+## THOUGHTS
+* Should we distinguish between men and women in couples?
+  - within couples, male h should increase fertility (via earnings)
+  - female h should decrease it (ditto?)
+  - see explore-sex-vs-pay.R
+"
+```
 
@@
 
-```{r plot-means-over-time, fig.height = 5, fig.cap = "Mean polygenic scores by birth year in UK Biobank.
+```{r plot-means-over-time, fig.height = 5.5, fig.cap = "Mean polygenic scores by birth year in UK Biobank.
 Points are means for 5-year intervals.
@@
       geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
-        fatten = 2) +
+        fatten = 1.25) +
       facet_wrap(~score_name, ncol = 5) +
@@
 
-Why do selection effects have *opposite* signs across older and younger
-mothers at first live birth, and opposite signs when AFLB is controlled
-for? A possible answer is given by the economic theory of fertility
-[@becker1960economic]. According to this, increases in potential earned
-income affect fertility via two opposing channels. There is an "income
-effect" by which children become more affordable, like any other good.
-Since childrearing has a cost in time, there is also a "substitution
-effect": the opportunity cost of childrearing increases if one's market wage is
-higher. The income effect would lead higher earners to have more children. The
-substitution effect would lead higher earners to have fewer children. It is
-often assumed that the substitution effect will dominate for lone parents, or
-those in unstable relationships, since they have less opportunity to share
-childcare responsibilities, while the income effect will dominate for couples
-who are able to reap the gains from specialization.
+Why do selection effects have *opposite* signs across older and younger mothers
+at first live birth, and opposite signs when AFLB is controlled for? A possible
+answer is given by the economic theory of fertility [@becker1960economic; @willis1973new].
+According to this, increases in potential earned income affect fertility via two
+opposing channels. There is an "income effect" by which children become more
+affordable, like any other good. There is also a "substitution effect": since
+childrearing has a cost in time, the opportunity cost of childrearing increases
+if one's market wage is higher. The income effect would lead higher earners to
+have more children. The substitution effect would lead higher earners to have
+fewer children. It is often assumed that the substitution effect will dominate
+for lone parents, or those in unstable relationships, since they have less
+opportunity to share childcare responsibilities, while the income effect will
+dominate for couples who are able to reap the gains from specialization.
 
@@
 
-# Conclusion
+# Discussion
 
@@
 selected for. However, this theory is less good at explaining variation
-in selection across the full range of scores, including physical measures, e.g. waist-hip ratio and BMI. Externalizing behaviour also does not explain
-that selection effects change sign among older parents. Overall, we believe
-that the economic theory is the leading explanation at present.
+in selection across the full range of scores, including physical measures, e.g.
+waist-hip ratio and BMI. Externalizing behaviour also does not explain
+why selection effects change sign among older parents. Overall, we believe
+that the economic theory is the leading explanation.
 
@@
 traits. Potentially, this could increase the health burden on modern societies,
-but that depends on effect sizes.
-
-Our results suggest that naïve estimates can be affected by sample ascertainment
-bias. This problem may be less serious in surveys which aim to be representative
-(as the UK Biobank does not). Nevertheless, not all respondents consent to the
-collection of genetic data: for instance, completion rates for genotype data in
-the US Health and Retirement Study  were around 80-85% [@hrs2020]. Researchers
-should be aware of the risks of ascertainment when studying contemporary 
-natural selection.
+but that depends on effect sizes. Our results suggest that naïve estimates can
+be affected by sample ascertainment bias. This problem may be less serious in
+surveys which aim to be representative (as the UK Biobank does not).
+Nevertheless, not all respondents consent to the collection of genetic data: for
+instance, completion rates for genotype data in the US Health and Retirement
+Study were around 80-85% [@hrs2020]. Researchers should be aware of the risks
+of ascertainment when studying modern natural selection.
 
@@
 @zenz2020]. Whilst we claim scientific legitimacy for our empirical results, we
-do not claim to speak with any special authority on how society should react to
+have no special authority to speak on how society should react to
 them. This can only be resolved by open debate, in which scientists take part as
@@
 
-\FloatBarrier
-## Controlling for principal components
-
-Polygenic scores could capture effects that are really due to population
-stratification, although would not change our results for natural
-selection of the scores. In \@ref(fig:plot-regs-controlled) we show
-results for selection on polygenic scores residualized for the top 100
-principal components of the genetic data, calculated within the UK
-Biobank population. (TODO Abdel details.)
-
-```{r calc-regs-controlled}
-
-drake::loadd(res_all)
-
-res_all_controlled <- res_all %>% 
-  select(reg.type, dep.var, term, estimate) %>% 
-  pivot_wider(names_from = c(reg.type, dep.var), values_from = estimate) %>% 
-  mutate(
-    consistent = sign(`raw_N siblings`) == sign(`raw_N children`) &
-                 sign(`raw_N siblings`) == sign(`controlled_N siblings`) &
-                 sign(`raw_N siblings`) == sign(`controlled_N children`)
-  )
-
-n_pgs  <- as.double(nrow(res_all_controlled))
-prop_consistent_controlled <- mean(res_all_controlled$consistent)
-controlled_smaller_sibs <- abs(res_all_controlled[["controlled_N siblings"]]) < 
-  abs(res_all_controlled[["raw_N siblings"]])
-controlled_smaller_children <- abs(res_all_controlled[["controlled_N children"]]) < 
-  abs(res_all_controlled[["raw_N children"]])
-median_prop_sibs <- median(abs(res_all_controlled[["controlled_N siblings"]]) /
-    abs(res_all_controlled[["raw_N siblings"]]))
-median_prop_children <- median(abs(res_all_controlled[["controlled_N children"]]) /
-    abs(res_all_controlled[["raw_N children"]]))
-
-```
-
-In siblings regressions, effect sizes are smaller when residualizing for
-principal components -- sometimes much smaller, as in the case of
-height. `r sum(controlled_smaller_sibs)` out of `r n_pgs` "controlled"
-effect sizes have a smaller absolute value than the corresponding "raw"
-effect size. The median proportion between raw and controlled effect
-sizes is `r median_prop_sibs`. Among the children regressions, this no
-longer holds. Effect sizes are barely affected by controlling for
-principal components.
-
-Overall, `r prop_consistent_controlled * 100` per cent of effect sizes
-are consistently signed across all four regressions (on children and
-siblings, and with and without residualization).
-
-```{r plot-regs-controlled, fig.cap = "Effects of residualized polygenic scores on number of siblings/children."}
-
-drake::loadd(res_all)
-
-n_regs <- as.double(nrow(res_all))
-res_all %>% 
-      mutate(
-        term    = pretty_names(term),
-        term    = reorder_within(term, estimate, dep.var, order_abs(1)),
-        dep.var = fct_inorder(dep.var) 
-      ) %>% 
-      ggplot(aes(y = term, x = estimate, fill = reg.type, 
-        colour = p.value < 0.05/{{n_regs}})) +
-        my_vline + 
-        facet_wrap(vars(dep.var), scales = "free_y") +
-        pval_geom_point(position = position_dodge(0.4)) +
-        my_labs() +
-        pval_scale() +
-        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
-        my_theme() +
-        scale_y_reordered()
-
-```
-
-```{r}
-drake::loadd(res_pcs)
-n_regs <- as.double(nrow(res_pcs))
-prop_sig <- res_pcs %>% 
-      group_by(dep.var) %>% 
-      summarize(sig = sum(p.value <= 0.05/nrow(res_pcs))) %>% 
-      pull(sig, name = dep.var)
-```
-
-To get a further insight into this we regress *siblings* and *children*
-on individual principal components. As Figure \@ref(fig:plot-pc-regs)
-shows, effects are larger and more significant in siblings regressions.
-`r prop_sig[["N siblings"]]` principal components significantly
-predicted number of siblings, while only `r prop_sig[["N children"]]`
-significantly predicted number of children.
-
-```{r plot-pc-regs, fig.cap = "Effect of principal components of genetic data on number of siblings/children.
-Absolute effect sizes are plotted.
-Each dot represents one bivariate regression.
-Points are jittered on the Y axis.", fig.height = 2}
-
-res_pcs %>% 
-      mutate(
-        estimate = abs(estimate),
-        dep.var  = fct_relevel(dep.var, "N children")
-      ) %>% 
-      ggplot(aes(estimate, dep.var, colour = p.value < 0.05/{{n_regs}})) + 
-        geom_jitter(height = 0.15, width = 0, alpha = 0.9) + 
-        scale_x_log10(labels = scales::label_number(accuracy = 0.0001)) +
-        labs(x = "Effect size (log scale)", y = "") + 
-        pval_scale(aesthetics = "colour")
-
-```
-
-\FloatBarrier
-## Selection over time
-
 ```{r calc-period-regs}
@@
 \@ref(tab:tbl-children-by-period) summarize the results. There is no
-change in natural selection in the parents' generation. In the
-respondents' generation, effect sizes were larger in absolute size among
-the late born for five PGS: autism, cognitive ability, EA2, EA3 and
-extraversion. Three PGS showed significant changes in sign: alcohol use
+change in natural selection within the parents' generation. In the
+respondents' generation, effect sizes were significantly larger in absolute size among the late-born for five PGS: autism, cognitive ability, EA2, EA3 and
+extraversion. These changes are inconsistent with the intergenerational change, 
+where effect sizes were if anything larger among the earlier, parents' generation.
+Three PGS showed significant changes in sign: alcohol use
 (positive to negative); conscientiousness (negative to positive); and
-type II diabetes (negative to positive).
+type II diabetes (negative to positive). Overall, there is weak evidence for
+change over time, and the clearest result is that the direction of selection is
+broadly consistent.
 
@@
       period_table("Change in effect sizes between early and late born respondents (regressions on number of children)")
+```
+
+\FloatBarrier
+## Controlling for principal components
+
+Polygenic scores could capture effects that are really due to population
+stratification, although would not change our results for natural
+selection of the scores. In \@ref(fig:plot-regs-controlled) we show
+results for selection on polygenic scores residualized for the top 100
+principal components of the genetic data, calculated within the UK
+Biobank population. (TODO Abdel details.)
+
+```{r calc-regs-controlled}
+
+drake::loadd(res_all)
+
+res_all_controlled <- res_all %>% 
+  select(reg.type, dep.var, term, estimate) %>% 
+  pivot_wider(names_from = c(reg.type, dep.var), values_from = estimate) %>% 
+  mutate(
+    consistent = sign(`raw_N siblings`) == sign(`raw_N children`) &
+                 sign(`raw_N siblings`) == sign(`controlled_N siblings`) &
+                 sign(`raw_N siblings`) == sign(`controlled_N children`)
+  )
+
+n_pgs  <- as.double(nrow(res_all_controlled))
+prop_consistent_controlled <- mean(res_all_controlled$consistent)
+controlled_smaller_sibs <- abs(res_all_controlled[["controlled_N siblings"]]) < 
+  abs(res_all_controlled[["raw_N siblings"]])
+controlled_smaller_children <- abs(res_all_controlled[["controlled_N children"]]) < 
+  abs(res_all_controlled[["raw_N children"]])
+median_prop_sibs <- median(abs(res_all_controlled[["controlled_N siblings"]]) /
+    abs(res_all_controlled[["raw_N siblings"]]))
+median_prop_children <- median(abs(res_all_controlled[["controlled_N children"]]) /
+    abs(res_all_controlled[["raw_N children"]]))
+
+```
+
+In siblings regressions, effect sizes are smaller when residualizing for
+principal components -- sometimes much smaller, as in the case of
+height. `r sum(controlled_smaller_sibs)` out of `r n_pgs` "controlled"
+effect sizes have a smaller absolute value than the corresponding "raw"
+effect size. The median proportion between raw and controlled effect
+sizes is `r median_prop_sibs`. Among the children regressions, this no
+longer holds. Effect sizes are barely affected by controlling for
+principal components.
+
+Overall, `r prop_consistent_controlled * 100` per cent of effect sizes
+are consistently signed across all four regressions (on children and
+siblings, and with and without residualization).
+
+```{r plot-regs-controlled, fig.cap = "Effects of residualized polygenic scores on number of siblings/children."}
+
+drake::loadd(res_all)
+
+n_regs <- as.double(nrow(res_all))
+res_all %>% 
+      mutate(
+        term    = pretty_names(term),
+        term    = reorder_within(term, estimate, dep.var, order_abs(1)),
+        dep.var = fct_inorder(dep.var) 
+      ) %>% 
+      ggplot(aes(y = term, x = estimate, fill = reg.type, 
+        colour = p.value < 0.05/{{n_regs}})) +
+        my_vline + 
+        facet_wrap(vars(dep.var), scales = "free_y") +
+        pval_geom_point(position = position_dodge(0.4)) +
+        my_labs() +
+        pval_scale() +
+        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
+        my_theme() +
+        scale_y_reordered()
+
+```
+
+```{r}
+drake::loadd(res_pcs)
+n_regs <- as.double(nrow(res_pcs))
+prop_sig <- res_pcs %>% 
+      group_by(dep.var) %>% 
+      summarize(sig = sum(p.value <= 0.05/nrow(res_pcs))) %>% 
+      pull(sig, name = dep.var)
+```
+
+To get a further insight into this we regress *siblings* and *children*
+on individual principal components. As Figure \@ref(fig:plot-pc-regs)
+shows, effects are larger and more significant in siblings regressions.
+`r prop_sig[["N siblings"]]` principal components significantly
+predicted number of siblings, while only `r prop_sig[["N children"]]`
+significantly predicted number of children.
+
+```{r plot-pc-regs, fig.cap = "Effect of principal components of genetic data on number of siblings/children.
+Absolute effect sizes are plotted.
+Each dot represents one bivariate regression.
+Points are jittered on the Y axis.", fig.height = 2}
+
+res_pcs %>% 
+      mutate(
+        estimate = abs(estimate),
+        dep.var  = fct_relevel(dep.var, "N children")
+      ) %>% 
+      ggplot(aes(estimate, dep.var, colour = p.value < 0.05/{{n_regs}})) + 
+        geom_jitter(height = 0.15, width = 0, alpha = 0.9) + 
+        scale_x_log10(labels = scales::label_number(accuracy = 0.0001)) +
+        labs(x = "Effect size (log scale)", y = "") + 
+        pval_scale(aesthetics = "colour")
+
 ```
@@
 Another way to examine the "earnings" theory of natural selection is to
-compare effect sizes of PGS with their genetic correlation with
+compare selection effects of PGS with their genetic correlation with
 educational attainment (EA3). Since EA3 strongly predicts earnings, if

### 2020-11-12 explore-sex-vs-pay.R

@@ -1,0 +1,90 @@
+
+map_dfr(score_names, 
+        ~ (paste("n_children ~ sex * fj_q + sex * fj_q:", .x) %>% 
+             as.formula %>% 
+             lm(data = famhist) %>% 
+             tidy()
+           ), .id = "score_name") -> tmp
+tmp %>% 
+  mutate(female = grepl("sex", term), pay = sub(".*?(\\d+%).*", "\\1", term)) %>% 
+  mutate(term_bit = gsub(".*:", "", term)) %>% 
+  filter(term_bit == score_name) %>% 
+  select(-term_bit) %>% 
+  ggplot(aes(pay, estimate, color = female, group = female)) + 
+    geom_line(position = "dodge") + 
+    facet_wrap(vars(score_name), ) +
+    geom_hline(yintercept = 0)
+
+map_dfr(score_names, 
+        ~ (paste("n_children ~ sex * ", .x) %>% 
+             as.formula %>% 
+             lm(data = famhist) %>% 
+             tidy()), .id = "score_name") -> tmp2
+map_dfr(score_names, 
+        ~ (paste("first_job_pay ~ factor(sex):", .x) %>% 
+             as.formula %>% lm(data = famhist) %>% 
+             tidy()), .id = "score_name") -> tmp3
+
+tmp2 <- left_join(tmp2, tmp3, by = c("score_name", "term"), suffix = c(".kids", ".pay"))
+tmp2 <- tmp2 %>% filter(term != "(Intercept)") %>% mutate(female = grepl("sex.0", term))
+ggplot(tmp2, aes(estimate.kids, estimate.pay, color = female)) + 
+  geom_point() +
+  geom_text(aes(label = score_name))
+
+
+# explore controlling for pay and/or education
+fml <- paste("n_children ~ ", paste(score_names, collapse = " + "))
+fml <- as.formula(fml)
+fml <- update(fml, .~.+sex+age_at_recruitment+I(age_at_recruitment^2))
+fml <- update(fml, . ~ . - EA2_noUKB - wc_combined - whr_combined - hip_combined - autism_2017 - body_fat)
+
+tmp <- lm(fml, data=famhist)
+tmp2 <- lm(update(fml, .~.+log(first_job_pay) + factor(edu_qual)), data= famhist)
+
+tmp <- tidy(tmp)
+tmp2 <- tidy(tmp2)
+tmp3 <- inner_join(tmp, tmp2, by = "term")
+
[…]

### 2020-11-13 why-negative-selection.Rmd
Rewrite discussion of economic theory. Add some more papers.

@@
 The direction of natural selection is reversed among older parents, or  controlling for age at first live birth.
-These patterns can be explained by economic theories of fertility, in which higher earnings can either increase or decrease fertility."
+These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility."
 bibliography: "negative-selection.bib"
@@
 
+We use data from UK Biobank. We compute 33 polygenic scores, normalized
+to mean 0, variance 1. Figure \@ref(fig:plot-means-over-time) plots mean
+polygenic scores by 5-year birth intervals. Several scores show
+consistent increases or declines over this 30-year period, of the order
+of 5% of a standard deviation.
+
+```{r plot-means-over-time, fig.height = 5.5, fig.cap = "Mean polygenic scores by birth year in UK Biobank.
+Points are means for 5-year intervals.
+Lines are 95% confidence intervals.
+Green lines show a significant linear increase over time (p < 0.05/33).
+Red lines show a significant decrease."}
+
+drake::loadd(pgs_over_time)
+
+pgs_over_time %>% 
+      mutate(score_name = pretty_names(score_name)) %>% 
+      ggplot(aes(YOB, group = score_name, colour = Change)) + 
+      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
+        fatten = 1.25) +
+      facet_wrap(~score_name, ncol = 5) +
+      my_hline +
+      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
+      theme(
+        strip.text  = element_text(size = 6), 
+        axis.text.x = element_text(size = 7),
+        legend.position = "none"
+      ) +
+      labs(y = "PGS", x = "Birth year (5 year intervals)")
+
+```
+
+These changes could reflect either natural selection within the UK
+population, or ascertainment bias within the sample. Respondents are
+higher income and better educated than the UK population, and they may
+also differ on other unobserved characteristics [@10.1093/aje/kwx246].
+Since richer and educated people also live longer, this bias might also
+increase with age.
+
+We next regress respondents'polygenic scores on their number of children
+($y_i$):
+
+```{=tex}
+\begin{equation}
+y_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
+\label{eq:regression}
+\end{equation}
+```
+
+```{r calc-sex-ratio}
+female_ratio <- mean(famhist$sex == 0)
+```
+
+The "selection effect", $\beta$, measures the strength of natural
+selection within the sample. In fact, since polygenic scores are
+normalized to mean 0, $\beta$ is the expected polygenic score among
+children of the sample [@beauchamp2016genetic]. To learn more about the
+underlying mechanisms, we split the sample, starting with basic
+demographic variables including education, income and sex. These are all
+potential sources of ascertainment bias: as well as the ascertainment
+for income and education, mentioned above, the sample sex ratio skews
+`r female_ratio * 100`% female.
+
+Figure \@ref(fig:plot-income-educ-level) plots selection effects for
+each polygenic score, grouping respondents by age of completing
+full-time education, and by household income. Effects are larger and
+more significant for the lowest income category, and for the lowest
+education category. Note that the overall effect is not a simple average
+of the effect among the different subgroups, because polygenic scores
+may also shift respondents between the subgroups. For example, a high
+PGS for educational attainment may correlate with fewer children among
+early school leavers, but may also increase the age at which a
+respondent leaves school.
+
+```{r plot-income-educ-level, fig.cap = "Selection effects by education and income", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}
+
+drake::loadd(res_edu)
+n_regs <- as.double(nrow(res_edu))
+
+res_edu %>% 
+      mutate(
+        term = pretty_names(term), 
+        term = fct_reorder(term, estimate, order_abs(1)),
+        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
+        "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
+      ) %>% 
+      ggplot(aes(y = term, x = estimate, colour = p.value < 0.05/{{n_regs}}, 
+        fill = `Age left FTE`)) +
+        my_vline +
+        pval_geom_point() + 
+        pval_scale() +
+        my_fill_scale(n = 3, direction = -1) +
+        my_labs() +
+        my_theme()
+
+drake::loadd(res_income)
+
+n_regs <- as.double(nrow(res_income))
+res_income %>% 
+      mutate(
+        term = pretty_names(term), 
+        term = fct_reorder(term, estimate, order_abs(1)),
+        Income = factor(income_cat, 
+          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
+        ),
+        Income = add_n(Income, "income_cat")
+      ) %>% 
+      ggplot(aes(y = term, x = estimate, fill = Income, 
+        colour = p.value < 0.05/{{n_regs}})) +
+        my_vline +
+        pval_geom_point(alpha = 0.8) + 
+        my_fill_scale(n = 5, direction = -1) +
+        pval_scale() +
+        my_labs() +
+        my_theme()     
+```
+
+```{r calc-income-edu-controlled}
+drake::loadd(res_income_controlled, res_edu_controlled)
+n_regs_ic <- nrow(res_income_controlled)
+n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
+
+n_regs_educ <- nrow(res_edu_controlled)
+n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
+```
+
+These results could be driven by age, if older respondents are poorer
+and less educated, and also more subject to selection on polygenic
+scores. However, if we rerun the regressions, interacting the polygenic
+score with income category and also with a quadratic in age, the
+interaction with income remains significant at 0.05/`r n_regs_ic` for
+`r n_sig_ic` out of `r n_regs_ic` regressions. Similarly if we interact
+the PGS with age of leaving full time education and a quadratic in age,
+the interaction with age leaving FTE remains significant at
+0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ`
+regressions.
+
+Selection effects are also different between men and women (Appendix
+Figure \@ref(fig:plot-sexes)). Differences are particularly large for
+educational attainment, height and MDD. Several PGS for mental illness
+and personality traits are more selected for (or less against) among
+women, including major depressive disorder (MDD), schizophrenia and
+neuroticism, while extraversion is more selected for among men. Waist
+circumference and waist-hip ratio are less selected for among women, and
+educational attainment is more selected against, though the difference
+is only significant for EA3.
+
+```{r calc-age-flb-cross}
+
+drake::loadd(res_age_flb_cross)
+
+cor_age_flb <- res_age_flb_cross %>% 
+      mutate(category = gsub(":.*", "", term)) %>% 
+      pivot_wider(score_name, names_from = category, values_from = estimate) %>%              {cor(.$`age_flb_cat10-22`, .$`age_flb_cat28-52`)}
+
+```
+
+We next focus on variables related to mate search and household
+formation. We split males and females by lifetime number of sexual
+partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
+For both sexes, selection effects are larger and more significant among
+those with more than 3 partners. Results are unchanged if we control for
+respondents' age (appendix Figure
+\@ref(fig:plot-n-partners-controlled)). Splitting respondents by whether
+they were living with a spouse or partner at the time of interview,
+effects are larger among those not living with a spouse or partner
+(Figure \@ref(fig:plot-n-partners)b). Lastly, we split female
+respondents by age at first live birth (AFLB).[^1] There is evidence for
+genetic effects on AFLB [@Barban_2016], and there is a close link
+between this variable and number of children born. Figure
+\@ref(fig:plot-n-partners)c shows effect sizes estimated separately for
+each tercile of AFLB. Several effects are strikingly different across
+terciles. ADHD and MDD are selected for amongst the youngest third of
+mothers, but selected against among the oldest two-thirds. Educational
+attainment is selected for among the oldest two-thirds of mothers, but
+is not significantly selected among the youngest third. Similarly,
+several PGS of body measurements are selected against only among older
+mothers. The correlation between effect sizes for the youngest and
+oldest terciles is `r cor_age_flb`. As before, recall that PGS may have
+two effects: changing the number of children within categories, and
+shifting respondents between categories.
+
+[^1]: This information is unavailable for men.
+
+```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth", fig.subcap = c("Number of partners", "Household type", "Age at first live birth"), fig.ncol = 1, out.height = c("35%", "27%", "27%"), fig.align = "center"}
+
+drake::loadd(res_partners)
+res_partners %<>%       
+      filter(grepl(":", term)) %>% 
+      mutate(`N partners` = ifelse(grepl("TRUE:", term), "<= 3", "> 3"))
+
+n_regs <- as.double(nrow(res_partners))
+res_partners %>% 
+      mutate(
+        score_name = pretty_names(score_name)
+      ) %>% 
+      ggplot(aes(
+        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        x      = estimate, 
+        fill   = `N partners`, 
+        colour = p.value < 0.05/{{n_regs}}
+      )) + 
+        my_vline +
+        pval_geom_point() +
+        facet_wrap(vars(sex), scales = "free_y") +
+        my_fill_scale(n = 2) +
+        pval_scale() +
+        my_theme(legend.position = "bottom") +
+        my_labs(x = "", y = "") + 
+        scale_y_reordered()
+
+drake::loadd(res_with_partner)
+res_with_partner %<>% filter(grepl(":", term))
+
+n_regs <- nrow(res_with_partner)
+hhtype_n <- n_in_regs("with_partner")
+
+res_with_partner %>% 
+      mutate(
+        Household = ifelse(grepl("TRUE", term), "With partner", 
+                      "Without partner"),
+        with_p    = as.character(grepl("TRUE", term)),
+        Household = sprintf("%s (N = %s)", Household, 
+                      prettyNum(hhtype_n[with_p], big.mark = ","))
+      ) %>% 
+      standard_ggplot(fill = Household, n_regs = n_regs, score_col = score_name)
+
+drake::loadd(res_age_flb_cross)
+
+# this counts each cross term as a separate test:
+n_regs <- as.double(nrow(res_age_flb_cross))
+
+res_age_flb_cross %>% 
+      mutate(
+        score_name = pretty_names(score_name),
+        score_name = fct_reorder(score_name, estimate, order_abs(1)),
+        term       = gsub("age_flb_cat(.*):.*", "\\1", term),
+        term       = add_n(term, "age_flb_cat")
+      ) %>% 
+      ggplot(aes(estimate, score_name, fill = term, 
+        colour = p.value < 0.05/{{n_regs}})) +
+        my_vline +
+        pval_geom_point() +
+        pval_scale() +
+        my_fill_scale(n = 3, direction = -1, 
+          guide = guide_legend(title = "Age at first live birth")) +
+        my_labs() +
+        my_theme()
+```
+
+```{r calc-corr-age-flb}
+
+drake::loadd(res_age_flb)
+drake::loadd(res_sex)
+
+res_sex_comparison <- res_sex %>% 
+      filter(sex == "Female") %>% 
+      mutate(score_name = term) %>% 
+      select(score_name, term, estimate:conf.high)
+
+res_age_flb %<>% filter(term != "age_flb")
+
+res_combined <- bind_rows(
+        raw      = res_sex_comparison, 
+        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
+        .id      = "reg.type"
+      ) %>% 
+      arrange(score_name)
+
+raw_flb_comparison <- res_combined %>% 
+      pivot_wider(
+        id_cols     = score_name, 
+        names_from  = reg.type, 
+        values_from = estimate
+      )
+
+cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
+opp_signed <- sum(
+        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
+      )
+
+```
+
+To investigate this further, we regress number of children on PGS
+*controlling* for AFLB (appendix Figure \@ref(fig:plot-age-flb)). In
+`r opp_signed` out of `r length(score_names)` cases, effects
+change sign when controls are added. The correlation between effect
+sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`.
+Selection effects appear to come through two opposing channels: an
+effect on AFLB, and an opposite-signed effect on number of children 
+controlling for AFLB.
+
+```{r calc-res-weighted}
+
+drake::loadd(res_wt_flb_weights)
+drake::loadd(res_wt_msoa_weights)
+drake::loadd(res_wt_age_qual_weights)
+drake::loadd(res_sibs_parent_weights)
+drake::loadd(res_all)
+drake::loadd(res_sex)
+
+res_wt_combined <- bind_rows(
+  None                     = res_all %>% 
+                               filter(reg.type == "raw", dep.var == "N children") %>% 
+                               select(-reg.type, -dep.var, -score_name),
+  Geographical             = res_wt_msoa_weights,
+  `Age/Qualification`      = res_wt_age_qual_weights,
+  `Age/Qual/AFLB`  = res_wt_flb_weights,
+  .id = "Weights"
+)
+
+res_wt_cb_wide <- res_wt_combined %>% 
+                    tidyr::pivot_wider(
+                      id_cols     = term,
+                      names_from  = Weights,
+                      values_from = estimate
+                    ) %>% 
+                    rowwise() %>% 
+                    mutate(
+                      consistent = all(c_across(-term) > 0) ||
+                        all(c_across(-term) < 0)
+                    ) %>% 
+                    ungroup() %>% 
+                    left_join(
+                      res_wt_combined %>% 
+                        filter(Weights == "None") %>% 
+                        select(term, p.value),
+                      by = "term"
+                    ) %>% 
+                    left_join(
+                      res_sex %>% 
+                        filter(sex == "Female") %>% 
+                        select(term, estimate_females = estimate),
+                      by = "term"
+                    )
+
+rel_sizes <- res_wt_cb_wide %>% 
+                filter(
+                  consistent, 
+                  p.value < 0.05/length(score_names)
+                ) %>% 
+                mutate(
+                  Geographical = Geographical/None,
+                  `Age/Qualification` = `Age/Qualification`/None,
+                  `Age/Qual/AFLB` = `Age/Qual/AFLB`/estimate_females
+                ) %>% 
+                select(-consistent, -None, -p.value, -estimate_females) %>% 
+                mutate(
+                  term = pretty_names(term)
+                )
+
+rel_sizes_summary <- rel_sizes %>% 
+                summarize(
+                  across(-term, ~ c(mean(.x), median(.x)))
+                ) %>% 
+                mutate(
+                  term = c("Mean", "Median")
+                )
+
+mean_rel <- c(rel_sizes_summary[1, 1:3])
+```
+
+Since UK Biobank subjects are not representative of the wider population, our
+results suggest that naïve estimates of natural selection are at risk of
+ascertainment bias. To correct for this, we weight participants using data from
+the 2011 Census and from the 2006 General Household Survey (GHS). We try three
+alternative weighting schemes: geographical weighting by census MSOA, age and
+presence/absence of a partner; weighting by age and highest educational
+qualification; and age, highest qualification, and age at first live birth, for
+women only. Mean effect sizes across all polygenic scores are increased by 
+`r mean_rel$Geographical` (geographical weighting), 
+`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
+`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
+further affected by weighting on other demographic variables. Since the Biobank
+sample seems to be ascertained in ways that reduce estimates of selection
+effects, we suspect that our weighted estimates remain conservative.
+
+The UK Biobank data contains information on respondents' number of
+siblings. Since respondents' polygenic scores are equal in expectation
+to the mean scores of their parents, we can use this to look at selection 
+effects in the parents' generation. We estimate equation (\ref{eq:regression})
+using *number of siblings* (including the respondent) as the dependent variable.
+
+The parents' generation has a further source of ascertainment bias: sampling
+parents of respondents overweights parents who have many children. For instance,
+parents of three children will have, on average, three times more children
+represented in UK Biobank than parents of one child. Parents of no children will
+by definition not be represented. To compensate, we reweight our preferred
+weightings (Age/Qualification) by the inverse of *number of siblings*. We then
+regress this on our polygenic scores. Selection effects are highly correlated across
+the two generations and most share the same sign (appendix Figure \@ref(fig:plot-res-sibs)). Effect sizes appear larger in the parents' generation. 
+We treat this result cautiously, because when we
+split respondents up by year of birth, we find few differences in effect
+sizes between early- and late-born respondents, for either generation. In other
+words, since estimated effect sizes change when we change the dependent
+variable, but do not change over time within either dependent variable,
+the explanation may be due to remaining ascertainment bias within our
+sample.
+
 ```{r NOTES-TODO, eval=FALSE}
@@
 
-* Happel et al. 1994
+* Happel et al. 1984
    - childbirth is chosen to minimize loss of earnings from interrupted work.
@@
     have the same effect on fertility - i.e. less skills = higher fertility
+
+- Baudin et al. 2015
+  * look at childlessness vs. fertility (extensive/intensive margin)
+  * some people are too poor to afford children
+  * 'a husband alleviates the time cost of having children'. But for some groups (see appendix)
+    there is a +ve income/fertility gradient
+  * fertility decreases with female educn due to subst effect
+  * baby boom and bust in US: a decline in childlessness caused by poverty, then an increase
+    caused by substn effect ('opportunity cost')
+  * fertility decreases with educn both for single mothers and married women
+    - slightly (25%) higher decrease per year of educn (Table 2) for single mothers
 
@@
 
-We use data from UK Biobank. We compute 33 polygenic scores, normalized
-to mean 0, variance 1. Figure \@ref(fig:plot-means-over-time) plots mean
-polygenic scores by 5-year birth intervals. Several scores show
-consistent increases or declines over this 30-year period, of the order
-of 5% of a standard deviation.
-
-```{r plot-means-over-time, fig.height = 5.5, fig.cap = "Mean polygenic scores by birth year in UK Biobank.
-Points are means for 5-year intervals.
-Lines are 95% confidence intervals.
-Green lines show a significant linear increase over time (p < 0.05/33).
-Red lines show a significant decrease."}
-
-drake::loadd(pgs_over_time)
-
-pgs_over_time %>% 
-      mutate(score_name = pretty_names(score_name)) %>% 
-      ggplot(aes(YOB, group = score_name, colour = Change)) + 
-      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
-        fatten = 1.25) +
-      facet_wrap(~score_name, ncol = 5) +
-      my_hline +
-      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
-      theme(
-        strip.text  = element_text(size = 6), 
-        axis.text.x = element_text(size = 7),
-        legend.position = "none"
-      ) +
-      labs(y = "PGS", x = "Birth year (5 year intervals)")
-
-```
-
-These changes could reflect either natural selection within the UK
-population, or ascertainment bias within the sample. Respondents are
-higher income and better educated than the UK population, and they may
-also differ on other unobserved characteristics [@10.1093/aje/kwx246].
-Since richer and educated people also live longer, this bias might also
-increase with age.
-
-We next regress respondents'polygenic scores on their number of children
-($y_i$):
-
-```{=tex}
-\begin{equation}
-y_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
-\label{eq:regression}
-\end{equation}
-```
-
-```{r calc-sex-ratio}
-female_ratio <- mean(famhist$sex == 0)
-```
-
-The "selection effect", $\beta$, measures the strength of natural
-selection within the sample. In fact, since polygenic scores are
-normalized to mean 0, $\beta$ is the expected polygenic score among
-children of the sample [@beauchamp2016genetic]. To learn more about the
-underlying mechanisms, we split the sample, starting with basic
-demographic variables including education, income and sex. These are all
-potential sources of ascertainment bias: as well as the ascertainment
-for income and education, mentioned above, the sample sex ratio skews
-`r female_ratio * 100`% female.
-
-Figure \@ref(fig:plot-income-educ-level) plots selection effects for
-each polygenic score, grouping respondents by age of completing
-full-time education, and by household income. Effects are larger and
-more significant for the lowest income category, and for the lowest
-education category. Note that the overall effect is not a simple average
-of the effect among the different subgroups, because polygenic scores
-may also shift respondents between the subgroups. For example, a high
-PGS for educational attainment may correlate with fewer children among
-early school leavers, but may also increase the age at which a
-respondent leaves school.
-
-```{r plot-income-educ-level, fig.cap = "Selection effects by education and income", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}
-
-drake::loadd(res_edu)
-n_regs <- as.double(nrow(res_edu))
-
-res_edu %>% 
-      mutate(
-        term = pretty_names(term), 
-        term = fct_reorder(term, estimate, order_abs(1)),
-        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
-        "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
-      ) %>% 
-      ggplot(aes(y = term, x = estimate, colour = p.value < 0.05/{{n_regs}}, 
-        fill = `Age left FTE`)) +
-        my_vline +
-        pval_geom_point() + 
-        pval_scale() +
-        my_fill_scale(n = 3, direction = -1) +
-        my_labs() +
-        my_theme()
-
-drake::loadd(res_income)
-
-n_regs <- as.double(nrow(res_income))
-res_income %>% 
-      mutate(
-        term = pretty_names(term), 
-        term = fct_reorder(term, estimate, order_abs(1)),
-        Income = factor(income_cat, 
-          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
-        ),
-        Income = add_n(Income, "income_cat")
-      ) %>% 
-      ggplot(aes(y = term, x = estimate, fill = Income, 
-        colour = p.value < 0.05/{{n_regs}})) +
-        my_vline +
-        pval_geom_point(alpha = 0.8) + 
-        my_fill_scale(n = 5, direction = -1) +
-        pval_scale() +
-        my_labs() +
-        my_theme()     
-```
-
-```{r calc-income-edu-controlled}
-drake::loadd(res_income_controlled, res_edu_controlled)
-n_regs_ic <- nrow(res_income_controlled)
-n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
-
-n_regs_educ <- nrow(res_edu_controlled)
-n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
-```
-
-These results could be driven by age, if older respondents are poorer
-and less educated, and also more subject to selection on polygenic
-scores. However, if we rerun the regressions, interacting the polygenic
-score with income category and also with a quadratic in age, the
-interaction with income remains significant at 0.05/`r n_regs_ic` for
-`r n_sig_ic` out of `r n_regs_ic` regressions. Similarly if we interact
-the PGS with age of leaving full time education and a quadratic in age,
-the interaction with age leaving FTE remains significant at
-0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ`
-regressions.
-
-Selection effects are also different between men and women (Appendix
-Figure \@ref(fig:plot-sexes)). Differences are particularly large for
-educational attainment, height and MDD. Several PGS for mental illness
-and personality traits are more selected for (or less against) among
-women, including major depressive disorder (MDD), schizophrenia and
-neuroticism, while extraversion is more selected for among men. Waist
-circumference and waist-hip ratio are less selected for among women, and
-educational attainment is more selected against, though the difference
-is only significant for EA3.
-
-```{r calc-age-flb-cross}
-
-drake::loadd(res_age_flb_cross)
-
-cor_age_flb <- res_age_flb_cross %>% 
-      mutate(category = gsub(":.*", "", term)) %>% 
-      pivot_wider(score_name, names_from = category, values_from = estimate) %>%              {cor(.$`age_flb_cat10-22`, .$`age_flb_cat28-52`)}
-
-```
-
-We next focus on variables related to mate search and household
-formation. We split males and females by lifetime number of sexual
-partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
-For both sexes, selection effects are larger and more significant among
-those with more than 3 partners. Results are unchanged if we control for
-respondents' age (appendix Figure
-\@ref(fig:plot-n-partners-controlled)). Splitting respondents by whether
-they were living with a spouse or partner at the time of interview,
-effects are larger among those not living with a spouse or partner
-(Figure \@ref(fig:plot-n-partners)b). Lastly, we split female
-respondents by age at first live birth (AFLB).[^1] There is evidence for
-genetic effects on AFLB [@Barban_2016], and there is a close link
-between this variable and number of children born. Figure
-\@ref(fig:plot-n-partners)c shows effect sizes estimated separately for
-each tercile of AFLB. Several effects are strikingly different across
-terciles. ADHD and MDD are selected for amongst the youngest third of
-mothers, but selected against among the oldest two-thirds. Educational
-attainment is selected for among the oldest two-thirds of mothers, but
-is not significantly selected among the youngest third. Similarly,
-several PGS of body measurements are selected against only among older
-mothers. The correlation between effect sizes for the youngest and
-oldest terciles is `r cor_age_flb`. As before, recall that PGS may have
-two effects: changing the number of children within categories, and
-shifting respondents between categories.
-
-[^1]: This information is unavailable for men.
-
-```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth", fig.subcap = c("Number of partners", "Household type", "Age at first live birth"), fig.ncol = 1, out.height = c("35%", "27%", "27%"), fig.align = "center"}
-
-drake::loadd(res_partners)
-res_partners %<>%       
-      filter(grepl(":", term)) %>% 
-      mutate(`N partners` = ifelse(grepl("TRUE:", term), "<= 3", "> 3"))
-
-n_regs <- as.double(nrow(res_partners))
-res_partners %>% 
-      mutate(
-        score_name = pretty_names(score_name)
-      ) %>% 
-      ggplot(aes(
-        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        x      = estimate, 
-        fill   = `N partners`, 
-        colour = p.value < 0.05/{{n_regs}}
-      )) + 
-        my_vline +
-        pval_geom_point() +
-        facet_wrap(vars(sex), scales = "free_y") +
-        my_fill_scale(n = 2) +
-        pval_scale() +
-        my_theme(legend.position = "bottom") +
-        my_labs(x = "", y = "") + 
-        scale_y_reordered()
-
-drake::loadd(res_with_partner)
-res_with_partner %<>% filter(grepl(":", term))
-
-n_regs <- nrow(res_with_partner)
-hhtype_n <- n_in_regs("with_partner")
-
-res_with_partner %>% 
-      mutate(
-        Household = ifelse(grepl("TRUE", term), "With partner", 
-                      "Without partner"),
-        with_p    = as.character(grepl("TRUE", term)),
-        Household = sprintf("%s (N = %s)", Household, 
-                      prettyNum(hhtype_n[with_p], big.mark = ","))
-      ) %>% 
-      standard_ggplot(fill = Household, n_regs = n_regs, score_col = score_name)
-
-drake::loadd(res_age_flb_cross)
-
-# this counts each cross term as a separate test:
-n_regs <- as.double(nrow(res_age_flb_cross))
-
-res_age_flb_cross %>% 
-      mutate(
-        score_name = pretty_names(score_name),
-        score_name = fct_reorder(score_name, estimate, order_abs(1)),
-        term       = gsub("age_flb_cat(.*):.*", "\\1", term),
-        term       = add_n(term, "age_flb_cat")
-      ) %>% 
-      ggplot(aes(estimate, score_name, fill = term, 
-        colour = p.value < 0.05/{{n_regs}})) +
-        my_vline +
-        pval_geom_point() +
-        pval_scale() +
-        my_fill_scale(n = 3, direction = -1, 
-          guide = guide_legend(title = "Age at first live birth")) +
-        my_labs() +
-        my_theme()
-```
-
-```{r calc-corr-age-flb}
-
-drake::loadd(res_age_flb)
-drake::loadd(res_sex)
-
-res_sex_comparison <- res_sex %>% 
-      filter(sex == "Female") %>% 
-      mutate(score_name = term) %>% 
-      select(score_name, term, estimate:conf.high)
-
-res_age_flb %<>% filter(term != "age_flb")
-
-res_combined <- bind_rows(
-        raw      = res_sex_comparison, 
-        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
-        .id      = "reg.type"
-      ) %>% 
-      arrange(score_name)
-
-raw_flb_comparison <- res_combined %>% 
-      pivot_wider(
-        id_cols     = score_name, 
-        names_from  = reg.type, 
-        values_from = estimate
-      )
-
-cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
-opp_signed <- sum(
-        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
-      )
-
-```
-
-To investigate this further, we regress number of children on PGS
-*controlling* for AFLB (appendix Figure \@ref(fig:plot-age-flb)). In
-`r opp_signed` out of `r length(score_names)` cases, effects
-change sign when controls are added. The correlation between effect
-sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`.
-Selection effects appear to come through two opposing channels: an
-effect on AFLB, and an opposite-signed effect on number of children 
-controlling for AFLB.
-
-```{r calc-res-weighted}
-
-drake::loadd(res_wt_flb_weights)
-drake::loadd(res_wt_msoa_weights)
-drake::loadd(res_wt_age_qual_weights)
-drake::loadd(res_sibs_parent_weights)
-drake::loadd(res_all)
-drake::loadd(res_sex)
-
-res_wt_combined <- bind_rows(
-  None                     = res_all %>% 
-                               filter(reg.type == "raw", dep.var == "N children") %>% 
-                               select(-reg.type, -dep.var, -score_name),
-  Geographical             = res_wt_msoa_weights,
-  `Age/Qualification`      = res_wt_age_qual_weights,
-  `Age/Qual/AFLB`  = res_wt_flb_weights,
-  .id = "Weights"
-)
-
-res_wt_cb_wide <- res_wt_combined %>% 
-                    tidyr::pivot_wider(
-                      id_cols     = term,
-                      names_from  = Weights,
-                      values_from = estimate
-                    ) %>% 
-                    rowwise() %>% 
-                    mutate(
-                      consistent = all(c_across(-term) > 0) ||
-                        all(c_across(-term) < 0)
-                    ) %>% 
-                    ungroup() %>% 
-                    left_join(
-                      res_wt_combined %>% 
-                        filter(Weights == "None") %>% 
-                        select(term, p.value),
-                      by = "term"
-                    ) %>% 
-                    left_join(
-                      res_sex %>% 
-                        filter(sex == "Female") %>% 
-                        select(term, estimate_females = estimate),
-                      by = "term"
-                    )
-
-rel_sizes <- res_wt_cb_wide %>% 
-                filter(
-                  consistent, 
-                  p.value < 0.05/length(score_names)
-                ) %>% 
-                mutate(
-                  Geographical = Geographical/None,
-                  `Age/Qualification` = `Age/Qualification`/None,
-                  `Age/Qual/AFLB` = `Age/Qual/AFLB`/estimate_females
-                ) %>% 
-                select(-consistent, -None, -p.value, -estimate_females) %>% 
-                mutate(
-                  term = pretty_names(term)
-                )
-
-rel_sizes_summary <- rel_sizes %>% 
-                summarize(
-                  across(-term, ~ c(mean(.x), median(.x)))
-                ) %>% 
-                mutate(
-                  term = c("Mean", "Median")
-                )
-
-mean_rel <- c(rel_sizes_summary[1, 1:3])
-```
-
-Since UK Biobank subjects are not representative of the wider population, our
-results suggest that naïve estimates of natural selection are at risk of
-ascertainment bias. To correct for this, we weight participants using data from
-the 2011 Census and from the 2006 General Household Survey (GHS). We try three
-alternative weighting schemes: geographical weighting by census MSOA, age and
-presence/absence of a partner; weighting by age and highest educational
-qualification; and age, highest qualification, and age at first live birth, for
-women only. Mean effect sizes across all polygenic scores are increased by 
-`r mean_rel$Geographical` (geographical weighting), 
-`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
-`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
-further affected by weighting on other demographic variables. Since the Biobank
-sample seems to be ascertained in ways that reduce estimates of selection
-effects, we suspect that our weighted estimates remain conservative.
-
-The UK Biobank data contains information on respondents' number of
-siblings. Since respondents' polygenic scores are equal in expectation
-to the mean scores of their parents, we can use this to look at selection 
-effects in the parents' generation. We estimate equation (\ref{eq:regression})
-using *number of siblings* (including the respondent) as the dependent variable.
-
-The parents' generation has a further source of ascertainment bias: sampling
-parents of respondents overweights parents who have many children. For instance,
-parents of three children will have, on average, three times more children
-represented in UK Biobank than parents of one child. Parents of no children will
-by definition not be represented. To compensate, we reweight our preferred
-weightings (Age/Qualification) by the inverse of *number of siblings*. We then
-regress this on our polygenic scores. Selection effects are highly correlated across
-the two generations and most share the same sign (appendix Figure \@ref(fig:plot-res-sibs)). Effect sizes appear larger in the parents' generation. 
-We treat this result cautiously, because when we
-split respondents up by year of birth, we find few differences in effect
-sizes between early- and late-born respondents, for either generation. In other
-words, since estimated effect sizes change when we change the dependent
-variable, but do not change over time within either dependent variable,
-the explanation may be due to remaining ascertainment bias within our
-sample.
-
 Why do selection effects have *opposite* signs across older and younger mothers
 at first live birth, and opposite signs when AFLB is controlled for? A possible
-answer is given by the economic theory of fertility [@becker1960economic; @willis1973new].
+answer is given by the economic theory of fertility 
+[@becker1960economic; @willis1973new; @becker1976child].
 According to this, increases in potential earned income affect fertility via two
@@
 have more children. The substitution effect would lead higher earners to have
-fewer children. It is often assumed that the substitution effect will dominate
-for lone parents, or those in unstable relationships, since they have less
-opportunity to share childcare responsibilities, while the income effect will
-dominate for couples who are able to reap the gains from specialization.
-
-If so, then genetic characteristics which affect one's earnings
-potential in the labour market may lead to opposing effects on
-fertility. Genetic variants which improve one's earnings may increase
-fertility among couples, but decrease fertility among single parents.
-They may also decrease the likelihood of becoming a young mother, if
-this is correlated with single motherhood. Lastly, if lifetime number of
-sexual partners predicts single parent versus two-parent status, the
-theory could also explain the different effect sizes across people with
-more and fewer sexual partners.
-
-This explanation predicts that the strength of polygenic scores' effects
-on fertility will correlate with the strength of their relationship with
-earnings. To test this, we calculate the correlation of each PGS with
-estimated earnings from respondent's first job[^earnings-est], and with education
-level, a phenotype which predicts earnings. We then relate these
-measurements to effect sizes on number of children. There is a clear
-negative relationship between a score's effect on fertility, and its
-correlation with earnings or education. If we control for age at first live
-birth, the relationship changes sign and becomes positive  (Figure
-\@ref(fig:plot-cors-earnings-educ)).
+fewer children. If so, then genetic variants which affect earnings
+potential in the labour market may cause opposing effects on
+fertility. The income effect will cause natural selection in favour of 
+earnings-increasing variants. The substitution effect will do the reverse.
+
+This theory predicts that genetic variants will be selected for (or against) in
+proportion to their effect on earnings. Figure
+\@ref(fig:plot-cors-earnings-educ) plots selection effects on each PGS against
+that PGS's correlation with earnings in a respondent's first job[^earnings-est],
+and against its correlation with educational attainment, a predictor of lifetime
+earnings. The raw relationships (left column) are strongly negative. If we plot
+selection effects controlling for age at first live birth (right column), the
+effect reverses and becomes positive. This is evidence that the labour market
+plays an important role in natural selection, with substitution effects
+dominating overall.
 
@@
 
-```{r plot-cors-earnings-educ, fig.cap="Effect sizes on number of children, with and without controls for age at first live birth, by correlations with earnings and education.
+```{r plot-cors-earnings-educ, fig.cap="Selection effects, with and without controls for age at first live birth, by correlations with earnings and educational attainment.
 Each point represents one PGS."}
@@
 # TODO: use age_left_edu when this includes college
-cors_educ <-   cor(famhist[score_names], famhist$edu_qual, use = "pairwise")
+# TODO:
+cors_educ <-   cor(famhist[score_names], famhist$age_fulltime_edu, use = "pairwise")
 
@@
         my_hline +
-        labs(x = "Effect on number of children", y = "Correlation") +
+        labs(x = "Selection effect", y = "Correlation") +
         scale_colour_brewer(type = "qual", palette = 2) +
@@
 ```
+
+Suppose that people are aware of their endowments of human capital, and 
+that education and innate human capital are complements. They choose how long
+to stay in education. After leaving education they enter the labour market and
+form families. If so, then people with less human capital will leave education 
+earlier and have their first child earlier. Whilst in the labour market, higher
+earners will have more children (i.e. income effects dominate). These 
+assumptions will lead to the pattern we observe where earnings-increasing
+variants lead to later first births (and fewer children overall), but more
+children for any given AFLB.
+
+The theory can also explain why natural selection is weaker among higher-income
+parents. @becker1976child show that if parents care about child quality as well
+as quantity, under certain assumptions the income effect will be stronger at
+higher income levels; this can lead to a U-shaped relationship between income
+and fertility.
+
+Lastly, it is often assumed that the substitution effect will dominate
+for lone parents, or those in unstable relationships, since they have less
+opportunity to share childcare responsibilities, while the income effect will
+dominate for couples who are able to reap the gains from specialization. If so,
+then earnings-increasing genetic variants may decrease fertility more among single 
+parents than among couples, as in our data. The same idea could explain our
+results for lifetime number of sexual partners, if this is associated with
+relationship stability.
+
+Empirical work shows similar relationships between *earnings* or *education* and
+fertility, to the relationships shown here between *genetics* and fertility.
+Higher income is associated with lower fertility [@becker1960economic;
+@jones2006economic].
+@cohen2013financial find that income decreases fertility at low income levels
+but increases it at higher income levels. In a US sample, fertility decreases
+with education faster among single mothers than married mothers 
+[@baudin2015fertility].
+
+```{r calc-res-earn-educ}
+
+drake::loadd(res_ee_control)
+drake::loadd(res_all)
+
+res_ee_control %<>% filter(term == score_name)
+n_05 <- sum(res_ee_control$p.value < 0.05)
+n_05_33 <- sum(res_ee_control$p.value < 0.05/33)
+
+```
+
+The economic theory predicts that natural selection on polygenic scores is mediatet
+by labour market earnings. As a final test of this,
+we re-estimate equation (\@ref(eq:regression)) for each PGS, 
+controlling for earnings in first job and education levels. Effect sizes are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only `r n_05_33` 
+out of 33 PGS are significant at $p = 0.05/33$.
 
@@
 have no special authority to speak on how society should react to
-them. This can only be resolved by open debate, in which scientists take part as
+them. This is a matter for open debate, in which scientists take part as
 equals among many. We therefore choose not to discuss policy implications here.
@@
 
+\FloatBarrier
+## Controlling for earnings and education
+
+```{r plot-res-earn-educ, fig.cap = "Selection effects controlling for earnings in first job (estimated by median earnings from ASHE 2007 for the SOC2000 job code) and education (left education before 16, 16-18, or after 18).
+Raw effects are shown for comparison.
+Lines are 95% confidence intervals uncorrected for multiple testing."}
+
+drake::loadd(res_ee_control)
+drake::loadd(res_all)
+
+res_ee_control %<>% filter(term == score_name)
+
+res_all %<>% 
+  filter(term == score_name, reg.type == "raw", dep.var == "N children") %>% 
+  select(-dep.var, -reg.type)
+
+res_compare <- bind_rows(
+                 Raw = res_all, 
+                 Controlled = res_ee_control, 
+                 .id = "Regression"
+               )
+
+standard_ggplot(res_compare, fill_col = Regression, n_regs = 33) +
+  geom_linerange(
+    aes(
+      xmin = estimate - 2 * std.error, 
+      xmax = estimate + 2 * std.error
+    ), 
+    alpha = .5
+  )
+
+```
+
+\FloatBarrier
 ## Genetic correlations with EA3
@@
 
+\FloatBarrier
 ## Number of children

### 2020-11-13 _drake.R

@@ -352,6 +352,41 @@ plan <- drake_plan(
     res
   },
   
+  res_with_partner_narrow = {
+    mf_pairs_tmp <- mf_pairs %>% select(ID.m, ID.f, n_children.m, n_children.f)
+    famhist_tmp <- famhist %>% 
+                      filter(sex == 1) %>% 
+                      select(f.eid, with_partner, n_children) %>% 
+                      left_join(mf_pairs_tmp, by = c("f.eid" = "ID.m"))
+    famhist_tmp2 <- famhist %>% 
+                      filter(sex == 0) %>% 
+                      select(f.eid, with_partner, n_children) %>% 
+                      left_join(mf_pairs_tmp, by = c("f.eid" = "ID.f"))
+    famhist_tmp  %<>% bind_rows(famhist_tmp2)
+    
+    famhist_tmp %<>% 
+          filter(n_children.f == n_children.m, with_partner) %>% 
+          group_by(f.eid) %>% 
+          filter(n() == 1)
+    
+    famhist_tmp$with_partner_narrow <- TRUE
+    famhist_tmp %<>% select(f.eid, with_partner_narrow)
+    
+    famhist %<>% left_join(famhist_tmp)
+    famhist %<>% mutate(with_partner_narrow = ! is.na(with_partner_narrow))
+    
+    res <- map_dfr(score_names,
+            ~run_regs_fml(
+              "n_children ~ with_partner_narrow + {score_name}:with_partner_narrow",
+              score_name = .x,
+              famhist    = famhist
+            ),
+            .id = "score_name"
+          )
+    res %<>% filter(term != "(Intercept)")
+    res
+  },
+  
   res_edu = {
     subsets <- rlang::exprs(
             age_fte_cat == "< 16",
@@ -447,6 +482,15 @@ plan <- drake_plan(
           mutate(sex = ifelse(subset == "sex == 1", "Male", "Female")) %>%
           select(-row_number, -subset) %>% 
           filter(! grepl("YOB", term))
+  },
+  
+  res_ee_control = {
+    res_ee_control <- map_dfr(score_names, 
+                              ~ run_regs_fml(
+                                fml = "n_children ~ {score_name} + age_fte_cat + first_job_pay",
+                                score_name = .x,
+                                famhist    = famhist
+                              ), .id = "score_name")
   },
   
   res_together = {

### 2020-11-13 why-negative-selection.Rmd
Add discussion of changes between generations.

@@
 
-Geneticists have long hypothesized that natural selection is taking
-place in modern human populations. Recent work confirms this using
-genome-wide analysis [@Barban_2016; @beauchamp2016genetic;
-@kong2017selection; @sanjak2018evidence]. In particular, genetic
-variants associated with higher educational attainment are being
-selected against, although effect sizes appear small.
-
-As yet we know little about the *mechanisms* behind natural selection in
-contemporary populations. This paper uses data from UK Biobank to learn
-more. We test for natural selection on 33 different polygenic scores. We
-extend the analysis over two generations, using data on respondents'
-number of siblings as well as their number of children to reconstruct
-two generations of family size. This is interesting because consistent
-natural selection over multiple generations could lead to substantive
-effects in the long run. Most importantly, we examine the strength of
-natural selection in different subgroups of the population, in order to
-uncover patterns that can help illuminate the mechanisms involved.
-
-We use data from UK Biobank. We compute 33 polygenic scores, normalized
-to mean 0, variance 1. Figure \@ref(fig:plot-means-over-time) plots mean
-polygenic scores by 5-year birth intervals. Several scores show
-consistent increases or declines over this 30-year period, of the order
-of 5% of a standard deviation.
-
-```{r plot-means-over-time, fig.height = 5.5, fig.cap = "Mean polygenic scores by birth year in UK Biobank.
-Points are means for 5-year intervals.
-Lines are 95% confidence intervals.
-Green lines show a significant linear increase over time (p < 0.05/33).
-Red lines show a significant decrease."}
-
-drake::loadd(pgs_over_time)
-
-pgs_over_time %>% 
-      mutate(score_name = pretty_names(score_name)) %>% 
-      ggplot(aes(YOB, group = score_name, colour = Change)) + 
-      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
-        fatten = 1.25) +
-      facet_wrap(~score_name, ncol = 5) +
-      my_hline +
-      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
-      theme(
-        strip.text  = element_text(size = 6), 
-        axis.text.x = element_text(size = 7),
-        legend.position = "none"
-      ) +
-      labs(y = "PGS", x = "Birth year (5 year intervals)")
-
-```
-
-These changes could reflect either natural selection within the UK
-population, or ascertainment bias within the sample. Respondents are
-higher income and better educated than the UK population, and they may
-also differ on other unobserved characteristics [@10.1093/aje/kwx246].
-Since richer and educated people also live longer, this bias might also
-increase with age.
-
-We next regress respondents'polygenic scores on their number of children
-($y_i$):
-
-```{=tex}
-\begin{equation}
-y_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
-\label{eq:regression}
-\end{equation}
-```
-
-```{r calc-sex-ratio}
-female_ratio <- mean(famhist$sex == 0)
-```
-
-The "selection effect", $\beta$, measures the strength of natural
-selection within the sample. In fact, since polygenic scores are
-normalized to mean 0, $\beta$ is the expected polygenic score among
-children of the sample [@beauchamp2016genetic]. To learn more about the
-underlying mechanisms, we split the sample, starting with basic
-demographic variables including education, income and sex. These are all
-potential sources of ascertainment bias: as well as the ascertainment
-for income and education, mentioned above, the sample sex ratio skews
-`r female_ratio * 100`% female.
-
-Figure \@ref(fig:plot-income-educ-level) plots selection effects for
-each polygenic score, grouping respondents by age of completing
-full-time education, and by household income. Effects are larger and
-more significant for the lowest income category, and for the lowest
-education category. Note that the overall effect is not a simple average
-of the effect among the different subgroups, because polygenic scores
-may also shift respondents between the subgroups. For example, a high
-PGS for educational attainment may correlate with fewer children among
-early school leavers, but may also increase the age at which a
-respondent leaves school.
-
-```{r plot-income-educ-level, fig.cap = "Selection effects by education and income", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}
-
-drake::loadd(res_edu)
-n_regs <- as.double(nrow(res_edu))
-
-res_edu %>% 
-      mutate(
-        term = pretty_names(term), 
-        term = fct_reorder(term, estimate, order_abs(1)),
-        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
-        "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
-      ) %>% 
-      ggplot(aes(y = term, x = estimate, colour = p.value < 0.05/{{n_regs}}, 
-        fill = `Age left FTE`)) +
-        my_vline +
-        pval_geom_point() + 
-        pval_scale() +
-        my_fill_scale(n = 3, direction = -1) +
-        my_labs() +
-        my_theme()
-
-drake::loadd(res_income)
-
-n_regs <- as.double(nrow(res_income))
-res_income %>% 
-      mutate(
-        term = pretty_names(term), 
-        term = fct_reorder(term, estimate, order_abs(1)),
-        Income = factor(income_cat, 
-          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
-        ),
-        Income = add_n(Income, "income_cat")
-      ) %>% 
-      ggplot(aes(y = term, x = estimate, fill = Income, 
-        colour = p.value < 0.05/{{n_regs}})) +
-        my_vline +
-        pval_geom_point(alpha = 0.8) + 
-        my_fill_scale(n = 5, direction = -1) +
-        pval_scale() +
-        my_labs() +
-        my_theme()     
-```
-
-```{r calc-income-edu-controlled}
-drake::loadd(res_income_controlled, res_edu_controlled)
-n_regs_ic <- nrow(res_income_controlled)
-n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
-
-n_regs_educ <- nrow(res_edu_controlled)
-n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
-```
-
-These results could be driven by age, if older respondents are poorer
-and less educated, and also more subject to selection on polygenic
-scores. However, if we rerun the regressions, interacting the polygenic
-score with income category and also with a quadratic in age, the
-interaction with income remains significant at 0.05/`r n_regs_ic` for
-`r n_sig_ic` out of `r n_regs_ic` regressions. Similarly if we interact
-the PGS with age of leaving full time education and a quadratic in age,
-the interaction with age leaving FTE remains significant at
-0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ`
-regressions.
-
-Selection effects are also different between men and women (Appendix
-Figure \@ref(fig:plot-sexes)). Differences are particularly large for
-educational attainment, height and MDD. Several PGS for mental illness
-and personality traits are more selected for (or less against) among
-women, including major depressive disorder (MDD), schizophrenia and
-neuroticism, while extraversion is more selected for among men. Waist
-circumference and waist-hip ratio are less selected for among women, and
-educational attainment is more selected against, though the difference
-is only significant for EA3.
-
-```{r calc-age-flb-cross}
-
-drake::loadd(res_age_flb_cross)
-
-cor_age_flb <- res_age_flb_cross %>% 
-      mutate(category = gsub(":.*", "", term)) %>% 
-      pivot_wider(score_name, names_from = category, values_from = estimate) %>%              {cor(.$`age_flb_cat10-22`, .$`age_flb_cat28-52`)}
-
-```
-
-We next focus on variables related to mate search and household
-formation. We split males and females by lifetime number of sexual
-partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
-For both sexes, selection effects are larger and more significant among
-those with more than 3 partners. Results are unchanged if we control for
-respondents' age (appendix Figure
-\@ref(fig:plot-n-partners-controlled)). Splitting respondents by whether
-they were living with a spouse or partner at the time of interview,
-effects are larger among those not living with a spouse or partner
-(Figure \@ref(fig:plot-n-partners)b). Lastly, we split female
-respondents by age at first live birth (AFLB).[^1] There is evidence for
-genetic effects on AFLB [@Barban_2016], and there is a close link
-between this variable and number of children born. Figure
-\@ref(fig:plot-n-partners)c shows effect sizes estimated separately for
-each tercile of AFLB. Several effects are strikingly different across
-terciles. ADHD and MDD are selected for amongst the youngest third of
-mothers, but selected against among the oldest two-thirds. Educational
-attainment is selected for among the oldest two-thirds of mothers, but
-is not significantly selected among the youngest third. Similarly,
-several PGS of body measurements are selected against only among older
-mothers. The correlation between effect sizes for the youngest and
-oldest terciles is `r cor_age_flb`. As before, recall that PGS may have
-two effects: changing the number of children within categories, and
-shifting respondents between categories.
-
-[^1]: This information is unavailable for men.
-
-```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth", fig.subcap = c("Number of partners", "Household type", "Age at first live birth"), fig.ncol = 1, out.height = c("35%", "27%", "27%"), fig.align = "center"}
-
-drake::loadd(res_partners)
-res_partners %<>%       
-      filter(grepl(":", term)) %>% 
-      mutate(`N partners` = ifelse(grepl("TRUE:", term), "<= 3", "> 3"))
-
-n_regs <- as.double(nrow(res_partners))
-res_partners %>% 
-      mutate(
-        score_name = pretty_names(score_name)
-      ) %>% 
-      ggplot(aes(
-        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        x      = estimate, 
-        fill   = `N partners`, 
-        colour = p.value < 0.05/{{n_regs}}
-      )) + 
-        my_vline +
-        pval_geom_point() +
-        facet_wrap(vars(sex), scales = "free_y") +
-        my_fill_scale(n = 2) +
-        pval_scale() +
-        my_theme(legend.position = "bottom") +
-        my_labs(x = "", y = "") + 
-        scale_y_reordered()
-
-drake::loadd(res_with_partner)
-res_with_partner %<>% filter(grepl(":", term))
-
-n_regs <- nrow(res_with_partner)
-hhtype_n <- n_in_regs("with_partner")
-
-res_with_partner %>% 
-      mutate(
-        Household = ifelse(grepl("TRUE", term), "With partner", 
-                      "Without partner"),
-        with_p    = as.character(grepl("TRUE", term)),
-        Household = sprintf("%s (N = %s)", Household, 
-                      prettyNum(hhtype_n[with_p], big.mark = ","))
-      ) %>% 
-      standard_ggplot(fill = Household, n_regs = n_regs, score_col = score_name)
-
-drake::loadd(res_age_flb_cross)
-
-# this counts each cross term as a separate test:
-n_regs <- as.double(nrow(res_age_flb_cross))
-
-res_age_flb_cross %>% 
-      mutate(
-        score_name = pretty_names(score_name),
-        score_name = fct_reorder(score_name, estimate, order_abs(1)),
-        term       = gsub("age_flb_cat(.*):.*", "\\1", term),
-        term       = add_n(term, "age_flb_cat")
-      ) %>% 
-      ggplot(aes(estimate, score_name, fill = term, 
-        colour = p.value < 0.05/{{n_regs}})) +
-        my_vline +
-        pval_geom_point() +
-        pval_scale() +
-        my_fill_scale(n = 3, direction = -1, 
-          guide = guide_legend(title = "Age at first live birth")) +
-        my_labs() +
-        my_theme()
-```
-
-```{r calc-corr-age-flb}
-
-drake::loadd(res_age_flb)
-drake::loadd(res_sex)
-
-res_sex_comparison <- res_sex %>% 
-      filter(sex == "Female") %>% 
-      mutate(score_name = term) %>% 
-      select(score_name, term, estimate:conf.high)
-
-res_age_flb %<>% filter(term != "age_flb")
-
-res_combined <- bind_rows(
-        raw      = res_sex_comparison, 
-        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
-        .id      = "reg.type"
-      ) %>% 
-      arrange(score_name)
-
-raw_flb_comparison <- res_combined %>% 
-      pivot_wider(
-        id_cols     = score_name, 
-        names_from  = reg.type, 
-        values_from = estimate
-      )
-
-cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
-opp_signed <- sum(
-        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
-      )
-
-```
-
-To investigate this further, we regress number of children on PGS
-*controlling* for AFLB (appendix Figure \@ref(fig:plot-age-flb)). In
-`r opp_signed` out of `r length(score_names)` cases, effects
-change sign when controls are added. The correlation between effect
-sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`.
-Selection effects appear to come through two opposing channels: an
-effect on AFLB, and an opposite-signed effect on number of children 
-controlling for AFLB.
-
-```{r calc-res-weighted}
-
-drake::loadd(res_wt_flb_weights)
-drake::loadd(res_wt_msoa_weights)
-drake::loadd(res_wt_age_qual_weights)
-drake::loadd(res_sibs_parent_weights)
-drake::loadd(res_all)
-drake::loadd(res_sex)
-
-res_wt_combined <- bind_rows(
-  None                     = res_all %>% 
-                               filter(reg.type == "raw", dep.var == "N children") %>% 
-                               select(-reg.type, -dep.var, -score_name),
-  Geographical             = res_wt_msoa_weights,
-  `Age/Qualification`      = res_wt_age_qual_weights,
-  `Age/Qual/AFLB`  = res_wt_flb_weights,
-  .id = "Weights"
-)
-
-res_wt_cb_wide <- res_wt_combined %>% 
-                    tidyr::pivot_wider(
-                      id_cols     = term,
-                      names_from  = Weights,
-                      values_from = estimate
-                    ) %>% 
-                    rowwise() %>% 
-                    mutate(
-                      consistent = all(c_across(-term) > 0) ||
-                        all(c_across(-term) < 0)
-                    ) %>% 
-                    ungroup() %>% 
-                    left_join(
-                      res_wt_combined %>% 
-                        filter(Weights == "None") %>% 
-                        select(term, p.value),
-                      by = "term"
-                    ) %>% 
-                    left_join(
-                      res_sex %>% 
-                        filter(sex == "Female") %>% 
-                        select(term, estimate_females = estimate),
-                      by = "term"
-                    )
-
-rel_sizes <- res_wt_cb_wide %>% 
-                filter(
-                  consistent, 
-                  p.value < 0.05/length(score_names)
-                ) %>% 
-                mutate(
-                  Geographical = Geographical/None,
-                  `Age/Qualification` = `Age/Qualification`/None,
-                  `Age/Qual/AFLB` = `Age/Qual/AFLB`/estimate_females
-                ) %>% 
-                select(-consistent, -None, -p.value, -estimate_females) %>% 
-                mutate(
-                  term = pretty_names(term)
-                )
-
-rel_sizes_summary <- rel_sizes %>% 
-                summarize(
-                  across(-term, ~ c(mean(.x), median(.x)))
-                ) %>% 
-                mutate(
-                  term = c("Mean", "Median")
-                )
-
-mean_rel <- c(rel_sizes_summary[1, 1:3])
-```
-
-Since UK Biobank subjects are not representative of the wider population, our
-results suggest that naïve estimates of natural selection are at risk of
-ascertainment bias. To correct for this, we weight participants using data from
-the 2011 Census and from the 2006 General Household Survey (GHS). We try three
-alternative weighting schemes: geographical weighting by census MSOA, age and
-presence/absence of a partner; weighting by age and highest educational
-qualification; and age, highest qualification, and age at first live birth, for
-women only. Mean effect sizes across all polygenic scores are increased by 
-`r mean_rel$Geographical` (geographical weighting), 
-`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
-`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
-further affected by weighting on other demographic variables. Since the Biobank
-sample seems to be ascertained in ways that reduce estimates of selection
-effects, we suspect that our weighted estimates remain conservative.
-
-The UK Biobank data contains information on respondents' number of
-siblings. Since respondents' polygenic scores are equal in expectation
-to the mean scores of their parents, we can use this to look at selection 
-effects in the parents' generation. We estimate equation (\ref{eq:regression})
-using *number of siblings* (including the respondent) as the dependent variable.
-
-The parents' generation has a further source of ascertainment bias: sampling
-parents of respondents overweights parents who have many children. For instance,
-parents of three children will have, on average, three times more children
-represented in UK Biobank than parents of one child. Parents of no children will
-by definition not be represented. To compensate, we reweight our preferred
-weightings (Age/Qualification) by the inverse of *number of siblings*. We then
-regress this on our polygenic scores. Selection effects are highly correlated across
-the two generations and most share the same sign (appendix Figure \@ref(fig:plot-res-sibs)). Effect sizes appear larger in the parents' generation. 
-We treat this result cautiously, because when we
-split respondents up by year of birth, we find few differences in effect
-sizes between early- and late-born respondents, for either generation. In other
-words, since estimated effect sizes change when we change the dependent
-variable, but do not change over time within either dependent variable,
-the explanation may be due to remaining ascertainment bias within our
-sample.
-
 ```{r NOTES-TODO, eval=FALSE}
@@
 
-Why do selection effects have *opposite* signs across older and younger mothers
-at first live birth, and opposite signs when AFLB is controlled for? A possible
-answer is given by the economic theory of fertility 
-[@becker1960economic; @willis1973new; @becker1976child].
-According to this, increases in potential earned income affect fertility via two
-opposing channels. There is an "income effect" by which children become more
-affordable, like any other good. There is also a "substitution effect": since
+Geneticists have long hypothesized that natural selection is taking
+place in modern human populations. Recent work confirms this using
+genome-wide analysis [@Barban_2016; @beauchamp2016genetic;
+@kong2017selection; @sanjak2018evidence]. In particular, genetic
+variants associated with higher educational attainment are being
+selected against, although effect sizes appear small.
+
+As yet we know little about the *mechanisms* behind natural selection in
+contemporary populations. This paper uses data from UK Biobank to learn
+more. We test for natural selection on 33 different polygenic scores. We
+extend the analysis over two generations, using data on respondents'
+number of siblings as well as their number of children to reconstruct
+two generations of family size. This is interesting because consistent
+natural selection over multiple generations could lead to substantive
+effects in the long run. Most importantly, we examine the strength of
+natural selection in different subgroups of the population, in order to
+uncover patterns that can help illuminate the mechanisms involved.
+
+We use data from UK Biobank. We compute 33 polygenic scores, normalized
+to mean 0, variance 1. Figure \@ref(fig:plot-means-over-time) plots mean
+polygenic scores by 5-year birth intervals. Several scores show
+consistent increases or declines over this 30-year period, of the order
+of 5% of a standard deviation.
+
+```{r plot-means-over-time, fig.height = 5.5, fig.cap = "Mean polygenic scores by birth year in UK Biobank.
+Points are means for 5-year intervals.
+Lines are 95% confidence intervals.
+Green lines show a significant linear increase over time (p < 0.05/33).
+Red lines show a significant decrease."}
+
+drake::loadd(pgs_over_time)
+
+pgs_over_time %>% 
+      mutate(score_name = pretty_names(score_name)) %>% 
+      ggplot(aes(YOB, group = score_name, colour = Change)) + 
+      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
+        fatten = 1.25) +
+      facet_wrap(~score_name, ncol = 5) +
+      my_hline +
+      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
+      theme(
+        strip.text  = element_text(size = 6), 
+        axis.text.x = element_text(size = 7),
+        legend.position = "none"
+      ) +
+      labs(y = "PGS", x = "Birth year (5 year intervals)")
+
+```
+
+These changes could reflect either natural selection within the UK
+population, or ascertainment bias within the sample. Respondents are
+higher income and better educated than the UK population, and they may
+also differ on other unobserved characteristics [@10.1093/aje/kwx246].
+Since richer and educated people also live longer, this bias might also
+increase with age.
+
+We next regress respondents'polygenic scores on their number of children
+($y_i$):
+
+```{=tex}
+\begin{equation}
+y_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
+\label{eq:regression}
+\end{equation}
+```
+
+```{r calc-sex-ratio}
+female_ratio <- mean(famhist$sex == 0)
+```
+
+The "selection effect", $\beta$, measures the strength of natural
+selection within the sample. In fact, since polygenic scores are
+normalized to mean 0, $\beta$ is the expected polygenic score among
+children of the sample [@beauchamp2016genetic]. To learn more about the
+underlying mechanisms, we split the sample, starting with basic
+demographic variables including education, income and sex. These are all
+potential sources of ascertainment bias: as well as the ascertainment
+for income and education, mentioned above, the sample sex ratio skews
+`r female_ratio * 100`% female.
+
+Figure \@ref(fig:plot-income-educ-level) plots selection effects for
+each polygenic score, grouping respondents by age of completing
+full-time education, and by household income. Effects are larger and
+more significant for the lowest income category, and for the lowest
+education category. Note that the overall effect is not a simple average
+of the effect among the different subgroups, because polygenic scores
+may also shift respondents between the subgroups. For example, a high
+PGS for educational attainment may correlate with fewer children among
+early school leavers, but may also increase the age at which a
+respondent leaves school.
+
+```{r plot-income-educ-level, fig.cap = "Selection effects by education and income", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}
+
+drake::loadd(res_edu)
+n_regs <- as.double(nrow(res_edu))
+
+res_edu %>% 
+      mutate(
+        term = pretty_names(term), 
+        term = fct_reorder(term, estimate, order_abs(1)),
+        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
+        "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
+      ) %>% 
+      ggplot(aes(y = term, x = estimate, colour = p.value < 0.05/{{n_regs}}, 
+        fill = `Age left FTE`)) +
+        my_vline +
+        pval_geom_point() + 
+        pval_scale() +
+        my_fill_scale(n = 3, direction = -1) +
+        my_labs() +
+        my_theme()
+
+drake::loadd(res_income)
+
+n_regs <- as.double(nrow(res_income))
+res_income %>% 
+      mutate(
+        term = pretty_names(term), 
+        term = fct_reorder(term, estimate, order_abs(1)),
+        Income = factor(income_cat, 
+          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
+        ),
+        Income = add_n(Income, "income_cat")
+      ) %>% 
+      ggplot(aes(y = term, x = estimate, fill = Income, 
+        colour = p.value < 0.05/{{n_regs}})) +
+        my_vline +
+        pval_geom_point(alpha = 0.8) + 
+        my_fill_scale(n = 5, direction = -1) +
+        pval_scale() +
+        my_labs() +
+        my_theme()     
+```
+
+```{r calc-income-edu-controlled}
+drake::loadd(res_income_controlled, res_edu_controlled)
+n_regs_ic <- nrow(res_income_controlled)
+n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
+
+n_regs_educ <- nrow(res_edu_controlled)
+n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
+```
+
+These results could be driven by age, if older respondents are poorer
+and less educated, and also more subject to selection on polygenic
+scores. However, if we rerun the regressions, interacting the polygenic
+score with income category and also with a quadratic in age, the
+interaction with income remains significant at 0.05/`r n_regs_ic` for
+`r n_sig_ic` out of `r n_regs_ic` regressions. Similarly if we interact
+the PGS with age of leaving full time education and a quadratic in age,
+the interaction with age leaving FTE remains significant at
+0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ`
+regressions.
+
+Selection effects are also different between men and women (Appendix
+Figure \@ref(fig:plot-sexes)). Differences are particularly large for
+educational attainment, height and MDD. Several PGS for mental illness
+and personality traits are more selected for (or less against) among
+women, including major depressive disorder (MDD), schizophrenia and
+neuroticism, while extraversion is more selected for among men. Waist
+circumference and waist-hip ratio are less selected for among women, and
+educational attainment is more selected against, though the difference
+is only significant for EA3.
+
+```{r calc-age-flb-cross}
+
+drake::loadd(res_age_flb_cross)
+
+cor_age_flb <- res_age_flb_cross %>% 
+      mutate(category = gsub(":.*", "", term)) %>% 
+      pivot_wider(score_name, names_from = category, values_from = estimate) %>%              {cor(.$`age_flb_cat10-22`, .$`age_flb_cat28-52`)}
+
+```
+
+We next focus on variables related to mate search and household
+formation. We split males and females by lifetime number of sexual
+partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
+For both sexes, selection effects are larger and more significant among
+those with more than 3 partners. Results are unchanged if we control for
+respondents' age (appendix Figure
+\@ref(fig:plot-n-partners-controlled)). Splitting respondents by whether
+they were living with a spouse or partner at the time of interview,
+effects are larger among those not living with a spouse or partner
+(Figure \@ref(fig:plot-n-partners)b). Lastly, we split female
+respondents by age at first live birth (AFLB).[^1] There is evidence for
+genetic effects on AFLB [@Barban_2016], and there is a close link
+between this variable and number of children born. Figure
+\@ref(fig:plot-n-partners)c shows effect sizes estimated separately for
+each tercile of AFLB. Several effects are strikingly different across
+terciles. ADHD and MDD are selected for amongst the youngest third of
+mothers, but selected against among the oldest two-thirds. Educational
+attainment is selected for among the oldest two-thirds of mothers, but
+is not significantly selected among the youngest third. Similarly,
+several PGS of body measurements are selected against only among older
+mothers. The correlation between effect sizes for the youngest and
+oldest terciles is `r cor_age_flb`. As before, recall that PGS may have
+two effects: changing the number of children within categories, and
+shifting respondents between categories.
+
+[^1]: This information is unavailable for men.
+
+```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth", fig.subcap = c("Number of partners", "Household type", "Age at first live birth"), fig.ncol = 1, out.height = c("35%", "27%", "27%"), fig.align = "center"}
+
+drake::loadd(res_partners)
+res_partners %<>%       
+      filter(grepl(":", term)) %>% 
+      mutate(`N partners` = ifelse(grepl("TRUE:", term), "<= 3", "> 3"))
+
+n_regs <- as.double(nrow(res_partners))
+res_partners %>% 
+      mutate(
+        score_name = pretty_names(score_name)
+      ) %>% 
+      ggplot(aes(
+        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        x      = estimate, 
+        fill   = `N partners`, 
+        colour = p.value < 0.05/{{n_regs}}
+      )) + 
+        my_vline +
+        pval_geom_point() +
+        facet_wrap(vars(sex), scales = "free_y") +
+        my_fill_scale(n = 2) +
+        pval_scale() +
+        my_theme(legend.position = "bottom") +
+        my_labs(x = "", y = "") + 
+        scale_y_reordered()
+
+drake::loadd(res_with_partner)
+res_with_partner %<>% filter(grepl(":", term))
+
+n_regs <- nrow(res_with_partner)
+hhtype_n <- n_in_regs("with_partner")
+
+res_with_partner %>% 
+      mutate(
+        Household = ifelse(grepl("TRUE", term), "With partner", 
+                      "Without partner"),
+        with_p    = as.character(grepl("TRUE", term)),
+        Household = sprintf("%s (N = %s)", Household, 
+                      prettyNum(hhtype_n[with_p], big.mark = ","))
+      ) %>% 
+      standard_ggplot(fill = Household, n_regs = n_regs, score_col = score_name)
+
+drake::loadd(res_age_flb_cross)
+
+# this counts each cross term as a separate test:
+n_regs <- as.double(nrow(res_age_flb_cross))
+
+res_age_flb_cross %>% 
+      mutate(
+        score_name = pretty_names(score_name),
+        score_name = fct_reorder(score_name, estimate, order_abs(1)),
+        term       = gsub("age_flb_cat(.*):.*", "\\1", term),
+        term       = add_n(term, "age_flb_cat")
+      ) %>% 
+      ggplot(aes(estimate, score_name, fill = term, 
+        colour = p.value < 0.05/{{n_regs}})) +
+        my_vline +
+        pval_geom_point() +
+        pval_scale() +
+        my_fill_scale(n = 3, direction = -1, 
+          guide = guide_legend(title = "Age at first live birth")) +
+        my_labs() +
+        my_theme()
+```
+
+```{r calc-corr-age-flb}
+
+drake::loadd(res_age_flb)
+drake::loadd(res_sex)
+
+res_sex_comparison <- res_sex %>% 
+      filter(sex == "Female") %>% 
+      mutate(score_name = term) %>% 
+      select(score_name, term, estimate:conf.high)
+
+res_age_flb %<>% filter(term != "age_flb")
+
+res_combined <- bind_rows(
+        raw      = res_sex_comparison, 
+        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
+        .id      = "reg.type"
+      ) %>% 
+      arrange(score_name)
+
+raw_flb_comparison <- res_combined %>% 
+      pivot_wider(
+        id_cols     = score_name, 
+        names_from  = reg.type, 
+        values_from = estimate
+      )
+
+cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
+opp_signed <- sum(
+        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
+      )
+
+```
+
+To investigate this further, we regress number of children on PGS
+*controlling* for AFLB (appendix Figure \@ref(fig:plot-age-flb)). In
+`r opp_signed` out of `r length(score_names)` cases, effects
+change sign when controls are added. The correlation between effect
+sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`.
+Selection effects appear to come through two opposing channels: an
+effect on AFLB, and an opposite-signed effect on number of children 
+controlling for AFLB.
+
+```{r calc-res-weighted}
+
+drake::loadd(res_wt_flb_weights)
+drake::loadd(res_wt_msoa_weights)
+drake::loadd(res_wt_age_qual_weights)
+drake::loadd(res_sibs_parent_weights)
+drake::loadd(res_all)
+drake::loadd(res_sex)
+
+res_wt_combined <- bind_rows(
+  None                     = res_all %>% 
+                               filter(reg.type == "raw", dep.var == "N children") %>% 
+                               select(-reg.type, -dep.var, -score_name),
+  Geographical             = res_wt_msoa_weights,
+  `Age/Qualification`      = res_wt_age_qual_weights,
+  `Age/Qual/AFLB`  = res_wt_flb_weights,
+  .id = "Weights"
+)
+
+res_wt_cb_wide <- res_wt_combined %>% 
+                    tidyr::pivot_wider(
+                      id_cols     = term,
+                      names_from  = Weights,
+                      values_from = estimate
+                    ) %>% 
+                    rowwise() %>% 
+                    mutate(
+                      consistent = all(c_across(-term) > 0) ||
+                        all(c_across(-term) < 0)
+                    ) %>% 
+                    ungroup() %>% 
+                    left_join(
+                      res_wt_combined %>% 
+                        filter(Weights == "None") %>% 
+                        select(term, p.value),
+                      by = "term"
+                    ) %>% 
+                    left_join(
+                      res_sex %>% 
+                        filter(sex == "Female") %>% 
+                        select(term, estimate_females = estimate),
+                      by = "term"
+                    )
+
+rel_sizes <- res_wt_cb_wide %>% 
+                filter(
+                  consistent, 
+                  p.value < 0.05/length(score_names)
+                ) %>% 
+                mutate(
+                  Geographical = Geographical/None,
+                  `Age/Qualification` = `Age/Qualification`/None,
+                  `Age/Qual/AFLB` = `Age/Qual/AFLB`/estimate_females
+                ) %>% 
+                select(-consistent, -None, -p.value, -estimate_females) %>% 
+                mutate(
+                  term = pretty_names(term)
+                )
+
+rel_sizes_summary <- rel_sizes %>% 
+                summarize(
+                  across(-term, ~ c(mean(.x), median(.x)))
+                ) %>% 
+                mutate(
+                  term = c("Mean", "Median")
+                )
+
+mean_rel <- c(rel_sizes_summary[1, 1:3])
+```
+
+Since UK Biobank subjects are not representative of the wider population, our
+results suggest that naïve estimates of natural selection are at risk of
+ascertainment bias. To correct for this, we weight participants using data from
+the 2011 Census and from the 2006 General Household Survey (GHS). We try three
+alternative weighting schemes: geographical weighting by census MSOA, age and
+presence/absence of a partner; weighting by age and highest educational
+qualification; and age, highest qualification, and age at first live birth, for
+women only. Mean effect sizes across all polygenic scores are increased by 
+`r mean_rel$Geographical` (geographical weighting), 
+`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
+`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
+further affected by weighting on other demographic variables. Since the Biobank
+sample seems to be ascertained in ways that reduce estimates of selection
+effects, we suspect that our weighted estimates remain conservative.
+
+The UK Biobank data contains information on respondents' number of
+siblings. Since respondents' polygenic scores are equal in expectation
+to the mean scores of their parents, we can use this to look at selection 
+effects in the parents' generation. We estimate equation (\ref{eq:regression})
+using *number of siblings* (including the respondent) as the dependent variable.
+
+The parents' generation has a further source of ascertainment bias: sampling
+parents of respondents overweights parents who have many children. For instance,
+parents of three children will have, on average, three times more children
+represented in UK Biobank than parents of one child. Parents of no children will
+by definition not be represented. To compensate, we reweight our preferred
+weightings (Age/Qualification) by the inverse of *number of siblings*. We then
+regress this on our polygenic scores. Selection effects are highly correlated
+across the two generations, and most share the same sign (appendix Figure
+\@ref(fig:plot-res-sibs)). Effect sizes appear larger in the parents'
+generation. We treat this result cautiously, because when we split respondents
+up by year of birth, we find few differences in effect sizes between early- and
+late-born respondents, for either generation. In other words, since estimated
+effect sizes change when we change the dependent variable, but do not change
+over time within either dependent variable, the explanation may be due to
+remaining ascertainment bias within our sample.
+
+Although the direction of selection effects does not change between the generations,
+there are some differences. In particular, after we residualize polygenic scores
+on the first 100 principal components of the array data, selection effects on 
+*number of siblings* are reduced by about ten percent on average, whereas effects on
+*number of children* barely change (appendix Figure 
+\@ref(fig:plot-regs-controlled-pcs)). One possible explanation for this is that
+earlier fertility is driven more by poverty, which is geographically clustered
+and therefore correlates with the broad-scale genetic variation captured by
+principal components. (TODO: insurance motive.) Another difference is that
+in the parents' generation, raw selection effects and effects controlling for
+AFLB are positively correlated, whereas they are negatively correlated in
+the respondents' generation.[^educ-exception] These differences suggest that
+the mechanisms of natural selection may have changed over time, even though
+the outcomes are broadly the same.
+
+[^educ-exception] Educational attainment is an exception. It is selected 
+against in the parents' generation, but selected for if we control for AFLB,
+which is the same pattern as for respondents.
+
+Our empirical results can be explained by the economic
+theory of fertility [@becker1960economic; @willis1973new; @becker1976child].
+According to this, increases in a person's wage affect their fertility via two
+opposing channels. There is an *income effect* by which children become more
+affordable, like any other good. There is also a *substitution effect*: since
 childrearing has a cost in time, the opportunity cost of childrearing increases
@@
 have more children. The substitution effect would lead higher earners to have
-fewer children. If so, then genetic variants which affect earnings
-potential in the labour market may cause opposing effects on
-fertility. The income effect will cause natural selection in favour of 
-earnings-increasing variants. The substitution effect will do the reverse.
+fewer children. If so, then genetic variants which affect earnings potential in
+the labour market may cause opposing effects on fertility. The income effect
+will cause natural selection in favour of earnings-increasing variants. The
+substitution effect will do the reverse.
+
+```{r calc-res-earn-educ}
+
+drake::loadd(res_ee_control)
+drake::loadd(res_all)
+
+res_ee_control %<>% filter(term == score_name)
+n_05 <- sum(res_ee_control$p.value < 0.05)
+n_05_33 <- sum(res_ee_control$p.value < 0.05/33)
+
+```
 
@@
 plays an important role in natural selection, with substitution effects
-dominating overall.
+dominating overall. We also re-estimate equation \@ref(eq:regression) for each
+PGS, controlling for earnings in first job and education levels. Effect sizes
+are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only
+`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$, suggesting that
+earnings and education indeed mediate the effect of PGS on fertility.
 
@@
 
-Suppose that people are aware of their endowments of human capital, and 
-that education and innate human capital are complements. They choose how long
-to stay in education. After leaving education they enter the labour market and
-form families. If so, then people with less human capital will leave education 
-earlier and have their first child earlier. Whilst in the labour market, higher
-earners will have more children (i.e. income effects dominate). These 
-assumptions will lead to the pattern we observe where earnings-increasing
+This theory can explain why selection effects have opposite signs when AFLB
+is controlled for. Suppose that people are aware of their endowments of human
+capital, and that education and innate human capital are complements. They
+choose how long to stay in education. After leaving education they enter the
+labour market and form families. If so, then people with less human capital will
+leave education earlier and have their first child earlier. This is a pure
+substitution effect: the income effect plays no role in the decision to leave
+education, since people are not yet earning wages. Whilst in the labour
+market, higher earners will have more children (i.e. income effects dominate).
+These assumptions will lead to the pattern we observe: earnings-increasing
 variants lead to later first births (and fewer children overall), but more
@@
 and fertility.
+
+* TODO: more.
 
@@
 
-Empirical work shows similar relationships between *earnings* or *education* and
+Previous work shows similar relationships between *earnings* or *education* and
 fertility, to the relationships shown here between *genetics* and fertility.
@@
 [@baudin2015fertility].
-
-```{r calc-res-earn-educ}
-
-drake::loadd(res_ee_control)
-drake::loadd(res_all)
-
-res_ee_control %<>% filter(term == score_name)
-n_05 <- sum(res_ee_control$p.value < 0.05)
-n_05_33 <- sum(res_ee_control$p.value < 0.05/33)
-
-```
-
-The economic theory predicts that natural selection on polygenic scores is mediatet
-by labour market earnings. As a final test of this,
-we re-estimate equation (\@ref(eq:regression)) for each PGS, 
-controlling for earnings in first job and education levels. Effect sizes are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only `r n_05_33` 
-out of 33 PGS are significant at $p = 0.05/33$.
 
@@
 stratification, although would not change our results for natural
-selection of the scores. In \@ref(fig:plot-regs-controlled) we show
-results for selection on polygenic scores residualized for the top 100
+selection of the scores. Figure \@ref(fig:plot-regs-controlled-pcs) shows
+selection effects on polygenic scores residualized for the top 100
 principal components of the genetic data, calculated within the UK
-Biobank population. (TODO Abdel details.)
+Biobank population.
 
@@
 
-```{r plot-regs-controlled, fig.cap = "Effects of residualized polygenic scores on number of siblings/children."}
+```{r plot-regs-controlled-pcs, fig.cap = "Effects of residualized polygenic scores on number of siblings/children."}
 
@@
 
-```{r plot-age-birth-parents, fig.cap = "Effect sizes on number of siblings controlling for parents' age at birth, eldest siblings"}
+```{r plot-age-birth-parents, fig.cap = "Selection effects controlling for parents' age at birth, eldest siblings"}
 

### 2020-11-13 why-negative-selection.Rmd
Tweaks and literature

@@
 ---
-title: "Natural Selection in the 20th Century"
+title: "Economic Theory of Fertility Explains Modern Natural Selection"
 author: "David Hugh-Jones, Abdel Abdellaoui"
@@
 Selection effects appear consistent over time.
-The strength of natural selection on a polygenic score can be predicted by its correlation with earnings and education.
-Effects are larger among low-income and less educated individuals, younger parents, and people not living with a partner.
-Effect sizes are larger when we correct for ascertainment bias.
+Polygenic scores which predict lower earnings and education are selected for.
+Effects are larger among low-income and less educated people, younger parents, and people not living with a partner.
+Overall effect sizes are substantially bigger when we correct for ascertainment bias.
 The direction of natural selection is reversed among older parents, or controlling for age at first live birth.
@@
 qualification; and age, highest qualification, and age at first live birth, for
-women only. Mean effect sizes across all polygenic scores are increased by 
-`r mean_rel$Geographical` (geographical weighting), 
+women only. Mean effect sizes across all polygenic scores are increased by a
+factor of `r mean_rel$Geographical` (geographical weighting), 
 `r mean_rel[["Age/Qualification"]]` (age/qualification) or 
@@
 further affected by weighting on other demographic variables. Since the Biobank
-sample seems to be ascertained in ways that reduce estimates of selection
-effects, we suspect that our weighted estimates remain conservative.
+sample seems to be ascertained in ways that shrink estimates of selection
+effects, we suspect that our weighted estimates are still conservative.
 
@@
 using *number of siblings* (including the respondent) as the dependent variable.
-
 The parents' generation has a further source of ascertainment bias: sampling
@@
 by definition not be represented. To compensate, we reweight our preferred
-weightings (Age/Qualification) by the inverse of *number of siblings*. We then
-regress this on our polygenic scores. Selection effects are highly correlated
-across the two generations, and most share the same sign (appendix Figure
-\@ref(fig:plot-res-sibs)). Effect sizes appear larger in the parents'
-generation. We treat this result cautiously, because when we split respondents
-up by year of birth, we find few differences in effect sizes between early- and
-late-born respondents, for either generation. In other words, since estimated
-effect sizes change when we change the dependent variable, but do not change
-over time within either dependent variable, the explanation may be due to
-remaining ascertainment bias within our sample.
+weightings (Age/Qualification) by the inverse of *number of siblings*.
+
+Selection effects are highly correlated across the two generations, and most
+share the same sign (appendix Figure \@ref(fig:plot-res-sibs)). Effect size
+estimates are larger for the parents' generation. We treat this result cautiously,
+because when we split respondents up by year of birth, we find few differences
+in effect sizes between early- and late-born respondents, for either generation.
+In other words, since estimated effect sizes change when we change the dependent
+variable, but do not change over time within either dependent variable, this
+may be due to remaining ascertainment bias within the sample.
 
 Although the direction of selection effects does not change between the generations,
-there are some differences. In particular, after we residualize polygenic scores
-on the first 100 principal components of the array data, selection effects on 
-*number of siblings* are reduced by about ten percent on average, whereas effects on
+there are other differences. After we residualize polygenic scores on the first
+100 principal components of the array data, selection effects on *number of
+siblings* are reduced by about ten percent on average, whereas effects on
 *number of children* barely change (appendix Figure
-\@ref(fig:plot-regs-controlled-pcs)). One possible explanation for this is that
-earlier fertility is driven more by poverty, which is geographically clustered
-and therefore correlates with the broad-scale genetic variation captured by
-principal components. (TODO: insurance motive.) Another difference is that
-in the parents' generation, raw selection effects and effects controlling for
-AFLB are positively correlated, whereas they are negatively correlated in
-the respondents' generation.[^educ-exception] These differences suggest that
-the mechanisms of natural selection may have changed over time, even though
-the outcomes are broadly the same.
-
-[^educ-exception] Educational attainment is an exception. It is selected 
+\@ref(fig:plot-regs-controlled-pcs)). This could be because
+earlier fertility is driven more by geographically clustered deprivation [e.g.
+via an insurance motive, @rendall1998old], which may correlate with the
+broad-scale genetic variation captured by principal components. Another
+difference is that in the parents' generation, raw selection effects and effects
+controlling for AFLB are positively correlated, whereas they are negatively
+correlated in the respondents' generation.[^educ-exception] These differences
+suggest that the mechanisms of natural selection may have changed over time,
+even though the outcomes are broadly the same.
+
+[^educ-exception]: Educational attainment is an exception. It is selected 
 against in the parents' generation, but selected for if we control for AFLB,
@@
 selection effects controlling for age at first live birth (right column), the
-effect reverses and becomes positive. This is evidence that the labour market
-plays an important role in natural selection, with substitution effects
-dominating overall. We also re-estimate equation \@ref(eq:regression) for each
+effect reverses and becomes positive. Thus, the labour market appears to
+play an important role in natural selection. Substitution effects
+dominate income effects overall, in accordance with the widespread association 
+between income and lower fertility [@becker1960economic;@jones2006economic].
+
+[^earnings-est]: Earnings are estimated from the 2007 Annual Survey of Hours and
+Earnings, using the SOC 2000 job code.
+
+We also re-estimate equation \@ref(eq:regression) for each
 PGS, controlling for earnings in first job and education levels. Effect sizes
@@
 
-[^earnings-est]: Earnings are estimated from the 2007 Annual Survey of Hours and
-Earnings, using the SOC 2000 job code.
-
 ```{r plot-cors-earnings-educ, fig.cap="Selection effects, with and without controls for age at first live birth, by correlations with earnings and educational attainment.
@@
 # TODO: use age_left_edu when this includes college
-# TODO:
+
 cors_educ <-   cor(famhist[score_names], famhist$age_fulltime_edu, use = "pairwise")
@@
 
-This theory can explain why selection effects have opposite signs when AFLB
-is controlled for. Suppose that people are aware of their endowments of human
+This theory can explain why selection effects have opposite signs when AFLB is
+controlled for. Suppose that people are aware of their endowments of human
 capital, and that education and innate human capital are complements. They
@@
 labour market and form families. If so, then people with less human capital will
-leave education earlier and have their first child earlier. This is a pure
-substitution effect: the income effect plays no role in the decision to leave
-education, since people are not yet earning wages. Whilst in the labour
-market, higher earners will have more children (i.e. income effects dominate).
-These assumptions will lead to the pattern we observe: earnings-increasing
-variants lead to later first births (and fewer children overall), but more
-children for any given AFLB.
+leave education earlier and have their first child earlier [cf.
+@caucutt2002women; @monstad2008education]. This is a pure substitution effect: the income effect plays
+no role in the decision to leave education, since people are not yet earning
+wages.[^no-borrow] Whilst in the labour market, higher earners will have more
+children (i.e. income effects dominate). These assumptions will lead to the
+pattern we observe: earnings-increasing variants lead to later first births (and
+fewer children overall), but more children for any given AFLB.
+
+[^no-borrow]: This argument assumes that capital market imperfections hinder
+people from borrowing against future income.
 
@@
 higher income levels; this can lead to a U-shaped relationship between income
-and fertility.
-
-* TODO: more.
-
-Lastly, it is often assumed that the substitution effect will dominate
-for lone parents, or those in unstable relationships, since they have less
-opportunity to share childcare responsibilities, while the income effect will
-dominate for couples who are able to reap the gains from specialization. If so,
-then earnings-increasing genetic variants may decrease fertility more among single 
-parents than among couples, as in our data. The same idea could explain our
-results for lifetime number of sexual partners, if this is associated with
-relationship stability.
-
-Previous work shows similar relationships between *earnings* or *education* and
-fertility, to the relationships shown here between *genetics* and fertility.
-Higher income is associated with lower fertility [@becker1960economic;
-@jones2006economic].
-@cohen2013financial find that income decreases fertility at low income levels
-but increases it at higher income levels. In a US sample, fertility decreases
-with education faster among single mothers than married mothers 
-[@baudin2015fertility].
+and fertility. Empirically, @cohen2013financial find that income decreases 
+fertility at low income levels but increases it at higher income levels.
+
+Lastly, it is often assumed that the substitution effect will dominate for lone
+parents, or those in unstable relationships, since they have less opportunity to
+share childcare responsibilities, while the income effect will dominate for
+couples who are able to reap the gains from specialization. Indeed, US fertility
+decreases faster with education among single mothers than married mothers
+[@baudin2015fertility].If so, then earnings-increasing genetic variants may
+decrease fertility more among single parents than among couples, as in our data.
+The same idea could explain our results for lifetime number of sexual partners,
+if this is associated with relationship stability.
 

### 2020-11-13 why-negative-selection.Rmd
p.value to p in graphs

@@
         ) %>% 
+        rename(p = p.value) %>% 
         ggplot(aes(estimate, {{score_col}}, fill = {{fill_col}}, ...,
-          colour = p.value < 0.05/{{n_regs}})) +
+          colour = p < 0.05/{{n_regs}})) +
           pval_geom_point() +
@@
       ) %>% 
-      ggplot(aes(y = term, x = estimate, colour = p.value < 0.05/{{n_regs}}, 
+      rename(p = p.value) %>% 
+      ggplot(aes(y = term, x = estimate, colour = p < 0.05/{{n_regs}}, 
         fill = `Age left FTE`)) +
@@
       ) %>% 
+      rename(p = p.value) %>% 
       ggplot(aes(y = term, x = estimate, fill = Income, 
-        colour = p.value < 0.05/{{n_regs}})) +
+        colour = p < 0.05/{{n_regs}})) +
         my_vline +
@@
       filter(grepl(":", term)) %>% 
-      mutate(`N partners` = ifelse(grepl("TRUE:", term), "<= 3", "> 3"))
+      mutate(`N partners` = ifelse(grepl("TRUE:", term), "3 or less", "4 or more"))
 
@@
       ) %>% 
+      rename(p = p.value) %>% 
       ggplot(aes(
@@
         fill   = `N partners`, 
-        colour = p.value < 0.05/{{n_regs}}
+        colour = p < 0.05/{{n_regs}}
       )) + 
@@
       ) %>% 
+      rename(p = p.value) %>% 
       ggplot(aes(estimate, score_name, fill = term, 
-        colour = p.value < 0.05/{{n_regs}})) +
+        colour = p < 0.05/{{n_regs}})) +
         my_vline +
@@
       ) %>% 
+      rename(p = p.value) %>% 
       ggplot(aes(y = term, x = estimate, fill = sex, 
-            colour = p.value < 0.05/{{n_regs}})) +
+            colour = p < 0.05/{{n_regs}})) +
         my_vline +
@@
       ) %>% 
+      rename(p = p.value) %>% 
       ggplot(aes(y = score_name, x = estimate, fill = Regression, 
-        color = p.value < 0.05/{{n_regs}})) + 
+        color = p < 0.05/{{n_regs}})) + 
         geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
@@
   ) %>% 
-  ggplot(aes(x = estimate, y = term, shape = Weights, fill = Weights, colour = p.value < 0.05/{{n_regs}})) +
+  rename(p = p.value) %>% 
+  ggplot(aes(
+    x = estimate, 
+    y = term, 
+    shape = Weights, 
+    fill = Weights, 
+    colour = p < 0.05/{{n_regs}}
+  )) +
   geom_point(size = 1.5, stroke = 0.5) +
@@
       ) %>% 
+      rename(p = p.value) %>% 
       ggplot(aes(y = term, x = estimate, fill = reg.type, 
-        colour = p.value < 0.05/{{n_regs}})) +
+        colour = p < 0.05/{{n_regs}})) +
         my_vline + 
@@
       ) %>% 
-      ggplot(aes(estimate, dep.var, colour = p.value < 0.05/{{n_regs}})) + 
+      rename(p = p.value) %>% 
+      ggplot(aes(estimate, dep.var, colour = p < 0.05/{{n_regs}})) + 
         geom_jitter(height = 0.15, width = 0, alpha = 0.9) + 
@@
     ) %>% 
+    rename(p = p.value) %>% 
     ggplot(aes(y = term, x = estimate, fill = Control, 
-      colour = p.value < 0.05/{{n_regs}})) + 
+      colour = p < 0.05/{{n_regs}})) + 
       pval_geom_point(alpha = 0.8) +  
@@
       ) %>% 
-      ggplot(aes(estimate, term, colour = p.value < 0.05 / {{n_regs}})) + 
+      rename(p = p.value) %>% 
+      ggplot(aes(estimate, term, colour = p < 0.05 / {{n_regs}})) + 
       my_vline +
@@
       ) %>% 
+      rename(p = p.value) %>% 
       ggplot(aes(estimate, term, fill = `Dependent variable`, 
-        colour = p.value < 0.05 / {{n_regs}})) + 
+        colour = p < 0.05 / {{n_regs}})) + 
         my_vline +
@@
       ) %>% 
+      rename(p = p.value) %>% 
       ggplot(aes(y = reorder_within(term, estimate, dep.var, order_abs(1)), 
-        x = estimate, color = p.value < 0.05/{{n_regs}}, fill  = PCs)) + 
+        x = estimate, color = p < 0.05/{{n_regs}}, fill  = PCs)) + 
         pval_geom_point(alpha = 0.8) +

### 2020-11-16 why-negative-selection.Rmd
tweaks from abdel comments

@@
 title: "Economic Theory of Fertility Explains Modern Natural Selection"
-author: "David Hugh-Jones, Abdel Abdellaoui"
-date: "`r Sys.Date()`"
+author: "David Hugh-Jones (D.Hugh-Jones@uea.ac.uk), Abdel Abdellaoui (a.abdellaoui@gmail.com)"
+date: "`r format(Sys.Date(), '%d %B %Y')`"
 abstract: "Natural selection has been documented in contemporary humans, but little is known about the mechanisms behind it.
@@
 Overall effect sizes are substantially bigger when we correct for ascertainment bias.
-The direction of natural selection is reversed among older parents, or controlling for age at first live birth.
+The direction of natural selection is reversed among older parents, or after controlling for age at first live birth.
 These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility."
@@
     dpw_substance_use       = "Drinks per week",
-    EA2_noUKB               = "Educ. attainment 2 (no UK)",
+    EA2_noUKB               = "Educ. attainment 2 (no UKBB)",
     EA3_excl_23andMe_UK     = "Educ. attainment 3 (no UK)",
@@
 
-Geneticists have long hypothesized that natural selection is taking
-place in modern human populations. Recent work confirms this using
-genome-wide analysis [@Barban_2016; @beauchamp2016genetic;
-@kong2017selection; @sanjak2018evidence]. In particular, genetic
-variants associated with higher educational attainment are being
-selected against, although effect sizes appear small.
+Living organisms evolve through natural selection, in which allele frequencies
+change in the population through differential reproduction rates. Geneticists
+have long hypothesized that natural selection is taking place in modern human
+populations. Recent work confirms this using genome-wide analysis [@Barban_2016;
+@beauchamp2016genetic;
+@kong2017selection; @sanjak2018evidence]. In particular, genetic variants
+associated with higher educational attainment are being selected against,
+although effect sizes appear small.
 
 As yet we know little about the *mechanisms* behind natural selection in
-contemporary populations. This paper uses data from UK Biobank to learn
-more. We test for natural selection on 33 different polygenic scores. We
-extend the analysis over two generations, using data on respondents'
-number of siblings as well as their number of children to reconstruct
-two generations of family size. This is interesting because consistent
-natural selection over multiple generations could lead to substantive
-effects in the long run. Most importantly, we examine the strength of
-natural selection in different subgroups of the population, in order to
-uncover patterns that can help illuminate the mechanisms involved.
+contemporary human populations. This study uses data from UK Biobank to learn
+more. We test for natural selection on 33 different polygenic scores. We extend
+the analysis over two generations, using data on respondents' number of siblings
+as well as their number of children. This is interesting because consistent
+natural selection over multiple generations could lead to substantive effects in
+the long run. Most importantly, we examine reproductive rates in different
+subgroups of the population, in order to uncover patterns that can help
+illuminate the mechanisms behind modern natural selection.
 
@@
 oldest terciles is `r cor_age_flb`. As before, recall that PGS may have
-two effects: changing the number of children within categories, and
-shifting respondents between categories.
+two effects: changing the number of children within each category, and changing
+the number of children by shifting respondents between categories.
 
@@
 controlled for. Suppose that people are aware of their endowments of human
-capital, and that education and innate human capital are complements. They
+capital, i.e. skills and characteristics that are valuable in the labour market.
+Suppose also that education and innate human capital are complements. They
 choose how long to stay in education. After leaving education they enter the
@@
 children (i.e. income effects dominate). These assumptions will lead to the
-pattern we observe: earnings-increasing variants lead to later first births (and
-fewer children overall), but more children for any given AFLB.
+pattern we observe: earnings-increasing variants will predict later first births (and
+fewer children overall), but they will also predict more children within any given
+age at first live birth.
 
@@
 
-```{r plot-age-flb, fig.cap = "Selection effects controlling for age at first live birth (women only). Effect sizes for women without controls are shown for comparison"}
+```{r plot-age-flb, fig.cap = "Selection effects controlling for age at first live birth (women only). Effect sizes for women only without controls are shown for comparison."}
 
@@
 
-```{r plot-res-wt, fig.cap = "Effects of polygenic scores on fertility. Weighted regressions."}
+```{r plot-res-wt, fig.cap = "Selection effects: weighted regressions."}
 
@@
 
-```{r plot-res-sibs, fig.cap = "Selection effects, respondents' parents vs. respondents"}
+```{r plot-res-sibs, fig.cap = "Selection effects, respondents' parents vs. respondents."}
 drake::loadd(res_sibs_parent_weights)
@@
       count() %>% 
-      period_table("Change in effect sizes between parents of early and late born respondents (regressions on number of siblings)")
+      period_table("Change in selection effects between parents of early and late born respondents (regressions on number of siblings)")
 
@@
       count() %>% 
-      period_table("Change in effect sizes between early and late born respondents (regressions on number of children)")
+      period_table("Change in selection effects between early and late born respondents (regressions on number of children)")
 ```
@@
 
-```{r plot-regs-controlled-pcs, fig.cap = "Effects of residualized polygenic scores on number of siblings/children."}
+```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using residualized polygenic scores on number of siblings/children."}
 
@@
 
-```{r plot-pc-regs, fig.cap = "Effect of principal components of genetic data on number of siblings/children.
+```{r plot-pc-regs, fig.cap = "Selection effects of principal components of genetic data.
 Absolute effect sizes are plotted.
@@
 
-```{r plot-n-partners-controlled, fig.cap = "Effect sizes on number of siblings interacted with number of sexual partners (median split at 3), respondent's age and age squared."}
+```{r plot-n-partners-controlled, fig.cap = "Selection effects by number of sexual partners (median split at 3), controlling for respondent's age and age squared."}
 drake::loadd(res_partners_controlled)
@@
 
-```{r plot-age-birth-parents, fig.cap = "Selection effects controlling for parents' age at birth, eldest siblings"}
+```{r plot-age-birth-parents, fig.cap = "Selection effects (parents' generation) among eldest siblings, controlling for parents' age at birth."}
 
@@
 
-```{r plot-age-flb-dv, fig.cap = "Effects of polygenic scores on age at first live birth"}
+```{r plot-age-flb-dv, fig.cap = "Effects of polygenic scores on age at first live birth."}
 
@@
 
-```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at own birth, eldest siblings"}
+```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at respondent's birth, eldest siblings."}
 
@@
         "Dependent variable" = dplyr::recode(dep.var, 
-          "fath_age_birth" = "Father's age", 
-          "moth_age_birth" = "Mother's age"
+          "fath_age_birth" = "Father's AFLB", 
+          "moth_age_birth" = "Mother's AFLB"
         )
@@
 
-```{r plot-rgs-by-effect-size, fig.cap = "Effect size on n siblings by genetic correlation with EA3"}
+```{r plot-rgs-by-effect-size, fig.cap = "Selection effects plotted against genetic correlation with EA3."}
 
@@
 cor(combined_results[-1], use = "complete")
-
-```
-
-\FloatBarrier
-## Number of children
-
-Figure \@ref(fig:plot-n-children-bar) shows the full distribution of
-number of children born for different ventiles of the EA3 polygenic
-score. The strongest relationship seems to be for having 0 children
-versus 1 or more.
-
-```{r plot-n-children-bar, fig.cap = "Number of children by ventiles of EA3 PGS"}
-famhist %>% 
-      select(n_children, EA3_excl_23andMe_UK, sex) %>% 
-      mutate(
-        "EA3 Ventile" = chop_equally(EA3_excl_23andMe_UK, 20, labels = 1:20),
-        "N Children"  = santoku::chop(n_children, 0:4, labels = c(0:3, "4+")),
-        Sex           = factor(sex, levels = 0:1, labels = c("Female", "Male"))
-        ) %>% 
-      filter(! is.na(`EA3 Ventile`), ! is.na(`N Children`)) %>% 
-      ggplot(aes(`EA3 Ventile`, fill = `N Children`, group = `N Children`)) + 
-        geom_bar(position = position_fill(reverse = TRUE), alpha = 0.7) + 
-        facet_wrap(vars(Sex)) +
-        scale_fill_viridis_d(option = "A") +
-        scale_y_continuous(n.breaks = 10, labels = scales::percent) +
-        theme(legend.position = "bottom")
 
@@
 
-```{r plot-together, fig.cap = "Partial correlations with number of children"}
+```{r plot-together, fig.cap = "Partial correlations of PGS with fertility, from a single regression."}
 

### 2020-11-18 why-negative-selection.Rmd
Limit to over-45s. Run with_partner regression by sex.

@@
 Polygenic scores which predict lower earnings and education are selected for.
-Effects are larger among low-income and less educated people, younger parents, and people not living with a partner.
-Overall effect sizes are substantially bigger when we correct for ascertainment bias.
+Effects are concentrated among low-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner.
+Overall effect sizes are substantially larger after correcting for ascertainment bias.
 The direction of natural selection is reversed among older parents, or after controlling for age at first live birth.
@@
 n_in_regs <- function (var, dv = "n_children") {
+  fh_subset <- famhist
+  if (dv == "n_children") fh_subset %<>% filter(age_at_recruitment >= 45)
   # all PGS have the same NA pattern, so we just use one
-  tbl <- table(famhist[
-                  ! is.na(famhist[dv]) & ! is.na(famhist$whr_combined), 
+  tbl <- table(fh_subset[
+                  ! is.na(fh_subset[dv]) & ! is.na(fh_subset$whr_combined), 
                   var
                 ])
+
   c(tbl)
@@
 
+* Check each result with a manually done regression!
 * If we weight with the GHS, and plot the means of figure 1 using this w
@@
 * Reweight other regressions (e.g. the omnibus regression in the appendix)?
-* Look at geography, esp. in siblings regressions.
-* We say we don't see change over time. But, we do see the effect of
-  controlling for AFLB changing over generations.
-* Beauchamp 2016 used the HRS. He reckons it's representative. Is it? How do
-  his effect sizes compare?
-  - he also worries about the quality/quantity tradeoff, which might mean that
-    people with fewer children have more grandchildren (than you'd expect).
-    As we have the two generations, we can test this. (Actually, there's a
-    positive correlation of about 0.1)
+* Look at geography, esp. in siblings regressions 
+  - maybe Townshend f.189? Not sure essential
 * Another point is that negative selection only among poor will increase
@@
   project to define with_partner
+  - `with_partner` is fine, but if you want to have pairs who had the same children
+  you need to do more
 * If you take "year of last job started" and see if it is after/before
@@
   choice to work is endogenous.
-* Put mediating effects of pay and education in there (see explore-sex-vs-pay.R)
-  - in a multivariate regression, most effects vanish once we control for these, including EA3
+* double check which scores are reverse coded!
 
@@
   - see explore-sex-vs-pay.R
+
+## RIVAL MECHANISMS
+* Lynn 2011 mentions two:
+  - social class gradient on ability to delay gratification
+  - modern society protects the weak (Darwin 1871, Spencer 1874)
+    NB this latter doesn't predict more fertility among "dysgenic" ppl, just
+    equal fertility and then genetic drift does its work
+    - examples include modern sanitation, post-1800
+    - and monogamy!
 "
@@
 natural selection over multiple generations could lead to substantive effects in
-the long run. Most importantly, we examine reproductive rates in different
+the long run.
+
+Most importantly, we examine reproductive rates in different
 subgroups of the population, in order to uncover patterns that can help
-illuminate the mechanisms behind modern natural selection.
-
-We use data from UK Biobank. We compute 33 polygenic scores, normalized
+illuminate the mechanisms behind modern natural selection. We find that 
+across the board, polygenic scores have stronger relationships with fertility 
+in specific subgroups of the population. Furthermore, in some subgroups, 
+the direction of relationships is actually reversed.
+
+These patterns can be explained by economic theories of fertility
+[@becker1960economic]. In these, higher potential earnings have two opposite
+effects on fertility: a fertility-increasing *income effect* (higher income
+makes children more affordable), and a fertility-lowering *substitution effect*
+(time spent on childrearing has a higher cost in foregone earnings). Our
+results suggest that the substitution effect dominates for single parents,
+while the income effect is stronger for couples.
+
+# Data and methods
+
+We use data from UK Biobank. Our sample consists of white British (as defined
+by genetic estimated ancestry). For regressions on number of children we 
+limit the sample to people over 45, since most fertility is completed by this age.
+
+We compute 33 polygenic scores, normalized
 to mean 0, variance 1. Figure \@ref(fig:plot-means-over-time) plots mean
@@
 educational attainment is more selected against, though the difference
-is only significant for EA3.
+is only significant for EA3. One possible reason for these sex differences is
+that polygenic scores may affect fertility via mate search.
 
@@
 
-```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth", fig.subcap = c("Number of partners", "Household type", "Age at first live birth"), fig.ncol = 1, out.height = c("35%", "27%", "27%"), fig.align = "center"}
+```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth", fig.subcap = c("Lifetime number of sexual partners", "Household type", "Age at first live birth, females"), fig.ncol = 1, out.height = c("32%", "32%", "25%"), fig.align = "center"}
+
+sexes_theme <-  my_theme(
+                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
+                  legend.position      = "bottom", 
+                  legend.box.spacing   = unit(0, "lines"),
+                  legend.box.margin    = margin(),
+                  legend.box.just      = "left",
+                  legend.justification = "left",
+                  legend.text          = element_text(size = 7),
+                  legend.title         = element_text(size = 8)
+                )
+sexes_guides <- guides(
+                  # fill  = guide_legend(title.position = "top"),
+                  # color = guide_legend(title.position = "top")
+                )
 
@@
         pval_scale() +
-        my_theme(legend.position = "bottom") +
+        sexes_theme +
         my_labs(x = "", y = "") + 
-        scale_y_reordered()
-
-drake::loadd(res_with_partner)
-res_with_partner %<>% filter(grepl(":", term))
-
-n_regs <- nrow(res_with_partner)
-hhtype_n <- n_in_regs("with_partner")
-
-res_with_partner %>% 
-      mutate(
+        scale_y_reordered() +
+        sexes_guides
+
+drake::loadd(res_with_partner_sex)
+res_with_partner_sex %<>% filter(grepl(":", term))
+
+n_regs <- as.double(nrow(res_with_partner_sex))
+
+res_with_partner_sex %>% 
+      mutate(
+        score_name = pretty_names(score_name),
         Household = ifelse(grepl("TRUE", term), "With partner", 
-                      "Without partner"),
-        with_p    = as.character(grepl("TRUE", term)),
-        Household = sprintf("%s (N = %s)", Household, 
-                      prettyNum(hhtype_n[with_p], big.mark = ","))
-      ) %>% 
-      standard_ggplot(fill = Household, n_regs = n_regs, score_col = score_name)
+                      "Without partner")
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(
+        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        x      = estimate, 
+        fill   = Household, 
+        colour = p < 0.05/{{n_regs}}
+      )) + 
+        my_vline +
+        pval_geom_point() +
+        facet_wrap(vars(sex), scales = "free_y") +
+        my_fill_scale(n = 2) +
+        pval_scale() +
+        sexes_theme +
+        my_labs(x = "", y = "") + 
+        scale_y_reordered() +
+        sexes_guides
 
@@
 against in the parents' generation, but selected for if we control for AFLB,
-which is the same pattern as for respondents.
+which is the same pattern we observe among the respondents' generation.
+
+```{r calc-siblings-children}
+sib_chn_cor <- cor(famhist$n_sibs, famhist$n_children, use = "complete")
+```
+
+Lastly, the siblings data lets us check for a "quantity-quality tradeoff" between
+number of children and number of grandchildren (for the parents' generation). We
+do not find any: in fact, the correlation between *number of siblings* and 
+*number of children* is `r sib_chn_cor`.
 
@@
 play an important role in natural selection. Substitution effects
-dominate income effects overall, in accordance with the widespread association 
+dominate income effects overall, which fits the empirical association 
 between income and lower fertility [@becker1960economic;@jones2006economic].
@@
 are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only
-`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$, suggesting that
+`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$. This suggests that
 earnings and education indeed mediate the effect of PGS on fertility.
@@
 decreases faster with education among single mothers than married mothers
-[@baudin2015fertility].If so, then earnings-increasing genetic variants may
-decrease fertility more among single parents than among couples, as in our data.
-The same idea could explain our results for lifetime number of sexual partners,
-if this is associated with relationship stability.
+[@baudin2015fertility]. This can explain why earnings-increasing genetic variants 
+decrease fertility more among single parents than among couples. The same logic 
+could explain our results for lifetime number of sexual partners, if this is 
+associated with relationship stability.
 
@@
 sexual partners. Indeed, among older parents selection effects are actually
-reversed. Furthermore, the size of selection effects correlates with the
-association between a PGS and labour market earnings. The economic theory of
-fertility provides a parsimonious explanation for this.
+reversed. Furthermore, the size of selection effects on a polygenic score
+correlates with that score's association with labour market earnings. The 
+economic theory of fertility provides a parsimonious explanation for this.
 
@@
 in selection across the full range of scores, including physical measures, e.g.
-waist-hip ratio and BMI. Externalizing behaviour also does not explain
+waist-hip ratio and BMI (see also appendix Figure \@ref(fig:plot-together)). 
+Externalizing behaviour also does not explain
 why selection effects change sign among older parents. Overall, we believe
@@
 selected against (selected for). In addition, many of the phenotypes under
-positive selection are linked to poor health, or would be seen as undesirable to
-have by many people. For example, most people would probably prefer to have high
-educational attainment, a low risk of major depressive disorder, and a low risk
-of coronary artery disease, but natural selection is pushing against these
-traits. Potentially, this could increase the health burden on modern societies,
-but that depends on effect sizes. Our results suggest that naïve estimates can
-be affected by sample ascertainment bias. This problem may be less serious in
-surveys which aim to be representative (as the UK Biobank does not).
-Nevertheless, not all respondents consent to the collection of genetic data: for
-instance, completion rates for genotype data in the US Health and Retirement
-Study were around 80-85% [@hrs2020]. Researchers should be aware of the risks
-of ascertainment when studying modern natural selection.
+positive selection are linked to poor health, or are what many people would see
+as undesirable to have. For example, most people would probably prefer to have
+high educational attainment, a low risk of major depressive disorder, and a low
+risk of coronary artery disease, but natural selection is pushing against
+genetics linked to these traits. Potentially, this could increase the health
+burden on modern populations, but that depends on effect sizes. Our results
+suggest that naïve estimates can be affected by sample ascertainment bias. This
+problem may be less serious in surveys which aim to be representative (as the UK
+Biobank does not). Nevertheless, not all respondents consent to the collection
+of genetic data: for instance, completion rates for genotype data in the US
+Health and Retirement Study were around 80-85% [@hrs2020]. Researchers should be
+aware of the risks of ascertainment when studying modern natural selection.
 
@@
 
-Research on natural selection is controversial. It brings to mind eugenics,
-which has been used to justify mass murder and human rights violations, in the
-past and possibly the present [@mao1998chinese; @sleeboom2011genetic;
-@zenz2020]. Whilst we claim scientific legitimacy for our empirical results, we
-have no special authority to speak on how society should react to
-them. This is a matter for open debate, in which scientists take part as
-equals among many. We therefore choose not to discuss policy implications here.
+Any model of fertility is implicitly a model of natural selection. But so far,
+the economic and human genetics literatures have developed in parallel.
+Integrating the two could deepen our understanding of modern natural selection.
+Economics possesses a range of theoretical models on the effects of skills,
+education and income [see @hotz1997economics, @lundberg2007american]. One
+perennial problem is how to test these theories in a world where education,
+labour and marriage markets all interact. Genetic data, such as polygenic
+scores, could help to pin down causality. Conversely, theory and empirical
+results from economics can shine a light on the mechanisms behind natural
+selection.
 

### 2020-11-18 _drake.R

@@ -147,7 +147,8 @@ plan <- drake_plan(
     res_chn <- run_regs_basic(
             dep_var     = "n_children", 
             score_names = score_names, 
-            famhist     = famhist
+            famhist     = famhist,
+            subset      = quote(age_at_recruitment >= 45)
           )
     dplyr::bind_rows(
             "N siblings" = res_sibs, 
@@ -166,7 +167,8 @@ plan <- drake_plan(
     res_chn_pcs <- run_regs_pcs(
       dep_var     = "n_children", 
       famhist     = famhist,
-      pcs         = pcs
+      pcs         = pcs,
+      subset      = quote(age_at_recruitment >= 45)
     )
     dplyr::bind_rows(
       "N siblings" = res_sibs_pcs, 
@@ -178,9 +180,10 @@ plan <- drake_plan(
   res_wt =  target(
               map_dfr(score_names, 
                 run_regs_weighted,
-                famhist = famhist,
+                famhist     = famhist,
                 weight_data = weighting_scheme,
-                dep.var = "n_children"
+                dep.var     = "n_children",
+                subset      = quote(age_at_recruitment >= 45)
               ),
               transform = map(
                 weighting_scheme = !! weighting_schemes
@@ -198,7 +201,8 @@ plan <- drake_plan(
                               run_regs_weighted,
                               famhist     = famhist %>% filter(n_children > 0),
                               weight_data = age_qual_weights,
-                              dep.var     = "n_children"
+                              dep.var     = "n_children",
+                              subset      = quote(age_at_recruitment >= 45)
                             ),
   
   res_period =  target(
@@ -224,13 +228,17 @@ plan <- drake_plan(
             score_name = score_names, 
             subset     = sexes
           ) %>% 
-            pmap_dfr(run_regs_subset, famhist = famhist)
+            pmap_dfr(
+              run_regs_subset, 
+              famhist = famhist %>% filter(age_at_recruitment >= 45)
+            )
     res_sex$sex <- ifelse(res_sex$subset == "sex == 0", "Female", "Male") 
     
     int_pval <- map_dfr(score_names, ~ run_regs_fml(
                     fml        = "n_children ~ {score_name}*I(sex==0)",
                     score_name = .x,
-                    famhist    = famhist
+                    famhist    = famhist,
+                    subset     = quote(age_at_recruitment >= 45)
                   ),
                   .id = "score_name"
                 ) %>% 
[…]

### 2020-11-19 why-negative-selection.Rmd
add manual checks of every result. Use Anova to check for interaction effects when controlling for age.

@@
   fh_subset <- famhist
-  if (dv == "n_children") fh_subset %<>% filter(age_at_recruitment >= 45)
+  if (dv == "n_children") fh_subset %<>% filter(kids_ss)
   # all PGS have the same NA pattern, so we just use one
@@
 
+* Limit max no. of children in regressions (use an "in_children" bool)
+  - not necessary. Very few have abnormally many.
 * Check each result with a manually done regression!
+* Consider doing separate regressions for "any children" vs. "n children | 1 or more"
+  - heckit too slow and gives crazy answers, just use plm/ols
+  - some variables do look different
 * If we weight with the GHS, and plot the means of figure 1 using this w
@@
 the interaction with age leaving FTE remains significant at
-0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ`
-regressions.
+0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.
 
@@
 res_partners_controlled %<>%       
-      filter(term != "(Intercept)", term != "lo_partnersTRUE") %>% 
-      mutate(`N partners` = ifelse(grepl("TRUE:", term), "<= 3", "> 3"))
+      mutate(`N partners` = ifelse(grepl("TRUE", term), "<= 3", "> 3"))
 

### 2020-11-19 _drake.R

@@ -46,6 +46,7 @@ weighting_schemes <- rlang::syms(c("flb_weights", "age_qual_weights",
 period_weight_syms <- rlang::syms((c("age_qual_weights", "parent_weights")))
 
 plan <- drake_plan(
+  
   score_names  = {
     score_names <- sub(
       ".*UKB\\.AMC\\.(.*?)\\..*", 
@@ -148,7 +149,7 @@ plan <- drake_plan(
             dep_var     = "n_children", 
             score_names = score_names, 
             famhist     = famhist,
-            subset      = quote(age_at_recruitment >= 45)
+            subset      = quote(kids_ss)
           )
     dplyr::bind_rows(
             "N siblings" = res_sibs, 
@@ -168,7 +169,7 @@ plan <- drake_plan(
       dep_var     = "n_children", 
       famhist     = famhist,
       pcs         = pcs,
-      subset      = quote(age_at_recruitment >= 45)
+      subset      = quote(kids_ss)
     )
     dplyr::bind_rows(
       "N siblings" = res_sibs_pcs, 
@@ -183,7 +184,7 @@ plan <- drake_plan(
                 famhist     = famhist,
                 weight_data = weighting_scheme,
                 dep.var     = "n_children",
-                subset      = quote(age_at_recruitment >= 45)
+                subset      = quote(kids_ss)
               ),
               transform = map(
                 weighting_scheme = !! weighting_schemes
@@ -202,7 +203,7 @@ plan <- drake_plan(
                               famhist     = famhist %>% filter(n_children > 0),
                               weight_data = age_qual_weights,
                               dep.var     = "n_children",
-                              subset      = quote(age_at_recruitment >= 45)
+                              subset      = quote(kids_ss)
                             ),
   
   res_period =  target(
@@ -230,7 +231,7 @@ plan <- drake_plan(
           ) %>% 
             pmap_dfr(
               run_regs_subset, 
-              famhist = famhist %>% filter(age_at_recruitment >= 45)
+              famhist = famhist %>% filter(kids_ss)
             )
     res_sex$sex <- ifelse(res_sex$subset == "sex == 0", "Female", "Male") 
     
@@ -238,7 +239,7 @@ plan <- drake_plan(
                     fml        = "n_children ~ {score_name}*I(sex==0)",
                     score_name = .x,
                     famhist    = famhist,
-                    subset     = quote(age_at_recruitment >= 45)
+                    subset     = quote(kids_ss)
                   ),
                   .id = "score_name"
                 ) %>% 
[…]

### 2020-11-19 manual-result-tests.R

@@ -1,0 +1,419 @@
+
+# manual tests for each result
+
+library(drake)
+library(dplyr)
+library(testthat)
+
+loadd(score_names)
+loadd(famhist)
+
+lm_coef <- function (fml, target = 2, data = famhist %>% filter(kids_ss)) {
+  mod <- lm(fml, data = data)
+  coef(mod)[target]
+}
+
+estimate_for <- function (data, score_name, term) {
+  if (! missing(term)) data$estimate[data$term == term] else 
+        data$estimate[data$score_name == score_name]
+}
+
+test_that("res_all", {
+  loadd(res_all)
+  loadd(resid_scores)
+  
+  expect_equivalent(
+    res_all %>% 
+      filter(dep.var == "N children", reg.type == "raw") %>% 
+      estimate_for("ADHD_2017"),
+    lm_coef(n_children ~ ADHD_2017)
+  )
+  
+  expect_equivalent(
+    res_all %>% 
+      filter(dep.var == "N siblings", reg.type == "raw") %>% 
+      estimate_for("alcohol_schumann"),
+    lm_coef(n_sibs ~ alcohol_schumann, data = famhist)
+  )
+  
+  
+  fh2 <- cbind(famhist, resid_scores)
+  expect_equivalent(
+    res_all %>% 
+      filter(dep.var == "N children", reg.type == "controlled") %>% 
+      estimate_for("height_combined"),
+    lm_coef(n_children ~ height_combined_resid, data = fh2 %>% filter(kids_ss))
+  )
+  
+  expect_equivalent(
+    res_all %>% 
+      filter(dep.var == "N siblings", reg.type == "controlled") %>% 
+      estimate_for("height_combined"),
+    lm_coef(n_sibs ~ height_combined_resid, data = fh2)
+  )
+})
+
+
+test_that("res_pcs", {
+  loadd(res_pcs)
+  loadd(pcs)
+  fh <- left_join(famhist, pcs, by = c("f.eid" = "IID"))
+  expect_equivalent(
+    res_pcs %>% filter(dep.var == "N children", term == "PC33") %>% pull(estimate),
+    lm_coef(n_children ~ PC33, data = fh %>% filter(kids_ss))
+  )
+  
+  expect_equivalent(
+    res_pcs %>% filter(dep.var == "N siblings", term == "PC45") %>% pull(estimate),
+    lm_coef(n_sibs ~ PC45, data = fh)
+  )
+})
+
+
+test_that("res_wt_flb_weights", {
[…]

### 2020-11-19 regressions.R

@@ -7,7 +7,9 @@ suppressPackageStartupMessages({
   library(tidyr)
   library(dplyr)
   library(lmtest)
+  loadNamespace("tibble")
   loadNamespace("ggplot2")
+  loadNamespace("sampleSelection")
 })
 
 calc_pgs_over_time <- function (famhist, score_names) {
@@ -77,7 +79,7 @@ run_regs_basic <- function (dep_var, score_names, famhist, subset = NULL) {
 run_regs_pcs <- function (dep_var, famhist, pcs, subset = NULL) {
   pc_names <- grep("PC", names(pcs), value = TRUE)
   
-  fh_short <- join_famhist_pcs(famhist[c("f.eid", dep_var, "age_at_recruitment")], pcs)
+  fh_short <- join_famhist_pcs(famhist[c("f.eid", dep_var, "kids_ss")], pcs)
   
   run_reg_pc <- function(pc_name) {
     fml <- as.formula(glue("{dep_var} ~ {pc_name}"))
@@ -108,7 +110,7 @@ run_regs_period <- function (children, score_name, famhist, weight_data) {
   famhist <- inner_join(famhist, weight_data, by = "f.eid")
   
   dep_var  <- if (children) "n_children" else "n_sibs"
-  subset   <- if (children) quote(age_at_recruitment >= 45) else NULL
+  subset   <- if (children) quote(kids_ss) else NULL
   famhist$year_split <- famhist$YOB >= 1950
   # evaluates to 1950:
   # median(famhist$YOB, na.rm = TRUE)
@@ -152,7 +154,7 @@ run_regs_fml <- function(fml, ..., subset = NULL, famhist) {
 }
 
 
-run_regs_mnlogit <- function(score_names, fhl_mlogit) {
+run_regs_mnlogit <- function (score_names, fhl_mlogit) {
   loadNamespace("mnlogit")
   run_1_mnlogit <- function(score_name) {
     fml <- as.formula(glue::glue("n_ch_fac ~ 1 | {score_name}"))
@@ -164,3 +166,12 @@ run_regs_mnlogit <- function(score_names, fhl_mlogit) {
   lapply(score_names, run_1_mnlogit)
 }
 
+
+run_age_anova <- function (score_name, control, famhist) {
+  fml <- as.formula(glue::glue(
+    "n_children ~ {score_name}*({control} + 
+          age_at_recruitment + I(age_at_recruitment^2))"))
[…]

### 2020-11-19 why-negative-selection.Rmd
remove partners-controlled regression until you can figure out how to do it right

@@
 For both sexes, selection effects are larger and more significant among
-those with more than 3 partners. Results are unchanged if we control for
-respondents' age (appendix Figure
-\@ref(fig:plot-n-partners-controlled)). Splitting respondents by whether
+those with more than 3 partners. Splitting respondents by whether
 they were living with a spouse or partner at the time of interview,
@@
 
-## Selection by number of partners, controlling for age
-
-Figure \@ref(fig:plot-n-partners-controlled) shows selection effects by
-lifetime number of sexual partners. Age, age squared, and interactions
-of age and age squared with PGS are controlled for.
-
-```{r plot-n-partners-controlled, fig.cap = "Selection effects by number of sexual partners (median split at 3), controlling for respondent's age and age squared."}
-drake::loadd(res_partners_controlled)
-res_partners_controlled %<>%       
-      mutate(`N partners` = ifelse(grepl("TRUE", term), "<= 3", "> 3"))
-
-n_regs <- as.double(nrow(res_partners_controlled))
-res_partners_controlled %>% 
-      mutate(
-        score_name = pretty_names(score_name)
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(
-        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        x      = estimate, 
-        fill   = `N partners`, 
-        colour = p < 0.05/{{n_regs}}
-      )) + 
-        my_vline +
-        pval_geom_point() +
-        facet_wrap(vars(sex), scales = "free_y") +
-        my_fill_scale(n = 2) +
-        pval_scale() +
-        my_theme(legend.position = "bottom") +
-        my_labs(x = "", y = "") + 
-        scale_y_reordered()
-
-```
-
 ## Selection controlling for age at first live birth: respondents' parents

### 2020-11-19 _drake.R

@@ -497,28 +497,6 @@ plan <- drake_plan(
     res %>% filter(grepl(":factor\\(age_fte_cat\\)", term))
   },
   
-  res_partners_controlled = {
-    sexes <- rlang::exprs(sex == 0, sex == 1)
-    pars <- expand_grid(subset = sexes, score_name = score_names)
-    res <- pmap_dfr(pars,
-            ~ run_regs_fml(
-              fml = "n_children ~ age_at_recruitment + I(age_at_recruitment^2) + lo_partners +
-                {score_name}:lo_partners + 
-                {score_name}:(age_at_recruitment + I(age_at_recruitment^2))",
-              score_name = .y,
-              famhist    = famhist %>% filter(kids_ss),
-              subset     = .x
-            ),
-              .id = "row_number"
-          )
-    pars$row_number <- as.character(seq_len(nrow(pars)))
-    
-    left_join(res, pars, by = "row_number") %>%
-          mutate(sex = ifelse(subset == "sex == 1", "Male", "Female")) %>%
-          select(-row_number, -subset) %>% 
-          filter(grepl("lo_partners.*:", term))
-  },
-  
   res_ee_control = {
     map_dfr(score_names, 
               ~ run_regs_fml(

### 2020-11-19 why-negative-selection.Rmd
correct text

@@
 change in natural selection within the parents' generation. In the
-respondents' generation, effect sizes were significantly larger in absolute size among the late-born for five PGS: autism, cognitive ability, EA2, EA3 and
+respondents' generation, effect sizes were significantly larger in absolute size among the late-born for four PGS: cognitive ability, EA2, EA3 and
 extraversion. These changes are inconsistent with the intergenerational change, 

### 2020-11-30 why-negative-selection.Rmd
Use shared infrastructure from import-ukbb-data; add regression controlling for risk attitude

@@
     wrap: 72
+  chunk_output_type: console
 ---
@@
 capital, i.e. skills and characteristics that are valuable in the labour market.
-Suppose also that education and innate human capital are complements. They
-choose how long to stay in education. After leaving education they enter the
+Suppose also that education and innate human capital are complements. Individuals
+choose how long to stay in education; after leaving education they enter the
 labour market and form families. If so, then people with less human capital will
@@
 
+```{r calc-risk-attitude}
+
+drake::loadd(res_risk_control)
+drake::loadd(res_all)
+
+n_risk_sig <- res_risk_control %>% 
+            filter(term == 'f.2040.0.0') %>% 
+            {sum(.$p.value <= 0.05/33)}
+
+n_pgs_sig <- res_risk_control %>% 
+            filter(term == score_name) %>% 
+            {sum(.$p.value <= 0.05/33)}
+
+res_compare <- res_all %>% 
+  filter(dep.var=='N children', reg.type == 'raw') %>% 
+  left_join(res_risk_control %>% filter(term == score_name), by = 'score_name')
+
+ratio <- res_compare$estimate.y/res_compare$estimate.x
+n_less_sig <- sum(res_compare$p.value.y <= 0.05/33 & res_compare$p.value.x > 0.05/33)
+```
+
+An alternative theory is that traits selected for are linked to externalizing
+behaviour, risk-seeking and low time discount rates, via the channel of early
+sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
+for this: scores which might plausibly be linked to externalizing behaviour,
+like ADHD and age at smoking initiation, are selected for. However, this theory
+is less good at explaining variation in selection across the full range of
+scores, including physical measures, e.g. waist-hip ratio and BMI (see also
+appendix Figure \@ref(fig:plot-together)). Externalizing behaviour also does not
+explain why selection effects change sign among older parents.
+
+We test the alternative theory directly by re-estimating equation
+\@ref(eq:regression) controlling for a measure of risk attitude (Biobank field
+2040). While risk attitude is always a highly significant predictor of number of
+children, it has little impact on the effect size or significance of the PGS.
+The median ratio of PGS effect sizes between regressions with and without
+controls is `r median(ratio)`; all PGS which are significant at $p \le 0.05/33$
+in uncontrolled regressions remain so when controlling for risk attitude. Thus,
+although risk attitude does predict number of children, it appears not to
+mediate selection effects. Overall, we believe that the economic theory is the
+leading explanation.
+
 # Discussion
@@
 economic theory of fertility provides a parsimonious explanation for this.
-
-An alternative theory is that traits selected for are linked to
-externalizing behaviour and low time discount rates, via the channel of
-early sexual behaviour [@Mills2020.05.06.081273]. The data here provide
-some support for this: scores which might plausibly be linked to
-externalizing behaviour, like ADHD and age at smoking initiation, are
-selected for. However, this theory is less good at explaining variation
-in selection across the full range of scores, including physical measures, e.g.
-waist-hip ratio and BMI (see also appendix Figure \@ref(fig:plot-together)). 
-Externalizing behaviour also does not explain
-why selection effects change sign among older parents. Overall, we believe
-that the economic theory is the leading explanation.
 

### 2020-11-30 _drake.R

@@ -5,29 +5,17 @@ suppressPackageStartupMessages({
   library(purrr)
   library(tidyr)
   library(readr)
+  loadNamespace("future")
 })
 
-
-source("make-data.R")
+source("make-data.R") 
+# infrastructure available at github.com/hughjonesd-private/import-ukbb-data
+source("~/import-ukbb-data/import-ukbb-data.R")
 source("make-census-weights.R")
 source("regressions.R")
 source("weight-data.R")
 
-data_dir           <- "../negative-selection-data"
-pgs_dir            <- file.path(data_dir, "polygenic_scores/")
-ph_file            <- file.path(data_dir, "UKB.EA_pheno.coordinates.QC.csv")
-pcs_file           <- file.path(data_dir, "UKB.HM3.100PCs.40310.txt")
-famhist_file       <- file.path(data_dir, "david.family_history.traits.out.csv")
-famhist2_file      <- file.path(data_dir, "david.family_history.traits.20042020.out.csv")
-famhist3_file      <- file.path(data_dir, "david.family_history.traits.05052020.out.csv")
-famhist4_file      <- file.path(data_dir, "david.family_history.traits.16052020.out.csv")
-famhist5_file      <- file.path(data_dir, "david.family_history.traits.18052020.out.csv")
-famhist6_file      <- file.path(data_dir, "david.family_history.traits.17062020.out.csv")
-famhist7_file      <- file.path(data_dir, "david.birthinfo.traits.14072020.out.csv")
-famhist8_file      <- file.path(data_dir, "david.traits.03112020.out.csv")
-rgs_file           <- file.path(data_dir, "EA3_rgs.10052019.rgs.csv")
-mf_pairs_file      <- file.path(data_dir, "spouse_pair_info", 
-                        "UKB_out.mf_pairs_rebadged.csv")
+
 nomis_file         <- file.path(data_dir, "nomis-2011-statistics.csv")
 ghs_file           <- file.path(data_dir, "UKDA-5804-stata8", "stata8", 
                         "Ghs06client.dta")
@@ -36,85 +24,73 @@ DC1108EW_msoa_file <- file.path(data_dir, "DC1108EW-msoas.csv")
 DC5202SC_file      <- file.path(data_dir, "DC5202SC.csv")
 msoa_shapefile     <- file.path(data_dir, "infuse_msoa_lyr_2011_clipped", 
                         "infuse_msoa_lyr_2011_clipped.shp")
[…]

### 2020-11-30 make-census-weights.R

@@ -123,7 +123,7 @@ find_containing_msoas <- function (famhist, msoa_shapefile) {
 make_census_msoa <- function (census_msoa_file) {
   census_msoa <-  readr::read_csv(census_msoa_file) 
   census_msoa %<>% 
-        select(
+        dplyr::select(
           -matches("Total"), 
           -matches("All categories"), 
           -matches("All persons")

### 2020-11-30 make-data.R

@@ -4,9 +4,9 @@ suppressPackageStartupMessages({
   library(magrittr)
   library(dplyr)
   library(rlang)
-  library(santoku)
   library(haven)
   library(sf)
+  loadNamespace("santoku")
   loadNamespace("ggplot2")
   loadNamespace("car") # very annoying if it overwrites recode
   loadNamespace("matrixStats")
@@ -14,96 +14,14 @@ suppressPackageStartupMessages({
   loadNamespace("readxl")
 })
 
-# utility function:
-negative_to_na <- function (x) {
-  x[x < 0] <- NA
-  x
-}
-
-make_pcs <- function (pcs_file) {
-  pcs <- read_table(pcs_file)
-  pc_names <- grep("PC", names(pcs), value = TRUE)
-  pcs[pc_names] <- scale(pcs[pc_names])
-  pcs
-}
-
 
 join_famhist_pcs <- function (famhist, pcs) {
-  left_join(famhist, pcs, by = c("f.eid" = "IID"))
+  left_join(famhist, pcs, by = c("f.eid" = "eid"))
 }
 
 
 join_famhist_resid_scores <- function (famhist, resid_scores) {
   cbind(famhist, resid_scores)
-}
-
-
-make_resid_scores <- function (famhist, pcs, score_names) {
-  famhist <- join_famhist_pcs(famhist, pcs)
-  resid_scores <- data.frame(dummy = numeric(nrow(famhist))) 
-  
-  for (score_name in score_names) {
-    resid_fml <- paste(score_name, "~", paste0("PC", 1:40, collapse = " + "))
-    resid_score <- resid(lm(as.formula(resid_fml), famhist, 
-          na.action = na.exclude))
-    resid_scores[[paste0(score_name, "_resid")]] <- resid_score
-  }
-  
-  resid_scores$dummy <- NULL
-  resid_scores
-}
-
-
-make_famhist <- function (
-        ph_file, 
-        famhist_file,
-        famhist2_file,
-        famhist3_file,
-        famhist4_file,
-        famhist5_file,
-        famhist6_file,
-        famhist7_file,
-        famhist8_file,
-        pgs_dir
-      ) {
-  ph <- read_csv(ph_file, col_types = cols(
-    geno_measurement_plate = col_skip(),
-    geno_measurement_well = col_skip(),
-    .default = col_double()
-  ))
[…]

### 2020-11-30 regressions.R

@@ -14,6 +14,7 @@ suppressPackageStartupMessages({
 
 calc_pgs_over_time <- function (famhist, score_names) {
   pgs_over_time <- famhist %>% 
+        select(YOB, all_of(score_names)) %>% 
         group_by(
           YOB = santoku::chop(YOB, 
             seq(1940, 1970, 5), 
@@ -22,7 +23,9 @@ calc_pgs_over_time <- function (famhist, score_names) {
           ) 
         ) %>% 
         filter(! is.na(YOB)) %>% 
-        summarize(across(all_of(score_names), ggplot2::mean_cl_normal, na.rm = TRUE)) %>% 
+        summarize(across(all_of(score_names), 
+          ggplot2::mean_cl_normal, na.rm = TRUE
+        )) %>% 
         pivot_longer(-YOB, names_to = "score_name", values_to = "score") %>% 
         mutate(
           score_lo = score$ymin,

### 2020-12-01 why-negative-selection.Rmd
Add townsend deprivation regressions for parents

@@
 
-n_in_regs <- function (var, dv = "n_children") {
-  fh_subset <- famhist
+n_in_regs <- function (var, dv = "n_children", data = famhist) {
+  fh_subset <- data
   if (dv == "n_children") fh_subset %<>% filter(kids_ss)
@@
 
-add_n <- function (var, famhist_var = var, dv = "n_children") {
+add_n <- function (var, famhist_var = var, dv = "n_children", data = famhist) {
   var <- as.factor(var)
-  n <- n_in_regs(famhist_var, dv)
+  n <- n_in_regs(famhist_var, dv, data)
   levels(var) <- sprintf("%s (N = %s)", levels(var), 
@@
 
+```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by Townsend deprivation score quintile", fig.align = "center"}
+
+drake::loadd(res_townsend_parents)
+drake::loadd(famhist_townsend_71)
+n_regs <- nrow(res_townsend_parents)
+
+res_townsend_parents %>% 
+  mutate(
+    `Townsend quintile` = sub("Quin71(\\d):.*", "\\1", term),
+    add_n(`Townsend quintile`, "Quin71", dv = "n_sibs", data = famhist_townsend_71)
+  ) %>% 
+  standard_ggplot(`Townsend quintile`, n_regs = n_regs, score_col = score_name, order_idx = 5)
+
+rm(famhist_townsend_71)
+```
+
+We also check whether selection effects differ by socio-economic status in the
+parents' generation. We have no information about parents' income, so we use
+the 1971 Townsend deprivation score of respondents' birthplace as a proxy. Results
+(Figure \@ref(fig:plot-siblings-townsend)) show the same pattern as for respondents:
+effect sizes are larger and more significant in the most deprived areas.
+
 ```{r calc-siblings-children}
@@
 results. Effect sizes are very similar, whether controlling for father's
-or mother's age. Unlike for the respondents' generation, effect sizes
-are positively correlated with the effect sizes from bivariate
-regressions without the age at birth control (father's age at birth:
-$\rho$ `r cor_raw_fath`; mother's age at birth: $\rho$
-`r cor_raw_moth`).
+or mother's age. As in the respondents' generation, effect sizes
+are negatively correlated with the effect sizes from bivariate
+regressions without the age at birth control, though the correlation
+is smaller (father's age at birth: $\rho$ `r cor_raw_fath`; mother's age at
+birth: $\rho$ `r cor_raw_moth`).
 

### 2020-12-01 _drake.R

@@ -24,7 +24,7 @@ DC1108EW_msoa_file <- file.path(data_dir, "DC1108EW-msoas.csv")
 DC5202SC_file      <- file.path(data_dir, "DC5202SC.csv")
 msoa_shapefile     <- file.path(data_dir, "infuse_msoa_lyr_2011_clipped", 
                         "infuse_msoa_lyr_2011_clipped.shp")
-
+dep_data_dir       <- file.path(data_dir, "deprivation-data")
 
 weighting_scheme_syms <- rlang::syms(c("flb_weights", "age_qual_weights", 
                         "msoa_weights"))
@@ -63,6 +63,11 @@ plan <- drake_plan(
                    find_containing_msoas(famhist, msoa_shapefile), 
                    format = "fst_tbl"
                  ),
+  
+  famhist_townsend_71 = target(
+                          add_deprivation_data(famhist, dep_data_dir),
+                          format = "fst_tbl"
+                        ),
   
   # mlogit now uses dfidx to create a "long" dataset, but mnlogit 
   # (used below) doesn't update. Skipping for now.
@@ -287,6 +292,19 @@ plan <- drake_plan(
     res$id <- NULL
     res %<>% filter(term != "(Intercept)")
     
+    res
+  },
+  
+  res_townsend_parents = {
+    res <-  map_dfr(score_names,
+              ~run_regs_fml(
+                "n_sibs ~ Quin71 + {score_name}:Quin71",
+                score_name = .x,
+                famhist    = famhist_townsend_71
+              ),
+              .id = "score_name"
+            )
+    res %<>% filter(grepl(":", term))
     res
   },
   

### 2020-12-01 why-negative-selection.Rmd
Tweak _drake; add within-family regression on income (unused at present); add drake-runner script for job; remove parallelism again

@@
 
-```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by Townsend deprivation score quintile", fig.align = "center"}
+```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by area Townsend deprivation quintile", fig.align = "center"}
 
@@
     `Townsend quintile` = sub("Quin71(\\d):.*", "\\1", term),
-    add_n(`Townsend quintile`, "Quin71", dv = "n_sibs", data = famhist_townsend_71)
-  ) %>% 
-  standard_ggplot(`Townsend quintile`, n_regs = n_regs, score_col = score_name, order_idx = 5)
+    `Townsend quintile` = add_n(`Townsend quintile`, "Quin71", dv = "n_sibs", 
+                                  data = famhist_townsend_71)
+  ) %>% 
+  standard_ggplot(fill_col    = `Townsend quintile`,
+                    n_regs    = n_regs, 
+                    score_col = score_name, 
+                    order_idx = 5
+                  )
 
@@
 We also check whether selection effects differ by socio-economic status in the
-parents' generation. We have no information about parents' income, so we use
-the 1971 Townsend deprivation score of respondents' birthplace as a proxy. Results
-(Figure \@ref(fig:plot-siblings-townsend)) show the same pattern as for respondents:
-effect sizes are larger and more significant in the most deprived areas.
+parents' generation. We have no information about parents' income, so we use the
+1971 Townsend deprivation score of respondents' birthplace as a proxy
+[@townsend1987deprivation]. Results (Figure \@ref(fig:plot-siblings-townsend))
+show the same pattern as for respondents: effect sizes are larger and more
+significant in the most deprived areas.
 
@@
 drake::loadd(res_age_flb)
+drake::loadd(ashe_income)
+
+famhist <- add_ashe_income(famhist, ashe_income)
 
@@
 
-```{r plot-res-earn-educ, fig.cap = "Selection effects controlling for earnings in first job (estimated by median earnings from ASHE 2007 for the SOC2000 job code) and education (left education before 16, 16-18, or after 18).
+```{r plot-res-earn-educ, fig.cap = "Selection effects controlling for earnings in first job (estimated by mean earnings from ASHE 2007 for the SOC2000 job code) and education (left education before 16, 16-18, or after 18).
 Raw effects are shown for comparison.

### 2020-12-01 drake-runner.R

@@ -1,0 +1,1 @@
+drake::r_make()

### 2020-12-01 regressions.R

@@ -10,6 +10,8 @@ suppressPackageStartupMessages({
   loadNamespace("tibble")
   loadNamespace("ggplot2")
   loadNamespace("sampleSelection")
+  loadNamespace("Formula")
+  loadNamespace("fixest")
 })
 
 calc_pgs_over_time <- function (famhist, score_names) {
@@ -177,4 +179,41 @@ run_age_anova <- function (score_name, control, famhist) {
   res <- lm(fml, data = famhist, subset = kids_ss) %>% 
            car::Anova() %>% 
            tidy()
+}
+
+
+run_reg_income_dv <- function (famhist, score_names, ashe_income, age_qual_weights) {
+  famhist %<>% add_ashe_income(ashe_income)
+  famhist %<>% left_join(age_qual_weights, by = "f.eid")
+  
+  most_score_names <- setdiff(score_names, c("EA2_noUKB", "whr_combined", 
+                                "wc_combined", "hip_combined", "body_fat"))
+  
+  fml <- paste(most_score_names, collapse = " + ")
+  fml <- glue::glue("log(cur_job_pay) ~ age_at_recruitment + ",
+                      "I(age_at_recruitment^2) + {fml} | factor(sib_group)")
+  fml <- Formula::as.Formula(fml)
+  
+  res <- fixest::feols(fml, data = famhist, weights = ~ weights)
+  
+  # predict manually, using causal coefficients from within-family regs
+  # and choosing intercept to minimize squared error
+  scores <- famhist %>% 
+              mutate(agesq = age_at_recruitment^2) %>% 
+              select(age_at_recruitment, agesq, all_of(most_score_names)) %>% 
+              as.matrix()
+  pay_hat <- c(scores %*% coef(res))
+  alpha <- mean(log(famhist$cur_job_pay) - pay_hat, na.rm = TRUE)
+  pay_hat <- pay_hat + alpha
+  
+  famhist$pay_hat <- pay_hat
+}
+
+run_income_dist <- function (famhist) {
+  # predict individual income from PGS
+  # 3 densities:
+  # - empirically from last_job_pay, weighted by e.g. age_qual
+  # - predicted from reg of last_job_pay on PGS, same weighting
+  #   - use within-family regressions, then work out the intercept separately
+  # - predicted from regression, weighting multiplied by n_children
 }

### 2020-12-01 why-negative-selection.Rmd
Add plot of mean PSEA by income; remove "causality" from appendix; nicer font

@@
   chunk_output_type: console
+mainfont: Times
 ---
@@
   - not necessary. Very few have abnormally many.
-* Check each result with a manually done regression!
 * Consider doing separate regressions for "any children" vs. "n children | 1 or more"
@@
 * Reweight other regressions (e.g. the omnibus regression in the appendix)?
-* Look at geography, esp. in siblings regressions 
-  - maybe Townshend f.189? Not sure essential
 * Another point is that negative selection only among poor will increase
     genetic inequality.
-* When we have better definition of "partner", bring it across from the other
-  project to define with_partner
-  - `with_partner` is fine, but if you want to have pairs who had the same children
-  you need to do more
+  - one way to look at this: regress PGS on e.g. income or Quin71
+  - then weight by n_children and see the change
+  - needs to be comparable in some sense though, across generations
+  - e.g. quintile of where born, unweighted; vs quintile of where now,
+    weighted by children
+  - perhaps simpler is just to run regressions with/without weights
+    by n_children. This gives you the "effect" of natural selection
 * If you take "year of last job started" and see if it is after/before
@@
 In short, it is probably too early to tell whether modern natural
-selection is substantively important or not.
+selection has a substantively important effect on populations' mean 
+polygenic scores.
+
+Because selection effects are concentrated in poorer populations, they also increase
+inequality with respect to polygenic scores. For example, Figure 
+\@ref(fig:plot-mean-EA3-by-income) graphs children's mean polygenic scores for 
+educational attainment (EA3), in households of different income groups. The blue
+bars show the actual means. The grey bars show the hypothetical means if all 
+households had equal numbers of children. Natural selection against EA3, which
+lowers the mean, is stronger at the bottom of the income distribution, and this
+increases the differences between groups.
+
+```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean PGS for educational attainment (EA3) of children by household income group.
+Blue is actual.
+Grey is hypothetical in the absence of selection effects.", fig.width = 4}
+income_EA3 <- famhist %>% 
+      filter(! is.na(n_children), ! is.na(income_cat)) %>% 
+      mutate(
+        `Household income` = factor(income_cat, 
+                          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", 
+                                      "> £100K"))
+      ) %>% 
+      group_by(`Household income`) %>% 
+      summarize(
+        `Without selection` = mean(EA3_excl_23andMe_UK, na.rm = TRUE),
+        Actual              = weighted.mean(EA3_excl_23andMe_UK, n_children, 
+                                              na.rm = TRUE)
+      ) %>% 
+      tidyr::pivot_longer(-`Household income`) %>% 
+      rename(
+        `Children's mean EA3`     = value
+      )
+
+income_EA3 %>% 
+              ggplot(aes(`Household income`, `Children's mean EA3`, fill = name)) +
+                geom_col(position = "dodge") +
+                scale_fill_manual(values = c("navy", "grey60")) +
+                theme(
+                  legend.background = element_rect(fill = "white", colour = NA),
+                  legend.title = element_blank(), 
+                  legend.position = c(.25, .85)
+                )
+```
 
@@
 
-## Causality
-
-Different polygenic scores are correlated. Table
-\@ref(tab:tbl-pgs-corrs) shows the top correlations in the sample.
-Because of this, bivariate correlations between PGS and number of
-children might be driven by other genetic scores. To explore which
-polygenic scores are driving negative selection, we run a single omnibus
-regression of *number of children* on all the PGS. We exclude EA2,
-waist-hip ratio, waist-circumference, and "Hip combined" since they are
-highly correlated with other scores, which could make our estimates
-unstable. Figure \@ref(fig:plot-together) shows the results.
-Interestingly, several PGS remain independently significant, although
-effect sizes are reduced.
-
-```{r tbl-pgs-corrs}
-cor_tbl <- cor(famhist[score_names], use = "pair") %>% 
-      as_tibble() %>% 
-      mutate(score_name = score_names, .before = 1) %>% 
-      pivot_longer(cols = -score_name, values_to = "Correlation") %>% 
-      filter(score_name < name) %>% 
-      arrange(desc(Correlation)) %>% 
-      slice(1:10) %>% 
-      mutate(across(c(score_name, name), pretty_names))
-
-as_hux(cor_tbl, add_colnames = TRUE) %>% 
-      set_contents(1, 1:2, c("PGS", "PGS")) %>% 
-      theme_basic(header_col = FALSE) %>% 
-      set_bottom_border(1, everywhere, 0.4) %>% 
-      set_bottom_border(final(1), everywhere, 0.4) %>% 
-      set_tb_padding(4) %>% 
-      set_caption("Top 10 correlations between polygenic scores")
-
-```
-
-```{r plot-together, fig.cap = "Partial correlations of PGS with fertility, from a single regression."}
-
-loadd(res_together)
-n_regs <- as.double(nrow(res_together))
-res_together %>% 
-      mutate(
-        term = pretty_names(term),
-        dep.var = dplyr::recode(dep.var, 
-          "n_children" = "N Children", "n_sibs" = "N Siblings"),
-        dep.var = fct_relevel(dep.var, "N Siblings")
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(y = reorder_within(term, estimate, dep.var, order_abs(1)), 
-        x = estimate, color = p < 0.05/{{n_regs}}, fill  = PCs)) + 
-        pval_geom_point(alpha = 0.8) +
-        my_vline +
-        facet_wrap(vars(dep.var), scales = "free_y") +
-        pval_scale() +
-        my_fill_scale(n = 2) +
-        my_labs() + 
-        scale_y_reordered() +
-        theme(legend.position = "bottom")
-```
-
 \FloatBarrier

### 2020-12-01 why-negative-selection.Rmd
trivia

@@
 own birth, and on own age at first live birth, have a correlation of
-`r cor_flb_fath`; for mother's age and own age it is `r cor_flb_moth`
+`r cor_flb_fath`; for mother's age and own age it is `r cor_flb_moth`.
 

### 2020-12-01 why-negative-selection.Rmd
tweaks and fixes

@@
 library(huxtable)
+
+# nasty hack to make add_ashe_income work later
+source("~/import-ukbb-data/import-ukbb-data.R")
 
@@
 
-```{r plot-means-over-time, fig.height = 5.5, fig.cap = "Mean polygenic scores by birth year in UK Biobank.
+```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores by birth year in UK Biobank.
 Points are means for 5-year intervals.
 Lines are 95% confidence intervals.
-Green lines show a significant linear increase over time (p < 0.05/33).
-Red lines show a significant decrease."}
+Green triangles show a significant linear increase over time (p < 0.05/33).
+Red squares show a significant decrease."}
 
@@
       mutate(score_name = pretty_names(score_name)) %>% 
-      ggplot(aes(YOB, group = score_name, colour = Change)) + 
+      ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
       geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
@@
       scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
+      scale_shape_manual(values = c("-" = "square", "+" = "triangle", "o" = "circle open")) +
       theme(
@@
 may also shift respondents between the subgroups. For example, a high
-PGS for educational attainment may correlate with fewer children among
-early school leavers, but may also increase the age at which a
+PGS for educational attainment may predict having fewer children among
+early school leavers, but it may also increase the age at which a
 respondent leaves school.
 
-```{r plot-income-educ-level, fig.cap = "Selection effects by education and income", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}
+```{r plot-income-educ-level, fig.cap = "Selection effects by education and income.
+Each point represents a single bivariate regression of number of children on a PGS", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}
 
@@
                   legend.box.spacing   = unit(0, "lines"),
-                  legend.box.margin    = margin(),
+                  legend.box.margin    = margin(-5, 0, 0, -5),
+                  legend.margin        = margin(0, 0, 0, 0),
                   legend.box.just      = "left",
@@
         scale_colour_brewer(type = "qual", palette = 2) +
-        theme(legend.position = "none")
+        theme(
+          legend.position = "none",
+          panel.spacing = unit(14, "pt")
+        )
 
@@
 
-```{r plot-res-wt, fig.cap = "Selection effects: weighted regressions."}
+```{r plot-res-wt, fig.cap = "Selection effects: weighted regressions"}
 
@@
     "Age/Qual/AFLB" = "circle filled"
-  ))
+  )) + 
+  labs(x = "Effect size", y = "")
 
@@
 polygenic scores. For comparison, we show regressions on the respondents'
-generation (*number of children*), excluding respondents with no
-children.
+generation (*number of children*), where respondents with no
+children have been excluded.
 

### 2020-12-08 why-negative-selection.Rmd
Work on text. Add summary of inequality effects.

@@
 ---
-title: "Economic Theory of Fertility Explains Modern Natural Selection"
+title: "Labor Market Income and Substitution Effects Explain Modern Natural Selection"
 author: "David Hugh-Jones (D.Hugh-Jones@uea.ac.uk), Abdel Abdellaoui (a.abdellaoui@gmail.com)"
@@
 We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank.
-Selection effects appear consistent over time.
-Polygenic scores which predict lower earnings and education are selected for.
-Effects are concentrated among low-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner.
-Overall effect sizes are substantially larger after correcting for ascertainment bias.
+Consistently over time, polygenic scores which predict lower (higher) earnings and education are selected for (against).
+Selection effects are concentrated among low-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner.
+Effect sizes are substantially larger after correcting for ascertainment bias.
 The direction of natural selection is reversed among older parents, or after controlling for age at first live birth.
-These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility."
+These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market."
 bibliography: "negative-selection.bib"
@@
 * Reweight other regressions (e.g. the omnibus regression in the appendix)?
-* Another point is that negative selection only among poor will increase
-    genetic inequality.
-  - one way to look at this: regress PGS on e.g. income or Quin71
-  - then weight by n_children and see the change
-  - needs to be comparable in some sense though, across generations
-  - e.g. quintile of where born, unweighted; vs quintile of where now,
-    weighted by children
-  - perhaps simpler is just to run regressions with/without weights
-    by n_children. This gives you the "effect" of natural selection
+
 * If you take "year of last job started" and see if it is after/before
@@
 
-As yet we know little about the *mechanisms* behind natural selection in
-contemporary human populations. This study uses data from UK Biobank to learn
-more. We test for natural selection on 33 different polygenic scores. We extend
-the analysis over two generations, using data on respondents' number of siblings
-as well as their number of children. This is interesting because consistent
-natural selection over multiple generations could lead to substantive effects in
-the long run.
-
-Most importantly, we examine reproductive rates in different
-subgroups of the population, in order to uncover patterns that can help
-illuminate the mechanisms behind modern natural selection. We find that 
-across the board, polygenic scores have stronger relationships with fertility 
-in specific subgroups of the population. Furthermore, in some subgroups, 
-the direction of relationships is actually reversed.
+As yet we know little about the mechanisms behind these effects. This study uses
+data from UK Biobank to learn more. We test for natural selection on 33
+different polygenic scores by estimating their correlation with fertility. We
+extend the analysis over two generations, using data on respondents' number of
+siblings as well as their number of children. This is interesting because
+consistent natural selection over multiple generations could lead to substantive
+effects in the long run. Most importantly, we examine reproductive rates in
+different subgroups of the population, in order to uncover patterns that can
+help illuminate the mechanisms behind modern natural selection.
+
+We find selection effects on many polygenic scores. Effects are consistent
+across generations. The strength of natural selection on a polygenic score is
+associated with that score's correlation with education and earnings: scores
+that predict lower education and earnings are being selected for. Also, across
+the board, polygenic scores have stronger relationships with fertility among
+specific subgroups. Selection effects are stronger among people with lower
+income and less education, younger parents, people not living with a partner,
+and people with many lifetime sexual partners. Outside these groups, effects are
+weaker and often statistically insignificant. In fact, in some subgroups, the
+direction of selection is actually reversed: polygenic scores predicting
+higher education and earnings are associated with *higher* fertility.
 
@@
 by genetic estimated ancestry). For regressions on number of children we 
-limit the sample to people over 45, since most fertility is completed by this age.
-
-We compute 33 polygenic scores, normalized
+limit the sample to people over 45, since most fertility is completed by this age. We compute 33 polygenic scores, normalized
 to mean 0, variance 1. Figure \@ref(fig:plot-means-over-time) plots mean
-polygenic scores by 5-year birth intervals. Several scores show
+polygenic scores in the sample by 5-year birth intervals. Several scores show
 consistent increases or declines over this 30-year period, of the order
@@
 
-These changes could reflect either natural selection within the UK
-population, or ascertainment bias within the sample. Respondents are
+These changes could reflect natural selection within the UK
+population, but also ascertainment bias within the sample. Respondents are
 higher income and better educated than the UK population, and they may
@@
 
-We next regress respondents'polygenic scores on their number of children
-($y_i$):
+To test directly for natural selection, we regress respondents'polygenic scores
+on their number of children ($y_i$):
 
@@
 selection within the sample. In fact, since polygenic scores are
-normalized to mean 0, $\beta$ is the expected polygenic score among
-children of the sample [@beauchamp2016genetic]. To learn more about the
+normalized, $\beta$ is the expected polygenic score among
+children of the sample [@beauchamp2016genetic].^[normalization] To learn more about the
 underlying mechanisms, we split the sample, starting with basic
@@
 `r female_ratio * 100`% female.
+
+^[normalization]: The selection effect $\beta$ equals $Cov(Y, PGS)/Var(PGS)$
+where $Y$ is the number of children. Since PGS are normalized to variance 1,
+this reduces to to $Cov(Y, PGS) \equiv E(Y PGS) - E(Y) E(PGS)$, which in turn reduces to $E(Y PGS)$ as $E(PGS) = 0$. This is the polygenic score weighted
+by the number of children, which is the average polygenic score in the 
+next generation.
 
@@
 ```{r calc-income-edu-controlled}
+
 drake::loadd(res_income_controlled, res_edu_controlled)
@@
 women, including major depressive disorder (MDD), schizophrenia and
-neuroticism, while extraversion is more selected for among men. Waist
-circumference and waist-hip ratio are less selected for among women, and
-educational attainment is more selected against, though the difference
+neuroticism, while extraversion is more selected for among men. PGS for waist
+circumference and waist-hip ratio are less selected for among women, and PGS for
+educational attainment are more selected against, though the difference
 is only significant for EA3. One possible reason for these sex differences is
-that polygenic scores may affect fertility via mate search.
+that polygenic scores may affect fertility via success in marriage markets, and
+men and women may value different characteristics in these.
 
@@
 
-We next focus on variables related to mate search and household
-formation. We split males and females by lifetime number of sexual
+We next focus on variables related to household type and reproductive
+strategy. We split males and females by lifetime number of sexual
 partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
 For both sexes, selection effects are larger and more significant among
-those with more than 3 partners. Splitting respondents by whether
-they were living with a spouse or partner at the time of interview,
-effects are larger among those not living with a spouse or partner
+those with more than 3 partners. Next we split respondents by whether
+they were living with a spouse or partner at the time of interview.
+Effects are larger among those not living with a spouse or partner
 (Figure \@ref(fig:plot-n-partners)b). Lastly, we split female
@@
 sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`.
-Selection effects appear to come through two opposing channels: an
+Thus, selection effects seem to come through two opposing channels: an
 effect on AFLB, and an opposite-signed effect on number of children 
@@
 show the same pattern as for respondents: effect sizes are larger and more
-significant in the most deprived areas.
+often significant in the most deprived areas.
 
@@
 
-Our empirical results can be explained by the economic
-theory of fertility [@becker1960economic; @willis1973new; @becker1976child].
-According to this, increases in a person's wage affect their fertility via two
-opposing channels. There is an *income effect* by which children become more
-affordable, like any other good. There is also a *substitution effect*: since
-childrearing has a cost in time, the opportunity cost of childrearing increases
-if one's market wage is higher. The income effect would lead higher earners to
-have more children. The substitution effect would lead higher earners to have
-fewer children. If so, then genetic variants which affect earnings potential in
-the labour market may cause opposing effects on fertility. The income effect
-will cause natural selection in favour of earnings-increasing variants. The
-substitution effect will do the reverse.
-
-```{r calc-res-earn-educ}
-
-drake::loadd(res_ee_control)
-drake::loadd(res_all)
-
-res_ee_control %<>% filter(term == score_name)
-n_05 <- sum(res_ee_control$p.value < 0.05)
-n_05_33 <- sum(res_ee_control$p.value < 0.05/33)
-
-```
-
-This theory predicts that genetic variants will be selected for (or against) in
-proportion to their effect on earnings. Figure
-\@ref(fig:plot-cors-earnings-educ) plots selection effects on each PGS against
-that PGS's correlation with earnings in a respondent's first job[^earnings-est],
-and against its correlation with educational attainment, a predictor of lifetime
-earnings. The raw relationships (left column) are strongly negative. If we plot
-selection effects controlling for age at first live birth (right column), the
-effect reverses and becomes positive. Thus, the labour market appears to
-play an important role in natural selection. Substitution effects
-dominate income effects overall, which fits the empirical association 
-between income and lower fertility [@becker1960economic;@jones2006economic].
-
-[^earnings-est]: Earnings are estimated from the 2007 Annual Survey of Hours and
-Earnings, using the SOC 2000 job code.
-
-We also re-estimate equation \@ref(eq:regression) for each
-PGS, controlling for earnings in first job and education levels. Effect sizes
-are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only
-`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$. This suggests that
-earnings and education indeed mediate the effect of PGS on fertility.
-
-```{r plot-cors-earnings-educ, fig.cap="Selection effects, with and without controls for age at first live birth, by correlations with earnings and educational attainment.
-Each point represents one PGS."}
-
-drake::loadd(res_all)
-drake::loadd(res_age_flb)
-drake::loadd(ashe_income)
-
-famhist <- add_ashe_income(famhist, ashe_income)
-
-cors_income <- cor(famhist[score_names], famhist$first_job_pay, use = "pairwise")
-# TODO: use age_left_edu when this includes college
-
-cors_educ <-   cor(famhist[score_names], famhist$age_fulltime_edu, use = "pairwise")
-
-effect_size <-  res_all %>% 
-                  filter(dep.var == "N children", reg.type == "raw") %>% 
-                  pull(estimate)
-
-effect_size_flb <-  res_age_flb %>% 
-                      filter(term != "age_flb") %>% 
-                      pull(estimate)
-
-dfr <- data.frame(
-        Correlation  = rep(c("Earnings", "Education"), each = 33, 2),
-        cor          = rep(c(cors_income[,1], cors_educ[,1]), 2),
-        effect_size  = c(rep(effect_size, 2), rep(effect_size_flb, 2)),
-        Controls     = rep(c("None", 
-                        "Age at first live birth"), each = 66)
-      )
-
-dfr %>% 
-      mutate(
-        Controls = fct_relevel(Controls, "None")
-      ) %>% 
-      ggplot(aes(effect_size, cor, colour = Correlation)) + 
-        geom_point() +
-        facet_grid(Correlation ~ Controls, labeller = label_both) + 
-        my_vline + 
-        my_hline +
-        labs(x = "Selection effect", y = "Correlation") +
-        scale_colour_brewer(type = "qual", palette = 2) +
-        theme(
-          legend.position = "none",
-          panel.spacing = unit(14, "pt")
-        )
-
-```
+# Economic theory and natural selection
+
+These results show that selection effects are weaker, absent, or even reversed
+among some subgroups of the population. This can potentially be explained by the
+economic theory of fertility [@becker1960economic; @willis1973new;
+@becker1976child]. According to this, increases in a person's wage affect their
+fertility via two opposing channels. There is an *income effect* by which
+children become more affordable, like any other good. There is also a
+*substitution effect*: since childrearing has a cost in time, the opportunity
+cost of childrearing increases if one's market wage is higher. The income effect
+would lead higher earners to have more children. The substitution effect would
+lead higher earners to have fewer children. If so, then genetic variants which
+affect earnings potential in the labour market may cause opposing effects on
+fertility. The income effect will cause natural selection in favour of
+earnings-increasing variants. The substitution effect will do the reverse.
 
@@
 capital, i.e. skills and characteristics that are valuable in the labour market.
-Suppose also that education and innate human capital are complements. Individuals
-choose how long to stay in education; after leaving education they enter the
-labour market and form families. If so, then people with less human capital will
-leave education earlier and have their first child earlier [cf.
-@caucutt2002women; @monstad2008education]. This is a pure substitution effect: the income effect plays
-no role in the decision to leave education, since people are not yet earning
-wages.[^no-borrow] Whilst in the labour market, higher earners will have more
-children (i.e. income effects dominate). These assumptions will lead to the
-pattern we observe: earnings-increasing variants will predict later first births (and
-fewer children overall), but they will also predict more children within any given
-age at first live birth.
+Suppose also that education and innate human capital are complements.
+Individuals choose how long to stay in education; after leaving education they
+enter the labour market and form families. If so, then people with less human
+capital will leave education earlier and have their first child earlier [cf.
+@caucutt2002women; @monstad2008education]. This is a pure substitution effect:
+the income effect plays no role in the decision to leave education, since people
+are not yet earning wages.[^no-borrow] Whilst in the labour market, higher
+earners will have more children (i.e. income effects dominate). These
+assumptions will lead to the pattern we observe: earnings-increasing variants
+will predict later first births (and fewer children overall), but they will also
+predict more children within any given age at first live birth.
 
@@
 
+```{r plot-cors-earnings-educ, fig.cap="Selection effects, with and without controls for age at first live birth, by correlations with earnings and educational attainment.
+Each point represents one PGS."}
+
+drake::loadd(res_all)
+drake::loadd(res_age_flb)
+drake::loadd(ashe_income)
+
+famhist <- add_ashe_income(famhist, ashe_income)
+
+cors_income <- cor(famhist[score_names], famhist$first_job_pay, use = "pairwise")
+# TODO: use age_left_edu when this includes college
+
+cors_educ <-   cor(famhist[score_names], famhist$age_fulltime_edu, use = "pairwise")
+
+effect_size <-  res_all %>% 
+                  filter(dep.var == "N children", reg.type == "raw") %>% 
+                  pull(estimate)
+
+effect_size_flb <-  res_age_flb %>% 
+                      filter(term != "age_flb") %>% 
+                      pull(estimate)
+
+dfr <- data.frame(
+        PGS          = rep(score_names, 4),
+        Correlation  = rep(c("Earnings", "Education"), each = 33, 2),
+        cor          = rep(c(cors_income[,1], cors_educ[,1]), 2),
+        effect_size  = c(rep(effect_size, 2), rep(effect_size_flb, 2)),
+        Controls     = rep(c("None", 
+                        "Age at first live birth"), each = 66)
+      )
+
+dfr %>% 
+      mutate(
+        Controls = fct_relevel(Controls, "None"),
+        PGS      = ifelse(abs(effect_size) > 0.035, PGS, ""),
+        PGS      = sub("(^\\w*?)_.*", "\\1", PGS)
+      ) %>% 
+      ggplot(aes(effect_size, cor, colour = Correlation)) + 
+        geom_point() +
+        geom_text(aes(label = PGS), size = 3, colour = "black", 
+                    check_overlap = TRUE, nudge_x = 0.012) +
+        facet_grid(Correlation ~ Controls, labeller = label_both) + 
+        my_vline + 
+        my_hline +
+        labs(x = "Selection effect", y = "Correlation") +
+        scale_colour_brewer(type = "qual", palette = 2) +
+        coord_cartesian(clip = "off") +
+        theme(
+          legend.position = "none",
+          panel.spacing = unit(14, "pt")
+        )
+
+```
+
+We test this explanation in two ways. First, the economic theory predicts that
+genetic variants will be selected for (or against) in proportion to their effect
+on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
+on each PGS against that PGS's correlation with earnings in a respondent's first
+job[^earnings-est], and against its correlation with educational attainment, a
+predictor of lifetime earnings. The raw relationships (left column) are strongly
+negative. If we plot selection effects controlling for age at first live birth
+(right column), the effect reverses and becomes positive. Thus, the labour
+market appears to play an important role in natural selection. Substitution
+effects dominate income effects overall, which fits the known association
+between income and lower fertility [@becker1960economic;@jones2006economic].
+
+[^earnings-est]: Earnings are estimated from the 2007 Annual Survey of Hours and
+Earnings, using the SOC 2000 job code.
+
+```{r calc-res-earn-educ}
+
+drake::loadd(res_ee_control)
+drake::loadd(res_all)
+
+res_ee_control %<>% filter(term == score_name)
+n_05 <- sum(res_ee_control$p.value < 0.05)
+n_05_33 <- sum(res_ee_control$p.value < 0.05/33)
+
+```
+
+Second, we re-estimate equation \@ref(eq:regression) for each
+PGS, controlling for earnings in first job and education levels. Effect sizes
+are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only
+`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$. This suggests that
+earnings and education indeed mediate the effect of PGS on fertility.
+
 ```{r calc-risk-attitude}
@@
 is less good at explaining variation in selection across the full range of
-scores, including physical measures, e.g. waist-hip ratio and BMI (see also
-appendix Figure \@ref(fig:plot-together)). Externalizing behaviour also does not
-explain why selection effects change sign among older parents.
-
-We test the alternative theory directly by re-estimating equation
-\@ref(eq:regression) controlling for a measure of risk attitude (Biobank field
-2040). While risk attitude is always a highly significant predictor of number of
-children, it has little impact on the effect size or significance of the PGS.
-The median ratio of PGS effect sizes between regressions with and without
-controls is `r median(ratio)`; all PGS which are significant at $p \le 0.05/33$
-in uncontrolled regressions remain so when controlling for risk attitude. Thus,
-although risk attitude does predict number of children, it appears not to
-mediate selection effects. Overall, we believe that the economic theory is the
-leading explanation.
+scores, including physical measures, e.g. waist-hip ratio and BMI. Externalizing
+behaviour also does not explain why selection effects should work in the
+opposite direction among older parents, whereas the economic theory can explain
+this via income effects, as described above. We test the alternative theory
+directly by re-estimating equation \@ref(eq:regression) controlling for a
+measure of risk attitude (Biobank field 2040). While risk attitude is always a
+highly significant predictor of number of children, it has little impact on the
+effect size or significance of the PGS. The median ratio of PGS effect sizes
+between regressions with and without controls is `r median(ratio)`; all PGS
+which are significant at $p \le 0.05/33$ in uncontrolled regressions remain so
+when controlling for risk attitude. Thus, although risk attitude does predict
+number of children, it appears not to mediate selection effects. Overall, we
+believe that the economic theory is the leading explanation.
 
@@
 problem may be less serious in surveys which aim to be representative (as the UK
-Biobank does not). Nevertheless, not all respondents consent to the collection
-of genetic data: for instance, completion rates for genotype data in the US
-Health and Retirement Study were around 80-85% [@hrs2020]. Researchers should be
-aware of the risks of ascertainment when studying modern natural selection.
-
-In addition, we do not know whether estimated effect sizes of natural
-selection will increase or decrease as more accurate polygenic scores
-are produced. We are also unsure whether genetic variants underlying other
-phenotypes will show a similar pattern of natural selection as these ones. 
-In short, it is probably too early to tell whether modern natural
-selection has a substantively important effect on populations' mean 
-polygenic scores.
-
-Because selection effects are concentrated in poorer populations, they also increase
+Biobank does not). However, there is still scope for bias, since not all
+respondents consent to the collection of genetic data. For instance, completion
+rates for genotype data in the US Health and Retirement Study were around 80-85%
+[@hrs2020]. Researchers should be aware of the risks of ascertainment when
+studying modern natural selection.
+
+We also do not know how estimated effect sizes of natural selection will change
+as more accurate measures of genetic variation are produced. And we are unsure
+whether genetic variants underlying other phenotypes will show a similar pattern
+of natural selection to those studied here. In short, it is probably too early
+to tell whether modern natural selection has a substantively important effect on
+population genetics. Nevertheless, we note that selection effects on our measured
+polygenic scores are still relatively small, even after reweighting to account for ascertainment bias.
+
+Because selection effects are concentrated in poorer groups, they also increase
 inequality with respect to polygenic scores. For example, Figure
-\@ref(fig:plot-mean-EA3-by-income) graphs children's mean polygenic scores for 
-educational attainment (EA3), in households of different income groups. The blue
-bars show the actual means. The grey bars show the hypothetical means if all 
-households had equal numbers of children. Natural selection against EA3, which
-lowers the mean, is stronger at the bottom of the income distribution, and this
-increases the differences between groups.
+\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
+attainment (EA3), for children in households of different income groups. The
+blue bars show the actual means, i.e. parents' mean polygenic score weighted by
+number of children. The grey bars show the hypothetical means if all households
+had equal numbers of children. Natural selection against EA3, which lowers the
+mean, is stronger at the bottom of the income distribution, and this increases
+the differences between groups.
 
@@
 
+## Effects on inequality
+
+Table \@ref(tab:tbl-inequality-effects) estimates differences in children's mean polygenic
+scores between the highest and lowest income groups. Column "Actual" shows parents' scores weighted by Age/Qualification times number of children. 
+Column "Without selection" shows scores weighted by Age/Qualification only, i.e. if all couples had the same number of children.
+
+```{r tbl-inequality-effects}
+
+drake::loadd(age_qual_weights)
+
+ineq <- famhist %>% 
+          left_join(age_qual_weights, by = "f.eid") %>% 
+          filter(! is.na(n_children), ! is.na(income_cat), ! is.na(weights)) %>% 
+          group_by(income_cat) %>% 
+          summarize(across(all_of(score_names), 
+                    list(
+                      unw = ~ weighted.mean(.x, weights, na.rm = TRUE),
+                      wtd = ~ weighted.mean(.x, weights * n_children, 
+                                              na.rm = TRUE)
+                    )
+          )) %>% 
+          tidyr::pivot_longer(-income_cat, 
+                                names_to = c("score", ".value"), 
+                                names_pattern = "(.*)_(.+?)$"
+                              )
+
+ineq_comp <-  left_join(
+                ineq %>% filter(income_cat == 1),
+                ineq %>% filter(income_cat == 5),
+                by = "score"
+              )
+
+ineq_comp %<>% mutate(
+                 diff_children = wtd.y - wtd.x,
+                 diff_hypoth   = unw.y - unw.x
+               )
+
+ineq_comp %>% 
+      select(
+        PGS                 = score, 
+        `With selection`    = diff_children, 
+        `Without selection` = diff_hypoth
+      ) %>% 
+      mutate(
+        PGS = pretty_names(PGS)
+      ) %>% 
+      arrange(desc(Actual)) %>% 
+      as_hux() %>% 
+      set_align(everywhere, -1, "right") %>% 
+      set_caption("Differences in polygenic scores between highest and lowest income group") %>% 
+      theme_compact()
+```
+
+\FloatBarrier
+
 # References

### 2020-12-10 why-negative-selection.Rmd
Use residualized scores as the default. Add materials and methods section. Final tweaks for first draft.

@@
 
-# nasty hack to make add_ashe_income work later
-source("~/import-ukbb-data/import-ukbb-data.R")
+# nasty hack to make add_ashe_income work later:
+if (! exists("add_ashe_income")) source("~/import-ukbb-data/import-ukbb-data.R")
 
@@
 }
+
+comma_num <- function (x) prettyNum(x, big.mark = ",")
 
@@
 
-# Data and methods
-
-We use data from UK Biobank. Our sample consists of white British (as defined
-by genetic estimated ancestry). For regressions on number of children we 
-limit the sample to people over 45, since most fertility is completed by this age. We compute 33 polygenic scores, normalized
-to mean 0, variance 1. Figure \@ref(fig:plot-means-over-time) plots mean
-polygenic scores in the sample by 5-year birth intervals. Several scores show
-consistent increases or declines over this 30-year period, of the order
-of 5% of a standard deviation.
+# Results
+
+Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores in the
+sample by 5-year birth intervals. Several scores show consistent increases or
+declines over this 30-year period, of the order of 5% of a standard
+deviation.These changes could reflect natural selection within the UK
+population, but also ascertainment bias within the sample. Respondents are
+higher income and better educated than the UK population, and they may
+also differ on other unobserved characteristics [@10.1093/aje/kwx246].
+Since richer and educated people also live longer, this bias might also
+increase with age.
 
@@
 
-These changes could reflect natural selection within the UK
-population, but also ascertainment bias within the sample. Respondents are
-higher income and better educated than the UK population, and they may
-also differ on other unobserved characteristics [@10.1093/aje/kwx246].
-Since richer and educated people also live longer, this bias might also
-increase with age.
-
 To test directly for natural selection, we regress respondents'polygenic scores
@@
 
-The "selection effect", $\beta$, measures the strength of natural
-selection within the sample. In fact, since polygenic scores are
-normalized, $\beta$ is the expected polygenic score among
-children of the sample [@beauchamp2016genetic].^[normalization] To learn more about the
-underlying mechanisms, we split the sample, starting with basic
-demographic variables including education, income and sex. These are all
-potential sources of ascertainment bias: as well as the ascertainment
-for income and education, mentioned above, the sample sex ratio skews
-`r female_ratio * 100`% female.
-
-^[normalization]: The selection effect $\beta$ equals $Cov(Y, PGS)/Var(PGS)$
+The "selection effect", $\beta$, measures the strength of natural selection
+within the sample. In fact, since polygenic scores are normalized, $\beta$ is
+the expected polygenic score among children of the sample
+[@beauchamp2016genetic].[^normalization] To learn more about the underlying
+mechanisms, we split the sample, starting with basic demographic variables
+including education, income and sex. These are all potential sources of
+ascertainment bias: as well as the ascertainment for income and education,
+mentioned above, the sample sex ratio skews `r female_ratio * 100`% female.
+
+[^normalization]: The selection effect $\beta$ equals $Cov(Y, PGS)/Var(PGS)$
 where $Y$ is the number of children. Since PGS are normalized to variance 1,
-this reduces to to $Cov(Y, PGS) \equiv E(Y PGS) - E(Y) E(PGS)$, which in turn reduces to $E(Y PGS)$ as $E(PGS) = 0$. This is the polygenic score weighted
-by the number of children, which is the average polygenic score in the 
-next generation.
+this reduces to to $Cov(Y, PGS) \equiv E(Y PGS) - E(Y) E(PGS)$, which in turn
+reduces to $E(Y PGS)$ as $E(PGS) = 0$. This is the polygenic score weighted by
+the number of children, which is the average polygenic score in the next
+generation.
 
@@
 
+## Correcting for sampling bias
+
 ```{r calc-res-weighted}
@@
   None                = res_all %>% 
-                               filter(reg.type == "raw", dep.var == "N children") %>% 
+                          filter(
+                            reg.type == "controlled", 
+                            dep.var == "N children"
+                          ) %>% 
                           select(-reg.type, -dep.var, -score_name),
@@
 results suggest that naïve estimates of natural selection are at risk of
-ascertainment bias. To correct for this, we weight participants using data from
-the 2011 Census and from the 2006 General Household Survey (GHS). We try three
-alternative weighting schemes: geographical weighting by census MSOA, age and
+ascertainment bias. To correct for this, we weight participants using population data. We try three
+alternative weighting schemes: weighting by geography, age and
 presence/absence of a partner; weighting by age and highest educational
 qualification; and age, highest qualification, and age at first live birth, for
-women only. Mean effect sizes across all polygenic scores are increased by a
-factor of `r mean_rel$Geographical` (geographical weighting), 
+women only. Figure \@ref(fig:plot-res-wt) plots selection effects 
+among the entire sample, estimated with the three weighting schemes. 
+Mean effect sizes across all polygenic scores are
+increased by a factor of `r mean_rel$Geographical` (geographical weighting), 
 `r mean_rel[["Age/Qualification"]]` (age/qualification) or 
@@
 
-The UK Biobank data contains information on respondents' number of
-siblings. Since respondents' polygenic scores are equal in expectation
-to the mean scores of their parents, we can use this to look at selection 
-effects in the parents' generation. We estimate equation (\ref{eq:regression})
-using *number of siblings* (including the respondent) as the dependent variable.
-The parents' generation has a further source of ascertainment bias: sampling
-parents of respondents overweights parents who have many children. For instance,
-parents of three children will have, on average, three times more children
-represented in UK Biobank than parents of one child. Parents of no children will
-by definition not be represented. To compensate, we reweight our preferred
-weightings (Age/Qualification) by the inverse of *number of siblings*.
-
-Selection effects are highly correlated across the two generations, and most
-share the same sign (appendix Figure \@ref(fig:plot-res-sibs)). Effect size
-estimates are larger for the parents' generation. We treat this result cautiously,
-because when we split respondents up by year of birth, we find few differences
-in effect sizes between early- and late-born respondents, for either generation.
-In other words, since estimated effect sizes change when we change the dependent
-variable, but do not change over time within either dependent variable, this
-may be due to remaining ascertainment bias within the sample.
-
-Although the direction of selection effects does not change between the generations,
-there are other differences. After we residualize polygenic scores on the first
-100 principal components of the array data, selection effects on *number of
-siblings* are reduced by about ten percent on average, whereas effects on
-*number of children* barely change (appendix Figure
-\@ref(fig:plot-regs-controlled-pcs)). This could be because
-earlier fertility is driven more by geographically clustered deprivation [e.g.
-via an insurance motive, @rendall1998old], which may correlate with the
-broad-scale genetic variation captured by principal components. Another
-difference is that in the parents' generation, raw selection effects and effects
-controlling for AFLB are positively correlated, whereas they are negatively
-correlated in the respondents' generation.[^educ-exception] These differences
-suggest that the mechanisms of natural selection may have changed over time,
-even though the outcomes are broadly the same.
-
-[^educ-exception]: Educational attainment is an exception. It is selected 
-against in the parents' generation, but selected for if we control for AFLB,
-which is the same pattern we observe among the respondents' generation.
-
-```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by area Townsend deprivation quintile", fig.align = "center"}
-
-drake::loadd(res_townsend_parents)
-drake::loadd(famhist_townsend_71)
-n_regs <- nrow(res_townsend_parents)
-
-res_townsend_parents %>% 
-  mutate(
-    `Townsend quintile` = sub("Quin71(\\d):.*", "\\1", term),
-    `Townsend quintile` = add_n(`Townsend quintile`, "Quin71", dv = "n_sibs", 
-                                  data = famhist_townsend_71)
-  ) %>% 
-  standard_ggplot(fill_col    = `Townsend quintile`,
-                    n_regs    = n_regs, 
-                    score_col = score_name, 
-                    order_idx = 5
-                  )
-
-rm(famhist_townsend_71)
-```
-
-We also check whether selection effects differ by socio-economic status in the
-parents' generation. We have no information about parents' income, so we use the
-1971 Townsend deprivation score of respondents' birthplace as a proxy
-[@townsend1987deprivation]. Results (Figure \@ref(fig:plot-siblings-townsend))
-show the same pattern as for respondents: effect sizes are larger and more
-often significant in the most deprived areas.
-
-```{r calc-siblings-children}
-sib_chn_cor <- cor(famhist$n_sibs, famhist$n_children, use = "complete")
-```
-
-Lastly, the siblings data lets us check for a "quantity-quality tradeoff" between
-number of children and number of grandchildren (for the parents' generation). We
-do not find any: in fact, the correlation between *number of siblings* and 
-*number of children* is `r sib_chn_cor`.
-
-# Economic theory and natural selection
-
-These results show that selection effects are weaker, absent, or even reversed
-among some subgroups of the population. This can potentially be explained by the
-economic theory of fertility [@becker1960economic; @willis1973new;
-@becker1976child]. According to this, increases in a person's wage affect their
-fertility via two opposing channels. There is an *income effect* by which
-children become more affordable, like any other good. There is also a
-*substitution effect*: since childrearing has a cost in time, the opportunity
-cost of childrearing increases if one's market wage is higher. The income effect
-would lead higher earners to have more children. The substitution effect would
-lead higher earners to have fewer children. If so, then genetic variants which
-affect earnings potential in the labour market may cause opposing effects on
-fertility. The income effect will cause natural selection in favour of
-earnings-increasing variants. The substitution effect will do the reverse.
-
-This theory can explain why selection effects have opposite signs when AFLB is
-controlled for. Suppose that people are aware of their endowments of human
-capital, i.e. skills and characteristics that are valuable in the labour market.
-Suppose also that education and innate human capital are complements.
-Individuals choose how long to stay in education; after leaving education they
-enter the labour market and form families. If so, then people with less human
-capital will leave education earlier and have their first child earlier [cf.
-@caucutt2002women; @monstad2008education]. This is a pure substitution effect:
-the income effect plays no role in the decision to leave education, since people
-are not yet earning wages.[^no-borrow] Whilst in the labour market, higher
-earners will have more children (i.e. income effects dominate). These
-assumptions will lead to the pattern we observe: earnings-increasing variants
-will predict later first births (and fewer children overall), but they will also
-predict more children within any given age at first live birth.
-
-[^no-borrow]: This argument assumes that capital market imperfections hinder
-people from borrowing against future income.
-
-The theory can also explain why natural selection is weaker among higher-income
-parents. @becker1976child show that if parents care about child quality as well
-as quantity, under certain assumptions the income effect will be stronger at
-higher income levels; this can lead to a U-shaped relationship between income
-and fertility. Empirically, @cohen2013financial find that income decreases 
-fertility at low income levels but increases it at higher income levels.
-
-Lastly, it is often assumed that the substitution effect will dominate for lone
-parents, or those in unstable relationships, since they have less opportunity to
-share childcare responsibilities, while the income effect will dominate for
-couples who are able to reap the gains from specialization. Indeed, US fertility
-decreases faster with education among single mothers than married mothers
-[@baudin2015fertility]. This can explain why earnings-increasing genetic variants 
-decrease fertility more among single parents than among couples. The same logic 
-could explain our results for lifetime number of sexual partners, if this is 
-associated with relationship stability.
-
-```{r plot-cors-earnings-educ, fig.cap="Selection effects, with and without controls for age at first live birth, by correlations with earnings and educational attainment.
-Each point represents one PGS."}
-
-drake::loadd(res_all)
-drake::loadd(res_age_flb)
-drake::loadd(ashe_income)
-
-famhist <- add_ashe_income(famhist, ashe_income)
-
-cors_income <- cor(famhist[score_names], famhist$first_job_pay, use = "pairwise")
-# TODO: use age_left_edu when this includes college
-
-cors_educ <-   cor(famhist[score_names], famhist$age_fulltime_edu, use = "pairwise")
-
-effect_size <-  res_all %>% 
-                  filter(dep.var == "N children", reg.type == "raw") %>% 
-                  pull(estimate)
-
-effect_size_flb <-  res_age_flb %>% 
-                      filter(term != "age_flb") %>% 
-                      pull(estimate)
-
-dfr <- data.frame(
-        PGS          = rep(score_names, 4),
-        Correlation  = rep(c("Earnings", "Education"), each = 33, 2),
-        cor          = rep(c(cors_income[,1], cors_educ[,1]), 2),
-        effect_size  = c(rep(effect_size, 2), rep(effect_size_flb, 2)),
-        Controls     = rep(c("None", 
-                        "Age at first live birth"), each = 66)
-      )
-
-dfr %>% 
-      mutate(
-        Controls = fct_relevel(Controls, "None"),
-        PGS      = ifelse(abs(effect_size) > 0.035, PGS, ""),
-        PGS      = sub("(^\\w*?)_.*", "\\1", PGS)
-      ) %>% 
-      ggplot(aes(effect_size, cor, colour = Correlation)) + 
-        geom_point() +
-        geom_text(aes(label = PGS), size = 3, colour = "black", 
-                    check_overlap = TRUE, nudge_x = 0.012) +
-        facet_grid(Correlation ~ Controls, labeller = label_both) + 
-        my_vline + 
-        my_hline +
-        labs(x = "Selection effect", y = "Correlation") +
-        scale_colour_brewer(type = "qual", palette = 2) +
-        coord_cartesian(clip = "off") +
-        theme(
-          legend.position = "none",
-          panel.spacing = unit(14, "pt")
-        )
-
-```
-
-We test this explanation in two ways. First, the economic theory predicts that
-genetic variants will be selected for (or against) in proportion to their effect
-on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
-on each PGS against that PGS's correlation with earnings in a respondent's first
-job[^earnings-est], and against its correlation with educational attainment, a
-predictor of lifetime earnings. The raw relationships (left column) are strongly
-negative. If we plot selection effects controlling for age at first live birth
-(right column), the effect reverses and becomes positive. Thus, the labour
-market appears to play an important role in natural selection. Substitution
-effects dominate income effects overall, which fits the known association
-between income and lower fertility [@becker1960economic;@jones2006economic].
-
-[^earnings-est]: Earnings are estimated from the 2007 Annual Survey of Hours and
-Earnings, using the SOC 2000 job code.
-
-```{r calc-res-earn-educ}
-
-drake::loadd(res_ee_control)
-drake::loadd(res_all)
-
-res_ee_control %<>% filter(term == score_name)
-n_05 <- sum(res_ee_control$p.value < 0.05)
-n_05_33 <- sum(res_ee_control$p.value < 0.05/33)
-
-```
-
-Second, we re-estimate equation \@ref(eq:regression) for each
-PGS, controlling for earnings in first job and education levels. Effect sizes
-are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only
-`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$. This suggests that
-earnings and education indeed mediate the effect of PGS on fertility.
-
-```{r calc-risk-attitude}
-
-drake::loadd(res_risk_control)
-drake::loadd(res_all)
-
-n_risk_sig <- res_risk_control %>% 
-            filter(term == 'f.2040.0.0') %>% 
-            {sum(.$p.value <= 0.05/33)}
-
-n_pgs_sig <- res_risk_control %>% 
-            filter(term == score_name) %>% 
-            {sum(.$p.value <= 0.05/33)}
-
-res_compare <- res_all %>% 
-  filter(dep.var=='N children', reg.type == 'raw') %>% 
-  left_join(res_risk_control %>% filter(term == score_name), by = 'score_name')
-
-ratio <- res_compare$estimate.y/res_compare$estimate.x
-n_less_sig <- sum(res_compare$p.value.y <= 0.05/33 & res_compare$p.value.x > 0.05/33)
-```
-
-An alternative theory is that traits selected for are linked to externalizing
-behaviour, risk-seeking and low time discount rates, via the channel of early
-sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
-for this: scores which might plausibly be linked to externalizing behaviour,
-like ADHD and age at smoking initiation, are selected for. However, this theory
-is less good at explaining variation in selection across the full range of
-scores, including physical measures, e.g. waist-hip ratio and BMI. Externalizing
-behaviour also does not explain why selection effects should work in the
-opposite direction among older parents, whereas the economic theory can explain
-this via income effects, as described above. We test the alternative theory
-directly by re-estimating equation \@ref(eq:regression) controlling for a
-measure of risk attitude (Biobank field 2040). While risk attitude is always a
-highly significant predictor of number of children, it has little impact on the
-effect size or significance of the PGS. The median ratio of PGS effect sizes
-between regressions with and without controls is `r median(ratio)`; all PGS
-which are significant at $p \le 0.05/33$ in uncontrolled regressions remain so
-when controlling for risk attitude. Thus, although risk attitude does predict
-number of children, it appears not to mediate selection effects. Overall, we
-believe that the economic theory is the leading explanation.
-
-# Discussion
-
-Previous work has documented natural selection in modern populations on variants
-underlying many polygenic scores [@beauchamp2016genetic; @kong2017selection;
-@sanjak2018evidence]. We show that natural selection effects are highly
-concentrated among specific subgroups of the population, including poorer
-people, less educated people, younger parents and those with more lifetime
-sexual partners. Indeed, among older parents selection effects are actually
-reversed. Furthermore, the size of selection effects on a polygenic score
-correlates with that score's association with labour market earnings. The 
-economic theory of fertility provides a parsimonious explanation for this.
-
-PGS which correlate with high (low) earnings and more (less) education are being
-selected against (selected for). In addition, many of the phenotypes under
-positive selection are linked to poor health, or are what many people would see
-as undesirable to have. For example, most people would probably prefer to have
-high educational attainment, a low risk of major depressive disorder, and a low
-risk of coronary artery disease, but natural selection is pushing against
-genetics linked to these traits. Potentially, this could increase the health
-burden on modern populations, but that depends on effect sizes. Our results
-suggest that naïve estimates can be affected by sample ascertainment bias. This
-problem may be less serious in surveys which aim to be representative (as the UK
-Biobank does not). However, there is still scope for bias, since not all
-respondents consent to the collection of genetic data. For instance, completion
-rates for genotype data in the US Health and Retirement Study were around 80-85%
-[@hrs2020]. Researchers should be aware of the risks of ascertainment when
-studying modern natural selection.
-
-We also do not know how estimated effect sizes of natural selection will change
-as more accurate measures of genetic variation are produced. And we are unsure
-whether genetic variants underlying other phenotypes will show a similar pattern
-of natural selection to those studied here. In short, it is probably too early
-to tell whether modern natural selection has a substantively important effect on
-population genetics. Nevertheless, we note that selection effects on our measured
-polygenic scores are still relatively small, even after reweighting to account for ascertainment bias.
-
-Because selection effects are concentrated in poorer groups, they also increase
-inequality with respect to polygenic scores. For example, Figure
-\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
-attainment (EA3), for children in households of different income groups. The
-blue bars show the actual means, i.e. parents' mean polygenic score weighted by
-number of children. The grey bars show the hypothetical means if all households
-had equal numbers of children. Natural selection against EA3, which lowers the
-mean, is stronger at the bottom of the income distribution, and this increases
-the differences between groups.
-
-```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean PGS for educational attainment (EA3) of children by household income group.
-Blue is actual.
-Grey is hypothetical in the absence of selection effects.", fig.width = 4}
-income_EA3 <- famhist %>% 
-      filter(! is.na(n_children), ! is.na(income_cat)) %>% 
-      mutate(
-        `Household income` = factor(income_cat, 
-                          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", 
-                                      "> £100K"))
-      ) %>% 
-      group_by(`Household income`) %>% 
-      summarize(
-        `Without selection` = mean(EA3_excl_23andMe_UK, na.rm = TRUE),
-        Actual              = weighted.mean(EA3_excl_23andMe_UK, n_children, 
-                                              na.rm = TRUE)
-      ) %>% 
-      tidyr::pivot_longer(-`Household income`) %>% 
-      rename(
-        `Children's mean EA3`     = value
-      )
-
-income_EA3 %>% 
-              ggplot(aes(`Household income`, `Children's mean EA3`, fill = name)) +
-                geom_col(position = "dodge") +
-                scale_fill_manual(values = c("navy", "grey60")) +
-                theme(
-                  legend.background = element_rect(fill = "white", colour = NA),
-                  legend.title = element_blank(), 
-                  legend.position = c(.25, .85)
-                )
-```
-
-Any model of fertility is implicitly a model of natural selection. But so far,
-the economic and human genetics literatures have developed in parallel.
-Integrating the two could deepen our understanding of modern natural selection.
-Economics possesses a range of theoretical models on the effects of skills,
-education and income [see @hotz1997economics, @lundberg2007american]. One
-perennial problem is how to test these theories in a world where education,
-labour and marriage markets all interact. Genetic data, such as polygenic
-scores, could help to pin down causality. Conversely, theory and empirical
-results from economics can shine a light on the mechanisms behind natural
-selection.
-
-\FloatBarrier
-# Appendix
-
-## Natural selection by sex
-
-```{r plot-sexes, fig.cap = "Selection effects by sex. Lines show significant differences."}
-
-drake::loadd(res_sex)
-n_regs <- as.double(nrow(res_sex))
-
-res_sex_wide <- res_sex %>% 
-      select(term, estimate, sex, diff.p.value) %>% 
-      pivot_wider(names_from = "sex", values_from = "estimate") %>% 
-      mutate(
-        min_est = pmin(Female, Male),
-        max_est = pmax(Female, Male),
-        term = pretty_names(term),
-        term = fct_reorder(term, Female)
-      )
-
-res_sex %>% 
-      mutate(
-        term = pretty_names(term),
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(y = term, x = estimate, fill = sex, 
-            colour = p < 0.05/{{n_regs}})) +
-        my_vline +
-        geom_linerange(data = res_sex_wide, 
-          aes(
-            y      = term,
-            x      = NULL,
-            fill   = NULL,
-            xmin   = min_est, 
-            xmax   = max_est, 
-            colour = diff.p.value < 0.05/{{n_regs}}
-          ), 
-        group = "") +
-        pval_geom_point() +
-        pval_scale() +
-        my_fill_scale(n = 2) +
-        my_labs() +
-        my_theme()
-
-```
-
-\FloatBarrier
-## Controlling for AFLB
-
-```{r plot-age-flb, fig.cap = "Selection effects controlling for age at first live birth (women only). Effect sizes for women only without controls are shown for comparison."}
-
-n_regs <- as.double(nrow(res_age_flb))
-
-res_combined %>% 
-      mutate(
-        score_name = pretty_names(score_name),
-        score_name = fct_reorder(score_name, estimate, order_abs(2)),
-        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for AFLB")
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
-        color = p < 0.05/{{n_regs}})) + 
-        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
-        pval_geom_point() +
-        pval_scale() +
-        scale_fill_brewer(type = "qual", palette = 3) +
-        my_labs() +
-        my_theme()
-```
-
-\FloatBarrier
-## Weighted regressions
-
-Figure \@ref(fig:plot-res-wt) plots effect sizes among the entire
-sample, using our three weighting schemes. We include unweighted regressions
-for comparison. Table \@ref(tab:tbl-res-wt) gives effect sizes as a
-proportion of the unweighted effect size.
-
-```{r tbl-res-wt}
-
-rel_sizes %>% 
-      arrange(desc(Geographical)) %>% 
-      bind_rows(rel_sizes_summary) %>% 
-      rename(PGS = term) %>% 
-      as_hux() %>% 
-      insert_row("", "Weighting", fill = "") %>% 
-      set_caption(
-        "Weighted effect sizes as a proportion of unweighted"
-      ) %>% 
-      merge_cells(1, 2:4) %>% 
-      set_align(1, 2, "centre") %>% 
-      set_header_rows(1, TRUE) %>% 
-      theme_compact() %>% 
-      style_headers(bold = TRUE) %>% 
-      set_bottom_border(1, 2:4) %>% 
-      set_bottom_border(2, everywhere) %>% 
-      set_italic(final(2), 1) %>% 
-      set_font_size(8) %>% 
-      set_width(0.7) %>% # ensures line wrapping below
-      set_col_width(c(.34, .22, .22, .22)) %>% 
-      add_footnote(paste0(
-        "Only consistent and significant (when unweighted) estimates ",
-        "are shown.\n",
-        "Age/Qual/AFLB as a proportion of unweighted regressions ",
-        "for females only."),
-        font_size    = 6, 
-        top_padding  = 0,
-        left_padding = 0
-      )
-
-```
-
-```{r plot-res-wt, fig.cap = "Selection effects: weighted regressions"}
+```{r plot-res-wt, fig.cap = "Selection effects: weighted regressions. Unweighted estimates are shown for comparison."}
 
@@
 
+## Selection in the parents' generation
+
+The UK Biobank data contains information on respondents' number of
+siblings. Since respondents' polygenic scores are equal in expectation
+to the mean scores of their parents, we can use this to look at selection 
+effects in the parents' generation. We estimate equation (\ref{eq:regression})
+using *number of siblings* (including the respondent) as the dependent variable.
+The parents' generation has a further source of ascertainment bias: sampling
+parents of respondents overweights parents who have many children. For instance,
+parents of three children will have, on average, three times more children
+represented in UK Biobank than parents of one child. Parents of no children will
+by definition not be represented. To compensate, we reweight our preferred
+weightings (Age/Qualification) by the inverse of *number of siblings*.
+
+Selection effects are highly correlated across the two generations, and most
+share the same sign (appendix Figure \@ref(fig:plot-res-sibs)). Effect size
+estimates are larger for the parents' generation. We treat this result cautiously,
+because when we split respondents up by year of birth, we find few differences
+in effect sizes between early- and late-born respondents, for either generation.
+In other words, since estimated effect sizes change when we change the dependent
+variable, but do not change over time within either dependent variable, this
+may be due to remaining ascertainment bias within the sample.
+
+Although the direction of selection effects does not change between the
+generations, there are other differences. Compared to residualized polygenic
+scores, selection effects on unresidualized scores are about ten percent higher
+on average for *n siblings*, whereas effects for *number of children* barely
+change (appendix Figure \@ref(fig:plot-regs-controlled-pcs)). This could be
+because earlier fertility is driven more by geographically clustered deprivation
+[e.g. via an insurance motive, @rendall1998old], which may correlate with the
+broad-scale genetic variation captured by principal components.
+
+We also check whether selection effects differ by socio-economic status in the
+parents' generation. We have no information about parents' income, so we use the
+1971 Townsend deprivation score of respondents' birthplace as a proxy
+[@townsend1987deprivation]. Results (appendix Figure
+\@ref(fig:plot-siblings-townsend)) show the same pattern as for respondents:
+effect sizes are larger and more often significant in the most deprived areas.
+
+```{r calc-siblings-children}
+sib_chn_cor <- cor(famhist$n_sibs, famhist$n_children, use = "complete")
+```
+
+Lastly, the siblings data lets us check for a "quantity-quality tradeoff" between
+number of children and number of grandchildren (for the parents' generation). We
+do not find any: in fact, the correlation between *number of siblings* and 
+*number of children* is positive ($\rho = `r sib_chn_cor`$).
+
+## Economic fertility theory and natural selection
+
+These results show that selection effects are weaker, absent, or even reversed
+among some subgroups of the population. This can potentially be explained by the
+economic theory of fertility [@becker1960economic; @willis1973new;
+@becker1976child]. According to this, increases in a person's wage affect their
+fertility via two opposing channels. There is an *income effect* by which
+children become more affordable, like any other good. There is also a
+*substitution effect*: since childrearing has a cost in time, the opportunity
+cost of childrearing increases if one's market wage is higher. The income effect
+would lead higher earners to have more children. The substitution effect would
+lead higher earners to have fewer children. If so, then genetic variants which
+affect earnings potential in the labour market may cause opposing effects on
+fertility. The income effect will cause natural selection in favour of
+earnings-increasing variants. The substitution effect will do the reverse.
+
+This theory can explain why selection effects have opposite signs when AFLB is
+controlled for. Suppose that people are aware of their endowments of human
+capital, i.e. skills and characteristics that are valuable in the labour market.
+Suppose also that education and innate human capital are complements.
+Individuals choose how long to stay in education; after leaving education they
+enter the labour market and form families. If so, then people with less human
+capital will leave education earlier and have their first child earlier [cf.
+@caucutt2002women; @monstad2008education]. This is a pure substitution effect:
+the income effect plays no role in the decision to leave education, since people
+are not yet earning wages.[^no-borrow] Whilst in the labour market, higher
+earners will have more children (i.e. income effects dominate). These
+assumptions will lead to the pattern we observe: earnings-increasing variants
+will predict later first births (and fewer children overall), but they will also
+predict more children within any given age at first live birth.
+
+[^no-borrow]: This argument assumes that capital market imperfections hinder
+people from borrowing against future income.
+
+The theory can also explain why natural selection is weaker among higher-income
+parents. @becker1976child show that if parents care about child quality as well
+as quantity, under certain assumptions the income effect will be stronger at
+higher income levels; this can lead to a U-shaped relationship between income
+and fertility. Empirically, @cohen2013financial find that income decreases
+fertility at low income levels but increases it at higher income levels. If, in
+our data, at high income levels income and substitution effects are roughly
+balanced , then income-linked polygenic scores will be neither selected for nor
+against.
+
+Lastly, it is often assumed that the substitution effect will dominate for lone
+parents, or those in unstable relationships, since they have less opportunity to
+share childcare responsibilities, while the income effect will dominate for
+couples who are able to reap the gains from specialization. Indeed, US fertility
+decreases faster with education among single mothers than married mothers
+[@baudin2015fertility]. This can explain why earnings-increasing genetic variants 
+decrease fertility more among single parents than among couples. The same logic 
+could explain our results for lifetime number of sexual partners, if this is 
+associated with relationship stability.
+
+```{r plot-cors-earnings-educ, fig.cap="Selection effects, with and without controls for age at first live birth, by correlations with earnings and educational attainment.
+Each point represents one PGS."}
+
+drake::loadd(res_all)
+drake::loadd(res_age_flb)
+drake::loadd(ashe_income)
+
+famhist <- add_ashe_income(famhist, ashe_income)
+
+cors_income <- cor(famhist[score_names], famhist$first_job_pay, use = "pairwise")
+# TODO: use age_left_edu when this includes college
+
+cors_educ <-   cor(famhist[score_names], famhist$age_fulltime_edu, use = "pairwise")
+
+effect_size <-  res_all %>% 
+                  filter(dep.var == "N children", reg.type == "controlled") %>% 
+                  pull(estimate)
+
+effect_size_flb <-  res_age_flb %>% 
+                      filter(term != "age_flb") %>% 
+                      pull(estimate)
+
+dfr <- data.frame(
+        PGS          = rep(score_names, 4),
+        Correlation  = rep(c("Earnings", "Education"), each = 33, 2),
+        cor          = rep(c(cors_income[,1], cors_educ[,1]), 2),
+        effect_size  = c(rep(effect_size, 2), rep(effect_size_flb, 2)),
+        Controls     = rep(c("None", 
+                        "Age at first live birth"), each = 66)
+      )
+
+dfr %>% 
+      mutate(
+        Controls = fct_relevel(Controls, "None"),
+        PGS      = ifelse(abs(effect_size) > 0.035, PGS, ""),
+        PGS      = sub("(^\\w*?)_.*", "\\1", PGS)
+      ) %>% 
+      ggplot(aes(effect_size, cor, colour = Correlation)) + 
+        geom_point() +
+        geom_text(aes(label = PGS), size = 3, colour = "black", 
+                    check_overlap = TRUE, nudge_x = 0.005) +
+        facet_grid(Correlation ~ Controls, labeller = label_both,
+                   scales = "free") + 
+        my_vline + 
+        my_hline +
+        labs(x = "Selection effect", y = "") +
+        coord_cartesian(clip = "off") +
+        scale_colour_brewer(type = "qual", palette = 2) +
+        theme(
+          legend.position = "none",
+          panel.spacing = unit(20, "pt")
+        )
+
+```
+
+## Testing the theory
+
+We test this explanation in two ways. First, the economic theory predicts that
+genetic variants will be selected for (or against) in proportion to their effect
+on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
+on each PGS against that PGS's correlation with earnings in a respondent's first
+job, and against its correlation with educational attainment, a
+predictor of lifetime earnings. The raw relationships (left column) are strongly
+negative. If we plot selection effects controlling for age at first live birth
+(right column), the effect reverses and becomes positive. Thus, the labour
+market appears to play an important role in natural selection. Substitution
+effects dominate income effects overall, which fits the known association
+between income and lower fertility [@becker1960economic;@jones2006economic].
+
+```{r calc-res-earn-educ}
+
+drake::loadd(res_ee_control)
+drake::loadd(res_all)
+
+res_ee_control %<>% filter(term == score_name)
+n_05 <- sum(res_ee_control$p.value < 0.05)
+n_05_33 <- sum(res_ee_control$p.value < 0.05/33)
+
+```
+
+Second, we re-estimate equation \@ref(eq:regression) for each
+PGS, controlling for earnings in first job and education levels. Effect sizes
+are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only
+`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$. This suggests that
+earnings and education indeed mediate the effect of PGS on fertility.
+
+```{r calc-risk-attitude}
+
+drake::loadd(res_risk_control)
+drake::loadd(res_all)
+
+n_risk_sig <- res_risk_control %>% 
+            filter(term == 'f.2040.0.0') %>% 
+            {sum(.$p.value <= 0.05/33)}
+
+n_pgs_sig <- res_risk_control %>% 
+            filter(term == score_name) %>% 
+            {sum(.$p.value <= 0.05/33)}
+
+res_compare <- res_all %>% 
+  filter(dep.var == "N children", reg.type == "controlled") %>% 
+  left_join(res_risk_control %>% filter(term == score_name), by = "score_name")
+
+ratio <- res_compare$estimate.y/res_compare$estimate.x
+n_less_sig <- sum(res_compare$p.value.y <= 0.05/33 & res_compare$p.value.x > 0.05/33)
+stopifnot(n_less_sig == 0)
+```
+
+An alternative theory is that traits selected for are linked to externalizing
+behaviour, risk-seeking and low time discount rates, via the channel of early
+sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
+for this: scores which might plausibly be linked to externalizing behaviour,
+like ADHD and age at smoking initiation, are selected for. However, this theory
+is less good at explaining variation in selection across the full range of
+scores, including physical measures, e.g. waist-hip ratio and BMI. Externalizing
+behaviour also does not explain why selection effects should work in the
+opposite direction among older parents, whereas the economic theory can explain
+this via income effects, as described above. We test the alternative theory
+directly by re-estimating equation \@ref(eq:regression) controlling for a
+measure of risk attitude (Biobank field 2040). While risk attitude is always a
+highly significant predictor of number of children, it has little impact on the
+effect size or significance of the PGS. The median ratio of PGS effect sizes
+between regressions with and without controls is `r median(ratio)`; all PGS
+which are significant at $p \le 0.05/33$ in uncontrolled regressions remain so
+when controlling for risk attitude. Thus, although risk attitude does predict
+number of children, it appears not to mediate selection effects. Overall, we
+believe that the economic theory is the most likely explanation.
+
+# Discussion
+
+Previous work has documented natural selection in modern populations on variants
+underlying many polygenic scores [@beauchamp2016genetic; @kong2017selection;
+@sanjak2018evidence]. We show that correlations between polygenic scores and
+fertility are highly concentrated among specific subgroups of the population,
+including poorer people, less educated people, younger parents and those with
+more lifetime sexual partners. Indeed, among older parents selection effects are
+actually reversed. Furthermore, the size of selection effects on a polygenic
+score correlates with that score's association with labour market earnings. The
+economic theory of fertility provides a parsimonious explanation for this.
+
+PGS which correlate with high (low) earnings and more (less) education are being
+selected against (selected for). In addition, many of the phenotypes under
+positive selection are linked to poor health, or are what many people would see
+as undesirable to have. For example, most people would probably prefer to have
+high educational attainment, a low risk of major depressive disorder, and a low
+risk of coronary artery disease, but natural selection is pushing against
+genetics linked to these traits. Potentially, this could increase the health
+burden on modern populations, but that depends on effect sizes. Our results
+suggest that naïve estimates can be affected by sample ascertainment bias. This
+problem may be less serious in surveys which aim to be representative (as the UK
+Biobank does not). However, there is still scope for bias, since not all
+respondents consent to the collection of genetic data. For instance, completion
+rates for genotype data in the US Health and Retirement Study were around 80-85%
+[@hrs2020]. Researchers should be aware of the risks of ascertainment when
+studying modern natural selection.
+
+We also do not know how estimated effect sizes of natural selection will change
+as more accurate measures of genetic variation are produced. And we are unsure
+whether genetic variants underlying other phenotypes will show a similar pattern
+of natural selection to those studied here. In short, it is probably too early
+to tell whether modern natural selection has a substantively important effect on
+population genetics. Nevertheless, selection effects on our measured
+polygenic scores are still relatively small, even after reweighting to 
+account for ascertainment bias.
+
+Because selection effects are concentrated in poorer groups, they also increase
+inequality with respect to polygenic scores. For example, Figure
+\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
+attainment (EA3), for children in households of different income groups. The
+blue bars show the actual means, i.e. parents' mean polygenic score weighted by
+number of children. The grey bars show the hypothetical means if all households
+had equal numbers of children. Natural selection against EA3, which lowers the
+mean, is stronger at the bottom of the income distribution, and this increases
+the differences between groups.
+
+```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean PGS for educational attainment (EA3) of children by household income group.
+Blue is actual.
+Grey is hypothetical in the absence of selection effects.", fig.width = 4}
+income_EA3 <- famhist %>% 
+      filter(! is.na(n_children), ! is.na(income_cat)) %>% 
+      mutate(
+        `Household income` = factor(income_cat, 
+                          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", 
+                                      "> £100K"))
+      ) %>% 
+      group_by(`Household income`) %>% 
+      summarize(
+        `Without selection` = mean(EA3_excl_23andMe_UK, na.rm = TRUE),
+        Actual              = weighted.mean(EA3_excl_23andMe_UK, n_children, 
+                                              na.rm = TRUE)
+      ) %>% 
+      tidyr::pivot_longer(-`Household income`) %>% 
+      rename(
+        `Children's mean EA3`     = value
+      )
+
+income_EA3 %>% 
+              ggplot(aes(`Household income`, `Children's mean EA3`, fill = name)) +
+                geom_col(position = "dodge") +
+                scale_fill_manual(values = c("navy", "grey60")) +
+                theme(
+                  legend.background = element_rect(fill = "white", colour = NA),
+                  legend.title = element_blank(), 
+                  legend.position = c(.25, .85)
+                )
+```
+
+Any model of fertility is implicitly a model of natural selection. But so far,
+the economic and human genetics literatures have developed in parallel.
+Integrating the two could deepen our understanding of modern natural selection.
+Economics possesses a range of theoretical models on the effects of skills,
+education and income [see @hotz1997economics, @lundberg2007american]. One
+perennial problem is how to test these theories in a world where education,
+labour and marriage markets all interact. Genetic data, such as polygenic
+scores, could help to pin down causality. Conversely, theory and empirical
+results from economics can shine a light on the mechanisms behind natural
+selection.
+
+# Materials and methods
+
+We use participant data from UK Biobank [@sudlow2015uk], which has received
+ethical approval from the National Health Service North
+West Centre for Research Ethics Committee (reference: 11/NW/0382).
+We limit the sample to white British participants, as defined by
+genetic estimated ancestry and self-identified ethnic group, giving a sample
+size of `r comma_num(nrow(famhist))`. For regressions on number of children we use participants over 45, since most fertility is completed by this age. This
+gives a sample size of `r comma_num(nrow(famhist %>% filter(kids_ss)))`.
+
+We computed 33 polygenic scores. The effect size estimates were obtained from
+GWASs that were chosen to not have included the UK Biobank dataset, to avoid
+overestimation of the genetic predisposition of a trait. The polygenic scores
+were computed using the summary-data based best linear unbiased prediction
+(SBLUP) approach. As a reference sample for the linkage disequilibrium, we used
+a random sample of 10,000 unrelated individuals from UK Biobank, imputed using
+the Haplotype Reference Consortium panel. Scores were normalized to mean 0,
+variance 1, and residualized on the first 100 principal components of array
+data, which were calculated as in @abdellaoui2019genetic.
+
+Earnings in first job are estimated from mean earnings in the 2007 Annual Survey
+of Hours and Earnings, using the SOC 2000 job code (Biobank field 22617).
+
+Population data for weighting is taken from the 2011 UK Census and the 2006
+General Household Survey (GHS). Weighting for Age/Qualification and 
+Age/Qualification/AFLB weights was done using marginal totals from a linear
+model, using the `calibrate()` function in
+the R "survey" package [@lumley2020survey]. Geographical weighting was done with
+iterative post-stratification using the `rake()` function, on Census Middle Layer
+Super Output Areas, sex and presence/absence of a partner.
+
+\FloatBarrier
+\newpage
+# Appendix
+
+## Natural selection by sex
+
+```{r plot-sexes, fig.cap = "Selection effects by sex. Lines show significant differences."}
+
+drake::loadd(res_sex)
+n_regs <- as.double(nrow(res_sex))
+
+res_sex_wide <- res_sex %>% 
+      select(term, estimate, sex, diff.p.value) %>% 
+      pivot_wider(names_from = "sex", values_from = "estimate") %>% 
+      mutate(
+        min_est = pmin(Female, Male),
+        max_est = pmax(Female, Male),
+        term = pretty_names(term),
+        term = fct_reorder(term, Female)
+      )
+
+res_sex %>% 
+      mutate(
+        term = pretty_names(term),
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(y = term, x = estimate, fill = sex, 
+            colour = p < 0.05/{{n_regs}})) +
+        my_vline +
+        geom_linerange(data = res_sex_wide, 
+          aes(
+            y      = term,
+            x      = NULL,
+            fill   = NULL,
+            xmin   = min_est, 
+            xmax   = max_est, 
+            colour = diff.p.value < 0.05/{{n_regs}}
+          ), 
+        group = "") +
+        pval_geom_point() +
+        pval_scale() +
+        my_fill_scale(n = 2) +
+        my_labs() +
+        my_theme()
+
+```
+
+\FloatBarrier
+## Controlling for AFLB
+
+```{r plot-age-flb, fig.cap = "Selection effects controlling for age at first live birth (women only). Effect sizes for women only without controls are shown for comparison."}
+
+n_regs <- as.double(nrow(res_age_flb))
+
+res_combined %>% 
+      mutate(
+        score_name = pretty_names(score_name),
+        score_name = fct_reorder(score_name, estimate, order_abs(2)),
+        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for AFLB")
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
+        color = p < 0.05/{{n_regs}})) + 
+        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
+        pval_geom_point() +
+        pval_scale() +
+        scale_fill_brewer(type = "qual", palette = 3) +
+        my_labs() +
+        my_theme()
+```
+
+\FloatBarrier
+## Weighted regressions
+
+Table \@ref(tab:tbl-res-wt) gives effect sizes as a proportion of the unweighted
+effect size, for all polygenic scores which are consistently signed and which
+are significantly different from zero in unweighted regressions.
+
+```{r tbl-res-wt}
+
+rel_sizes %>% 
+      arrange(desc(Geographical)) %>% 
+      bind_rows(rel_sizes_summary) %>% 
+      rename(PGS = term) %>% 
+      as_hux() %>% 
+      insert_row("", "Weighting", fill = "") %>% 
+      set_caption(
+        "Weighted effect sizes as a proportion of unweighted effect sizes."
+      ) %>% 
+      merge_cells(1, 2:4) %>% 
+      set_align(1, 2, "centre") %>% 
+      set_header_rows(1, TRUE) %>% 
+      theme_compact() %>% 
+      set_number_format(-1, -1, 2) %>% 
+      style_headers(bold = TRUE) %>% 
+      set_bottom_border(1, 2:4) %>% 
+      set_bottom_border(2, everywhere) %>% 
+      set_italic(final(2), 1) %>% 
+      set_font_size(8) %>% 
+      set_width(0.7) %>% # ensures line wrapping below
+      set_col_width(c(.34, .22, .22, .22)) %>% 
+      add_footnote(paste0(
+        "Only consistently-signed and significant (when unweighted) ",
+        "estimates are shown. ",
+        "Age/Qual/AFLB as a proportion of unweighted regressions ",
+        "including females only."),
+        top_padding  = 0,
+        left_padding = 0
+      )
+
+```
+
 \FloatBarrier
@@
 Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings* on
-polygenic scores. For comparison, we show regressions on the respondents'
-generation (*number of children*), where respondents with no
-children have been excluded.
+polygenic scores. By definition, people who had no children cannot be included in
+this data. For a cleaner comparison, we rerun regressions on the respondents'
+generation (*number of children*) after excluding respondents with no
+children.
 
@@
       count() %>% 
-      period_table("Change in selection effects between parents of early and late born respondents (regressions on number of siblings)")
+      period_table("Change in selection effects between parents of early and late born respondents (regressions on number of siblings).")
 
@@
       count() %>% 
-      period_table("Change in selection effects between early and late born respondents (regressions on number of children)")
+      period_table("Change in selection effects between early and late born respondents (regressions on number of children).")
+```
+
+Figure \@ref(fig:plot-siblings-townsend) plots effects on *n siblings* by
+Townsend deprivation quintile of birth area.
+
+```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by Townsend deprivation quintile of birth area", fig.align = "center"}
+
+drake::loadd(res_townsend_parents)
+drake::loadd(famhist_townsend_71)
+n_regs <- nrow(res_townsend_parents)
+
+res_townsend_parents %>% 
+  mutate(
+    `Townsend quintile` = sub("Quin71(\\d):.*", "\\1", term),
+    `Townsend quintile` = add_n(`Townsend quintile`, "Quin71", dv = "n_sibs", 
+                                  data = famhist_townsend_71)
+  ) %>% 
+  standard_ggplot(fill_col    = `Townsend quintile`,
+                    n_regs    = n_regs, 
+                    score_col = score_name, 
+                    order_idx = 5
+                  )
+
+rm(famhist_townsend_71)
 ```
@@
 \FloatBarrier
-## Controlling for principal components
-
-Polygenic scores could capture effects that are really due to population
-stratification, although would not change our results for natural
-selection of the scores. Figure \@ref(fig:plot-regs-controlled-pcs) shows
-selection effects on polygenic scores residualized for the top 100
-principal components of the genetic data, calculated within the UK
-Biobank population.
+## Selection effects on raw polygenic scores
 
@@
 
-In siblings regressions, effect sizes are smaller when residualizing for
-principal components -- sometimes much smaller, as in the case of
-height. `r sum(controlled_smaller_sibs)` out of `r n_pgs` "controlled"
-effect sizes have a smaller absolute value than the corresponding "raw"
+Figure \@ref(fig:plot-regs-controlled-pcs) compares
+selection effects on polygenic scores residualized for the top 100
+principal components of the genetic data, to selection effects on raw, unresidualized polygenic
+scores. In siblings regressions, effect sizes are larger for raw scores
+principal components -- sometimes much larger, as in the case of
+height. `r sum(controlled_smaller_sibs)` out of `r n_pgs` "raw"
+effect sizes have a larger absolute value than the corresponding "residualized"
 effect size. The median proportion between raw and controlled effect
@@
 
-```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using residualized polygenic scores on number of siblings/children."}
+```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores on number of siblings/children."}
 
@@
         term     = reorder_within(term, estimate, dep.var, order_abs(1)),
-        dep.var = fct_inorder(dep.var) 
+        dep.var  = fct_inorder(dep.var),
+        reg.type = fct_recode(reg.type, 
+                               Residualized   = "controlled", 
+                               Unresidualized = "raw"
+                             )
       ) %>% 
@@
         facet_wrap(vars(dep.var), scales = "free_y") +
-        pval_geom_point(position = position_dodge(0.4)) +
+        pval_geom_point(position = position_dodge(0.4), alpha = 0.8) +
         my_labs() +
@@
         my_theme() +
-        scale_y_reordered()
+        scale_y_reordered() +
+        theme(
+          axis.text.x = element_text(size = 8)
+        )
 
@@
 
-To get a further insight into this we regress *siblings* and *children*
+To get a further insight into this we regress *n siblings* and *n children*
 on individual principal components. As Figure \@ref(fig:plot-pc-regs)
@@
 
-res_sibs_raw <- filter(res_all, dep.var == "N siblings", reg.type == "raw")
-compare_abp_raw <- res_age_birth_parents %>% 
+res_sibs_unc <- filter(res_all, dep.var == "N siblings", reg.type == "controlled")
+compare_abp_unc <- res_age_birth_parents %>% 
       filter(! term %in% c("fath_age_birth", "moth_age_birth")) %>% 
-      left_join(res_sibs_raw, by = "term")
-
-cor_raw_fath <- with(
-  compare_abp_raw %>% filter(control == "fath_age_birth"), 
+      left_join(res_sibs_unc, by = "term")
+
+cor_unc_fath <- with(
+  compare_abp_unc %>% filter(control == "fath_age_birth"), 
   cor(estimate.x, estimate.y))
-cor_raw_moth <- with(
-  compare_abp_raw %>% filter(control == "moth_age_birth"), 
+cor_unc_moth <- with(
+  compare_abp_unc %>% filter(control == "moth_age_birth"), 
   cor(estimate.x, estimate.y))
@@
 regressions without the age at birth control, though the correlation
-is smaller (father's age at birth: $\rho$ `r cor_raw_fath`; mother's age at
-birth: $\rho$ `r cor_raw_moth`).
+is smaller (father's age at birth: $\rho$ `r cor_unc_fath`; mother's age at
+birth: $\rho$ `r cor_unc_moth`).
 
@@
 res_all %<>% 
-  filter(term == score_name, reg.type == "raw", dep.var == "N children") %>% 
+  filter(term == score_name, reg.type == "controlled", dep.var == "N children") %>% 
   select(-dep.var, -reg.type)
@@
 
-TODO: ref source of data.
-
 ```{r plot-rgs-by-effect-size, fig.cap = "Selection effects plotted against genetic correlation with EA3."}
+
+# TODO: ref source of data.
 
@@
 rgs <- left_join(rgs, 
-        res_all %>% filter(reg.type == "raw", dep.var == "N children"), 
+        res_all %>% filter(reg.type == "controlled", dep.var == "N children"), 
         by = c("p2" = "term"))
@@
 
-```{r correlate-effect-sizes, eval = FALSE}
-combined_results <- res %>% 
-      mutate(PCs = ifelse(reg.type=="raw", "no", "yes")) %>% 
-      dplyr::select(-reg.type) %>% 
-      {bind_rows(bivariate = ., multivariate = all_res, .id = "reg.type")} %>% 
-      arrange(term, reg.type, PCs) %>% 
-      dplyr::select(term, reg.type, PCs, estimate) %>% 
-      pivot_wider(names_from = c(reg.type, PCs), values_from = estimate)
-
-cor(combined_results[-1], use = "complete")
-
-```
-
 \FloatBarrier
@@
 
-Table \@ref(tab:tbl-inequality-effects) estimates differences in children's mean polygenic
-scores between the highest and lowest income groups. Column "Actual" shows parents' scores weighted by Age/Qualification times number of children. 
-Column "Without selection" shows scores weighted by Age/Qualification only, i.e. if all couples had the same number of children.
+Table \@ref(tab:tbl-inequality-effects) estimates differences in children's mean
+polygenic scores between the highest and lowest income groups. Column "Actual"
+shows respondents' scores weighted by Age/Qualification times number of
+children. Column "Without selection" shows scores weighted by Age/Qualification
+only, i.e. if all couples had the same number of children.
 
@@
       ) %>% 
-      arrange(desc(Actual)) %>% 
+      arrange(desc(`With selection`)) %>% 
       as_hux() %>% 
       set_align(everywhere, -1, "right") %>% 
-      set_caption("Differences in polygenic scores between highest and lowest income group") %>% 
+      set_number_format(everywhere, -1, 3) %>% 
+      set_caption("Differences in polygenic scores between highest and lowest income group.") %>% 
       theme_compact()
@@
 \FloatBarrier
-
+\newpage
 # References

### 2020-12-10 _drake.R

@@ -50,14 +50,27 @@ plan <- drake_plan(
   
   ashe_income  = import_ashe_income(file_in(!! ashe_income_file)),
   
-  famhist      =  target({
-                      famhist <- clean_famhist(famhist_raw, score_names, sib_groups)
-                      famhist$kids_ss <- famhist$age_at_recruitment >= 45
-                      famhist
+  famhist_no_resid      =  target({
+                      fnr <- clean_famhist(famhist_raw, score_names, sib_groups)
+
+                      fnr$kids_ss <- fnr$age_at_recruitment >= 45
+                      fnr
                     },
                     format = "fst_tbl"
                   ),
   
+  famhist = target({
+                     famhist <- join_famhist_resid_scores(famhist_no_resid, 
+                                                            resid_scores)
+                     # use residualized scores as defaults:
+                     names(famhist)[names(famhist) %in% score_names] %<>%
+                       paste0("_raw")
+                     names(famhist) <- gsub("_resid$", "", names(famhist))
+                     famhist
+                   },
+                    format = "fst_tbl"
+                  ),
+
   famhist_msoa = target(
                    find_containing_msoas(famhist, msoa_shapefile), 
                    format = "fst_tbl"
@@ -73,8 +86,11 @@ plan <- drake_plan(
   # fhl_mlogit   =  make_famhist_long_mlogit(famhist, score_names),
   
   resid_scores = target({
-                     resid_scores <- compute_resid_scores(famhist, pcs, score_names)
-                     resid_scores <- subset_resid_scores(resid_scores, famhist, score_names)
+                     resid_scores <- compute_resid_scores(famhist_no_resid, 
+                                                            pcs, score_names)
+                     resid_scores <- subset_resid_scores(resid_scores, 
+                                                           famhist_no_resid, 
+                                                           score_names)
                      resid_scores
                    },
                    format = "fst_tbl"
@@ -113,7 +129,7 @@ plan <- drake_plan(
   
   mf_pairs =  target({
                 mf_pairs <- make_mf_pairs(file_in(!! mf_pairs_file), famhist, 
-                                            resid_scores, ashe_income)
+                                            resid_scores = NULL, ashe_income)
                 mf_pairs <- filter_mf_pairs(mf_pairs)
                 mf_pairs
               },
@@ -123,7 +139,6 @@ plan <- drake_plan(
[…]

### 2020-12-10 make-data.R

@@ -21,7 +21,7 @@ join_famhist_pcs <- function (famhist, pcs) {
 
 
 join_famhist_resid_scores <- function (famhist, resid_scores) {
-  cbind(famhist, resid_scores)
+  cbind(famhist, resid_scores %>% select(-f.eid))
 }
 
 

### 2020-12-10 manual-result-tests.R

@@ -26,30 +26,30 @@ test_that("res_all", {
     res_all %>% 
       filter(dep.var == "N children", reg.type == "raw") %>% 
       estimate_for("ADHD_2017"),
-    lm_coef(n_children ~ ADHD_2017)
+    lm_coef(n_children ~ ADHD_2017_raw)
   )
   
   expect_equivalent(
     res_all %>% 
       filter(dep.var == "N siblings", reg.type == "raw") %>% 
       estimate_for("alcohol_schumann"),
-    lm_coef(n_sibs ~ alcohol_schumann, data = famhist)
+    lm_coef(n_sibs ~ alcohol_schumann_raw, data = famhist)
   )
   
   
-  fh2 <- cbind(famhist, resid_scores)
+  fh2 <- cbind(famhist, resid_scores[-1])
   expect_equivalent(
     res_all %>% 
       filter(dep.var == "N children", reg.type == "controlled") %>% 
       estimate_for("height_combined"),
-    lm_coef(n_children ~ height_combined_resid, data = fh2 %>% filter(kids_ss))
+    lm_coef(n_children ~ height_combined, data = fh2 %>% filter(kids_ss))
   )
   
   expect_equivalent(
     res_all %>% 
       filter(dep.var == "N siblings", reg.type == "controlled") %>% 
       estimate_for("height_combined"),
-    lm_coef(n_sibs ~ height_combined_resid, data = fh2)
+    lm_coef(n_sibs ~ height_combined, data = fh2)
   )
 })
 
@@ -57,7 +57,7 @@ test_that("res_all", {
 test_that("res_pcs", {
   loadd(res_pcs)
   loadd(pcs)
-  fh <- left_join(famhist, pcs, by = c("f.eid" = "IID"))
+  fh <- left_join(famhist, pcs, by = c("f.eid" = "eid"))
   expect_equivalent(
     res_pcs %>% filter(dep.var == "N children", term == "PC33") %>% pull(estimate),
     lm_coef(n_children ~ PC33, data = fh %>% filter(kids_ss))
@@ -409,7 +409,12 @@ test_that("res_partners_controlled", {
 
 test_that("res_ee_control", {
   loadd(res_ee_control)
-  coef <- lm_coef(n_children ~ dpw_substance_use + age_fte_cat + first_job_pay)
+  loadd(ashe_income)
+  source("~/import-ukbb-data/import-ukbb-data.R")
+  fh <- add_ashe_income(famhist, ashe_income)
[…]

### 2020-12-10 regressions.R

@@ -59,17 +59,16 @@ calc_pgs_over_time <- function (famhist, score_names) {
 run_regs_basic <- function (dep_var, score_names, famhist, subset = NULL) {
   
   run_reg <- function (score_name) {
-    fml <- as.formula(glue("{dep_var} ~ {score_name}"))
+    fml <- as.formula(glue("{dep_var} ~ {score_name}_raw"))
     reg_bv <- tidy(lm(fml, famhist, subset = eval(subset)), conf.int = TRUE) 
+    reg_bv <- mutate(reg_bv, term = gsub("_raw", "", term))
     reg_bv
   }
   basic_res <- map_dfr(score_names, run_reg, .id = "score_name")
   
   run_resid_reg <- function (score_name) {
-    score_resid <- paste0(score_name, "_resid")
-    fml_resid <- as.formula(glue("{dep_var} ~ {score_resid}"))
+    fml_resid <- as.formula(glue("{dep_var} ~ {score_name}"))
     reg_resid <- tidy(lm(fml_resid, famhist, subset = eval(subset)), conf.int = TRUE) 
-    reg_resid <- mutate(reg_resid, term = gsub("_resid", "", term))
     reg_resid
   }
   resid_res <- map_dfr(score_names, run_resid_reg, .id = "score_name")

### 2020-12-16 why-negative-selection.Rmd
Abdel's comments, restyling

@@
 title: "Labor Market Income and Substitution Effects Explain Modern Natural Selection"
-author: "David Hugh-Jones (D.Hugh-Jones@uea.ac.uk), Abdel Abdellaoui (a.abdellaoui@gmail.com)"
+author: "David Hugh-Jones \\thanks{Corresponding author.
+School of Economics, University of East Anglia.
+Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Amsterdam UMC.
+Email: a.abdellaoui@amsterdamumc.nl}"
 date: "`r format(Sys.Date(), '%d %B %Y')`"
 abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it.
-We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank.
+We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry).
 Consistently over time, polygenic scores which predict lower (higher) earnings and education are selected for (against).
@@
 The direction of natural selection is reversed among older parents, or after controlling for age at first live birth.
-These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market."
+These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market.
+We find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects and that are strongly related to the socio-economic system."
 bibliography: "negative-selection.bib"
@@
   - \usepackage{placeins}
+  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
+  - \usepackage{titlesec}
+  - \titleformat*{\section}{\bfseries\sffamily\LARGE}
+  - \titleformat*{\subsection}{\bfseries\sffamily\itshape\Large}
+  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
 editor_options: 
@@
 
-As yet we know little about the mechanisms behind these effects. This study uses
-data from UK Biobank to learn more. We test for natural selection on 33
+As yet we know little about the social mechanisms behind these effects. This study uses
+data from UK Biobank [@bycroft2018uk] to learn more. We test for natural selection on 33
 different polygenic scores by estimating their correlation with fertility. We
@@
 
-We find selection effects on many polygenic scores. Effects are consistent
+We find selection effects on many polygenic scores. Effects are largely consistent
 across generations. The strength of natural selection on a polygenic score is
@@
 the board, polygenic scores have stronger relationships with fertility among
-specific subgroups. Selection effects are stronger among people with lower
-income and less education, younger parents, people not living with a partner,
+specific subgroups. Selection effects are stronger in groups with lower
+income and less education, among younger parents, people not living with a partner,
 and people with many lifetime sexual partners. Outside these groups, effects are
-weaker and often statistically insignificant. In fact, in some subgroups, the
-direction of selection is actually reversed: polygenic scores predicting
+weaker and often statistically insignificant. In some subgroups, the
+direction of selection is even reversed: polygenic scores predicting
 higher education and earnings are associated with *higher* fertility.
 
-These patterns can be explained by economic theories of fertility
+These patterns are in line with economic theories of fertility
 [@becker1960economic]. In these, higher potential earnings have two opposite
@@
 declines over this 30-year period, of the order of 5% of a standard
-deviation.These changes could reflect natural selection within the UK
-population, but also ascertainment bias within the sample. Respondents are
-higher income and better educated than the UK population, and they may
-also differ on other unobserved characteristics [@10.1093/aje/kwx246].
-Since richer and educated people also live longer, this bias might also
+deviation. These changes could reflect natural selection within the UK
+population, but also emigration, or ascertainment bias within the sample. 
+Respondents have higher income and are better educated than the UK population, 
+and they may also differ on other unobserved characteristics [@10.1093/aje/kwx246].
+Since richer and educated people also live longer, this bias could also
 increase with age.
 
-```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores by birth year in UK Biobank.
+```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank.
 Points are means for 5-year intervals.
@@
 
-To test directly for natural selection, we regress respondents'polygenic scores
-on their number of children ($y_i$):
+To test directly for natural selection, we regress respondents' polygenic scores
+(PGS) on their number of children ($y_i$):
 
@@
 
-The "selection effect", $\beta$, measures the strength of natural selection
+The "selection effect", $\beta$, reflects the strength of natural selection
 within the sample. In fact, since polygenic scores are normalized, $\beta$ is
@@
 generation.
+
+```{r plot-income-educ-level, fig.cap = "Selection effects by education and income.
+Each point represents a single bivariate regression of number of children on a PGS", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}
+
+drake::loadd(res_edu)
+n_regs <- as.double(nrow(res_edu))
+
+res_edu %>% 
+      mutate(
+        term = pretty_names(term), 
+        term = fct_reorder(term, estimate, order_abs(1)),
+        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
+        "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(y = term, x = estimate, colour = p < 0.05/{{n_regs}}, 
+        fill = `Age left FTE`)) +
+        my_vline +
+        pval_geom_point() + 
+        pval_scale() +
+        my_fill_scale(n = 3, direction = -1) +
+        my_labs() +
+        my_theme()
+
+drake::loadd(res_income)
+
+n_regs <- as.double(nrow(res_income))
+res_income %>% 
+      mutate(
+        term = pretty_names(term), 
+        term = fct_reorder(term, estimate, order_abs(1)),
+        Income = factor(income_cat, 
+          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
+        ),
+        Income = add_n(Income, "income_cat")
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(y = term, x = estimate, fill = Income, 
+        colour = p < 0.05/{{n_regs}})) +
+        my_vline +
+        pval_geom_point(alpha = 0.8) + 
+        my_fill_scale(n = 5, direction = -1) +
+        pval_scale() +
+        my_labs() +
+        my_theme()
+```
 
@@
 
-```{r plot-income-educ-level, fig.cap = "Selection effects by education and income.
-Each point represents a single bivariate regression of number of children on a PGS", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}
-
-drake::loadd(res_edu)
-n_regs <- as.double(nrow(res_edu))
-
-res_edu %>% 
-      mutate(
-        term = pretty_names(term), 
-        term = fct_reorder(term, estimate, order_abs(1)),
-        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
-        "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(y = term, x = estimate, colour = p < 0.05/{{n_regs}}, 
-        fill = `Age left FTE`)) +
-        my_vline +
-        pval_geom_point() + 
-        pval_scale() +
-        my_fill_scale(n = 3, direction = -1) +
-        my_labs() +
-        my_theme()
-
-drake::loadd(res_income)
-
-n_regs <- as.double(nrow(res_income))
-res_income %>% 
-      mutate(
-        term = pretty_names(term), 
-        term = fct_reorder(term, estimate, order_abs(1)),
-        Income = factor(income_cat, 
-          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
-        ),
-        Income = add_n(Income, "income_cat")
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(y = term, x = estimate, fill = Income, 
-        colour = p < 0.05/{{n_regs}})) +
-        my_vline +
-        pval_geom_point(alpha = 0.8) + 
-        my_fill_scale(n = 5, direction = -1) +
-        pval_scale() +
-        my_labs() +
-        my_theme()
-```
-
 ```{r calc-income-edu-controlled}
@@
 
-These results could be driven by age, if older respondents are poorer
-and less educated, and also more subject to selection on polygenic
+These results could be driven by age, if older respondents have lower income
+and are less educated, and also more subject to selection on polygenic
 scores. However, if we rerun the regressions, interacting the polygenic
@@
 educational attainment are more selected against, though the difference
-is only significant for EA3. One possible reason for these sex differences is
+is only significant for the educational attainment PGS of @lee2018gene
+("EA3"). One possible reason for these sex differences is
 that polygenic scores may affect fertility via success in marriage markets, and
@@
 mothers. The correlation between effect sizes for the youngest and
-oldest terciles is `r cor_age_flb`. As before, recall that PGS may have
-two effects: changing the number of children within each category, and changing
-the number of children by shifting respondents between categories.
+oldest terciles is `r cor_age_flb`. As before, recall that polygenic scores may 
+have two effects: changing the number of children within each category, and 
+changing the number of children by shifting respondents between categories.
 
 [^1]: This information is unavailable for men.
+
+\newgeometry{left=1cm,right=1cm,top=1cm,bottom=1cm}
 
@@
 
+\restoregeometry
+
 ```{r calc-corr-age-flb}
@@
 To investigate this further, we regress number of children on PGS
-*controlling* for AFLB (appendix Figure \@ref(fig:plot-age-flb)). In
+*controlling* for AFLB (Figure \@ref(fig:plot-age-flb)). In
 `r opp_signed` out of `r length(score_names)` cases, effects
@@
 controlling for AFLB.
+
+```{r plot-age-flb, fig.cap = "Selection effects controlling for age at first live birth (women only). Effect sizes without controls (for women only) are shown for comparison."}
+
+n_regs <- as.double(nrow(res_age_flb))
+
+res_combined %>% 
+      mutate(
+        score_name = pretty_names(score_name),
+        score_name = fct_reorder(score_name, estimate, order_abs(2)),
+        Regression = ifelse(reg.type == "raw", "Uncontrolled", "Controlled for AFLB")
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
+        color = p < 0.05/{{n_regs}})) + 
+        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
+        pval_geom_point() +
+        pval_scale() +
+        scale_fill_brewer(type = "qual", palette = 3) +
+        my_labs() +
+        my_theme()
+```
 
@@
 
-n_regs <- length(score_names)
+n_regs <- as.double(length(score_names))
+
 res_wt_combined %>% 
@@
 using *number of siblings* (including the respondent) as the dependent variable.
-The parents' generation has a further source of ascertainment bias: sampling
+The parents' generation has an additional source of ascertainment bias: sampling
 parents of respondents overweights parents who have many children. For instance,
@@
 Selection effects are highly correlated across the two generations, and most
-share the same sign (appendix Figure \@ref(fig:plot-res-sibs)). Effect size
+share the same sign (Appendix Figure \@ref(fig:plot-res-sibs)). Effect size
 estimates are larger for the parents' generation. We treat this result cautiously,
@@
 on average for *n siblings*, whereas effects for *number of children* barely
-change (appendix Figure \@ref(fig:plot-regs-controlled-pcs)). This could be
+change (Appendix Figure \@ref(fig:plot-regs-controlled-pcs)). This could be
 because earlier fertility is driven more by geographically clustered deprivation
 [e.g. via an insurance motive, @rendall1998old], which may correlate with the
-broad-scale genetic variation captured by principal components.
+broad-scale genetic variation captured by principal components, such as ancestry.
 
@@
 1971 Townsend deprivation score of respondents' birthplace as a proxy
-[@townsend1987deprivation]. Results (appendix Figure
+[@townsend1987deprivation]. Results (Appendix Figure
 \@ref(fig:plot-siblings-townsend)) show the same pattern as for respondents:
@@
 ```{r calc-siblings-children}
-sib_chn_cor <- cor(famhist$n_sibs, famhist$n_children, use = "complete")
+sib_chn_cor <- cor.test(famhist$n_sibs, famhist$n_children, use = "complete")
+stopifnot(sib_chn_cor$p.value < 2e-16)
 ```
@@
 do not find any: in fact, the correlation between *number of siblings* and 
-*number of children* is positive ($\rho = `r sib_chn_cor`$).
+*number of children* is positive ($\rho$ = `r sib_chn_cor$estimate`, 
+$p < 2 \times 10^{-17}$).
 
@@
 economic theory of fertility [@becker1960economic; @willis1973new;
-@becker1976child]. According to this, increases in a person's wage affect their
+@becker1976child]. According to this theory, increases in a person's wage affect their
 fertility via two opposing channels. There is an *income effect* by which
@@
 
-The theory can also explain why natural selection is weaker among higher-income
-parents. @becker1976child show that if parents care about child quality as well
-as quantity, under certain assumptions the income effect will be stronger at
-higher income levels; this can lead to a U-shaped relationship between income
-and fertility. Empirically, @cohen2013financial find that income decreases
-fertility at low income levels but increases it at higher income levels. If, in
-our data, at high income levels income and substitution effects are roughly
-balanced , then income-linked polygenic scores will be neither selected for nor
-against.
+The theory can also explain why the natural selection pressure is weaker among
+higher-income parents. @becker1976child show that if parents care about child
+quality as well as quantity, under certain assumptions the income effect will be
+stronger at higher income levels; this can lead to a U-shaped relationship
+between income and fertility. Empirically, @cohen2013financial find that income
+decreases fertility at low income levels but increases it at higher income
+levels. If, in our data, at high income levels income and substitution effects
+are roughly balanced , then income-linked polygenic scores will be neither
+selected for nor against.
 
@@
 Second, we re-estimate equation \@ref(eq:regression) for each
-PGS, controlling for earnings in first job and education levels. Effect sizes
-are generally reduced (appendix Figure \@ref(fig:plot-res-earn-educ)), and only
+polygenic score, controlling for earnings in first job and education levels. Effect sizes
+are generally reduced (Appendix Figure \@ref(fig:plot-res-earn-educ)), and only
 `r n_05_33` out of 33 PGS are significant at $p = 0.05/33$. This suggests that
-earnings and education indeed mediate the effect of PGS on fertility.
+earnings and education indeed mediate the effect of polygenic scores on fertility.
 
@@
 directly by re-estimating equation \@ref(eq:regression) controlling for a
-measure of risk attitude (Biobank field 2040). While risk attitude is always a
+measure of risk attitude (UK Biobank field 2040). While risk attitude is always a
 highly significant predictor of number of children, it has little impact on the
@@
 Previous work has documented natural selection in modern populations on variants
-underlying many polygenic scores [@beauchamp2016genetic; @kong2017selection;
+underlying polygenic traits [@beauchamp2016genetic; @kong2017selection;
 @sanjak2018evidence]. We show that correlations between polygenic scores and
 fertility are highly concentrated among specific subgroups of the population,
-including poorer people, less educated people, younger parents and those with
-more lifetime sexual partners. Indeed, among older parents selection effects are
+including people with lower income, lower education, younger first parenthood, 
+and more lifetime sexual partners. Indeed, among older parents, selection effects are
 actually reversed. Furthermore, the size of selection effects on a polygenic
@@
 
-PGS which correlate with high (low) earnings and more (less) education are being
+Polygenic scores which correlate with high (low) earnings and more (less) education are being
 selected against (selected for). In addition, many of the phenotypes under
-positive selection are linked to poor health, or are what many people would see
+positive selection are linked to disease risk, or are what many people would see
 as undesirable to have. For example, most people would probably prefer to have
@@
 risk of coronary artery disease, but natural selection is pushing against
-genetics linked to these traits. Potentially, this could increase the health
+genes associated with these traits. Potentially, this could increase the health
 burden on modern populations, but that depends on effect sizes. Our results
@@
 whether genetic variants underlying other phenotypes will show a similar pattern
-of natural selection to those studied here. In short, it is probably too early
-to tell whether modern natural selection has a substantively important effect on
-population genetics. Nevertheless, selection effects on our measured
-polygenic scores are still relatively small, even after reweighting to 
-account for ascertainment bias.
-
-Because selection effects are concentrated in poorer groups, they also increase
+of natural selection to those studied here. In addition, genetic effects on
+educational attainment have been shown to be inflated in population-based
+samples as compared to within-family designs, likely because of indirect genetic 
+effects, gene-environment correlations, and/or assortative mating 
+[@lee2018gene; @selzam2019comparing; @kong2018nature]. 
+In short, it is probably too early to tell whether modern natural selection has a
+substantively important effect on the genetic make-up of the population.
+Nevertheless, we note that selection effects on our measured polygenic scores
+are still relatively small, even after reweighting to account for ascertainment
+bias.
+
+Because selection effects are concentrated in poorer groups, they may also increase
 inequality with respect to polygenic scores. For example, Figure
@@
 number of children. The grey bars show the hypothetical means if all households
-had equal numbers of children. Natural selection against EA3, which lowers the
-mean, is stronger at the bottom of the income distribution, and this increases
-the differences between groups.
+had equal numbers of children. Natural selection against genes associated with 
+educational attainment, which lowers the mean, is stronger at the bottom of the 
+income distribution, and this increases the differences between groups.
 
@@
 
-Any model of fertility is implicitly a model of natural selection. But so far,
+Any model of fertility is implicitly a model of natural selection, but so far,
 the economic and human genetics literatures have developed in parallel.
-Integrating the two could deepen our understanding of modern natural selection.
+Integrating the two could deepen our understanding of natural selection in 
+modern societies.
 Economics possesses a range of theoretical models on the effects of skills,
-education and income [see @hotz1997economics, @lundberg2007american]. One
+education and income [see @hotz1997economics; @lundberg2007american]. One
 perennial problem is how to test these theories in a world where education,
 labour and marriage markets all interact. Genetic data, such as polygenic
-scores, could help to pin down causality. Conversely, theory and empirical
-results from economics can shine a light on the mechanisms behind natural
-selection.
+scores, could help to pin down directions of causality. Conversely, theory 
+and empirical results from economics can shine a light on the mechanisms 
+behind natural selection.
 
@@
 
-We use participant data from UK Biobank [@sudlow2015uk], which has received
-ethical approval from the National Health Service North
-West Centre for Research Ethics Committee (reference: 11/NW/0382).
-We limit the sample to white British participants, as defined by
-genetic estimated ancestry and self-identified ethnic group, giving a sample
-size of `r comma_num(nrow(famhist))`. For regressions on number of children we use participants over 45, since most fertility is completed by this age. This
-gives a sample size of `r comma_num(nrow(famhist %>% filter(kids_ss)))`.
-
-We computed 33 polygenic scores. The effect size estimates were obtained from
-GWASs that were chosen to not have included the UK Biobank dataset, to avoid
-overestimation of the genetic predisposition of a trait. The polygenic scores
-were computed using the summary-data based best linear unbiased prediction
-(SBLUP) approach. As a reference sample for the linkage disequilibrium, we used
-a random sample of 10,000 unrelated individuals from UK Biobank, imputed using
-the Haplotype Reference Consortium panel. Scores were normalized to mean 0,
-variance 1, and residualized on the first 100 principal components of array
-data, which were calculated as in @abdellaoui2019genetic.
+We use participant data from UK Biobank [@bycroft2018uk], which has received
+ethical approval from the National Health Service North West Centre for Research
+Ethics Committee (reference: 11/NW/0382). We limit the sample to white British
+participants of European descent, as defined by genetic estimated ancestry and
+self-identified ethnic group [@abdellaoui2019genetic], giving a sample size of
+`r comma_num(nrow(famhist))`. For regressions on number of children we use
+participants over 45, since most fertility is completed by this age. This gives
+a sample size of `r comma_num(nrow(famhist %>% filter(kids_ss)))`.
+
+Polygenic scores were computed by summing the alleles across ~1.3 million 
+genetic variants weighted by their effect sizes as estimated in 33 genome-wide 
+association studies (GWASs) that excluded UK Biobank. To control for population 
+stratification, we corrected the polygenic scores for 100 principal components (PCs). 
+To compute polygenic scores and PCs, the same procedures were followed as 
+described in @abdellaoui2019genetic.
 
@@
 
+# Acknowledgements
+
+Abdel Abdellaoui is supported by the Foundation Volksbond Rotterdam and by ZonMw grant 849200011 from The Netherlands Organisation for Health Research and Development.
+This study was conducted using UK Biobank resources under application number 40310.
+
 \FloatBarrier
-\newpage
+\clearpage
 # Appendix
@@
 \FloatBarrier
-## Controlling for AFLB
-
-```{r plot-age-flb, fig.cap = "Selection effects controlling for age at first live birth (women only). Effect sizes for women only without controls are shown for comparison."}
-
-n_regs <- as.double(nrow(res_age_flb))
-
-res_combined %>% 
-      mutate(
-        score_name = pretty_names(score_name),
-        score_name = fct_reorder(score_name, estimate, order_abs(2)),
-        Regression = ifelse(reg.type == "raw", "Bivariate", "Controlled for AFLB")
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
-        color = p < 0.05/{{n_regs}})) + 
-        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
-        pval_geom_point() +
-        pval_scale() +
-        scale_fill_brewer(type = "qual", palette = 3) +
-        my_labs() +
-        my_theme()
-```
-
-\FloatBarrier
+\clearpage
 ## Weighted regressions
@@
 \FloatBarrier
+\clearpage
 ## Parents' generation
@@
 To check whether effect sizes were changing over time, we ran
-regressions interacting PGS with birth year, median split at 1950
+regressions interacting polygenic scores with birth year, median split at 1950
 ("early born" versus "late born"). We use both *number of children* and
@@
 \@ref(tab:tbl-children-by-period) summarize the results. There is no
-change in natural selection within the parents' generation. In the
-respondents' generation, effect sizes were significantly larger in absolute size among the late-born for four PGS: cognitive ability, EA2, EA3 and
+evidence for change in natural selection within the parents' generation. In the
+respondents' generation, effect sizes were significantly larger in absolute size 
+among the late-born for four polygenic scores: cognitive ability, EA2, EA3 and
 extraversion. These changes are inconsistent with the intergenerational change, 
 where effect sizes were if anything larger among the earlier, parents' generation.
-Three PGS showed significant changes in sign: alcohol use
+Three scores showed significant changes in sign: alcohol use
 (positive to negative); conscientiousness (negative to positive); and
@@
 
+\clearpage
 \FloatBarrier
@@
 
-Figure \@ref(fig:plot-regs-controlled-pcs) compares
-selection effects on polygenic scores residualized for the top 100
-principal components of the genetic data, to selection effects on raw, unresidualized polygenic
-scores. In siblings regressions, effect sizes are larger for raw scores
-principal components -- sometimes much larger, as in the case of
-height. `r sum(controlled_smaller_sibs)` out of `r n_pgs` "raw"
-effect sizes have a larger absolute value than the corresponding "residualized"
-effect size. The median proportion between raw and controlled effect
-sizes is `r median_prop_sibs`. Among the children regressions, this no
-longer holds. Effect sizes are barely affected by controlling for
-principal components.
+Figure \@ref(fig:plot-regs-controlled-pcs) compares selection effects on
+polygenic scores residualized for the top 100 principal components of the
+genetic data, to selection effects on raw, unresidualized polygenic scores. In
+siblings regressions, effect sizes are larger for raw scores  -- sometimes much
+larger, as in the case of height. `r sum(controlled_smaller_sibs)` out of 
+`r n_pgs` "raw" effect sizes have a larger absolute value than the corresponding
+"residualized" effect size. The median proportion between raw and controlled
+effect sizes is `r median_prop_sibs`. Among the children regressions, this no
+longer holds. Effect sizes are barely affected by controlling for principal
+components.
 
@@
 
-```{r plot-pc-regs, fig.cap = "Selection effects of principal components of genetic data.
+```{r plot-pc-regs, fig.cap = "Selection effects of 100 principal components of genetic data.
 Absolute effect sizes are plotted.
@@
       ggplot(aes(estimate, dep.var, colour = p < 0.05/{{n_regs}})) + 
-        geom_jitter(height = 0.15, width = 0, alpha = 0.9) + 
+        geom_jitter(height = 0.15, width = 0, alpha = 0.8) + 
         scale_x_log10(labels = scales::label_number(accuracy = 0.0001)) +
@@
 
+\clearpage
+\FloatBarrier
 ## Selection controlling for age at first live birth: respondents' parents
@@
 are negatively correlated with the effect sizes from bivariate
-regressions without the age at birth control, though the correlation
-is smaller (father's age at birth: $\rho$ `r cor_unc_fath`; mother's age at
-birth: $\rho$ `r cor_unc_moth`).
+regressions without the age at birth control (father's age at birth: 
+$\rho$ `r cor_unc_fath`; mother's age at birth: $\rho$ `r cor_unc_moth`).
 
@@
 
-## Effects of PGS on age at first live birth
+\clearpage
+\FloatBarrier
+## Effects of polygenic scores on age at first live birth
 
@@
 
-Our results suggest that polygenic scores may directly correlate with
-age at first live birth. Figure \@ref(fig:plot-age-flb-dv) plots
-estimated effect sizes from bivariate regressions for respondents, and
-Figure \@ref(fig:plot-age-birth-parents-dv) does the same for their
-parents. Effect sizes are reasonably large. They are also very highly
-correlated across generations. Effect sizes of PGS on father's age at
-own birth, and on own age at first live birth, have a correlation of
-`r cor_flb_fath`; for mother's age and own age it is `r cor_flb_moth`.
+Our results suggest that polygenic scores may directly correlate with age at
+first live birth. Figure \@ref(fig:plot-age-flb-dv) plots estimated effect sizes
+from bivariate regressions for respondents, and Figure
+\@ref(fig:plot-age-birth-parents-dv) does the same for their parents. Effect
+sizes are reasonably large. They are also very highly correlated across
+generations. Effect sizes of polygenic scores on father's age at own birth, and
+on own age at first live birth, have a correlation of `r cor_flb_fath`; for
+mother's age and own age it is `r cor_flb_moth`.
 
@@
 \FloatBarrier
+\clearpage
 ## Controlling for earnings and education
@@
 \FloatBarrier
+\clearpage
 ## Genetic correlations with EA3
@@
 
+\clearpage
 \FloatBarrier
-
 ## Effects on inequality
@@
       set_caption("Differences in polygenic scores between highest and lowest income group.") %>% 
-      theme_compact()
-```
-
+      theme_compact() %>% 
+      set_bottom_border(final(1), everywhere)
+```
+
+\clearpage
 \FloatBarrier
-\newpage
 # References

### 2020-12-16 why-negative-selection.Rmd
More style and layout tweaks.

@@
 ---
-title: "Labor Market Income and Substitution Effects Explain Modern Natural Selection"
+title: "Labor Market Income and Substitution Effects Explain Contemporary Natural Selection"
 author: "David Hugh-Jones \\thanks{Corresponding author.
@@
 Effects are larger among those not living with a spouse or partner
-(Figure \@ref(fig:plot-n-partners)b). Lastly, we split female
-respondents by age at first live birth (AFLB).[^1] There is evidence for
-genetic effects on AFLB [@Barban_2016], and there is a close link
-between this variable and number of children born. Figure
-\@ref(fig:plot-n-partners)c shows effect sizes estimated separately for
-each tercile of AFLB. Several effects are strikingly different across
-terciles. ADHD and MDD are selected for amongst the youngest third of
+(Figure \@ref(fig:plot-n-partners)b).
+
+Lastly, we split female respondents by age at first live birth
+(AFLB).[^aflb-no-men] There is evidence for genetic effects on AFLB
+[@Barban_2016], and there is a close link between this variable and number of
+children born. Figure \@ref(fig:plot-age-flb)a shows effect sizes estimated
+separately for each tercile of AFLB. Several effects are strikingly different
+across terciles. ADHD and MDD are selected for amongst the youngest third of
 mothers, but selected against among the oldest two-thirds. Educational
-attainment is selected for among the oldest two-thirds of mothers, but
-is not significantly selected among the youngest third. Similarly,
-several PGS of body measurements are selected against only among older
-mothers. The correlation between effect sizes for the youngest and
-oldest terciles is `r cor_age_flb`. As before, recall that polygenic scores may 
-have two effects: changing the number of children within each category, and 
-changing the number of children by shifting respondents between categories.
-
-[^1]: This information is unavailable for men.
-
-\newgeometry{left=1cm,right=1cm,top=1cm,bottom=1cm}
-
-```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth", fig.subcap = c("Lifetime number of sexual partners", "Household type", "Age at first live birth, females"), fig.ncol = 1, out.height = c("32%", "32%", "25%"), fig.align = "center"}
+attainment is selected for among the oldest two-thirds of mothers, but is not
+significantly selected among the youngest third. Similarly, several PGS of body
+measurements are selected against only among older mothers. The correlation
+between effect sizes for the youngest and oldest terciles is `r cor_age_flb`. As
+before, recall that polygenic scores may have two effects: changing the number
+of children within each category, and changing the number of children by
+shifting respondents between categories.
+
+[^aflb-no-men]: This information is unavailable for men.
+
+```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth", fig.subcap = c("Lifetime number of sexual partners", "Household type", "Age at first live birth, females"), fig.ncol = 1, fig.align = "center"}
 
@@
                   legend.box.just      = "left",
-                  legend.justification = "left",
-                  legend.text          = element_text(size = 7),
-                  legend.title         = element_text(size = 8)
+                  legend.justification = "left"
                 )
@@
 
+```
+
+```{r calc-corr-age-flb}
+
+drake::loadd(res_age_flb)
+drake::loadd(res_sex)
+
+res_sex_comparison <- res_sex %>% 
+      filter(sex == "Female") %>% 
+      mutate(score_name = term) %>% 
+      select(score_name, term, estimate:conf.high)
+
+res_age_flb %<>% filter(term != "age_flb")
+
+res_combined <- bind_rows(
+        raw      = res_sex_comparison, 
+        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
+        .id      = "reg.type"
+      ) %>% 
+      arrange(score_name)
+
+raw_flb_comparison <- res_combined %>% 
+      pivot_wider(
+        id_cols     = score_name, 
+        names_from  = reg.type, 
+        values_from = estimate
+      )
+
+cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
+opp_signed <- sum(
+        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
+      )
+
+```
+
+To investigate this further, we regress number of children on PGS
+*controlling* for AFLB (Figure \@ref(fig:plot-age-flb)b). In
+`r opp_signed` out of `r length(score_names)` cases, effects
+change sign when controls are added. The correlation between effect
+sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`.
+Thus, selection effects seem to come through two opposing channels: an
+effect on AFLB, and an opposite-signed effect on number of children 
+controlling for AFLB.
+
+```{r plot-age-flb, fig.cap = "Selection effects by age at first live birth, and controlling for age at first live birth (women only).", fig.ncol = 1,  fig.subcap = c("Effect sizes by age at first live birth terciles", "Effect sizes controlling for age at first live birth.
+Effect sizes without controls (for women only) shown for comparison."), fig.align = "center"}
+
 drake::loadd(res_age_flb_cross)
@@
         my_theme()
-```
-
-\restoregeometry
-
-```{r calc-corr-age-flb}
-
-drake::loadd(res_age_flb)
-drake::loadd(res_sex)
-
-res_sex_comparison <- res_sex %>% 
-      filter(sex == "Female") %>% 
-      mutate(score_name = term) %>% 
-      select(score_name, term, estimate:conf.high)
-
-res_age_flb %<>% filter(term != "age_flb")
-
-res_combined <- bind_rows(
-        raw      = res_sex_comparison, 
-        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
-        .id      = "reg.type"
-      ) %>% 
-      arrange(score_name)
-
-raw_flb_comparison <- res_combined %>% 
-      pivot_wider(
-        id_cols     = score_name, 
-        names_from  = reg.type, 
-        values_from = estimate
-      )
-
-cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
-opp_signed <- sum(
-        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
-      )
-
-```
-
-To investigate this further, we regress number of children on PGS
-*controlling* for AFLB (Figure \@ref(fig:plot-age-flb)). In
-`r opp_signed` out of `r length(score_names)` cases, effects
-change sign when controls are added. The correlation between effect
-sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`.
-Thus, selection effects seem to come through two opposing channels: an
-effect on AFLB, and an opposite-signed effect on number of children 
-controlling for AFLB.
-
-```{r plot-age-flb, fig.cap = "Selection effects controlling for age at first live birth (women only). Effect sizes without controls (for women only) are shown for comparison."}
 
@@
         color = p < 0.05/{{n_regs}})) + 
-        geom_vline(xintercept = 0, linetype = "dashed", color = "grey20") +
+        my_vline +
         pval_geom_point() +
         pval_scale() +
-        scale_fill_brewer(type = "qual", palette = 3) +
+        my_fill_scale(n = 2, direction = -1) +
         my_labs() +
@@
 
+\FloatBarrier
 ## Correcting for sampling bias

### 2020-12-16 why-negative-selection.Rmd
More tweaks.

@@
       )
+options(huxtable.long_minus = TRUE)
 
@@
                  diff_children = wtd.y - wtd.x,
-                 diff_hypoth   = unw.y - unw.x
+                 diff_hypoth   = unw.y - unw.x,
+                 pct_change    = 100 * (diff_children - diff_hypoth)/diff_hypoth
                )
@@
         PGS                 = score, 
+        `Without selection` = diff_hypoth,
         `With selection`    = diff_children, 
-        `Without selection` = diff_hypoth
+        `% change`          = pct_change
       ) %>% 
@@
       ) %>% 
-      arrange(desc(`With selection`)) %>% 
+      arrange(desc(`% change`)) %>% 
       as_hux() %>% 
       set_align(everywhere, -1, "right") %>% 
-      set_number_format(everywhere, -1, 3) %>% 
+      set_number_format(everywhere, 2:3, 3) %>% 
+      set_number_format(everywhere, 4, 1) %>% 
       set_caption("Differences in polygenic scores between highest and lowest income group.") %>% 

### 2020-12-16 why-negative-selection.Rmd
Tweaks

@@
 ```{r plot-cors-earnings-educ, fig.cap="Selection effects, with and without controls for age at first live birth, by correlations with earnings and educational attainment.
-Each point represents one PGS."}
+Each point represents one PGS.", fig.height = 4}
 

### 2020-12-17 why-negative-selection.Rmd
Improve intro and conclusion. Mention Fermi paradox.

@@
 Living organisms evolve through natural selection, in which allele frequencies
-change in the population through differential reproduction rates. Geneticists
+change in the population through differential reproduction rates. Studying the
+mechanisms behind natural selection can help us better understand how
+individual differences in complex traits and disease risk arise. Geneticists
 have long hypothesized that natural selection is taking place in modern human
-populations. Recent work confirms this using genome-wide analysis [@Barban_2016;
-@beauchamp2016genetic;
-@kong2017selection; @sanjak2018evidence]. In particular, genetic variants
-associated with higher educational attainment are being selected against,
-although effect sizes appear small.
-
-As yet we know little about the social mechanisms behind these effects. This study uses
-data from UK Biobank [@bycroft2018uk] to learn more. We test for natural selection on 33
-different polygenic scores by estimating their correlation with fertility. We
-extend the analysis over two generations, using data on respondents' number of
-siblings as well as their number of children. This is interesting because
-consistent natural selection over multiple generations could lead to substantive
-effects in the long run. Most importantly, we examine reproductive rates in
-different subgroups of the population, in order to uncover patterns that can
-help illuminate the mechanisms behind modern natural selection.
+populations [@galton1869hereditary;@darwin1871the]; recent work confirms this 
+using genome-wide analysis
+[@Barban_2016;@beauchamp2016genetic;@kong2017selection;@sanjak2018evidence]. In
+particular, genetic variants associated with higher educational attainment are
+being selected against, although effect sizes appear small.
+
+As yet we know little about the social mechanisms behind these effects. This
+study uses data from UK Biobank [@bycroft2018uk] to learn more. We test for
+natural selection on 33 different polygenic scores by estimating their
+correlation with fertility. We extend the analysis over two generations, using
+data on respondents' number of siblings as well as their number of children.
+This is interesting because consistent natural selection over multiple
+generations could lead to substantive effects in the long run. Most importantly,
+we examine reproductive rates in different subgroups of the population, in order
+to uncover patterns that can help illuminate the mechanisms behind modern
+natural selection.
 
@@
 educational attainment, which lowers the mean, is stronger at the bottom of the 
-income distribution, and this increases the differences between groups.
+income distribution, and this increases the differences between groups.  Overall,
+natural selection increases inequality for 29 out of 33 polygenic scores 
+(Appendix Table \@ref(tab:tbl-inequality-effects)).
 
@@
 
+Existing evidence on human natural selection has led some to "biocosmic
+pessimism", including speculation that the Fermi paradox may be resolved by
+dysgenic selection among alien species [@sarraf2019modernity]. Others are more
+sanguine, and argue that natural selection's effects are outweighed by
+environmental improvements, such as those underlying (e.g.) the Flynn effect
+[@flynn1987massive]. The evidence here may add some nuance to this debate.
+Patterns of natural selection have been relatively consistent across the past two
+generations, but they are not the outcome of a universal, society-wide phenomenon.
+Instead they result from a sum of opposing forces, operating in different
+parts of society and pulling in different directions.
+
 Any model of fertility is implicitly a model of natural selection, but so far,
@@
 Integrating the two could deepen our understanding of natural selection in
-modern societies.
-Economics possesses a range of theoretical models on the effects of skills,
-education and income [see @hotz1997economics; @lundberg2007american]. One
-perennial problem is how to test these theories in a world where education,
-labour and marriage markets all interact. Genetic data, such as polygenic
-scores, could help to pin down directions of causality. Conversely, theory 
-and empirical results from economics can shine a light on the mechanisms 
-behind natural selection.
+modern societies. Economics possesses a range of theoretical models on the
+effects of skills, education and income [see @hotz1997economics;
+@lundberg2007american]. One perennial problem is how to test these theories in a
+world where education, labour and marriage markets all interact. Genetic data,
+such as polygenic scores, could help to pin down the direction of causality.
+Conversely, theory and empirical results from economics can shine a light on the
+mechanisms behind natural selection, and thereby on the nature of individual
+differences in complex traits and disease risk.
 
@@
 polygenic scores between the highest and lowest income groups. Column "Actual"
-shows respondents' scores weighted by Age/Qualification times number of
-children. Column "Without selection" shows scores weighted by Age/Qualification
+uses respondents' scores weighted by Age/Qualification times number of
+children. Column "Without selection" uses scores weighted by Age/Qualification
 only, i.e. if all couples had the same number of children.

### 2020-12-18 why-negative-selection.Rmd
move inequality table to _drake

@@
 
-drake::loadd(age_qual_weights)
-
-ineq <- famhist %>% 
-          left_join(age_qual_weights, by = "f.eid") %>% 
-          filter(! is.na(n_children), ! is.na(income_cat), ! is.na(weights)) %>% 
-          group_by(income_cat) %>% 
-          summarize(across(all_of(score_names), 
-                    list(
-                      unw = ~ weighted.mean(.x, weights, na.rm = TRUE),
-                      wtd = ~ weighted.mean(.x, weights * n_children, 
-                                              na.rm = TRUE)
-                    )
-          )) %>% 
-          tidyr::pivot_longer(-income_cat, 
-                                names_to = c("score", ".value"), 
-                                names_pattern = "(.*)_(.+?)$"
-                              )
+drake::loadd(res_ineq)
 
 ineq_comp <-  left_join(
-                ineq %>% filter(income_cat == 1),
-                ineq %>% filter(income_cat == 5),
+                res_ineq %>% filter(income_cat == 1),
+                res_ineq %>% filter(income_cat == 5),
                 by = "score"

### 2020-12-18 _drake.R

@@ -533,6 +533,24 @@ plan <- drake_plan(
               .id = "score_name")
   },
   
+  res_ineq = {
+    famhist %>% 
+          left_join(age_qual_weights, by = "f.eid") %>% 
+          filter(! is.na(n_children), ! is.na(income_cat), ! is.na(weights)) %>% 
+          group_by(income_cat) %>% 
+          summarize(across(all_of(score_names), 
+                    list(
+                      unw = ~ weighted.mean(.x, weights, na.rm = TRUE),
+                      wtd = ~ weighted.mean(.x, weights * n_children, 
+                                              na.rm = TRUE)
+                    )
+          )) %>% 
+          tidyr::pivot_longer(-income_cat, 
+                                names_to = c("score", ".value"), 
+                                names_pattern = "(.*)_(.+?)$"
+                              ) 
+  },
+  
   # res_margins = {
   #    res_extensive <- map_dfr(score_names, 
   #                       ~ run_regs_fml(

### 2020-12-18 why-negative-selection.Rmd
Use polygenic scores not PGS

@@
 ```{r plot-income-educ-level, fig.cap = "Selection effects by education and income.
-Each point represents a single bivariate regression of number of children on a PGS", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}
+Each point represents a single bivariate regression of number of children on a polygenic score", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}
 
@@
 
-Figure \@ref(fig:plot-income-educ-level) plots selection effects for
-each polygenic score, grouping respondents by age of completing
-full-time education, and by household income. Effects are larger and
-more significant for the lowest income category, and for the lowest
-education category. Note that the overall effect is not a simple average
-of the effect among the different subgroups, because polygenic scores
-may also shift respondents between the subgroups. For example, a high
-PGS for educational attainment may predict having fewer children among
-early school leavers, but it may also increase the age at which a
-respondent leaves school.
+Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
+polygenic score, grouping respondents by age of completing full-time education,
+and by household income. Effects are larger and more significant for the lowest
+income category, and for the lowest education category. Note that the overall
+effect is not a simple average of the effect among the different subgroups,
+because polygenic scores may also shift respondents between the subgroups. For
+example, a high polygenic score for educational attainment may predict having
+fewer children among early school leavers, but it may also increase the age at
+which a respondent leaves school.
 
@@
 
-Selection effects are also different between men and women (Appendix
-Figure \@ref(fig:plot-sexes)). Differences are particularly large for
-educational attainment, height and MDD. Several PGS for mental illness
-and personality traits are more selected for (or less against) among
-women, including major depressive disorder (MDD), schizophrenia and
-neuroticism, while extraversion is more selected for among men. PGS for waist
-circumference and waist-hip ratio are less selected for among women, and PGS for
-educational attainment are more selected against, though the difference
-is only significant for the educational attainment PGS of @lee2018gene
-("EA3"). One possible reason for these sex differences is
-that polygenic scores may affect fertility via success in marriage markets, and
-men and women may value different characteristics in these.
+Selection effects are also different between men and women (Appendix Figure
+\@ref(fig:plot-sexes)). Differences are particularly large for educational
+attainment, height and MDD. Several polygenic scores for mental illness and
+personality traits are more selected for (or less against) among women,
+including major depressive disorder (MDD), schizophrenia and neuroticism, while
+extraversion is more selected for among men. Scores for waist circumference and
+waist-hip ratio are less selected for among women, and scores for educational
+attainment are more selected against, though the difference is only significant
+for the "EA3" educational attainment PGS of @lee2018gene . One possible reason
+for these sex differences is that polygenic scores may affect fertility via
+success in marriage markets, and men and women may value different
+characteristics in these.
 
@@
 attainment is selected for among the oldest two-thirds of mothers, but is not
-significantly selected among the youngest third. Similarly, several PGS of body
-measurements are selected against only among older mothers. The correlation
-between effect sizes for the youngest and oldest terciles is `r cor_age_flb`. As
-before, recall that polygenic scores may have two effects: changing the number
-of children within each category, and changing the number of children by
-shifting respondents between categories.
+significantly selected among the youngest third. Similarly, several polygenic
+scores for body measurements are selected against only among older mothers. The
+correlation between effect sizes for the youngest and oldest terciles is `r
+cor_age_flb`. As before, recall that polygenic scores may have two effects:
+changing the number of children within each category, and changing the number of
+children by shifting respondents between categories.
 
@@
 
-To investigate this further, we regress number of children on PGS
-*controlling* for AFLB (Figure \@ref(fig:plot-age-flb)b). In
-`r opp_signed` out of `r length(score_names)` cases, effects
-change sign when controls are added. The correlation between effect
-sizes controlling for AFLB, and raw effect sizes, is `r cor_raw_flb`.
-Thus, selection effects seem to come through two opposing channels: an
-effect on AFLB, and an opposite-signed effect on number of children 
+To investigate this further, we regress *number of children* on polygenic scores
+*controlling* for AFLB (Figure \@ref(fig:plot-age-flb)b). In `r opp_signed` out
+of `r length(score_names)` cases, effects change sign when controls are added.
+The correlation between effect sizes controlling for AFLB, and raw effect sizes,
+is `r cor_raw_flb`. Thus, selection effects seem to come through two opposing
+channels: an effect on AFLB, and an opposite-signed effect on number of children
 controlling for AFLB.
@@
 ```{r plot-cors-earnings-educ, fig.cap="Selection effects, with and without controls for age at first live birth, by correlations with earnings and educational attainment.
-Each point represents one PGS.", fig.height = 4}
+Each point represents one polygenic score.", fig.height = 4}
 
@@
 on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
-on each PGS against that PGS's correlation with earnings in a respondent's first
+on each polygenic score against that score's correlation with earnings in a respondent's first
 job, and against its correlation with educational attainment, a
@@
 
-Second, we re-estimate equation \@ref(eq:regression) for each
-polygenic score, controlling for earnings in first job and education levels. Effect sizes
-are generally reduced (Appendix Figure \@ref(fig:plot-res-earn-educ)), and only
-`r n_05_33` out of 33 PGS are significant at $p = 0.05/33$. This suggests that
-earnings and education indeed mediate the effect of polygenic scores on fertility.
+Second, we re-estimate equation \@ref(eq:regression) for each polygenic score,
+controlling for earnings in first job and education levels. Effect sizes are
+generally reduced (Appendix Figure \@ref(fig:plot-res-earn-educ)), and only 
+`r n_05_33` out of 33 scores are significant at $p = 0.05/33$. This suggests that
+earnings and education indeed mediate the effect of polygenic scores on
+fertility.
 
@@
 highly significant predictor of number of children, it has little impact on the
-effect size or significance of the PGS. The median ratio of PGS effect sizes
-between regressions with and without controls is `r median(ratio)`; all PGS
+effect size or significance of the polygenic scores. The median ratio of effect sizes
+between regressions with and without controls is `r median(ratio)`; all scores
 which are significant at $p \le 0.05/33$ in uncontrolled regressions remain so

### 2020-12-18 why-natural-selection.Rmd
Add purifying selection (quadratic) regressions. Rename file.

--- 2020-12-18 why-negative-selection.Rmd
@@
 sample seems to be ascertained in ways that shrink estimates of selection
-effects, we suspect that our weighted estimates are still conservative.
+effects, we suspect that our weighted estimates are still conservative.[^purifying]
+
+[^purifying]: We also check for purifying selection by estimating (\@ref(eq:regression)) with a quadratic term in $PGS_i$. See Appendix figure \@ref(fig:plot-purifying).
 
@@
 \clearpage
+## Purifying and diversifying selection
+
+We rerun equation (\@ref(eq:regression)), adding a quadratic term in $PGS_i$,
+and using our age/qualification weights. Scores for hip circumference and drinks
+per week show significant purifying selection ($p < 0.05/33$, negative
+coefficient). Scores for educational attainment (EA2 and EA3) show significant
+diversifying selection ($p < 0.05/33$, positive coefficient), which reduces the
+strength of selection against educational attainment at very high levels of the 
+PGS. Figure
+\@ref(fig:plot-purifying) plots predicted number of children against polygenic
+score from these regressions.
+
+```{r plot-purifying, fig.align = "center", fig.cap = "Purifying/diversifying selection: predicted number of children by polygenic score", fig.height = 5}
+
+drake::loadd(res_quadratic)
+sig <- res_quadratic %>% 
+         filter(grepl("^2", term, fixed = TRUE)) %>% 
+         filter(p.value < 0.05/33) %>% 
+         pull(score_name)
+
+stopifnot(sig == c("dpw_substance_use", "EA2_noUKB", "EA3_excl_23andMe_UK",
+                     "hip_combined"))
+
+sig_all_terms <- res_quadratic %>% filter(score_name %in% sig)
+
+op <- par(mfrow = c(2,2))
+for (score in sig) {
+  coef <- sig_all_terms %>% filter(score_name == score) %>% pull(estimate)
+  curve(
+         coef[1] + coef[2] * x + coef[3] * x^2,
+         xlim = c(-4, 4), 
+         xlab = pretty_names(score),
+         ylab = "Predicted N children"
+       )
+}
+par(op)
+
+```
+
+\FloatBarrier
+\clearpage
 ## Parents' generation

### 2020-12-18 _drake.R

@@ -157,6 +157,19 @@ plan <- drake_plan(
           )
   },
   
+  res_quadratic = {
+    famhist %<>% inner_join(age_qual_weights, by = "f.eid")
+    map_dfr(score_names, 
+              ~run_regs_fml(
+                             "n_children ~ {score_name} + I({score_name}^2)", 
+                             score_name = .x,
+                             famhist    = famhist,
+                             weights    = quote(weights)
+                           ),
+              .id = "score_name"
+            )
+  },
+  
   res_pcs = {
     pc_names <- grep("PC", names(pcs), value = TRUE)
     res_sibs_pcs <- run_regs_pcs( 
@@ -623,8 +636,8 @@ plan <- drake_plan(
   
   report =  {
               rmarkdown::render(
-                input       = knitr_in("why-negative-selection.Rmd"), 
-                output_file = file_out("why-negative-selection.pdf"), 
+                input       = knitr_in("why-natural-selection.Rmd"), 
+                output_file = file_out("why-natural-selection.pdf"), 
                 quiet       = TRUE
               )
             }

### 2020-12-18 regressions.R

@@ -148,10 +148,10 @@ run_regs_subset <- function(score_name, subset, famhist) {
 }
 
 
-run_regs_fml <- function(fml, ..., subset = NULL, famhist) {
+run_regs_fml <- function(fml, ..., subset = NULL, famhist, weights = NULL) {
   glue_args <- list(...)
   fml <- as.formula(glue::glue_data(fml, .x = glue_args))
-  mod <- lm(fml, famhist, subset = eval(subset))
+  mod <- lm(fml, famhist, subset = eval(subset), weights = eval(weights))
   res <- tidy(mod, conf.int = TRUE)
   attr(res, "call") <- mod$call
   res

### 2020-12-18 why-natural-selection.Rmd
Tweaks

@@
 
-[^purifying]: We also check for purifying selection by estimating (\@ref(eq:regression)) with a quadratic term in $PGS_i$. See Appendix figure \@ref(fig:plot-purifying).
+[^purifying]: We also check for purifying selection by estimating \@ref(eq:regression) with a quadratic term. See Appendix Figure \@ref(fig:plot-purifying).
 
@@
 
-We rerun equation (\@ref(eq:regression)), adding a quadratic term in $PGS_i$,
+We rerun equation \@ref(eq:regression), adding a quadratic term in $PGS_i$,
 and using our age/qualification weights. Scores for hip circumference and drinks

### 2020-12-18 why-natural-selection.Rmd
Compare effect sizes with and without childless. Be more tentative about any conclusions from comparing generations.

@@
 Selection effects are highly correlated across the two generations, and most
-share the same sign (Appendix Figure \@ref(fig:plot-res-sibs)). Effect size
-estimates are larger for the parents' generation. We treat this result cautiously,
-because when we split respondents up by year of birth, we find few differences
-in effect sizes between early- and late-born respondents, for either generation.
-In other words, since estimated effect sizes change when we change the dependent
-variable, but do not change over time within either dependent variable, this
-may be due to remaining ascertainment bias within the sample.
+share the same sign (Appendix Figure \@ref(fig:plot-res-sibs)). Absolute effect
+size estimates are larger for the parents' generation. We treat this result
+cautiously, because when we split respondents up by year of birth, we find few
+differences in effect sizes between early- and late-born respondents, for either
+generation. In other words, since estimated effect sizes change across
+"generations", but do not change over time within either generation, the change
+may be due to remaining ascertainment bias within the sample or other effects.
+In particular, effect sizes in both generations may depend on polygenic scores'
+correlation with childlessness, and we cannot estimate this for the parents'
+generation.
 
@@
 
-Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings* on
-polygenic scores. By definition, people who had no children cannot be included in
-this data. For a cleaner comparison, we rerun regressions on the respondents'
-generation (*number of children*) after excluding respondents with no
-children.
+```{r calc-res-sibs}
+drake::loadd(res_wt_age_qual_weights)
+drake::loadd(res_children_comparison)
+
+cmp <- left_join(res_wt_age_qual_weights, res_children_comparison, by = "term")
+
+cmp %>%
+      mutate(
+        term = fct_reorder(term, estimate.x, order_abs(1)),
+        term = pretty_names(term)
+      ) %>%
+      ggplot(aes(y = term)) +
+        geom_pointrange(aes(x = estimate.x, xmin = conf.low.x,
+                            xmax = conf.high.x), color = "red", alpha = .8) +
+        geom_pointrange(aes(x = estimate.y, xmin = conf.low.y,
+                            xmax = conf.high.y), color = "blue", alpha = .8)
+
+cor_inc_vs_excl_childless <- cor(cmp$estimate.x, cmp$estimate.y)
+n_smaller <- sum(abs(cmp$estimate.y) < abs(cmp$estimate.x))
+
+pct_diff <- abs(cmp$estimate.y)/abs(cmp$estimate.x) - 1
+median_pct_diff <- median(pct_diff)
+
+```
+
+Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings*, i.e.
+parents' number of children, on polygenic scores. By definition, members of the
+parents' generation who had no children cannot be included in this data. For a
+clean comparison with the respondents' generation, we rerun regressions on
+*number of children* excluding those with no children, and show
+results in the figure.
+
+Excluding childless people in the parents' generation could bias our estimates.
+To learn about this, we compare effect sizes excluding and including childless
+people in the *current* generation. The correlation
+between the two sets of effect sizes is `r cor_inc_vs_excl_childless`. So, patterns across
+different scores are broadly similar whether the childless are counted or not. 
+Unfortunately, however, absolute effect sizes are smaller when the
+childless are excluded, for `r n_smaller` out of 33 scores; the median
+percentage change is `r percent(median_pct_diff)`. The fact that childless
+people have such a strong effect on estimates makes it hard to compare absolute
+effect sizes across generations. In particular, since the parents' generation has
+a different distribution of numbers of children, childless people may have had 
+more or less effect in that generation.
 
@@
 
-To check whether effect sizes were changing over time, we ran
+As an alternative approach, , we run
 regressions interacting polygenic scores with birth year, median split at 1950
@@
 \@ref(tab:tbl-children-by-period) summarize the results. There is no
-evidence for change in natural selection within the parents' generation. In the
+evidence for changes in selection effects within the parents' generation. In the
 respondents' generation, effect sizes were significantly larger in absolute size 
-among the late-born for four polygenic scores: cognitive ability, EA2, EA3 and
+among the later-born for four polygenic scores: cognitive ability, EA2, EA3 and
 extraversion. These changes are inconsistent with the intergenerational change, 
-where effect sizes were if anything larger among the earlier, parents' generation.
+where estimated effect sizes were larger among the earlier, parents' generation.
 Three scores showed significant changes in sign: alcohol use
 (positive to negative); conscientiousness (negative to positive); and
-type II diabetes (negative to positive). Overall, there is weak evidence for
-change over time, and the clearest result is that the direction of selection is
-broadly consistent.
+type II diabetes (negative to positive).
+
+Overall, there is weak evidence for change over time. The clearest results are
+that (a) the direction of selection, and (b) the pattern of relative effect sizes
+across scores, are broadly consistent over time.
 

### 2020-12-18 why-natural-selection.Rmd
tweaks

@@
 
-cmp %>%
-      mutate(
-        term = fct_reorder(term, estimate.x, order_abs(1)),
-        term = pretty_names(term)
-      ) %>%
-      ggplot(aes(y = term)) +
-        geom_pointrange(aes(x = estimate.x, xmin = conf.low.x,
-                            xmax = conf.high.x), color = "red", alpha = .8) +
-        geom_pointrange(aes(x = estimate.y, xmin = conf.low.y,
-                            xmax = conf.high.y), color = "blue", alpha = .8)
-
 cor_inc_vs_excl_childless <- cor(cmp$estimate.x, cmp$estimate.y)

### 2020-12-18 why-natural-selection.Rmd
Corrections about changes over time.

@@
 
+stopifnot(sum(summary_period$Change_Parents == "Insignificant") == 33)
+stopifnot(sum(summary_period$Change_Respondents == "Size increasing") == 4)
+stopifnot(sum(summary_period$Change_Respondents == "Change sign") == 1)
+stopifnot(sum(summary_period$Change_Respondents == "Insignificant") == 28)
 ```
@@
 where estimated effect sizes were larger among the earlier, parents' generation.
-Three scores showed significant changes in sign: alcohol use
-(positive to negative); conscientiousness (negative to positive); and
-type II diabetes (negative to positive).
+One score, conscientiousness, showed a significant change in sign, from 
+negative to positive effects on fertility.
 

### 2020-12-19 why-natural-selection.Rmd
Abdel tweaks

@@
 author: "David Hugh-Jones \\thanks{Corresponding author.
-School of Economics, University of East Anglia.
-Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Amsterdam UMC.
+School of Economics, University of East Anglia, Norwich, UK.
+Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands.
 Email: a.abdellaoui@amsterdamumc.nl}"
@@
 abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it.
-We test for natural selection on 33 polygenic scores over two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry).
+We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry).
 Consistently over time, polygenic scores which predict lower (higher) earnings and education are selected for (against).
-Selection effects are concentrated among low-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner.
+Selection effects are concentrated among lower-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner.
 Effect sizes are substantially larger after correcting for ascertainment bias.
 The direction of natural selection is reversed among older parents, or after controlling for age at first live birth.
-These patterns can be explained by economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market.
+These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market.
 We find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects and that are strongly related to the socio-economic system."
@@
   - \usepackage{titlesec}
-  - \titleformat*{\section}{\bfseries\sffamily\LARGE}
-  - \titleformat*{\subsection}{\bfseries\sffamily\itshape\Large}
+  - \titleformat*{\section}{\sffamily\LARGE}
+  - \titleformat*{\subsection}{\sffamily\itshape\Large}
   - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
@@
   chunk_output_type: console
-mainfont: Times
+mainfont: Baskerville
+sansfont: "Gill Sans"
 ---
@@
 change in the population through differential reproduction rates. Studying the
-mechanisms behind natural selection can help us better understand how
-individual differences in complex traits and disease risk arise. Geneticists
-have long hypothesized that natural selection is taking place in modern human
-populations [@galton1869hereditary;@darwin1871the]; recent work confirms this 
-using genome-wide analysis
+mechanisms behind natural selection can help us better understand how individual
+differences in complex traits and disease risk arise. Recent work confirms that
+natural selection is taking place in modern human populations, using genome-wide
+analysis
 [@Barban_2016;@beauchamp2016genetic;@kong2017selection;@sanjak2018evidence]. In
@@
 income and less education, among younger parents, people not living with a partner,
-and people with many lifetime sexual partners. Outside these groups, effects are
+and people with more lifetime sexual partners. Outside these groups, effects are
 weaker and often statistically insignificant. In some subgroups, the
@@
 
-Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores in the
-sample by 5-year birth intervals. Several scores show consistent increases or
-declines over this 30-year period, of the order of 5% of a standard
+We created polygenic scores for 33 traits in `r nrow(famhist)` individuals,
+corrected for ancestry using 100 genetic principal components (see Materials and
+Methods). Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores in
+the sample by 5-year birth intervals. Several scores show consistent increases
+or declines over this 30-year period, of the order of 5% of a standard
 deviation. These changes could reflect natural selection within the UK
@@
 Respondents have higher income and are better educated than the UK population,
-and they may also differ on other unobserved characteristics [@10.1093/aje/kwx246].
-Since richer and educated people also live longer, this bias could also
-increase with age.
+and they may also differ on other unobserved characteristics
+[@10.1093/aje/kwx246]. Since richer and educated people also live longer, this
+bias could also increase with age.
 
@@
 
-To test directly for natural selection, we regress respondents' polygenic scores
-(PGS) on their number of children ($y_i$):
+To test directly for natural selection, we regress respondents' number of
+children ($y_i$) on their polygenic scores (PGS):
 
@@
 
-These results could be driven by age, if older respondents have lower income
-and are less educated, and also more subject to selection on polygenic
+These results could be explained by age, if older respondents have lower income
+and are less educated, and also show more natural selection on polygenic
 scores. However, if we rerun the regressions, interacting the polygenic
@@
 the PGS with age of leaving full time education and a quadratic in age,
-the interaction with age leaving FTE remains significant at
+the interaction with age leaving full time education remains significant at
 0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.
@@
 attainment are more selected against, though the difference is only significant
-for the "EA3" educational attainment PGS of @lee2018gene . One possible reason
-for these sex differences is that polygenic scores may affect fertility via
-success in marriage markets, and men and women may value different
+for the ("EA3") educational attainment score of @lee2018gene. One
+possible reason for these sex differences is that polygenic scores may affect
+fertility via success in marriage markets, and men and women may value different
 characteristics in these.
@@
 scores for body measurements are selected against only among older mothers. The
-correlation between effect sizes for the youngest and oldest terciles is `r
-cor_age_flb`. As before, recall that polygenic scores may have two effects:
-changing the number of children within each category, and changing the number of
-children by shifting respondents between categories.
+correlation between effect sizes for the youngest and oldest terciles is 
+`r cor_age_flb`.
 
@@
 channels: an effect on AFLB, and an opposite-signed effect on number of children
-controlling for AFLB.
+controlling for AFLB. Note that we do not claim that AFLB is exogenous to an
+individual's polygenic scores -- indeed, our analyses show that it is not
+(Appendix Figure \@ref(fig:plot-age-flb-dv)). Rather, we argue that AFLB mediates
+part of the relationship between polygenic scores and fertility, and that
+once this is controlled for, the remaining part of the relationship has the
+opposite sign.
+
+AFLB mediates
+the relationship between polygenic scores and fertility, and that after this
+effect has been taken account of, there is a further, opposite-signed relationship.
 
@@
 
-Since UK Biobank subjects are not representative of the wider population, our
-results suggest that naïve estimates of natural selection are at risk of
-ascertainment bias. To correct for this, we weight participants using population data. We try three
-alternative weighting schemes: weighting by geography, age and
-presence/absence of a partner; weighting by age and highest educational
+Since UK Biobank participants are not representative of the wider population,
+our results suggest that naïve estimates of natural selection are at risk of
+ascertainment bias. To correct for this, we weight participants using population
+data. We try three alternative weighting schemes: weighting by geography, age
+and presence/absence of a partner; weighting by age and highest educational
 qualification; and age, highest qualification, and age at first live birth, for
-women only. Figure \@ref(fig:plot-res-wt) plots selection effects 
-among the entire sample, estimated with the three weighting schemes. 
+women only. Figure \@ref(fig:plot-res-wt) plots selection effects among the
+entire sample, estimated with the three weighting schemes.
 Mean effect sizes across all polygenic scores are
@@
 `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
-further affected by weighting on other demographic variables. Since the Biobank
+further affected by weighting on other demographic variables. Since the UK Biobank
 sample seems to be ascertained in ways that shrink estimates of selection
@@
   )) + 
-  labs(x = "Effect size", y = "")
+  labs(x = "Effect size", y = "") +
+  my_theme()
 
@@
 Although the direction of selection effects does not change between the
-generations, there are other differences. Compared to residualized polygenic
-scores, selection effects on unresidualized scores are about ten percent higher
-on average for *n siblings*, whereas effects for *number of children* barely
-change (Appendix Figure \@ref(fig:plot-regs-controlled-pcs)). This could be
-because earlier fertility is driven more by geographically clustered deprivation
-[e.g. via an insurance motive, @rendall1998old], which may correlate with the
-broad-scale genetic variation captured by principal components, such as ancestry.
+generations, there are other differences. Compared to our standard polygenic
+scores (residualized on 100 principal components of genetic data) 
+selection effects on unresidualized scores are about ten percent higher
+on average for *number of siblings*, whereas effects for *number of children*
+barely change (Appendix Figure \@ref(fig:plot-regs-controlled-pcs)). This could
+be because earlier fertility is driven more by geographically clustered
+deprivation [e.g. via an insurance motive, @rendall1998old], which may correlate
+with the broad-scale genetic variation captured by principal components, such as
+ancestry.
 
@@
 These results show that selection effects are weaker, absent, or even reversed
-among some subgroups of the population. This can potentially be explained by the
-economic theory of fertility [@becker1960economic; @willis1973new;
-@becker1976child]. According to this theory, increases in a person's wage affect their
-fertility via two opposing channels. There is an *income effect* by which
-children become more affordable, like any other good. There is also a
-*substitution effect*: since childrearing has a cost in time, the opportunity
-cost of childrearing increases if one's market wage is higher. The income effect
-would lead higher earners to have more children. The substitution effect would
-lead higher earners to have fewer children. If so, then genetic variants which
-affect earnings potential in the labour market may cause opposing effects on
-fertility. The income effect will cause natural selection in favour of
-earnings-increasing variants. The substitution effect will do the reverse.
+among some subgroups of the population. One possible explanation for this is
+given by the economic theory of fertility 
+[@becker1960economic; @willis1973new;@becker1976child]. According to this theory,
+increases in a person's wage affect their fertility via two opposing channels.
+There is an *income effect* by which children become more affordable, like any
+other good. There is also a *substitution effect*: since childrearing has a cost
+in time, the opportunity cost of childrearing increases if one's market wage is
+higher. The income effect would lead higher earners to have more children. The
+substitution effect would lead higher earners to have fewer children. If so,
+then genetic variants which affect earnings potential in the labour market may
+cause opposing effects on fertility. The income effect will cause natural
+selection in favour of earnings-increasing variants. The substitution effect
+will do the reverse.
 
@@
 share childcare responsibilities, while the income effect will dominate for
-couples who are able to reap the gains from specialization. Indeed, US fertility
+couples who are able to reap the gains from specialization [@becker1981treatise]. Indeed, US fertility
 decreases faster with education among single mothers than married mothers
@@
 for this: scores which might plausibly be linked to externalizing behaviour,
-like ADHD and age at smoking initiation, are selected for. However, this theory
-is less good at explaining variation in selection across the full range of
-scores, including physical measures, e.g. waist-hip ratio and BMI. Externalizing
-behaviour also does not explain why selection effects should work in the
-opposite direction among older parents, whereas the economic theory can explain
-this via income effects, as described above. We test the alternative theory
-directly by re-estimating equation \@ref(eq:regression) controlling for a
-measure of risk attitude (UK Biobank field 2040). While risk attitude is always a
-highly significant predictor of number of children, it has little impact on the
-effect size or significance of the polygenic scores. The median ratio of effect sizes
-between regressions with and without controls is `r median(ratio)`; all scores
-which are significant at $p \le 0.05/33$ in uncontrolled regressions remain so
-when controlling for risk attitude. Thus, although risk attitude does predict
-number of children, it appears not to mediate selection effects. Overall, we
-believe that the economic theory is the most likely explanation.
+like ADHD and (younger) age at smoking initiation, are selected for. However,
+this theory is less good at explaining variation in selection across the full
+range of scores, including physical measures, e.g. waist-hip ratio and BMI.
+Externalizing behaviour also does not explain why selection effects should work
+in the opposite direction among older parents, whereas the economic theory can
+explain this via income effects, as described above. We test the alternative
+theory directly by re-estimating equation \@ref(eq:regression) controlling for a
+measure of risk attitude (UK Biobank field 2040). While risk attitude is always
+a highly significant predictor of number of children, it has little impact on
+the effect size or significance of the polygenic scores. The median ratio of
+effect sizes between regressions with and without controls is `r median(ratio)`;
+all scores which are significant at $p \le 0.05/33$ in uncontrolled regressions
+remain so when controlling for risk attitude. Thus, although risk attitude does
+predict number of children, it appears not to mediate selection effects.
+Overall, we believe that the economic theory is the most likely explanation.
 
@@
 
-```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean PGS for educational attainment (EA3) of children by household income group.
+```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean polygenic score for educational attainment (EA3) of children by household income group.
 Blue is actual.
@@
 
-Abdel Abdellaoui is supported by the Foundation Volksbond Rotterdam and by ZonMw grant 849200011 from The Netherlands Organisation for Health Research and Development.
-This study was conducted using UK Biobank resources under application number 40310.
+AA is supported by the Foundation Volksbond Rotterdam and by ZonMw grant 849200011 from The Netherlands Organisation for Health Research and Development.
+This study was conducted using UK Biobank resources under application numbers 40310 and 19127.
 
@@
 different scores are broadly similar whether the childless are counted or not. 
-Unfortunately, however, absolute effect sizes are smaller when the
+However, absolute effect sizes are smaller when the
 childless are excluded, for `r n_smaller` out of 33 scores; the median
 percentage change is `r percent(median_pct_diff)`. The fact that childless
-people have such a strong effect on estimates makes it hard to compare absolute
+people have such a strong effect on estimates makes it hard to compare total
 effect sizes across generations. In particular, since the parents' generation has
@@
 
-As an alternative approach, , we run
-regressions interacting polygenic scores with birth year, median split at 1950
-("early born" versus "late born"). We use both *number of children* and
-*number of siblings* as a dependent variable. We weight using
-age/qualification cells, and further adjust for selection in the
-parents' generation (see above).
+As an alternative approach, we run regressions interacting polygenic scores with
+birth year, median split at 1950 ("early born" versus "late born"). We use both
+*number of children* and *number of siblings* as a dependent variable. We weight
+using age/qualification cells, and further adjust for selection in the parents'
+generation (see above).
 
@@
         my_vline + 
-        facet_wrap(vars(dep.var), scales = "free_y") +
+        facet_wrap(vars(dep.var), scales = "free") +
         pval_geom_point(position = position_dodge(0.4), alpha = 0.8) +
@@
       pval_scale() +
-      my_labs()
+      my_labs() +
+      my_theme()
 
@@
 first live birth. Figure \@ref(fig:plot-age-flb-dv) plots estimated effect sizes
-from bivariate regressions for respondents, and Figure
-\@ref(fig:plot-age-birth-parents-dv) does the same for their parents. Effect
-sizes are reasonably large. They are also very highly correlated across
-generations. Effect sizes of polygenic scores on father's age at own birth, and
-on own age at first live birth, have a correlation of `r cor_flb_fath`; for
-mother's age and own age it is `r cor_flb_moth`.
+from bivariate regressions for respondents. Figure
+\@ref(fig:plot-age-birth-parents-dv) does the same for their parents, using only
+eldest siblings.[^only-eldest] Effect sizes are reasonably large. They are also
+highly correlated across generations. Effect sizes of polygenic scores on
+father's age at own birth, and on own age at first live birth, have a
+correlation of `r cor_flb_fath`; for mother's age and own age it is 
+`r cor_flb_moth`.
+
+[^only-eldest]: Parental AFLB can only be calculated for this group.
 
@@
         my_fill_scale(n = 2) +
-        my_labs()
+        my_labs() +
+        my_theme()
 
@@
 
-Another way to examine the "earnings" theory of natural selection is to
-compare selection effects of PGS with their genetic correlation with
-educational attainment (EA3). Since EA3 strongly predicts earnings, if
-earnings drives differences in fertility, we'd expect a correlation
-between the two sets of results. Figure
-\@ref(fig:plot-rgs-by-effect-size) shows this is so.
+Another way to examine the "earnings" theory of natural selection is to compare
+selection effects of PGS with their genetic correlation with educational
+attainment (EA3). Since EA3 strongly predicts earnings, if earnings drives
+differences in fertility, we'd expect a correlation between the two sets of
+results. Figure \@ref(fig:plot-rgs-by-effect-size) shows this is so. Genetic
+correlations were calculated using LD score regression from GWAS summary
+statistics.
 
@@
         my_hline +
-        labs(x = "rG with EA3", y = "Effect on number of children")
+        labs(x = "Genetic correlation with EA3", y = "Effect on number of children")
 
@@
 Table \@ref(tab:tbl-inequality-effects) estimates differences in children's mean
-polygenic scores between the highest and lowest income groups. Column "Actual"
-uses respondents' scores weighted by Age/Qualification times number of
-children. Column "Without selection" uses scores weighted by Age/Qualification
-only, i.e. if all couples had the same number of children.
+polygenic scores between the highest and lowest income groups. Column "With
+selection" uses respondents' scores weighted by Age/Qualification times number
+of children. Column "Without selection" uses scores weighted by
+Age/Qualification only, i.e. if all couples had the same number of children.
 
@@
       set_caption("Differences in polygenic scores between highest and lowest income group.") %>% 
+      set_width(0.6) %>% 
       theme_compact() %>% 

### 2020-12-20 why-natural-selection.Rmd
bugfix

@@
 mainfont: Baskerville
+mathfont: Baskerville
 sansfont: "Gill Sans"
@@
 
-AFLB mediates
-the relationship between polygenic scores and fertility, and that after this
-effect has been taken account of, there is a further, opposite-signed relationship.
-
 ```{r plot-age-flb, fig.cap = "Selection effects by age at first live birth, and controlling for age at first live birth (women only).", fig.ncol = 1,  fig.subcap = c("Effect sizes by age at first live birth terciles", "Effect sizes controlling for age at first live birth.

### 2020-12-20 why-natural-selection.Rmd
Add, but don't use, inequality calculation for parents' generation

@@
 generations, there are other differences. Compared to our standard polygenic
-scores (residualized on 100 principal components of genetic data) 
+scores (residualized on 100 principal components of genetic data), 
 selection effects on unresidualized scores are about ten percent higher
@@
 \@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
-attainment (EA3), for children in households of different income groups. The
+attainment (EA3) among children from households of different income groups. The
 blue bars show the actual means, i.e. parents' mean polygenic score weighted by

### 2020-12-20 _drake.R

@@ -551,10 +551,16 @@ plan <- drake_plan(
           left_join(age_qual_weights, by = "f.eid") %>% 
           filter(! is.na(n_children), ! is.na(income_cat), ! is.na(weights)) %>% 
           group_by(income_cat) %>% 
+          mutate(
+            parent_weights = weights/n_sibs,
+            parent_weights = ifelse(is.na(parent_weights), 0, parent_weights)
+          ) %>% 
           summarize(across(all_of(score_names), 
                     list(
-                      unw = ~ weighted.mean(.x, weights, na.rm = TRUE),
-                      wtd = ~ weighted.mean(.x, weights * n_children, 
+                      unw     = ~ weighted.mean(.x, weights, na.rm = TRUE),
+                      wtd     = ~ weighted.mean(.x, weights * n_children, 
+                                              na.rm = TRUE),
+                      parents = ~ weighted.mean(.x, parent_weights, 
                                               na.rm = TRUE)
                     )
           )) %>% 

### 2020-12-21 why-natural-selection.Rmd
tweaks

@@
 
-Figure \@ref(fig:plot-siblings-townsend) plots effects on *n siblings* by
+Figure \@ref(fig:plot-siblings-townsend) plots effects on *number of siblings* by
 Townsend deprivation quintile of birth area.
@@
       set_caption("Differences in polygenic scores between highest and lowest income group.") %>% 
-      set_width(0.6) %>% 
       theme_compact() %>% 
+      set_width(0.9) %>% 
+      set_col_width(c(.35, .15, .15 , .15)) %>% 
       set_bottom_border(final(1), everywhere)

### 2020-12-21 why-natural-selection.Rmd
Last tweaks. (I hope.)

@@
 ---
-title: "Labor Market Income and Substitution Effects Explain Contemporary Natural Selection"
+title: "Natural selection in contemporary humans is linked to income and substitution effects"
 author: "David Hugh-Jones \\thanks{Corresponding author.
@@
 
-We created polygenic scores for 33 traits in `r nrow(famhist)` individuals,
-corrected for ancestry using 100 genetic principal components (see Materials and
-Methods). Figure \@ref(fig:plot-means-over-time) plots mean polygenic scores in
-the sample by 5-year birth intervals. Several scores show consistent increases
-or declines over this 30-year period, of the order of 5% of a standard
-deviation. These changes could reflect natural selection within the UK
-population, but also emigration, or ascertainment bias within the sample.
+We created polygenic scores for 33 traits in `r comma_num(nrow(famhist))`
+individuals, corrected for ancestry using 100 genetic principal components (see
+Materials and Methods). Figure \@ref(fig:plot-means-over-time) plots mean
+polygenic scores in the sample by 5-year birth intervals. Several scores show
+consistent increases or declines over this 30-year period, of the order of 5% of
+a standard deviation. These changes could reflect natural selection within the
+UK population, but also emigration, or ascertainment bias within the sample.
 Respondents have higher income and are better educated than the UK population,
@@
 ```{r plot-income-educ-level, fig.cap = "Selection effects by education and income.
-Each point represents a single bivariate regression of number of children on a polygenic score", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}
+Each point represents a single bivariate regression of number of children on a polygenic score.", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}
 
@@
 
-```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth", fig.subcap = c("Lifetime number of sexual partners", "Household type", "Age at first live birth, females"), fig.ncol = 1, fig.align = "center"}
+```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth.", fig.subcap = c("Lifetime number of sexual partners", "Household type", "Age at first live birth, females"), fig.ncol = 1, fig.align = "center"}
 
@@
 To investigate this further, we regress *number of children* on polygenic scores
-*controlling* for AFLB (Figure \@ref(fig:plot-age-flb)b). In `r opp_signed` out
-of `r length(score_names)` cases, effects change sign when controls are added.
-The correlation between effect sizes controlling for AFLB, and raw effect sizes,
-is `r cor_raw_flb`. Thus, selection effects seem to come through two opposing
-channels: an effect on AFLB, and an opposite-signed effect on number of children
-controlling for AFLB. Note that we do not claim that AFLB is exogenous to an
-individual's polygenic scores -- indeed, our analyses show that it is not
-(Appendix Figure \@ref(fig:plot-age-flb-dv)). Rather, we argue that AFLB mediates
-part of the relationship between polygenic scores and fertility, and that
-once this is controlled for, the remaining part of the relationship has the
-opposite sign.
+*controlling* for AFLB, again among females (Figure \@ref(fig:plot-age-flb)b).
+In `r opp_signed` out of `r length(score_names)` cases, effects change sign when
+controls are added. The correlation between effect sizes controlling for AFLB,
+and raw effect sizes, is `r cor_raw_flb`. Thus, selection effects seem to come
+through two opposing channels: an effect on AFLB, and an opposite-signed effect
+on number of children controlling for AFLB. Note that we do not claim that AFLB
+is exogenous to an individual's polygenic scores -- indeed, our analyses show
+that it is not (Appendix Figure \@ref(fig:plot-age-flb-dv)). Rather, we argue
+that AFLB mediates part of the relationship between polygenic scores and
+fertility, and that once this is controlled for, the remaining part of the
+relationship has the opposite sign.
 
@@
 Externalizing behaviour also does not explain why selection effects should work
-in the opposite direction among older parents, whereas the economic theory can
+in the opposite direction among older mothers, whereas the economic theory can
 explain this via income effects, as described above. We test the alternative
@@
 including people with lower income, lower education, younger first parenthood, 
-and more lifetime sexual partners. Indeed, among older parents, selection effects are
+and more lifetime sexual partners. Indeed, among older mothers, selection effects are
 actually reversed. Furthermore, the size of selection effects on a polygenic
@@
 world where education, labour and marriage markets all interact. Genetic data,
-such as polygenic scores, could help to pin down the direction of causality.
+such as polygenic scores, could help to pin down the direction of causality,
+for example via Mendelian randomization [@davey2003mendelian].
 Conversely, theory and empirical results from economics can shine a light on the
@@
 
-```{r plot-sexes, fig.cap = "Selection effects by sex. Lines show significant differences."}
+```{r plot-sexes, fig.cap = "Selection effects by sex. Highlighted lines are significant differences. Highlighted points are significantly different from 0."}
 
@@
 
-```{r plot-purifying, fig.align = "center", fig.cap = "Purifying/diversifying selection: predicted number of children by polygenic score", fig.height = 5}
+```{r plot-purifying, fig.align = "center", fig.cap = "Purifying/diversifying selection: predicted number of children by polygenic score.", fig.height = 5}
 
@@
 
-```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by Townsend deprivation quintile of birth area", fig.align = "center"}
+```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by Townsend deprivation quintile of birth area.", fig.align = "center"}
 
@@
         reg.type = fct_recode(reg.type, 
-                               Residualized   = "controlled", 
+                               `Resid. for 100 PCs`  = "controlled", 
                                Unresidualized        = "raw"
@@
         theme(
-          axis.text.x = element_text(size = 8)
+          axis.text.x  = element_text(size = 7),
+          legend.text  = element_text(size = 8),
+          legend.title = element_text(size = 9)
         )
@@
 
-Another way to examine the "earnings" theory of natural selection is to compare
-selection effects of PGS with their genetic correlation with educational
-attainment (EA3). Since EA3 strongly predicts earnings, if earnings drives
-differences in fertility, we'd expect a correlation between the two sets of
-results. Figure \@ref(fig:plot-rgs-by-effect-size) shows this is so. Genetic
-correlations were calculated using LD score regression from GWAS summary
-statistics.
-
 ```{r plot-rgs-by-effect-size, fig.cap = "Selection effects plotted against genetic correlation with EA3."}
@@
 
+rgs_cor <- cor.test(~ rg + estimate, 
+                      data = rgs %>%
+                               filter(
+                                 p2 != "EA3_excl_23andMe_UK", 
+                                 p2 != "EA2_noUKB"
+                               )
+                    )
 rgs %>% 
@@
 
+Another way to examine the "earnings" theory of natural selection is to compare
+selection effects of polygenic scores with their genetic correlation with educational
+attainment (EA3). Since EA3 strongly predicts earnings, if earnings drives
+differences in fertility, we'd expect a correlation between the two sets of
+results. Figure \@ref(fig:plot-rgs-by-effect-size) shows this is so: the 
+correlation is `r rgs_cor$estimate`. Genetic
+correlations were calculated using LD score regression from GWAS summary
+statistics.
+
 \clearpage
@@
       theme_compact() %>% 
-      set_width(0.9) %>% 
-      set_col_width(c(.35, .15, .15 , .15)) %>% 
-      set_bottom_border(final(1), everywhere)
+      set_width(0.75) %>% 
+      set_col_width(c(.4, .2, .2, .2)) %>% 
+      set_bottom_border(final(1), everywhere) %>% 
+      set_position("center")
 ```

### 2020-12-22 make-data.R
Fix reverse-coded rgs

@@ -162,6 +162,10 @@ make_rgs <- function (rgs_file) {
     autism_2017.ipsych.pgc     = "autism_2017",
   ) %>% {sub("\\.GPC\\.23andme$", "", .)}
   
+  reverse_coded <- c("ai_substance_use", "dpw_substance_use", 
+                       "cpd_substance_use", "si_substance_use")
+  rgs$rg[rgs$p2 %in% reverse_coded] <- rgs$rg[rgs$p2 %in% reverse_coded] * -1
+  
   rgs
 }
 

### 2020-12-22 why-natural-selection.Rmd
Fix reverse-coded rgs
Tweaks

@@
 
-```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners, household type, and age at first live birth.", fig.subcap = c("Lifetime number of sexual partners", "Household type", "Age at first live birth, females"), fig.ncol = 1, fig.align = "center"}
+```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and household type.", fig.subcap = c("Lifetime number of sexual partners", "Household type"), fig.ncol = 1, fig.align = "center"}
 
@@
 
-```{r plot-age-flb, fig.cap = "Selection effects by age at first live birth, and controlling for age at first live birth (women only).", fig.ncol = 1,  fig.subcap = c("Effect sizes by age at first live birth terciles", "Effect sizes controlling for age at first live birth.
+```{r plot-age-flb, fig.cap = "Selection effects by age at first live birth, and controlling for age at first live birth (women only).", fig.ncol = 1,  fig.subcap = c("Effect sizes by age at first live birth terciles (women only)", "Effect sizes controlling for age at first live birth.
 Effect sizes without controls (for women only) shown for comparison."), fig.align = "center"}
@@
 results. Figure \@ref(fig:plot-rgs-by-effect-size) shows this is so: the 
-correlation is `r rgs_cor$estimate`. Genetic
+correlation, after excluding EA2, is `r rgs_cor$estimate`. Genetic
 correlations were calculated using LD score regression from GWAS summary

### 2021-01-11 why-natural-selection.Rmd
Tweaks for UKBB; formatting.

@@
 date: "`r format(Sys.Date(), '%d %B %Y')`"
-abstract: "Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it.
+abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it.
 We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry).
@@
 Effect sizes are substantially larger after correcting for ascertainment bias.
-The direction of natural selection is reversed among older parents, or after controlling for age at first live birth.
+The direction of natural selection is reversed among older parents (22+), or after controlling for age at first live birth.
 These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market.
-We find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects and that are strongly related to the socio-economic system."
+We find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
 bibliography: "negative-selection.bib"
@@
 (time spent on childrearing has a higher cost in foregone earnings). Our
-results suggest that the substitution effect dominates for single parents,
-while the income effect is stronger for couples.
+results suggest that the substitution effect dominates for single parents and
+younger parents, while among couples and older parents (22+) effects are more
+evenly balanced.
 
@@
 
-```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and household type.", fig.subcap = c("Lifetime number of sexual partners", "Household type"), fig.ncol = 1, fig.align = "center"}
+```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and household type.", fig.subcap = c("Lifetime number of sexual partners", "Household type"), fig.ncol = 1, fig.align = "center", fig.height = 4}
 
@@
                   legend.box.spacing   = unit(0, "lines"),
-                  legend.box.margin    = margin(-5, 0, 0, -5),
+                  legend.box.margin    = margin(-5, 0, 0, -85),
                   legend.margin        = margin(0, 0, 0, 0),
                   legend.box.just      = "left",
-                  legend.justification = "left"
+                  legend.justification = "left",
+                  legend.spacing       = unit(1.5, "lines")
                 )
 sexes_guides <- guides(
-                  # fill  = guide_legend(title.position = "top"),
-                  # color = guide_legend(title.position = "top")
+                  # fill  = guide_legend(ncol = 1),
+                  # color = guide_legend(ncol = 1)
                 )
@@
 job, and against its correlation with educational attainment, a
-predictor of lifetime earnings. The raw relationships (left column) are strongly
+predictor of lifetime earnings.^[animation] The raw relationships (left column) are strongly
 negative. If we plot selection effects controlling for age at first live birth
@@
 between income and lower fertility [@becker1960economic;@jones2006economic].
+
+^[animation]: See also the supplementary animation at https://xxx.xxx.xxx
 
@@
 including people with lower income, lower education, younger first parenthood, 
-and more lifetime sexual partners. Indeed, among older mothers, selection effects are
+and more lifetime sexual partners. Indeed, among older mothers (22+), selection effects are
 actually reversed. Furthermore, the size of selection effects on a polygenic
@@
 
-Because selection effects are concentrated in poorer groups, they may also increase
-inequality with respect to polygenic scores. For example, Figure
+Because selection effects are concentrated in lower-income groups, they may also
+increase inequality with respect to polygenic scores. For example, Figure
 \@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
@@
 educational attainment, which lowers the mean, is stronger at the bottom of the
-income distribution, and this increases the differences between groups.  Overall,
-natural selection increases inequality for 29 out of 33 polygenic scores 
-(Appendix Table \@ref(tab:tbl-inequality-effects)).
+income distribution, and this increases the differences between groups.
+Overall, natural selection increases inequality for 29 out of 33 polygenic
+scores (Appendix Table \@ref(tab:tbl-inequality-effects)). Since many polygenic
+scores are predictive of disease risk, this increase in inequality could
+potentially increase health inequalities. In general, the demographic history of
+anatomically modern humans is related to disease risk [@benton2021influence];
+understanding the role of contemporary natural selection may help in the design
+of health policies.
 

### 2021-01-12 why-natural-selection.Rmd
Final tweaks

@@
 abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it.
+Studying these can help us understand how the genetic architecture of health outcomes arise.
 We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry).
-Consistently over time, polygenic scores which predict lower (higher) earnings and education are selected for (against).
-Selection effects are concentrated among lower-income and less educated people, younger parents, people with more lifetime sexual partners, and people not living with a partner.
-Effect sizes are substantially larger after correcting for ascertainment bias.
+Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against).
+Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner.
 The direction of natural selection is reversed among older parents (22+), or after controlling for age at first live birth.
@@
 mechanisms behind natural selection can help us better understand how individual
-differences in complex traits and disease risk arise. Recent work confirms that
+differences in complex traits and disease risk arise [@benton2021influence]. Recent work confirms that
 natural selection is taking place in modern human populations, using genome-wide
@@
 and they may also differ on other unobserved characteristics
-[@10.1093/aje/kwx246]. Since richer and educated people also live longer, this
+[@10.1093/aje/kwx246]. Since richer and more educated people also live longer, this
 bias could also increase with age.
@@
 on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
-on each polygenic score against that score's correlation with earnings in a respondent's first
-job, and against its correlation with educational attainment, a
-predictor of lifetime earnings.^[animation] The raw relationships (left column) are strongly
-negative. If we plot selection effects controlling for age at first live birth
-(right column), the effect reverses and becomes positive. Thus, the labour
-market appears to play an important role in natural selection. Substitution
-effects dominate income effects overall, which fits the known association
-between income and lower fertility [@becker1960economic;@jones2006economic].
-
-^[animation]: See also the supplementary animation at https://xxx.xxx.xxx
+on each polygenic score against that score's correlation with earnings in a
+respondent's first job, and against its correlation with educational attainment,
+a predictor of lifetime earnings. (See also the Supplementary Animations for the
+uncorrected effects per AFLB group, which show the correlation reversing after
+age 22.) The raw relationships (left column) are strongly negative. If we plot
+selection effects controlling for age at first live birth (right column), the
+effect reverses and becomes positive. Thus, the labour market appears to play an
+important role in natural selection. Substitution effects dominate income
+effects overall, which fits the known association between income and lower
+fertility [@becker1960economic;@jones2006economic].
 
@@
 
-Polygenic scores which correlate with high (low) earnings and more (less) education are being
-selected against (selected for). In addition, many of the phenotypes under
-positive selection are linked to disease risk, or are what many people would see
-as undesirable to have. For example, most people would probably prefer to have
-high educational attainment, a low risk of major depressive disorder, and a low
-risk of coronary artery disease, but natural selection is pushing against
-genes associated with these traits. Potentially, this could increase the health
-burden on modern populations, but that depends on effect sizes. Our results
-suggest that naïve estimates can be affected by sample ascertainment bias. This
-problem may be less serious in surveys which aim to be representative (as the UK
-Biobank does not). However, there is still scope for bias, since not all
-respondents consent to the collection of genetic data. For instance, completion
-rates for genotype data in the US Health and Retirement Study were around 80-85%
-[@hrs2020]. Researchers should be aware of the risks of ascertainment when
-studying modern natural selection.
+Polygenic scores which correlate with high (low) earnings and more (less)
+education are being selected against (selected for). In addition, many of the
+phenotypes under positive selection are linked to disease risk, or are what many
+people would see as undesirable to have. For example, most people would probably
+prefer to have high educational attainment, a low risk of ADHD and major
+depressive disorder, and a low risk of coronary artery disease, but natural
+selection is pushing against genes associated with these traits. Potentially,
+this could increase the health burden on modern populations, but that depends on
+effect sizes. Our results suggest that naïve estimates can be affected by sample
+ascertainment bias. This problem may be less serious in surveys which aim to be
+representative (as the UK Biobank does not). However, there is still scope for
+bias, since not all respondents consent to the collection of genetic data. For
+instance, completion rates for genotype data in the US Health and Retirement
+Study were around 80-85% [@hrs2020]. Researchers should be aware of the risks of
+ascertainment when studying modern natural selection.
 
@@
 scores are predictive of disease risk, this increase in inequality could
-potentially increase health inequalities. In general, the demographic history of
+potentially increase health inequalities. In general, the evolutionary history of
 anatomically modern humans is related to disease risk [@benton2021influence];
-understanding the role of contemporary natural selection may help in the design
-of health policies.
+understanding the role of contemporary natural selection may aid in mapping the
+genetic architecture of current health disparities.
 

### 2021-01-12 why-natural-selection.Rmd
small abstract tweak

@@
 abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it.
-Studying these can help us understand how the genetic architecture of health outcomes arise.
 We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry).
@@
 These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market.
-We find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
+Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
 bibliography: "negative-selection.bib"

### 2021-04-13 model-appendix.Rmd
Include model in main paper

\externaldocument{why-natural-selection}

\section*{Appendix}

\section*{Solution for the one-period model}

Differentiating and setting $\frac{dU}{dN}=0$ gives the first order
condition for an optimal choice of children $N^{*}>0$:

\[
\frac{bW}{(W(1-bN^{*}))^{\sigma}}\ge a\textrm{, with equality if }N^{*}>0.
\]
Rearranging gives
\begin{equation}
N^{*}=\max\left\{ \frac{1}{b}\left(1-\left(\frac{b}{a}\right)^{1/\sigma}W^{(1-\sigma)/\sigma}\right),0\right\} .\label{eq:N-one-period}
\end{equation}
Note that when $\sigma<1$, for high enough $W$, $N^{*}=0$. Differentiating
gives the effect of wages on fertility for $N^{*}>0$. This is also
the fertility-human capital relationship:

\begin{equation}
\frac{dN^{*}}{dh}=\frac{dN^{*}}{dW}=-\frac{1}{b}\left(\frac{b}{a}\right)^{1/\sigma}\frac{1-\sigma}{\sigma}W^{(1-2\sigma)/\sigma}.\label{eq:DNdW-one-period}
\end{equation}
This is negative if $\sigma<1$. Also, 
\[
\frac{d^{2}N^{*}}{dW^{2}}=-\frac{1}{b}\left(\frac{b}{a}\right)^{1/\sigma}\frac{1-\sigma}{\sigma}\frac{1-2\sigma}{\sigma}W^{(1-3\sigma)/\sigma}
\]
For $0.5<\sigma<1$, this is positive, so the effect of fertility
on wages shrinks towards zero as wages increase (and becomes 0 when
$N^{*}=0$). Next, we consider the time cost of children $b$:
\[
\frac{d^{2}N^{*}}{dWdb}=-\left(\frac{1}{a}\right)^{1/\sigma}\left(\frac{1-\sigma}{\sigma}\right)^{2}(Wb)^{(1-2\sigma)/\sigma}<0.
\]

Lastly we consider the effect of $a$. From (\ref{eq:N-one-period}),
$N^{*}$ is increasing in $a$. Differentiating (\ref{eq:DNdW-one-period})
by $a$ gives
\[
\frac{d^{2}N^{*}}{dadW}=b^{1/\sigma-1}\frac{1-\sigma}{\sigma^{2}}W^{(1-2\sigma)/\sigma}a^{-1/\sigma-1}
\]
which is positive for $\sigma<1$.

\section*{Solution for the two-period model}

Period 1 and period 2 income are:
\begin{align}
Y_{1} & =1-s-bN_{1}\label{eq:Y1}\\
Y_{2} & =w(s,h)(1-bN_{2})\label{eq:Y2}
\end{align}

Write the Lagrangian of utility $U$ (\ref{eq:U}) as 
\[
\mathcal{L}(N_{1},N_{2},s)=u(Y_{1})+u(Y_{2})+a(N_{1}+N_{2})+\lambda_{1}N_{1}+\lambda_{2}N_{2}+\lambda_{3}(\frac{1}{b}-N_{2})+\mu s
\]

Lemma \ref{lemma-concavity} below shows that if $\sigma>0.5$, this
problem is globally concave, guaranteeing that the first order conditions
identify a unique solution. We assume $\sigma>0.5$ from here on.

Plugging (\ref{eq:Y1}) and (\ref{eq:Y2}) into the above, we can
derive the Karush-Kuhn-Tucker conditions for an optimum $(N_{1}^{*},N_{2}^{*},s^{*})$
as:
\begin{align}
\frac{d\mathcal{L}}{dN_{1}}=-bY_{1}^{-\sigma}+a+\lambda_{1} & =0\textrm{, with }\lambda_{1}=0\textrm{ if }N_{1}^{*}>0;\label{eq:kkt1}\\
\frac{d\mathcal{L}}{dN_{2}}=-bs^{*}hY_{2}^{-\sigma}+a+\lambda_{2} & -\lambda_{3}=0\textrm{, with }\lambda_{2}=0\textrm{ if }N_{2}^{*}>0,\lambda_{3}=0\textrm{ if }N_{2}^{*}<\frac{1}{b};\label{eq:kkt2}\\
\frac{d\mathcal{L}}{ds}=-Y_{1}^{-\sigma}+h(1-bN_{2}^{*})Y_{2}^{-\sigma}+\mu & =0;\label{eq:kkt3}\\
N_{1}^{*},N_{2}^{*},s^{*},\lambda_{1},\lambda_{2},\lambda_{3},\mu & \ge0;N_{2}^{*}\le\frac{1}{b}.
\end{align}
Note that the Inada condition ($\lim_{x\rightarrow0}u'(x)=\infty$)
for period 1 rules out $s^{*}=1$ and $N_{1}=1/b$, so we need not
impose these constraints explicitly. Also, so long as $N_{2}^{*}<1/b$,
the same condition rules out $s^{*}=0$. We consider four cases,
of which only three can occur.

\subsection*{Case 1: $N_{1}^{*}>0,N_{2}^{*}>0$}

Rearranging (\ref{eq:kkt1}), (\ref{eq:kkt2}) and (\ref{eq:kkt3})
gives:
\begin{align}
N_{1}^{*} & =\frac{1}{b}\left(1-s^{*}-\left(\frac{b}{a}\right)^{1/\sigma}\right);\label{eq:N11}\\
N_{2}^{*} & =\frac{1}{b}\left(1-\left(\frac{b}{a}\right)^{1/\sigma}(s^{*}h)^{(1-\sigma)/\sigma}\right);\label{eq:N21}\\
s^{*} & =\frac{1-bN_{1}^{*}}{1+\left((1-bN_{2}^{*})h\right)^{1-1/\sigma}}.
\end{align}
Plugging the expressions for $N_{1}^{*}$ and $N_{2}^{*}$ into $s^{*}$gives
\[
s^{*}=\frac{s^{*}+\left(\frac{b}{a}\right)^{1/\sigma}}{1+\left(\left(\frac{b}{a}\right)^{1/\sigma}s^{*(1-\sigma)/\sigma}h{}^{1/\sigma}\right)^{1-1/\sigma}}
\]
which simplifies to
\begin{equation}
s^{*}=\left(\frac{b}{a}\right)^{1/(2\sigma-1)}h^{(1-\sigma)/(2\sigma-1)}.\label{eq:s1}
\end{equation}

Plugging the above into (\ref{eq:N11}) and (\ref{eq:N21}) gives:
\begin{align*}
N_{1}^{*} & =\frac{1}{b}\left(1-\left(\frac{b}{a}\right)^{1/(2\sigma-1)}h^{(1-\sigma)/(2\sigma-1)}-\left(\frac{b}{a}\right)^{1/\sigma}\right);\\
N_{2}^{*} & =\frac{1}{b}\left(1-\left(\frac{b}{a}\right)^{1/(2\sigma-1)}h^{(1-\sigma)/(2\sigma-1)}\right).
\end{align*}
Note that that $N_{1}^{*}<N_{2}^{*}$. For these both to be positive
requires low values of $h$ if $\sigma<1$ and high values of $h$
if $\sigma>1$. Also:
\[
w(s^{*},h)\equiv s^{*}h=\left(\frac{b}{a}\right)^{1/(2\sigma-1)}h^{\sigma/(2\sigma-1)}.
\]

Observe that $w(s^{*},h)$ is increasing in $h$ for $\sigma>0.5$,
and convex iff $0.5<\sigma<1$.

While $N_{1}^{*}$ and $N_{2}^{*}$ are positive, they have the same
derivative with respect to $h$:
\begin{equation}
\frac{dN_{t}^{*}}{dh}=-\frac{1}{b}\left(\frac{b}{a}\right)^{1/(2\sigma-1)}\frac{1-\sigma}{2\sigma-1}h^{(1-\sigma)/(2\sigma-1)-1}\label{eq:dNdh1}
\end{equation}
Examining this and expression (\ref{eq:s1}) gives:
\begin{lem}
\label{case-1-lemma}For $\sigma<1$, case 1 holds for $h$ low enough,
and in case 1, $N_{1}^{*}$ and $N_{2}^{*}$ decrease in $h$, while
$s^{*}$ increases in $h$.

For $\sigma>1$, case 1 holds for $h$ high enough, and in case 1
$N_{1}^{*}$ and $N_{2}^{*}$ increase in $h$, while $s^{*}$ decreases
in $h$.

$N_{t}^{*}$ is convex in $h$ for $\sigma>2/3$, and concave otherwise.
$s^{*}$ is convex in $h$ if $\sigma<2/3$, and concave otherwise.
\end{lem}

\subsection*{Case 2: $N_{1}^{*}=0,N_{2}^{*}>0$}

Replace $N_{1}^{*}=0$ into the first order condition for $s^{*}$
from (\ref{eq:kkt3}), and rearrange to give:
\[
s^{*}=\frac{1}{1+\left((1-bN_{2})h\right)^{1-1/\sigma}}.
\]

Now since $N_{2}^{*}>0$, we can rearrange (\ref{eq:kkt2}) to give
\begin{equation}
N_{2}^{*}=\frac{1}{b}\left(1-\left(\frac{b}{a}\right)^{1/\sigma}(s^{*}h)^{(1-\sigma)/\sigma}\right).\label{eq:N22}
\end{equation}

Plugging this into $s^{*}$ gives
\begin{align*}
s^{*} & =\frac{1}{1+\left(\frac{bh}{a}\right)^{(\sigma-1)/\sigma^{2}}(s^{*}){}^{-(1-\sigma)^{2}/\sigma^{2}}}
\end{align*}
which can be rearranged to
\begin{equation}
(1-s^{*})(s^{*})^{(1-2\sigma)/\sigma^{2}}=\left(\frac{a}{bh}\right)^{(1-\sigma)/\sigma^{2}}.\label{eq:estar2}
\end{equation}
Differentiate the left hand side of the above to get
\begin{align}
 & \frac{1-2\sigma}{\sigma^{2}}(1-s^{*})(s^{*})^{(1-2\sigma)/\sigma^{2}-1}-(s^{*})^{(1-2\sigma)/\sigma^{2}}\nonumber \\
= & \frac{1-2\sigma}{\sigma^{2}}(s^{*})^{(1-2\sigma)/\sigma^{2}-1}-\frac{\sigma^{2}+1-2\sigma}{\sigma^{2}}(s^{*})^{(1-2\sigma)/\sigma^{2}}\nonumber \\
= & \frac{1-2\sigma}{\sigma^{2}}(s^{*})^{(1-2\sigma)/\sigma^{2}-1}-\frac{(1-\sigma)^{2}}{\sigma^{2}}(s^{*})^{(1-2\sigma)/\sigma^{2}}.\label{eq:destar-lhs}
\end{align}
This is negative if and only if
\begin{align*}
s^{*} & >\frac{1-2\sigma}{(1-\sigma)^{2}}
\end{align*}
which is always true since $\sigma>0.5$. Note also that since $\sigma>0.5$,
then the left hand side of (\ref{eq:estar2}) approaches infinity
as $s^{*}\rightarrow0$ and approaches 0 as $s^{*}\rightarrow1$.
Thus, (\ref{eq:estar2}) implicitly defines the unique solution for
$s^{*}$.

Conditions (\ref{eq:kkt1}) and (\ref{eq:kkt2}) give the bounds for
this case. When $\sigma<1$, (\ref{eq:kkt1}) puts a minimum on $h$
and (\ref{eq:kkt2}) puts a maximum on $h$. The situation is reversed
for $\sigma>1$.

To find how $s^{*}$ changes with $h$, note that the right hand side
of the above decreases in $h$ for $\sigma<1$, and increases in $h$
for $\sigma>1$. Putting these facts together: for $\sigma<1$, when
$h$ increases the RHS of (\ref{eq:estar2}) decreases, hence the
LHS decreases and $s^{*}$ increases, i.e. $s^{*}$ is increasing
in $h$. For $\sigma>1$, $s^{*}$ is decreasing in $h$.

To find how $N_{2}^{*}$ changes with $h$, we differentiate (\ref{eq:N22}):

\begin{align}
\frac{dN_{2}^{*}}{dh} & =-\frac{1}{b}\left(\frac{b}{a}\right)^{1/\sigma}\frac{1-\sigma}{\sigma}(s^{*}h)^{(1-2\sigma)/\sigma}(s^{*}+h\frac{ds^{*}}{dh})\label{eq:dNdh2}
\end{align}
which is negative for $\sigma<1$, since $\frac{ds^{*}}{dh}>0$ in
this case.

Differentiating again:

\begin{align*}
\frac{d^{2}N_{2}}{dh^{2}} & =-X[\frac{1-2\sigma}{\sigma}(s^{*}h)^{(1-3\sigma)/\sigma}(s^{*}+h\frac{ds^{*}}{dh})^{2}+(s^{*}h)^{(1-2\sigma)/\sigma}(2\frac{ds^{*}}{dh}+h\frac{d^{2}s^{*}}{dh^{2}})]\\
 & =X(s^{*}h)^{(1-3\sigma)/\sigma}[\frac{2\sigma-1}{\sigma}(s^{*}+h\frac{ds^{*}}{dh})^{2}-(s^{*}h)(2\frac{ds^{*}}{dh}+h\frac{d^{2}s^{*}}{dh^{2}})]
\end{align*}
where $X=\frac{1}{b}\left(\frac{b}{a}\right)^{1/\sigma}\frac{1-\sigma}{\sigma}>0$.
Note that $\frac{d^{2}N_{2}}{dh^{2}}$ is continuous in $\sigma$
around $\sigma=1$. Note also from (\ref{eq:estar2}) that for $\sigma=1$,
$s^{*}$ becomes constant in $\sigma$. The term in square brackets
then reduces to $(s^{*})^{2}>0$. Putting these facts together, for
$\sigma$ sufficiently close to 1, $\frac{d^{2}N_{2}^{*}}{dh^{2}}>0$,
i.e. $N_{2}^{*}$ is convex in $h$.

Summarizing:
\begin{lem}
\label{case-2-lemma}Case 2 holds for intermediate values of $h$.
In case 2: for $\sigma<1$, $s^{*}$ is increasing in $h$ and $N_{2}^{*}$
is decreasing in $h$. For $\sigma>1$, $s^{*}$is decreasing in $h$.
For $\sigma$ close enough to 1, $N_{2}^{*}$ is convex in $h$.
\end{lem}

\subsection*{Case 3: $N_{1}^{*}=0,N_{2}^{*}=0$}

We can solve for $s^{*}$ by substituting values of $Y_{1}$ and $Y_{2}$
into (\ref{eq:kkt3}):
\[
-(1-s^{*}){}^{-\sigma}+h(s^{*}h)^{-\sigma}=0
\]
which rearranges to 
\begin{equation}
s^{*}=\frac{1}{1+h^{(\sigma-1)/\sigma}}.\label{eq:estar1}
\end{equation}

Conditions (\ref{eq:kkt1}) and (\ref{eq:kkt2}) become:
\begin{align*}
-b(1-s^{*})^{-\sigma}+a & \le0\\
-bs^{*}h(s^{*}h)^{-\sigma}+a & \le0
\end{align*}
equivalently
\begin{align*}
\frac{a}{b} & \le(1-s^{*})^{-\sigma}\\
\frac{a}{b} & \le s^{*}h(s^{*}h)^{-\sigma}
\end{align*}
which can both be satisfied for $a/b$ close enough to zero. Note
from (\ref{eq:estar1}) that as $h\rightarrow\infty$, $s^{*}$ increases
towards 1 for $\sigma<1$, and decreases towards 0 for $\sigma>1$.
Note also that the right hand side of the first inequality above approaches
infinity as $s^{*}\rightarrow1$, therefore also as $h\rightarrow\infty$
for $\sigma<1$. Rewrite the second inequality as
\[
\frac{a}{b}<(s^{*}h)^{1-\sigma}=\left(\frac{h}{1+h^{(\sigma-1)/\sigma}}\right)^{1-\sigma}=\left(h^{-1}+h^{-1/\sigma}\right)^{\sigma-1}
\]
and note that again, as $h\rightarrow\infty$, the RHS increases towards
infinity for $\sigma<1$, and decreases towards zero otherwise. Thus,
for $\sigma<1$, both equations will be satisfied for $h$ high enough.
For $\sigma>1$, they will be satisfied for $h$ low enough. Summarizing
\begin{lem}
\label{case-3-lemma}For $\sigma<1$, case 3 holds for $h$ high enough,
and in case 3, $s^{*}$ increases in $h$. For $\sigma>1$, case 3
holds for $h$ low enough and $s^{*}$ decreases in $h$.
\end{lem}

\subsection*{Case 4: $N_{1}^{*}>0,N_{2}^{*}=0$}

Rearranging the first order conditions (\ref{eq:kkt1}) and (\ref{eq:kkt2})
for $N_{1}^{*}$ and $N_{2}^{*}$ gives
\begin{align*}
\frac{a}{b} & =(1-s^{*}-bN_{1}^{*})^{-\sigma}\\
\frac{a}{b} & \le s^{*}hY_{2}^{-\sigma}
\end{align*}
hence
\begin{align*}
(1-s^{*}-bN_{1}^{*})^{-\sigma} & \le s^{*}hY_{2}^{-\sigma}=(s^{*}h)^{1-\sigma}\\
\Leftrightarrow(1-s^{*}-bN_{1}^{*})^{\sigma} & \ge(s^{*}h)^{\sigma-1}\\
\Leftrightarrow1-s^{*}-bN_{1}^{*} & \ge(s^{*}h)^{1-1/\sigma}
\end{align*}

Now rearrange the first order condition for $s^{*}$ from (\ref{eq:kkt3}),
noting that since $N_{2}^{*}=0$, $s^{*}>0$ by the Inada condition.
\begin{align*}
h^{1/\sigma-1}(1-s^{*}-bN_{1}^{*}) & =s^{*}\\
1-s^{*}-bN_{1}^{*} & =s^{*}h^{1-1/\sigma}
\end{align*}
This, combined with the previous inequality, implies 
\begin{align*}
(s^{*}h)^{1-1/\sigma} & \le s^{*}h^{1-1/\sigma}\\
\Leftrightarrow(s^{*})^{-1/\sigma} & \le1
\end{align*}
which cannot hold since $0<s^{*}<1$.

\subsection*{Comparative statics}

We can now examine how the fertility-human capital relationship 
\[
\frac{dN^{*}}{dh},\textrm{ where }N^{*}\equiv N_{1}^{*}+N_{2}^{*},
\]

changes with respect to other parameters. We focus on the case $\sigma<1$,
since it gives the closest match to our observations, and since it
also generates ``reasonable'' predictions in other areas, e.g. that
education levels increase with human capital. Figure \ref{fig:theory}
shows how $N^{*}$ changes with $h$ for $a=0.4,b=0.25,\sigma=0.7$.

\begin{figure}
\begin{centering}
\includegraphics[width=9cm, height=6cm]{N-plot}
\par\end{centering}
\caption{\label{fig:theory}Fertility vs. human capital in the two-period model
with $a=0.4,b=0.25,\sigma=0.7$.}
\end{figure}

From Lemmas \ref{case-1-lemma}, \ref{case-2-lemma} and \ref{case-3-lemma},
as $h$ increases we move from $N_{1}^{*},N_{2}^{*}>0$ to $N_{1}^{*}=0,N_{2}^{*}>0$
to $N_{1}^{*}=N_{2}^{*}=0$. Furthermore, for $\sigma>2/3$, $N_{1}^{*}$
and $N_{2}^{*}$ are convex in $h$ when they are both positive, and
for $\sigma$ close enough to 1, $N_{2}^{*}$ is convex in $h$ when
$N_{1}^{*}=0$. To show global concavity of $N^{*}$ in $h$ as $\sigma\nearrow1$,
all that remains is to check that the derivative is increasing around
the points where these 3 regions meet. That is trivially satisfied
where $N_{2}^{*}$ becomes 0, since thereafter $\frac{dN^{*}}{dh}$
is zero. The derivative as $N_{1}^{*}$ approaches zero is twice the
expression in (\ref{eq:dNdh1}):
\begin{equation}
-\frac{2}{b}\left(\frac{b}{a}\right)^{1/(2\sigma-1)}\frac{1-\sigma}{2\sigma-1}h^{(1-\sigma)/(2\sigma-1)-1}\label{eq:dNonleft}
\end{equation}
and the derivative to the right of this point is given by (\ref{eq:dNdh2}):
\begin{equation}
-\frac{1}{b}\left(\frac{b}{a}\right)^{1/\sigma}\frac{1-\sigma}{\sigma}(s^{*}h)^{(1-2\sigma)/\sigma}(s^{*}+h\frac{ds^{*}}{dh})\label{eq:dNonright}
\end{equation}

We want to prove that the former is larger in magnitude (i.e. more
negative). Dividing (\ref{eq:dNonleft}) by (\ref{eq:dNonright})
gives
\[
2\frac{\sigma}{2\sigma-1}\left(\frac{b}{a}\right)^{(1-\sigma)/(\sigma(2\sigma-1))}\frac{h^{(1-\sigma)^{2}/(\sigma(2\sigma-1))}}{s^{*}(s^{*}+h\frac{ds^{*}}{dh})}
\]

Examining (\ref{eq:estar2}) shows that as $\sigma\rightarrow1$,
$s^{*}\rightarrow0.5$ and $\frac{ds^{*}}{dh}\rightarrow0$, and therefore
the above approaches
\[
2\frac{1}{(0.5)^{2}}=8.
\]

Thus, as $\sigma\nearrow1$, $N^{*}$ is guaranteed to be globally
convex in $h$.

We can now prove the facts stated in the main text.

\textbf{Fact 1}: for $\sigma<1$, total fertility $N^{*}\equiv N_{1}^{*}+N_{2}^{*}$
is decreasing in human capital $h$.

Furthermore, for $\sigma$ close enough to 1, fertility is convex
in human capital, i.e.

\textbf{Fact 2 part 1}: the fertility-human capital relationship is
closer to 0 at high levels of $h$.

For $\sigma<1$, education levels $s^{*}$ increase in $h$, and so
therefore do equilibrium wages $w(s^{*},h)$. This, plus fact 1, gives:

\textbf{Fact 2 part 2}: for $\sigma<1$ and close to 1, the fertility-human
capital relationship is weaker among higher earners.

\textbf{Fact 4}: for $\sigma<1$ and close to 1, the fertility-human
capital relationship is weaker at high levels of education.

Next, we compare people who start fertility early ($N_{1}^{*}>0$)
versus those who start fertility late ($N_{1}^{*}=0$). Again, for
$\sigma<1$ the former group have lower $h$ than the latter group.
Thus we have:

\textbf{Fact 5}: for $\sigma<1$ and close to 1, the fertility-human
capital relationship is weaker among those who start fertility late.

Lastly, we prove fact 3. Differentiating $dN_{t}^{*}/dh$ in (\ref{eq:dNdh1})
with respect to $b$, for when $N_{1}^{*}>0$ gives:
\begin{align*}
\frac{d^{2}N_{t}^{*}}{dhdb} & =\frac{2\sigma-2}{2\sigma-1}b^{(3-4\sigma)/(2\sigma-1)}\left(\frac{1}{a}\right)^{1/(2\sigma-1)}\frac{1-\sigma}{2\sigma-1}h^{(\sigma-1)^{2}/(\sigma(2\sigma-1))}
\end{align*}

which is negative for $0.5<\sigma<1$. When $N_{1}^{*}=0$, differentiating
$dN_{2}^{*}/dh$ in (\ref{eq:dNdh2}) gives:

\[
\frac{d^{2}N_{2}^{*}}{dhdb}=-\frac{1-\sigma}{\sigma}b^{(1-2\sigma)/\sigma}\left(\frac{1}{a}\right)^{1/\sigma}\frac{1-\sigma}{\sigma}(s^{*}h)^{(1-2\sigma)/\sigma}(s^{*}+h\frac{ds^{*}}{dh})
\]
which again is negative for $\sigma<1$. Therefore:

\textbf{Fact 3}: for $\sigma<1$, the fertility-human capital relationship
is more negative when the burden of childcare $b$ is larger.

\subsection*{Concavity}
\begin{lem}
\label{lemma-concavity}For $\sigma>0.5$, $U$ in equation (\ref{eq:U})
is concave in $N_{1},N_{2}$ and $s$.
\end{lem}
\begin{proof}
We examine the Hessian matrix of utility in each period. Note that
period 1 utility is constant in $N_{2}$ and period 2 utility is constant
in $N_{1}$. For period 1 the Hessian with respect to $N_{1}$and
$s$ is: 
\[
\left[\begin{array}{cc}
d^{2}u/dN_{1}^{2} & d^{2}u/dsdN_{1}\\
d^{2}u/dsdN_{1} & d^{2}u/ds^{2}
\end{array}\right]=\left[\begin{array}{cc}
-\sigma b^{2} & -\sigma b\\
-\sigma b & -\sigma
\end{array}\right]Y_{1}^{-\sigma-1}
\]
with determinant 
\[
(\sigma^{2}b^{2}-\sigma^{2}b^{2})Y_{1}^{-2\sigma-2}=0.
\]
Thus, first period utility is weakly concave. For period 2 with respect
to $N_{2}$ and $s$, the Hessian is:

\[
\left[\begin{array}{cc}
d^{2}u/dN_{2}^{2} & d^{2}u/dsdN_{2}\\
d^{2}u/dsdN_{2} & d^{2}u/ds^{2}dN_{2}
\end{array}\right]=\left[\begin{array}{cc}
-\sigma(bsh)^{2}Y_{2}^{-\sigma-1} & -(1-\sigma)bhY_{2}^{-\sigma}\\
-(1-\sigma)bhY_{2}^{-\sigma} & -\sigma[h(1-bN_{2}^{*})]^{2}Y_{2}^{-\sigma-1}
\end{array}\right]
\]
with determinant
\begin{align*}
 & (-\sigma(bsh)^{2}Y_{2}^{-\sigma-1})(-\sigma[h(1-bN_{2}^{*})]^{2}Y_{2}^{-\sigma-1})-(-(1-\sigma)bhY_{2}^{-\sigma})^{2}\\
= & \sigma^{2}(bsh)^{2}[h(1-bN_{2}^{*})]^{2}Y_{2}^{-2\sigma-2}-(1-\sigma)^{2}(bh)^{2}Y_{2}^{-2\sigma}\\
= & \sigma^{2}(bh)^{2}Y_{2}^{-2\sigma}-(1-\sigma)^{2}(bh)^{2}Y_{2}^{-2\sigma}\textrm{, using that }Y_{2}=(sh)(1-bN)\\
= & (bh)^{2}Y_{2}^{-2\sigma}(\sigma^{2}-(1-\sigma)^{2})
\end{align*}
which is positive if and only if $\sigma>0.5$. Thus, if $\sigma>0.5$
then the Hessian is negative definite and thus utility is concave;
this combined with weak concavity of period 1, and linearity of $a(N_{1}+N_{2})$,
shows that (\ref{eq:U}) is concave.
\end{proof}

### 2021-04-13 why-natural-selection.Rmd
update paper
Include model in main paper

--- 2021-01-12 why-natural-selection.Rmd
@@
 ---
-title: "Natural selection in contemporary humans is linked to income and substitution effects"
+title: "Natural selection in contemporary humans is driven by correlations with human capital"
 author: "David Hugh-Jones \\thanks{Corresponding author.
@@
   - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
+  - \usepackage{amsthm}
+  - \theoremstyle{plain}
+  - \newtheorem{lem}{\protect\lemmaname}
+  - \providecommand{\lemmaname}{Lemma}
+  - \usepackage{xr-hyper}
 editor_options: 
@@
 then genetic variants which affect earnings potential in the labour market may
-cause opposing effects on fertility. The income effect will cause natural
+cause opposing effects on fertility. The income effect will lead to natural
 selection in favour of earnings-increasing variants. The substitution effect
 will do the reverse.
+
+To demonstrate this, we write down a simple model of 
+fertility choices. $h$ is an individual's level of human capital, which 
+(for now) we identify with his or her wage $W$. Raising a child takes time $b$. People maximize utility
+from the number of children $N$, and income $Y\equiv(1-bN)W$. An
+individual's payoff is
+
+\[
+U(N)=u(Y)+aN.
+\]
+
+Here $a$ captures the strength of preference for children. $u(\cdot)$
+is concave and increasing. We treat $N$ as continuous, in line with
+the literature: this can be thought of as the expected number of children
+among similar people. The marginal benefit of an extra child is
+$\frac{dU}{dN}=-bWu'(Y)+a$. The effect of an increase in wages (or human capital) 
+on this marginal benefit is
+
+\[
+\frac{d^{2}U}{dNdh}=\underbrace{-bu'(Y)}_{\textrm{Substitution effect}}\underbrace{-bYu''(Y)}_{\textrm{Income effect}}.
+\]
+
+The *substitution effect* is negative and reflects that when
+wages increase, time devoted to childcare costs more in foregone income.
+The positive *income effect* reflects that when wage income is
+higher, the marginal loss of wages from children is less painful.
+Letting $u(\cdot)$ take the constant relative risk aversion (CRRA)
+form $u(y)=\frac{y^{1-\sigma}-1}{1-\sigma}$, where $\sigma>0$ captures
+risk aversion in income, we solve for optimal fertility $N^{*}$,
+and examine the *fertility-human capital relationship*, $\frac{dN^{*}}{dh}$.
+For moderate levels of risk aversion ($0.5<\sigma<1$):
+
+\begin{enumerate}
+\item The fertility-human capital relationship is negative: $\frac{dN^{*}}{dh}<0$.
+\item The relationship is weaker (closer to zero) at higher wages and/or
+levels of human capital.
+\item The relationship is more negative when the time burden of children,
+$b$, is larger.
+\end{enumerate}
+
+To examine education and fertility timing, we extend the model to
+two periods. For convenience we ignore time discounting, and assume
+that credit markets are imperfect so that agents cannot borrow. Write
+
+\begin{equation}
+U(N_{1},N_{2})=u(Y_{1})+u(Y_{2})+a(N_{1}+N_{2})\label{eq:U}
+\end{equation}
+
+Instead of identifying human capital $h$ with wages, we now allow
+individuals to choose a level of education $s\in[0,1]$, which has
+a time cost in period 1. Education increases period 2 wages, and is
+complementary to human capital $h>0$. For period 2 wages we assume
+the simple functional form $w(s,h)=sh$. We normalize period 1 wages
+to 1. Maintaining the assumption of CRRA utility, we examine total
+fertility $N^{*}\equiv N_{1}^{*}+N_{2}^{*}$. For $0.5<\sigma<1$,
+facts 1-3 continue to hold (fact 2 in a neighborhood of $\sigma=1$).
+In addition, for $\sigma$ in a neighborhood of 1:
+
+\begin{enumerate}
+\item [4.]\setcounter{enumi}{4}The fertility-human capital relationship
+is weaker at higher levels of education $s$.
+\item The relationship is weaker among those who start fertility in period
+2 ($N_{1}^{*}=0$) than among those who start fertility in period
+1 ($N_{1}^{*}>0$).
+\end{enumerate}
+
+Facts 1-5 match our empirical results. In this model, polygenic scores which
+correlate with human capital would have a negative correlation with fertility at
+low income and education levels, and a weaker or zero correlation at higher
+levels (Figure \@ref(fig:plot-income-educ-level)). For scores which are highly
+correlated with human capital, the correlation would also be weaker at higher
+levels of the score itself, giving rise to diversifying selection (Appendix
+Figure \@ref(fig:plot-purifying)). The correlation would be more negative when
+the time burden of children $b$ is high, e.g. for single parents (Figure
+\@ref(fig:plot-n-partners)). Lastly, the correlation would be weaker among those
+starting fertility later (Figure \@ref(fig:plot-age-flb)).
+
+An alternative theory is that polygenic scores correlate with the
+motivation to have children (parameter $a$ in the model). However, this theory
+would not explain why our results are weaker at higher incomes and education levels.
+Indeed, in the one-period model, while $\frac{dN^{*}}{da}>0$, this
+relationship actually becomes stronger at higher wages.
+
+The above does not provide a complete theory of fertility. Rather,
+it shows that a relatively simple economic model can explain many
+of our results. In the appendix, we discuss other models of fertility
+from the economic literature, as well as causal evidence for the relationship
+between human capital, income and fertility.
 
@@
 
-AA is supported by the Foundation Volksbond Rotterdam and by ZonMw grant 849200011 from The Netherlands Organisation for Health Research and Development.
-This study was conducted using UK Biobank resources under application numbers 40310 and 19127.
+AA is supported by the Foundation Volksbond Rotterdam and by ZonMw grant 
+849200011 from The Netherlands Organisation for Health Research and Development. 
+This study was conducted using UK Biobank resources under application numbers 
+40310 and 19127.
 
@@
 \FloatBarrier
+
+# Proofs
+
+```{r include-model-appendix, child="model-appendix.Rmd"}
+
+```
+
+\clearpage
+\FloatBarrier
 # References

### 2021-04-13 why-natural-selection.Rmd
responses to PNAS review

@@
 
-## LITERATURE NOTES
-
-* Lundberg and Pollak JEP 2007:
-    - several models where nonmarital fertility is more likely for poorer women
-    - Burdett and Ermisch 2002, Willis 1999, Neal 2004
-    - Unmarried parents may not be able to credibly commit to each other
-
-* Happel et al. 1984
-   - childbirth is chosen to minimize loss of earnings from interrupted work.
-   - early childbirth if your skills decay a lot  while childrearing. (No point
-     gaining them then losing them.) Late childbirth otherwise.
-   - late childbirth if you have acquired skills before marriage.
-   - Capital-market imperfections make it wise to delay childbirth (till man has
-     high income to cope while wife is at home). More so for men's wages which
-     rise faster.
-   - empirically: (among married couples) 
-     - high-skill occupation women have children later. Higher-earning men have 
-     them earlier.
-
-- Caucutt et al. 2001
-  - high waged and high educated women (of recent cohorts) more likely to
-    delay childbirth
-
-- Cohen et al. 2013
-  - analyse effect of Israeli subsidy
-  - along the way, find that income affects fertility negatively at low income,
-    positively at high income, supporting Becker and Tomes 76
-
-- Becker and Tomes 1976
-  - investment in children quality and quantity when quality must be equal
-    across children; and some quality comes from an innate endowment.
-  - "observed" quantity-of-children income elasticity rises as income rises
-  - while "observed" quality-of-children elasticity falls as income rises, 
-    b/c more of total quality must be paid for by parents themselves. 
-  - the need to pay more for quality results in lower quantity, since total cost is 
-    quantity X quality
-  - something fishy about argument somehow...
-
-- Lindo 2008
-  - men who lose a job have children sooner, but fewer children in total
-  - controls for all skills effects
-  - so overall effect of income is positive, but timing changes
-
-- Monstad et al. 2008
-  - using an education reform to vary education
-  - no evidence of a causal effect on fertility, but births start later
-  * maybe the usual correlation reflects underlying skills. Your data could
-    test this.
-
-- Autor et al. 2017
-  - the China shock (exogenous shock to income) 
-  - they use Chinese import penetration by industry (instrumented using
-    other countries) times industry employment shares; and separate it for
-    males and females
-  - shocks to male employment decreased fertility and marriage; 
-    increased unmarried mothers
-  - shocks to female employment *increase* marriage and motherhood
-  * this roughly seems to be the consensus
-  * Whereas, you find that male and female *skills* (as 'instrumented' by PGS)
-    have the same effect on fertility - i.e. less skills = higher fertility
-
-- Baudin et al. 2015
-  * look at childlessness vs. fertility (extensive/intensive margin)
-  * some people are too poor to afford children
-  * 'a husband alleviates the time cost of having children'. But for some groups (see appendix)
-    there is a +ve income/fertility gradient
-  * fertility decreases with female educn due to subst effect
-  * baby boom and bust in US: a decline in childlessness caused by poverty, then an increase
-    caused by substn effect ('opportunity cost')
-  * fertility decreases with educn both for single mothers and married women
-    - slightly (25%) higher decrease per year of educn (Table 2) for single mothers
-
 # WHAT MODEL FITS YOUR DATA
@@
 
+```{r calc-sex-ratio}
+female_ratio <- mean(famhist$sex == 0)
+```
+
 We created polygenic scores for 33 traits in `r comma_num(nrow(famhist))`
@@
 [@10.1093/aje/kwx246]. Since richer and more educated people also live longer, this
-bias could also increase with age.
+bias could also increase with age. Lastly, the sample sex ratio skews 
+`r female_ratio * 100`% female.
 
@@
 
-```{r calc-sex-ratio}
-female_ratio <- mean(famhist$sex == 0)
-```
-
 The "selection effect", $\beta$, reflects the strength of natural selection
@@
 the expected polygenic score among children of the sample
-[@beauchamp2016genetic].[^normalization] To learn more about the underlying
-mechanisms, we split the sample, starting with basic demographic variables
-including education, income and sex. These are all potential sources of
-ascertainment bias: as well as the ascertainment for income and education,
-mentioned above, the sample sex ratio skews `r female_ratio * 100`% female.
+[@beauchamp2016genetic].[^normalization]
 
@@
 
-```{r plot-income-educ-level, fig.cap = "Selection effects by education and income.
-Each point represents a single bivariate regression of number of children on a polygenic score.", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}
-
-drake::loadd(res_edu)
-n_regs <- as.double(nrow(res_edu))
-
-res_edu %>% 
-      mutate(
-        term = pretty_names(term), 
-        term = fct_reorder(term, estimate, order_abs(1)),
-        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
-        "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(y = term, x = estimate, colour = p < 0.05/{{n_regs}}, 
-        fill = `Age left FTE`)) +
-        my_vline +
-        pval_geom_point() + 
-        pval_scale() +
-        my_fill_scale(n = 3, direction = -1) +
-        my_labs() +
-        my_theme()
-
-drake::loadd(res_income)
-
-n_regs <- as.double(nrow(res_income))
-res_income %>% 
-      mutate(
-        term = pretty_names(term), 
-        term = fct_reorder(term, estimate, order_abs(1)),
-        Income = factor(income_cat, 
-          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
-        ),
-        Income = add_n(Income, "income_cat")
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(y = term, x = estimate, fill = Income, 
-        colour = p < 0.05/{{n_regs}})) +
-        my_vline +
-        pval_geom_point(alpha = 0.8) + 
-        my_fill_scale(n = 5, direction = -1) +
-        pval_scale() +
-        my_labs() +
-        my_theme()
-```
-
-Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
-polygenic score, grouping respondents by age of completing full-time education,
-and by household income. Effects are larger and more significant for the lowest
-income category, and for the lowest education category. Note that the overall
-effect is not a simple average of the effect among the different subgroups,
-because polygenic scores may also shift respondents between the subgroups. For
-example, a high polygenic score for educational attainment may predict having
-fewer children among early school leavers, but it may also increase the age at
-which a respondent leaves school.
-
-```{r calc-income-edu-controlled}
-
-drake::loadd(res_income_controlled, res_edu_controlled)
-n_regs_ic <- nrow(res_income_controlled)
-n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
-
-n_regs_educ <- nrow(res_edu_controlled)
-n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
-```
-
-These results could be explained by age, if older respondents have lower income
-and are less educated, and also show more natural selection on polygenic
-scores. However, if we rerun the regressions, interacting the polygenic
-score with income category and also with a quadratic in age, the
-interaction with income remains significant at 0.05/`r n_regs_ic` for
-`r n_sig_ic` out of `r n_regs_ic` regressions. Similarly if we interact
-the PGS with age of leaving full time education and a quadratic in age,
-the interaction with age leaving full time education remains significant at
-0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.
-
-Selection effects are also different between men and women (Appendix Figure
-\@ref(fig:plot-sexes)). Differences are particularly large for educational
-attainment, height and MDD. Several polygenic scores for mental illness and
-personality traits are more selected for (or less against) among women,
-including major depressive disorder (MDD), schizophrenia and neuroticism, while
-extraversion is more selected for among men. Scores for waist circumference and
-waist-hip ratio are less selected for among women, and scores for educational
-attainment are more selected against, though the difference is only significant
-for the ("EA3") educational attainment score of @lee2018gene. One
-possible reason for these sex differences is that polygenic scores may affect
-fertility via success in marriage markets, and men and women may value different
-characteristics in these.
-
-```{r calc-age-flb-cross}
-
-drake::loadd(res_age_flb_cross)
-
-cor_age_flb <- res_age_flb_cross %>% 
-      mutate(category = gsub(":.*", "", term)) %>% 
-      pivot_wider(score_name, names_from = category, values_from = estimate) %>%              {cor(.$`age_flb_cat10-22`, .$`age_flb_cat28-52`)}
-
-```
-
-We next focus on variables related to household type and reproductive
-strategy. We split males and females by lifetime number of sexual
-partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
-For both sexes, selection effects are larger and more significant among
-those with more than 3 partners. Next we split respondents by whether
-they were living with a spouse or partner at the time of interview.
-Effects are larger among those not living with a spouse or partner
-(Figure \@ref(fig:plot-n-partners)b).
-
-Lastly, we split female respondents by age at first live birth
-(AFLB).[^aflb-no-men] There is evidence for genetic effects on AFLB
-[@Barban_2016], and there is a close link between this variable and number of
-children born. Figure \@ref(fig:plot-age-flb)a shows effect sizes estimated
-separately for each tercile of AFLB. Several effects are strikingly different
-across terciles. ADHD and MDD are selected for amongst the youngest third of
-mothers, but selected against among the oldest two-thirds. Educational
-attainment is selected for among the oldest two-thirds of mothers, but is not
-significantly selected among the youngest third. Similarly, several polygenic
-scores for body measurements are selected against only among older mothers. The
-correlation between effect sizes for the youngest and oldest terciles is 
-`r cor_age_flb`.
-
-[^aflb-no-men]: This information is unavailable for men.
-
-```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and household type.", fig.subcap = c("Lifetime number of sexual partners", "Household type"), fig.ncol = 1, fig.align = "center", fig.height = 4}
-
-sexes_theme <-  my_theme(
-                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
-                  legend.position      = "bottom", 
-                  legend.box.spacing   = unit(0, "lines"),
-                  legend.box.margin    = margin(-5, 0, 0, -85),
-                  legend.margin        = margin(0, 0, 0, 0),
-                  legend.box.just      = "left",
-                  legend.justification = "left",
-                  legend.spacing       = unit(1.5, "lines")
-                )
-sexes_guides <- guides(
-                  # fill  = guide_legend(ncol = 1),
-                  # color = guide_legend(ncol = 1)
-                )
-
-drake::loadd(res_partners)
-res_partners %<>%       
-      filter(grepl(":", term)) %>% 
-      mutate(`N partners` = ifelse(grepl("TRUE:", term), "3 or less", "4 or more"))
-
-n_regs <- as.double(nrow(res_partners))
-res_partners %>% 
-      mutate(
-        score_name = pretty_names(score_name)
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(
-        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        x      = estimate, 
-        fill   = `N partners`, 
-        colour = p < 0.05/{{n_regs}}
-      )) + 
-        my_vline +
-        pval_geom_point() +
-        facet_wrap(vars(sex), scales = "free_y") +
-        my_fill_scale(n = 2) +
-        pval_scale() +
-        sexes_theme +
-        my_labs(x = "", y = "") + 
-        scale_y_reordered() +
-        sexes_guides
-
-drake::loadd(res_with_partner_sex)
-res_with_partner_sex %<>% filter(grepl(":", term))
-
-n_regs <- as.double(nrow(res_with_partner_sex))
-
-res_with_partner_sex %>% 
-      mutate(
-        score_name = pretty_names(score_name),
-        Household = ifelse(grepl("TRUE", term), "With partner", 
-                      "Without partner")
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(
-        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        x      = estimate, 
-        fill   = Household, 
-        colour = p < 0.05/{{n_regs}}
-      )) + 
-        my_vline +
-        pval_geom_point() +
-        facet_wrap(vars(sex), scales = "free_y") +
-        my_fill_scale(n = 2) +
-        pval_scale() +
-        sexes_theme +
-        my_labs(x = "", y = "") + 
-        scale_y_reordered() +
-        sexes_guides
-
-```
-
-```{r calc-corr-age-flb}
-
-drake::loadd(res_age_flb)
-drake::loadd(res_sex)
-
-res_sex_comparison <- res_sex %>% 
-      filter(sex == "Female") %>% 
-      mutate(score_name = term) %>% 
-      select(score_name, term, estimate:conf.high)
-
-res_age_flb %<>% filter(term != "age_flb")
-
-res_combined <- bind_rows(
-        raw      = res_sex_comparison, 
-        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
-        .id      = "reg.type"
-      ) %>% 
-      arrange(score_name)
-
-raw_flb_comparison <- res_combined %>% 
-      pivot_wider(
-        id_cols     = score_name, 
-        names_from  = reg.type, 
-        values_from = estimate
-      )
-
-cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
-opp_signed <- sum(
-        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
-      )
-
-```
-
-To investigate this further, we regress *number of children* on polygenic scores
-*controlling* for AFLB, again among females (Figure \@ref(fig:plot-age-flb)b).
-In `r opp_signed` out of `r length(score_names)` cases, effects change sign when
-controls are added. The correlation between effect sizes controlling for AFLB,
-and raw effect sizes, is `r cor_raw_flb`. Thus, selection effects seem to come
-through two opposing channels: an effect on AFLB, and an opposite-signed effect
-on number of children controlling for AFLB. Note that we do not claim that AFLB
-is exogenous to an individual's polygenic scores -- indeed, our analyses show
-that it is not (Appendix Figure \@ref(fig:plot-age-flb-dv)). Rather, we argue
-that AFLB mediates part of the relationship between polygenic scores and
-fertility, and that once this is controlled for, the remaining part of the
-relationship has the opposite sign.
-
-```{r plot-age-flb, fig.cap = "Selection effects by age at first live birth, and controlling for age at first live birth (women only).", fig.ncol = 1,  fig.subcap = c("Effect sizes by age at first live birth terciles (women only)", "Effect sizes controlling for age at first live birth.
-Effect sizes without controls (for women only) shown for comparison."), fig.align = "center"}
-
-drake::loadd(res_age_flb_cross)
-
-# this counts each cross term as a separate test:
-n_regs <- as.double(nrow(res_age_flb_cross))
-
-res_age_flb_cross %>% 
-      mutate(
-        score_name = pretty_names(score_name),
-        score_name = fct_reorder(score_name, estimate, order_abs(1)),
-        term       = gsub("age_flb_cat(.*):.*", "\\1", term),
-        term       = add_n(term, "age_flb_cat")
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(estimate, score_name, fill = term, 
-        colour = p < 0.05/{{n_regs}})) +
-        my_vline +
-        pval_geom_point() +
-        pval_scale() +
-        my_fill_scale(n = 3, direction = -1, 
-          guide = guide_legend(title = "Age at first live birth")) +
-        my_labs() +
-        my_theme()
-
-n_regs <- as.double(nrow(res_age_flb))
-
-res_combined %>% 
-      mutate(
-        score_name = pretty_names(score_name),
-        score_name = fct_reorder(score_name, estimate, order_abs(2)),
-        Regression = ifelse(reg.type == "raw", "Uncontrolled", "Controlled for AFLB")
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
-        color = p < 0.05/{{n_regs}})) + 
-        my_vline +
-        pval_geom_point() +
-        pval_scale() +
-        my_fill_scale(n = 2) +
-        my_labs() +
-        my_theme()
-```
-
-\FloatBarrier
-## Correcting for sampling bias
-
 ```{r calc-res-weighted}
@@
 Since UK Biobank participants are not representative of the wider population,
-our results suggest that naïve estimates of natural selection are at risk of
+naïve estimates of natural selection are at risk of
 ascertainment bias. To correct for this, we weight participants using population
@@
 `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
-further affected by weighting on other demographic variables. Since the UK Biobank
-sample seems to be ascertained in ways that shrink estimates of selection
-effects, we suspect that our weighted estimates are still conservative.[^purifying]
+further affected by weighting on other demographic variables.[^purifying]
 
@@
 
-```{r plot-res-wt, fig.cap = "Selection effects: weighted regressions. Unweighted estimates are shown for comparison."}
+```{r plot-res-wt, fig.cap = "Selection effects: weighted and unweighted regressions.  Each point represents a single bivariate regression of number of children on a polygenic score."}
 
@@
 
+To understand the reasons for this ascertainment bias, and to learn about the
+mechanisms underlying natural selection, we split the sample, starting with
+basic demographic variables including education, income and sex. These are all
+potential sources of ascertainment bias, as mentioned above.
+
+```{r plot-income-educ-level, fig.cap = "Selection effects by education and income.", fig.subcap = c("Age left full-time education", "Household income"), fig.ncol = 1}
+
+drake::loadd(res_edu)
+n_regs <- as.double(nrow(res_edu))
+
+res_edu %>% 
+      mutate(
+        term = pretty_names(term), 
+        term = fct_reorder(term, estimate, order_abs(1)),
+        "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
+        "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(y = term, x = estimate, colour = p < 0.05/{{n_regs}}, 
+        fill = `Age left FTE`)) +
+        my_vline +
+        pval_geom_point() + 
+        pval_scale() +
+        my_fill_scale(n = 3, direction = -1) +
+        my_labs() +
+        my_theme()
+
+drake::loadd(res_income)
+
+n_regs <- as.double(nrow(res_income))
+res_income %>% 
+      mutate(
+        term = pretty_names(term), 
+        term = fct_reorder(term, estimate, order_abs(1)),
+        Income = factor(income_cat, 
+          labels = c("< £18K", "£18-30K", "£31-51K", "£52-100K", "> £100K")
+        ),
+        Income = add_n(Income, "income_cat")
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(y = term, x = estimate, fill = Income, 
+        colour = p < 0.05/{{n_regs}})) +
+        my_vline +
+        pval_geom_point(alpha = 0.8) + 
+        my_fill_scale(n = 5, direction = -1) +
+        pval_scale() +
+        my_labs() +
+        my_theme()
+```
+
+Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
+polygenic score, grouping respondents by age of completing full-time education,
+and by household income. Effects are larger and more significant for the lowest
+income category, and for the lowest education category. Note that the overall
+effect is not a simple average of the effect among the different subgroups,
+because polygenic scores may also shift respondents between the subgroups. For
+example, a high polygenic score for educational attainment may predict having
+fewer children among early school leavers, but it may also increase the age at
+which a respondent leaves school.
+
+```{r calc-income-edu-controlled}
+
+drake::loadd(res_income_controlled, res_edu_controlled)
+n_regs_ic <- nrow(res_income_controlled)
+n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
+
+n_regs_educ <- nrow(res_edu_controlled)
+n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
+```
+
+These results could be explained by age, if older respondents have lower income
+and are less educated, and also show more natural selection on polygenic
+scores. However, if we rerun the regressions, interacting the polygenic
+score with income category and also with a quadratic in age, the
+interaction with income remains significant at 0.05/`r n_regs_ic` for
+`r n_sig_ic` out of `r n_regs_ic` regressions. Similarly if we interact
+the PGS with age of leaving full time education and a quadratic in age,
+the interaction with age leaving full time education remains significant at
+0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.
+
+Selection effects are also different between men and women (Appendix Figure
+\@ref(fig:plot-sexes)). Differences are particularly large for educational
+attainment, height and MDD. Several polygenic scores for mental illness and
+personality traits are more selected for (or less against) among women,
+including major depressive disorder (MDD), schizophrenia and neuroticism, while
+extraversion is more selected for among men. Scores for waist circumference and
+waist-hip ratio are less selected for among women, and scores for educational
+attainment are more selected against, though the difference is only significant
+for the ("EA3") educational attainment score of @lee2018gene. One
+possible reason for these sex differences is that polygenic scores may affect
+fertility via success in marriage markets, and men and women may value different
+characteristics in these.
+
+```{r calc-age-flb-cross}
+
+drake::loadd(res_age_flb_cross)
+
+cor_age_flb <- res_age_flb_cross %>% 
+      mutate(category = gsub(":.*", "", term)) %>% 
+      pivot_wider(score_name, names_from = category, values_from = estimate) %>%              {cor(.$`age_flb_cat10-22`, .$`age_flb_cat28-52`)}
+
+```
+
+We next focus on variables related to household type and reproductive
+strategy. We split males and females by lifetime number of sexual
+partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
+For both sexes, selection effects are larger and more significant among
+those with more than 3 partners. Next we split respondents by whether
+they were living with a spouse or partner at the time of interview.
+Effects are larger among those not living with a spouse or partner
+(Figure \@ref(fig:plot-n-partners)b).
+
+Lastly, we split female respondents by age at first live birth
+(AFLB).[^aflb-no-men] There is evidence for genetic effects on AFLB
+[@Barban_2016], and there is a close link between this variable and number of
+children born. Figure \@ref(fig:plot-age-flb)a shows effect sizes estimated
+separately for each tercile of AFLB. Several effects are strikingly different
+across terciles. ADHD and MDD are selected for amongst the youngest third of
+mothers, but selected against among the oldest two-thirds. Educational
+attainment is selected for among the oldest two-thirds of mothers, but is not
+significantly selected among the youngest third. Similarly, several polygenic
+scores for body measurements are selected against only among older mothers. The
+correlation between effect sizes for the youngest and oldest terciles is 
+`r cor_age_flb`.
+
+[^aflb-no-men]: This information is unavailable for men.
+
+```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and household type.", fig.subcap = c("Lifetime number of sexual partners", "Household type"), fig.ncol = 1, fig.align = "center", fig.height = 4}
+
+sexes_theme <-  my_theme(
+                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
+                  legend.position      = "bottom", 
+                  legend.box.spacing   = unit(0, "lines"),
+                  legend.box.margin    = margin(-5, 0, 0, -85),
+                  legend.margin        = margin(0, 0, 0, 0),
+                  legend.box.just      = "left",
+                  legend.justification = "left",
+                  legend.spacing       = unit(1.5, "lines")
+                )
+sexes_guides <- guides(
+                  # fill  = guide_legend(ncol = 1),
+                  # color = guide_legend(ncol = 1)
+                )
+
+drake::loadd(res_partners)
+res_partners %<>%       
+      filter(grepl(":", term)) %>% 
+      mutate(`N partners` = ifelse(grepl("TRUE:", term), "3 or less", "4 or more"))
+
+n_regs <- as.double(nrow(res_partners))
+res_partners %>% 
+      mutate(
+        score_name = pretty_names(score_name)
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(
+        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        x      = estimate, 
+        fill   = `N partners`, 
+        colour = p < 0.05/{{n_regs}}
+      )) + 
+        my_vline +
+        pval_geom_point() +
+        facet_wrap(vars(sex), scales = "free_y") +
+        my_fill_scale(n = 2) +
+        pval_scale() +
+        sexes_theme +
+        my_labs(x = "", y = "") + 
+        scale_y_reordered() +
+        sexes_guides
+
+drake::loadd(res_with_partner_sex)
+res_with_partner_sex %<>% filter(grepl(":", term))
+
+n_regs <- as.double(nrow(res_with_partner_sex))
+
+res_with_partner_sex %>% 
+      mutate(
+        score_name = pretty_names(score_name),
+        Household = ifelse(grepl("TRUE", term), "With partner", 
+                      "Without partner")
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(
+        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        x      = estimate, 
+        fill   = Household, 
+        colour = p < 0.05/{{n_regs}}
+      )) + 
+        my_vline +
+        pval_geom_point() +
+        facet_wrap(vars(sex), scales = "free_y") +
+        my_fill_scale(n = 2) +
+        pval_scale() +
+        sexes_theme +
+        my_labs(x = "", y = "") + 
+        scale_y_reordered() +
+        sexes_guides
+
+```
+
+```{r calc-corr-age-flb}
+
+drake::loadd(res_age_flb)
+drake::loadd(res_sex)
+
+res_sex_comparison <- res_sex %>% 
+      filter(sex == "Female") %>% 
+      mutate(score_name = term) %>% 
+      select(score_name, term, estimate:conf.high)
+
+res_age_flb %<>% filter(term != "age_flb")
+
+res_combined <- bind_rows(
+        raw      = res_sex_comparison, 
+        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
+        .id      = "reg.type"
+      ) %>% 
+      arrange(score_name)
+
+raw_flb_comparison <- res_combined %>% 
+      pivot_wider(
+        id_cols     = score_name, 
+        names_from  = reg.type, 
+        values_from = estimate
+      )
+
+cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
+opp_signed <- sum(
+        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
+      )
+
+```
+
+To investigate this further, we regress *number of children* on polygenic scores
+*controlling* for AFLB, again among females (Figure \@ref(fig:plot-age-flb)b).
+In `r opp_signed` out of `r length(score_names)` cases, effects change sign when
+controls are added. The correlation between effect sizes controlling for AFLB,
+and raw effect sizes, is `r cor_raw_flb`. Thus, selection effects seem to come
+through two opposing channels: an effect on AFLB, and an opposite-signed effect
+on number of children controlling for AFLB. Note that we do not claim that AFLB
+is exogenous to an individual's polygenic scores -- indeed, our analyses show
+that it is not (Appendix Figure \@ref(fig:plot-age-flb-dv)). Rather, we argue
+that AFLB mediates part of the relationship between polygenic scores and
+fertility, and that once this is controlled for, the remaining part of the
+relationship has the opposite sign.
+
+```{r plot-age-flb, fig.cap = "Selection effects by age at first live birth, and controlling for age at first live birth (women only).", fig.ncol = 1,  fig.subcap = c("Effect sizes by age at first live birth terciles (women only)", "Effect sizes controlling for age at first live birth.
+Effect sizes without controls (for women only) shown for comparison."), fig.align = "center"}
+
+drake::loadd(res_age_flb_cross)
+
+# this counts each cross term as a separate test:
+n_regs <- as.double(nrow(res_age_flb_cross))
+
+res_age_flb_cross %>% 
+      mutate(
+        score_name = pretty_names(score_name),
+        score_name = fct_reorder(score_name, estimate, order_abs(1)),
+        term       = gsub("age_flb_cat(.*):.*", "\\1", term),
+        term       = add_n(term, "age_flb_cat")
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(estimate, score_name, fill = term, 
+        colour = p < 0.05/{{n_regs}})) +
+        my_vline +
+        pval_geom_point() +
+        pval_scale() +
+        my_fill_scale(n = 3, direction = -1, 
+          guide = guide_legend(title = "Age at first live birth")) +
+        my_labs() +
+        my_theme()
+
+n_regs <- as.double(nrow(res_age_flb))
+
+res_combined %>% 
+      mutate(
+        score_name = pretty_names(score_name),
+        score_name = fct_reorder(score_name, estimate, order_abs(2)),
+        Regression = ifelse(reg.type == "raw", "Uncontrolled", "Controlled for AFLB")
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
+        color = p < 0.05/{{n_regs}})) + 
+        my_vline +
+        pval_geom_point() +
+        pval_scale() +
+        my_fill_scale(n = 2) +
+        my_labs() +
+        my_theme()
+```
+
+\FloatBarrier
+
 ## Selection in the parents' generation
@@
 
-An alternative theory is that polygenic scores correlate with the
-motivation to have children (parameter $a$ in the model). However, this theory
-would not explain why our results are weaker at higher incomes and education levels.
-Indeed, in the one-period model, while $\frac{dN^{*}}{da}>0$, this
-relationship actually becomes stronger at higher wages.
-
-The above does not provide a complete theory of fertility. Rather,
-it shows that a relatively simple economic model can explain many
-of our results. In the appendix, we discuss other models of fertility
-from the economic literature, as well as causal evidence for the relationship
-between human capital, income and fertility.
-
-This theory can explain why selection effects have opposite signs when AFLB is
-controlled for. Suppose that people are aware of their endowments of human
-capital, i.e. skills and characteristics that are valuable in the labour market.
-Suppose also that education and innate human capital are complements.
-Individuals choose how long to stay in education; after leaving education they
-enter the labour market and form families. If so, then people with less human
-capital will leave education earlier and have their first child earlier [cf.
-@caucutt2002women; @monstad2008education]. This is a pure substitution effect:
-the income effect plays no role in the decision to leave education, since people
-are not yet earning wages.[^no-borrow] Whilst in the labour market, higher
-earners will have more children (i.e. income effects dominate). These
-assumptions will lead to the pattern we observe: earnings-increasing variants
-will predict later first births (and fewer children overall), but they will also
-predict more children within any given age at first live birth.
-
-[^no-borrow]: This argument assumes that capital market imperfections hinder
-people from borrowing against future income.
-
-The theory can also explain why the natural selection pressure is weaker among
-higher-income parents. @becker1976child show that if parents care about child
-quality as well as quantity, under certain assumptions the income effect will be
-stronger at higher income levels; this can lead to a U-shaped relationship
-between income and fertility. Empirically, @cohen2013financial find that income
-decreases fertility at low income levels but increases it at higher income
-levels. If, in our data, at high income levels income and substitution effects
-are roughly balanced , then income-linked polygenic scores will be neither
-selected for nor against.
-
-Lastly, it is often assumed that the substitution effect will dominate for lone
-parents, or those in unstable relationships, since they have less opportunity to
-share childcare responsibilities, while the income effect will dominate for
-couples who are able to reap the gains from specialization [@becker1981treatise]. Indeed, US fertility
-decreases faster with education among single mothers than married mothers
-[@baudin2015fertility]. This can explain why earnings-increasing genetic variants 
-decrease fertility more among single parents than among couples. The same logic 
-could explain our results for lifetime number of sexual partners, if this is 
-associated with relationship stability.
+```{r, eval = FALSE}
+"Other models of the literature
+
+## LITERATURE NOTES
+
+* Lundberg and Pollak JEP 2007:
+    - several models where nonmarital fertility is more likely for poorer women
+    - Burdett and Ermisch 2002, Willis 1999, Neal 2004
+    - Unmarried parents may not be able to credibly commit to each other
+
+* Happel et al. 1984
+   - childbirth is chosen to minimize loss of earnings from interrupted work.
+   - early childbirth if your skills decay a lot  while childrearing. (No point
+     gaining them then losing them.) Late childbirth otherwise.
+   - late childbirth if you have acquired skills before marriage.
+   - Capital-market imperfections make it wise to delay childbirth (till man has
+     high income to cope while wife is at home). More so for men's wages which
+     rise faster.
+   - empirically: (among married couples) 
+     - high-skill occupation women have children later. Higher-earning men have 
+     them earlier.
+
+- Caucutt et al. 2001
+  - high waged and high educated women (of recent cohorts) more likely to
+    delay childbirth
+
+- Cohen et al. 2013
+  - analyse effect of Israeli subsidy
+  - along the way, find that income affects fertility negatively at low income,
+    positively at high income, supporting Becker and Tomes 76
+
+- Becker and Tomes 1976
+  - investment in children quality and quantity when quality must be equal
+    across children; and some quality comes from an innate endowment.
+  - "observed" quantity-of-children income elasticity rises as income rises
+  - while "observed" quality-of-children elasticity falls as income rises, 
+    b/c more of total quality must be paid for by parents themselves. 
+  - the need to pay more for quality results in lower quantity, since total cost is 
+    quantity X quality
+  - something fishy about argument somehow...
+
+- Lindo 2008
+  - men who lose a job have children sooner, but fewer children in total
+  - controls for all skills effects
+  - so overall effect of income is positive, but timing changes
+
+- Monstad et al. 2008
+  - using an education reform to vary education
+  - no evidence of a causal effect on fertility, but births start later
+  * maybe the usual correlation reflects underlying skills. Your data could
+    test this.
+
+- Autor et al. 2017
+  - the China shock (exogenous shock to income) 
+  - they use Chinese import penetration by industry (instrumented using
+    other countries) times industry employment shares; and separate it for
+    males and females
+  - shocks to male employment decreased fertility and marriage; 
+    increased unmarried mothers
+  - shocks to female employment *increase* marriage and motherhood
+  * this roughly seems to be the consensus
+  * Whereas, you find that male and female *skills* (as 'instrumented' by PGS)
+    have the same effect on fertility - i.e. less skills = higher fertility
+
+- Baudin et al. 2015
+  * look at childlessness vs. fertility (extensive/intensive margin)
+  * some people are too poor to afford children
+  * 'a husband alleviates the time cost of having children'. But for some groups (see appendix)
+    there is a +ve income/fertility gradient
+  * fertility decreases with female educn due to subst effect
+  * baby boom and bust in US: a decline in childlessness caused by poverty, then an increase
+    caused by substn effect ('opportunity cost')
+  * fertility decreases with educn both for single mothers and married women
+    - slightly (25%) higher decrease per year of educn (Table 2) for single mothers
+
+```
+
+The above does not provide a complete theory of fertility. Rather, it shows that
+a relatively simple economic model can explain many of our results. Other
+empirical work in economics also supports our mechanism.
+@caucutt2002women and @monstad2008education show that education and skills
+affect age at first birth and fertility. Income decreases fertility at low
+income levels, but increases it at higher income levels [@cohen2013financial].
+US fertility decreases faster with education among single mothers than married
+mothers [@baudin2015fertility], in line with our fact 3 and as predicted by
+@becker1981treatise.
 
@@
             filter(term == 'f.2040.0.0') %>% 
-            {sum(.$p.value <= 0.05/33)}
+            {sum(.$p.value <= .Machine$double.eps)}
 
@@
 
-An alternative theory is that traits selected for are linked to externalizing
-behaviour, risk-seeking and low time discount rates, via the channel of early
-sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
-for this: scores which might plausibly be linked to externalizing behaviour,
-like ADHD and (younger) age at smoking initiation, are selected for. However,
-this theory is less good at explaining variation in selection across the full
-range of scores, including physical measures, e.g. waist-hip ratio and BMI.
-Externalizing behaviour also does not explain why selection effects should work
-in the opposite direction among older mothers, whereas the economic theory can
-explain this via income effects, as described above. We test the alternative
-theory directly by re-estimating equation \@ref(eq:regression) controlling for a
-measure of risk attitude (UK Biobank field 2040). While risk attitude is always
-a highly significant predictor of number of children, it has little impact on
-the effect size or significance of the polygenic scores. The median ratio of
-effect sizes between regressions with and without controls is `r median(ratio)`;
-all scores which are significant at $p \le 0.05/33$ in uncontrolled regressions
-remain so when controlling for risk attitude. Thus, although risk attitude does
-predict number of children, it appears not to mediate selection effects.
-Overall, we believe that the economic theory is the most likely explanation.
+An alternative theory is that polygenic scores correlate with the motivation to
+have children (parameter $a$ in the model; @jones2008fertility). However, this
+theory would not explain why our results are weaker at higher incomes and
+education levels. Indeed, in the one-period model, while $\frac{dN^{*}}{da}>0$,
+this relationship actually becomes stronger at higher wages.
+
+Another alternative is that traits under selection are linked to externalizing
+behaviour and risk-seeking. This might be captured by our risk-aversion 
+parameter $\sigma$, which affects fertility by changing the marginal utility of income, but a more direct channel is risky sexual behaviour
+[@Mills2020.05.06.081273]. The data here provide some support for this: scores
+which might plausibly be linked to externalizing behaviour, like ADHD and
+(younger) age at smoking initiation, are selected for. However, this theory is
+less good at explaining variation in selection across the full range of scores,
+including physical measures such as waist-hip ratio and BMI. We test this theory
+directly by re-estimating equation \@ref(eq:regression) controlling for a
+measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
+sizes between regressions with and without controls is `r median(ratio)`; all
+scores which are significant at $p \le 0.05/33$ in uncontrolled regressions
+remain so when controlling for risk attitude. This non-result could simply
+reflect the imprecision of the risk aversion measure, which is a single yes/no
+question. However, the measure *does* predict the overall number of children,
+highly significantly ($p \le 2^{-16}$, `r n_risk_sig`/33 regressions). Given that,
+and our large N, we believe that the non-result is real: while risk aversion does
+predict fertility in the sample, it is not an important channel for natural 
+selection.
 
@@
 education are being selected against (selected for). In addition, many of the
-phenotypes under positive selection are linked to disease risk, or are what many
-people would see as undesirable to have. For example, most people would probably
-prefer to have high educational attainment, a low risk of ADHD and major
-depressive disorder, and a low risk of coronary artery disease, but natural
-selection is pushing against genes associated with these traits. Potentially,
-this could increase the health burden on modern populations, but that depends on
-effect sizes. Our results suggest that naïve estimates can be affected by sample
-ascertainment bias. This problem may be less serious in surveys which aim to be
-representative (as the UK Biobank does not). However, there is still scope for
-bias, since not all respondents consent to the collection of genetic data. For
-instance, completion rates for genotype data in the US Health and Retirement
-Study were around 80-85% [@hrs2020]. Researchers should be aware of the risks of
-ascertainment when studying modern natural selection.
+phenotypes under positive selection are linked to disease risk. Most people
+would probably prefer to have high educational attainment, a low risk of ADHD
+and major depressive disorder, and a low risk of coronary artery disease, but
+natural selection is pushing against genes associated with these traits.
+Potentially, this could increase the health burden on modern populations, but
+that depends on effect sizes. Our results suggest that naïve estimates can be
+affected by sample ascertainment bias. This problem may be less serious in
+surveys which aim to be representative (as the UK Biobank does not). However,
+there is still scope for bias, since not all respondents consent to the
+collection of genetic data. For instance, completion rates for genotype data in
+the US Health and Retirement Study were around 80-85% [@hrs2020]. Researchers
+should be aware of the risks of ascertainment when studying modern natural
+selection.
 
@@
 
+```{r calc-inequality}
+drake::loadd(res_ineq)
+
+ineq_comp <-  left_join(
+                res_ineq %>% filter(income_cat == 1),
+                res_ineq %>% filter(income_cat == 5),
+                by = "score"
+              )
+
+ineq_comp %<>% mutate(
+                 diff_children = wtd.y - wtd.x,
+                 diff_hypoth   = unw.y - unw.x,
+                 pct_change    = 100 * (diff_children - diff_hypoth)/diff_hypoth
+               )
+
+median_pct_change <- median(ineq_comp$pct_change)
+n_increases <- sum(ineq_comp$pct_change > 0)
+
+```
+
 Because selection effects are concentrated in lower-income groups, they may also
@@
 educational attainment, which lowers the mean, is stronger at the bottom of the
-income distribution, and this increases the differences between groups.
-Overall, natural selection increases inequality for 29 out of 33 polygenic
-scores (Appendix Table \@ref(tab:tbl-inequality-effects)). Since many polygenic
-scores are predictive of disease risk, this increase in inequality could
-potentially increase health inequalities. In general, the evolutionary history of
+income distribution, and this increases the differences between groups. Overall,
+natural selection increases inequality for `r n_increases` out of 33 polygenic
+scores, with a median increase of `r median_pct_change`% in the difference
+between highest and lowest income groups (Appendix Table
+\@ref(tab:tbl-inequality-effects)). These are substantial effects, and since
+many polygenic scores are predictive of disease risk, they could potentially
+increase health inequalities. In general, the evolutionary history of
 anatomically modern humans is related to disease risk [@benton2021influence];
@@
 Existing evidence on human natural selection has led some to "biocosmic
-pessimism", including speculation that the Fermi paradox may be resolved by
-dysgenic selection among alien species [@sarraf2019modernity]. Others are more
-sanguine, and argue that natural selection's effects are outweighed by
-environmental improvements, such as those underlying (e.g.) the Flynn effect
-[@flynn1987massive]. The evidence here may add some nuance to this debate.
-Patterns of natural selection have been relatively consistent across the past two
-generations, but they are not the outcome of a universal, society-wide phenomenon.
-Instead they result from a sum of opposing forces, operating in different
-parts of society and pulling in different directions.
+pessimism" [@sarraf2019modernity]. Others are more sanguine, and argue that
+natural selection's effects are outweighed by environmental improvements, such
+as those underlying the Flynn effect [@flynn1987massive]. The evidence
+here may add some nuance to this debate. Patterns of natural selection have been
+relatively consistent across the past two generations, but they are not the
+outcome of a universal, society-wide phenomenon. Instead they result from a sum
+of opposing forces, operating in different parts of society and pulling in
+different directions.
 
@@
 
-drake::loadd(res_ineq)
-
-ineq_comp <-  left_join(
-                res_ineq %>% filter(income_cat == 1),
-                res_ineq %>% filter(income_cat == 5),
-                by = "score"
-              )
-
-ineq_comp %<>% mutate(
-                 diff_children = wtd.y - wtd.x,
-                 diff_hypoth   = unw.y - unw.x,
-                 pct_change    = 100 * (diff_children - diff_hypoth)/diff_hypoth
-               )
-
 ineq_comp %>% 

### 2021-04-14 model-appendix.Rmd
responses to PNAS review
Address comments in PNAS review - new title - Use standard_ggplot more - Add confidence intervals - Main plot to start - Discuss endogeneity - Add computation of CARA utility

@@
-\externaldocument{why-natural-selection}
-
-\section*{Appendix}
-
 \section*{Solution for the one-period model}
@@
 \end{align}
-Plugging the expressions for $N_{1}^{*}$ and $N_{2}^{*}$ into $s^{*}$gives
+Plugging the expressions for $N_{1}^{*}$ and $N_{2}^{*}$ into $s^{*}$ gives
 \[
@@
 
-Conditions (\ref{eq:kkt1}) and (\ref{eq:kkt2}) give the bounds for
-this case. When $\sigma<1$, (\ref{eq:kkt1}) puts a minimum on $h$
-and (\ref{eq:kkt2}) puts a maximum on $h$. The situation is reversed
-for $\sigma>1$.
-
 To find how $s^{*}$ changes with $h$, note that the right hand side
@@
 i.e. $N_{2}^{*}$ is convex in $h$.
+
+Conditions (\ref{eq:kkt1}) and (\ref{eq:kkt2}) give the bounds for
+this case. When $\sigma<1$, (\ref{eq:kkt1}) puts a minimum on $h$
+and (\ref{eq:kkt2}) puts a maximum on $h$. The situation is reversed
+for $\sigma>1$.
 
@@
 
+\begin{lem}
+For $\sigma < 1$ in a neighbourhood of 1, $N^*$ is globally convex in $h$.
+\end{lem}
+\begin{proof}
 From Lemmas \ref{case-1-lemma}, \ref{case-2-lemma} and \ref{case-3-lemma},
@@
 for $\sigma$ close enough to 1, $N_{2}^{*}$ is convex in $h$ when
-$N_{1}^{*}=0$. To show global concavity of $N^{*}$ in $h$ as $\sigma\nearrow1$,
-all that remains is to check that the derivative is increasing around
+$N_{1}^{*}=0$. All that remains is to check that the derivative is increasing around
 the points where these 3 regions meet. That is trivially satisfied
@@
 \]
+\end{proof}
 
-Thus, as $\sigma\nearrow1$, $N^{*}$ is guaranteed to be globally
-convex in $h$.
-
-We can now prove the facts stated in the main text.
+We can now gather the facts stated in the main text.
 

### 2021-04-14 why-natural-selection.Rmd
Address comments in PNAS review - new title - Use standard_ggplot more - Add confidence intervals - Main plot to start - Discuss endogeneity - Add computation of CARA utility

@@
 ---
-title: "Natural selection in contemporary humans is driven by correlations with human capital"
+title: "Human capital drives natural selection in contemporary humans"
 author: "David Hugh-Jones \\thanks{Corresponding author.
@@
   - \providecommand{\lemmaname}{Lemma}
-  - \usepackage{xr-hyper}
 editor_options: 
@@
 theme_set(theme_minimal())
+
 my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dotted")
@@
         order_idx = 1, 
-        n_cats    = NULL
+        fill_direction = 1,
+        n_cats    = NULL,
+        conf_int  = NULL
       ) {
   if (! missing(score_col)) score_col <- enquo(score_col)
+
   mfc <- missing(fill_col)
@@
 
+  if (is.null(conf_int)) conf_int <- "conf.low" %in% names(dfr)
+
   n_regs <- as.double(n_regs)
+
+  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL
 
@@
         rename(p = p.value) %>% 
-        ggplot(aes(estimate, {{score_col}}, fill = {{fill_col}}, ...,
-          colour = p < 0.05/{{n_regs}})) +
-          pval_geom_point() +
+        ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
+                     fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
           my_vline + 
+          confint_geom_segment +
+          pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
           pval_scale() +
-          my_fill_scale(n = n_cats) +
+          my_fill_scale(n = n_cats, direction = fill_direction) +
           my_labs() +
-          my_theme()
+          my_theme() +
+          scale_linetype_manual(name = "", values = c("95% c.i." = 1))
+
 }
@@
 
-pval_geom_point <- function (size = 1.5, stroke = 0.5, ...) {
+pval_geom_point <- function (size = 1.5, stroke = 0.5, fill = NULL, ...) {
+  if (is.null(fill)) {
     geom_point(shape = 21, size = size, stroke = stroke, ...)
+  } else {
+    geom_point(shape = 21, size = size, stroke = stroke, fill = fill, ...)
   }
+}
+
+confint_geom_segment <- function () {
+  geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i."), 
+                    color = "grey60", alpha = 0.25, size = 1.1, 
+                    lineend = "round")
+}
 
 my_fill_scale <- function (n, aesthetics = "fill", ...) {
-  if (n == 2) {
-    scale_colour_manual(values = c("steelblue4", "#D63C3C"), 
-          aesthetics = aesthetics, ...)
-    # scale_color_brewer(type = "qual", palette = 3, aesthetics = aesthetics, ...)
+
+  fill_colors <- c("steelblue4", "#D63C3C")
+
+  if (n <= 2) {
+    dots <- list(...)
+    dots$direction <- NULL
+    do.call(scale_colour_manual, c(list(values = fill_colors[1:n],
+                                        aesthetics = aesthetics), 
+                                   dots))
   } else {
@@
 "
+
+# TODO response to PNAS
+
+* Add CIs - done.
+* Clarify collider bias/endogeneity - done.
+* Formalize theory - done.
+  - Comment on results where effects are actually reversed. Could this
+  happen if \sigma were higher in period 2? Or if it increased with income?
+  Could maybe do a computational example.
+  - One possibility is CARA utility. Solving it computationally
+* Clarify causality?
+  - Helps to deal with the "collider bias" argument of reviewer
+  - Might also be a better test of mediation via human capital
+  - One other possibility. If we use only people with first profession
+    teacher, that (as it happens) shuts down the EA3-income correlation.
+    It also shuts down the EA3-fertility relationship, suggesting that
+    income is indeed the mediator.
+* Control for PGS for fertility
+* Clarify "what we learn about genetic links to fertility by using economic theory to interpret relationships between polygenic scores and fertility."
+  - Point is that we can explain how natural selection is occurring.
+* "Discuss empirical predictions for AFLB grounded in dynamic theory"
+  - hmm, tricky, maybe we can include it in discussion of economic results at
+  end of formal section.
+* Why don't women see stronger effects? But note, they do: not significant
+  after correction for multiple testing, but a 25% difference.
+
 # Open questions/TODO
@@
 To test directly for natural selection, we regress respondents' number of
-children ($y_i$) on their polygenic scores (PGS):
+children ($N_i$) on their polygenic scores (PGS):
 
@@
 \begin{equation}
-y_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
+N_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
 \label{eq:regression}
@@
 
-[^normalization]: The selection effect $\beta$ equals $Cov(Y, PGS)/Var(PGS)$
-where $Y$ is the number of children. Since PGS are normalized to variance 1,
-this reduces to to $Cov(Y, PGS) \equiv E(Y PGS) - E(Y) E(PGS)$, which in turn
-reduces to $E(Y PGS)$ as $E(PGS) = 0$. This is the polygenic score weighted by
-the number of children, which is the average polygenic score in the next
-generation.
+[^normalization]: The selection effect $\beta$ equals $Cov(N, PGS)/Var(PGS)$
+where $N$ is the number of children. Since PGS are normalized to variance 1 and
+mean 0, this reduces to 
+$Cov(N, PGS) \equiv E(N\ PGS) - E(N) E(PGS) = E(N\ PGS)$. This is the polygenic 
+score weighted by the number of children, which is the average polygenic score 
+in the next generation.
 
@@
 `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
-further affected by weighting on other demographic variables.[^purifying]
-
-[^purifying]: We also check for purifying selection by estimating \@ref(eq:regression) with a quadratic term. See Appendix Figure \@ref(fig:plot-purifying).
+further affected by weighting on other demographic variables.
+
+We also check for purifying and diversifying selection by 
+estimating \@ref(eq:regression) with a quadratic term. See Appendix Figure \@ref(fig:plot-purifying). In particular, we find diversifying selection for
+educational attainment polygenic scores: at higher values of these scores, 
+the effect on fertility is closer to zero.
 
@@
 basic demographic variables including education, income and sex. These are all
-potential sources of ascertainment bias, as mentioned above.
+potential sources of ascertainment bias, as mentioned above. We then re-estimate
+\@ref(eq:regression) within each subgroup.
+
+We emphasize two points. First, this exercise does
+*not* apportion the total correlation with fertility between subgroups. 
+That is because the polygenic score may also predict the probability of being in 
+each subgroup. For example, the EA3 polygenic score for educational attainment not only affects
+fertility among university graduates, and among non-graduates; it also changes
+a person's chance of attending university. Second, we do *not* claim that our
+observed correlations reflect causal effects of polygenic scores on fertility.
+Indeed, polygenic scores correlate with many sources of environmental variation 
+that could affect fertility. However, natural selection is a correlational
+phenomenon, not a causal one. That is, if a polygenic score correlates with 
+low fertility, it is being selected against, whatever the underlying causal
+relationship. Differences in equation \@ref(eq:regression) across subgroups can 
+be informative about how natural selection works, because they constrain the set 
+of possible explanations to those that generate these differences.
 
@@
       mutate(
-        term = pretty_names(term), 
-        term = fct_reorder(term, estimate, order_abs(1)),
         "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
@@
       ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(y = term, x = estimate, colour = p < 0.05/{{n_regs}}, 
-        fill = `Age left FTE`)) +
-        my_vline +
-        pval_geom_point() + 
-        pval_scale() +
-        my_fill_scale(n = 3, direction = -1) +
-        my_labs() +
-        my_theme()
+      standard_ggplot(n_regs = n_regs, fill_col = `Age left FTE`, 
+                        fill_direction = -1)
 
@@
       mutate(
-        term = pretty_names(term), 
-        term = fct_reorder(term, estimate, order_abs(1)),
         Income = factor(income_cat, 
@@
       ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(y = term, x = estimate, fill = Income, 
-        colour = p < 0.05/{{n_regs}})) +
-        my_vline +
-        pval_geom_point(alpha = 0.8) + 
-        my_fill_scale(n = 5, direction = -1) +
-        pval_scale() +
-        my_labs() +
-        my_theme()
+      standard_ggplot(n_regs = n_regs, fill_col = `Income`, 
+                        fill_direction = -1)
 ```
@@
 and by household income. Effects are larger and more significant for the lowest
-income category, and for the lowest education category. Note that the overall
-effect is not a simple average of the effect among the different subgroups,
-because polygenic scores may also shift respondents between the subgroups. For
-example, a high polygenic score for educational attainment may predict having
-fewer children among early school leavers, but it may also increase the age at
-which a respondent leaves school.
+income category, and for the lowest education category.
 
@@
 and are less educated, and also show more natural selection on polygenic
-scores. However, if we rerun the regressions, interacting the polygenic
+scores. However, when we rerun the regressions, interacting the polygenic
 score with income category and also with a quadratic in age, the
@@
 extraversion is more selected for among men. Scores for waist circumference and
-waist-hip ratio are less selected for among women, and scores for educational
-attainment are more selected against, though the difference is only significant
-for the ("EA3") educational attainment score of @lee2018gene. One
+waist-hip ratio are less selected for among women. One
 possible reason for these sex differences is that polygenic scores may affect
@@
                   legend.justification = "left",
-                  legend.spacing       = unit(1.5, "lines")
+                  legend.spacing       = unit(1, "lines")
                 )
@@
         y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
         x      = estimate, 
@@
         my_vline +
+        confint_geom_segment() +
         pval_geom_point() +
@@
         scale_y_reordered() +
+        scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
         sexes_guides
@@
       mutate(
-        score_name = pretty_names(score_name),
-        score_name = fct_reorder(score_name, estimate, order_abs(1)),
-        term       = gsub("age_flb_cat(.*):.*", "\\1", term),
-        term       = add_n(term, "age_flb_cat")
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(estimate, score_name, fill = term, 
-        colour = p < 0.05/{{n_regs}})) +
-        my_vline +
-        pval_geom_point() +
-        pval_scale() +
-        my_fill_scale(n = 3, direction = -1, 
-          guide = guide_legend(title = "Age at first live birth")) +
-        my_labs() +
-        my_theme()
+        `Age at first live birth` = gsub("age_flb_cat(.*):.*", "\\1", term),
+        `Age at first live birth` = add_n(`Age at first live birth`, "age_flb_cat")
+      ) %>% 
+      standard_ggplot(score_col = score_name, 
+                      fill_col = `Age at first live birth`, n_regs = n_regs,
+                      fill_direction = -1)
 
@@
       ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(y = score_name, x = estimate, fill = Regression, 
-        color = p < 0.05/{{n_regs}})) + 
-        my_vline +
-        pval_geom_point() +
-        pval_scale() +
-        my_fill_scale(n = 2) +
-        my_labs() +
-        my_theme()
+      standard_ggplot(score_col = score_name, 
+                      fill_col = Regression, n_regs = n_regs)
+
 ```
@@
 *number of children* is positive ($\rho$ = `r sib_chn_cor$estimate`, 
-$p < 2 \times 10^{-17}$).
-
-## Economic fertility theory and natural selection
+$p < 2 \times 10^{-16}$).
+
+## Human capital and natural selection
 
@@
 
-```{r, eval = FALSE}
+Our results are actually stronger than this, in that correlations with
+fertility are reversed at higher AFLB. This could occur if relative risk 
+aversion increased at higher income levels.
+
+```{r literature, eval = FALSE}
 "Other models of the literature
@@
 reflect the imprecision of the risk aversion measure, which is a single yes/no
-question. However, the measure *does* predict the overall number of children,
-highly significantly ($p \le 2^{-16}$, `r n_risk_sig`/33 regressions). Given that,
-and our large N, we believe that the non-result is real: while risk aversion does
-predict fertility in the sample, it is not an important channel for natural 
-selection.
+question. However, the risk aversion measure *does* predict the overall number
+of children, highly significantly ($p \le 2 \times 10^{-16}$, `r n_risk_sig`/33
+regressions). Given that, and the statistical power we get from our large N, we 
+believe that the non-result is real: while risk aversion does predict fertility 
+in the sample, it is not an important channel for natural selection.
 
@@
 relatively consistent across the past two generations, but they are not the
-outcome of a universal, society-wide phenomenon. Instead they result from a sum
-of opposing forces, operating in different parts of society and pulling in
+outcome of a universal, society-wide phenomenon. Instead they result from  
+opposing forces, operating in different parts of society and pulling in
 different directions.
@@
 
-```{r plot-sexes, fig.cap = "Selection effects by sex. Highlighted lines are significant differences. Highlighted points are significantly different from 0."}
+```{r plot-sexes, fig.cap = "Selection effects by sex. Highlighted lines are significant differences at p < 0.05/33. Highlighted points are significantly different from 0 at p < 0.05/66."}
 
@@
       ) 
+n_comps <-as.double(nrow(res_sex_wide))
 
@@
             xmax   = max_est, 
-            colour = diff.p.value < 0.05/{{n_regs}}
+            colour = diff.p.value < 0.05/{{n_comps}}
           ), 
@@
         pval_geom_point() +
-        pval_scale() +
+        pval_scale(name = "p < 0.05 (corrected)") +
         my_fill_scale(n = 2) +
@@
     mutate(
-      term    = pretty_names(term),
-      term    = fct_reorder(term, estimate, order_abs(2)),
       Control = ifelse(control == "fath_age_birth", 
@@
     ) %>% 
-    rename(p = p.value) %>% 
-    ggplot(aes(y = term, x = estimate, fill = Control, 
-      colour = p < 0.05/{{n_regs}})) + 
-      pval_geom_point(alpha = 0.8) +  
-      my_vline +
-      my_fill_scale(2) +
-      pval_scale() +
-      my_labs() +
-      my_theme()
+    standard_ggplot(fill_col = Control, n_regs = n_regs, order_idx = 2)
 
@@
 
-res_age_flb_dv %>% 
-      mutate(
-        term = pretty_names(term),
-        term = fct_reorder(term, estimate, order_abs(1))
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(estimate, term, colour = p < 0.05 / {{n_regs}})) + 
-      my_vline +
-      pval_geom_point(fill = "steelblue4") +
-      pval_scale() + 
-      my_labs() +
-      my_theme()
+res_age_flb_dv %>% standard_ggplot(n_regs = n_regs)
 
@@
       mutate(
-        term = pretty_names(term),
-        term = fct_reorder(term, estimate, order_abs(1)),
         "Dependent variable" = dplyr::recode(dep.var, 
@@
       ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(estimate, term, fill = `Dependent variable`, 
-        colour = p < 0.05 / {{n_regs}})) + 
-        my_vline +
-        pval_geom_point() +
-        pval_scale() +
-        my_fill_scale(n = 2) +
-        my_labs() +
-        my_theme()
+      standard_ggplot(fill_col = `Dependent variable`, n_regs = n_regs)
 
@@
 ```{r plot-res-earn-educ, fig.cap = "Selection effects controlling for earnings in first job (estimated by mean earnings from ASHE 2007 for the SOC2000 job code) and education (left education before 16, 16-18, or after 18).
-Raw effects are shown for comparison.
-Lines are 95% confidence intervals uncorrected for multiple testing."}
+Raw effects are shown for comparison."}
 
@@
 
-standard_ggplot(res_compare, fill_col = Regression, n_regs = 33) +
-  geom_linerange(
-    aes(
-      xmin = estimate - 2 * std.error, 
-      xmax = estimate + 2 * std.error
-    ), 
-    alpha = .5
-  )
+standard_ggplot(res_compare, fill_col = Regression, n_regs = 33)
 

### 2021-04-14 model-julia.jl

@@ -13,7 +13,7 @@ model = Model(Ipopt.Optimizer)
 set_optimizer_attribute(model, "print_level", 0)
 @variable(model, 0 <= N₁)
 @variable(model, 0 <= N₂)
-@variable(model, 0 <= s)
+@variable(model, 0 <= s <= 1)
 @NLparameter(model, h == 1)
 @NLparameter(model, a == 0.1)
 @NLparameter(model, b == 0.1)

### 2021-04-15 model-appendix.Rmd
Describe how money costs can generate our changing signs for early/late first births. Add Julia code to compute equilibrium with money cost (and for CARA; not used now).

@@
 since it gives the closest match to our observations, and since it
-also generates ``reasonable'' predictions in other areas, e.g. that
+also generates "reasonable" predictions in other areas, e.g. that
 education levels increase with human capital. Figure \ref{fig:theory}
@@
 
+\subsection*{Including a money cost}
+
+The model can be extended by adding a money cost $m$ per child. Utility is
+then
+
+\[
+U = u(1 - s - bN_1 - mN_1) + u(w(s,h)(1 - bN_2) - mN_2) + a(N_1 + N_2)
+\]
+
+Figure \ref{fig:N-plot-with-m} shows a computed example with 
+$a = 0.4, b = 0.175, \sigma = 0.7, m = 0.075$. Fertility first declines steeply with
+human capital, then rises. In addition, for parents with low AFLB ($N_1 > 0$), the
+fertility-human capital relationship is negative, while for parents with 
+higher AFLB ($N_1 = 0$) it is positive.
+
+\begin{figure}
+\begin{centering}
+\includegraphics[width=9cm, height=6cm]{N-plot-with-m}
+\par\end{centering}
+\caption{\label{fig:N-plot-with-m}Fertility vs. human capital in the two-period model
+with money costs of children. $a=0.4,b=0.175,\sigma=0.7, m = 0.075$.}
+\end{figure}
+
 \subsection*{Concavity}

### 2021-04-15 why-natural-selection.Rmd
Describe how money costs can generate our changing signs for early/late first births. Add Julia code to compute equilibrium with money cost (and for CARA; not used now).

@@
 Our results are actually stronger than this, in that correlations with
-fertility are reversed at higher AFLB. This could occur if relative risk 
-aversion increased at higher income levels.
+fertility are reversed at higher AFLB. This could be accommodated in the model 
+if children have a money cost as well as a time cost, since money costs are less
+burdensome at higher wages, i.e. in period 2 for those with high human capital. Appendix Figure \@ref(fig:N-plot-with-m) shows an example.
 

### 2021-04-15 compute-CARA.jl

@@ -1,40 +1,0 @@
-
-
-using JuMP, Ipopt, Plots
-
-model = Model(Ipopt.Optimizer)
-set_optimizer_attribute(model, "print_level", 0)
-
-@variable(model, 0 <= N₁)
-@variable(model, 0 <= N₂)
-@variable(model, 0 <= s <= 1)
-
-@NLparameter(model, h == 1)
-@NLparameter(model, a == 0.1)
-@NLparameter(model, b == 0.1)
-@NLparameter(model, σ == 0.8)
-
-@NLconstraint(model, 1 - b * N₂ >= 0)
-@NLconstraint(model, 1 - s - b * N₁ >= 0)
-
-@NLobjective(model, Max,
-          - exp(-σ*(1-s-b*N₁)) 
-          - exp(-σ * (s*h)*(1-b*N₂)) 
-          + a*(N₁ + N₂)
-       )
-
-hhs = 0.1:0.1:2
-ipopt_result = Array{Float64}(undef, 3, length(hhs))
-
-for (ix, hh) in enumerate(hhs) 
-    set_value(h, hh)
-    optimize!(model)
-    ipopt_result[:, ix] = [value(N₁), value(N₂), value(s)]
-end
-
-
-plot(hhs, ipopt_result[1,:] + ipopt_result[2,:], label = "N")
-plot!(hhs, ipopt_result[1,:], label = "N1", linestyle = :dash)
-plot!(hhs, ipopt_result[2,:], label = "N2", linestyle = :dash)
-plot!(hhs, ipopt_result[3,:], label = "s", color = "black", 
-    linestyle = :dot)

### 2021-04-15 model-julia.jl

@@ -17,11 +17,15 @@ set_optimizer_attribute(model, "print_level", 0)
 @NLparameter(model, h == 1)
 @NLparameter(model, a == 0.1)
 @NLparameter(model, b == 0.1)
+@NLparameter(model, m == 0.1)
 @NLparameter(model, σ == 0.8)
 
+@NLconstraint(model, 1 - s - b * N₁ - m * N₁ >= 0)
+@NLconstraint(model, (s * h) * (1 - b * N₂) - m * N₂ >= 0)
+
 @NLobjective(model, Max, 
-    (1 - s - b * N₁)^(1 - σ)/(1 - σ) +
-    ((s * h) * (1 - b * N₂))^(1 - σ)/(1 - σ) + 
+    (1 - s - b * N₁ - m * N₁)^(1 - σ)/(1 - σ) +
+    ((s * h) * (1 - b * N₂) - m * N₂)^(1 - σ)/(1 - σ) + 
     a * (N₁ + N₂)
 )
 

### 2021-04-15 why-natural-selection.Rmd
small tweaks

@@
 * Formalize theory - done.
-  - Comment on results where effects are actually reversed. Could this
-  happen if \sigma were higher in period 2? Or if it increased with income?
-  Could maybe do a computational example.
-  - One possibility is CARA utility. Solving it computationally
+  - Comment on results where effects are actually reversed. This works
+    nicely if you throw in a money cost. Done computational example.
 * Clarify causality?
@@
   after correction for multiple testing, but a 25% difference. 
+  - I think we leave this for reviewers.
 
@@
 
-We also check for purifying and diversifying selection by 
-estimating \@ref(eq:regression) with a quadratic term. See Appendix Figure \@ref(fig:plot-purifying). In particular, we find diversifying selection for
-educational attainment polygenic scores: at higher values of these scores, 
-the effect on fertility is closer to zero.
+We also check for purifying and diversifying selection by estimating
+\@ref(eq:regression) with a quadratic term. In particular, we find diversifying
+selection for educational attainment polygenic scores: at lower values of these
+scores, the negative effect on fertility is larger (Appendix Figure
+\@ref(fig:plot-purifying)).
 
@@
       rename(p = p.value) %>% 
+      # can't use standard_ggplot because we have non-standard geom_point()
       ggplot(aes(
@@
         y      = term, 
+        yend   = term,
         shape  = Weights, 
@@
       )) +
+      my_vline +
       geom_point(size = 1.5, stroke = 0.5) +
-  my_vline +
+      confint_geom_segment() +
       pval_scale() +
@@
       labs(x = "Effect size", y = "") +
+      scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
       my_theme()

### 2021-04-21 why-natural-selection.Rmd
More about human capital; simplify graphs; add quadratic regression on n_sibs

@@
   - I think we leave this for reviewers.
+* Quadratic regressions for parents - DONE
 
@@
 
+res_significance <- res_wt_combined %>% 
+                      group_by(term) %>% 
+                      summarize(
+                        all = all(p.value < 0.05/33), 
+                        any = any(p.value < 0.05/33)
+                      )
+res_sig_all <- sum(res_significance$all)
+res_sig_any <- sum(res_significance$any)
+
 res_wt_cb_wide <- res_wt_combined %>% 
@@
 
-Since UK Biobank participants are not representative of the wider population,
-naïve estimates of natural selection are at risk of
-ascertainment bias. To correct for this, we weight participants using population
+To correct for ascertainment bias, we weight participants using population
 data. We try three alternative weighting schemes: weighting by geography, age
@@
 women only. Figure \@ref(fig:plot-res-wt) plots selection effects among the
-entire sample, estimated with the three weighting schemes.
-Mean effect sizes across all polygenic scores are
-increased by a factor of `r mean_rel$Geographical` (geographical weighting), 
-`r mean_rel[["Age/Qualification"]]` (age/qualification) or 
-`r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB). Estimates might be
-further affected by weighting on other demographic variables.
+entire sample, estimated with the three weighting schemes, and also with uniform
+weights. Effects for `r res_sig_all` out of 33 polygenic scores are significant
+at $p < 0.05/33$ under all four weighting schemes, with `r res_sig_any` scores
+reaching significance under at least one scheme. Weighting makes
+a large difference to estimates: mean effect sizes across all
+polygenic scores are increased by a factor of `r mean_rel$Geographical`
+(geographical weighting), `r mean_rel[["Age/Qualification"]]`
+(age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
+Estimates might be further affected by weighting on other demographic variables.
 
@@
 \@ref(eq:regression) with a quadratic term. In particular, we find diversifying
-selection for educational attainment polygenic scores: at lower values of these
-scores, the negative effect on fertility is larger (Appendix Figure
+selection for educational attainment polygenic scores: at higher values of these
+scores, the negative effect on fertility is smaller (Appendix Figure
 \@ref(fig:plot-purifying)).
@@
 *not* apportion the total correlation with fertility between subgroups. 
-That is because the polygenic score may also predict the probability of being in 
+That is because the polygenic score may also affect the probability of being in 
 each subgroup. For example, the EA3 polygenic score for educational attainment not only affects
@@
 that could affect fertility. However, natural selection is a correlational
-phenomenon, not a causal one. That is, if a polygenic score correlates with 
+phenomenon, not a causal one: if a polygenic score correlates with 
 low fertility, it is being selected against, whatever the underlying causal
-relationship. Differences in equation \@ref(eq:regression) across subgroups can 
-be informative about how natural selection works, because they constrain the set 
-of possible explanations to those that generate these differences.
+relationship. Irrespective of causality, differences in equation
+\@ref(eq:regression) across subgroups can inform us about how natural
+selection works, by constraining the set of possible explanations.
 
@@
 
+```{r calc-income-edu-controlled}
+
+drake::loadd(res_income_controlled, res_edu_controlled)
+n_regs_ic <- nrow(res_income_controlled)
+n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
+
+n_regs_educ <- nrow(res_edu_controlled)
+n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
+```
+
 Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
@@
 and by household income. Effects are larger and more significant for the lowest
-income category, and for the lowest education category.
-
-```{r calc-income-edu-controlled}
-
-drake::loadd(res_income_controlled, res_edu_controlled)
-n_regs_ic <- nrow(res_income_controlled)
-n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
-
-n_regs_educ <- nrow(res_edu_controlled)
-n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
-```
-
-These results could be explained by age, if older respondents have lower income
-and are less educated, and also show more natural selection on polygenic
-scores. However, when we rerun the regressions, interacting the polygenic
-score with income category and also with a quadratic in age, the
-interaction with income remains significant at 0.05/`r n_regs_ic` for
-`r n_sig_ic` out of `r n_regs_ic` regressions. Similarly if we interact
-the PGS with age of leaving full time education and a quadratic in age,
-the interaction with age leaving full time education remains significant at
-0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.
+income category, and for the lowest education category.  These results could be
+explained by age, if older respondents have lower income and are less educated,
+and also show more natural selection on polygenic scores. However, when we rerun
+the regressions, interacting the polygenic score with income category and also
+with a quadratic in age, the interaction with income remains significant at
+$p < 0.05/`r n_regs_ic`$ for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
+if we interact the PGS with age of leaving full time education and a quadratic
+in age, the interaction with age leaving full time education remains significant
+at $p < 0.05/`r n_regs_educ`$ for `r n_sig_educ` out of `r n_regs_educ` regressions.
 
@@
 For both sexes, selection effects are larger and more significant among
-those with more than 3 partners. Next we split respondents by whether
+those with more than 3 lifetime partners. Next we split respondents by whether
 they were living with a spouse or partner at the time of interview.
@@
 
-To investigate this further, we regress *number of children* on polygenic scores
+To investigate this further, we estimate equation \@ref(eq:regression)
 *controlling* for AFLB, again among females (Figure \@ref(fig:plot-age-flb)b).
@@
 through two opposing channels: an effect on AFLB, and an opposite-signed effect
-on number of children controlling for AFLB. Note that we do not claim that AFLB
-is exogenous to an individual's polygenic scores -- indeed, our analyses show
+on number of children controlling for AFLB. Again, we do not claim that AFLB
+is exogenous to an individual's polygenic scores. Indeed, our analyses show
 that it is not (Appendix Figure \@ref(fig:plot-age-flb-dv)). Rather, we argue
@@
 These results show that selection effects are weaker, absent, or even reversed
-among some subgroups of the population. One possible explanation for this is
-given by the economic theory of fertility 
+among some subgroups of the population. A possible explanation for this comes 
+from the economic theory of fertility 
 [@becker1960economic; @willis1973new;@becker1976child]. According to this theory,
@@
 in time, the opportunity cost of childrearing increases if one's market wage is
-higher. The income effect would lead higher earners to have more children. The
-substitution effect would lead higher earners to have fewer children. If so,
-then genetic variants which affect earnings potential in the labour market may
-cause opposing effects on fertility. The income effect will lead to natural
-selection in favour of earnings-increasing variants. The substitution effect
-will do the reverse.
-
-To demonstrate this, we write down a simple model of 
-fertility choices. $h$ is an individual's level of human capital, which 
-(for now) we identify with his or her wage $W$. Raising a child takes time $b$. People maximize utility
-from the number of children $N$, and income $Y\equiv(1-bN)W$. An
-individual's payoff is
+higher. The income effect leads higher earners to have more children. The
+substitution effect leads them to have fewer children.
+
+Suppose that certain genetic variants correlate with human capital -- skills or 
+other characteristics that affect an individual's earnings in the labour market.
+These variants may then be associated with opposing effects on fertility. The income
+effect will lead to natural selection in favour of earnings-increasing variants
+(or variants that are merely associated with higher earnings). The substitution
+effect will do the reverse.
+
+To show this, we write down a simple model of fertility choices. $h$ is
+an individual's level of human capital. For now, we simply identify this with
+his or her wage $W$. Raising a child takes time $b$. People maximize utility
+from the number of children $N$, and income $Y\equiv(1-bN)W$. An individual's
+payoff is
 
@@
 among similar people. The marginal benefit of an extra child is
-$\frac{dU}{dN}=-bWu'(Y)+a$. The effect of an increase in wages (or human capital) 
-on this marginal benefit is
+$\frac{dU}{dN}=-bWu'(Y)+a$. The effect of an increase in human capital on this
+marginal benefit is
 
@@
 
-The *substitution effect* is negative and reflects that when
-wages increase, time devoted to childcare costs more in foregone income.
-The positive *income effect* reflects that when wage income is
-higher, the marginal loss of wages from children is less painful.
-Letting $u(\cdot)$ take the constant relative risk aversion (CRRA)
-form $u(y)=\frac{y^{1-\sigma}-1}{1-\sigma}$, where $\sigma>0$ captures
-risk aversion in income, we solve for optimal fertility $N^{*}$,
-and examine the *fertility-human capital relationship*, $\frac{dN^{*}}{dh}$.
-For moderate levels of risk aversion ($0.5<\sigma<1$):
+The *substitution effect* is negative and reflects that when wages increase,
+time devoted to childcare costs more in foregone income. The positive *income
+effect* depends on the curvature of the utility function, and reflects that when
+income is higher, the marginal loss of income from children is less painful.
+Letting $u(\cdot)$ take the constant relative risk aversion (CRRA) form
+$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
+in income, we solve for optimal fertility $N^{*}$, and examine the
+*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For moderate levels
+of risk aversion ($0.5<\sigma<1$):
 
 \begin{enumerate}
-\item The fertility-human capital relationship is negative: $\frac{dN^{*}}{dh}<0$.
+\item The fertility-human capital relationship is negative: $\frac{dN^{*}}{dh} < 0$.
 \item The relationship is weaker (closer to zero) at higher wages and/or
@@
 \begin{equation}
-U(N_{1},N_{2})=u(Y_{1})+u(Y_{2})+a(N_{1}+N_{2})\label{eq:U}
+U(N_{1},N_{2}) = u(Y_{1}) + u(Y_{2}) + aN_{1} + aN_{2})\label{eq:U}
 \end{equation}
@@
 Instead of identifying human capital $h$ with wages, we now allow
-individuals to choose a level of education $s\in[0,1]$, which has
+individuals to choose a level of education $s \in [0,1]$, which has
 a time cost in period 1. Education increases period 2 wages, and is
-complementary to human capital $h>0$. For period 2 wages we assume
-the simple functional form $w(s,h)=sh$. We normalize period 1 wages
+complementary to human capital $h > 0$. For period 2 wages we assume
+the simple functional form $w(s,h) = sh$. We normalize period 1 wages
 to 1. Maintaining the assumption of CRRA utility, we examine total
-fertility $N^{*}\equiv N_{1}^{*}+N_{2}^{*}$. For $0.5<\sigma<1$,
-facts 1-3 continue to hold (fact 2 in a neighborhood of $\sigma=1$).
+fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$. For $0.5 < \sigma < 1$,
+facts 1-3 continue to hold (fact 2 in a neighborhood of $\sigma = 1$).
 In addition, for $\sigma$ in a neighborhood of 1:
@@
 \item The relationship is weaker among those who start fertility in period
-2 ($N_{1}^{*}=0$) than among those who start fertility in period
-1 ($N_{1}^{*}>0$).
+2 ($N_{1}^{*} = 0$) than among those who start fertility in period
+1 ($N_{1}^{*} > 0$).
 \end{enumerate}
@@
 low income and education levels, and a weaker or zero correlation at higher
-levels (Figure \@ref(fig:plot-income-educ-level)). For scores which are highly
-correlated with human capital, the correlation would also be weaker at higher
-levels of the score itself, giving rise to diversifying selection (Appendix
+levels (Figures \@ref(fig:plot-income-educ-level)a and \@ref(fig:plot-income-educ-level)b). Among scores which are highly
+correlated with human capital, the correlation might also be weaker at higher
+levels of the score itself, giving rise to diversifying selection, as we observe
+for scores for educational attainment (Appendix
 Figure \@ref(fig:plot-purifying)). The correlation would be more negative when
@@
 
-The above does not provide a complete theory of fertility. Rather, it shows that
-a relatively simple economic model can explain many of our results. Other
-empirical work in economics also supports our mechanism.
+The above does not aim to provide a complete theory of fertility. Rather, it
+shows that a relatively simple economic model can explain many of our results.
+Other empirical work in economics also supports our mechanism.
 @caucutt2002women and @monstad2008education show that education and skills
@@
 
-```{r plot-cors-earnings-educ, fig.cap="Selection effects, with and without controls for age at first live birth, by correlations with earnings and educational attainment.
-Each point represents one polygenic score.", fig.height = 4}
+```{r plot-cors-earnings-educ, fig.cap="Selection effects by correlations with earnings and educational attainment. Each point represents one polygenic score.", fig.height = 4}
 
@@
       ) %>% 
+      filter(
+        Controls == "None"
+      ) %>% 
       ggplot(aes(effect_size, cor, colour = Correlation)) + 
@@
                     check_overlap = TRUE, nudge_x = 0.005) +
-        facet_grid(Correlation ~ Controls, labeller = label_both,
-                   scales = "free") + 
+        facet_wrap( ~ Correlation, scales = "free") + 
         my_vline + 
         my_hline +
-        labs(x = "Selection effect", y = "") +
+        labs(x = "Selection effect", y = "Correlation") +
         coord_cartesian(clip = "off") +
@@
 on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
-on each polygenic score against that score's correlation with earnings in a
-respondent's first job, and against its correlation with educational attainment,
-a predictor of lifetime earnings. (See also the Supplementary Animations for the
+on each polygenic score against that score's correlation with two measures of
+human capital: earnings in a respondent's first job, and educational attainment.
+The relationships are strongly negative. Thus, human capital appears to be relevant
+to natural selection. Substitution effects dominate income effects overall, which fits the known association between income and lower fertility 
+[@becker1960economic;@jones2006economic]. The correlations reverse when we control
+for age at first live birth, suggesting that within AFLB categories, the income effect dominates. (See also the Supplementary Animations for the
 uncorrected effects per AFLB group, which show the correlation reversing after
-age 22.) The raw relationships (left column) are strongly negative. If we plot
-selection effects controlling for age at first live birth (right column), the
-effect reverses and becomes positive. Thus, the labour market appears to play an
-important role in natural selection. Substitution effects dominate income
-effects overall, which fits the known association between income and lower
-fertility [@becker1960economic;@jones2006economic].
+age 22.)
 
@@
 generally reduced (Appendix Figure \@ref(fig:plot-res-earn-educ)), and only 
-`r n_05_33` out of 33 scores are significant at $p = 0.05/33$. This suggests that
+`r n_05_33` out of 33 scores are significant at $p < 0.05/33$. This suggests that
 earnings and education indeed mediate the effect of polygenic scores on
@@
 An alternative theory is that polygenic scores correlate with the motivation to
-have children (parameter $a$ in the model; @jones2008fertility). However, this
+have children (parameter $a$ in the model; cf. @jones2008fertility). However, this
 theory would not explain why our results are weaker at higher incomes and
-education levels. Indeed, in the one-period model, while $\frac{dN^{*}}{da}>0$,
-this relationship actually becomes stronger at higher wages.
+education levels.[^note]
+
+[^note]: Indeed, in the one-period model above, while $\frac{dN^{*}}{da} > 0$,
+this relationship becomes stronger, not weaker, at higher wages.
 
 Another alternative is that traits under selection are linked to externalizing
-behaviour and risk-seeking. This might be captured by our risk-aversion 
-parameter $\sigma$, which affects fertility by changing the marginal utility of income, but a more direct channel is risky sexual behaviour
+behaviour and risk-seeking. This might be partially captured by our risk-aversion 
+parameter $\sigma$, which affects fertility by changing the marginal utility of 
+income, but a more direct channel is risky sexual behaviour
 [@Mills2020.05.06.081273]. The data here provide some support for this: scores
@@
 sizes between regressions with and without controls is `r median(ratio)`; all
-scores which are significant at $p \le 0.05/33$ in uncontrolled regressions
+scores which are significant at $p < 0.05/33$ in uncontrolled regressions
 remain so when controlling for risk attitude. This non-result could simply
@@
 question. However, the risk aversion measure *does* predict the overall number
-of children, highly significantly ($p \le 2 \times 10^{-16}$, `r n_risk_sig`/33
-regressions). Given that, and the statistical power we get from our large N, we 
-believe that the non-result is real: while risk aversion does predict fertility 
+of children, highly significantly ($p < 2 \times 10^{-16}$, `r n_risk_sig`/33
+regressions). Given that, and the statistical power we get from our sample size, 
+we believe that the non-result is real: while risk aversion does predict fertility 
 in the sample, it is not an important channel for natural selection.
@@
 including people with lower income, lower education, younger first parenthood,
-and more lifetime sexual partners. Indeed, among older mothers (22+), selection effects are
-actually reversed. Furthermore, the size of selection effects on a polygenic
-score correlates with that score's association with labour market earnings. The
-economic theory of fertility provides a parsimonious explanation for this.
+and more lifetime sexual partners. Among older mothers (22+), selection effects
+are reversed. Furthermore, the size of selection effects on a polygenic score
+correlates with that score's association with labour market earnings. The
+economic theory of fertility provides a parsimonious explanation for these findings.
+Because of the substitution effect of earnings on fertility, scores are selected
+for when they correlate with low human capital, and this effect is stronger at 
+lower levels of income and education.
 
@@
 between highest and lowest income groups (Appendix Table
-\@ref(tab:tbl-inequality-effects)). These are substantial effects, and since
-many polygenic scores are predictive of disease risk, they could potentially
-increase health inequalities. In general, the evolutionary history of
-anatomically modern humans is related to disease risk [@benton2021influence];
-understanding the role of contemporary natural selection may aid in mapping the
-genetic architecture of current health disparities.
+\@ref(tab:tbl-inequality-effects)). Inequalities in polygenic scores are
+believed to be important for understanding class mobility and structure 
+[see e.g. @belsky2018genetic, @Rimfeld_2018]. If so, these changes are substantial
+effects. Since many polygenic scores are predictive of disease risk, they could
+also potentially increase health inequalities. In general, the evolutionary
+history of anatomically modern humans is related to disease risk
+[@benton2021influence]; understanding the role of contemporary natural selection
+may aid in mapping the genetic architecture of current health disparities.
 
@@
 
+We also checked for purifying selection in the parents' generation, using
+age/qualification weights multipled by number of siblings. Scores for EA2 and
+EA3 again show significant diversifying selection ($p < 0.05/33$, positive coefficient). Other scores including hip circumference and drinks per week 
+are not significant.
+
 ```{r plot-purifying, fig.align = "center", fig.cap = "Purifying/diversifying selection: predicted number of children by polygenic score.", fig.height = 5}
@@
 sig_all_terms <- res_quadratic %>% filter(score_name %in% sig)
+
+drake::loadd(res_sibs_quadratic)
+sig_sibs <- res_sibs_quadratic %>% 
+         filter(grepl("^2", term, fixed = TRUE)) %>% 
+         filter(p.value < 0.05/33) %>% 
+         pull(score_name)
+stopifnot(sig_sibs == c("EA2_noUKB", "EA3_excl_23andMe_UK"))
 

### 2021-04-21 _drake.R

@@ -170,6 +170,19 @@ plan <- drake_plan(
             )
   },
   
+  res_sibs_quadratic = {
+    famhist %<>% inner_join(parent_weights, by = "f.eid")
+    map_dfr(score_names, 
+            ~run_regs_fml(
+              "n_sibs ~ {score_name} + I({score_name}^2)", 
+              score_name = .x,
+              famhist    = famhist,
+              weights    = quote(weights)
+            ),
+            .id = "score_name"
+    )
+  },
+  
   res_pcs = {
     pc_names <- grep("PC", names(pcs), value = TRUE)
     res_sibs_pcs <- run_regs_pcs( 
@@ -534,16 +547,16 @@ plan <- drake_plan(
             .id = "score_name")
   },
   
-    res_risk_control = {
-      # f.2040.0.0 is risk attitude (via questionnaire)
-      map_dfr(score_names, 
-                ~ run_regs_fml(
-                  fml = "n_children ~ {score_name} + f.2040.0.0",
-                  score_name = .x,
-                  famhist    = famhist,
-                  subset     = quote(kids_ss)
-                ), 
-              .id = "score_name")
+  res_risk_control = {
+    # f.2040.0.0 is risk attitude (via questionnaire)
+    map_dfr(score_names, 
+              ~ run_regs_fml(
+                fml = "n_children ~ {score_name} + f.2040.0.0",
+                score_name = .x,
+                famhist    = famhist,
+                subset     = quote(kids_ss)
+              ), 
+            .id = "score_name")
   },
   
   res_ineq = {

### 2021-04-21 why-natural-selection.Rmd
Remove the one-period model results.

@@
 among similar people. The marginal benefit of an extra child is
-$\frac{dU}{dN}=-bWu'(Y)+a$. The effect of an increase in human capital on this
+$\frac{dU}{dN} = -bWu'(Y)+a$. The effect of an increase in human capital on this
 marginal benefit is
@@
 income is higher, the marginal loss of income from children is less painful.
-Letting $u(\cdot)$ take the constant relative risk aversion (CRRA) form
-$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
-in income, we solve for optimal fertility $N^{*}$, and examine the
-*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For moderate levels
-of risk aversion ($0.5<\sigma<1$):
-
-\begin{enumerate}
-\item The fertility-human capital relationship is negative: $\frac{dN^{*}}{dh} < 0$.
-\item The relationship is weaker (closer to zero) at higher wages and/or
-levels of human capital.
-\item The relationship is more negative when the time burden of children,
-$b$, is larger.
-\end{enumerate}
 
@@
 
-Instead of identifying human capital $h$ with wages, we now allow
+Instead of identifying human capital with wages, we now allow
 individuals to choose a level of education $s \in [0,1]$, which has
-a time cost in period 1. Education increases period 2 wages, and is
-complementary to human capital $h > 0$. For period 2 wages we assume
+a time cost in period 1. Education is
+complementary to human capital $h > 0$, and increases period 2 wages, which take 
 the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
-to 1. Maintaining the assumption of CRRA utility, we examine total
-fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$. For $0.5 < \sigma < 1$,
-facts 1-3 continue to hold (fact 2 in a neighborhood of $\sigma = 1$).
-In addition, for $\sigma$ in a neighborhood of 1:
-
-\begin{enumerate}
-\item [4.]\setcounter{enumi}{4}The fertility-human capital relationship
-is weaker at higher levels of education $s$.
-\item The relationship is weaker among those who start fertility in period
+to 1. Letting $u(\cdot)$ take the constant relative risk aversion (CRRA) form
+$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
+in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
+*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$,
+and in a neighbourhood of $\sigma = 1$:
+
+1. The fertility-human capital relationship is negative: $\frac{dN^{*}}{dh} < 0$.
+2. The relationship is weaker (closer to zero) at higher wages and/or
+  levels of human capital.
+3. The relationship is more negative when the time burden of children,
+  $b$, is larger.
+4. The relationship is weaker at higher levels of education $s$.
+5. The relationship is weaker among those who start fertility in period
   2 ($N_{1}^{*} = 0$) than among those who start fertility in period
   1 ($N_{1}^{*} > 0$).
-\end{enumerate}
-
-Facts 1-5 match our empirical results. In this model, polygenic scores which
-correlate with human capital would have a negative correlation with fertility at
-low income and education levels, and a weaker or zero correlation at higher
-levels (Figures \@ref(fig:plot-income-educ-level)a and \@ref(fig:plot-income-educ-level)b). Among scores which are highly
-correlated with human capital, the correlation might also be weaker at higher
-levels of the score itself, giving rise to diversifying selection, as we observe
-for scores for educational attainment (Appendix
-Figure \@ref(fig:plot-purifying)). The correlation would be more negative when
-the time burden of children $b$ is high, e.g. for single parents (Figure
+
+Facts 1-5 parallel our empirical results.[^one-period-results] In this model,
+polygenic scores that correlate with human capital would have a negative
+correlation with fertility at low income and education levels, and a weaker or
+zero correlation at higher levels (Figures \@ref(fig:plot-income-educ-level)a
+and \@ref(fig:plot-income-educ-level)b). Among scores which are highly
+correlated with human capital, the fertility correlation might also be weaker at
+higher levels of the score itself, giving rise to diversifying selection, as we
+observe for scores for educational attainment (Appendix Figure
+\@ref(fig:plot-purifying)). The correlation would be more negative when the time
+burden of children $b$ is high, e.g. for single parents (Figure
 \@ref(fig:plot-n-partners)). Lastly, the correlation would be weaker among those
-starting fertility later (Figure \@ref(fig:plot-age-flb)).
-
-Our results are actually stronger than this, in that correlations with
-fertility are reversed at higher AFLB. This could be accommodated in the model 
-if children have a money cost as well as a time cost, since money costs are less
-burdensome at higher wages, i.e. in period 2 for those with high human capital. Appendix Figure \@ref(fig:N-plot-with-m) shows an example.
+starting fertility later (Figure \@ref(fig:plot-age-flb)).[^money-costs]
+
+[^one-period-results]: Facts 1-3 also hold in the one-period model.
+
+[^money-costs]: Our results are actually stronger than this, in that
+correlations with fertility are reversed at higher AFLB. This prediction can be
+accommodated in the model if children have a money cost as well as a time cost. 
+Appendix Figure \@ref(fig:N-plot-with-m) shows an example.
 
@@
 
-```{r plot-cors-earnings-educ, fig.cap="Selection effects by correlations with earnings and educational attainment. Each point represents one polygenic score.", fig.height = 4}
+```{r plot-cors-earnings-educ, fig.cap="Selection effects by correlations with earnings and educational attainment. Each point represents one polygenic score."}
 
@@
 We test this explanation in two ways. First, the economic theory predicts that
-genetic variants will be selected for (or against) in proportion to their effect
-on earnings. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
+genetic variants will be selected for (or against) in proportion to their correlation with human capital. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
 on each polygenic score against that score's correlation with two measures of
@@
 theory would not explain why our results are weaker at higher incomes and
-education levels.[^note]
-
-[^note]: Indeed, in the one-period model above, while $\frac{dN^{*}}{da} > 0$,
-this relationship becomes stronger, not weaker, at higher wages.
-
-Another alternative is that traits under selection are linked to externalizing
-behaviour and risk-seeking. This might be partially captured by our risk-aversion 
-parameter $\sigma$, which affects fertility by changing the marginal utility of 
-income, but a more direct channel is risky sexual behaviour
-[@Mills2020.05.06.081273]. The data here provide some support for this: scores
-which might plausibly be linked to externalizing behaviour, like ADHD and
-(younger) age at smoking initiation, are selected for. However, this theory is
-less good at explaining variation in selection across the full range of scores,
-including physical measures such as waist-hip ratio and BMI. We test this theory
-directly by re-estimating equation \@ref(eq:regression) controlling for a
-measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
-sizes between regressions with and without controls is `r median(ratio)`; all
-scores which are significant at $p < 0.05/33$ in uncontrolled regressions
-remain so when controlling for risk attitude. This non-result could simply
-reflect the imprecision of the risk aversion measure, which is a single yes/no
-question. However, the risk aversion measure *does* predict the overall number
-of children, highly significantly ($p < 2 \times 10^{-16}$, `r n_risk_sig`/33
-regressions). Given that, and the statistical power we get from our sample size, 
-we believe that the non-result is real: while risk aversion does predict fertility 
-in the sample, it is not an important channel for natural selection.
+education levels.[^note] Another alternative is that traits under selection are
+linked to externalizing behaviour and risk-seeking. This might be partially
+captured by our risk-aversion parameter $\sigma$, which affects fertility by
+changing the marginal utility of income, but a more direct channel is risky
+sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
+for this: scores which might plausibly be linked to externalizing behaviour,
+like ADHD and (younger) age at smoking initiation, are selected for. However,
+this theory is less good at explaining variation in selection across the full
+range of scores, including physical measures such as waist-hip ratio and BMI. We
+test this theory directly by re-estimating equation \@ref(eq:regression)
+controlling for a measure of risk attitude (UK Biobank field 2040).  The median
+ratio of effect sizes between regressions with and without controls is 
+`r median(ratio)`; all scores which are significant at $p < 0.05/33$ in
+uncontrolled regressions remain so when controlling for risk attitude. This
+non-result could simply reflect the imprecision of the risk aversion measure,
+which is a single yes/no question. However, the risk aversion measure *does*
+predict the overall number of children, highly significantly ($p < 2 \times
+10^{-16}$, `r n_risk_sig`/33 regressions). Given that, and the statistical power
+we get from our sample size, we believe that the non-result is real: while risk
+aversion does predict fertility in the sample, it is not an important channel
+for natural selection.
+
+[^note]: Indeed in the one-period model, while $\frac{dN^{*}}{da} > 0$,
+this effect becomes stronger, not weaker, at higher wages.
 

### 2021-07-16 why-natural-selection.Rmd
New fertility prs; use income_cat not first_job_pay as a control; plan within-sib regressions with/without controls; comments from Abdel; table to explain theory and empirics together.

@@
 ---
-title: "Human capital drives natural selection in contemporary humans"
+title: "Human capital mediates natural selection in contemporary humans"
 author: "David Hugh-Jones \\thanks{Corresponding author.
 School of Economics, University of East Anglia, Norwich, UK.
-Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands.
-Email: a.abdellaoui@amsterdamumc.nl}"
+Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands.}"
 date: "`r format(Sys.Date(), '%d %B %Y')`"
@@
 mechanisms behind natural selection can help us better understand how individual
-differences in complex traits and disease risk arise [@benton2021influence]. Recent work confirms that
-natural selection is taking place in modern human populations, using genome-wide
-analysis
+differences in complex traits and disease risk arise [@benton2021influence].
+Recent work confirms that natural selection is taking place in modern human
+populations, using genome-wide analysis
 [@Barban_2016;@beauchamp2016genetic;@kong2017selection;@sanjak2018evidence]. In
@@
 
-To test directly for natural selection, we regress respondents' number of
+To test more directly for natural selection, we regress respondents' number of
 children ($N_i$) on their polygenic scores (PGS):
@@
 the expected polygenic score among children of the sample
-[@beauchamp2016genetic].[^normalization]
+[@beauchamp2016genetic].[^normalization] Note that equation 
+\eqref{eq:regression} does not include many possible environmental and
+genetic confounds that could affect fertility, and as a result, $\beta$ is not
+an estimate of the causal effect of a polygenic score on fertility. However, 
+natural selection is a correlational phenomenon, not a causal one: if a 
+polygenic score correlates with low fertility, it is being selected against, 
+whatever the underlying causal relationship.
 
@@
 To correct for ascertainment bias, we weight participants using population
-data. We try three alternative weighting schemes: weighting by geography, age
-and presence/absence of a partner; weighting by age and highest educational
-qualification; and age, highest qualification, and age at first live birth, for
-women only. Figure \@ref(fig:plot-res-wt) plots selection effects among the
+data. We try three alternative weighting schemes: (1) "Geographical" weighting,
+by geography, age and presence/absence of a partner; (2) age and highest educational
+qualification; and (3) for women only, age, highest qualification, and age at 
+first live birth. Figure \@ref(fig:plot-res-wt) plots selection effects among the
 entire sample, estimated with the three weighting schemes, and also with uniform
@@
 
-We also check for purifying and diversifying selection by estimating
+We also check for balancing and diversifying selection by estimating
 \@ref(eq:regression) with a quadratic term. In particular, we find diversifying
@@
 
-To understand the reasons for this ascertainment bias, and to learn about the
-mechanisms underlying natural selection, we split the sample, starting with
+To understand the reasons for the ascertainment bias in our sample, and to learn 
+about the mechanisms underlying natural selection, we split the sample, starting with
 basic demographic variables including education, income and sex. These are all
@@
 
-We emphasize two points. First, this exercise does
-*not* apportion the total correlation with fertility between subgroups. 
+We emphasize that this exercise does
+*not* apportion the total correlation with fertility into subgroups. 
 That is because the polygenic score may also affect the probability of being in 
@@
 fertility among university graduates, and among non-graduates; it also changes
-a person's chance of attending university. Second, we do *not* claim that our
-observed correlations reflect causal effects of polygenic scores on fertility.
-Indeed, polygenic scores correlate with many sources of environmental variation 
-that could affect fertility. However, natural selection is a correlational
-phenomenon, not a causal one: if a polygenic score correlates with 
-low fertility, it is being selected against, whatever the underlying causal
-relationship. Irrespective of causality, differences in equation
+a person's chance of attending university. Nevertheless, differences in equation
 \@ref(eq:regression) across subgroups can inform us about how natural
-selection works, by constraining the set of possible explanations.
+selection works, by constraining the set of possible mechanisms.
 
@@
 \[
-U(N)=u(Y)+aN.
+U = u(Y)+aN.
 \]
 
-Here $a$ captures the strength of preference for children. $u(\cdot)$
-is concave and increasing. We treat $N$ as continuous, in line with
+Here $a$ captures the strength of preference for children. $u(\cdot)$ captures
+the taste for income, and is concave and increasing. We treat $N$ as continuous, in line with
 the literature: this can be thought of as the expected number of children
@@
 \begin{equation}
-U(N_{1},N_{2}) = u(Y_{1}) + u(Y_{2}) + aN_{1} + aN_{2})\label{eq:U}
+U(N_{1},N_{2}) = u(Y_{1}) + u(Y_{2}) + aN_{1} + aN_{2}\label{eq:U}
 \end{equation}
@@
 *fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$,
-and in a neighbourhood of $\sigma = 1$:
-
-1. The fertility-human capital relationship is negative: $\frac{dN^{*}}{dh} < 0$.
-2. The relationship is weaker (closer to zero) at higher wages and/or
-  levels of human capital.
-3. The relationship is more negative when the time burden of children,
-  $b$, is larger.
-4. The relationship is weaker at higher levels of education $s$.
-5. The relationship is weaker among those who start fertility in period
-  2 ($N_{1}^{*} = 0$) than among those who start fertility in period
-  1 ($N_{1}^{*} > 0$).
-
-Facts 1-5 parallel our empirical results.[^one-period-results] In this model,
-polygenic scores that correlate with human capital would have a negative
-correlation with fertility at low income and education levels, and a weaker or
-zero correlation at higher levels (Figures \@ref(fig:plot-income-educ-level)a
-and \@ref(fig:plot-income-educ-level)b). Among scores which are highly
-correlated with human capital, the fertility correlation might also be weaker at
-higher levels of the score itself, giving rise to diversifying selection, as we
-observe for scores for educational attainment (Appendix Figure
-\@ref(fig:plot-purifying)). The correlation would be more negative when the time
-burden of children $b$ is high, e.g. for single parents (Figure
-\@ref(fig:plot-n-partners)). Lastly, the correlation would be weaker among those
-starting fertility later (Figure \@ref(fig:plot-age-flb)).[^money-costs]
-
-[^one-period-results]: Facts 1-3 also hold in the one-period model.
-
-[^money-costs]: Our results are actually stronger than this, in that
-correlations with fertility are reversed at higher AFLB. This prediction can be
+and in a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical 
+predictions, along with corresponding empirical results for the correlation between
+polygenic scores and fertility.[^one-period-results]
+
+[^one-period-results]: Predictions 1-3 also hold in the one-period model. Our empirical 
+results are actually stronger than prediction 5, in that
+correlations with fertility are *reversed* at higher AFLB. This prediction can be
 accommodated in the model if children have a money cost as well as a time cost. 
 Appendix Figure \@ref(fig:N-plot-with-m) shows an example.
+
+------------------------------------------------------------------------------------
+    **Theory**                                        **Empirical results**
+--- ------------------------------------------------- ------------------------------
+1.  The fertility-human capital                       Figures \@ref(fig:plot-means-over-time) 
+    relationship is negative:                         and \@ref(fig:plot-res-wt).
+    $\frac{dN^{*}}{dh} < 0$.
+
+2.  The fertility-human capital                       Figure \@ref(fig:plot-income-educ-level)a. 
+    relationship is weaker (closer to zero)           Selection effects are also weaker at higher
+    at higher wages and/or levels of                  polygenic scores for EA (Appendix Figure
+    human capital.                                    \@ref(fig:plot-purifying)).
+
+3.  The relationship is more negative when            Stronger effects for single parents
+    the time burden of children, $b$, is              (Figure \@ref(fig:plot-n-partners)).
+    larger.
+
+4.  The relationship is weaker at higher              Figure \@ref(fig:plot-income-educ-level)b.
+    levels of education $s$.
+
+5.  The relationship is weaker among those            Effects weaker among those starting 
+    who start fertility in period 2                   fertility later 
+    ($N_{1}^{*} = 0$) than among those who            (Figure \@ref(fig:plot-age-flb)a).
+    start fertility in period 1 
+    ($N_{1}^{*} > 0$).
+----------------------------------------------------------------------------------------
+
+Table: (\#tab:theory) Predictions from the theoretical model and corresponding 
+empirical results.
 
@@
 US fertility decreases faster with education among single mothers than married
-mothers [@baudin2015fertility], in line with our fact 3 and as predicted by
+mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
 @becker1981treatise.
@@
 We test this explanation in two ways. First, the economic theory predicts that
-genetic variants will be selected for (or against) in proportion to their correlation with human capital. Figure \@ref(fig:plot-cors-earnings-educ) plots selection effects
-on each polygenic score against that score's correlation with two measures of
-human capital: earnings in a respondent's first job, and educational attainment.
-The relationships are strongly negative. Thus, human capital appears to be relevant
-to natural selection. Substitution effects dominate income effects overall, which fits the known association between income and lower fertility 
-[@becker1960economic;@jones2006economic]. The correlations reverse when we control
-for age at first live birth, suggesting that within AFLB categories, the income effect dominates. (See also the Supplementary Animations for the
-uncorrected effects per AFLB group, which show the correlation reversing after
-age 22.)
+genetic variants will be selected for (or against) in proportion to their
+correlation with human capital. Figure \@ref(fig:plot-cors-earnings-educ) plots
+selection effects on each polygenic score against that score's correlation with
+two measures of human capital: earnings in a respondent's first job, and
+educational attainment. The relationships are strongly negative. Thus, human
+capital appears to be relevant to natural selection. Substitution effects
+dominate income effects overall, which fits the known association between income
+and lower fertility [@becker1960economic;@jones2006economic]. The correlations
+reverse when we control for age at first live birth, suggesting that within AFLB
+categories, the income effect dominates. (See also the Supplementary Animations
+for the uncorrected effects per AFLB group, which show the correlation reversing
+after age 22.)
 
@@
 controlling for earnings in first job and education levels. Effect sizes are
-generally reduced (Appendix Figure \@ref(fig:plot-res-earn-educ)), and only 
-`r n_05_33` out of 33 scores are significant at $p < 0.05/33$. This suggests that
-earnings and education indeed mediate the effect of polygenic scores on
-fertility.
+generally reduced (Appendix Figure \@ref(fig:plot-res-earn-educ)), with
+large reductions for polygenic scores for educational attainment. This 
+suggests that earnings and education indeed mediate the effect of 
+polygenic scores on fertility.
+
+* We estimate the causal effect of each PGS on fertility using within-sibling
+  regressions.
+  - These are causal due to lottery of meiosis
+  - Also do the fertility PGS
+
+* We also estimate the effect of PSEA and fertility PGS on education and
+  earnings, again within-siblings
+  - confirm PSEA affects both; does fertility? What if we put both in together?
+* Lastly we again estimate effect of PSEAs on fertility, 
+  controlling for education and earnings in first job. 
+  - We cannot exclude some reverse causality from EA3 -> fertility -> education/income
+  - Our theory indeed assumes that people simultaneously choose N and s, while taking into
+    account their effect on Y. We don't make claims about exogenous (chance) variation
+    in Y or in N. But note that exog variation in N1 (i.e. no choice, treat
+    it as fixed to different values) would affect s. OTOH, if EA3 affects $a$ and
+    thereby changes things, you don't get predictions 2 or 4.
+
+* Results:
+  - in bivariate within-sib regs, extraversion, EA2/3, ADHD and some others are
+    weakly significant (only at .05!)
+  - adding education as a control barely changes effect sizes
+  - adding income as a control increases effect sizes (of EA at least)
+  - by contrast, in pooled regs, adding education shrinks effect sizes, and
+    adding income does nothing.
+  - relevant: effect of EA2 on education is halved within-sib, and effect of
+    education on fertility is halved within-sib
+  - whereas, effect of EA2 on income is halved within-sib, but effect of income
+    on fertility is multiplied by 10!! within-sib. Much of this is because
+    of year of birth confound i.e. declining fertility and also retired people
+    have smaller income
 
@@
 changing the marginal utility of income, but a more direct channel is risky
-sexual behaviour [@Mills2020.05.06.081273]. The data here provide some support
+sexual behaviour [@mills2021identification]. The data here provide some support
 for this: scores which might plausibly be linked to externalizing behaviour,
@@
 education are being selected against (selected for). In addition, many of the
-phenotypes under positive selection are linked to disease risk. Most people
+phenotypes under positive selection are linked to disease risk. Many people
 would probably prefer to have high educational attainment, a low risk of ADHD
@@
 between highest and lowest income groups (Appendix Table
-\@ref(tab:tbl-inequality-effects)). Inequalities in polygenic scores are
-believed to be important for understanding class mobility and structure 
-[see e.g. @belsky2018genetic, @Rimfeld_2018]. If so, these changes are substantial
-effects. Since many polygenic scores are predictive of disease risk, they could
-also potentially increase health inequalities. In general, the evolutionary
-history of anatomically modern humans is related to disease risk
-[@benton2021influence]; understanding the role of contemporary natural selection
-may aid in mapping the genetic architecture of current health disparities.
+\@ref(tab:tbl-inequality-effects)). If inequalities in polygenic scores are
+important for understanding class mobility and structure [@belsky2018genetic;
+@Rimfeld_2018], then these increases are substantial effects. Since many
+polygenic scores are predictive of disease risk, they could also potentially
+increase health inequalities. In general, the evolutionary history of
+anatomically modern humans is related to disease risk [@benton2021influence];
+understanding the role of contemporary natural selection may aid in mapping the
+genetic architecture of current health disparities.
 
@@
 \clearpage
-## Purifying and diversifying selection
+## Balancing and diversifying selection
 
@@
 and using our age/qualification weights. Scores for hip circumference and drinks
-per week show significant purifying selection ($p < 0.05/33$, negative
+per week show significant balancing selection ($p < 0.05/33$, negative
 coefficient). Scores for educational attainment (EA2 and EA3) show significant
@@
 
-We also checked for purifying selection in the parents' generation, using
-age/qualification weights multipled by number of siblings. Scores for EA2 and
-EA3 again show significant diversifying selection ($p < 0.05/33$, positive coefficient). Other scores including hip circumference and drinks per week 
-are not significant.
+We also checked for balancing selection in the parents' generation, using
+age/qualification weights multiplied by the inverse of number of siblings. 
+Scores for EA2 and EA3 again show significant diversifying selection 
+($p < 0.05/33$, positive coefficient). Other scores including hip circumference 
+and drinks per week are not significant.
 

### 2021-07-16 make-data.R

@@ -301,3 +301,15 @@ add_deprivation_data <- function (famhist, dep_data_dir) {
   
   famhist
 }
+
+
+add_fertility_prs <- function (famhist, fertility_data_dir) {
+  fertility_prs <- readr::read_table2(
+                     file.path(fertility_data_dir, 
+                               "UKB.AMC.NEB_POOLED.HM3.EUR.SBLUP.10k.csv")
+  )
+  famhist %<>% left_join(fertility_prs, by = c("f.eid" = "FID")) %>% 
+               rename(fertility_prs = SCORE) %>%
+               mutate(fertility_prs = c(scale(fertility_prs)))
+  famhist
+}

### 2021-07-16 regressions.R

@@ -216,3 +216,23 @@ run_income_dist <- function (famhist) {
   #   - use within-family regressions, then work out the intercept separately
   # - predicted from regression, weighting multiplied by n_children
 }
+
+run_reg_fe_fertility <- function (famhist, score_names) {
+  fmls <- paste("n_children ~", score_names, "| sib_group + YOB + sex")
+  fmls <- purrr::map(fmls, as.formula)
+  mods <- purrr::map(fmls, fixest::feols, data = famhist, 
+                       subset = famhist$kids_ss & ! is.na(famhist$first_job_pay), 
+                       note = FALSE)
+  tidied_raw <- purrr::map_dfr(mods, tidy, conf.int = TRUE)
+
+  fmls <- paste("n_children ~", score_names, "+ age_fte_cat + first_job_pay | sib_group + YOB + sex")
+  fmls <- purrr::map(fmls, as.formula)
+  mods <- purrr::map(fmls, fixest::feols, data = famhist, subset = famhist$kids_ss, 
+                       note = FALSE)
+  tidied_mediators <- purrr::map_dfr(mods, tidy, conf.int = TRUE)
+  tidied_mediators %<>% dplyr::filter(! grepl("age_fte_cat|first_job_pay", term))
+  
+  
+  tidied_all <- bind_rows(Raw = tidied_raw, Controlled = tidied_mediators, 
+                            .id = "reg.type")
+}

### 2021-07-22 why-natural-selection.Rmd
Report results from within-siblings regressions

@@
 
-```{r calc-res-earn-educ}
-
-drake::loadd(res_ee_control)
-drake::loadd(res_all)
-
-res_ee_control %<>% filter(term == score_name)
-n_05 <- sum(res_ee_control$p.value < 0.05)
-n_05_33 <- sum(res_ee_control$p.value < 0.05/33)
-
-```
-
 Second, we re-estimate equation \@ref(eq:regression) for each polygenic score,
@@
 
-* We estimate the causal effect of each PGS on fertility using within-sibling
-  regressions.
-  - These are causal due to lottery of meiosis
-  - Also do the fertility PGS
-
-* We also estimate the effect of PSEA and fertility PGS on education and
-  earnings, again within-siblings
-  - confirm PSEA affects both; does fertility? What if we put both in together?
-* Lastly we again estimate effect of PSEAs on fertility, 
-  controlling for education and earnings in first job. 
-  - We cannot exclude some reverse causality from EA3 -> fertility -> education/income
-  - Our theory indeed assumes that people simultaneously choose N and s, while taking into
-    account their effect on Y. We don't make claims about exogenous (chance) variation
-    in Y or in N. But note that exog variation in N1 (i.e. no choice, treat
-    it as fixed to different values) would affect s. OTOH, if EA3 affects $a$ and
-    thereby changes things, you don't get predictions 2 or 4.
+```{r calc-fe-fertility}
+drake::loadd(res_fe_fertility)
+drake::loadd(res_all)
+
+res_all %<>% filter(dep.var == "N children", reg.type == "controlled")
+res_fe_fertility %<>% filter(Regression == "Raw")
+res_all %<>% left_join(res_fe_fertility, by = "term")
+
+cor_pool_within <- cor.test(res_all$estimate.x, res_all$estimate.y)
+```
+
+One interpretation of these results is that natural selection on polygenic
+scores is driven by their correlation with human capital, which affects
+fertility as our model describes. This does not necessarily imply that polygenic
+scores are causal of either human capital or fertility. Instead they may simply
+correlate with different family environments, for example via "genetic nurture"
+effects [@kong2018nature] or population stratification. To test whether
+polygenic scores cause fertility, we estimate equation \@ref(eq:regression)
+among full siblings, controlling for sibling group fixed effects (Appendix
+Figure \@ref(fig:plot-res-fe-fertility)). Results are never significant after
+Bonferroni correction. The correlation between pooled and within-sibling effect
+sizes is `r cor_pool_within$estimate` (p = `r cor_pool_within$p.value`), and within-sibling effects are about 1/3 smaller. Thus, there is only weak evidence for causal
+links from polygenic scores to fertility, with at least some of the effect
+coming from gene-environment correlation. If we re-estimate the within-sibling
+regressions adding a control for education, effect sizes barely change. This
+suggests that our measures of education only mediate between-sibling-group
+differences in fertility; among siblings, other causal channels are more
+important.
 
@@
     have smaller income
+  - within-sib and pooled coefs for PGS are highly correlated; within coefs
+    are about 67% of the size.
+
+* Interpretation:
+  - some families are better educated and have fewer children
+  - they also have selected-against PG scores like what we see
+  - scores aren't necessarily causal for children, they may be
+    associated
+  - the between-family effect goes through education, but not the 
+    within-family effect (cf. the paper on causality)
 
@@
 \clearpage
+## Within-siblings regressions
+
+```{r plot-res-fe-fertility, fig.cap = "Selection effects controlling for sibling-group fixed effects.
+\"Controlled\" regressions add a control for education (left education before 16, 16-18, or after 18)."}
+
+drake::loadd(res_fe_fertility)
+
+standard_ggplot(res_fe_fertility, fill_col = Regression, n_regs = 33, fill_direction = -1)
+
+```
+
+\FloatBarrier
+\clearpage
 ## Genetic correlations with EA3

### 2021-07-22 _drake.R

@@ -587,7 +587,7 @@ plan <- drake_plan(
                               ) 
   },
   
-  # res_fe_fertility = run_reg_fe_fertility(famhist, score_names),
+  res_fe_fertility = run_reg_fe_fertility(famhist, score_names),
   
   # res_margins = {
   #    res_extensive <- map_dfr(score_names, 

### 2021-07-22 regressions.R

@@ -218,21 +218,21 @@ run_income_dist <- function (famhist) {
 }
 
 run_reg_fe_fertility <- function (famhist, score_names) {
-  fmls <- paste("n_children ~", score_names, "| sib_group + YOB + sex")
+  fmls <- paste("n_children ~", score_names, "| sib_group")
   fmls <- purrr::map(fmls, as.formula)
   mods <- purrr::map(fmls, fixest::feols, data = famhist, 
-                       subset = famhist$kids_ss & ! is.na(famhist$first_job_pay), 
+                       subset = famhist$kids_ss, 
                        note = FALSE)
   tidied_raw <- purrr::map_dfr(mods, tidy, conf.int = TRUE)
 
-  fmls <- paste("n_children ~", score_names, "+ age_fte_cat + first_job_pay | sib_group + YOB + sex")
+  fmls <- paste("n_children ~", score_names, "+ age_fte_cat | sib_group")
   fmls <- purrr::map(fmls, as.formula)
   mods <- purrr::map(fmls, fixest::feols, data = famhist, subset = famhist$kids_ss, 
                        note = FALSE)
   tidied_mediators <- purrr::map_dfr(mods, tidy, conf.int = TRUE)
-  tidied_mediators %<>% dplyr::filter(! grepl("age_fte_cat|first_job_pay", term))
+  tidied_mediators %<>% dplyr::filter(! grepl("age_fte_cat", term))
   
   
   tidied_all <- bind_rows(Raw = tidied_raw, Controlled = tidied_mediators, 
-                            .id = "reg.type")
+                            .id = "Regression")
 }

### 2021-07-23 why-natural-selection.Rmd
Describe within-sibling results.

@@
 
+* Only use education, not first job pay, in "controlled" regressions; check
+  description is correct - done.
 * Add CIs - done.
@@
   - Comment on results where effects are actually reversed. This works
-    nicely if you throw in a money cost. Done computational example.
-* Clarify causality?
+    nicely if you throw in a money cost - done computational example.
+* Clarify causality? - done...
   - Helps to deal with the "collider bias" argument of reviewer
@@
     income is indeed the mediator.
-* Control for PGS for fertility
 * Clarify "what we learn about genetic links to fertility by using economic theory to interpret relationships between polygenic scores and fertility."
@@
   end of formal section.
+  - think our theory with 2 periods probably covers this
 * Why don't women see stronger effects? But note, they do: not significant
@@
 Second, we re-estimate equation \@ref(eq:regression) for each polygenic score,
-controlling for earnings in first job and education levels. Effect sizes are
-generally reduced (Appendix Figure \@ref(fig:plot-res-earn-educ)), with
+controlling for education levels. Effect sizes are
+generally reduced (Appendix Figure \@ref(fig:plot-res-educ)), with
 large reductions for polygenic scores for educational attainment. This 
-suggests that earnings and education indeed mediate the effect of 
-polygenic scores on fertility.
+suggests that education indeed mediates the effect of polygenic scores on fertility.
 
@@
 
-cor_pool_within <- cor.test(res_all$estimate.x, res_all$estimate.y)
+lm_pool_w <- tidy(lm(estimate.y ~ estimate.x, res_all))
+coef_pool_w <- lm_pool_w[[2, "estimate"]]
+se_pool_w <- lm_pool_w[[2, "std.error"]]
 ```
@@
 polygenic scores cause fertility, we estimate equation \@ref(eq:regression)
-among full siblings, controlling for sibling group fixed effects (Appendix
-Figure \@ref(fig:plot-res-fe-fertility)). Results are never significant after
-Bonferroni correction. The correlation between pooled and within-sibling effect
-sizes is `r cor_pool_within$estimate` (p = `r cor_pool_within$p.value`), and within-sibling effects are about 1/3 smaller. Thus, there is only weak evidence for causal
-links from polygenic scores to fertility, with at least some of the effect
-coming from gene-environment correlation. If we re-estimate the within-sibling
-regressions adding a control for education, effect sizes barely change. This
-suggests that our measures of education only mediate between-sibling-group
-differences in fertility; among siblings, other causal channels are more
+among full siblings, controlling for a full set of sibling group dummies
+(Appendix Figure \@ref(fig:plot-res-fe-fertility)). All within-sibling effects
+are insignificant after Bonferroni correction. Effect sizes are positively
+correlated with effect sizes from the pooled model, and about 1/3 smaller
+(regressing within-sibling on pooled effect sizes, b = `r coef_pool_w`, se = 
+`r se_pool_w` ). Thus, there is only weak evidence that polygenic scores are
+causally linked to fertility, and selection effects are at least partially
+driven by gene-environment correlation. Also, if we add a control for education 
+to the within-sibling regressions, effect sizes barely change. This
+suggests that our measure of education only mediates differences in fertility
+between sibling groups; among siblings, other causal channels are more
 important.
-
-* Results:
-  - in bivariate within-sib regs, extraversion, EA2/3, ADHD and some others are
-    weakly significant (only at .05!)
-  - adding education as a control barely changes effect sizes
-  - adding income as a control increases effect sizes (of EA at least)
-  - by contrast, in pooled regs, adding education shrinks effect sizes, and
-    adding income does nothing.
-  - relevant: effect of EA2 on education is halved within-sib, and effect of
-    education on fertility is halved within-sib
-  - whereas, effect of EA2 on income is halved within-sib, but effect of income
-    on fertility is multiplied by 10!! within-sib. Much of this is because
-    of year of birth confound i.e. declining fertility and also retired people
-    have smaller income
-  - within-sib and pooled coefs for PGS are highly correlated; within coefs
-    are about 67% of the size.
-
-* Interpretation:
-  - some families are better educated and have fewer children
-  - they also have selected-against PG scores like what we see
-  - scores aren't necessarily causal for children, they may be
-    associated
-  - the between-family effect goes through education, but not the 
-    within-family effect (cf. the paper on causality)
 
@@
 predict the overall number of children, highly significantly ($p < 2 \times
-10^{-16}$, `r n_risk_sig`/33 regressions). Given that, and the statistical power
+10^{-16}$ in `r n_risk_sig`/33 regressions). Given that, and the statistical power
 we get from our sample size, we believe that the non-result is real: while risk
@@
 
-```{r plot-res-earn-educ, fig.cap = "Selection effects controlling for earnings in first job (estimated by mean earnings from ASHE 2007 for the SOC2000 job code) and education (left education before 16, 16-18, or after 18).
-Raw effects are shown for comparison."}
+```{r plot-res-educ, fig.cap = "Selection effects controlling for education (left education before 16, 16-18, or after 18). Raw effects are shown for comparison."}
 
@@
 
-```{r plot-res-fe-fertility, fig.cap = "Selection effects controlling for sibling-group fixed effects.
-\"Controlled\" regressions add a control for education (left education before 16, 16-18, or after 18)."}
+```{r plot-res-fe-fertility, fig.cap = "Selection effects controlling for sibling-group fixed effects, with and without a control for education (left education before 16, 16-18, or after 18)."}
 

### 2021-07-23 _drake.R

@@ -540,10 +540,9 @@ plan <- drake_plan(
   },
   
   res_ee_control = {
-    famhist <- add_ashe_income(famhist, ashe_income)
     map_dfr(score_names, 
               ~ run_regs_fml(
-                fml = "n_children ~ {score_name} + age_fte_cat + factor(income_cat)",
+                fml = "n_children ~ {score_name} + age_fte_cat",
                 score_name = .x,
                 famhist    = famhist,
                 subset     = quote(kids_ss)

### 2021-07-26 why-natural-selection.Rmd
speed up res_income by not repeatedly filtering on kids_ss; tweak explanation of within-siblings results

@@
 among full siblings, controlling for a full set of sibling group dummies
-(Appendix Figure \@ref(fig:plot-res-fe-fertility)). All within-sibling effects
-are insignificant after Bonferroni correction. Effect sizes are positively
-correlated with effect sizes from the pooled model, and about 1/3 smaller
-(regressing within-sibling on pooled effect sizes, b = `r coef_pool_w`, se = 
-`r se_pool_w` ). Thus, there is only weak evidence that polygenic scores are
-causally linked to fertility, and selection effects are at least partially
-driven by gene-environment correlation. Also, if we add a control for education 
-to the within-sibling regressions, effect sizes barely change. This
-suggests that our measure of education only mediates differences in fertility
-between sibling groups; among siblings, other causal channels are more
-important.
+(Appendix Figure \@ref(fig:plot-res-fe-fertility)). On the one hand, all 
+within-sibling effects are insignificant after Bonferroni correction. On the
+other hand, effect sizes are positively correlated with effect sizes from the
+pooled model, and about 1/3 smaller (regressing within-sibling on pooled effect
+sizes, b = `r coef_pool_w`, se = `r se_pool_w` ). We view this as providing weak
+evidence that polygenic scores are causally linked to fertility, with at least
+some role for correlated environments. Also, if we add a control for education
+to the within-sibling regressions, effect sizes barely change. This suggests
+that our measure of education only mediates differences in fertility between
+sibling groups; among siblings, other causal channels are more important.
 

### 2021-07-26 _drake.R

@@ -504,15 +504,17 @@ plan <- drake_plan(
           ) 
     # keep variables in this order (it matters to .id):
     pars <- expand_grid(subset = subsets, score_name = score_names) 
+    fh_tmp <- famhist %>% filter(kids_ss)
     res_income <- pmap_dfr(pars,
             ~ run_regs_fml(
               "n_children ~ {score_name}", 
               score_name = .y, 
               subset     = .x,
-              famhist    = famhist %>% filter(kids_ss)
+              famhist    = fh_tmp
             ),
             .id = "row_number"
           ) 
+    rm(fh_tmp)
     pars$row_number <- as.character(seq_len(nrow(pars)))
     res_income %>% left_join(pars, by = "row_number") %>% 
           filter(term != "(Intercept)") %>% 

### 2021-08-09 why-natural-selection.Rmd
Final tweaks pre-NHB submission

--- 2021-07-23 why-natural-selection.Rmd
@@
     relationship is weaker (closer to zero)           Selection effects are also weaker at higher
-    at higher wages and/or levels of                  polygenic scores for EA (Appendix Figure
-    human capital.                                    \@ref(fig:plot-purifying)).
+    at higher wages and/or levels of                  polygenic scores for 
+    human capital.                                    educational attainment (Appendix Figure
+                                                      \@ref(fig:plot-purifying)).
 
@@
 
-One interpretation of these results is that natural selection on polygenic
-scores is driven by their correlation with human capital, which affects
-fertility as our model describes. This does not necessarily imply that polygenic
-scores are causal of either human capital or fertility. Instead they may simply
-correlate with different family environments, for example via "genetic nurture"
-effects [@kong2018nature] or population stratification. To test whether
-polygenic scores cause fertility, we estimate equation \@ref(eq:regression)
-among full siblings, controlling for a full set of sibling group dummies
-(Appendix Figure \@ref(fig:plot-res-fe-fertility)). All within-sibling effects
-are insignificant after Bonferroni correction. Effect sizes are positively
-correlated with effect sizes from the pooled model, and about 1/3 smaller
-(regressing within-sibling on pooled effect sizes, b = `r coef_pool_w`, se = 
-`r se_pool_w` ). Thus, there is only weak evidence that polygenic scores are
-causally linked to fertility, and selection effects are at least partially
-driven by gene-environment correlation. Also, if we add a control for education 
-to the within-sibling regressions, effect sizes barely change. This
-suggests that our measure of education only mediates differences in fertility
-between sibling groups; among siblings, other causal channels are more
-important.
+These results support our theory that natural selection on polygenic scores is
+driven by their correlation with human capital. This need not imply
+that a given polygenic score is causal of either fertility or human capital. Instead
+scores may simply correlate with causally relevant environments, for example via
+"genetic nurture" effects [@kong2018nature] or population stratification. To
+test whether polygenic scores cause fertility, we estimate equation
+\@ref(eq:regression) among full siblings, controlling for a full set of sibling
+group dummies (Appendix Figure \@ref(fig:plot-res-fe-fertility)). With a reduced
+sample size, all within-sibling effects are insignificant after Bonferroni
+correction. However, effect sizes are positively correlated with effect sizes
+from the pooled model, and about 1/3 smaller (regressing within-sibling on
+pooled effect sizes, $b$ = `r coef_pool_w`). This attenuation is broadly in line
+with the ~50% decrease in heritability in within-sibling GWASs on age at first
+birth and educational attainment [@Howe2021.03.05.433935]. We see these results
+as providing weak evidence that polygenic scores cause fertility, with effects
+being at least partly driven by gene-environment correlations. We also reran
+within-siblings regressions adding a control for education. Most effect sizes
+barely change, suggesting that our measure of education does not in general
+mediate differences in fertility among siblings. However, for educational
+attainment polygenic scores (EA2 and EA3), effect sizes are reduced by about 20%.
 
@@
 for this: scores which might plausibly be linked to externalizing behaviour,
-like ADHD and (younger) age at smoking initiation, are selected for. However,
+like ADHD and younger age at smoking initiation, are selected for. However,
 this theory is less good at explaining variation in selection across the full
@@
 predict the overall number of children, highly significantly ($p < 2 \times
-10^{-16}$ in `r n_risk_sig`/33 regressions). Given that, and the statistical power
+10^{-16}$ in `r n_risk_sig` out of 33 regressions). Given that, and the statistical power
 we get from our sample size, we believe that the non-result is real: while risk
@@
 effects, gene-environment correlations, and/or assortative mating 
-[@lee2018gene; @selzam2019comparing; @kong2018nature]. 
+[@lee2018gene; @selzam2019comparing; @kong2018nature; @Howe2021.03.05.433935]. 
 In short, it is probably too early to tell whether modern natural selection has a
@@
 important for understanding class mobility and structure [@belsky2018genetic;
-@Rimfeld_2018], then these increases are substantial effects. Since many
+@Rimfeld_2018], then these increases are substantially important. Since many
 polygenic scores are predictive of disease risk, they could also potentially

### 2021-08-10 model-appendix.Rmd
Final final tweaks including new end of intro

@@
 
-We can now gather the facts stated in the main text.
+We can now gather the theoretical predictions stated in Table \ref{tab:theory}.
 
-\textbf{Fact 1}: for $\sigma<1$, total fertility $N^{*}\equiv N_{1}^{*}+N_{2}^{*}$
+\textbf{Prediction 1}: for $\sigma<1$, total fertility $N^{*}\equiv N_{1}^{*}+N_{2}^{*}$
 is decreasing in human capital $h$.
@@
 
-\textbf{Fact 2 part 1}: the fertility-human capital relationship is
+\textbf{Prediction 2 part 1}: the fertility-human capital relationship is
 closer to 0 at high levels of $h$.
@@
 
-\textbf{Fact 2 part 2}: for $\sigma<1$ and close to 1, the fertility-human
+\textbf{Prediction 2 part 2}: for $\sigma<1$ and close to 1, the fertility-human
 capital relationship is weaker among higher earners.
 
-\textbf{Fact 4}: for $\sigma<1$ and close to 1, the fertility-human
+\textbf{Prediction 4}: for $\sigma<1$ and close to 1, the fertility-human
 capital relationship is weaker at high levels of education.
@@
 
-\textbf{Fact 5}: for $\sigma<1$ and close to 1, the fertility-human
+\textbf{Prediction 5}: for $\sigma<1$ and close to 1, the fertility-human
 capital relationship is weaker among those who start fertility late.
 
-Lastly, we prove fact 3. Differentiating $dN_{t}^{*}/dh$ in (\ref{eq:dNdh1})
+Lastly, we prove prediction 3. Differentiating $dN_{t}^{*}/dh$ in (\ref{eq:dNdh1})
 with respect to $b$, for when $N_{1}^{*}>0$ gives:
@@
 
-\textbf{Fact 3}: for $\sigma<1$, the fertility-human capital relationship
+\textbf{Prediction 3}: for $\sigma<1$, the fertility-human capital relationship
 is more negative when the burden of childcare $b$ is larger.

### 2021-08-10 why-natural-selection.Rmd
Final final tweaks including new end of intro

@@
 The direction of natural selection is reversed among older parents (22+), or after controlling for age at first live birth.
-These patterns are in line with economic theories of fertility, in which higher earnings may either increase or decrease fertility via income and substitution effects in the labour market.
+These patterns are in line with economic theories of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market.
 Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
@@
 makes children more affordable), and a fertility-lowering *substitution effect*
-(time spent on childrearing has a higher cost in foregone earnings). Our
-results suggest that the substitution effect dominates for single parents and
-younger parents, while among couples and older parents (22+) effects are more
-evenly balanced.
+(time spent on childrearing has a higher cost in foregone earnings). Our results
+suggest that the substitution effect dominates for single parents and younger
+parents, while among couples and older parents (22+) effects are more evenly
+balanced. Thus, contemporary natural selection on polygenic scores can be
+explained by their correlation with earnings-increasing human capital. Below, we 
+show that a simple model of human capital, education and fertility choices 
+can give rise to our key empirical results.
 
@@
 qualification; and (3) for women only, age, highest qualification, and age at 
-first live birth. Figure \@ref(fig:plot-res-wt) plots selection effects among the
+first live birth (AFLB). Figure \@ref(fig:plot-res-wt) plots selection effects among the
 entire sample, estimated with the three weighting schemes, and also with uniform
@@
 Suppose that certain genetic variants correlate with human capital -- skills or
-other characteristics that affect an individual's earnings in the labour market.
-These variants may then be associated with opposing effects on fertility. The income
-effect will lead to natural selection in favour of earnings-increasing variants
-(or variants that are merely associated with higher earnings). The substitution
-effect will do the reverse.
+other characteristics that affect an individual's earnings in the labour market
+[@mincer1958investment; @becker1964human]. These variants may then be associated
+with opposing effects on fertility. The income effect will lead to natural
+selection in favour of earnings-increasing variants (or variants that are merely
+associated with higher earnings). The substitution effect will do the reverse.
 
@@
 in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
-*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$,
-and in a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical 
+*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$
+in a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical 
 predictions, along with corresponding empirical results for the correlation between
@@
 "genetic nurture" effects [@kong2018nature] or population stratification. To 
-test whether polygenic scores cause fertility, we estimate equation
+investigate the presence of these indirect genetic effects, we estimate equation
 \@ref(eq:regression) among full siblings, controlling for a full set of sibling
@@
 mediate differences in fertility among siblings. However, for educational
-attainment polygenic scores (EA2 and EA3), effect sizes are reduced by about 20%.
+attainment polygenic scores, effect sizes are reduced by about 20%.
 
@@
 Blue is actual.
-Grey is hypothetical in the absence of selection effects.", fig.width = 4}
+Grey is hypothetical in the absence of selection effects.
+Respondents weighted by Age/Qualification.", fig.width = 4}
 income_EA3 <- famhist %>% 

### 2021-08-10 why-natural-selection.Rmd
add line numbering

@@
   - \providecommand{\lemmaname}{Lemma}
+  - \usepackage{lineno}
+  - \linenumbers
 editor_options: 

### 2021-08-10 why-natural-selection.Rmd
BLIND MANUSCRIPT

@@
 title: "Human capital mediates natural selection in contemporary humans"
-author: "David Hugh-Jones \\thanks{Corresponding author.
-School of Economics, University of East Anglia, Norwich, UK.
-Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands.}"
 date: "`r format(Sys.Date(), '%d %B %Y')`"
@@
 participants of European descent, as defined by genetic estimated ancestry and
-self-identified ethnic group [@abdellaoui2019genetic], giving a sample size of
+self-identified ethnic group, giving a sample size of
 `r comma_num(nrow(famhist))`. For regressions on number of children we use
@@
 stratification, we corrected the polygenic scores for 100 principal components (PCs).
-To compute polygenic scores and PCs, the same procedures were followed as 
-described in @abdellaoui2019genetic.
 
@@
 Super Output Areas, sex and presence/absence of a partner.
-
-# Acknowledgements
-
-AA is supported by the Foundation Volksbond Rotterdam and by ZonMw grant 
-849200011 from The Netherlands Organisation for Health Research and Development. 
-This study was conducted using UK Biobank resources under application numbers 
-40310 and 19127.
 

### 2021-12-01 why-natural-selection.Rmd
Focus on motivating the theory as a reason for sample-splitting. Reorder intro to put theory first.

@@
 
-As yet we know little about the social mechanisms behind these effects. This
-study uses data from UK Biobank [@bycroft2018uk] to learn more. We test for
+As yet we know little about the social mechanisms behind these effects. A
+potential explanation for them comes from the economic theory of fertility first
+presented by [@becker1960economic]. In this theory, higher potential earnings
+have two opposite effects on fertility: a fertility-increasing *income effect*
+(higher income makes children more affordable), and a fertility-lowering
+*substitution effect* (time spent on childrearing has a higher cost in foregone
+earnings). Thus, an individual's *human capital* -- skills and personality
+traits which are valuable in labour markets -- can increase or decrease his or
+her fertility. This can lead to selection on inherited
+variants which are linked to human capital. Furthermore, the economic theory predicts
+systematic variation in the relative strength of the income and substitution
+effects across different social groups.
+
+This
+study uses data from UK Biobank [@bycroft2018uk] to learn more about contemporary
+natural selection. We test for
 natural selection on 33 different polygenic scores by estimating their
@@
 This is interesting because consistent natural selection over multiple
-generations could lead to substantive effects in the long run. Most importantly,
-we examine reproductive rates in different subgroups of the population, in order
-to uncover patterns that can help illuminate the mechanisms behind modern
-natural selection.
-
+generations could lead to substantive effects in the long run. We then examine
+reproductive rates in different subgroups of the population. 
 We find selection effects on many polygenic scores. Effects are largely consistent
-across generations. The strength of natural selection on a polygenic score is
-associated with that score's correlation with education and earnings: scores
-that predict lower education and earnings are being selected for. Also, across
-the board, polygenic scores have stronger relationships with fertility among
-specific subgroups. Selection effects are stronger in groups with lower
+across generations.
+
+Next, we examine correlations with fertility in different subgroups. Across
+the board, selection effects are stronger in groups with lower
 income and less education, among younger parents, people not living with a partner,
@@
 weaker and often statistically insignificant. In some subgroups, the
-direction of selection is even reversed: polygenic scores predicting
-higher education and earnings are associated with *higher* fertility.
-
-These patterns are in line with economic theories of fertility
-[@becker1960economic]. In these, higher potential earnings have two opposite
-effects on fertility: a fertility-increasing *income effect* (higher income
-makes children more affordable), and a fertility-lowering *substitution effect*
-(time spent on childrearing has a higher cost in foregone earnings). Our results
-suggest that the substitution effect dominates for single parents and younger
-parents, while among couples and older parents (22+) effects are more evenly
-balanced. Thus, contemporary natural selection on polygenic scores can be
-explained by their correlation with earnings-increasing human capital. Below, we 
-show that a simple model of human capital, education and fertility choices 
-can give rise to our key empirical results.
+direction of selection is even reversed.
+
+We then show that a parsimonious model of human capital, education and fertility
+choices can give rise to these empirical results. To test the theory, we show
+that polygenic scores' correlation with fertility is associated with their
+correlation with education and earnings. Furthermore, controlling for education
+and earnings reduces the correlation with fertility, suggesting that human
+capital mediates natural selection. Thus, contemporary natural selection on
+polygenic scores can be explained by their correlation with earnings-increasing
+human capital.
 
@@
 
-To understand the reasons for the ascertainment bias in our sample, and to learn 
-about the mechanisms underlying natural selection, we split the sample, starting with
-basic demographic variables including education, income and sex. These are all
-potential sources of ascertainment bias, as mentioned above. We then re-estimate
-\@ref(eq:regression) within each subgroup.
-
-We emphasize that this exercise does
-*not* apportion the total correlation with fertility into subgroups. 
-That is because the polygenic score may also affect the probability of being in 
-each subgroup. For example, the EA3 polygenic score for educational attainment not only affects
-fertility among university graduates, and among non-graduates; it also changes
-a person's chance of attending university. Nevertheless, differences in equation
-\@ref(eq:regression) across subgroups can inform us about how natural
-selection works, by constraining the set of possible mechanisms.
+To motivate our economic model, we split the sample by demographic
+and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
+We examine socio-demographic variables including income, education and sex, and
+family structure variables including age at first live birth, presence of a partner, and
+lifetime number of sexual partners.
+We emphasize that these categories are not necessarily exogenous to the polygenic
+scores. For example, both in the data and in our theoretical model, age at first
+live birth is a choice variable, which is endogenous to human-capital-related 
+polygenic scores. Nevertheless, differences in equation
+\@ref(eq:regression) across subgroups constrain the set of possible explanations
+for natural selection. As we show below, the economic theory predicts 
+the patterns in the data, while alternative explanations do not.
 
@@
 
-We rerun equation \@ref(eq:regression), adding a quadratic term in $PGS_i$,
+Balancing selection reduces variance in the trait under selection, 
+while diversifying selection increases variance. To check for these,
+we rerun equation \@ref(eq:regression), adding a quadratic term in $PGS_i$,
 and using our age/qualification weights. Scores for hip circumference and drinks

### 2021-12-01 why-natural-selection.Rmd
Focus on motivating the theory as a reason for sample-splitting. Reorder intro to put theory first.

@@
 reproductive rates in different subgroups of the population.
+
 We find selection effects on many polygenic scores. Effects are largely consistent
-across generations.
-
-Next, we examine correlations with fertility in different subgroups. Across
-the board, selection effects are stronger in groups with lower
+across generations. The strength of natural selection on a polygenic score is
+associated with that score's correlation with education and earnings: scores
+that predict lower education and earnings are being selected for. Also, across
+the board, polygenic scores have stronger relationships with fertility among
+specific subgroups. Selection effects are stronger in groups with lower
 income and less education, among younger parents, people not living with a partner,
@@
 weaker and often statistically insignificant. In some subgroups, the
-direction of selection is even reversed.
-
-We then show that a parsimonious model of human capital, education and fertility
-choices can give rise to these empirical results. To test the theory, we show
-that polygenic scores' correlation with fertility is associated with their
-correlation with education and earnings. Furthermore, controlling for education
-and earnings reduces the correlation with fertility, suggesting that human
-capital mediates natural selection. Thus, contemporary natural selection on
-polygenic scores can be explained by their correlation with earnings-increasing
-human capital.
+direction of selection is even reversed: polygenic scores predicting
+higher education and earnings are associated with *higher* fertility.
+
+We then show that a parsimonious model of human capital, education and
+fertility choices can give rise to these empirical results. Thus, contemporary
+natural selection on polygenic scores can be explained by their correlation with
+earnings-increasing human capital.
 

### 2021-12-08 why-natural-selection.Rmd
Tweaks in response to NHB reviews. Rewrite comments on alternative theories.

--- 2021-12-01 why-natural-selection.Rmd
@@
 
-This
-study uses data from UK Biobank [@bycroft2018uk] to learn more about contemporary
-natural selection. We test for
-natural selection on 33 different polygenic scores by estimating their
-correlation with fertility. We extend the analysis over two generations, using
-data on respondents' number of siblings as well as their number of children.
-This is interesting because consistent natural selection over multiple
-generations could lead to substantive effects in the long run. We then examine
-reproductive rates in different subgroups of the population. 
-We find selection effects on many polygenic scores. Effects are largely consistent
-across generations.
+This study uses data from UK Biobank [@bycroft2018uk] to learn more about
+contemporary natural selection. We test for natural selection on 33 different
+polygenic scores by estimating their correlation with fertility. We extend the
+analysis over two generations, using data on respondents' number of siblings as
+well as their number of children. This is interesting because consistent natural
+selection over multiple generations could lead to substantive effects in the
+long run. We then examine reproductive rates in different subgroups of the
+population. We find selection effects on many polygenic scores. Effects are
+largely consistent across generations.
 
@@
 We created polygenic scores for 33 traits in `r comma_num(nrow(famhist))`
-individuals, corrected for ancestry using 100 genetic principal components (see
-Materials and Methods). Figure \@ref(fig:plot-means-over-time) plots mean
-polygenic scores in the sample by 5-year birth intervals. Several scores show
-consistent increases or declines over this 30-year period, of the order of 5% of
-a standard deviation. These changes could reflect natural selection within the
-UK population, but also emigration, or ascertainment bias within the sample.
-Respondents have higher income and are better educated than the UK population,
-and they may also differ on other unobserved characteristics
-[@10.1093/aje/kwx246]. Since richer and more educated people also live longer, this
-bias could also increase with age. Lastly, the sample sex ratio skews 
-`r female_ratio * 100`% female.
+individuals of European descent, corrected for ancestry using 100 genetic
+principal components (see Materials and Methods). Figure
+\@ref(fig:plot-means-over-time) plots mean polygenic scores in the sample by
+5-year birth intervals. Several scores show consistent increases or declines
+over this 30-year period, of the order of 5% of a standard deviation. These
+changes could reflect natural selection within the UK population, but also
+emigration, or ascertainment bias within the sample. Respondents have higher
+income and are better educated than the UK population, and they may also differ
+on other unobserved characteristics [@10.1093/aje/kwx246]. Since richer and more
+educated people also live longer, this bias could also increase with age.
+Lastly, the sample sex ratio skews `r female_ratio * 100`% female.
 
@@
 (age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
-Estimates might be further affected by weighting on other demographic variables.
+Estimates might be further affected by weighting on other demographic variables.[^weighting]
+
+[^weighting]: To make interpretation straightforward, we do not use weights in 
+results below unless noted. However, all our results are robust to using weights.
 
 We also check for balancing and diversifying selection by estimating
-\@ref(eq:regression) with a quadratic term. In particular, we find diversifying
-selection for educational attainment polygenic scores: at higher values of these
-scores, the negative effect on fertility is smaller (Appendix Figure
-\@ref(fig:plot-purifying)).
+\@ref(eq:regression) with a quadratic term.[^balance-diverse-def] In particular,
+we find diversifying selection for educational attainment polygenic scores: at
+higher values of these scores, the negative effect on fertility is smaller
+(Appendix Figure \@ref(fig:plot-purifying)).
+
+[^balance-diverse-def]: Balancing selection selects for intermediate values, 
+reducing the variance of the trait under selection. Diversifying selection
+selects for extreme values and increases the variance.
 
@@
 for natural selection. As we show below, the economic theory predicts 
-the patterns in the data, while alternative explanations do not.
+these differences, while alternative explanations do not.
 
@@
 his or her wage $W$. Raising a child takes time $b$. People maximize utility
-from the number of children $N$, and income $Y\equiv(1-bN)W$. An individual's
+from the number of children $N$ and from income $Y\equiv(1-bN)W$. An individual's
 payoff is
@@
 in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
-*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$
-in a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical 
-predictions, along with corresponding empirical results for the correlation between
-polygenic scores and fertility.[^one-period-results]
-
-[^one-period-results]: Predictions 1-3 also hold in the one-period model. Our empirical 
-results are actually stronger than prediction 5, in that
-correlations with fertility are *reversed* at higher AFLB. This prediction can be
-accommodated in the model if children have a money cost as well as a time cost. 
-Appendix Figure \@ref(fig:N-plot-with-m) shows an example.
+*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$ in
+a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical
+predictions, along with corresponding empirical results for the correlation
+between polygenic scores and fertility.[^one-period-results] The key insight of
+the model is that for middling levels of risk aversion, the substitution effect
+dominates at low income levels, but is balanced by the income effect as income
+increases.
+
+[^one-period-results]: Predictions 1-3 also hold in the one-period model with
+CRRA utility. Our empirical results are actually stronger than prediction 5, in
+that correlations with fertility are *reversed* at higher AFLB. This prediction
+can be accommodated in the model if children have a money cost as well as a time
+cost. Appendix Figure \@ref(fig:N-plot-with-m) shows an example.
 
@@
 The above does not aim to provide a complete theory of fertility. Rather, it
-shows that a relatively simple economic model can explain many of our results.
+shows that a simple economic model can explain many of our results.
 Other empirical work in economics also supports our mechanism.
@@
 
-An alternative theory is that polygenic scores correlate with the motivation to
-have children (parameter $a$ in the model; cf. @jones2008fertility). However, this
-theory would not explain why our results are weaker at higher incomes and
-education levels.[^note] Another alternative is that traits under selection are
-linked to externalizing behaviour and risk-seeking. This might be partially
-captured by our risk-aversion parameter $\sigma$, which affects fertility by
-changing the marginal utility of income, but a more direct channel is risky
-sexual behaviour [@mills2021identification]. The data here provide some support
-for this: scores which might plausibly be linked to externalizing behaviour,
-like ADHD and younger age at smoking initiation, are selected for. However,
-this theory is less good at explaining variation in selection across the full
-range of scores, including physical measures such as waist-hip ratio and BMI. We
-test this theory directly by re-estimating equation \@ref(eq:regression)
-controlling for a measure of risk attitude (UK Biobank field 2040).  The median
-ratio of effect sizes between regressions with and without controls is 
-`r median(ratio)`; all scores which are significant at $p < 0.05/33$ in
-uncontrolled regressions remain so when controlling for risk attitude. This
-non-result could simply reflect the imprecision of the risk aversion measure,
-which is a single yes/no question. However, the risk aversion measure *does*
-predict the overall number of children, highly significantly ($p < 2 \times
-10^{-16}$ in `r n_risk_sig` out of 33 regressions). Given that, and the statistical power
-we get from our sample size, we believe that the non-result is real: while risk
-aversion does predict fertility in the sample, it is not an important channel
-for natural selection.
+We consider three alternative theories that might explain our results. The first
+is that welfare benefits which incentivize child-bearing might be taken up
+among low-income individuals. However, our effect sizes appear unchanged
+over a large span of twentieth-century history (tables \@ref(tab:tbl-siblings-by-period) and 
+\@ref(tab:tbl-children-by-period)), during which
+government spending on child-related benefits varied considerably
+[@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
+affect fertility [@gauthier2007impact]. A second alternative theory is that
+polygenic scores correlate with the motivation to have children, i.e. parameter $a$
+in the model [cf. @jones2008fertility]. This theory would not explain why our
+results are weaker at higher incomes and education levels.[^note] A third
+alternative is that traits under selection are linked to externalizing behaviour
+and risk-seeking. This might be partially captured by our risk-aversion
+parameter $\sigma$, which affects fertility by changing the marginal utility of
+income, but a more direct channel is risky sexual behaviour
+[@mills2021identification]. The data here provide some support for this: scores
+which might plausibly be linked to externalizing behaviour, like ADHD and
+younger age at smoking initiation, are selected for. However, this theory is
+less good at explaining variation in selection across the full range of scores,
+including physical measures such as waist-hip ratio and BMI. We test this theory
+directly by re-estimating equation \@ref(eq:regression) controlling for a
+measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
+sizes between regressions with and without controls is `r median(ratio)`; all
+scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
+so when controlling for risk attitude. This non-result could simply reflect the
+imprecision of the risk aversion measure, which is a single yes/no question.
+However, the risk aversion measure *does* predict the overall number of
+children, highly significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of
+33 regressions). Given that, and the statistical power we get from our sample
+size, we believe that the non-result is real: while risk aversion does predict
+fertility in the sample, it is not an important channel for natural selection.
 
@@
 \clearpage
-## Parents' generation
+## Parents' generation {#sec-parents}
 
@@
 childless are excluded, for `r n_smaller` out of 33 scores; the median
-percentage change is `r percent(median_pct_diff)`. The fact that childless
-people have such a strong effect on estimates makes it hard to compare total
-effect sizes across generations. In particular, since the parents' generation has
-a different distribution of numbers of children, childless people may have had 
-more or less effect in that generation.
+percentage change is `r percent(median_pct_diff)`.
 
@@
 ```
+
+The fact that childless people have such a strong effect on estimates makes it
+hard to compare total effect sizes across generations. In particular, since the
+parents' generation has a different distribution of numbers of children,
+childless people may have had more or less effect in that generation. Another
+issue is that we are in effect estimating parents' polygenic scores by the
+scores of their children. This introduces some noise into our independent
+variable, which might lead to errors-in-variables and bias coefficients towards
+zero.
 

### 2021-12-10 why-natural-selection.Rmd
Use correlations with income category, not differences between highest and lowest

@@
 
-ineq_comp <-  left_join(
-                res_ineq %>% filter(income_cat == 1),
-                res_ineq %>% filter(income_cat == 5),
-                by = "score"
-              )
-
-ineq_comp %<>% mutate(
-                 diff_children = wtd.y - wtd.x,
-                 diff_hypoth   = unw.y - unw.x,
-                 pct_change    = 100 * (diff_children - diff_hypoth)/diff_hypoth
-               )
-
-median_pct_change <- median(ineq_comp$pct_change)
-n_increases <- sum(ineq_comp$pct_change > 0)
-
+n_increases <- sum(res_ineq$ratio > 1)
+median_pct_change <- (median(res_ineq$ratio) - 1) * 100
 ```
@@
 income distribution, and this increases the differences between groups. Overall,
-natural selection increases inequality for `r n_increases` out of 33 polygenic
-scores, with a median increase of `r median_pct_change`% in the difference
-between highest and lowest income groups (Appendix Table
-\@ref(tab:tbl-inequality-effects)). If inequalities in polygenic scores are
+natural selection increases the correlation of polygenic scores with income 
+for `r n_increases` out of 33 polygenic
+scores, with a median increase in the correlation of `r median_pct_change`% 
+(Appendix Table \@ref(tab:tbl-inequality-effects)). If inequalities in polygenic scores are
 important for understanding class mobility and structure [@belsky2018genetic;
-@Rimfeld_2018], then these increases are substantially important. Since many
-polygenic scores are predictive of disease risk, they could also potentially
-increase health inequalities. In general, the evolutionary history of
-anatomically modern humans is related to disease risk [@benton2021influence];
-understanding the role of contemporary natural selection may aid in mapping the
-genetic architecture of current health disparities.
+@Rimfeld_2018; @harden2021genetic], then these increases are substantially
+important. Since many polygenic scores are predictive of disease risk, they
+could also potentially increase health inequalities. In general, the
+evolutionary history of anatomically modern humans is related to disease risk
+[@benton2021influence]; understanding the role of contemporary natural selection
+may aid in mapping the genetic architecture of current health disparities.
 
@@
 
-Table \@ref(tab:tbl-inequality-effects) estimates differences in children's mean
-polygenic scores between the highest and lowest income groups. Column "With
-selection" uses respondents' scores weighted by Age/Qualification times number
+Table \@ref(tab:tbl-inequality-effects) shows correlations between children's
+polygenic scores and household income (UKB data field 738). Column "With
+selection" uses respondents' scores weighted by Age/Qualification, times number
 of children. Column "Without selection" uses scores weighted by
-Age/Qualification only, i.e. if all couples had the same number of children.
+Age/Qualification only, i.e. if all respondents had the same number of children,
+equivalently the correlation among respondents.
 
@@
 
-ineq_comp %>% 
+res_ineq %>% 
       select(
         PGS                      = score, 
-        `Without selection` = diff_hypoth,
-        `With selection`    = diff_children, 
-        `% change`          = pct_change
+        `Cor. without selection` =cf,
+        `Cor. with selection`    = actual, 
+        `Ratio`                  = ratio
       ) %>% 
@@
       ) %>% 
-      arrange(desc(`% change`)) %>% 
+      arrange(desc(`Ratio`)) %>% 
       as_hux() %>% 
@@
       set_number_format(everywhere, 2:3, 3) %>% 
-      set_number_format(everywhere, 4, 1) %>% 
-      set_caption("Differences in polygenic scores between highest and lowest income group.") %>% 
+      set_number_format(everywhere, 4, 2) %>% 
+      set_caption("Correlations of polygenic scores with income group.") %>% 
       theme_compact() %>% 

### 2021-12-10 _drake.R

@@ -565,101 +565,9 @@ plan <- drake_plan(
             .id = "score_name")
   },
   
-  res_ineq = {
-    famhist %>% 
-          left_join(age_qual_weights, by = "f.eid") %>% 
-          filter(! is.na(n_children), ! is.na(income_cat), ! is.na(weights)) %>% 
-          group_by(income_cat) %>% 
-          mutate(
-            parent_weights = weights/n_sibs,
-            parent_weights = ifelse(is.na(parent_weights), 0, parent_weights)
-          ) %>% 
-          summarize(across(all_of(score_names), 
-                    list(
-                      unw     = ~ weighted.mean(.x, weights, na.rm = TRUE),
-                      wtd     = ~ weighted.mean(.x, weights * n_children, 
-                                              na.rm = TRUE),
-                      parents = ~ weighted.mean(.x, parent_weights, 
-                                              na.rm = TRUE)
-                    )
-          )) %>% 
-          tidyr::pivot_longer(-income_cat, 
-                                names_to = c("score", ".value"), 
-                                names_pattern = "(.*)_(.+?)$"
-                              ) 
-  },
+  res_ineq = run_cor_income(famhist, score_names, age_qual_weights),
   
   res_fe_fertility = run_reg_fe_fertility(famhist, score_names),
-  
-  # res_margins = {
-  #    res_extensive <- map_dfr(score_names, 
-  #                       ~ run_regs_fml(
-  #                         fml = "I(n_children > 0) ~ factor(sex) + {score_name}:factor(sex)",
-  #                         score_name = .x,
-  #                         famhist    = famhist,
-  #                         subset     = quote(kids_ss)
-  #                       ), 
-  #                     .id = "score_name")
-  #    res_intensive <- map_dfr(score_names, 
-  #                       ~ run_regs_fml(
-  #                         fml = "n_children ~ factor(sex) + {score_name}:factor(sex)",
-  #                         score_name = .x,
-  #                         famhist    = famhist,
-  #                         subset     = quote(kids_ss & n_children > 0)
-  #                       ), 
-  #                     .id = "score_name")
-  #    res <- bind_rows(
-  #                      extensive = res_extensive, 
-  #                      intensive = res_intensive, 
-  #                      .id = "type"
[…]

### 2021-12-10 regressions.R

@@ -12,6 +12,7 @@ suppressPackageStartupMessages({
   loadNamespace("sampleSelection")
   loadNamespace("Formula")
   loadNamespace("fixest")
+  loadNamespace("psych")
 })
 
 calc_pgs_over_time <- function (famhist, score_names) {
@@ -235,4 +236,32 @@ run_reg_fe_fertility <- function (famhist, score_names) {
   
   tidied_all <- bind_rows(Raw = tidied_raw, Controlled = tidied_mediators, 
                             .id = "Regression")
+}
+
+run_cor_income <- function (famhist, score_names, age_qual_weights) {
+  famhist <- famhist %>% 
+               left_join(age_qual_weights, by = "f.eid") %>% 
+               filter(
+                 ! is.na(n_children), 
+                 ! is.na(income_cat), 
+                 ! is.na(weights)
+               ) %>%
+               mutate(
+                 child_weights = weights * n_children
+               )
+    
+  cors_counterfactual <- purrr::map(score_names, ~{
+    psych::cor.wt(famhist[c("income_cat", .x)], w = famhist$weights)
+  })
+  cors_actual <- purrr::map(score_names, ~{
+    psych::cor.wt(famhist[c("income_cat", .x)], w = famhist$child_weights)
+  })
+  
+  cors_counterfactual <- purrr::map_dbl(cors_counterfactual, ~ .x$r[2,1])
+  cors_actual <- purrr::map_dbl(cors_actual, ~ .x$r[2,1])
+  
+  res <- tibble(score = score_names, actual = cors_actual, cf = cors_counterfactual)
+  res <- res %>% mutate(ratio = actual/cf)
+  
+  res
 }

### 2021-12-10 why-natural-selection.Rmd
Reorder income table; discuss in text

@@
 median_pct_change <- (median(res_ineq$ratio) - 1) * 100
+
+cor_ea3 <- res_ineq[res_ineq$score == "EA3_excl_23andMe_UK", ]
+cor_ea3$increase_pct <- (cor_ea3$ratio - 1) * 100
 ```
@@
 educational attainment, which lowers the mean, is stronger at the bottom of the
-income distribution, and this increases the differences between groups. Overall,
-natural selection increases the correlation of polygenic scores with income 
-for `r n_increases` out of 33 polygenic
-scores, with a median increase in the correlation of `r median_pct_change`% 
-(Appendix Table \@ref(tab:tbl-inequality-effects)). If inequalities in polygenic scores are
+income distribution, and this increases the differences between groups.
+Overall, natural selection increases the correlation of polygenic scores with
+income for `r n_increases` out of 33 polygenic scores, with a median
+proportionate increase of `r median_pct_change`% (Appendix Table
+\@ref(tab:tbl-inequality-effects)). If inequalities in polygenic scores are
 important for understanding class mobility and structure [@belsky2018genetic;
-@Rimfeld_2018; @harden2021genetic], then these increases are substantially
-important. Since many polygenic scores are predictive of disease risk, they
-could also potentially increase health inequalities. In general, the
-evolutionary history of anatomically modern humans is related to disease risk
-[@benton2021influence]; understanding the role of contemporary natural selection
-may aid in mapping the genetic architecture of current health disparities.
+@Rimfeld_2018; @harden2021genetic], then these increases are substantive. Since
+many polygenic scores are predictive of disease risk, they could also
+potentially increase health inequalities. In general, the evolutionary history
+of anatomically modern humans is related to disease risk [@benton2021influence];
+understanding the role of contemporary natural selection may aid in mapping the
+genetic architecture of current health disparities.
 
@@
 of children. Column "Without selection" uses scores weighted by
-Age/Qualification only, i.e. if all respondents had the same number of children,
-equivalently the correlation among respondents.
+Age/Qualification only, i.e. the counterfactual correlation if all respondents
+had the same number of children (equivalently, the correlation in the
+respondents' own generation).
 
@@
         PGS                      = score, 
-        `Cor. without selection` =cf,
         `Cor. with selection`    = actual, 
+        `Cor. without selection` = cf,
         `Ratio`                  = ratio
@@
       ) %>% 
-      arrange(desc(`Ratio`)) %>% 
+      arrange(desc(`Cor. with selection`)) %>% 
       as_hux() %>% 

### 2021-12-14 why-natural-selection.Rmd
Add analysis of natural selection's effect on true PSEA's correlation with income; tweak intro; add function to do mediation analysis using 1973 ROSLA as an exogenous shift to education

@@
 library(huxtable)
+loadNamespace("scales")
 
@@
 selection over multiple generations could lead to substantive effects in the
-long run. We then examine reproductive rates in different subgroups of the
-population. We find selection effects on many polygenic scores. Effects are
-largely consistent across generations.
-
-Next, we examine correlations with fertility in different subgroups. Across
+long run. Next, we examine correlations with fertility in different subgroups. Across
 the board, selection effects are stronger in groups with lower
@@
 human capital.
+
+Lastly, we discuss the effects of natural selection. While measured effects on
+mean genetic scores are small, natural selection may substantially increase the correlation
+between genetic scores and income, i.e. the "genetic lottery" [@harden2021genetic].
 
@@
 cor_ea3$increase_pct <- (cor_ea3$ratio - 1) * 100
+
+drake::loadd(res_ineq_ea3)
+
+r2_income_true_psea <- scales::percent(res_ineq_ea3$r2_income_true_psea, 
+                                         accuracy = 0.1)
+r2_income_true_psea_wt <- scales::percent(res_ineq_ea3$r2_income_true_psea_wt, 
+                                            accuracy = 0.1)
 ```
@@
 proportionate increase of `r median_pct_change`% (Appendix Table
-\@ref(tab:tbl-inequality-effects)). If inequalities in polygenic scores are
-important for understanding class mobility and structure [@belsky2018genetic;
-@Rimfeld_2018; @harden2021genetic], then these increases are substantive. Since
-many polygenic scores are predictive of disease risk, they could also
-potentially increase health inequalities. In general, the evolutionary history
-of anatomically modern humans is related to disease risk [@benton2021influence];
-understanding the role of contemporary natural selection may aid in mapping the
-genetic architecture of current health disparities.
+\@ref(tab:tbl-inequality-effects)).
+
+If inequalities in polygenic scores are important for understanding social
+structure and mobility [@belsky2018genetic;
+@Rimfeld_2018; @harden2021genetic], then these increases are substantive. A
+back-of-the-envelope calculation, using estimated heritability for educational
+attainment of 0.4 [@branigan2013variation], gives the proportion of variance in
+income attributable to the "true" polygenic score for educational attainment as
+`r r2_income_true_psea` among respondents. Counterfactually holding environmental
+conditions constant, natural selection would raise this proportion to 
+`r r2_income_true_psea_wt` among respondents' children. Similarly, since many
+polygenic scores are predictive of disease risk, they could potentially increase
+health inequalities. In general, the evolutionary history of anatomically modern
+humans is related to disease risk [@benton2021influence]; understanding the role
+of contemporary natural selection may aid in mapping the genetic architecture of
+current health disparities.
 
@@
 
+## Estimating the effect of natural selection on the contribution of PSEA to inequality
+
+```{r calc-load-ineq-ea3}
+
+res <- drake::readd(res_ineq_ea3)
+
+prop_effect_selection <- res$r2_income_ea3_wt/res$r2_income_ea3 - 1
+prop_effect_selection <- scales::percent(prop_effect_selection, accuracy = 0.1)
+
+prop_wt_fe <- res$r2_income_ea3_fe_wt/res$r2_income_ea3_fe - 1
+prop_wt_fe <- scales::percent(prop_wt_fe, accuracy = 0.1)
+
+res <- lapply(res, format, digits = 3)
+
+```
+
+From the meta-analysis of twin studies in @branigan2013variation, we take the
+heritability of educational attainment to be around 0.4. This is the variance of
+educational attainment explained by the "true" polygenic score for educational
+attainment $p^{*}$, i.e. the best linear unbiased predictor of educational
+attainment from genetic data. We write the measured polygenic
+score of educational attainment, "EA3" from @lee2018gene, as
+\[
+p=p^{*}+\eta.
+\]
+
+For realized educational attainment $EA$, write:
+\[
+EA = \alpha+\beta p^{*} + \varepsilon = \alpha + \beta(p+\eta) + \varepsilon
+\]
+We assume the classical errors-in-variables model, i.e. that $\eta$
+is independent of both $p$ and $\varepsilon$. Under these conditions,
+the ratio between the estimated $R^{2}$ in a regression of $p$ on
+$EA$, and the true $R^{2}$ of $p^{*}$ on $EA$, satisfies: 
+\begin{equation}
+\frac{\textrm{plim}\hat{R}^{2}}{R^{2}}=\frac{\sigma_{p}^{2}-\sigma_{\eta}^{2}}{\sigma_{p}^{2}}=\frac{\sigma_{p^{*}}^{2}}{\sigma_{p}^{2}}\equiv\lambda,\label{eq:r2-ratio}
+\end{equation}
+as we show below. The right hand side is the "reliability ratio".
+We regress educational attainment on EA3 in our sample, weighting
+by our Age/Qualification weights, to give 
+$\textrm{plim}\hat{R}^2 = `r res$r2_edu_ea3`$. Plugging this
+into the above, with $R^2 = h^2\approx0.4$, gives $\lambda=`r res$lambda`$.
+
+We now consider the $R^2$ of $p$ on income, that is, the proportion
+of variance in income that is explained by the true polygenic score
+of educational attainment. Applying (\ref{eq:r2-ratio}) again gives
+that
+\begin{equation}
+\frac{\textrm{plim}\hat{R}_{\textrm{income}}^{2}}{R_{\textrm{income}}^{2}}=\lambda.\label{eq:r2-ratio-income}
+\end{equation}
+
+We compute $\hat{R}_{\textrm{income}}^{2}$by regressing income group on EA3,
+first weighting by age/qualification, and then by age/qualification weights
+times respondent's number of children. The estimated $R^2$ are 
+`r res$r2_income_ea3` and `r res$r2_income_ea3_wt` respectively. Applying
+(\ref{eq:r2-ratio-income}) to both estimates gives the proportion of variance in
+income explained by true polygenic score for education attainment (a) among
+respondents as `r res$r2_income_true_psea`; (b) accounting for natural selection
+as `r res$r2_income_true_psea_wt`. The second figure gives the counterfactual
+estimate of $R_{\textrm{income}}^{2}$ after one generation of natural selection,
+holding environmental conditions unchanged.
+
+The errors-in-variables conditions are quite restrictive. In particular, they
+rule out that the unmeasured part of PSEA ($-\eta$) correlates with the error 
+term $\varepsilon$, e.g. with environmental causes of educational attainment. 
+If unmeasured PSEA correlates positively with $\varepsilon$, then the
+reliability coefficient $\lambda$ will be lower than calculated. However, the
+proportional effect of natural selection on the PSEA-income relationship will
+remain the same, so long as $\lambda$ dues not differ between the two sets of
+weighted regressions.
+
+The estimate above uses a pooled regression of income on EA3, which may include
+confounding environmental variation such as genetic nurture. Within-siblings
+regression would give a smaller estimate of the effect of EA3. However, this
+would also affect the reliability of the regression [@griliches1986errors],
+making it hard to back out estimates of the true $R^2$. Nevertheless, we
+can calculate the proportional increase in $R^2$ due to natural selection, again 
+so long as the reliability coefficient is unaffected by
+the weighting scheme. Running within-siblings regressions of EA3 on income group
+gives within-$R^2$ estimates of `r res$r2_income_ea3_fe` (weighted by
+age and qualification) and `r res$r2_income_ea3_fe_wt` (multiplying weights by number of children). Thus, this procedure estimates the proportional change in $R^2$ 
+as `r prop_wt_fe`.
+
+### Proof of \@ref(eq:r2-ratio)
+
+Write 
+\begin{align}
+\hat{R}^{2} & =1-\frac{\sum_{i=1}^{N}(\hat{\alpha}+\hat{\beta}p_{i}-EA_{i})^{2}}{\sum_{i=1}^{N}(EA_{i}-\bar{EA})^{2}}\label{eq:r-squared}
+\end{align}
+
+Using the fact that $\hat{\alpha}=\bar{EA}-\hat{\beta}\bar{p}$, where bars signify
+sample averages, we can simplify the numerator in the above:
+
+\[
+\sum_{i=1}^{N}(\hat{\alpha}+\hat{\beta}p_{i}-EA_{i})^{2}=\sum_{i=1}^{N}(\bar{EA}-\hat{\beta}\bar{p}+\hat{\beta}p_{i}-EA_{i})^{2}=\sum_{i=1}^{N}(\hat{\beta}(p_{i}-\bar{p})-(EA_{i}-\bar{EA}))^{2}
+\]
+
+Now, we can compute the following probability limit:
+
+\begin{align}
+ & \text{plim}\frac{\sum_{i=1}^{N}(\hat{\beta}(p_{i}-\bar{p})-(EA_{i}-\bar{EA}))^{2}}{N}\nonumber \\
+= & \text{plim}\frac{\sum_{i=1}^{N}\hat{\beta}^{2}(p_{i}-\bar{p})^{2}+(EA_{i}-\bar{EA})^{2}-2\hat{\beta}(p_{i}-\bar{p})(EA_{i}-\bar{EA})}{N}\nonumber \\
+= & \beta^{2}\lambda^{2}\sigma_{p}^{2}+\sigma_{EA}^{2}-2\beta\lambda\sigma_{p,EA},\label{eq:plim}
+\end{align}
+
+using the standard errors-in-variables formula $\textrm{plim}\hat{\beta}=\lambda\beta$,
+and where $\sigma_{p,EA}$ is the covariance between $p$ and $EA$.
+
+Write
+\begin{align*}
+\sigma_{p,EA} & =\textrm{Cov}(p^{*}+\eta,\alpha+\beta p^{*}+\varepsilon)\\
+ & =0+\beta\sigma_{p^{*}}^{2}+\beta\textrm{Cov}(\eta,p^{*})+\textrm{Cov}(\eta,\varepsilon)\\
+ & =\beta\sigma_{p^{*}}^{2}
+\end{align*}
+
+using the errors-in-variables assumption that $\eta$ is independent
+of $p^{*}$ and $\varepsilon$. Plug this into (\ref{eq:plim}) to
+give
+\begin{align*}
+ & \text{plim}\frac{\sum_{i=1}^{N}(\hat{\beta}(p_{i}-\bar{p})-(EA_{i}-\bar{EA}))^{2}}{N}\\
+= & \beta^{2}\lambda^{2}\sigma_{p}^{2}+\sigma_{EA}^{2}-2\beta^{2}\lambda\sigma_{p^{*}}^{2}\\
+= & \beta^{2}\lambda\sigma_{p^{*}}^{2}+\sigma_{EA}^{2}-2\beta^{2}\lambda\sigma_{p^{*}}^{2}\textrm{, using \ensuremath{\sigma_{p^{*}}^{2}=\lambda\sigma_{p}^{2}},}\\
+= & \sigma_{EA}^{2}-\beta^{2}\lambda\sigma_{p^{*}}^{2}
+\end{align*}
+
+Inserting this into (\ref{eq:r-squared}) gives
+\begin{align*}
+\textrm{plim}\hat{R}^{2} & =1-\frac{\sigma_{EA}^{2}}{\sigma_{EA}^{2}}+\frac{\beta^{2}\lambda\sigma_{p^{*}}^{2}}{\sigma_{EA}^{2}}\\
+ & =\lambda\frac{\beta^{2}\sigma_{p^{*}}^{2}}{\sigma_{EA}^{2}}\\
+ & =\lambda R^{2}
+\end{align*}
+since for bivariate regression $R^{2}$ is the square of the correlation
+between $EA$ and $p^{*}$, which can be written $r=$$\beta\frac{\sigma_{p^{*}}^{2}}{\sigma_{EA}\sigma_{p^{*}}}$.
+We thank an anonymous commenter on https://stats.stackexchange.com/questions/238878/how-do-errors-in-variables-affect-the-r2
+for much of this proof.
+
 \clearpage
@@
 
-# Proofs
+# Model proofs
 

### 2021-12-14 _drake.R

@@ -567,6 +567,8 @@ plan <- drake_plan(
   
   res_ineq = run_cor_income(famhist, score_names, age_qual_weights),
   
+  res_ineq_ea3 = run_ineq_ea3_calcs(famhist, age_qual_weights, h2 = 0.4),
+  
   res_fe_fertility = run_reg_fe_fertility(famhist, score_names),
   
   report =  {

### 2021-12-14 regressions.R

@@ -242,6 +242,7 @@ run_cor_income <- function (famhist, score_names, age_qual_weights) {
   famhist <- famhist %>% 
                left_join(age_qual_weights, by = "f.eid") %>% 
                filter(
+                 kids_ss,
                  ! is.na(n_children), 
                  ! is.na(income_cat), 
                  ! is.na(weights)
@@ -264,4 +265,98 @@ run_cor_income <- function (famhist, score_names, age_qual_weights) {
   res <- res %>% mutate(ratio = actual/cf)
   
   res
+}
+
+run_ineq_ea3_calcs <- function (famhist, age_qual_weights, h2) {
+  
+  famhist <- famhist %>% filter(kids_ss)
+  famhist <- left_join(famhist, age_qual_weights, by = "f.eid")
+  famhist$child_weights <- famhist$n_children * famhist$weights
+  
+  res <- list()
+  
+  reg_edu_ea3 <- lm(age_fulltime_edu ~ EA3_excl_23andMe_UK, 
+                    data = famhist, weights = weights)
+  
+  res$r2_edu_ea3 <- summary(reg_edu_ea3)$r.squared
+  res$lambda <- res$r2_edu_ea3/h2
+  
+  reg_income_ea3 <- lm(income_cat ~ EA3_excl_23andMe_UK, 
+                       data = famhist, weights = weights)
+  reg_income_ea3_wt <- lm(income_cat ~ EA3_excl_23andMe_UK, 
+                         data = famhist, weights = child_weights)
+  
+  res$r2_income_ea3 <- summary(reg_income_ea3)$r.squared
+  res$r2_income_ea3_wt <- summary(reg_income_ea3_wt)$r.squared
+  
+  res$r2_income_true_psea <- res$r2_income_ea3 / res$lambda
+  res$r2_income_true_psea_wt <- res$r2_income_ea3_wt / res$lambda
+  
+  reg_income_ea3_fe <- fixest::feols(income_cat ~ EA3_excl_23andMe_UK | sib_group, 
+                                     data = famhist, weights = famhist$weights)
+  reg_income_ea3_fe_wt <- fixest::feols(income_cat ~ EA3_excl_23andMe_UK | sib_group, 
+                                     data = famhist, weights = famhist$child_weights)
+  res$r2_income_ea3_fe <- fixest::r2(reg_income_ea3_fe, type = "wr2")
[…]

### 2021-12-15 why-natural-selection.Rmd
More of a mediation analysis

@@
 
-Second, we re-estimate equation \@ref(eq:regression) for each polygenic score,
-controlling for education levels. Effect sizes are
-generally reduced (Appendix Figure \@ref(fig:plot-res-educ)), with
-large reductions for polygenic scores for educational attainment. This 
-suggests that education indeed mediates the effect of polygenic scores on fertility.
+Second we test whether educational attainment mediates the correlation between
+polygenic scores and fertility.
 
@@
 \clearpage
-## Controlling for earnings and education
-
-```{r plot-res-educ, fig.cap = "Selection effects controlling for education (left education before 16, 16-18, or after 18). Raw effects are shown for comparison."}
-
-drake::loadd(res_ee_control)
-drake::loadd(res_all)
-
-res_ee_control %<>% filter(term == score_name)
-
-res_all %<>% 
-  filter(term == score_name, reg.type == "controlled", dep.var == "N children") %>% 
-  select(-dep.var, -reg.type)
-
-res_compare <- bind_rows(
-                 Raw = res_all, 
-                 Controlled = res_ee_control, 
-                 .id = "Regression"
-               )
-
-standard_ggplot(res_compare, fill_col = Regression, n_regs = 33)
-
-```
+## Mediation analysis
+
+We run a standard mediation analysis in the framework of @baron1986moderator. We estimate 
+\begin{align}
+N_i & = \alpha + \beta PGS_i + \gamma EA_i + X_i\mu + \varepsilon_i \\
+EA_i & = \delta + \zeta PGS_i + X_i\mu + \eta_i
+\end{align}
+where $N_i$ is fertility, $PGS_i$ is each polygenic score, $EA_i$ is educational
+attainment (age of leaving fulltime education), and $X_i$ is a vector of
+controls. The total effect of $PGS$ on number of children is $\beta + \gamma
+\zeta$. The "indirect effect" mediated by $EA$ is $\gamma \zeta$. The standard
+error of the indirect effect can be calculated as
+
+\[
+\sqrt{\hat\gamma^2 \hat\sigma_\zeta^2 + \hat\zeta^2\hat\sigma_\gamma^2}
+\]
+
+where $\hat\sigma_\zeta$ is the standard error of $\hat\zeta$, etc.
+
+We are not concerned about endogeneity of polygenic scores, since our theory
+requires only that these correlate with human capital. However, endogeneity of
+educational attainment is a concern. For example, educational attainment might
+correlate with unobserved characteristics that affect one's fertility, e.g. via
+success in marriage markets. To partially address this, we include
+controls for age, sex, fluid IQ and height in $X$.
+
+```{r tbl-res-mediation}
+
+drake::loadd(res_mediation)
+
+res_mediation %<>% mutate(
+                     prop = estimate_ind/estimate_total,
+                     # 0.025 not 0.05 because two-sided
+                     sig  = abs(statistic_ind) > qnorm(1 - 0.025/33),
+                     star = ifelse(sig, "*", "")
+                   )
+
+n_positive <- sum(res_mediation$prop > 0)
+n_pos_sig <- sum(res_mediation$sig & res_mediation$prop > 0)
+median_prop <- median(res_mediation$prop)
+
+res_mediation %>%
+      arrange(desc(estimate_total)) %>%
+      transmute(
+        PGS = pretty_names(term),
+        `Total effect` = estimate_total,
+        `Indirect effect` = estimate_ind,
+        `95% c.i.` = glue::glue(
+                     "[{estimate_ind - 1.96 * se_ind}, {estimate_ind + 1.96 * se_ind}]"
+                     ),
+        `Indirect/total` = prop,
+        ` ` = star
+      ) %>%
+      as_hux() %>%
+      set_align(everywhere, -1, "right") %>% 
+      set_number_format(-1, 2:4, 4) %>% 
+      set_number_format(-1, 5, fmt_percent()) %>%
+      set_caption("Mediation analysis") %>% 
+      theme_compact() %>% 
+      set_width(0.9) %>%
+      set_col_width(c(.3, .125, .125, .3, .1, .05)) %>% 
+      set_bottom_border(final(), everywhere) %>%
+      add_footnote("* indirect effect p < 0.05/33", font_size = 8)
+```
+
+Table \@ref(tab:tbl-res-mediation) shows results. `r n_pos_sig` scores are 
+significantly greater than 0 at $p = 0.05/33$.
 
@@
 \begin{equation}
-\frac{\textrm{plim}\hat{R}^{2}}{R^{2}}=\frac{\sigma_{p}^{2}-\sigma_{\eta}^{2}}{\sigma_{p}^{2}}=\frac{\sigma_{p^{*}}^{2}}{\sigma_{p}^{2}}\equiv\lambda,\label{eq:r2-ratio}
+\frac{\textrm{plim}\hat{R}^{2}}{R^{2}} 
+= \frac{\sigma_{p}^{2} - \sigma_{\eta}^{2}}{\sigma_{p}^{2}}
+= \frac{\sigma_{p^{*}}^{2}}{\sigma_{p}^{2}}
+\equiv \lambda, \label{eq:r2-ratio}
 \end{equation}

### 2021-12-15 _drake.R

@@ -542,16 +542,7 @@ plan <- drake_plan(
     res %>% filter(grepl(":factor\\(age_fte_cat\\)", term))
   },
   
-  res_ee_control = {
-    map_dfr(score_names, 
-              ~ run_regs_fml(
-                fml = "n_children ~ {score_name} + age_fte_cat",
-                score_name = .x,
-                famhist    = famhist,
-                subset     = quote(kids_ss)
-              ), 
-            .id = "score_name")
-  },
+  res_mediation = run_mediation(famhist, score_names),
   
   res_risk_control = {
     # f.2040.0.0 is risk attitude (via questionnaire)

### 2021-12-30 why-natural-selection.Rmd
Shorten the parents section; discuss mediation analysis; add TOC to appendix; explain choice of PGS.

@@
 Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner.
-The direction of natural selection is reversed among older parents (22+), or after controlling for age at first live birth.
+The direction of natural selection is reversed among older parents, or after controlling for age at first live birth.
 These patterns are in line with economic theories of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market.
@@
   - \usepackage{lineno}
+  - \usepackage{etoc}
   - \linenumbers
@@
 options(dplyr.summarise.inform = FALSE) # shut up about summarize
+
 knitr::opts_chunk$set(
@@
       )
+
+knitr::knit_hooks$set(
+  inline = function (x) {
+    if (is.numeric(x)) x <- as.character(round(x, getOption("digits")))
+    x <- gsub("-", "\u2212", x)
+    paste(as.character(x), collapse = ", ")
+  }
+)
+
 options(huxtable.long_minus = TRUE)
-
 huxtable::set_default_properties(latex_float = "h!")
@@
 [@beauchamp2016genetic].[^normalization] Note that equation
-\eqref{eq:regression} does not include many possible environmental and
-genetic confounds that could affect fertility, and as a result, $\beta$ is not
-an estimate of the causal effect of a polygenic score on fertility. However, 
-natural selection is a correlational phenomenon, not a causal one: if a 
-polygenic score correlates with low fertility, it is being selected against, 
-whatever the underlying causal relationship.
+\eqref{eq:regression} does not control for many environmental and genetic
+factors that could affect fertility, and as a result, $\beta$ is not an estimate
+of the causal effect of a polygenic score on fertility. However, natural
+selection is a correlational phenomenon: polygenic scores which correlate with
+high fertility are being selected for, whatever the underlying causal
+relationship.
 
@@
 
-\FloatBarrier
-
-## Selection in the parents' generation
-
-The UK Biobank data contains information on respondents' number of
-siblings. Since respondents' polygenic scores are equal in expectation
-to the mean scores of their parents, we can use this to look at selection 
-effects in the parents' generation. We estimate equation (\ref{eq:regression})
-using *number of siblings* (including the respondent) as the dependent variable.
-The parents' generation has an additional source of ascertainment bias: sampling
-parents of respondents overweights parents who have many children. For instance,
-parents of three children will have, on average, three times more children
-represented in UK Biobank than parents of one child. Parents of no children will
-by definition not be represented. To compensate, we reweight our preferred
-weightings (Age/Qualification) by the inverse of *number of siblings*.
-
-Selection effects are highly correlated across the two generations, and most
-share the same sign (Appendix Figure \@ref(fig:plot-res-sibs)). Absolute effect
-size estimates are larger for the parents' generation. We treat this result
-cautiously, because when we split respondents up by year of birth, we find few
-differences in effect sizes between early- and late-born respondents, for either
-generation. In other words, since estimated effect sizes change across
-"generations", but do not change over time within either generation, the change
-may be due to remaining ascertainment bias within the sample or other effects.
-In particular, effect sizes in both generations may depend on polygenic scores'
-correlation with childlessness, and we cannot estimate this for the parents'
-generation.
-
-Although the direction of selection effects does not change between the
-generations, there are other differences. Compared to our standard polygenic
-scores (residualized on 100 principal components of genetic data), 
-selection effects on unresidualized scores are about ten percent higher
-on average for *number of siblings*, whereas effects for *number of children*
-barely change (Appendix Figure \@ref(fig:plot-regs-controlled-pcs)). This could
-be because earlier fertility is driven more by geographically clustered
-deprivation [e.g. via an insurance motive, @rendall1998old], which may correlate
-with the broad-scale genetic variation captured by principal components, such as
-ancestry.
-
-We also check whether selection effects differ by socio-economic status in the
-parents' generation. We have no information about parents' income, so we use the
-1971 Townsend deprivation score of respondents' birthplace as a proxy
-[@townsend1987deprivation]. Results (Appendix Figure
-\@ref(fig:plot-siblings-townsend)) show the same pattern as for respondents:
-effect sizes are larger and more often significant in the most deprived areas.
-
 ```{r calc-siblings-children}
@@
 
-Lastly, the siblings data lets us check for a "quantity-quality tradeoff" between
-number of children and number of grandchildren (for the parents' generation). We
+We also examine selection effects among respondents' parents, using
+information on respondents' number of siblings. Effect sizes of polygenic
+scores are highly correlated across the two generations. Median-splitting
+respondents by year of birth, we find little evidence of change in effect sizes
+among either the parents' or the respondents' generation. We check whether 
+selection effects differ by socio-economic status in the
+parents' generation. We have no information about parents' income, so we use the
+1971 Townsend deprivation score of respondents' birthplace as a proxy
+[@townsend1987deprivation]. Results show the same pattern as for respondents:
+effect sizes are larger and more often significant in the most deprived areas
+(Appendix Figure \@ref(fig:plot-siblings-townsend)). Lastly, the siblings data 
+lets us check for a "quantity-quality tradeoff" between
+number of children (of respondents' parents) and number of grandchildren. We
 do not find any: in fact, the correlation between *number of siblings* and 
@@
 $p < 2 \times 10^{-16}$).
+
+\FloatBarrier
 
@@
 higher. The income effect leads higher earners to have more children. The
-substitution effect leads them to have fewer children.
+substitution effect leads them to have fewer.
 
@@
 Here $a$ captures the strength of preference for children. $u(\cdot)$ captures
-the taste for income, and is concave and increasing. We treat $N$ as continuous, in line with
-the literature: this can be thought of as the expected number of children
-among similar people. The marginal benefit of an extra child is
+the taste for income, and is increasing and concave. We treat $N$ as continuous,
+in line with the literature: this can be thought of as the expected number of
+children among similar people. The marginal benefit of an extra child is
 $\frac{dU}{dN} = -bWu'(Y)+a$. The effect of an increase in human capital on this
@@
 
-The above does not aim to provide a complete theory of fertility. Rather, it
-shows that a simple economic model can explain many of our results.
+Thus, a simple economic model can explain many of our results.
 Other empirical work in economics also supports our mechanism.
@@
 drake::loadd(res_age_flb)
-drake::loadd(ashe_income)
-
-famhist <- add_ashe_income(famhist, ashe_income)
-
-cors_income <- cor(famhist[score_names], famhist$first_job_pay, use = "pairwise")
-# TODO: use age_left_edu when this includes college
-
-cors_educ <-   cor(famhist[score_names], famhist$age_fulltime_edu, use = "pairwise")
+drake::loadd(res_cor_educ)
+drake::loadd(res_cor_income)
 
@@
         Correlation  = rep(c("Earnings", "Education"), each = 33, 2),
-        cor          = rep(c(cors_income[,1], cors_educ[,1]), 2),
+        cor          = rep(c(res_cor_income[,1], res_cor_educ[,1]), 2),
         effect_size  = c(rep(effect_size, 2), rep(effect_size_flb, 2)),
@@
 educational attainment. The relationships are strongly negative. Thus, human
-capital appears to be relevant to natural selection. Substitution effects
-dominate income effects overall, which fits the known association between income
-and lower fertility [@becker1960economic;@jones2006economic]. The correlations
-reverse when we control for age at first live birth, suggesting that within AFLB
-categories, the income effect dominates. (See also the Supplementary Animations
-for the uncorrected effects per AFLB group, which show the correlation reversing
-after age 22.)
-
-Second we test whether educational attainment mediates the correlation between
-polygenic scores and fertility.
+capital appears to be relevant to natural selection. The negative relationship
+suggests that substitution effects dominate income effects, which fits the known
+negative association between income and fertility
+[@becker1960economic;@jones2006economic]. The correlations reverse when we
+control for age at first live birth, suggesting that within AFLB categories, the
+income effect dominates. (See also the Supplementary Animations for the
+uncorrected effects per AFLB group, which show the correlation reversing after
+age 22.)
+
+```{r}
+drake::loadd(res_mediation)
+
+res_mediation %<>% mutate(
+                     prop = estimate_ind/estimate_total,
+                     # 0.025 not 0.05 because two-sided
+                     sig  = abs(statistic_ind) > qnorm(1 - 0.025/33),
+                     star = ifelse(sig, "*", "")
+                   )
+
+n_pos_sig <- sum(res_mediation$sig & res_mediation$prop > 0)
+
+```
+
+Second, we directly test whether the correlation between each polygenic score
+and fertility is mediated by educational attainment (Appendix Table
+\@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the indirect effect of
+the score on fertility *via* educational attainment is significantly different
+from zero ($p < 0.05/33$).
 
@@
 birth and educational attainment [@Howe2021.03.05.433935]. We see these results
-as providing weak evidence that polygenic scores cause fertility, with effects
+as providing tentative evidence that polygenic scores cause fertility, with effects
 being at least partly driven by gene-environment correlations. We also reran
@@
 
-We consider three alternative theories that might explain our results. The first
-is that welfare benefits which incentivize child-bearing might be taken up
+We consider three alternative theories that might explain our results. First,
+welfare benefits which incentivize child-bearing might be taken up more
 among low-income individuals. However, our effect sizes appear unchanged
@@
 so when controlling for risk attitude. This non-result could simply reflect the
-imprecision of the risk aversion measure, which is a single yes/no question.
-However, the risk aversion measure *does* predict the overall number of
+imprecision of the risk attitude measure, which is a single yes/no question.
+However, this measure *does* predict the overall number of
 children, highly significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of
 33 regressions). Given that, and the statistical power we get from our sample
-size, we believe that the non-result is real: while risk aversion does predict
+size, we believe that the non-result is real: while risk attitude does predict
 fertility in the sample, it is not an important channel for natural selection.
@@
 We also do not know how estimated effect sizes of natural selection will change
-as more accurate measures of genetic variation are produced. And we are unsure
-whether genetic variants underlying other phenotypes will show a similar pattern
-of natural selection to those studied here. In addition, genetic effects on
-educational attainment have been shown to be inflated in population-based
-samples as compared to within-family designs, likely because of indirect genetic 
-effects, gene-environment correlations, and/or assortative mating 
-[@lee2018gene; @selzam2019comparing; @kong2018nature; @Howe2021.03.05.433935]. 
-In short, it is probably too early to tell whether modern natural selection has a
-substantively important effect on the genetic make-up of the population.
-Nevertheless, we note that selection effects on our measured polygenic scores
-are still relatively small, even after reweighting to account for ascertainment
-bias.
+as more accurate measures of genetic variation are produced, or whether genetic
+variants underlying other phenotypes will show a similar pattern to those
+studied here. In addition, genetic effects on educational attainment have been
+shown to be inflated in population-based samples as compared to within-family
+designs, likely because of indirect genetic effects, gene-environment
+correlations, and/or assortative mating [@lee2018gene; @selzam2019comparing;
+@kong2018nature; @Howe2021.03.05.433935]. In short, it is probably too early to
+tell whether modern natural selection has a substantively important effect on
+population means of phenotypes under selection. Nevertheless, we note that selection
+effects on our measured polygenic scores are still relatively small, even after
+reweighting to account for ascertainment bias.
 
@@
 income attributable to the "true" polygenic score for educational attainment as
-`r r2_income_true_psea` among respondents. Counterfactually holding environmental
-conditions constant, natural selection would raise this proportion to 
-`r r2_income_true_psea_wt` among respondents' children. Similarly, since many
-polygenic scores are predictive of disease risk, they could potentially increase
-health inequalities. In general, the evolutionary history of anatomically modern
-humans is related to disease risk [@benton2021influence]; understanding the role
-of contemporary natural selection may aid in mapping the genetic architecture of
-current health disparities.
+`r r2_income_true_psea` among respondents. Natural selection would raise this
+proportion to `r r2_income_true_psea_wt` among respondents' children,
+counterfactually holding environmental conditions constant. Similarly, since
+many polygenic scores are predictive of disease risk, they could potentially
+increase health inequalities. In general, the evolutionary history of
+anatomically modern humans is related to disease risk [@benton2021influence];
+understanding the role of contemporary natural selection may aid in mapping the
+genetic architecture of current health disparities.
 
@@
 
-Polygenic scores were computed by summing the alleles across ~1.3 million 
+Polygenic scores were chosen so as to cover a reasonably broad range of traits,
+and based on the availability of a large and powerful GWAS which did not include
+UK Biobank. Scores were computed by summing the alleles across ~1.3 million
 genetic variants weighted by their effect sizes as estimated in 33 genome-wide
 association studies (GWASs) that excluded UK Biobank. To control for population
-stratification, we corrected the polygenic scores for 100 principal components (PCs).
+stratification, we corrected the polygenic scores for 100 principal components
+(PCs).
 
@@
 # Appendix
+
+\localtableofcontents
+\clearpage
 
@@
 
-Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings*, i.e.
-parents' number of children, on polygenic scores. By definition, members of the
-parents' generation who had no children cannot be included in this data. For a
-clean comparison with the respondents' generation, we rerun regressions on
-*number of children* excluding those with no children, and show
-results in the figure.
+The UK Biobank data contains information on respondents' number of
+siblings, i.e. their parents' number of children. Since respondents' polygenic 
+scores are equal in expectation
+to the mean scores of their parents, we can use this to look at selection 
+effects in the parents' generation. We estimate equation (\ref{eq:regression})
+using *number of siblings* (including the respondent) as the dependent variable.
+The parents' generation has an additional source of ascertainment bias: sampling
+parents of respondents overweights parents who have many children. For instance,
+parents of three children will have, on average, three times more children
+represented in UK Biobank than parents of one child. Parents of no children will
+by definition not be represented. To compensate, we reweight our preferred
+weightings (Age/Qualification) by the inverse of *number of siblings*.
+
+Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings* on
+polygenic scores. For a clean comparison with the respondents' generation, we
+rerun regressions on *number of children* excluding those with no children, and
+show results in the figure. Selection effects are highly correlated across the
+two generations, and most share the same sign. Absolute effect size estimates
+are larger for the parents' generation. We treat this result cautiously, because
+when we split respondents up by year of birth, we find few differences in effect
+sizes between early- and late-born respondents, for either generation. In other
+words, since estimated effect sizes change across "generations", but do not
+change over time within either generation, the change may be due to remaining
+ascertainment bias within the sample or other effects. In particular, effect
+sizes in both generations may depend on polygenic scores' correlation with
+childlessness, and we cannot estimate this for the parents' generation.
 
@@
 childless people may have had more or less effect in that generation. Another
-issue is that we are in effect estimating parents' polygenic scores by the
-scores of their children. This introduces some noise into our independent
+issue is that we are estimating parents' polygenic scores by the
+scores of their children. This introduces noise into our independent
 variable, which might lead to errors-in-variables and bias coefficients towards
@@
 drake::loadd(res_townsend_parents)
-drake::loadd(famhist_townsend_71)
 n_regs <- nrow(res_townsend_parents)
@@
     mutate(
-    `Townsend quintile` = sub("Quin71(\\d):.*", "\\1", term),
-    `Townsend quintile` = add_n(`Townsend quintile`, "Quin71", dv = "n_sibs", 
-                                  data = famhist_townsend_71)
+        n = prettyNum(n, big.mark = ","),
+        `Townsend quintile` = paste0(quintile, " (N = ", n, ")")
       ) %>% 
@@
 
-rm(famhist_townsend_71)
 ```
@@
 
-n_positive <- sum(res_mediation$prop > 0)
 n_pos_sig <- sum(res_mediation$sig & res_mediation$prop > 0)
-median_prop <- median(res_mediation$prop)
 
@@
 
-Table \@ref(tab:tbl-res-mediation) shows results. `r n_pos_sig` scores are 
-significantly greater than 0 at $p = 0.05/33$.
+Table \@ref(tab:tbl-res-mediation) shows results. For `r n_pos_sig` scores, the
+indirect effect on fertility via human capital is significantly greater than 0 
+at $p = 0.05/33$.
 
@@
 Age/Qualification only, i.e. the counterfactual correlation if all respondents
-had the same number of children (equivalently, the correlation in the
-respondents' own generation).
+had the same number of children.
 
@@
 proportional effect of natural selection on the PSEA-income relationship will
-remain the same, so long as $\lambda$ dues not differ between the two sets of
+remain the same, so long as $\lambda$ does not differ between the two sets of
 weighted regressions.
@@
 gives within-$R^2$ estimates of `r res$r2_income_ea3_fe` (weighted by
-age and qualification) and `r res$r2_income_ea3_fe_wt` (multiplying weights by number of children). Thus, this procedure estimates the proportional change in $R^2$ 
-as `r prop_wt_fe`.
+age and qualification) and `r res$r2_income_ea3_fe_wt` (multiplying weights by number of children).
+Thus, this procedure estimates the proportional change in $R^2_{income}$ due to natural selection as `r prop_wt_fe`.
 
@@
 
-# Model proofs
+## Model proofs
 
@@
 \FloatBarrier
+
 # References

### 2021-12-30 _drake.R

@@ -354,6 +354,16 @@ plan <- drake_plan(
               .id = "score_name"
             )
     res %<>% filter(grepl(":", term))
+    quintiles <- famhist_townsend_71 %>%
+                   filter(! is.na(n_sibs), ! is.na(whr_combined)) %>%
+                   group_by(Quin71) %>%
+                   count()
+    res %<>% 
+      mutate(
+        quintile = sub("Quin71(\\d):.*", "\\1", term)
+      ) %>%
+      left_join(quintiles, by = c("quintile" = "Quin71"))
+    
     res
   },
   
@@ -557,6 +567,15 @@ plan <- drake_plan(
   },
   
   res_ineq = run_cor_income(famhist, score_names, age_qual_weights),
+  
+  res_cor_income = {
+    famhist <- add_ashe_income(famhist, ashe_income)
+    cor(famhist[score_names], famhist$first_job_pay, use = "pairwise")
+  },
+  
+  res_cor_educ = {
+    cor(famhist[score_names], famhist$age_fulltime_edu, use = "pairwise")
+  },
   
   res_ineq_ea3 = run_ineq_ea3_calcs(famhist, age_qual_weights, h2 = 0.4),
   

### 2021-12-30 why-natural-selection.Rmd
Add control for risk attitude to mediation analysis

@@
 requires only that these correlate with human capital. However, endogeneity of
-educational attainment is a concern. For example, educational attainment might
-correlate with unobserved characteristics that affect one's fertility, e.g. via
-success in marriage markets. To partially address this, we include
-controls for age, sex, fluid IQ and height in $X$.
+educational attainment is a concern, i.e., educational attainment might
+correlate with unobserved characteristics that affect one's fertility (and which
+are not related to human capital). To partially address this, we include
+controls for age, sex, fluid IQ, height and risk attitude in $X$.
 

### 2021-12-30 why-natural-selection.Rmd
tweak intro

@@
 
-As yet we know little about the social mechanisms behind these effects. A
-potential explanation for them comes from the economic theory of fertility first
-presented by [@becker1960economic]. In this theory, higher potential earnings
+As yet we know little about the social mechanisms behind natural selection. A
+potential explanation comes from the economic theory of fertility [@becker1960economic]. In this theory, higher potential earnings
 have two opposite effects on fertility: a fertility-increasing *income effect*
@@
 
-We then show that a parsimonious model of human capital, education and fertility
+We then show that a simple model of human capital, education and fertility
 choices can give rise to these empirical results. To test the theory, we show
 that polygenic scores' correlation with fertility is associated with their
-correlation with education and earnings. Furthermore, controlling for education
-and earnings reduces the correlation with fertility, suggesting that human
-capital mediates natural selection. Thus, contemporary natural selection on
-polygenic scores can be explained by their correlation with earnings-increasing
-human capital.
+correlation with education and earnings. Furthermore, a mediation analysis shows
+that part of the correlation with fertility is indeed *via* education.
+Thus, contemporary natural selection on polygenic scores can be explained by their correlation with earnings-increasing human capital.
 
 Lastly, we discuss the effects of natural selection. While measured effects on
-mean genetic scores are small, natural selection may substantially increase the correlation
-between genetic scores and income, i.e. the "genetic lottery" [@harden2021genetic].
+mean genetic scores are small, natural selection may substantially increase the
+correlation between genetic scores and income, i.e. the "genetic lottery"
+[@harden2021genetic].
 

### 2022-01-02 why-natural-selection.Rmd
Run a single regression on all 29 scores simultaneousl within-siblings

@@
 
+educ_reduction <- with(res_fe_fertility, 
+       estimate[Regression == "Controlled" & term == "EA3_excl_23andMe_UK"] / 
+       estimate[Regression == "Raw" & term == "EA3_excl_23andMe_UK"]
+     )
+educ_reduction <- 1 - educ_reduction
+
 res_all %<>% filter(dep.var == "N children", reg.type == "controlled")
@@
 se_pool_w <- lm_pool_w[[2, "std.error"]]
+
 ```
@@
 These results support our theory that natural selection on polygenic scores is
-driven by their correlation with human capital. This need not imply
-that a given polygenic score is causal of either fertility or human capital. Instead
-scores may simply correlate with causally relevant environments, for example via
-"genetic nurture" effects [@kong2018nature] or population stratification. To 
-investigate the presence of these indirect genetic effects, we estimate equation
-\@ref(eq:regression) among full siblings, controlling for a full set of sibling
-group dummies (Appendix Figure \@ref(fig:plot-res-fe-fertility)). With a reduced
-sample size, all within-sibling effects are insignificant after Bonferroni
-correction. However, effect sizes are positively correlated with effect sizes
-from the pooled model, and about 1/3 smaller (regressing within-sibling on
-pooled effect sizes, $b$ = `r coef_pool_w`). This attenuation is broadly in line
-with the ~50% decrease in heritability in within-sibling GWASs on age at first
-birth and educational attainment [@Howe2021.03.05.433935]. We see these results
-as providing tentative evidence that polygenic scores cause fertility, with effects
-being at least partly driven by gene-environment correlations. We also reran
+driven by their correlation with human capital. We also test for causality by
+running within-siblings regressions (Appendix Figure
+\@ref(fig:plot-res-fe-fertility)). With a reduced sample size, all
+within-sibling effects are insignificant after Bonferroni correction. However,
+effect sizes are positively correlated with effect sizes from the pooled model,
+and about 50% smaller (regressing within-sibling on pooled effect sizes, $b$ =
+`r format(coef_pool_w, digits = 3)`). This attenuation is in line with the ~50%
+decrease in heritability in within-sibling GWASs on age at first birth and
+educational attainment [@Howe2021.03.05.433935]. We see these results as
+providing tentative evidence that polygenic scores cause fertility, with effects
+being partly driven by gene-environment correlations. We also reran
 within-siblings regressions adding a control for education. Most effect sizes
@@
 mediate differences in fertility among siblings. However, for educational
-attainment polygenic scores, effect sizes are reduced by about 20%.
+attainment polygenic scores, effect sizes are reduced by `r percent(educ_reduction)`.
 
@@
 
-```{r plot-res-fe-fertility, fig.cap = "Selection effects controlling for sibling-group fixed effects, with and without a control for education (left education before 16, 16-18, or after 18)."}
+```{r plot-res-fe-fertility, fig.cap = "Selection effects controlling for sibling-group fixed effects, with and without a control for education (left education before 16, 16-18, or after 18).
+Each set of results is from a single regression of 'n children' on 29 polygenic scores.
+Standard errors clustered by sibling group."}
 
 drake::loadd(res_fe_fertility)
-
-standard_ggplot(res_fe_fertility, fill_col = Regression, n_regs = 33, fill_direction = -1)
-
-```
+n_regs <- nrow(res_fe_fertility)/2 # Raw and Controlled treated separately
+
+standard_ggplot(res_fe_fertility, fill_col = Regression, n_regs = n_regs,
+                  fill_direction = -1)
+
+n_groups <- attr(res_fe_fertility, "n_groups")
+n <- attr(res_fe_fertility, "n")
+
+```
+
+We run a single regression on 29 polygenic scores within `r n_groups` sibling
+groups (N = `r n`). Thus, we control both for environmental confounds (since
+scores are randomly allocated within sib-groups by meiosis), and for genetic
+confounds captured by our polygenic scores. We remove four scores which
+correlate highly with other scores (educational attainment 2, hip circumference,
+waist circumference and waist-hip ratio). Figure \@ref(fig:plot-res-fe-fertility)
+shows the results.
 

### 2022-01-02 why-natural-selection.Rmd
Add a "further results" section in the appendix

@@
 effect sizes are larger and more often significant in the most deprived areas
-(Appendix Figure \@ref(fig:plot-siblings-townsend)). Lastly, the siblings data 
+(Appendix Figure \@ref(fig:plot-siblings-townsend)); effects change
+sign when controlling for AFLB (Appendix Figure \@ref(fig:plot-age-birth-parents)). Lastly, the siblings data 
 lets us check for a "quantity-quality tradeoff" between
@@
 \FloatBarrier
-## Selection effects on raw polygenic scores
-
-```{r calc-regs-controlled}
-
-drake::loadd(res_all)
-
-res_all_controlled <- res_all %>% 
-  select(reg.type, dep.var, term, estimate) %>% 
-  pivot_wider(names_from = c(reg.type, dep.var), values_from = estimate) %>% 
-  mutate(
-    consistent = sign(`raw_N siblings`) == sign(`raw_N children`) &
-                 sign(`raw_N siblings`) == sign(`controlled_N siblings`) &
-                 sign(`raw_N siblings`) == sign(`controlled_N children`)
-  )
-
-n_pgs  <- as.double(nrow(res_all_controlled))
-prop_consistent_controlled <- mean(res_all_controlled$consistent)
-controlled_smaller_sibs <- abs(res_all_controlled[["controlled_N siblings"]]) < 
-  abs(res_all_controlled[["raw_N siblings"]])
-controlled_smaller_children <- abs(res_all_controlled[["controlled_N children"]]) < 
-  abs(res_all_controlled[["raw_N children"]])
-median_prop_sibs <- median(abs(res_all_controlled[["controlled_N siblings"]]) /
-    abs(res_all_controlled[["raw_N siblings"]]))
-median_prop_children <- median(abs(res_all_controlled[["controlled_N children"]]) /
-    abs(res_all_controlled[["raw_N children"]]))
-
-```
-
-Figure \@ref(fig:plot-regs-controlled-pcs) compares selection effects on
-polygenic scores residualized for the top 100 principal components of the
-genetic data, to selection effects on raw, unresidualized polygenic scores. In
-siblings regressions, effect sizes are larger for raw scores  -- sometimes much
-larger, as in the case of height. `r sum(controlled_smaller_sibs)` out of 
-`r n_pgs` "raw" effect sizes have a larger absolute value than the corresponding
-"residualized" effect size. The median proportion between raw and controlled
-effect sizes is `r median_prop_sibs`. Among the children regressions, this no
-longer holds. Effect sizes are barely affected by controlling for principal
-components.
-
-Overall, `r prop_consistent_controlled * 100` per cent of effect sizes
-are consistently signed across all four regressions (on children and
-siblings, and with and without residualization).
-
-```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores on number of siblings/children."}
-
-drake::loadd(res_all)
-
-n_regs <- as.double(nrow(res_all))
-res_all %>% 
-      mutate(
-        term     = pretty_names(term),
-        term     = reorder_within(term, estimate, dep.var, order_abs(1)),
-        dep.var  = fct_inorder(dep.var),
-        reg.type = fct_recode(reg.type, 
-                               `Resid. for 100 PCs`  = "controlled", 
-                               Unresidualized        = "raw"
-                             )
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(y = term, x = estimate, fill = reg.type, 
-        colour = p < 0.05/{{n_regs}})) +
-        my_vline + 
-        facet_wrap(vars(dep.var), scales = "free") +
-        pval_geom_point(position = position_dodge(0.4), alpha = 0.8) +
-        my_labs() +
-        pval_scale() +
-        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
-        my_theme() +
-        scale_y_reordered() +
-        theme(
-          axis.text.x  = element_text(size = 7),
-          legend.text  = element_text(size = 8),
-          legend.title = element_text(size = 9)
-        )
-
-```
-
-```{r}
-drake::loadd(res_pcs)
-n_regs <- as.double(nrow(res_pcs))
-prop_sig <- res_pcs %>% 
-      group_by(dep.var) %>% 
-      summarize(sig = sum(p.value <= 0.05/nrow(res_pcs))) %>% 
-      pull(sig, name = dep.var)
-```
-
-To get a further insight into this we regress *n siblings* and *n children*
-on individual principal components. As Figure \@ref(fig:plot-pc-regs)
-shows, effects are larger and more significant in siblings regressions.
-`r prop_sig[["N siblings"]]` principal components significantly
-predicted number of siblings, while only `r prop_sig[["N children"]]`
-significantly predicted number of children.
-
-```{r plot-pc-regs, fig.cap = "Selection effects of 100 principal components of genetic data.
-Absolute effect sizes are plotted.
-Each dot represents one bivariate regression.
-Points are jittered on the Y axis.", fig.height = 2}
-
-res_pcs %>% 
-      mutate(
-        estimate = abs(estimate),
-        dep.var  = fct_relevel(dep.var, "N children")
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(estimate, dep.var, colour = p < 0.05/{{n_regs}})) + 
-        geom_jitter(height = 0.15, width = 0, alpha = 0.8) + 
-        scale_x_log10(labels = scales::label_number(accuracy = 0.0001)) +
-        labs(x = "Effect size (log scale)", y = "") + 
-        pval_scale(aesthetics = "colour")
-
-```
-
-\clearpage
-\FloatBarrier
 ## Selection controlling for age at first live birth: respondents' parents
@@
 shows the results.
-
-\FloatBarrier
-\clearpage
-## Genetic correlations with EA3
-
-```{r plot-rgs-by-effect-size, fig.cap = "Selection effects plotted against genetic correlation with EA3."}
-
-# TODO: ref source of data.
-
-drake::loadd(rgs)
-drake::loadd(res_all)
-
-rgs <- left_join(rgs, 
-        res_all %>% filter(reg.type == "controlled", dep.var == "N children"), 
-        by = c("p2" = "term"))
-
-rgs_cor <- cor.test(~ rg + estimate, 
-                      data = rgs %>%
-                               filter(
-                                 p2 != "EA3_excl_23andMe_UK", 
-                                 p2 != "EA2_noUKB"
-                               )
-                    )
-rgs %>% 
-      filter(p2 != "EA3_excl_23andMe_UK", p2 != "EA2_noUKB") %>% 
-      ggplot(aes(rg, estimate)) + 
-        geom_point() +
-        my_vline +
-        my_hline +
-        labs(x = "Genetic correlation with EA3", y = "Effect on number of children")
-
-```
-
-Another way to examine the "earnings" theory of natural selection is to compare
-selection effects of polygenic scores with their genetic correlation with educational
-attainment (EA3). Since EA3 strongly predicts earnings, if earnings drives
-differences in fertility, we'd expect a correlation between the two sets of
-results. Figure \@ref(fig:plot-rgs-by-effect-size) shows this is so: the 
-correlation, after excluding EA2, is `r rgs_cor$estimate`. Genetic
-correlations were calculated using LD score regression from GWAS summary
-statistics.
 
@@
 \FloatBarrier
-
-## Model proofs
-
-```{r include-model-appendix, child="model-appendix.Rmd"}
-
-```
+## Further results
+### Selection effects on raw polygenic scores
+
+```{r calc-regs-controlled}
+
+drake::loadd(res_all)
+
+res_all_controlled <- res_all %>% 
+  select(reg.type, dep.var, term, estimate) %>% 
+  pivot_wider(names_from = c(reg.type, dep.var), values_from = estimate) %>% 
+  mutate(
+    consistent = sign(`raw_N siblings`) == sign(`raw_N children`) &
+                 sign(`raw_N siblings`) == sign(`controlled_N siblings`) &
+                 sign(`raw_N siblings`) == sign(`controlled_N children`)
+  )
+
+n_pgs  <- as.double(nrow(res_all_controlled))
+prop_consistent_controlled <- mean(res_all_controlled$consistent)
+controlled_smaller_sibs <- abs(res_all_controlled[["controlled_N siblings"]]) < 
+  abs(res_all_controlled[["raw_N siblings"]])
+controlled_smaller_children <- abs(res_all_controlled[["controlled_N children"]]) < 
+  abs(res_all_controlled[["raw_N children"]])
+median_prop_sibs <- median(abs(res_all_controlled[["controlled_N siblings"]]) /
+    abs(res_all_controlled[["raw_N siblings"]]))
+median_prop_children <- median(abs(res_all_controlled[["controlled_N children"]]) /
+    abs(res_all_controlled[["raw_N children"]]))
+
+```
+
+Figure \@ref(fig:plot-regs-controlled-pcs) compares selection effects on
+polygenic scores residualized for the top 100 principal components of the
+genetic data, to selection effects on raw, unresidualized polygenic scores. In
+siblings regressions, effect sizes are larger for raw scores  -- sometimes much
+larger, as in the case of height. `r sum(controlled_smaller_sibs)` out of 
+`r n_pgs` "raw" effect sizes have a larger absolute value than the corresponding
+"residualized" effect size. The median proportion between raw and controlled
+effect sizes is `r median_prop_sibs`. Among the children regressions, this no
+longer holds. Effect sizes are barely affected by controlling for principal
+components.
+
+Overall, `r prop_consistent_controlled * 100` per cent of effect sizes
+are consistently signed across all four regressions (on children and
+siblings, and with and without residualization).
+
+```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores on number of siblings/children."}
+
+drake::loadd(res_all)
+
+n_regs <- as.double(nrow(res_all))
+res_all %>% 
+      mutate(
+        term     = pretty_names(term),
+        term     = reorder_within(term, estimate, dep.var, order_abs(1)),
+        dep.var  = fct_inorder(dep.var),
+        reg.type = fct_recode(reg.type, 
+                               `Resid. for 100 PCs`  = "controlled", 
+                               Unresidualized        = "raw"
+                             )
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(y = term, x = estimate, fill = reg.type, 
+        colour = p < 0.05/{{n_regs}})) +
+        my_vline + 
+        facet_wrap(vars(dep.var), scales = "free") +
+        pval_geom_point(position = position_dodge(0.4), alpha = 0.8) +
+        my_labs() +
+        pval_scale() +
+        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
+        my_theme() +
+        scale_y_reordered() +
+        theme(
+          axis.text.x  = element_text(size = 7),
+          legend.text  = element_text(size = 8),
+          legend.title = element_text(size = 9)
+        )
+
+```
+
+```{r}
+drake::loadd(res_pcs)
+n_regs <- as.double(nrow(res_pcs))
+prop_sig <- res_pcs %>% 
+      group_by(dep.var) %>% 
+      summarize(sig = sum(p.value <= 0.05/nrow(res_pcs))) %>% 
+      pull(sig, name = dep.var)
+```
+
+To get a further insight into this we regress *n siblings* and *n children*
+on individual principal components. As Figure \@ref(fig:plot-pc-regs)
+shows, effects are larger and more significant in siblings regressions.
+`r prop_sig[["N siblings"]]` principal components significantly
+predicted number of siblings, while only `r prop_sig[["N children"]]`
+significantly predicted number of children.
+
+```{r plot-pc-regs, fig.cap = "Selection effects of 100 principal components of genetic data.
+Absolute effect sizes are plotted.
+Each dot represents one bivariate regression.
+Points are jittered on the Y axis.", fig.height = 2}
+
+res_pcs %>% 
+      mutate(
+        estimate = abs(estimate),
+        dep.var  = fct_relevel(dep.var, "N children")
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(estimate, dep.var, colour = p < 0.05/{{n_regs}})) + 
+        geom_jitter(height = 0.15, width = 0, alpha = 0.8) + 
+        scale_x_log10(labels = scales::label_number(accuracy = 0.0001)) +
+        labs(x = "Effect size (log scale)", y = "") + 
+        pval_scale(aesthetics = "colour")
+
+```
+
+\FloatBarrier
+\clearpage
+
+### Genetic correlations with EA3
+
+```{r plot-rgs-by-effect-size, fig.cap = "Selection effects plotted against genetic correlation with EA3."}
+
+# TODO: ref source of data.
+
+drake::loadd(rgs)
+drake::loadd(res_all)
+
+rgs <- left_join(rgs, 
+        res_all %>% filter(reg.type == "controlled", dep.var == "N children"), 
+        by = c("p2" = "term"))
+
+rgs_cor <- cor.test(~ rg + estimate, 
+                      data = rgs %>%
+                               filter(
+                                 p2 != "EA3_excl_23andMe_UK", 
+                                 p2 != "EA2_noUKB"
+                               )
+                    )
+rgs %>% 
+      filter(p2 != "EA3_excl_23andMe_UK", p2 != "EA2_noUKB") %>% 
+      ggplot(aes(rg, estimate)) + 
+        geom_point() +
+        my_vline +
+        my_hline +
+        labs(x = "Genetic correlation with EA3", y = "Effect on number of children")
+
+```
+
+Another way to examine the "earnings" theory of natural selection is to compare
+selection effects of polygenic scores with their genetic correlation with educational
+attainment (EA3). Since EA3 strongly predicts earnings, if earnings drives
+differences in fertility, we'd expect a correlation between the two sets of
+results. Figure \@ref(fig:plot-rgs-by-effect-size) shows this is so: the 
+correlation, after excluding EA2, is `r rgs_cor$estimate`. Genetic
+correlations were calculated using LD score regression from GWAS summary
+statistics.
 
@@
 
+## Model proofs
+
+```{r include-model-appendix, child="model-appendix.Rmd"}
+
+```
+
+\clearpage
+\FloatBarrier
+
 # References

### 2022-01-02 why-natural-selection.Rmd
Run parental AFLB tercile regressions. Don't plot respondents regression controlling for AFLB.

@@
 live birth is a choice variable, which is endogenous to human-capital-related
-polygenic scores. Nevertheless, differences in equation
-\@ref(eq:regression) across subgroups constrain the set of possible explanations
-for natural selection. As we show below, the economic theory predicts 
-these differences, while alternative explanations do not.
+polygenic scores (Appendix Figure \@ref(fig:plot-age-flb-dv)). Nevertheless,
+differences in equation \@ref(eq:regression) across subgroups constrain the set
+of possible explanations for natural selection. As we show below, a simple
+economic model predicts the differences we find.
 
@@
 [@Barban_2016], and there is a close link between this variable and number of
-children born. Figure \@ref(fig:plot-age-flb)a shows effect sizes estimated
+children born. Figure \@ref(fig:plot-age-flb) shows effect sizes estimated
 separately for each tercile of AFLB. Several effects are strikingly different
@@
 
-To investigate this further, we estimate equation \@ref(eq:regression)
-*controlling* for AFLB, again among females (Figure \@ref(fig:plot-age-flb)b).
-In `r opp_signed` out of `r length(score_names)` cases, effects change sign when
-controls are added. The correlation between effect sizes controlling for AFLB,
-and raw effect sizes, is `r cor_raw_flb`. Thus, selection effects seem to come
-through two opposing channels: an effect on AFLB, and an opposite-signed effect
-on number of children controlling for AFLB. Again, we do not claim that AFLB
-is exogenous to an individual's polygenic scores. Indeed, our analyses show
-that it is not (Appendix Figure \@ref(fig:plot-age-flb-dv)). Rather, we argue
-that AFLB mediates part of the relationship between polygenic scores and
-fertility, and that once this is controlled for, the remaining part of the
-relationship has the opposite sign.
-
-```{r plot-age-flb, fig.cap = "Selection effects by age at first live birth, and controlling for age at first live birth (women only).", fig.ncol = 1,  fig.subcap = c("Effect sizes by age at first live birth terciles (women only)", "Effect sizes controlling for age at first live birth.
-Effect sizes without controls (for women only) shown for comparison."), fig.align = "center"}
+To investigate this further, we estimate equation \@ref(eq:regression) among
+females, *controlling* for AFLB. In `r opp_signed` out of 
+`r length(score_names)` cases, effects change sign when controls are added. The
+correlation between effect sizes controlling for AFLB, and raw effect sizes, is
+`r cor_raw_flb`. Thus, selection effects seem to come through two opposing
+channels: an effect on AFLB, and an opposite-signed effect on number of children
+after AFLB is controlled for.
+
+```{r plot-age-flb, fig.cap = "Selection effects by age at first live birth terciles (women only).", fig.align = "center"}
 
@@
 
-n_regs <- as.double(nrow(res_age_flb))
-
-res_combined %>% 
-      mutate(
-        score_name = pretty_names(score_name),
-        score_name = fct_reorder(score_name, estimate, order_abs(2)),
-        Regression = ifelse(reg.type == "raw", "Uncontrolled", "Controlled for AFLB")
-      ) %>% 
-      standard_ggplot(score_col = score_name, 
-                      fill_col = Regression, n_regs = n_regs)
-
 ```
@@
 
-We also examine selection effects among respondents' parents, using
-information on respondents' number of siblings. Effect sizes of polygenic
-scores are highly correlated across the two generations. Median-splitting
-respondents by year of birth, we find little evidence of change in effect sizes
-among either the parents' or the respondents' generation. We check whether 
-selection effects differ by socio-economic status in the
-parents' generation. We have no information about parents' income, so we use the
-1971 Townsend deprivation score of respondents' birthplace as a proxy
-[@townsend1987deprivation]. Results show the same pattern as for respondents:
-effect sizes are larger and more often significant in the most deprived areas
-(Appendix Figure \@ref(fig:plot-siblings-townsend)); effects change
-sign when controlling for AFLB (Appendix Figure \@ref(fig:plot-age-birth-parents)). Lastly, the siblings data 
-lets us check for a "quantity-quality tradeoff" between
-number of children (of respondents' parents) and number of grandchildren. We
-do not find any: in fact, the correlation between *number of siblings* and 
-*number of children* is positive ($\rho$ = `r sib_chn_cor$estimate`, 
-$p < 2 \times 10^{-16}$).
+We also examine selection effects among respondents' parents, using information
+on respondents' number of siblings. Effect sizes of polygenic scores are highly
+correlated across the two generations. Median-splitting respondents by year of
+birth, we find little evidence of change in effect sizes among either the
+parents' or the respondents' generation. We also check whether selection effects
+vary by socio-economic status and AFLB in the parents' generation. We have no
+information about parents' income, so we use the 1971 Townsend deprivation score
+of respondents' birthplace as a proxy [@townsend1987deprivation]. Results show
+the same pattern as for respondents. Effect sizes are larger and more often
+significant in the most deprived areas (Appendix Figure
+\@ref(fig:plot-siblings-townsend)). Effects are larger among younger parents,
+and change sign when controlling for AFLB, among both mothers and fathers
+(Appendix Figure \@ref(fig:plot-age-birth-parents)). Lastly, we check for a
+"quantity-quality tradeoff" between parents' number of children and number of
+grandchildren. We do not find any: in fact, the correlation between *number of
+siblings* and *number of children* is positive ($\rho$ = 
+`r sib_chn_cor$estimate`, $p < 2 \times 10^{-16}$).
 
@@
 the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
-to 1. Letting $u(\cdot)$ take the constant relative risk aversion (CRRA) form
+to 1. Letting $u(\cdot)$ take the constant relative risk aversion form
 $u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
@@
 [^one-period-results]: Predictions 1-3 also hold in the one-period model with
-CRRA utility. Our empirical results are actually stronger than prediction 5, in
-that correlations with fertility are *reversed* at higher AFLB. This prediction
-can be accommodated in the model if children have a money cost as well as a time
-cost. Appendix Figure \@ref(fig:N-plot-with-m) shows an example.
+constant relative risk aversion. Our empirical results are actually stronger
+than prediction 5, in that correlations with fertility are *reversed* at higher
+AFLB. This prediction can be accommodated in the model if children have a money
+cost as well as a time cost (Appendix Figure \@ref(fig:N-plot-with-m)).
 
@@
     who start fertility in period 2                   fertility later 
-    ($N_{1}^{*} = 0$) than among those who            (Figure \@ref(fig:plot-age-flb)a).
+    ($N_{1}^{*} = 0$) than among those who            (Figure \@ref(fig:plot-age-flb)).
     start fertility in period 1 
@@
 
+\clearpage
+\FloatBarrier
+## Parents: area deprivation
+
 Figure \@ref(fig:plot-siblings-townsend) plots effects on *number of siblings* by
@@
 \FloatBarrier
-## Selection controlling for age at first live birth: respondents' parents
+## Parents: age at first live birth
 
@@
 
-Among the parents' generation, we can control for age at first live
-birth using the subsets of respondents who reported their mother's or
-father's age, and who had no elder siblings. We run regressions on
-*number of siblings* on these subsets, controlling for either parent's
-age at their birth. Figure \@ref(fig:plot-age-birth-parents) shows the
-results. Effect sizes are very similar, whether controlling for father's
-or mother's age. As in the respondents' generation, effect sizes
-are negatively correlated with the effect sizes from bivariate
-regressions without the age at birth control (father's age at birth: 
-$\rho$ `r cor_unc_fath`; mother's age at birth: $\rho$ `r cor_unc_moth`).
+Among the parents' generation, we can control for age at first live birth using
+the subsets of respondents who reported their mother's or father's age, and who
+had no elder siblings. We run regressions on *number of siblings* on these
+subsets. Figure \@ref(fig:plot-age-birth-parents-cross) shows selection effects
+by terciles of age at first live birth, for mothers and fathers. As in the
+respondents' generation, effect sizes are larger for younger parents. This
+holds for both sexes.
+
+```{r plot-age-birth-parents-cross, fig.cap = "Selection effects (parents' generation) by age at first live birth terciles.", fig.subcap = c("Mothers", "Fathers"), fig.ncol = 1}
+
+drake::loadd(res_age_flb_mothers_cross)
+drake::loadd(res_age_flb_fathers_cross)
+
+famhist_bo1 <- famhist %>% 
+                 filter(birth_order == 1) %>%
+                 select(n_siblings, whr_combined, moth_age_flb_cat, 
+                          fath_age_flb_cat)
+
+# this counts each cross term as a separate test:
+n_regs <- as.double(nrow(res_age_flb_mothers_cross))
+
+res_age_flb_mothers_cross %>% 
+      mutate(
+        `Age at first live birth` = gsub("moth_age_birth_cat(.*):.*", "\\1", term),
+        `Age at first live birth` = add_n(`Age at first live birth`, 
+                                          "moth_age_flb_cat", dv = "n_sibs",
+                                          data = famhist_bo1)
+      ) %>% 
+      standard_ggplot(score_col = score_name, 
+                      fill_col = `Age at first live birth`, n_regs = n_regs,
+                      fill_direction = -1)
+
+# this counts each cross term as a separate test:
+n_regs <- as.double(nrow(res_age_flb_fathers_cross))
+
+res_age_flb_fathers_cross %>% 
+      mutate(
+        `Age at first live birth` = gsub("fath_age_birth_cat(.*):.*", "\\1", term),
+        `Age at first live birth` = add_n(`Age at first live birth`, 
+                                          "fath_age_flb_cat", dv = "n_sibs",
+                                          data = famhist_bo1)
+      ) %>% 
+      standard_ggplot(score_col = score_name, 
+                      fill_col = `Age at first live birth`, n_regs = n_regs,
+                      fill_direction = -1)
+
+rm(famhist_bo1)
+
+```
+
+Figure \@ref(fig:plot-age-birth-parents) shows the regressions controlling for
+either parent's age at their birth. Effect sizes are very similar, whether
+controlling for father's or mother's age. As in the respondents' generation,
+effect sizes are negatively correlated with the effect sizes from bivariate
+regressions without the age at birth control (father's age at birth: $\rho$ 
+`r cor_unc_fath`; mother's age at birth: $\rho$ `r cor_unc_moth`).
 

### 2022-01-02 _drake.R

@@ -7,6 +7,7 @@ suppressPackageStartupMessages({
   library(readr)
   library(dplyr)
   loadNamespace("future")
+  loadNamespace("santoku")
 })
 
 source("make-data.R") 
@@ -75,6 +76,19 @@ plan <- drake_plan(
                      names(famhist) <- gsub("_resid$", "", names(famhist))
                      
                      famhist <- add_fertility_prs(famhist, fertility_data_dir)
+                     
+                     # we put these here because they're tightly tied to this
+                     # particular project
+                     mab_terciles <- quantile(famhist$moth_age_birth[famhist$birth_order == 1], 
+                                              1:2/3, na.rm = TRUE)
+                     famhist$moth_age_birth_cat <- santoku::chop(famhist$moth_age_birth, 
+                                                                 mab_terciles, 
+                                                                 labels = lbl_discrete("-"))
+                     fab_terciles <- quantile(famhist$fath_age_birth[famhist$birth_order == 1], 
+                                              1:2/3, na.rm = TRUE)
+                     famhist$fath_age_birth_cat <- santoku::chop(famhist$fath_age_birth, 
+                                                                 fab_terciles, 
+                                                                 labels = lbl_discrete("-"))
                      
                      famhist
                    },
@@ -341,6 +355,36 @@ plan <- drake_plan(
     res$control    <- vars$control[as.numeric(res$id)]
     res$id <- NULL
     res %<>% filter(term != "(Intercept)")
+    
+    res
+  },
+  
+  res_age_flb_mothers_cross = {
+    res <- map_dfr(score_names, 
+                   ~run_regs_fml(
+                     fml        = "n_children ~ moth_age_birth_cat + {score_name}:moth_age_birth_cat",
+                     score_name = .x,
+                     famhist    = famhist,
+                     subset     = quote(birth_order == 1)
+                   ),
+                   .id = "score_name"
+    )
+    res %<>% filter(grepl(":", term))
+    
+    res
+  },
+  
+  res_age_flb_fathers_cross = {
+    res <- map_dfr(score_names, 
+                   ~run_regs_fml(
+                     fml        = "n_children ~ fath_age_birth_cat + {score_name}:fath_age_birth_cat",
+                     score_name = .x,
+                     famhist    = famhist,
+                     subset     = quote( birth_order == 1)
+                   ),
+                   .id = "score_name"
+    )
+    res %<>% filter(grepl(":", term))
     
     res
   },

### 2022-01-02 why-natural-selection.Rmd
bugfixes

@@
 ## Parents' generation {#sec-parents}
+### Selection effects
 
@@
 \FloatBarrier
-## Parents: area deprivation
+### Area deprivation
 
@@
 \FloatBarrier
-## Parents: age at first live birth
+### Age at first live birth
 
@@
                  filter(birth_order == 1) %>%
-                 select(n_siblings, whr_combined, moth_age_flb_cat, 
-                          fath_age_flb_cat)
+                 select(n_sibs, whr_combined, moth_age_birth_cat, 
+                          fath_age_birth_cat)
 
@@
         `Age at first live birth` = add_n(`Age at first live birth`, 
-                                          "moth_age_flb_cat", dv = "n_sibs",
+                                          "moth_age_birth_cat", dv = "n_sibs",
                                           data = famhist_bo1)
@@
         `Age at first live birth` = add_n(`Age at first live birth`, 
-                                          "fath_age_flb_cat", dv = "n_sibs",
+                                          "fath_age_birth_cat", dv = "n_sibs",
                                           data = famhist_bo1)
@@
 ```{r plot-res-fe-fertility, fig.cap = "Selection effects controlling for sibling-group fixed effects, with and without a control for education (left education before 16, 16-18, or after 18).
-Each set of results is from a single regression of 'n children' on 29 polygenic scores.
+Each set of results is from a single regression of \\emph{n children} on 29 polygenic scores.
 Standard errors clustered by sibling group."}

### 2022-01-03 why-natural-selection.Rmd
Separate out parent weights from age/qual weights. Rewrite PCs section in appendix based on new results.

@@
 
-```{r}
+To get a further insight into this we regress *n siblings* and *n children*
+on individual principal components. Figure \@ref(fig:plot-pc-regs)
+shows the results. Labels show the top principal components. These
+have larger effect sizes in siblings regressions. One possibility is that 
+the parents' generation was less geographically mobile, and so geographic
+patterns of childrearing were more correlated with principal components, which
+partly capture the location of people's ancestors.
+
+```{r plot-pc-regs, fig.cap = "Selection effects of 100 principal components of genetic data.
+Each dot represents one bivariate regression.
+Absolute effect sizes are plotted.
+Points are jittered on the Y axis.
+Top principal components are labelled."}
+
 drake::loadd(res_pcs)
 n_regs <- as.double(nrow(res_pcs))
-prop_sig <- res_pcs %>% 
-      group_by(dep.var) %>% 
-      summarize(sig = sum(p.value <= 0.05/nrow(res_pcs))) %>% 
-      pull(sig, name = dep.var)
-```
-
-To get a further insight into this we regress *n siblings* and *n children*
-on individual principal components. As Figure \@ref(fig:plot-pc-regs)
-shows, effects are larger and more significant in siblings regressions.
-`r prop_sig[["N siblings"]]` principal components significantly
-predicted number of siblings, while only `r prop_sig[["N children"]]`
-significantly predicted number of children.
-
-```{r plot-pc-regs, fig.cap = "Selection effects of 100 principal components of genetic data.
-Absolute effect sizes are plotted.
-Each dot represents one bivariate regression.
-Points are jittered on the Y axis.", fig.height = 2}
+
+pj <- position_jitter(height = 0.15, width = 0, seed = 12345)
 
@@
         estimate = abs(estimate),
-        dep.var  = fct_relevel(dep.var, "N children")
+        dep.var  = fct_relevel(dep.var, "N children"),
+        top_pcs  = ifelse(as.numeric(pc_name) <= 5, term, "")
       ) %>% 
@@
       ggplot(aes(estimate, dep.var, colour = p < 0.05/{{n_regs}})) + 
-        geom_jitter(height = 0.15, width = 0, alpha = 0.8) + 
+        geom_point(alpha = 0.8, position = pj) +
+        geom_text(aes(estimate, dep.var, label = top_pcs), position = pj, 
+                    hjust = "right", color = "black", size = 3) +
         scale_x_log10(labels = scales::label_number(accuracy = 0.0001)) +

### 2022-01-03 _drake.R

@@ -36,7 +36,7 @@ fertility_data_dir <- file.path(data_dir, "fertility_PRS")
 
 weighting_scheme_syms <- rlang::syms(c("flb_weights", "age_qual_weights", 
                         "msoa_weights"))
-period_weight_syms <- rlang::syms((c("age_qual_weights", "parent_weights")))
+period_weight_syms <- rlang::syms((c("age_qual_weights", "parent_aq_weights")))
 
 plan <- drake_plan(
   
@@ -94,6 +94,11 @@ plan <- drake_plan(
                    },
                     format = "fst_tbl"
                   ),
+  
+  famhist_pw = target(
+                 inner_join(famhist, parent_weights, by = "f.eid"),
+                 format = "fst_tbl"
+               ),
 
   famhist_msoa = target(
                    find_containing_msoas(famhist, msoa_shapefile), 
@@ -143,7 +148,9 @@ plan <- drake_plan(
   
   age_qual_weights = weight_by_census_age_qual(famhist, census_age_qual),
   
-  parent_weights = weight_parents(famhist, input_weights = age_qual_weights),
+  parent_weights = weight_parents(famhist),
+  
+  parent_aq_weights = weight_parents(famhist, input_weights = age_qual_weights),
   
   msoa_weights = weight_by_census_msoa(famhist, census_msoa, famhist_msoa),
   
@@ -162,7 +169,8 @@ plan <- drake_plan(
     res_sibs <- run_regs_basic(
             dep_var     = "n_sibs", 
             score_names = score_names,
-            famhist     = famhist
+            famhist     = famhist_pw,
+            weights     = famhist_pw$weights
           )
     res_chn <- run_regs_basic(
             dep_var     = "n_children", 
@@ -178,12 +186,11 @@ plan <- drake_plan(
   },
   
   res_quadratic = {
-    famhist %<>% inner_join(age_qual_weights, by = "f.eid")
     map_dfr(score_names, 
               ~run_regs_fml(
                              "n_children ~ {score_name} + I({score_name}^2)", 
                              score_name = .x,
-                             famhist    = famhist,
+                             famhist    = famhist_pw,
                              weights    = quote(weights)
                            ),
               .id = "score_name"
@@ -191,12 +198,11 @@ plan <- drake_plan(
   },
   
   res_sibs_quadratic = {
-    famhist %<>% inner_join(parent_weights, by = "f.eid")
     map_dfr(score_names, 
             ~run_regs_fml(
[…]

### 2022-01-03 regressions.R

@@ -57,11 +57,13 @@ calc_pgs_over_time <- function (famhist, score_names) {
 }
 
 
-run_regs_basic <- function (dep_var, score_names, famhist, subset = NULL) {
+run_regs_basic <- function (dep_var, score_names, famhist, subset = NULL, 
+                              weights = NULL) {
   
   run_reg <- function (score_name) {
     fml <- as.formula(glue("{dep_var} ~ {score_name}_raw"))
-    reg_bv <- tidy(lm(fml, famhist, subset = eval(subset)), conf.int = TRUE) 
+    reg_bv <- tidy(lm(fml, famhist, subset = eval(subset), 
+                        weights = weights), conf.int = TRUE) 
     reg_bv <- mutate(reg_bv, term = gsub("_raw", "", term))
     reg_bv
   }
@@ -69,7 +71,8 @@ run_regs_basic <- function (dep_var, score_names, famhist, subset = NULL) {
   
   run_resid_reg <- function (score_name) {
     fml_resid <- as.formula(glue("{dep_var} ~ {score_name}"))
-    reg_resid <- tidy(lm(fml_resid, famhist, subset = eval(subset)), conf.int = TRUE) 
+    reg_resid <- tidy(lm(fml_resid, famhist, subset = eval(subset), 
+                           weights = weights), conf.int = TRUE) 
     reg_resid
   }
   resid_res <- map_dfr(score_names, run_resid_reg, .id = "score_name")
@@ -81,14 +84,15 @@ run_regs_basic <- function (dep_var, score_names, famhist, subset = NULL) {
 }
 
 
-run_regs_pcs <- function (dep_var, famhist, pcs, subset = NULL) {
+run_regs_pcs <- function (dep_var, famhist, pcs, subset = NULL, weights = NULL) {
   pc_names <- grep("PC", names(pcs), value = TRUE)
   
   fh_short <- join_famhist_pcs(famhist[c("f.eid", dep_var, "kids_ss")], pcs)
   
   run_reg_pc <- function(pc_name) {
     fml <- as.formula(glue("{dep_var} ~ {pc_name}"))
-    reg <- tidy(lm(fml, fh_short, subset = eval(subset)), conf.int = TRUE) 
+    reg <- tidy(lm(fml, fh_short, subset = eval(subset), 
+                     weights = eval(weights)), conf.int = TRUE) 
     reg %<>% filter(term != "(Intercept)")
   }
   
[…]

### 2022-01-03 weight-data.R

@@ -38,8 +38,12 @@ weight_by_ghs <- function (fml, ghs_subset, famhist) {
 }
 
 
-weight_parents <- function (famhist, input_weights) {
-  famhist <- inner_join(famhist, input_weights, by = "f.eid")
+weight_parents <- function (famhist, input_weights = NULL) {
+  if (is.null(input_weights)) {
+    famhist$weights <- 1
+  } else {
+    famhist <- inner_join(famhist, input_weights, by = "f.eid")
+  }
   n_sibs_are_up_to_8 <- ! is.na(famhist$n_sibs) & famhist$n_sibs <= 8
   n_sibs_up_to_8 <- famhist$n_sibs[n_sibs_are_up_to_8]
   wts <- famhist$weights[n_sibs_are_up_to_8]

### 2022-01-03 why-natural-selection.Rmd
Mention weights in tables and figures.

@@
 We consider three alternative theories that might explain our results. First,
-welfare benefits which incentivize child-bearing might be taken up more
-among low-income individuals. However, our effect sizes appear unchanged
-over a large span of twentieth-century history (tables \@ref(tab:tbl-siblings-by-period) and 
-\@ref(tab:tbl-children-by-period)), during which
-government spending on child-related benefits varied considerably
-[@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
-affect fertility [@gauthier2007impact]. A second alternative theory is that
-polygenic scores correlate with the motivation to have children, i.e. parameter $a$
-in the model [cf. @jones2008fertility]. This theory would not explain why our
-results are weaker at higher incomes and education levels.[^note] A third
-alternative is that traits under selection are linked to externalizing behaviour
-and risk-seeking. This might be partially captured by our risk-aversion
-parameter $\sigma$, which affects fertility by changing the marginal utility of
-income, but a more direct channel is risky sexual behaviour
-[@mills2021identification]. The data here provide some support for this: scores
-which might plausibly be linked to externalizing behaviour, like ADHD and
-younger age at smoking initiation, are selected for. However, this theory is
-less good at explaining variation in selection across the full range of scores,
-including physical measures such as waist-hip ratio and BMI. We test this theory
-directly by re-estimating equation \@ref(eq:regression) controlling for a
-measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
-sizes between regressions with and without controls is `r median(ratio)`; all
-scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
-so when controlling for risk attitude. This non-result could simply reflect the
+welfare benefits which incentivize child-bearing might be taken up more among
+low-income individuals. However, our effect sizes appear unchanged over a large
+span of twentieth-century history (tables \@ref(tab:tbl-siblings-by-period) and
+\@ref(tab:tbl-children-by-period)), during which government spending on
+child-related benefits varied considerably [@hoc1999socsec]. In general, there
+is only weak evidence that welfare benefits affect fertility
+[@gauthier2007impact]. A second alternative theory is that polygenic scores
+correlate with the motivation to have children, i.e. parameter $a$ in the model
+[cf. @jones2008fertility]. This theory would not explain why selection effects
+are smaller at higher incomes and education levels.[^note] A third alternative
+is that traits under selection are linked to externalizing behaviour and
+risk-seeking. This might be partially captured by our risk-aversion parameter
+$\sigma$, which affects fertility by changing the marginal utility of income,
+but a more direct channel is risky sexual behaviour [@mills2021identification].
+The data here provide some support for this: scores which might plausibly be
+linked to externalizing behaviour, like ADHD and younger age at smoking
+initiation, are selected for. However, this theory is less good at explaining
+variation in selection across the full range of scores, including physical
+measures such as waist-hip ratio and BMI. We test this theory directly by
+re-estimating equation \@ref(eq:regression) controlling for a measure of risk
+attitude (UK Biobank field 2040).  The median ratio of effect sizes between
+regressions with and without controls is `r median(ratio)`; all scores which are
+significant at $p < 0.05/33$ in uncontrolled regressions remain so when
+controlling for risk attitude. This non-result could simply reflect the
 imprecision of the risk attitude measure, which is a single yes/no question.
-However, this measure *does* predict the overall number of
-children, highly significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of
-33 regressions). Given that, and the statistical power we get from our sample
-size, we believe that the non-result is real: while risk attitude does predict
-fertility in the sample, it is not an important channel for natural selection.
-
-[^note]: Indeed in the one-period model, while $\frac{dN^{*}}{da} > 0$,
-this effect becomes stronger, not weaker, at higher wages.
+However, this measure *does* predict the overall number of children, highly
+significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of 33 regressions).
+Given that, and the statistical power we get from our sample size, we believe
+that the non-result is real: while risk attitude does predict fertility in the
+sample, it is not an important channel for natural selection.
+[^note]: Indeed in the one-period model, while $\frac{dN^{*}}{da} > 0$, this
+effect becomes stronger, not weaker, at higher wages.
 
@@
 
-```{r plot-res-sibs, fig.cap = "Selection effects, respondents' parents vs. respondents."}
+```{r plot-res-sibs, fig.cap = "Selection effects, respondents' parents vs.
+respondents.
+Age/qualification weights.
+Parental generation weights multiplied by 1/number of siblings.
+Respondents' regression excludes childless respondents."}
 drake::loadd(res_sibs_parent_weights)
@@
       count() %>% 
-      period_table("Change in selection effects between parents of early and late born respondents (regressions on number of siblings).")
+      period_table("Change in selection effects between parents of early and late born respondents (regressions on number of siblings). Weights: Age/Qualification times 1/number of siblings.")
 
@@
       count() %>% 
-      period_table("Change in selection effects between early and late born respondents (regressions on number of children).")
+      period_table("Change in selection effects between early and late born respondents (regressions on number of children). Weights: Age/Qualification.")
 ```
@@
 
-```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by Townsend deprivation quintile of birth area.", fig.align = "center"}
+```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by Townsend deprivation quintile of birth area.
+Higher = more deprived.
+Estimates weighted by 1/number of siblings.", fig.align = "center"}
 
@@
 by terciles of age at first live birth, for mothers and fathers. As in the
-respondents' generation, effect sizes are larger for younger parents. This
-holds for both sexes.
-
-```{r plot-age-birth-parents-cross, fig.cap = "Selection effects (parents' generation) by age at first live birth terciles.", fig.subcap = c("Mothers", "Fathers"), fig.ncol = 1}
+respondents' generation, effect sizes are smaller, or even oppositely signed, for older parents. Importantly, this holds for both sexes.
+
+```{r plot-age-birth-parents-cross, fig.cap = "Selection effects (parents' generation) by age at first live birth terciles.
+Estimates weighted by 1/number of siblings.", fig.subcap = c("Mothers", "Fathers"), fig.ncol = 1}
 
@@
 
-```{r plot-age-birth-parents, fig.cap = "Selection effects (parents' generation) among eldest siblings, controlling for parents' age at birth."}
+```{r plot-age-birth-parents, fig.cap = "Selection effects (parents' generation) among eldest siblings, controlling for parents' age at birth. Estimates weighted by 1/number of siblings."}
 
@@
 
-```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at respondent's birth, eldest siblings."}
+```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at respondent's birth, eldest siblings. Estimates weighted by 1/number of siblings."}
 
@@
 
-```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores on number of siblings/children."}
+```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores on number of siblings/children. Sibling estimates weighted by 1/number of siblings."}
 
@@
 Each dot represents one bivariate regression.
+Sibling estimates weighted by 1/number of siblings.
 Absolute effect sizes are plotted.

### 2022-01-03 why-natural-selection.Rmd
Tweaks to text

@@
 traits which are valuable in labour markets -- can increase or decrease his or
-her fertility. This can lead to selection on inherited
-variants which are linked to human capital. Furthermore, the economic theory predicts
-systematic variation in the relative strength of the income and substitution
-effects across different social groups.
+her fertility. This can lead to selection on inherited variants which are linked
+to human capital. Furthermore, the economic theory predicts systematic variation
+in the relative strength of the income and substitution effects across different
+social groups.
 
@@
 selection over multiple generations could lead to substantive effects in the
-long run. Next, we examine correlations with fertility in different subgroups. Across
-the board, selection effects are stronger in groups with lower
-income and less education, among younger parents, people not living with a partner,
-and people with more lifetime sexual partners. Outside these groups, effects are
-weaker and often statistically insignificant. In some subgroups, the
-direction of selection is even reversed.
+long run. Next, we examine correlations with fertility in different subgroups.
+Across the board, selection effects are stronger in groups with lower income and
+less education, among younger parents, people not living with a partner, and
+people with more lifetime sexual partners. Outside these groups, effects are
+weaker and often statistically insignificant. In some subgroups, the direction
+of selection is even reversed.
 
@@
 changes could reflect natural selection within the UK population, but also
-emigration, or ascertainment bias within the sample. Respondents have higher
-income and are better educated than the UK population, and they may also differ
-on other unobserved characteristics [@10.1093/aje/kwx246]. Since richer and more
-educated people also live longer, this bias could also increase with age.
-Lastly, the sample sex ratio skews `r female_ratio * 100`% female.
+emigration, or ascertainment bias within the sample [@10.1093/aje/kwx246].
 
@@
 an individual's level of human capital. For now, we simply identify this with
-his or her wage $W$. Raising a child takes time $b$. People maximize utility
-from the number of children $N$ and from income $Y\equiv(1-bN)W$. An individual's
-payoff is
+his or her wage $W$. Raising a child takes time $b$. People maximize utility $U$
+from the number of children $N$ and from income $Y\equiv(1-bN)W$:
 
@@
                      # 0.025 not 0.05 because two-sided
-                     sig  = abs(statistic_ind) > qnorm(1 - 0.025/33),
-                     star = ifelse(sig, "*", "")
+                     sig  = abs(statistic_ind) > qnorm(1 - 0.025/33)
                    )
@@
 
-```
-
-Second, we directly test whether the correlation between each polygenic score
-and fertility is mediated by educational attainment (Appendix Table
-\@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the indirect effect of
-the score on fertility *via* educational attainment is significantly different
-from zero ($p < 0.05/33$).
+median_prop <- median(res_mediation$prop[res_mediation$prop > 0 & 
+                                           res_mediation$sig])
+
+```
+
+Second, we run a mediation analysis to directly test whether the correlation
+between each polygenic score and fertility is mediated by educational attainment
+(Appendix Table \@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the
+indirect effect of the score on fertility *via* educational attainment is
+significantly different from zero ($p < 0.05/33$). Among these scores, the
+median proportion of the total effect explained by the indirect effect is 
+`r percent(median_prop)`.
 
@@
 including people with lower income, lower education, younger first parenthood,
-and more lifetime sexual partners. Among older mothers (22+), selection effects
+and more lifetime sexual partners. Among mothers aged 22+, selection effects
 are reversed. Furthermore, the size of selection effects on a polygenic score
@@
 attainment of 0.4 [@branigan2013variation], gives the proportion of variance in
-income attributable to the "true" polygenic score for educational attainment as
-`r r2_income_true_psea` among respondents. Natural selection would raise this
-proportion to `r r2_income_true_psea_wt` among respondents' children,
+household income attributable to the "true" polygenic score for educational
+attainment as `r r2_income_true_psea` among respondents. Natural selection would
+raise this proportion to `r r2_income_true_psea_wt` among respondents' children,
 counterfactually holding environmental conditions constant. Similarly, since

### 2022-01-03 why-natural-selection.Rmd
Small tweaks

@@
 and change sign when controlling for AFLB, among both mothers and fathers
-(Appendix Figure \@ref(fig:plot-age-birth-parents)). Lastly, we check for a
+(Appendix Figures \@ref(fig:plot-age-birth-parents-cross), 
+\@ref(fig:plot-age-birth-parents)). Lastly, we check for a
 "quantity-quality tradeoff" between parents' number of children and number of
@@
 
-Balancing selection reduces variance in the trait under selection, 
-while diversifying selection increases variance. To check for these,
-we rerun equation \@ref(eq:regression), adding a quadratic term in $PGS_i$,
-and using our age/qualification weights. Scores for hip circumference and drinks
-per week show significant balancing selection ($p < 0.05/33$, negative
-coefficient). Scores for educational attainment (EA2 and EA3) show significant
-diversifying selection ($p < 0.05/33$, positive coefficient), which reduces the
-strength of selection against educational attainment at very high levels of the 
-PGS. Figure
+Balancing selection reduces variance in the trait under selection, while
+diversifying selection increases variance. To check for these, we rerun equation
+\@ref(eq:regression), adding a quadratic term in $PGS_i$, and using our
+age/qualification weights. Scores for hip circumference show significant
+balancing selection ($p < 0.05/33$, negative coefficient on quadratic term).
+Scores for educational attainment (EA2 and EA3) show significant diversifying
+selection ($p < 0.05/33$, positive coefficient), which reduces the strength of
+selection against educational attainment at very high levels of the PGS. Figure
 \@ref(fig:plot-purifying) plots predicted number of children against polygenic
@@
 age/qualification weights multiplied by the inverse of number of siblings.
-Scores for EA2 and EA3 again show significant diversifying selection 
-($p < 0.05/33$, positive coefficient). Other scores including hip circumference 
-and drinks per week are not significant.
-
-```{r plot-purifying, fig.align = "center", fig.cap = "Purifying/diversifying selection: predicted number of children by polygenic score.", fig.height = 5}
+Scores for EA2 and EA3 again show significant diversifying selection ($p <
+0.05/33$, positive coefficient). Other scores including hip circumference were 
+not significant.
+
+```{r plot-purifying, fig.align = "center", fig.cap = "Purifying/diversifying selection: predicted number of children by polygenic score.", fig.height = 3}
 
@@
 
-stopifnot(sig == c("dpw_substance_use", "EA2_noUKB", "EA3_excl_23andMe_UK",
-                     "hip_combined"))
+stopifnot(sig == c("EA2_noUKB", "EA3_excl_23andMe_UK", "hip_combined"))
 
@@
 
-op <- par(mfrow = c(2,2))
+op <- par(mfrow = c(1, 3))
 for (score in sig) {

### 2022-01-03 why-natural-selection.Rmd
Add proof that dN^*/da gets larger at higher levels of human capital

@@
 that part of the correlation with fertility is indeed *via* education.
-Thus, contemporary natural selection on polygenic scores can be explained by their correlation with earnings-increasing human capital.
+Thus, contemporary natural selection on polygenic scores can be explained by scores' correlation with earnings-increasing human capital.
 
@@
 [cf. @jones2008fertility]. This theory would not explain why selection effects
-are smaller at higher incomes and education levels.[^note] A third alternative
-is that traits under selection are linked to externalizing behaviour and
-risk-seeking. This might be partially captured by our risk-aversion parameter
-$\sigma$, which affects fertility by changing the marginal utility of income,
-but a more direct channel is risky sexual behaviour [@mills2021identification].
-The data here provide some support for this: scores which might plausibly be
-linked to externalizing behaviour, like ADHD and younger age at smoking
-initiation, are selected for. However, this theory is less good at explaining
-variation in selection across the full range of scores, including physical
-measures such as waist-hip ratio and BMI. We test this theory directly by
-re-estimating equation \@ref(eq:regression) controlling for a measure of risk
-attitude (UK Biobank field 2040).  The median ratio of effect sizes between
-regressions with and without controls is `r median(ratio)`; all scores which are
-significant at $p < 0.05/33$ in uncontrolled regressions remain so when
-controlling for risk attitude. This non-result could simply reflect the
+are smaller at higher incomes and education levels. (Indeed, in the model, the
+effect of $a$ on fertility becomes stronger at higher levels of human capital.) A
+third alternative is that traits under selection are linked to externalizing
+behaviour and risk-seeking. This might be partially captured by our
+risk-aversion parameter $\sigma$, which affects fertility by changing the
+marginal utility of income, but a more direct channel is risky sexual behaviour
+[@mills2021identification]. The data here provide some support for this: scores
+which might plausibly be linked to externalizing behaviour, like ADHD and
+younger age at smoking initiation, are selected for. However, this theory is
+less good at explaining variation in selection across the full range of scores,
+including physical measures such as waist-hip ratio and BMI. We test this theory
+directly by re-estimating equation \@ref(eq:regression) controlling for a
+measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
+sizes between regressions with and without controls is `r median(ratio)`; all
+scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
+so when controlling for risk attitude. This non-result could simply reflect the
 imprecision of the risk attitude measure, which is a single yes/no question.
@@
 sample, it is not an important channel for natural selection.
-[^note]: Indeed in the one-period model, while $\frac{dN^{*}}{da} > 0$, this
-effect becomes stronger, not weaker, at higher wages.
 
@@
 \[
-p=p^{*}+\eta.
+p = p^* + \eta.
 \]
@@
 \[
-EA = \alpha+\beta p^{*} + \varepsilon = \alpha + \beta(p+\eta) + \varepsilon
+EA = \alpha+\beta p^* + \varepsilon = \alpha + \beta(p - \eta) + \varepsilon
 \]
@@
 = \frac{\sigma_{p}^{2} - \sigma_{\eta}^{2}}{\sigma_{p}^{2}}
-= \frac{\sigma_{p^{*}}^{2}}{\sigma_{p}^{2}}
+= \frac{\sigma_{p^*}^{2}}{\sigma_{p}^{2}}
 \equiv \lambda, \label{eq:r2-ratio}
@@
 The errors-in-variables conditions are quite restrictive. In particular, they
-rule out that the unmeasured part of PSEA ($-\eta$) correlates with the error 
+rule out that the unmeasured part of PSEA ($\eta$) correlates with the error 
 term $\varepsilon$, e.g. with environmental causes of educational attainment. 

### 2022-01-10 why-natural-selection.Rmd
Tweaks for EHB submission

@@
 populations, using genome-wide analysis
-[@Barban_2016;@beauchamp2016genetic;@kong2017selection;@sanjak2018evidence]. In
+[@Barban_2016;@beauchamp2016genetic;@conley2016assortative;@kong2017selection;@sanjak2018evidence;@FIEDER202216]. In
 particular, genetic variants associated with higher educational attainment are
@@
 
-To correct for ascertainment bias, we weight participants using population
-data. We try three alternative weighting schemes: (1) "Geographical" weighting,
-by geography, age and presence/absence of a partner; (2) age and highest educational
-qualification; and (3) for women only, age, highest qualification, and age at 
-first live birth (AFLB). Figure \@ref(fig:plot-res-wt) plots selection effects among the
-entire sample, estimated with the three weighting schemes, and also with uniform
-weights. Effects for `r res_sig_all` out of 33 polygenic scores are significant
-at $p < 0.05/33$ under all four weighting schemes, with `r res_sig_any` scores
-reaching significance under at least one scheme. Weighting makes
-a large difference to estimates: mean effect sizes across all
+To correct for ascertainment bias, we weight participants using population data.
+We try three alternative weighting schemes: (1) "Geographical" weighting, by
+geography, age and presence/absence of a partner; (2) age and highest
+educational qualification; and (3) for women only, age, highest qualification,
+and age at first live birth (AFLB). Figure \@ref(fig:plot-res-wt) plots
+selection effects among the entire sample, estimated with the three weighting
+schemes, and also unweighted. Effects for `r res_sig_all` out of 33 polygenic
+scores are significant at $p < 0.05/33$ under all four weighting schemes, with
+`r res_sig_any` scores reaching significance under at least one scheme.
+Weighting makes a large difference to estimates: mean effect sizes across all
 polygenic scores are increased by a factor of `r mean_rel$Geographical`
@@
 (age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
-Estimates might be further affected by weighting on other demographic variables.[^weighting]
+Estimates might be further affected by weighting on other demographic
+variables.[^weighting]
 
 [^weighting]: To make interpretation straightforward, we do not use weights in 
-results below unless noted. However, all our results are robust to using weights.
+results below unless noted. However, all our qualitative results are robust to
+using weights.
 
@@
 a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical
-predictions, along with corresponding empirical results for the correlation
+predictions, along with our corresponding empirical results for the correlation
 between polygenic scores and fertility.[^one-period-results] The key insight of
@@
 cost as well as a time cost (Appendix Figure \@ref(fig:N-plot-with-m)).
+
+Thus, a simple economic model can explain many of our results.
+Other empirical work in economics also supports our mechanism.
+@caucutt2002women and @monstad2008education show that education and skills
+affect age at first birth and fertility. Income decreases fertility at low
+income levels, but increases it at higher income levels [@cohen2013financial].
+US fertility decreases faster with education among single mothers than married
+mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
+@becker1981treatise.
 
@@
 ```
-
-Thus, a simple economic model can explain many of our results.
-Other empirical work in economics also supports our mechanism.
-@caucutt2002women and @monstad2008education show that education and skills
-affect age at first birth and fertility. Income decreases fertility at low
-income levels, but increases it at higher income levels [@cohen2013financial].
-US fertility decreases faster with education among single mothers than married
-mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
-@becker1981treatise.
 
@@
 median proportion of the total effect explained by the indirect effect is 
-`r percent(median_prop)`.
+`r percent(median_prop)`. The educational attainment variable is a relatively
+crude measure of human capital: more accurate measures would likely
+explain more of the total effect.
 
@@
 These results support our theory that natural selection on polygenic scores is
-driven by their correlation with human capital. We also test for causality by
-running within-siblings regressions (Appendix Figure
+driven by their *correlation* with human capital. We also test whether polygenic
+scores *cause* fertility by running within-siblings regressions (Appendix Figure
 \@ref(fig:plot-res-fe-fertility)). With a reduced sample size, all
@@
 providing tentative evidence that polygenic scores cause fertility, with effects
-being partly driven by gene-environment correlations. We also reran
-within-siblings regressions adding a control for education. Most effect sizes
-barely change, suggesting that our measure of education does not in general
-mediate differences in fertility among siblings. However, for educational
-attainment polygenic scores, effect sizes are reduced by `r percent(educ_reduction)`.
+being partly driven by correlations with environmental variation in human
+capital. We also reran within-siblings regressions adding a control for
+education. Most effect sizes barely change, suggesting that our measure of
+education does not in general mediate differences in fertility among siblings.
+However, for the educational attainment polygenic score (EA3), effect sizes are
+reduced by `r percent(educ_reduction)`.
 
@@
       filter(dep.var == "N children", reg.type == "controlled") %>% 
-  left_join(res_risk_control %>% filter(term == score_name), by = "score_name")
+      left_join(res_risk_control %>% filter(term == score_name), 
+                  by = "score_name")
 
@@
 welfare benefits which incentivize child-bearing might be taken up more among
-low-income individuals. However, our effect sizes appear unchanged over a large
-span of twentieth-century history (tables \@ref(tab:tbl-siblings-by-period) and
-\@ref(tab:tbl-children-by-period)), during which government spending on
-child-related benefits varied considerably [@hoc1999socsec]. In general, there
-is only weak evidence that welfare benefits affect fertility
-[@gauthier2007impact]. A second alternative theory is that polygenic scores
-correlate with the motivation to have children, i.e. parameter $a$ in the model
-[cf. @jones2008fertility]. This theory would not explain why selection effects
-are smaller at higher incomes and education levels. (Indeed, in the model, the
-effect of $a$ on fertility becomes stronger at higher levels of human capital.) A
-third alternative is that traits under selection are linked to externalizing
-behaviour and risk-seeking. This might be partially captured by our
-risk-aversion parameter $\sigma$, which affects fertility by changing the
+low-income individuals. However, most effect sizes appear unchanged over a large
+span of twentieth-century history (Appendix Tables
+\@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)), during
+which government spending on child-related benefits varied considerably
+[@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
+affect fertility [@gauthier2007impact]. A second alternative theory is that
+polygenic scores correlate with the motivation to have children, i.e. parameter
+$a$ in the model [cf. @jones2008fertility]. This theory would not explain why
+selection effects are smaller at higher incomes and education levels. (Indeed,
+in the model, the effect of $a$ on fertility becomes stronger at higher levels
+of human capital.) A third alternative is that traits under selection are linked
+to externalizing behaviour and risk-seeking. This might be partially captured by
+our risk-aversion parameter $\sigma$, which affects fertility by changing the
 marginal utility of income, but a more direct channel is risky sexual behaviour
@@
 variants underlying other phenotypes will show a similar pattern to those
-studied here. In addition, genetic effects on educational attainment have been
-shown to be inflated in population-based samples as compared to within-family
-designs, likely because of indirect genetic effects, gene-environment
-correlations, and/or assortative mating [@lee2018gene; @selzam2019comparing;
-@kong2018nature; @Howe2021.03.05.433935]. In short, it is probably too early to
-tell whether modern natural selection has a substantively important effect on
-population means of phenotypes under selection. Nevertheless, we note that selection
-effects on our measured polygenic scores are still relatively small, even after
-reweighting to account for ascertainment bias.
+studied here. In addition, effects of polygenic scores may be inflated in
+population-based samples, because of indirect genetic effects, gene-environment
+correlations, and/or assortative mating [@lee2018gene;
+@selzam2019comparing; @kong2018nature; @Howe2021.03.05.433935], although we do
+not expect that this should impact their association with number of offspring,
+or the resulting changes in allele frequencies. In short, it is probably too
+early to tell whether modern natural selection has a substantively important
+effect on population means of phenotypes under selection. Nevertheless, we note
+that selection effects on our measured polygenic scores are still relatively
+small, even after reweighting to account for ascertainment bias.
 
@@
 income for `r n_increases` out of 33 polygenic scores, with a median
-proportionate increase of `r median_pct_change`% (Appendix Table
+percentage increase of `r median_pct_change`% (Appendix Table
 \@ref(tab:tbl-inequality-effects)).

### 2022-01-10 titlepage.Rmd
Add title page; number sections but not lines; reorganize sections a bit

---
title: "Human capital mediates natural selection in contemporary humans"
author: "David Hugh-Jones \\thanks{Corresponding author. School of Economics, University of East Anglia, Norwich, UK. Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands.}"
date: "`r format(Sys.Date(), '%d %B %Y')`"
abstract: "\\noindent Natural selection has been documented in contemporary humans, but  little is known about the mechanisms behind it. We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry). Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against). Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner. The direction of natural selection is reversed among older parents, or after controlling for age at first live birth. These patterns are in line with economic theories of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market. Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: true
header-includes:
  - \usepackage{subfig}
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{amsthm}
  - \theoremstyle{plain}
  - \newtheorem{lem}{\protect\lemmaname}
  - \providecommand{\lemmaname}{Lemma}
  - \usepackage{etoc}
editor_options: 
  markdown: 
    wrap: 72
  chunk_output_type: console
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
---

### 2022-01-10 why-natural-selection.Rmd
Add title page; number sections but not lines; reorganize sections a bit

@@
     latex_engine: xelatex
-    number_sections: false
+    number_sections: true
 header-includes:
@@
   - \providecommand{\lemmaname}{Lemma}
-  - \usepackage{lineno}
   - \usepackage{etoc}
-  - \linenumbers
 editor_options: 
@@
 
+# Introduction
+
 Living organisms evolve through natural selection, in which allele frequencies
@@
 
-## Human capital and natural selection
+# Human capital and natural selection
 
@@
 
-## Testing the theory
+# Testing the theory
 

### 2022-01-11 why-natural-selection.Rmd
Tweaks for AJHG: deblind; new cover letter

@@
 title: "Human capital mediates natural selection in contemporary humans"
+author: "David Hugh-Jones \\thanks{Corresponding author.
+School of Economics, University of East Anglia, Norwich, UK.
+Email: D.Hugh-Jones@uea.ac.uk}, Abdel Abdellaoui \\thanks{Department of Psychiatry, Amsterdam UMC, University of Amsterdam, Amsterdam, The Netherlands.
+Email: a.abdellaoui@amsterdamumc.nl}"
 date: "`r format(Sys.Date(), '%d %B %Y')`"
@@
     number_sections: true
+    keep_tex: true
 header-includes:
@@
   - \usepackage{etoc}
+  - \usepackage{lineno}
+  - \linenumbers
 editor_options: 
@@
 
+To compute polygenic scores and PCs, the same procedures were followed as 
+described in @abdellaoui2019genetic.
+
 Earnings in first job are estimated from mean earnings in the 2007 Annual Survey
@@
 Super Output Areas, sex and presence/absence of a partner.
+
+## Acknowledgements
+
+AA is supported by the Foundation Volksbond Rotterdam and by ZonMw grant 849200011 from The Netherlands Organisation for Health Research and Development.
+This study was conducted using UK Biobank resources under application numbers 40310 and 19127.
 

### 2022-01-16 why-natural-selection.Rmd
Small tweaks

--- 2022-01-10 why-natural-selection.Rmd
@@
 ```{r plot-res-fe-fertility, fig.cap = "Selection effects controlling for sibling-group fixed effects, with and without a control for education (left education before 16, 16-18, or after 18).
-Each set of results is from a single regression of \\emph{n children} on 29 polygenic scores.
+Each set of 29 results is from a single regression of \\emph{n children} on 29 polygenic scores.
 Standard errors clustered by sibling group."}

### 2022-01-19 why-natural-selection.Rmd
Add Age/Qualification weights to all regressions in this branch

--- 2022-01-11 why-natural-selection.Rmd
@@
 We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry).
-Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against).
+Consistently over time, polygenic scores that predict lower earnings, education and health also predict lower fertility.
 Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner.
@@
 Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
-bibliography: "negative-selection.bib"
+bibliography: "why-natural-selection.bib"
 output: 
@@
 loadNamespace("scales")
+loadNamespace("shades")
 
@@
         cache      = FALSE, 
-        error = TRUE,
+        error      = FALSE,
         fig.height = 3.5
@@
 my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
-pval_palette <- c("FALSE" = "grey65", "TRUE" = "#FAC32D")
+pval_palette <- c("FALSE" = "grey65", "TRUE" = "#f3b372")
 
@@
   if (is.null(conf_int)) conf_int <- "conf.low" %in% names(dfr)
+  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL
 
   n_regs <- as.double(n_regs)
-
-  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL
 
@@
     ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
-                     fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
+               shape = {{fill_col}}, ...)) +
     my_vline + 
     confint_geom_segment +
-          pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
-          pval_scale() +
-          my_fill_scale(n = n_cats, direction = fill_direction) +
+    geom_point(size = 1.8, aes(color = {{fill_col}})) +
+    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/33)) +
+    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction) +
+    scale_shape_manual(values = 1:n_cats + 14) +
     my_labs() +
     my_theme() +
-          scale_linetype_manual(name = "", values = c("95% c.i." = 1))
-
+    scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
+    scale_alpha_discrete(range = c(1, 0),
+                         guide = guide_legend(
+                           override.aes = list(
+                             shape = c(1, 16),
+                             color = "black",
+                             alpha = 1,
+                             size = 1.8
+                           )
+                         ))
 }
@@
 
-pval_geom_point <- function (size = 1.5, stroke = 0.5, fill = NULL, ...) {
+pval_geom_point <- function (size = 1.6, stroke = 0.8, fill = NULL, ...) {
   if (is.null(fill)) {
@@
 confint_geom_segment <- function () {
-  geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i."), 
+  geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i. uncorrected"), 
                     color = "grey60", alpha = 0.25, size = 1.1, 
@@
 
-  fill_colors <- c("steelblue4", "#D63C3C")
+  fill_colors <- c("steelblue4", "darkred")
 
@@
   } else {
-    scale_color_brewer(aesthetics = aesthetics, ...)
+    shades::lightness(scale_color_brewer(aesthetics = aesthetics, ...), 
+                        shades::scalefac(0.85))
   }
@@
 
-n_in_regs <- function (var, dv = "n_children", data = famhist) {
+n_in_regs <- function (var, dv = "RLRS", data = famhist) {
   fh_subset <- data
-  if (dv == "n_children") fh_subset %<>% filter(kids_ss)
+  if (dv == "RLRS") fh_subset %<>% filter(kids_ss)
   # all PGS have the same NA pattern, so we just use one
@@
 
-add_n <- function (var, famhist_var = var, dv = "n_children", data = famhist) {
+add_n <- function (var, famhist_var = var, dv = "RLRS", data = famhist) {
   var <- as.factor(var)
@@
 \normalem
-
-```{r NOTES-TODO, eval=FALSE}
-"
-
-# TODO response to PNAS
-
-* Only use education, not first job pay, in "controlled" regressions; check
-  description is correct - done.
-* Add CIs - done.
-* Clarify collider bias/endogeneity - done.
-* Formalize theory - done.
-  - Comment on results where effects are actually reversed. This works
-    nicely if you throw in a money cost - done computational example.
-* Clarify causality? - done...
-  - Helps to deal with the "collider bias" argument of reviewer
-  - Might also be a better test of mediation via human capital
-  - One other possibility. If we use only people with first profession
-    teacher, that (as it happens) shuts down the EA3-income correlation.
-    It also shuts down the EA3-fertility relationship, suggesting that
-    income is indeed the mediator.
-* Clarify "what we learn about genetic links to fertility by using economic theory to interpret relationships between polygenic scores and fertility."
-  - Point is that we can explain how natural selection is occurring.
-* "Discuss empirical predictions for AFLB grounded in dynamic theory"
-  - hmm, tricky, maybe we can include it in discussion of economic results at
-  end of formal section.
-  - think our theory with 2 periods probably covers this
-* Why don't women see stronger effects? But note, they do: not significant
-  after correction for multiple testing, but a 25% difference. 
-  - I think we leave this for reviewers.
-* Quadratic regressions for parents - DONE
-
-# Open questions/TODO
-
-* Limit max no. of children in regressions (use an "in_children" bool)
-  - not necessary. Very few have abnormally many.
-* Consider doing separate regressions for "any children" vs. "n children | 1 or more"
-  - heckit too slow and gives crazy answers, just use plm/ols
-  - some variables do look different
-* If we weight with the GHS, and plot the means of figure 1 using this w
-  weighting, most things stay the same - but the change in EA is reversed! EA
-  increases over time. Could this be because the sample includes many
-  relatively educated older people? If so, then mean effects are *overestimated*
-  by figure 1.
-* Reweight other regressions (e.g. the omnibus regression in the appendix)?
-
-* If you take "year of last job started" and see if it is after/before
-  the first child, you can figure out whether mothers keep working.
-  - better still, look at all jobs started and see if any are within e.g. 5 years of
-  first child's birth
-  - this could then be another way to get at income vs substitution effects. You'd
-  predict negative selection effects among mothers who worked. Of course the
-  choice to work is endogenous.
-* double check which scores are reverse coded!
-
-# WHAT MODEL FITS YOUR DATA
-## ASSUMPTIONS
-* Genetics produce human capital h. 
-* h and education are compliments in producing earnings
-* There are income and substitution effects on fertility:
-  - income effects are stronger at high income levels (Becker and Tomes 1976)
-    and among couples
-* People seek partners when in the labour market. Women match a person, observe
-  their quality, and decide whether to have a first child.
-* Then they (may) have more children, either with the partner or not 
-  (high quality partners stay around more)
-* High quality partners increase household income
-
-## RESULTS
-* People with low h are more likely to leave education early and enter the labour market
-  - because e and h are complements
-  - this is part of the substitution effect, as they expect to spend some time
-    raising children rather than being in work
-* At low values of h, increases in h reduce fertility (substitution effect dominates)
-  at higher values, smaller reductions or increases (income effect stronger)
-  - because h predicts income
-* In 1-parent households, increases in h reduce fertility (sub. effect), in 
-  2-parent households, smaller reductions or increases (income effect stronger)
-  - because h predicts income and because stronger income effect among couples
-* People with low h start having children earlier (b/c leave educ earlier)
-  - so part of the substitution effect comes via this
-  - controlling for this the income effect dominates
-  - hence increases in h reduce fertility less once we control for AFLB
-
-## THOUGHTS
-* Should we distinguish between men and women in couples?
-  - within couples, male h should increase fertility (via earnings)
-  - female h should decrease it (ditto?)
-  - see explore-sex-vs-pay.R
-
-## RIVAL MECHANISMS
-* Lynn 2011 mentions two:
-  - social class gradient on ability to delay gratification
-  - modern society protects the weak (Darwin 1871, Spencer 1874)
-    NB this latter doesn't predict more fertility among "dysgenic" ppl, just
-    equal fertility and then genetic drift does its work
-    - examples include modern sanitation, post-1800
-    - and monogamy!
-"
-```
 
@@
 mean genetic scores are small, natural selection may substantially increase the
-correlation between genetic scores and income, i.e. the "genetic lottery"
-[@harden2021genetic].
+correlation between genetic scores and income, increasing genetic differences 
+between different social groups, and thus making the "genetic lottery"
+[@harden2021genetic] more unfair.
 
 # Results
-
-```{r calc-sex-ratio}
-female_ratio <- mean(famhist$sex == 0)
-```
 
@@
 ```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank.
-Points are means for 5-year intervals.
+Circles are means for 5-year intervals.
 Lines are 95% confidence intervals.
-Green triangles show a significant linear increase over time (p < 0.05/33).
-Red squares show a significant decrease."}
+Solid circles show a significant linear increase (green) or decrease (red) over time (p < 0.05/33)."}
 
@@
       ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
-      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
-        fatten = 1.25) +
-      facet_wrap(~score_name, ncol = 5) +
+      geom_linerange(aes(ymin = score_lo, ymax = score_hi), color = "grey60", 
+                       size = 3) +
+      geom_point(aes(y = score), size = 2) +
+      facet_wrap(~score_name, ncol = 6) +
       my_hline +
-      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
-      scale_shape_manual(values = c("-" = "square", "+" = "triangle", "o" = "circle open")) +
+      scale_color_manual(values = c("-" = "darkred", "+" = "darkgreen", "o" = "black")) +
+      scale_shape_manual(values = c("-" = "circle", "+" = "circle", "o" = "circle open")) +
       theme(
+        line        = element_line(size = 0.5, lineend = "round"),
         strip.text  = element_text(size = 6), 
@@
 
-To test more directly for natural selection, we regress respondents' number of
-children ($N_i$) on their polygenic scores (PGS):
+To test more directly for natural selection, we regress respondents' relative 
+lifetime reproductive success (RLRS) on their polygenic scores (PGS):
 
@@
 \begin{equation}
-N_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
+\mathrm{RLRS}_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
 \label{eq:regression}
@@
 
-The "selection effect", $\beta$, reflects the strength of natural selection
-within the sample. In fact, since polygenic scores are normalized, $\beta$ is
-the expected polygenic score among children of the sample
-[@beauchamp2016genetic].[^normalization] Note that equation
-\eqref{eq:regression} does not control for many environmental and genetic
+RLRS is defined as respondents' number of children, divided by the mean number
+of children of people born in the same year. The "selection effect", $\beta$,
+reflects the strength of natural selection within the sample. In fact, since
+polygenic scores are normalized, $\beta$ is the expected polygenic score among
+children of the sample [@beauchamp2016genetic].[^normalization] Note that
+equation \eqref{eq:regression} does not control for many environmental and genetic
 factors that could affect fertility, and as a result, $\beta$ is not an estimate
@@
 
-[^normalization]: The selection effect $\beta$ equals $Cov(N, PGS)/Var(PGS)$
-where $N$ is the number of children. Since PGS are normalized to variance 1 and
-mean 0, this reduces to 
-$Cov(N, PGS) \equiv E(N\ PGS) - E(N) E(PGS) = E(N\ PGS)$. This is the polygenic 
-score weighted by the number of children, which is the average polygenic score 
-in the next generation.
+[^normalization]:The selection effect $\beta$ equals $Cov(RLRS, PGS)/Var(PGS)$.
+Since PGS are normalized to variance 1 and mean 0, this reduces to $Cov(RLRS,
+PGS) = E(RLRS\times PGS) - E(RLRS) E(PGS) = E(RLRS \times PGS)$. This is the
+polygenic score weighted by relative lifetime reproductive success, which is the
+average polygenic score in the next generation [@robertson1966mathematical].
 
@@
                             reg.type == "controlled", 
-                            dep.var == "N children"
+                            dep.var == "RLRS"
                           ) %>% 
@@
 scores are significant at $p < 0.05/33$ under all four weighting schemes, with
-`r res_sig_any` scores reaching significance under at least one scheme.
+`r res_sig_any` scores reaching significance under at least one scheme.[^balance-diverse-def]
 Weighting makes a large difference to estimates: mean effect sizes across all
@@
 
+[^balance-diverse-def]: We also check for stabilizing and disruptive selection by
+estimating \@ref(eq:regression) with a quadratic term. Stabilizing selection
+selects for intermediate values, while disruptive selection selects for extreme 
+values. In particular, we find disruptive selection for educational attainment
+polygenic scores: at higher values of these scores, the negative effect on
+fertility is smaller (Appendix Figure \@ref(fig:plot-purifying)).
+
 [^weighting]: To make interpretation straightforward, we do not use weights in 
@@
 
-We also check for balancing and diversifying selection by estimating
-\@ref(eq:regression) with a quadratic term.[^balance-diverse-def] In particular,
-we find diversifying selection for educational attainment polygenic scores: at
-higher values of these scores, the negative effect on fertility is smaller
-(Appendix Figure \@ref(fig:plot-purifying)).
-
-[^balance-diverse-def]: Balancing selection selects for intermediate values, 
-reducing the variance of the trait under selection. Diversifying selection
-selects for extreme values and increases the variance.
-
-```{r plot-res-wt, fig.cap = "Selection effects: weighted and unweighted regressions.  Each point represents a single bivariate regression of number of children on a polygenic score."}
+```{r plot-res-wt, fig.cap = "Selection effects: weighted and unweighted regressions.
+Each point represents a single bivariate regression of number of children on a polygenic score.
+P value threshold is 0.05, Bonferroni-corrected for multiple comparisons.
+Confidence intervals are uncorrected."}
 
@@
       my_vline +
-      geom_point(size = 1.5, stroke = 0.5) +
+      geom_point(size = 1.6, stroke = 0.8, alpha = 0.9) +
       confint_geom_segment() +
@@
       labs(x = "Effect size", y = "") +
-      scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
+      scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
       my_theme()
@@
 
-To motivate our economic model, we split the sample by demographic
-and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
-We examine socio-demographic variables including income, education and sex, and
-family structure variables including age at first live birth, presence of a partner, and
-lifetime number of sexual partners.
-We emphasize that these categories are not necessarily exogenous to the polygenic
-scores. For example, both in the data and in our theoretical model, age at first
-live birth is a choice variable, which is endogenous to human-capital-related
-polygenic scores (Appendix Figure \@ref(fig:plot-age-flb-dv)). Nevertheless,
-differences in equation \@ref(eq:regression) across subgroups constrain the set
-of possible explanations for natural selection. As we show below, a simple
-economic model predicts the differences we find.
+We now show the empirical puzzles which motivate our economic model. Each
+concerns differences in the strength of natural selection across different
+subgroups in the sample. Splitting the sample by demographic and social
+variables, we re-estimate \@ref(eq:regression) within each subgroup. We use
+socio-economic variables including income and education, and family structure
+variables including age at first live birth, presence of a partner, and lifetime
+number of sexual partners.
+
+Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
+polygenic score, grouping respondents by age of completing full-time education,
+and by household income. Effects are larger and more significant for the lowest
+income category, and for the lowest education category. These results are robust
+to controlling for respondents' age (Appendix section \@ref(sec:age-control)).
 
@@
 
-```{r calc-income-edu-controlled}
-
-drake::loadd(res_income_controlled, res_edu_controlled)
-n_regs_ic <- nrow(res_income_controlled)
-n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
-
-n_regs_educ <- nrow(res_edu_controlled)
-n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
-```
-
-Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
-polygenic score, grouping respondents by age of completing full-time education,
-and by household income. Effects are larger and more significant for the lowest
-income category, and for the lowest education category.  These results could be
-explained by age, if older respondents have lower income and are less educated,
-and also show more natural selection on polygenic scores. However, when we rerun
-the regressions, interacting the polygenic score with income category and also
-with a quadratic in age, the interaction with income remains significant at
-$p < 0.05/`r n_regs_ic`$ for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
-if we interact the PGS with age of leaving full time education and a quadratic
-in age, the interaction with age leaving full time education remains significant
-at $p < 0.05/`r n_regs_educ`$ for `r n_sig_educ` out of `r n_regs_educ` regressions.
-
-Selection effects are also different between men and women (Appendix Figure
-\@ref(fig:plot-sexes)). Differences are particularly large for educational
-attainment, height and MDD. Several polygenic scores for mental illness and
-personality traits are more selected for (or less against) among women,
-including major depressive disorder (MDD), schizophrenia and neuroticism, while
-extraversion is more selected for among men. Scores for waist circumference and
-waist-hip ratio are less selected for among women. One
-possible reason for these sex differences is that polygenic scores may affect
-fertility via success in marriage markets, and men and women may value different
-characteristics in these.
+Turning to family structure, we split respondents by lifetime number of sexual
+partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
+Selection effects are larger and more significant among
+those with more than 3 lifetime partners. Next we split respondents by whether
+they were living with a spouse or partner at the time of interview.
+Effects are larger among those not living with a spouse or partner
+(Figure \@ref(fig:plot-n-partners)b).[^sex]
+
+[^sex]: The same pattern holds if we analyse men and women separately (Appendix Figure
+\@ref(fig:plot-n-partners-sex)). We also directly compared selection effects 
+between men and women (Appendix Figure \@ref(fig:plot-sexes)).
+
+```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and presence of a partner.", fig.subcap = c("Lifetime number of sexual partners", "Presence of a partner"), fig.ncol = 1, fig.align = "center", fig.height = 4}
+
+drake::loadd(res_partners_joint)
+res_partners_joint %<>%       
+      filter(grepl(":", term)) %>% 
+      mutate(`N partners` = ifelse(grepl("TRUE:", term), "3 or less", "4 or more"))
+
+n_regs <- as.double(nrow(res_partners_joint))
+res_partners_joint %>% 
+      standard_ggplot(fill_col = `N partners`, n_regs = n_regs, 
+                        score_col = score_name)
+
+drake::loadd(res_with_partner)
+res_with_partner %<>% filter(grepl(":", term))
+
+n_regs <- as.double(nrow(res_with_partner))
+
+res_with_partner %>% 
+      mutate(
+        Household = ifelse(grepl("TRUE", term), "With partner", 
+                      "Without partner")
+      ) %>% 
+      standard_ggplot(fill_col = Household, n_regs = n_regs, 
+                        score_col = score_name)
+```
 
@@
       pivot_wider(score_name, names_from = category, values_from = estimate) %>%              {cor(.$`age_flb_cat10-22`, .$`age_flb_cat28-52`)}
-
-```
-
-We next focus on variables related to household type and reproductive
-strategy. We split males and females by lifetime number of sexual
-partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
-For both sexes, selection effects are larger and more significant among
-those with more than 3 lifetime partners. Next we split respondents by whether
-they were living with a spouse or partner at the time of interview.
-Effects are larger among those not living with a spouse or partner
-(Figure \@ref(fig:plot-n-partners)b).
-
-Lastly, we split female respondents by age at first live birth
-(AFLB).[^aflb-no-men] There is evidence for genetic effects on AFLB
-[@Barban_2016], and there is a close link between this variable and number of
-children born. Figure \@ref(fig:plot-age-flb) shows effect sizes estimated
-separately for each tercile of AFLB. Several effects are strikingly different
-across terciles. ADHD and MDD are selected for amongst the youngest third of
-mothers, but selected against among the oldest two-thirds. Educational
-attainment is selected for among the oldest two-thirds of mothers, but is not
-significantly selected among the youngest third. Similarly, several polygenic
-scores for body measurements are selected against only among older mothers. The
-correlation between effect sizes for the youngest and oldest terciles is 
-`r cor_age_flb`.
-
-[^aflb-no-men]: This information is unavailable for men.
-
-```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and household type.", fig.subcap = c("Lifetime number of sexual partners", "Household type"), fig.ncol = 1, fig.align = "center", fig.height = 4}
-
-sexes_theme <-  my_theme(
-                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
-                  legend.position      = "bottom", 
-                  legend.box.spacing   = unit(0, "lines"),
-                  legend.box.margin    = margin(-5, 0, 0, -85),
-                  legend.margin        = margin(0, 0, 0, 0),
-                  legend.box.just      = "left",
-                  legend.justification = "left",
-                  legend.spacing       = unit(1, "lines")
-                )
-sexes_guides <- guides(
-                  # fill  = guide_legend(ncol = 1),
-                  # color = guide_legend(ncol = 1)
-                )
-
-drake::loadd(res_partners)
-res_partners %<>%       
-      filter(grepl(":", term)) %>% 
-      mutate(`N partners` = ifelse(grepl("TRUE:", term), "3 or less", "4 or more"))
-
-n_regs <- as.double(nrow(res_partners))
-res_partners %>% 
-      mutate(
-        score_name = pretty_names(score_name)
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(
-        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        x      = estimate, 
-        fill   = `N partners`, 
-        colour = p < 0.05/{{n_regs}}
-      )) + 
-        my_vline +
-        confint_geom_segment() +
-        pval_geom_point() +
-        facet_wrap(vars(sex), scales = "free_y") +
-        my_fill_scale(n = 2) +
-        pval_scale() +
-        sexes_theme +
-        my_labs(x = "", y = "") + 
-        scale_y_reordered() +
-        scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
-        sexes_guides
-
-drake::loadd(res_with_partner_sex)
-res_with_partner_sex %<>% filter(grepl(":", term))
-
-n_regs <- as.double(nrow(res_with_partner_sex))
-
-res_with_partner_sex %>% 
-      mutate(
-        score_name = pretty_names(score_name),
-        Household = ifelse(grepl("TRUE", term), "With partner", 
-                      "Without partner")
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(
-        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        x      = estimate, 
-        fill   = Household, 
-        colour = p < 0.05/{{n_regs}}
-      )) + 
-        my_vline +
-        confint_geom_segment() +
-        pval_geom_point() +
-        facet_wrap(vars(sex), scales = "free_y") +
-        my_fill_scale(n = 2) +
-        pval_scale() +
-        sexes_theme +
-        my_labs(x = "", y = "") + 
-        scale_y_reordered() +
-        scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
-        sexes_guides
 
@@
 
-To investigate this further, we estimate equation \@ref(eq:regression) among
-females, *controlling* for AFLB. In `r opp_signed` out of 
-`r length(score_names)` cases, effects change sign when controls are added. The
-correlation between effect sizes controlling for AFLB, and raw effect sizes, is
-`r cor_raw_flb`. Thus, selection effects seem to come through two opposing
-channels: an effect on AFLB, and an opposite-signed effect on number of children
-after AFLB is controlled for.
+Lastly, we split female respondents by age at first live birth
+(AFLB).[^aflb-no-men] There is evidence for genetic effects on AFLB
+[@Barban_2016], and there is a close link between this variable and number of
+children born. Figure \@ref(fig:plot-age-flb) shows effect sizes estimated
+separately for each tercile of AFLB. Effects are strikingly different
+across terciles. Educational attainment, ADHD and MDD are selected for amongst
+the youngest third of mothers, but selected against among the oldest two-thirds.
+Similarly, several polygenic scores for body measurements are selected against
+only among older mothers. The correlation between effect sizes for the youngest
+and oldest terciles is `r cor_age_flb`. To investigate this further, we estimate
+equation \@ref(eq:regression) among females, *controlling* for AFLB. In 
+`r opp_signed` out of `r length(score_names)` cases, effects change sign when
+controls are added. The correlation between effect sizes controlling for AFLB,
+and raw effect sizes, is `r cor_raw_flb`. Thus, selection effects seem to come
+through two opposing channels: a correlation with AFLB, and an opposite-signed
+correlation with number of children after AFLB is controlled for.
+
+[^aflb-no-men]: AFLB is unavailable for men.
 
@@
 ```{r calc-siblings-children}
-sib_chn_cor <- cor.test(famhist$n_sibs, famhist$n_children, use = "complete")
+sib_chn_cor <- cor.test(famhist$RLRS_parents, famhist$RLRS, use = "complete")
 stopifnot(sib_chn_cor$p.value < 2e-16)
@@
 
+We emphasize that our categories are not necessarily exogenous to polygenic
+scores. For example -- both in the data (Appendix Figure
+\@ref(fig:plot-age-flb-dv)) and in our model -- age at first live birth is a
+choice variable, which is endogenous to an individual's human capital and
+related polygenic scores. Nevertheless, observed differences in equation
+\@ref(eq:regression) across subgroups constrain the set of possible
+theories; a good theory of contemporary natural selection needs to explain
+how these results arise. As we show below, a model based on the economic
+theory of fertility can do just that.
+
 We also examine selection effects among respondents' parents, using information
-on respondents' number of siblings. Effect sizes of polygenic scores are highly
-correlated across the two generations. Median-splitting respondents by year of
-birth, we find little evidence of change in effect sizes among either the
-parents' or the respondents' generation. We also check whether selection effects
-vary by socio-economic status and AFLB in the parents' generation. We have no
+on respondents' number of siblings to calculate parents' RLRS. Effect sizes of
+polygenic scores are highly correlated across the two generations (Appendix
+Figure \@ref(fig:plot-res-sibs)). Median-splitting respondents by year of birth,
+we find little evidence of change in effect sizes among the parents' generation.
+There is some evidence that selection effect sizes are increasing in the 
+respondents' generation, with 9 polygenic scores showing a significant increase. 
+We also check whether selection effects vary by
+socio-economic status and AFLB in the parents' generation. We have no
 information about parents' income, so we use the 1971 Townsend deprivation score
@@
 "quantity-quality tradeoff" between parents' number of children and number of
-grandchildren. We do not find any: in fact, the correlation between *number of
-siblings* and *number of children* is positive ($\rho$ = 
-`r sib_chn_cor$estimate`, $p < 2 \times 10^{-16}$).
+grandchildren. We do not find any: in fact, the correlation between respondents' 
+and parents' RLRS is positive ($\rho$ = `r sib_chn_cor$estimate`, 
+$p < 2 \times 10^{-16}$).
 
@@
 
-Suppose that certain genetic variants correlate with human capital -- skills or
+Suppose that certain genetic variants correlate with *human capital*: skills or
 other characteristics that affect an individual's earnings in the labour market
@@
 in line with the literature: this can be thought of as the expected number of
-children among similar people. The marginal benefit of an extra child is
-$\frac{dU}{dN} = -bWu'(Y)+a$. The effect of an increase in human capital on this
-marginal benefit is
+children among people with a given $a$, $b$ and $W$. The marginal
+benefit of an extra child is $\frac{dU}{dN} = -bWu'(Y)+a$. The effect of an
+increase in human capital on this marginal benefit is
 
@@
 
-Instead of identifying human capital with wages, we now allow
-individuals to choose a level of education $s \in [0,1]$, which has
-a time cost in period 1. Education is
-complementary to human capital $h > 0$, and increases period 2 wages, which take 
-the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
-to 1. Letting $u(\cdot)$ take the constant relative risk aversion form
-$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
-in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
-*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$ in
-a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical
-predictions, along with our corresponding empirical results for the correlation
-between polygenic scores and fertility.[^one-period-results] The key insight of
-the model is that for middling levels of risk aversion, the substitution effect
-dominates at low income levels, but is balanced by the income effect as income
-increases.
+Instead of identifying human capital with wages, we now allow individuals to
+choose a level of education $s \in [0,1]$, which has a time cost in period 1.
+Education is complementary to human capital $h > 0$, and increases period 2
+wages, which take the  simple functional form $w(s,h) = sh$. We normalize period
+1 wages to 1, and let $u(\cdot)$ take the constant relative risk aversion form
+$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$. $\sigma > 0$ measures the
+curvature of the utility function, i.e. the decline in marginal utility of
+income as income increases. We examine total fertility 
+$N^{*} = N_{1}^{*} + N_{2}^{*}$ and the *fertility-human capital relationship*,
+$\frac{dN^{*}}{dh}$. For $\sigma < 1$ in a neighbourhood of $\sigma = 1$, Table
+\@ref(tab:theory) shows five theoretical predictions, along with our
+corresponding empirical results for the correlation between polygenic scores and
+RLRS.[^one-period-results] The key insight of the model is that for middling
+levels of $\sigma$, the substitution effect dominates at low income levels, but
+is balanced by the income effect as income increases.
 
@@
 cost as well as a time cost (Appendix Figure \@ref(fig:N-plot-with-m)).
+
+------------------------------------------------------------------------------------
+    **Theory: the fertility-human capital         **Empirical results**
+    relationship is...**                                      
+--- --------------------------------------------- ----------------------------------
+1.  Negative: $\frac{dN^{*}}{dh} < 0$.            Figures \@ref(fig:plot-means-over-time) 
+                                                  and \@ref(fig:plot-res-wt).
+
+2.  Weaker (closer to zero) at higher wages       Figure 
+    and/or levels of human capital.               \@ref(fig:plot-income-educ-level)a.
+                                                  Selection effects are also weaker at
+                                                  higher polygenic scores for 
+                                                  educational attainment 
+                                                  (Appendix Figure
+                                                  \@ref(fig:plot-purifying)).
+
+3.  More negative when the time burden            Stronger effects for single parents
+    of children $b$ is larger.                    (Figure \@ref(fig:plot-n-partners)).
+
+4.  Weaker at higher levels of education $s$.     Figure \@ref(fig:plot-income-educ-level)b.
+
+5.  Weaker among those who start fertility        Effects weaker among those starting 
+    in period 2 ($N_{1}^{*} = 0$) than among      fertility later 
+    those who start fertility in period 1         (Figure \@ref(fig:plot-age-flb)).
+    ($N_{1}^{*} > 0$).
+
+------------------------------------------------------------------------------------
+
+Table: (\#tab:theory) Predictions from the theoretical model and corresponding 
+empirical results.
 
@@
 mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
-@becker1981treatise.
-
-------------------------------------------------------------------------------------
-    **Theory**                                        **Empirical results**
---- ------------------------------------------------- ------------------------------
-1.  The fertility-human capital                       Figures \@ref(fig:plot-means-over-time) 
-    relationship is negative:                         and \@ref(fig:plot-res-wt).
-    $\frac{dN^{*}}{dh} < 0$.
-
-2.  The fertility-human capital                       Figure \@ref(fig:plot-income-educ-level)a. 
-    relationship is weaker (closer to zero)           Selection effects are also weaker at higher
-    at higher wages and/or levels of                  polygenic scores for 
-    human capital.                                    educational attainment (Appendix Figure
-                                                      \@ref(fig:plot-purifying)).
-
-3.  The relationship is more negative when            Stronger effects for single parents
-    the time burden of children, $b$, is              (Figure \@ref(fig:plot-n-partners)).
-    larger.
-
-4.  The relationship is weaker at higher              Figure \@ref(fig:plot-income-educ-level)b.
-    levels of education $s$.
-
-5.  The relationship is weaker among those            Effects weaker among those starting 
-    who start fertility in period 2                   fertility later 
-    ($N_{1}^{*} = 0$) than among those who            (Figure \@ref(fig:plot-age-flb)).
-    start fertility in period 1 
-    ($N_{1}^{*} > 0$).
-----------------------------------------------------------------------------------------
-
-Table: (\#tab:theory) Predictions from the theoretical model and corresponding 
-empirical results.
-
-```{r literature, eval = FALSE}
-"Other models of the literature
-
-## LITERATURE NOTES
-
-* Lundberg and Pollak JEP 2007:
-    - several models where nonmarital fertility is more likely for poorer women
-    - Burdett and Ermisch 2002, Willis 1999, Neal 2004
-    - Unmarried parents may not be able to credibly commit to each other
-
-* Happel et al. 1984
-   - childbirth is chosen to minimize loss of earnings from interrupted work.
-   - early childbirth if your skills decay a lot  while childrearing. (No point
-     gaining them then losing them.) Late childbirth otherwise.
-   - late childbirth if you have acquired skills before marriage.
-   - Capital-market imperfections make it wise to delay childbirth (till man has
-     high income to cope while wife is at home). More so for men's wages which
-     rise faster.
-   - empirically: (among married couples) 
-     - high-skill occupation women have children later. Higher-earning men have 
-     them earlier.
-
-- Caucutt et al. 2001
-  - high waged and high educated women (of recent cohorts) more likely to
-    delay childbirth
-
-- Cohen et al. 2013
-  - analyse effect of Israeli subsidy
-  - along the way, find that income affects fertility negatively at low income,
-    positively at high income, supporting Becker and Tomes 76
-
-- Becker and Tomes 1976
-  - investment in children quality and quantity when quality must be equal
-    across children; and some quality comes from an innate endowment.
-  - "observed" quantity-of-children income elasticity rises as income rises
-  - while "observed" quality-of-children elasticity falls as income rises, 
-    b/c more of total quality must be paid for by parents themselves. 
-  - the need to pay more for quality results in lower quantity, since total cost is 
-    quantity X quality
-  - something fishy about argument somehow...
-
-- Lindo 2008
-  - men who lose a job have children sooner, but fewer children in total
-  - controls for all skills effects
-  - so overall effect of income is positive, but timing changes
-
-- Monstad et al. 2008
-  - using an education reform to vary education
-  - no evidence of a causal effect on fertility, but births start later
-  * maybe the usual correlation reflects underlying skills. Your data could
-    test this.
-
-- Autor et al. 2017
-  - the China shock (exogenous shock to income) 
-  - they use Chinese import penetration by industry (instrumented using
-    other countries) times industry employment shares; and separate it for
-    males and females
-  - shocks to male employment decreased fertility and marriage; 
-    increased unmarried mothers
-  - shocks to female employment *increase* marriage and motherhood
-  * this roughly seems to be the consensus
-  * Whereas, you find that male and female *skills* (as 'instrumented' by PGS)
-    have the same effect on fertility - i.e. less skills = higher fertility
-
-- Baudin et al. 2015
-  * look at childlessness vs. fertility (extensive/intensive margin)
-  * some people are too poor to afford children
-  * 'a husband alleviates the time cost of having children'. But for some groups (see appendix)
-    there is a +ve income/fertility gradient
-  * fertility decreases with female educn due to subst effect
-  * baby boom and bust in US: a decline in childlessness caused by poverty, then an increase
-    caused by substn effect ('opportunity cost')
-  * fertility decreases with educn both for single mothers and married women
-    - slightly (25%) higher decrease per year of educn (Table 2) for single mothers
-
-```
+@becker1981treatise. A related literature shows negative correlations
+between IQ and fertility [e.g. @lynn2004new; @reeve2018systematic].
 
@@
 effect_size <-  res_all %>% 
-                  filter(dep.var == "N children", reg.type == "controlled") %>% 
+                  filter(dep.var == "RLRS", reg.type == "controlled") %>% 
                   pull(estimate)
@@
       mutate(
-        Controls = fct_relevel(Controls, "None"),
-        PGS      = ifelse(abs(effect_size) > 0.035, PGS, ""),
+        PGS      = ifelse(abs(effect_size) > 0.02, PGS, ""),
         PGS      = sub("(^\\w*?)_.*", "\\1", PGS)
@@
         geom_text(aes(label = PGS), size = 3, colour = "black", 
-                    check_overlap = TRUE, nudge_x = 0.005) +
+                    check_overlap = TRUE, nudge_x = 0.002) +
         facet_wrap( ~ Correlation, scales = "free") + 
@@
 
-```{r}
+```{r calc-mediation}
 drake::loadd(res_mediation)
@@
 res_mediation %<>% mutate(
-                     prop = estimate_ind/estimate_total,
                      # 0.025 not 0.05 because two-sided
-                     sig  = abs(statistic_ind) > qnorm(1 - 0.025/33)
-                   )
-
-n_pos_sig <- sum(res_mediation$sig & res_mediation$prop > 0)
-
-median_prop <- median(res_mediation$prop[res_mediation$prop > 0 & 
-                                           res_mediation$sig])
-
-```
-
-Second, we run a mediation analysis to directly test whether the correlation
-between each polygenic score and fertility is mediated by educational attainment
-(Appendix Table \@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the
-indirect effect of the score on fertility *via* educational attainment is
-significantly different from zero ($p < 0.05/33$). Among these scores, the
-median proportion of the total effect explained by the indirect effect is 
-`r percent(median_prop)`. The educational attainment variable is a relatively
-crude measure of human capital: more accurate measures would likely
-explain more of the total effect.
-
-```{r calc-fe-fertility}
-drake::loadd(res_fe_fertility)
-drake::loadd(res_all)
-
-educ_reduction <- with(res_fe_fertility, 
-       estimate[Regression == "Controlled" & term == "EA3_excl_23andMe_UK"] / 
-       estimate[Regression == "Raw" & term == "EA3_excl_23andMe_UK"]
-     )
-educ_reduction <- 1 - educ_reduction
-
-res_all %<>% filter(dep.var == "N children", reg.type == "controlled")
-res_fe_fertility %<>% filter(Regression == "Raw")
-res_all %<>% left_join(res_fe_fertility, by = "term")
-
-lm_pool_w <- tidy(lm(estimate.y ~ estimate.x, res_all))
-coef_pool_w <- lm_pool_w[[2, "estimate"]]
-se_pool_w <- lm_pool_w[[2, "std.error"]]
-
-```
-
-These results support our theory that natural selection on polygenic scores is
-driven by their *correlation* with human capital. We also test whether polygenic
-scores *cause* fertility by running within-siblings regressions (Appendix Figure
-\@ref(fig:plot-res-fe-fertility)). With a reduced sample size, all
-within-sibling effects are insignificant after Bonferroni correction. However,
-effect sizes are positively correlated with effect sizes from the pooled model,
-and about 50% smaller (regressing within-sibling on pooled effect sizes, $b$ =
-`r format(coef_pool_w, digits = 3)`). This attenuation is in line with the ~50%
-decrease in heritability in within-sibling GWASs on age at first birth and
-educational attainment [@Howe2021.03.05.433935]. We see these results as
-providing tentative evidence that polygenic scores cause fertility, with effects
-being partly driven by correlations with environmental variation in human
-capital. We also reran within-siblings regressions adding a control for
-education. Most effect sizes barely change, suggesting that our measure of
-education does not in general mediate differences in fertility among siblings.
-However, for the educational attainment polygenic score (EA3), effect sizes are
-reduced by `r percent(educ_reduction)`.
+                     sig     = abs(statistic_ind) > qnorm(1 - 0.025/33),
+                     pos_sig = sig & prop_ind > 0
+                   )
+
+n_pos_sig <- sum(res_mediation$pos_sig)
+median_prop <- median(res_mediation$prop_ind[res_mediation$pos_sig])
+
+```
+
+Second, for the `r nrow(res_mediation)` scores where the bivariate correlation in
+\eqref{eq:regression} is significant at $p < 0.05/33$, we run a mediation
+analysis to directly test whether the correlation between each polygenic score
+and fertility is mediated by educational attainment (Appendix Table
+\@ref(tab:tbl-res-mediation)). Figure \@ref(fig:plot-mediation) shows estimated
+proportions explained along with bootstrap 95% confidence intervals
+(uncorrected; 100 bootstraps). For `r n_pos_sig` scores, the indirect effect of
+the score on fertility *via* educational attainment takes the same sign as the
+overall effect, and is significantly different from zero ($p < 0.05/33$). Among
+these scores, the median proportion of the total effect explained by the
+indirect effect is `r percent(median_prop)`. The educational attainment variable
+is a relatively crude measure of human capital: more accurate measures would
+likely explain more of the total effect.
+
+```{r plot-mediation, fig.align = "center", fig.cap = "Proportion of selection effect mediated by educational attainment, polygenic scores with significant selection effects"}
+
+res_mediation %>% 
+      mutate(
+        term = pretty_names(term),
+        term = fct_reorder(term, prop_ind, order_abs(1)),
+        `Indirect effect p < 0.05/33` = sig 
+      ) %>% 
+      ggplot(aes(y = term)) + 
+        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/33`), 
+                   width = .5) + 
+        geom_errorbar(aes(
+                          xmin = prop_ind_conf_low, 
+                          xmax = prop_ind_conf_high,
+                          linetype = "Bootstrap 95% c.i.",
+                        ),
+                        width = 0.3,
+                        color = "grey50"
+                      ) +
+        scale_fill_manual(values = pval_palette) +
+        my_vline +
+        scale_x_continuous(labels = scales::percent, n.breaks = 7) +
+        coord_cartesian(xlim = c(-0.5, 1)) +
+        labs(x = "Proportion mediated", y = "") +
+        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i." = 1))
+
+```
 
@@
 res_compare <- res_all %>% 
-      filter(dep.var == "N children", reg.type == "controlled") %>% 
+      filter(dep.var == "RLRS", reg.type == "controlled") %>% 
       left_join(res_risk_control %>% filter(term == score_name), 
@@
 welfare benefits which incentivize child-bearing might be taken up more among
-low-income individuals. However, most effect sizes appear unchanged over a large
-span of twentieth-century history (Appendix Tables
-\@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)), during
+low-income individuals. However, the majority of effect sizes appear unchanged
+over a large span of twentieth-century history (Appendix Table
+\@ref(tab:tbl-change-by-period)), during
 which government spending on child-related benefits varied considerably
 [@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
-affect fertility [@gauthier2007impact]. A second alternative theory is that
+affect fertility [@gauthier2007impact]. Future
+work could test this theory more explicitly. A second alternative theory is that
 polygenic scores correlate with the motivation to have children, i.e. parameter
 $a$ in the model [cf. @jones2008fertility]. This theory would not explain why
-selection effects are smaller at higher incomes and education levels. (Indeed,
+selection effects are smaller at higher incomes and education levels. Indeed,
 in the model, the effect of $a$ on fertility becomes stronger at higher levels
-of human capital.) A third alternative is that traits under selection are linked
+of human capital. A third alternative is that traits under selection are linked
 to externalizing behaviour and risk-seeking. This might be partially captured by
-our risk-aversion parameter $\sigma$, which affects fertility by changing the
-marginal utility of income, but a more direct channel is risky sexual behaviour
+our parameter $\sigma$, which can be interpreted as a measure of risk aversion
+over income. However, a more direct channel is risky sexual behaviour
 [@mills2021identification]. The data here provide some support for this: scores
@@
 are reversed. Furthermore, the size of selection effects on a polygenic score
-correlates with that score's association with labour market earnings. The
-economic theory of fertility provides a parsimonious explanation for these findings.
-Because of the substitution effect of earnings on fertility, scores are selected
-for when they correlate with low human capital, and this effect is stronger at 
-lower levels of income and education.
+correlates with that score's association with labour market earnings. Strikingly,
+some of these results were predicted by @fisher1930genetical, pp. 253-254.
+
+The economic theory of fertility provides a parsimonious explanation for these
+findings. Because of the substitution effect of earnings on fertility, scores
+are selected for when they correlate with low human capital, and this effect is
+stronger at lower levels of income and education.
 
@@
 that depends on effect sizes. Our results suggest that naïve estimates can be
-affected by sample ascertainment bias. This problem may be less serious in
-surveys which aim to be representative (as the UK Biobank does not). However,
-there is still scope for bias, since not all respondents consent to the
-collection of genetic data. For instance, completion rates for genotype data in
-the US Health and Retirement Study were around 80-85% [@hrs2020]. Researchers
-should be aware of the risks of ascertainment when studying modern natural
-selection.
+affected by sample ascertainment bias. There may be remaining sources of
+ascertainment bias after our weighting; if so, we expect that, like the sources
+of ascertainment we have controlled for, they probably bias our results towards
+zero. This problem may be less serious in surveys which aim to be representative
+(as the UK Biobank does not). However, there is still scope for bias, since not
+all respondents consent to the collection of genetic data. For instance,
+completion rates for genotype data in the US Health and Retirement Study were
+around 80-85% [@hrs2020]. Researchers should be aware of the risks of
+ascertainment when studying modern natural selection.
 
@@
 
-cor_ea3 <- res_ineq[res_ineq$score == "EA3_excl_23andMe_UK", ]
-cor_ea3$increase_pct <- (cor_ea3$ratio - 1) * 100
-
-drake::loadd(res_ineq_ea3)
-
-r2_income_true_psea <- scales::percent(res_ineq_ea3$r2_income_true_psea, 
-                                         accuracy = 0.1)
-r2_income_true_psea_wt <- scales::percent(res_ineq_ea3$r2_income_true_psea_wt, 
-                                            accuracy = 0.1)
 ```
@@
 percentage increase of `r median_pct_change`% (Appendix Table
-\@ref(tab:tbl-inequality-effects)).
-
-If inequalities in polygenic scores are important for understanding social
-structure and mobility [@belsky2018genetic;
-@Rimfeld_2018; @harden2021genetic], then these increases are substantive. A
-back-of-the-envelope calculation, using estimated heritability for educational
-attainment of 0.4 [@branigan2013variation], gives the proportion of variance in
-household income attributable to the "true" polygenic score for educational
-attainment as `r r2_income_true_psea` among respondents. Natural selection would
-raise this proportion to `r r2_income_true_psea_wt` among respondents' children,
-counterfactually holding environmental conditions constant. Similarly, since
-many polygenic scores are predictive of disease risk, they could potentially
-increase health inequalities. In general, the evolutionary history of
-anatomically modern humans is related to disease risk [@benton2021influence];
-understanding the role of contemporary natural selection may aid in mapping the
-genetic architecture of current health disparities.
+\@ref(tab:tbl-inequality-effects)).  If inequalities in polygenic scores are
+important for understanding social structure and mobility 
+[@belsky2018genetic; @Rimfeld_2018; @harden2021genetic], then these increases
+are substantive. Similarly, since many polygenic scores are predictive of
+disease risk, they could potentially increase health inequalities. In general,
+the evolutionary history of anatomically modern humans is related to disease
+risk [@benton2021influence]; understanding the role of contemporary natural
+selection may aid in mapping the genetic architecture of current health
+disparities.
 
@@
 `r comma_num(nrow(famhist))`. For regressions on number of children we use
-participants over 45, since most fertility is completed by this age. This gives
+participants over 50 (males)/45 (females), since most fertility is completed by this age. This gives
 a sample size of `r comma_num(nrow(famhist %>% filter(kids_ss)))`.
@@
 
+Code for this paper is available at 
+https://github.com/hughjonesd/why-natural-selection.
+
 ## Acknowledgements
@@
 
-## Natural selection by sex
+## Selection effects by sex
+
+Figure \@ref(fig:plot-sexes) plots selection effects by sex. Differences are
+particularly large for educational attainment, height and MDD. Several polygenic
+scores for mental illness and personality traits are more selected for (or less
+against) among women, including major depressive disorder (MDD), schizophrenia
+and neuroticism, while extraversion is more selected for among men. Scores for
+waist circumference and waist-hip ratio are less selected for among women. One
+possible reason for these sex differences is that polygenic scores may affect
+fertility via success in marriage markets, and men and women may value different
+characteristics in these.
 
@@
 \clearpage
-## Balancing and diversifying selection
-
-Balancing selection reduces variance in the trait under selection, while
-diversifying selection increases variance. To check for these, we rerun equation
+## Stabillizing and disruptive selection
+
+Stabilizing selection reduces variance in the trait under selection, while
+disruptive selection increases variance. To check for these, we rerun equation
 \@ref(eq:regression), adding a quadratic term in $PGS_i$, and using our
 age/qualification weights. Scores for hip circumference show significant
-balancing selection ($p < 0.05/33$, negative coefficient on quadratic term).
-Scores for educational attainment (EA2 and EA3) show significant diversifying
-selection ($p < 0.05/33$, positive coefficient), which reduces the strength of
-selection against educational attainment at very high levels of the PGS. Figure
-\@ref(fig:plot-purifying) plots predicted number of children against polygenic
-score from these regressions.
-
-We also checked for balancing selection in the parents' generation, using
+stabilizing selection ($p < 0.05/33$, negative coefficient on quadratic term).
+The EA2 score for educational attainment shows significant disruptive selection
+($p < 0.05/33$, positive coefficient), which reduces the strength of selection
+against educational attainment at very high levels of the PGS. (The quadratic on
+the EA3 score has a similar coefficient but is not significant at $p < 0.05/33$.)
+Figure \@ref(fig:plot-purifying) plots predicted number of children
+against polygenic score from these regressions.
+
+We also checked for stabilizing selection in the parents' generation, using
 age/qualification weights multiplied by the inverse of number of siblings.
-Scores for EA2 and EA3 again show significant diversifying selection ($p <
-0.05/33$, positive coefficient). Other scores including hip circumference were 
-not significant.
-
-```{r plot-purifying, fig.align = "center", fig.cap = "Purifying/diversifying selection: predicted number of children by polygenic score.", fig.height = 3}
+Scores for EA2 and EA3 show significant disruptive selection ($p < 0.05/33$, 
+positive coefficient on quadratic). Other scores including hip circumference 
+were not significant.
+
+*Weighted regressions only*: in addition to the above, 
+major depressive disorder shows diversifying selection
+and drinks per week show balancing selection.
+
+```{r plot-purifying, fig.align = "center", fig.cap = "Stabilizing/disruptive selection: predicted number of children by polygenic score.", fig.height = 3}
 
@@
 
-stopifnot(sig == c("EA2_noUKB", "EA3_excl_23andMe_UK", "hip_combined"))
+stopifnot(sig == c("EA2_noUKB", "hip_combined"))
 
@@
 
-op <- par(mfrow = c(1, 3))
+op <- par(mfrow = c(1, 2))
+
 for (score in sig) {
@@
          xlab = pretty_names(score),
-         ylab = "Predicted N children"
+         ylab = "Predicted RLRS"
        )
@@
 par(op)
+
+```
+
+\FloatBarrier
+\clearpage
+## Controlling for age {#sec:age-control}
+
+```{r calc-income-edu-controlled}
+
+drake::loadd(res_income_controlled, res_edu_controlled)
+n_regs_ic <- nrow(res_income_controlled)
+n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
+
+n_regs_educ <- nrow(res_edu_controlled)
+n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
+```
+
+Results in Figure \@ref(fig:plot-income-educ-level) could be
+explained by age, if older respondents have lower income and are less educated,
+and also show more natural selection on polygenic scores. However, when we rerun
+the regressions, interacting the polygenic score with income category and also
+with a quadratic in age, the interaction with income remains significant at
+$p < 0.05/`r n_regs_ic`$ for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
+if we interact the PGS with age of leaving full time education and a quadratic
+in age, the interaction with age leaving full time education remains significant
+at $p < 0.05/`r n_regs_educ`$ for `r n_sig_educ` out of `r n_regs_educ` regressions.
+
+\FloatBarrier
+\clearpage
+## Number of partners and presence of partner by sex
+
+Figure \@ref(fig:plot-n-partners-sex) splits up Figure \@ref(fig:plot-n-partners)
+by sex. The pattern of results is the same in both sexes: selection effects are stronger among those with more lifetime sexual partners, and among those not
+currently living with a partner.
+
+```{r plot-n-partners-sex, fig.cap = "Selection effects by number of sexual partners and presence of a partner, for men and women separately.", fig.subcap = c("Lifetime number of sexual partners", "Presence of a partner"), fig.ncol = 1, fig.align = "center", fig.height = 4}
+
+sexes_theme <-  my_theme(
+                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
+                  legend.position      = "bottom", 
+                  #legend.box.spacing   = unit(0, "lines"),
+                  legend.box.margin    = margin(-5, 0, 0, -85),
+                  legend.margin        = margin(0, 0, 0, 0),
+                  legend.box.just      = "left",
+                  legend.justification = "left",
+                  legend.spacing.y     = unit(0, "lines"),
+                  #legend.spacing       = unit(0.5, "lines"),
+                  legend.title         = element_text(size = 10)
+                  #legend.text          = element_text(size = 8)
+                )
+sexes_guides <- guides(
+                  fill  = guide_legend(ncol = 1),
+                  color = guide_legend(ncol = 1)
+                )
+
+drake::loadd(res_partners)
+res_partners %<>%       
+      filter(grepl(":", term)) %>% 
+      mutate(`N partners` = ifelse(grepl("TRUE:", term), "3 or less", "4 or more"))
+
+n_regs <- as.double(nrow(res_partners))
+res_partners %>% 
+      mutate(
+        score_name = pretty_names(score_name)
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(
+        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        x      = estimate, 
+        fill   = `N partners`, 
+        colour = p < 0.05/{{n_regs}}
+      )) + 
+        my_vline +
+        confint_geom_segment() +
+        pval_geom_point() +
+        facet_wrap(vars(sex), scales = "free_y") +
+        my_fill_scale(n = 2) +
+        pval_scale() +
+        sexes_theme +
+        my_labs(x = "", y = "") + 
+        scale_y_reordered() +
+        scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
+        sexes_guides
+
+drake::loadd(res_with_partner_sex)
+res_with_partner_sex %<>% filter(grepl(":", term))
+
+n_regs <- as.double(nrow(res_with_partner_sex))
+
+res_with_partner_sex %>% 
+      mutate(
+        score_name = pretty_names(score_name),
+        Household = ifelse(grepl("TRUE", term), "With partner", 
+                      "Without partner")
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(
+        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        x      = estimate, 
+        fill   = Household, 
+        colour = p < 0.05/{{n_regs}}
+      )) + 
+        my_vline +
+        confint_geom_segment() +
+        pval_geom_point() +
+        facet_wrap(vars(sex), scales = "free_y") +
+        my_fill_scale(n = 2) +
+        pval_scale() +
+        sexes_theme +
+        my_labs(x = "", y = "") + 
+        scale_y_reordered() +
+        scale_linetype_manual(
+                                name   = "", 
+                                values = c("95% c.i. uncorrected" = 1)
+                              ) +
+        sexes_guides
 
@@
 ## Parents' generation {#sec-parents}
-### Selection effects
+### Selection effects and change over time
 
@@
 
-The UK Biobank data contains information on respondents' number of
-siblings, i.e. their parents' number of children. Since respondents' polygenic 
-scores are equal in expectation
-to the mean scores of their parents, we can use this to look at selection 
-effects in the parents' generation. We estimate equation (\ref{eq:regression})
-using *number of siblings* (including the respondent) as the dependent variable.
-The parents' generation has an additional source of ascertainment bias: sampling
-parents of respondents overweights parents who have many children. For instance,
-parents of three children will have, on average, three times more children
-represented in UK Biobank than parents of one child. Parents of no children will
-by definition not be represented. To compensate, we reweight our preferred
-weightings (Age/Qualification) by the inverse of *number of siblings*.
-
-Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings* on
+The UK Biobank data contains information on respondents' number of siblings
+(including them), i.e. their parents' number of children. Since respondents'
+polygenic scores are equal in expectation to the mean scores of their parents,
+we can use this to look at selection effects in the parents' generation. We
+estimate equation (\ref{eq:regression}) using parents' RLRS as the dependent
+variable.[^parent-RLRS] The parents' generation has an additional source of ascertainment
+bias: sampling parents of respondents overweights parents who have many
+children. For instance, parents of three children will have, on average, three
+times more children represented in UK Biobank than parents of one child. Parents
+of no children will by definition not be represented. To compensate, we reweight
+our preferred weightings (Age/Qualification) by the inverse of *number of
+siblings*.
+
+[^parent-RLRS]: We don't have data on parents' year of birth for most 
+respondents. To create parents' RLRS, we divide respondents' number of siblings 
+by the average number of siblings of all respondents born in the same year,
+weighting the average by respondents' inverse of number of siblings to 
+compensate for ascertainment bias.
+
+Figure \@ref(fig:plot-res-sibs) shows regressions of parents' RLRS on
 polygenic scores. For a clean comparison with the respondents' generation, we
-rerun regressions on *number of children* excluding those with no children, and
+rerun regressions on respondents' RLRS excluding those with no children, and
 show results in the figure. Selection effects are highly correlated across the
@@
 are larger for the parents' generation. We treat this result cautiously, because
-when we split respondents up by year of birth, we find few differences in effect
-sizes between early- and late-born respondents, for either generation. In other
-words, since estimated effect sizes change across "generations", but do not
-change over time within either generation, the change may be due to remaining
-ascertainment bias within the sample or other effects. In particular, effect
-sizes in both generations may depend on polygenic scores' correlation with
+effect sizes in both generations may depend on polygenic scores' correlation with
 childlessness, and we cannot estimate this for the parents' generation.
 
-Excluding childless people in the parents' generation could bias our estimates.
-To learn about this, we compare effect sizes excluding and including childless
-people in the *current* generation. The correlation
-between the two sets of effect sizes is `r cor_inc_vs_excl_childless`. So, patterns across
+To learn more about this, we compare effect sizes excluding and including
+childless people in the *current* generation. The correlation between the two
+sets of effect sizes is `r cor_inc_vs_excl_childless`. So, patterns across
 different scores are broadly similar whether the childless are counted or not.
-However, absolute effect sizes are smaller when the
-childless are excluded, for `r n_smaller` out of 33 scores; the median
-percentage change is `r percent(median_pct_diff)`.
+However, absolute effect sizes are smaller when the childless are excluded, for
+`r n_smaller` out of 33 scores; the median percentage change is 
+`r percent(median_pct_diff)`.
 
@@
 
-stopifnot(sum(summary_period$Change_Parents == "Insignificant") == 33)
-stopifnot(sum(summary_period$Change_Respondents == "Size increasing") == 4)
+stopifnot(sum(summary_period$Change_Parents == "Insignificant") == 32)
+stopifnot(sum(summary_period$Change_Respondents == "Size increasing") == 9)
 stopifnot(sum(summary_period$Change_Respondents == "Change sign") == 1)
-stopifnot(sum(summary_period$Change_Respondents == "Insignificant") == 28)
+stopifnot(sum(summary_period$Change_Respondents == "Insignificant") == 23)
 ```
@@
 birth year, median split at 1950 ("early born" versus "late born"). We use both
-*number of children* and *number of siblings* as a dependent variable. We weight
+respondents' RLRS and parents' RLRS as a dependent variable. We weight
 using age/qualification cells, and further adjust for selection in the parents'
@@
 
-Tables \@ref(tab:tbl-siblings-by-period) and
-\@ref(tab:tbl-children-by-period) summarize the results. There is no
-evidence for changes in selection effects within the parents' generation. In the
-respondents' generation, effect sizes were significantly larger in absolute size 
-among the later-born for four polygenic scores: cognitive ability, EA2, EA3 and
-extraversion. These changes are inconsistent with the intergenerational change, 
+Table \@ref(tab:tbl-change-by-period) 
+summarizes the results. We report the number of scores showing significant
+changes over time (i.e. a significant interaction between polygenic score and
+the "late born" dummy): either a significant change in sign, a significant
+increase in size, or a significant decrease in size. There is little evidence
+for changes in selection effects within the parents' generation, with just one
+score showing a significant decrease in size. In the respondents' generation,
+effect sizes were significantly larger in absolute size among the later-born for
+nine polygenic scores: ADHD, age at menopause, caffeine consumption, cognitive
+ability, Coronary Artery Disease, EA2, EA3, extraversion and Major Depressive
+Disorder. These changes are inconsistent with the intergenerational change,
 where estimated effect sizes were larger among the earlier, parents' generation.
-One score, conscientiousness, showed a significant change in sign, from 
+One score, for alzheimer's disease, showed a significant change in sign, from
 negative to positive effects on fertility.
 
-Overall, there is weak evidence for change over time. The clearest results are
-that (a) the direction of selection, and (b) the pattern of relative effect sizes
-across scores, are broadly consistent over time.
-
-```{r tbl-siblings-by-period}
-
-period_table <- function (tbl, caption) {
-  tbl %>% 
+Overall, there is some evidence for an increase in the strength of selection in
+recent history. The clearest result is that the pattern of relative effect sizes
+across scores is broadly consistent over time.
+
+```{r tbl-change-by-period}
+
+tbl_resp <- summary_period %>% 
+              group_by(Change_Respondents) %>% 
+              count()
+
+tbl_par <- summary_period %>% 
+             group_by(Change_Parents) %>% 
+             count()
+
+tbl_both <- full_join(tbl_par, tbl_resp, 
+                        by = c("Change_Parents" = "Change_Respondents"))
+
+tbl_both %>% 
+            rename(
+              Change              = Change_Parents, 
+              `Parents' RLRS`     = n.x, 
+              `Respondents' RLRS` = n.y
+            ) %>% 
             as_hux() %>%
@@
             set_bottom_border(1, everywhere) %>% 
-
-      add_footnote(glue("Significance is measured at p < 0.05/{n_tests}"), 
+            add_footnote(glue("Significance is measured at p < 0.05/{n_tests}."), 
               border = 0.4, font_size = 8) %>% 
-      set_contents(1, everywhere, 
-        c("Change", "Number of scores")) %>% 
-      set_right_border(final(1), 1, 0) %>% 
             set_top_padding(1) %>% 
@@
             set_left_padding(everywhere, 1, 0) %>% 
-      set_right_padding(everywhere, 2, 0) %>% 
-      set_caption(caption) %>% 
+            set_right_padding(everywhere, 3, 0) %>% 
+            set_caption("Numbers of polygenic scores showing changes in selection effects between early and late born. 
+      Weights for respondents: Age/Qualification. Weights for parents: Age/Qualification times 1/number of siblings.") %>% 
             set_caption_width(1)
-}
-
-summary_period %>% 
-      group_by(Change_Parents) %>% 
-      count() %>% 
-      period_table("Change in selection effects between parents of early and late born respondents (regressions on number of siblings). Weights: Age/Qualification times 1/number of siblings.")
-
-```
-
-```{r tbl-children-by-period}
-summary_period %>% 
-      group_by(Change_Respondents) %>% 
-      count() %>% 
-      period_table("Change in selection effects between early and late born respondents (regressions on number of children). Weights: Age/Qualification.")
+
 ```
@@
 
-Figure \@ref(fig:plot-siblings-townsend) plots effects on *number of siblings* by
+Figure \@ref(fig:plot-siblings-townsend) plots effects on parents' RLRS by
 Townsend deprivation quintile of birth area.
 
-```{r plot-siblings-townsend, fig.cap = "Selection effects in the parents' generation by Townsend deprivation quintile of birth area.
+```{r plot-siblings-townsend, fig.cap = "Selection effects (parents' RLRS) by Townsend deprivation quintile of birth area.
 Higher = more deprived.
@@
 
+For comparison, Figure \@ref(fig:plot-townsend) plots effects on 
+respondents' RLRS by Townsend deprivation quintile of birth area.
+
+```{r plot-townsend, fig.cap = "Selection effects in the respondents' generation by Townsend deprivation quintile of birth area. Higher = more deprived.", fig.align = "center"}
+
+drake::loadd(res_townsend)
+n_regs <- nrow(res_townsend)
+
+res_townsend %>% 
+    mutate(
+        n = prettyNum(n, big.mark = ","),
+        `Townsend quintile` = paste0(quintile, " (N = ", n, ")")
+      ) %>% 
+  standard_ggplot(fill_col    = `Townsend quintile`,
+                    n_regs    = n_regs, 
+                    score_col = score_name, 
+                    order_idx = 5
+                  )
+
+```
+
 \clearpage
@@
 
-res_sibs_unc <- filter(res_all, dep.var == "N siblings", reg.type == "controlled")
+res_sibs_unc <- filter(res_all, dep.var == "RLRS_parents", reg.type == "controlled")
 compare_abp_unc <- res_age_birth_parents %>% 
@@
 the subsets of respondents who reported their mother's or father's age, and who
-had no elder siblings. We run regressions on *number of siblings* on these
+had no elder siblings. We run regressions on parents' RLRS on these
 subsets. Figure \@ref(fig:plot-age-birth-parents-cross) shows selection effects
@@
 
-```{r plot-age-birth-parents-cross, fig.cap = "Selection effects (parents' generation) by age at first live birth terciles.
+```{r plot-age-birth-parents-cross, fig.cap = "Selection effects (parents' RLRS) by age at first live birth terciles.
 Estimates weighted by 1/number of siblings.", fig.subcap = c("Mothers", "Fathers"), fig.ncol = 1}
@@
                  filter(birth_order == 1) %>%
-                 select(n_sibs, whr_combined, moth_age_birth_cat, 
+                 select(RLRS_parents, whr_combined, moth_age_birth_cat, 
                           fath_age_birth_cat)
@@
         `Age at first live birth` = add_n(`Age at first live birth`, 
-                                          "moth_age_birth_cat", dv = "n_sibs",
-                                          data = famhist_bo1)
+                                            "moth_age_birth_cat", 
+                                            dv   = "RLRS_parents",
+                                            data = famhist_bo1
+                                          )
       ) %>% 
@@
         `Age at first live birth` = add_n(`Age at first live birth`, 
-                                          "fath_age_birth_cat", dv = "n_sibs",
-                                          data = famhist_bo1)
+                                            "fath_age_birth_cat", 
+                                            dv   = "RLRS_parents",
+                                            data = famhist_bo1
+                                          )
       ) %>% 
@@
 
-We run a standard mediation analysis in the framework of @baron1986moderator. We estimate 
+We run a standard mediation analysis in the framework of @baron1986moderator.
+For each polygenic score where the bivariate correlation with RLRS is 
+significant at $p$ < 0.05/33, we estimate 
 \begin{align}
-N_i & = \alpha + \beta PGS_i + \gamma EA_i + X_i\mu + \varepsilon_i \\
+RLRS_i & = \alpha + \beta PGS_i + \gamma EA_i + X_i\mu + \varepsilon_i \\
 EA_i & = \delta + \zeta PGS_i + X_i\mu + \eta_i
 \end{align}
-where $N_i$ is fertility, $PGS_i$ is each polygenic score, $EA_i$ is educational
-attainment (age of leaving fulltime education), and $X_i$ is a vector of
-controls. The total effect of $PGS$ on number of children is $\beta + \gamma
-\zeta$. The "indirect effect" mediated by $EA$ is $\gamma \zeta$. The standard
-error of the indirect effect can be calculated as
+where $RLRS_i$ is relative lifetime reproductive success, $PGS_i$ is the
+polygenic score, $EA_i$ is educational attainment (age of leaving fulltime
+education), and $X_i$ is a vector of controls. The total effect of $PGS$ on
+number of children is $\beta + \gamma \zeta$. The "indirect effect" mediated by
+$EA$ is $\gamma \zeta$. The standard error of the indirect effect can be
+calculated as
 
@@
 are not related to human capital). To partially address this, we include
-controls for age, sex, fluid IQ, height and risk attitude in $X$.
+controls for age, sex, fluid IQ, height and risk attitude in $X$. Note that
+controlling for fluid IQ is conservative, in the sense that IQ might also measure
+human capital; nevertheless, we control for it because fluid IQ might affect
+fertility decisions via other channels than human capital.
 
@@
 res_mediation %<>% mutate(
-                     prop = estimate_ind/estimate_total,
                      # 0.025 not 0.05 because two-sided
@@
 
-n_pos_sig <- sum(res_mediation$sig & res_mediation$prop > 0)
+n_pos_sig <- sum(res_mediation$sig & res_mediation$prop_ind > 0)
 
@@
         `Total effect`    = estimate_total,
-        `Indirect effect` = estimate_ind,
-        `95% c.i.` = glue::glue(
-                     "[{estimate_ind - 1.96 * se_ind}, {estimate_ind + 1.96 * se_ind}]"
-                     ),
-        `Indirect/total` = prop,
-        ` ` = star
+        `Indirect effect` = paste(estimate_ind, star),
+        `Proportion (%)`      = prop_ind * 100,
+        `Proportion 95% c.i. (%)` = glue::glue(
+                     "[{prop_ind_conf_low * 100}, {prop_ind_conf_high * 100}]"
+                     )
       ) %>%
@@
       set_align(everywhere, -1, "right") %>% 
-      set_number_format(-1, 2:4, 4) %>% 
-      set_number_format(-1, 5, fmt_percent()) %>%
+      set_number_format(-1, 2:3, 4) %>% 
+      set_number_format(-1, 4:5, 1) %>% 
       set_caption("Mediation analysis") %>% 
       theme_compact() %>% 
-      set_width(0.9) %>%
-      set_col_width(c(.3, .125, .125, .3, .1, .05)) %>% 
+      set_width(0.95) %>%
+      set_col_width(c(.25, .15, .15, .15, .25)) %>% 
       set_bottom_border(final(), everywhere) %>%
-      add_footnote("* indirect effect p < 0.05/33", font_size = 8)
-```
-
-Table \@ref(tab:tbl-res-mediation) shows results. For `r n_pos_sig` scores, the
-indirect effect on fertility via human capital is significantly greater than 0 
-at $p = 0.05/33$.
+      set_font_size(9) %>% 
+      add_footnote("* p < 0.05/33.", font_size = 8)
+```
+
+Table \@ref(tab:tbl-res-mediation) shows results. For `r n_pos_sig` out of
+`r nrow(res_mediation)` scores, the indirect effect on fertility via human 
+capital is significantly different from 0 at $p = 0.05/33$ and has the
+same sign as the total effect. We also calculate
+the proportion of the total effect that is mediated via the indirect effect,
+along with uncorrected 95% confidence intervals (100 bootstraps).
 
@@
 
+```{r calc-fe-fertility}
+drake::loadd(res_fe_fertility)
+drake::loadd(res_all)
+
+educ_reduction <- with(res_fe_fertility, 
+       estimate[Regression == "Controlled" & term == "EA3_excl_23andMe_UK"] / 
+       estimate[Regression == "Raw" & term == "EA3_excl_23andMe_UK"]
+     )
+educ_reduction <- 1 - educ_reduction
+
+res_all %<>% filter(dep.var == "RLRS", reg.type == "controlled")
+res_fe_fertility %<>% filter(Regression == "Raw")
+res_all %<>% left_join(res_fe_fertility, by = "term")
+
+lm_pool_w <- tidy(lm(estimate.y ~ estimate.x, res_all))
+coef_pool_w <- lm_pool_w[[2, "estimate"]]
+se_pool_w <- lm_pool_w[[2, "std.error"]]
+
+```
+
+Results in the main text support our theory that natural selection on polygenic scores is
+driven by their *correlation* with human capital. Here, we test whether polygenic
+scores *cause* fertility by running within-siblings regressions (Figure
+\@ref(fig:plot-res-fe-fertility)). With a reduced sample size, all
+within-sibling effects are insignificant after Bonferroni correction. However,
+effect sizes are positively correlated with effect sizes from the pooled model,
+and about 50% smaller (regressing within-sibling on pooled effect sizes, $b$ =
+`r format(coef_pool_w, digits = 3)`). This attenuation is in line with the ~50%
+decrease in heritability in within-sibling GWASs on age at first birth and
+educational attainment [@Howe2021.03.05.433935]. We see these results as
+providing tentative evidence that polygenic scores cause fertility, with effects
+being partly driven by correlations with environmental variation in human
+capital. We also reran within-siblings regressions adding a control for
+education. Most effect sizes barely change, suggesting that our measure of
+education does not in general mediate differences in fertility among siblings.
+
 ```{r plot-res-fe-fertility, fig.cap = "Selection effects controlling for sibling-group fixed effects, with and without a control for education (left education before 16, 16-18, or after 18).
-Each set of results is from a single regression of \\emph{n children} on 29 polygenic scores.
+Each set of 29 results is from a single regression of RLRS on 29 polygenic scores.
 Standard errors clustered by sibling group."}
@@
 standard_ggplot(res_fe_fertility, fill_col = Regression, n_regs = n_regs,
-                  fill_direction = -1)
+                  fill_direction = -1) +
+  guides(alpha = guide_legend(
+                   override.aes = list(
+                             shape = 1,
+                             color = "black",
+                             alpha = 1,
+                             size = 1.8
+                   )
+                 )
+  )
 
@@
 
-## Estimating the effect of natural selection on the contribution of PSEA to inequality
-
-```{r calc-load-ineq-ea3}
-
-res <- drake::readd(res_ineq_ea3)
-
-prop_effect_selection <- res$r2_income_ea3_wt/res$r2_income_ea3 - 1
-prop_effect_selection <- scales::percent(prop_effect_selection, accuracy = 0.1)
-
-prop_wt_fe <- res$r2_income_ea3_fe_wt/res$r2_income_ea3_fe - 1
-prop_wt_fe <- scales::percent(prop_wt_fe, accuracy = 0.1)
-
-res <- lapply(res, format, digits = 3)
-
-```
-
-From the meta-analysis of twin studies in @branigan2013variation, we take the
-heritability of educational attainment to be around 0.4. This is the variance of
-educational attainment explained by the "true" polygenic score for educational
-attainment $p^{*}$, i.e. the best linear unbiased predictor of educational
-attainment from genetic data. We write the measured polygenic
-score of educational attainment, "EA3" from @lee2018gene, as
-\[
-p = p^* + \eta.
-\]
-
-For realized educational attainment $EA$, write:
-\[
-EA = \alpha+\beta p^* + \varepsilon = \alpha + \beta(p - \eta) + \varepsilon
-\]
-We assume the classical errors-in-variables model, i.e. that $\eta$
-is independent of both $p$ and $\varepsilon$. Under these conditions,
-the ratio between the estimated $R^{2}$ in a regression of $p$ on
-$EA$, and the true $R^{2}$ of $p^{*}$ on $EA$, satisfies: 
-\begin{equation}
-\frac{\textrm{plim}\hat{R}^{2}}{R^{2}} 
-= \frac{\sigma_{p}^{2} - \sigma_{\eta}^{2}}{\sigma_{p}^{2}}
-= \frac{\sigma_{p^*}^{2}}{\sigma_{p}^{2}}
-\equiv \lambda, \label{eq:r2-ratio}
-\end{equation}
-as we show below. The right hand side is the "reliability ratio".
-We regress educational attainment on EA3 in our sample, weighting
-by our Age/Qualification weights, to give 
-$\textrm{plim}\hat{R}^2 = `r res$r2_edu_ea3`$. Plugging this
-into the above, with $R^2 = h^2\approx0.4$, gives $\lambda=`r res$lambda`$.
-
-We now consider the $R^2$ of $p$ on income, that is, the proportion
-of variance in income that is explained by the true polygenic score
-of educational attainment. Applying (\ref{eq:r2-ratio}) again gives
-that
-\begin{equation}
-\frac{\textrm{plim}\hat{R}_{\textrm{income}}^{2}}{R_{\textrm{income}}^{2}}=\lambda.\label{eq:r2-ratio-income}
-\end{equation}
-
-We compute $\hat{R}_{\textrm{income}}^{2}$by regressing income group on EA3,
-first weighting by age/qualification, and then by age/qualification weights
-times respondent's number of children. The estimated $R^2$ are 
-`r res$r2_income_ea3` and `r res$r2_income_ea3_wt` respectively. Applying
-(\ref{eq:r2-ratio-income}) to both estimates gives the proportion of variance in
-income explained by true polygenic score for education attainment (a) among
-respondents as `r res$r2_income_true_psea`; (b) accounting for natural selection
-as `r res$r2_income_true_psea_wt`. The second figure gives the counterfactual
-estimate of $R_{\textrm{income}}^{2}$ after one generation of natural selection,
-holding environmental conditions unchanged.
-
-The errors-in-variables conditions are quite restrictive. In particular, they
-rule out that the unmeasured part of PSEA ($\eta$) correlates with the error 
-term $\varepsilon$, e.g. with environmental causes of educational attainment. 
-If unmeasured PSEA correlates positively with $\varepsilon$, then the
-reliability coefficient $\lambda$ will be lower than calculated. However, the
-proportional effect of natural selection on the PSEA-income relationship will
-remain the same, so long as $\lambda$ does not differ between the two sets of
-weighted regressions.
-
-The estimate above uses a pooled regression of income on EA3, which may include
-confounding environmental variation such as genetic nurture. Within-siblings
-regression would give a smaller estimate of the effect of EA3. However, this
-would also affect the reliability of the regression [@griliches1986errors],
-making it hard to back out estimates of the true $R^2$. Nevertheless, we
-can calculate the proportional increase in $R^2$ due to natural selection, again 
-so long as the reliability coefficient is unaffected by
-the weighting scheme. Running within-siblings regressions of EA3 on income group
-gives within-$R^2$ estimates of `r res$r2_income_ea3_fe` (weighted by
-age and qualification) and `r res$r2_income_ea3_fe_wt` (multiplying weights by number of children).
-Thus, this procedure estimates the proportional change in $R^2_{income}$ due to natural selection as `r prop_wt_fe`.
-
-### Proof of \@ref(eq:r2-ratio)
-
-Write 
-\begin{align}
-\hat{R}^{2} & =1-\frac{\sum_{i=1}^{N}(\hat{\alpha}+\hat{\beta}p_{i}-EA_{i})^{2}}{\sum_{i=1}^{N}(EA_{i}-\bar{EA})^{2}}\label{eq:r-squared}
-\end{align}
-
-Using the fact that $\hat{\alpha}=\bar{EA}-\hat{\beta}\bar{p}$, where bars signify
-sample averages, we can simplify the numerator in the above:
-
-\[
-\sum_{i=1}^{N}(\hat{\alpha}+\hat{\beta}p_{i}-EA_{i})^{2}=\sum_{i=1}^{N}(\bar{EA}-\hat{\beta}\bar{p}+\hat{\beta}p_{i}-EA_{i})^{2}=\sum_{i=1}^{N}(\hat{\beta}(p_{i}-\bar{p})-(EA_{i}-\bar{EA}))^{2}
-\]
-
-Now, we can compute the following probability limit:
-
-\begin{align}
- & \text{plim}\frac{\sum_{i=1}^{N}(\hat{\beta}(p_{i}-\bar{p})-(EA_{i}-\bar{EA}))^{2}}{N}\nonumber \\
-= & \text{plim}\frac{\sum_{i=1}^{N}\hat{\beta}^{2}(p_{i}-\bar{p})^{2}+(EA_{i}-\bar{EA})^{2}-2\hat{\beta}(p_{i}-\bar{p})(EA_{i}-\bar{EA})}{N}\nonumber \\
-= & \beta^{2}\lambda^{2}\sigma_{p}^{2}+\sigma_{EA}^{2}-2\beta\lambda\sigma_{p,EA},\label{eq:plim}
-\end{align}
-
-using the standard errors-in-variables formula $\textrm{plim}\hat{\beta}=\lambda\beta$,
-and where $\sigma_{p,EA}$ is the covariance between $p$ and $EA$.
-
-Write
-\begin{align*}
-\sigma_{p,EA} & =\textrm{Cov}(p^{*}+\eta,\alpha+\beta p^{*}+\varepsilon)\\
- & =0+\beta\sigma_{p^{*}}^{2}+\beta\textrm{Cov}(\eta,p^{*})+\textrm{Cov}(\eta,\varepsilon)\\
- & =\beta\sigma_{p^{*}}^{2}
-\end{align*}
-
-using the errors-in-variables assumption that $\eta$ is independent
-of $p^{*}$ and $\varepsilon$. Plug this into (\ref{eq:plim}) to
-give
-\begin{align*}
- & \text{plim}\frac{\sum_{i=1}^{N}(\hat{\beta}(p_{i}-\bar{p})-(EA_{i}-\bar{EA}))^{2}}{N}\\
-= & \beta^{2}\lambda^{2}\sigma_{p}^{2}+\sigma_{EA}^{2}-2\beta^{2}\lambda\sigma_{p^{*}}^{2}\\
-= & \beta^{2}\lambda\sigma_{p^{*}}^{2}+\sigma_{EA}^{2}-2\beta^{2}\lambda\sigma_{p^{*}}^{2}\textrm{, using \ensuremath{\sigma_{p^{*}}^{2}=\lambda\sigma_{p}^{2}},}\\
-= & \sigma_{EA}^{2}-\beta^{2}\lambda\sigma_{p^{*}}^{2}
-\end{align*}
-
-Inserting this into (\ref{eq:r-squared}) gives
-\begin{align*}
-\textrm{plim}\hat{R}^{2} & =1-\frac{\sigma_{EA}^{2}}{\sigma_{EA}^{2}}+\frac{\beta^{2}\lambda\sigma_{p^{*}}^{2}}{\sigma_{EA}^{2}}\\
- & =\lambda\frac{\beta^{2}\sigma_{p^{*}}^{2}}{\sigma_{EA}^{2}}\\
- & =\lambda R^{2}
-\end{align*}
-since for bivariate regression $R^{2}$ is the square of the correlation
-between $EA$ and $p^{*}$, which can be written $r=$$\beta\frac{\sigma_{p^{*}}^{2}}{\sigma_{EA}\sigma_{p^{*}}}$.
-We thank an anonymous commenter on https://stats.stackexchange.com/questions/238878/how-do-errors-in-variables-affect-the-r2
-for much of this proof.
-
 \clearpage
@@
   mutate(
-    consistent = sign(`raw_N siblings`) == sign(`raw_N children`) &
-                 sign(`raw_N siblings`) == sign(`controlled_N siblings`) &
-                 sign(`raw_N siblings`) == sign(`controlled_N children`)
+    consistent = sign(`raw_RLRS_parents`) == sign(`raw_RLRS`) &
+                 sign(`raw_RLRS_parents`) == sign(`controlled_RLRS_parents`) &
+                 sign(`raw_RLRS_parents`) == sign(`controlled_RLRS`)
   )
@@
 prop_consistent_controlled <- mean(res_all_controlled$consistent)
-controlled_smaller_sibs <- abs(res_all_controlled[["controlled_N siblings"]]) < 
-  abs(res_all_controlled[["raw_N siblings"]])
-controlled_smaller_children <- abs(res_all_controlled[["controlled_N children"]]) < 
-  abs(res_all_controlled[["raw_N children"]])
-median_prop_sibs <- median(abs(res_all_controlled[["controlled_N siblings"]]) /
-    abs(res_all_controlled[["raw_N siblings"]]))
-median_prop_children <- median(abs(res_all_controlled[["controlled_N children"]]) /
-    abs(res_all_controlled[["raw_N children"]]))
+controlled_smaller_sibs <- abs(res_all_controlled[["controlled_RLRS_parents"]]) < 
+  abs(res_all_controlled[["raw_RLRS"]])
+controlled_smaller_children <- abs(res_all_controlled[["controlled_RLRS"]]) < 
+  abs(res_all_controlled[["raw_RLRS"]])
+median_prop_sibs <- median(abs(res_all_controlled[["controlled_RLRS_parents"]]) /
+    abs(res_all_controlled[["raw_RLRS_parents"]]))
+median_prop_children <- median(abs(res_all_controlled[["controlled_RLRS"]]) /
+    abs(res_all_controlled[["raw_RLRS"]]))
 
@@
 
-```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores on number of siblings/children. Sibling estimates weighted by 1/number of siblings."}
+```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores on number of siblings/children. Sibling estimates weighted by 1/number of siblings.", fig.height = 6}
 
@@
         dep.var  = fct_inorder(dep.var),
+        dep.var  = fct_recode(dep.var, 
+                                `Respondents' RLRS` = "RLRS",
+                                `Parents' RLRS` = "RLRS_parents" 
+                              ),
         reg.type = fct_recode(reg.type, 
@@
         my_vline + 
-        facet_wrap(vars(dep.var), scales = "free") +
+        facet_grid(rows = vars(dep.var), scales = "free") +
         pval_geom_point(position = position_dodge(0.4), alpha = 0.8) +
@@
 
-To get a further insight into this we regress *n siblings* and *n children*
+To get a further insight into this we regress respondents' and parents' RLRS
 on individual principal components. Figure \@ref(fig:plot-pc-regs)
@@
         estimate = abs(estimate),
-        dep.var  = fct_relevel(dep.var, "N children"),
+        dep.var  = fct_relevel(dep.var, "RLRS"),
         top_pcs  = ifelse(as.numeric(pc_name) <= 5, term, "")
@@
 rgs <- left_join(rgs, 
-        res_all %>% filter(reg.type == "controlled", dep.var == "N children"), 
+        res_all %>% filter(reg.type == "controlled", dep.var == "RLRS"), 
         by = c("p2" = "term"))
@@
         my_hline +
-        labs(x = "Genetic correlation with EA3", y = "Effect on number of children")
+        labs(x = "Genetic correlation with EA3", y = "Effect on RLRS")
 

### 2022-01-19 _drake.R

@@ -95,12 +95,16 @@ plan <- drake_plan(
                     format = "fst_tbl"
                   ),
   famhist_kids = target(
-                   famhist %>% filter(kids_ss),
+                   famhist %>% 
+                             filter(kids_ss) %>%
+                     # added for weight-everything branch
+                             left_join(age_qual_weights, by = "f.eid"),
                    format = "fst_tbl"
                  ),
   
   famhist_pw = target(
-                 inner_join(famhist, parent_weights, by = "f.eid"),
+                 # use parent_aq_weights for weight-everything branch
+                 inner_join(famhist, parent_aq_weights, by = "f.eid"),
                  format = "fst_tbl"
                ),
 
@@ -130,7 +134,11 @@ plan <- drake_plan(
                    format = "fst_tbl"
                  ),
   
-  pgs_over_time = calc_pgs_over_time(famhist, score_names),
+  pgs_over_time = {
+                    # weight-everything: add weights and use them
+                    famhist %<>% left_join(age_qual_weights, by = "f.eid")
+                    calc_pgs_over_time(famhist, score_names)
+                  },
   
   census_age_qual =  make_census_age_qual(
                        file_in(!! DC1108EW_file),
@@ -179,12 +187,15 @@ plan <- drake_plan(
             dep_var     = "RLRS_parents", 
             score_names = score_names,
             famhist     = famhist_pw,
+            # weight-everything branch: these are now parent_aq_weights
             weights     = famhist_pw$weights
           )
     res_chn <- run_regs_basic(
             dep_var     = "RLRS", 
             score_names = score_names, 
-            famhist     = famhist_kids
+            famhist     = famhist_kids,
+            # added for weight-everything branch
+            weights     = famhist_kids$weights
           )
     dplyr::bind_rows(
             "RLRS_parents" = res_sibs, 
@@ -198,7 +209,10 @@ plan <- drake_plan(
               ~run_regs_fml(
                              "RLRS ~ {score_name} + I({score_name}^2)", 
                              score_name = .x,
-                             famhist    = famhist_kids
+                             famhist    = famhist_kids,
+                             # weight-everything: added weights
+                             # for some reason these need to be made explicit
[…]

### 2022-01-19 regressions.R

@@ -39,7 +39,8 @@ calc_pgs_over_time <- function (famhist, score_names) {
   
   run_reg <- function (score_name) {
     fml <- as.formula(glue("{score_name} ~ YOB"))
-    tidy(lm(fml, famhist))  
+    # weight-everything: use weights
+    tidy(lm(fml, famhist, weights = weights))  
   }
   time_regs <- map_dfr(score_names, run_reg, .id = "score_name")
   time_regs %<>% 
@@ -144,7 +145,8 @@ run_regs_period <- function (children, score_name, famhist, weight_data) {
 
 run_regs_subset <- function(score_name, subset, famhist) {
   fml <- as.formula(glue::glue("RLRS ~ {score_name}"))
-  mod <- lm(fml, famhist, subset = eval(subset))
+  # weights added for weight-everything branch
+  mod <- lm(fml, famhist, subset = eval(subset), weights = famhist$weights)
   res <- tidy(mod, conf.int = TRUE)
   res <- filter(res, term == {{score_name}})
   res$subset <- list(subset)
@@ -153,8 +155,8 @@ run_regs_subset <- function(score_name, subset, famhist) {
   res
 }
 
-
-run_regs_fml <- function(fml, ..., subset = NULL, famhist, weights = NULL) {
+# weights added for weight-everything branch
+run_regs_fml <- function(fml, ..., subset = NULL, famhist, weights = quote(weights)) {
   glue_args <- list(...)
   fml <- as.formula(glue::glue_data(fml, .x = glue_args))
   mod <- lm(fml, famhist, subset = eval(subset), weights = eval(weights))
@@ -199,7 +201,8 @@ run_age_anova <- function (score_name, control, famhist) {
   fml <- as.formula(glue::glue(
     "RLRS ~ {score_name}*({control} + 
           age_at_recruitment + I(age_at_recruitment^2))"))
-  res <- lm(fml, data = famhist) %>% 
+  # weight-everything branch: weights added
+  res <- lm(fml, data = famhist, weights = weights) %>% 
            car::Anova() %>% 
            tidy()
 }
@@ -247,8 +250,11 @@ run_reg_fe_fertility <- function (famhist, score_names) {
   fml_raw <- as.formula(fml_raw)
   fml_mediators <- as.formula(fml_mediators)
   
-  mod_raw <- fixest::feols(fml_raw, data = famhist, note = FALSE)
-  mod_mediators <- fixest::feols(fml_mediators, data = famhist, note = FALSE)
+  # weight-everything: weights added
[…]

### 2022-03-09 _drake.R
Cutoff males at 50 not 45

@@ -61,7 +61,7 @@ plan <- drake_plan(
   famhist_no_resid      =  target({
                       fnr <- clean_famhist(famhist_raw, score_names, sib_groups)
 
-                      fnr$kids_ss <- fnr$age_at_recruitment >= 45
+                      fnr$kids_ss <- fnr$age_at_recruitment >= ifelse(fnr$female, 45, 50)
                       fnr
                     },
                     format = "fst_tbl"

### 2022-03-11 genetic-exposure-to-a-trade-shock.Rmd

---
title: "Genetic exposure to a trade shock"
author: "David Hugh-Jones, Abdel Abdellaoui"
date: "`r format(Sys.Date(), '%d %B %Y')`"
bibliography: "negative-selection.bib"
output: 
  bookdown::pdf_document2:
    toc: false
    latex_engine: xelatex
    number_sections: true
    keep_tex: true
mainfont: Baskerville
mathfont: Baskerville
sansfont: "Gill Sans"
header-includes:
  - \usepackage{setspace}\doublespacing
  - \usepackage{placeins}
  - \usepackage[format=plain, labelfont={bf,it}, textfont=it]{caption}
  - \usepackage{titlesec}
  - \titleformat*{\section}{\sffamily\LARGE}
  - \titleformat*{\subsection}{\sffamily\itshape\Large}
---

```{r setup, include=FALSE}

remake_data <- FALSE

library(sf)
library(dplyr)

loadNamespace("drake")
loadNamespace("santoku")
loadNamespace("tidyr")
loadNamespace("haven")
loadNamespace("fixest")
loadNamespace("broom")
loadNamespace("scales")
loadNamespace("huxtable")

source("~/import-ukbb-data/import-ukbb-data.R")

knitr::opts_chunk$set(echo = FALSE)
```

```{r save-data, eval = remake_data}

drake::loadd(famhist)

fh <- famhist |>
        tidyr::drop_na(home_lon.0.0, home_lat.0.0) |>
        st_as_sf(coords = c("home_lon.0.0", "home_lat.0.0"))

rm(famhist)

projection <- '+proj=tmerc +lat_0=49 +lon_0=-2 +k=0.9996012717 +x_0=400000 +y_0=-100000 +ellps=airy +datum=OSGB36 +units=m +no_defs'
st_crs(fh) <- projection

nuts <- st_read("~/negative-selection-data/NUTS_Level_3_(January_2018)_Boundaries.geojson")
nuts <- st_transform(nuts, crs = projection)

col_stan <- haven::read_dta("~/negative-selection-data/colantone_stanig/Replication_DB_Regional.dta")
# make colantone & stanig data spatial using the nuts geojson:
col_stan <- left_join(nuts, col_stan, by = c("nuts318cd" = "nuts3"))

fh <- st_join(fh, col_stan, join = st_covered_by)

fh <- as.data.frame(fh)

fh$adult_yrs_1990 <- 1990 - fh$YOB - 18  # how many years have you been 18+ in 1990?
fh$dist_moved <- with(fh, sqrt((birth_lat.0.0 - home_lat.0.1)^2 + 
                                 (birth_lon.0.0 - home_lon.0.1)^2))

fh <- add_ashe_income(fh, drake::readd(ashe_income))
fh <- left_join(fh, drake::readd(age_qual_weights), by = "f.eid")

fh <- fh |> 
        mutate(
          num_sick = rowSums(across(starts_with("f.22660.0")) == 106, na.rm = TRUE),
          num_unem = rowSums(across(starts_with("f.22660.0")) == 107, na.rm = TRUE),
          num_pt   = rowSums(across(starts_with("f.22660.0")) == 101, na.rm = TRUE),
          num_oolm = num_sick + num_unem + num_pt,
          birth_decade  = santoku::chop_width(YOB, 10),
        )

# == long data on jobs ====

fh$f.22617.0.0 <- as.numeric(fh$f.22617.0.0)

# 22601: job coding
# 22602: year job started (negative numbers are NAs of different kinds)
# 22617: job coding (historical): this is the SOC 2000 mapping for 22601
# 22603: year job ended (-313 is ongoing at assessment)
# 22605: work hrs/week
# 22660: gap coding
# 22663: year gap started
# 22664: year gap ended

# ideally we might want to have jobs and gaps in the same columns and 
# different rows

fhl <- fh |> 
  select(f.eid, starts_with("f.226"), -f.22661.0.0) |> 
  tidyr::pivot_longer(
    starts_with("f.226"), 
    names_to = c(".value", "instance"),
    names_pattern = "(f\\.\\d+\\.0)\\.(\\d+)",
    values_drop_na = TRUE,
    names_transform = list(instance = as.integer),
    values_transform = list(f.22617.0 = as.character)
  )

fhl  <- left_join(fhl, drake::readd(ashe_income), by = c("f.22617.0" = "Code")) |>
  select(-Description)
fhl <- left_join(fhl, 
                 fh |> select(! starts_with("f."), f.eid, f.53.0.0),
                 by = "f.eid")

fhl <- fhl |>
         rename(
                 job_year_start   = f.22602.0,
                 gap_year_start   = f.22663.0,
                 job_hours_worked = f.22605.0
               ) |>
         mutate( 
               attend_year   = as.numeric(substring(f.53.0.0, 1, 4)),
               job_year_end  = recode(f.22603.0, `-313` = attend_year),
               gap_year_end  = recode(f.22664.0, `-313` = attend_year),
               job_post_90   = job_year_start >= 1990,
               gap_post_90   = gap_year_start >= 1990,
               job_length    = job_year_end - job_year_start + 1, # round up
               gap_length    = gap_year_end - gap_year_start + 1,
               job_age_start = job_year_start - YOB,
               gap_age_start = gap_year_start - YOB,
               birth_decade  = santoku::chop_width(YOB, 10),
               year_start_dm = job_year_start - mean(job_year_start, 
                                                       na.rm = TRUE)
             )

# presumably typos
fhl$job_length[fhl$job_length < 0] <- NA
fhl$gap_length[fhl$gap_length < 0] <- NA

fhl <- mutate(fhl,
                job_length_post_90 = case_when(
                    job_post_90 ~ job_length,
                    !job_post_90 & job_year_end >= 1990 ~ job_year_end - 1990,
                    TRUE ~ 0
                  ),
                gap_length_post_90 = case_when(
                  gap_post_90 ~ gap_length,
                  !gap_post_90 & gap_year_start >= 1990 ~ gap_year_end - 1990,
                  TRUE ~ 0
                )
              )

fhl_sums <- fhl |> 
  group_by(f.eid) |>
  summarize(
    job_years = sum(job_length, na.rm = TRUE),
    gap_years = sum(gap_length, na.rm = TRUE),
    job_years_post_90 = sum(job_length_post_90, na.rm = TRUE),
    gap_years_post_90 = sum(gap_length_post_90, na.rm = TRUE)
  )

fh <- left_join(fh, fhl_sums, by = "f.eid")

fhl_jobs <- fhl |> 
              select(-starts_with("gap")) |>
              rename_with(\(x) gsub("^job_", "", x))

fhl_gaps <- fhl |> 
              select(-starts_with("job")) |>
              rename_with(\(x) gsub("^gap_", "", x))

rm(fhl)

save(fh, fhl_jobs, fhl_gaps, file = "trade.Rdata")
```

```{r load-data, eval = ! remake_data}
load("trade.Rdata")
```

International trade can be mutually enriching. But trade shocks can also harm workers in affected industries. It's important to understand who is most vulnerable to trade shocks, as this can help policy-makers target aid.

Here, we investigate the shock to the UK economy caused by import competition from China since 1990. We use an unusual source of data: a large-scale health survey with half a million participants. By examining employment history data, we find that people in areas exposed to import competiton from China had jobs in lower-paid occupations, and these jobs did not last as long.

Then, we test whether genetics contributed to vulnerability to the China import shock. We interact our shock measure with individuals' polygenic scores for educational attainment, a DNA-derived measure which predicts people's education.

# Data and methods

```{r}
n_indivs <- nrow(fh[! is.na(fh$f.22601.0.0),])
n_jobs <- nrow(fhl_jobs[! is.na(fhl_jobs$year_start),])

mod_uni <- lm(university ~ EA3_excl_23andMe_UK, data = fh)
uni_prob <- scales::percent(coef(mod_uni)["EA3_excl_23andMe_UK"], 
                              suffix = "")
```

Data comes from UK Biobank, a health-centric survey with half a million participants. Individuals were drawn from the UK population born between 1937 and 1970. UK Biobank captures health measurements, interview data and genetic data from blood samples. A follow-up online survey recorded job history data for `r n_indivs` individuals, capturing `r n_jobs` job spells.

Using individuals' genetic data, we calculate their *polygenic score for educational attainment* (PSEA). A polygenic score sums estimated effects of numerous common variants (SNPs), to create an overall score that predicts an invidual characteristic - here, the number of years of education a person receives. We create our score using per-SNP statistics from @lee2018gene, recalculated excluding UK Biobank participants. The score is normalized to mean 0, variance 1. In our data, a 1 standard deviation increase in PSEA predicts a `r uni_prob` percentage point increase in an individual's chance of going to university.

Our measure of the China shock comes from @colantone2018a. For each industry, they calculate the change in Chinese imports over 1992-2007, divided by total labour employed in the industry before 1992. Then, for each UK NUTS-3 region, they weight these changes by the proportion of workers employed in each industry in the region. Thus, regions with a high proportion of workers in industries with a large change in Chinese imports are more exposed to the shock. In our panel, we interact this measure with a dummy variable which is 1 if a job (or job gap) started after 1990.

Workers in regions exposed to the China shock could be different, even genetically different [@abdellaoui2019genetic]. Also, some workers may migrate in response to the China shock, and stayers may be different from leavers [@abdellaoui2019genetic]. Lastly, UK industries where Chinese imports grow could be different in some way. To avoid these threats to identification, we do three things. First we include individual level fixed effects, controlling for all time-invariant characteristics of our subjects, including all genetic variation. Second, for our individual-level measure of exposure, we use the NUTS-3 region where subjects were born, not where they are living now. Third, following @colantone2018a and the literature, we instrument Chinese import growth to the UK with Chinese import growth to the USA.

UK Biobank data is not a random sample of the population. Participants were sampled from a set of geographical areas near assessment centres. Also, people who responded to the invitation letter were selected, being (e.g.) richer, better educated and healthier than the population. Arguably, this makes the data better for theory testing than for estimation of population effect sizes. To mitigate this weighting respondents by age and qualification to match the UK population, using data from the 2007 General Household Survey.

Our cases are individual job spells and gaps between employment. Our independent variables include log job length in years; log job pay; hours worked; and log gap length in years. UK Biobank data does not include information on job pay, but it does include the SOC2000 job code. We approximate pay by taking mean pay for the SOC2000 job code from the 2007 Annual Survey of Hours and Earnings... XXX. We include fixed effects for individual and for the year of start of each job/gap, as well as a quadratic in age.

# Results

```{r mods-basic}

# individual fixed effects. Nice. The import shock is absorbed by the FE:
mods <- list()
mods$`Log pay` <- fixest::feols(
                                 log(mean_pay) ~ age_start + age_start^2 |
                                 factor(f.eid) + factor(year_start) |
                                 i(post_90, import_shock, TRUE) ~ 
                                 i(post_90, instrument_for_shock, TRUE),
                                 data    = fhl_jobs,
                                 cluster = ~ nuts2,
                                 weights = ~ weights,
                                 notes   = ! knitr::is_latex_output()
                               )
mods$`Log job length`   <- update(mods$`Log pay`, log(length) ~ . )
mods$`Log gap length`   <- update(mods$`Log job length`, data = fhl_gaps)
mods$`Hours worked` <- update(mods$`Log pay`, hours_worked ~ . )

# could use col_stan but this is easier
sd_shock <- sd(fhl_jobs$import_shock, na.rm = TRUE)

effects_pct <- mods |>
           lapply(broom::tidy) |> 
           lapply(\(x) {
             shock_term <- grepl("post_90.*shock", x$term)
             exp(x$estimate[shock_term] * sd_shock) - 1
           }) |>
           lapply(scales::percent, accuracy = 0.1)

# for ease of use in visual editor
names(effects_pct) <- gsub(" ", "_", names(effects_pct))

coef_names <- c(
                 "Import shock x post 1990" = 
                   "fit_post_90::TRUE:import_shock"
               )
huxreg_note <- paste0("Standard errors clustered by NUTS-2 area. {stars}.",
                      "TODO: put in IV statistics.")
ht_mods <- huxtable::huxreg(mods,
                       coefs = coef_names,
                       note = huxreg_note
                     )

extra_cols <- huxtable::as_hux(matrix(
               c(
                 "Indiv. dummies", rep("Yes", length(mods)),
                 "Year dummies",  rep("Yes", length(mods)),
                 "Age quadratic", rep("Yes", length(mods))
               ), nrow = 3, byrow = TRUE
             ))

extra_cols <- extra_cols |>
                huxtable::set_bottom_border(3, -1) |>
                huxtable::set_align(1:3, -1, "right")

ht_mods <- huxtable::add_rows(ht_mods, extra_cols, after = 3)

ht_mods <- ht_mods |>
             huxtable::set_caption("Effects of import shock on jobs")

ht_mods
```

Table \@ref(tab:mods-basic) shows our central regressions. A one standard deviation increase in the value of the import shock led to a `r effects_pct$Log_pay` drop in job pay after 1990; job length dropped by `r effects_pct$Log_job_length` , while hours worked rose by `r effects_pct$Hours_worked` . There was no significant change to the length of employment gaps.

```{r mods-EA3-interaction}

mods_x_ea3 <- lapply(mods, update,
                      . ~ . + i(post_90, EA3_excl_23andMe_UK, TRUE)| 
                      . | 
                      . + .:EA3_excl_23andMe_UK ~ 
                      . + .:EA3_excl_23andMe_UK,
                      notes = ! knitr::is_latex_output()
                    )
coef_names_ix <- c(coef_names, 
  "Import shock x post 1990 x PSEA" =
    "fit_post_90::TRUE:import_shock:EA3_excl_23andMe_UK",
  "Post 1990 x PSEA" = 
    "post_90::TRUE:EA3_excl_23andMe_UK"
)

ht_mods_x_ea3 <- huxtable::huxreg(mods_x_ea3,
                   coefs = coef_names_ix,
                   note  = huxreg_note
                 )

ht_mods_x_ea3 <- huxtable::add_rows(ht_mods_x_ea3, extra_cols, after = 7)

ht_mods_x_ea3 <- ht_mods_x_ea3 |> 
                   huxtable::set_caption("Import shock effects by PSEA")

ht_mods_x_ea3

```

Table \@ref(tab:mods-EA3-interaction) interacts the post-1990 import shock with individual polygenic score for educational attainment (PSEA). The interactions are never significant, but standard errors are large, so we can neither reject the null nor confirm that estimates are close to zero.

\FloatBarrier

# Discussion

Our results contribute to two literatures. First, we capture individual-level vulnerability to trade shocks in a detailed survey.

Second, we use a macroeconomic shock to provide clean evidence of a gene-environment interaction. Gene-environment interactions are of interest to geneticists, but they are not easy to identify. Because people may select into different environments based on their genetics, supposed gene-environment interactions may actually result from gene-environment correlation. Here, our data is rich enough that we can include fixed effects, controlling completely for time-invariant individual characteristics including all forms of genetic variation. At the same time, our environmental variation derives from the economic rise of China, a historic macroeconomic change which is almost certainly independent of UK Biobank participants' genetics. We find little evidence of genetic vulnerability to trade shocks, i.e. of a gene-environment interaction.

# References

### 2022-03-14 response-to-reviewers-BE.Rmd
Response to BG reviewers: - add response document - move to RLRS not absolute N children   - some changes to over-time results - add analysis for respondents by Townsend deprivation - add bootstrap for effect of natural selection on income-EA3 R2, in FE model - add graph of mediation analysis - tweak plot style - other tweaks

---
title: "Response to reviewers"
author: "David Hugh-Jones and Abdel Abdellaoui"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output: 
  pdf_document:
    toc: false
    latex_engine: xelatex
header-includes:
  - \usepackage{subfig}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
---

```{r echo=FALSE}

suppressWarnings(suppressPackageStartupMessages(library(dplyr)))
library(ggplot2)
library(forcats)

drake::loadd(res_edu)

theme_set(theme_minimal())

order_abs <- function (n = 1) {
  function (x) x[n] 
}

pretty_names <- function (names) {
  pretty <- c(
    ADHD_2017               = "ADHD",
    age_at_menarche         = "Age at menarche",
    age_at_menopauze        = "Age at menopause",
    agreeableness           = "Agreeableness",
    ai_substance_use        = "Age at smoking initiation",
    alcohol_schumann        = "Alcohol use",
    alzheimer               = "Alzheimer",
    autism_2017             = "Autism",
    bipolar                 = "Bipolar",
    bmi_combined            = "BMI",
    body_fat                = "Body Fat",
    caffeine                = "Caffeine",
    cannabis                = "Cannabis (ever vs. never)",
    cognitive_ability       = "Cognitive Ability",
    conscientiousness       = "Conscientiousness",
    coronary_artery_disease = "Coronary Artery Disease",
    cpd_substance_use       = "Cigarettes per day",
    diagram_T2D             = "Type 2 Diabetes",
    dpw_substance_use       = "Drinks per week",
    EA2_noUKB               = "Educ. attainment 2 (no UKBB)",
    EA3_excl_23andMe_UK     = "Educ. attainment 3 (no UK)",
    eating_disorder         = "Eating disorder",
    extraversion            = "Extraversion",
    height_combined         = "Height",
    hip_combined            = "Hip circumference",
    MDD_PGC2_noUKB          = "Major Depressive Disorder",
    neuroticism             = "Neuroticism",
    openness                = "Openness",
    sc_substance_use        = "Smoking cessation",
    SCZ2                    = "Schizophrenia",
    si_substance_use        = "Smoking initiation",
    wc_combined             = "Waist circumference",
    whr_combined            = "Waist-hip ratio"
  )

  pretty[names]
}

add_n <- function (var, ...) {
  var <- as.factor(var)
  n <- c(88274, 99987, 154603)
  levels(var) <- sprintf("%s (N = %s)", levels(var), 
                         prettyNum(n[seq_len(nlevels(var))], big.mark = ","))
  # var_n <- sprintf("%s (N = %s)", var, prettyNum(n[var], big.mark = ","))
  # var_n[is.na(var)] <- NA
  var
}

my_hline <- geom_hline(yintercept = 0, colour = "grey20", linetype = "dotted")
my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
pval_palette <- c("FALSE" = "grey65", "TRUE" = "#FAC32D")

standard_ggplot <- function (
  dfr, 
  fill_col, 
  n_regs, 
  ...,
  score_col = quo(term), 
  order_idx = 1, 
  fill_direction = 1,
  n_cats    = NULL,
  conf_int  = NULL
) {
  if (! missing(score_col)) score_col <- enquo(score_col)

  mfc <- missing(fill_col)
  fill_col <- if (mfc) quo(NULL) else enquo(fill_col)
  if (missing(n_cats)) {
    n_cats <- if (mfc) 1 else length(unique(pull(dfr, {{fill_col}})))
  }

  if (is.null(conf_int)) conf_int <- "conf.low" %in% names(dfr)

  n_regs <- as.double(n_regs)

  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL

  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
    my_vline + 
    confint_geom_segment +
    pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
    pval_scale() +
    my_fill_scale(n = n_cats, direction = fill_direction) +
    my_labs() +
    my_theme() +
    scale_linetype_manual(name = "", values = c("95% c.i." = 1))
}

shapes_ggplot <- function (
  dfr, 
  fill_col, 
  n_regs, 
  ...,
  score_col = quo(term), 
  order_idx = 1, 
  fill_direction = 1,
  n_cats    = NULL,
  conf_int  = NULL
) {
  if (! missing(score_col)) score_col <- enquo(score_col)

  mfc <- missing(fill_col)
  fill_col <- if (mfc) quo(NULL) else enquo(fill_col)
  if (missing(n_cats)) {
    n_cats <- if (mfc) 1 else length(unique(pull(dfr, {{fill_col}})))
  }

  if (is.null(conf_int)) conf_int <- "conf.low" %in% names(dfr)

  n_regs <- as.double(n_regs)

  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL

  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               shape = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
    my_vline + 
    confint_geom_segment +
    geom_point(size = 1.8, alpha = 0.8) +
    scale_color_manual(values = c("FALSE" = "grey45", "TRUE" = "orange"), ...) +
    scale_shape_manual(values = 1:n_cats + 14) +
    my_labs() +
    my_theme() +
    scale_linetype_manual(name = "", values = c("95% c.i." = 1))

}

tweaked_ggplot <- function (
  dfr, 
  fill_col, 
  n_regs, 
  ...,
  score_col = quo(term), 
  order_idx = 1, 
  fill_direction = 1,
  n_cats    = NULL,
  conf_int  = NULL
) {
  if (! missing(score_col)) score_col <- enquo(score_col)

  mfc <- missing(fill_col)
  fill_col <- if (mfc) quo(NULL) else enquo(fill_col)
  if (missing(n_cats)) {
    n_cats <- if (mfc) 1 else length(unique(pull(dfr, {{fill_col}})))
  }

  if (is.null(conf_int)) conf_int <- "conf.low" %in% names(dfr)

  n_regs <- as.double(n_regs)

  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL

  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
    my_vline + 
    confint_geom_segment +
    pval_geom_point(size = 1.6, stroke = 0.8, fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
    scale_color_manual(values = c("FALSE" = "grey65", "TRUE" = "#f3b372")) +
    my_fill_scale(n = n_cats, direction = fill_direction) +
    my_labs() +
    my_theme() +
    scale_linetype_manual(name = "", values = c("95% c.i." = 1))
}

my_theme <- function (...) theme(axis.text.y = element_text(size = 7), ...)

pval_scale <- function (...) scale_color_manual(values = pval_palette, ...)

pval_geom_point <- function (size = 1.5, stroke = 0.5, fill = NULL, ...) {
  if (is.null(fill)) {
    geom_point(shape = 21, size = size, stroke = stroke, ...)
  } else {
    geom_point(shape = 21, size = size, stroke = stroke, fill = fill, ...)
  }
}

confint_geom_segment <- function () {
  geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i."), 
               color = "grey60", alpha = 0.25, size = 1.1, 
               lineend = "round")
}

my_fill_scale <- function (n, aesthetics = "fill", ...) {

  fill_colors <- c("steelblue4", "#D63C3C")

  if (n <= 2) {
    dots <- list(...)
    dots$direction <- NULL
    do.call(scale_colour_manual, c(list(values = fill_colors[1:n],
                                        aesthetics = aesthetics), 
                                   dots))
  } else {
    scale_color_brewer(aesthetics = aesthetics, ...)
  }
}

my_labs <- function (x = "Effect size", y = "", ...) labs(x = x, y = y, ...)

n_regs <- as.double(nrow(res_edu))
res_edu <- res_edu %>% 
  mutate(
    "Age left FTE" = fct_relevel(age_fte_cat, "< 16", "16-18", "> 18"),
    "Age left FTE" = add_n(`Age left FTE`, "age_fte_cat")
  )

```

Dear Professor Keller,

Thank you very much for the chance to revise our manuscript. We would also like to thank both reviewers for their comments. Below we deal with the points they raised.

## Editor

> In particular, Reviewer 2 makes several good points that I believe can improve the clarity and flow of the manuscript and I suggest you consider these carefully. Reviewer 1 was highly favorable but also makes several excellent suggestions -- including pointing out that a similar observation has been made (remarkably) almost 100 years ago by Fisher. This seems important to include.

See below.

## Reviewer 1

> This is an outstanding paper, perhaps the most important that will be published in the field of behavioral genetics this year.

Thank you 😊.

> Most curious is the failure to cite Fisher (1930).... The overlap in theory and observations with the present work is striking, so much so that the omission is equally striking.

We thank the reviewer for giving us this reference. Indeed it is striking. Fisher's theory about the positive correlation he mentions is that it's driven by selection bias: Yale parents who can afford to send 6 children to Yale are "more able" than those who can only send 1, and as a result, among Yale students, children from bigger families are smarter. This is related to the income effect in our theory. We now mention this work in our conclusion. XXX maybe do more.

> The authors do not have to respond to my next point. This paper already has a lot. But I am curious as to whether the authors have an opinion about Fisher's proposed solution to this problem (if it is a problem). He in essence suggested that the population be stratified into classes based on "human capital," as we might say now, and that within each stratum redistribution should be performed so as to reward parents of larger families. The basic idea is that if I have four children, while an academic colleague who is the same age as me has none (which has allowed him to publish twice as many papers), the tax-and-spend system will take away some of his income and give it to me, the total amount of redistribution being such as to remove the fundamental tradeoff. Can this work in theory? Can it be made politically feasible?

Indeed, we don't want to talk about policy in the paper - we think we do not yet know enough. For what it's worth, here are my personal thoughts as corresponding author. First, government schemes to encourage fertility have a poor track record, see e.g. Gauthier (2007). Second, my strong gut feeling is that it is wrong for governments to (bluntly) institute breeding programs for their citizens. I wrote more informally on this topic [here](https://wyclif.substack.com/p/the-natural-selection-paper-part-822?s=w).

> Another general point is that there is some uncertainty over whether the fertility-IQ correlation is negative in entire populations. Some suggest that its negative value in unrepresentative samples is the result of ascertainment bias (e.g., Higgins, Reed, & Reed, 1962; Kolk & Kieron, 2019). In fact, I have reviewed a manuscript where the authors declined using the UK Biobank for a particular analysis (while using it for all others) on the grounds that the fertility-IQ relationship cannot be accurately inferred in that dataset. My position is that the fertility-IQ correlation has been negative, at most times since relevant data have been collected, while being less so at certain atypical times (e.g., the Baby Boom) and in men. In the present work, the authors use sample reweighting to address ascertainment bias in the UK Biobank, but this is not guaranteed to be effective. A brief literature review addressing the sign of the fertility-IQ correlation, in most Western populations in recent decades, therefore seems warranted to provide some reassurance. My suggestion is to begin with Lynn and Van Court (2004), Meisenberg (2010), and Reeve, Heeney, and Woodley (2018).

We agree this is an important topic, but we are more focused on human capital and earnings than IQ specifically. (Of course the polygenic score for educational attainment does also predict IQ.) The negative fertility-income relationship, in developed countries, is a very well-established stylized fact, and we mention some papers that also show income is causal.

Regarding ascertainment bias, the reviewer is right that there are no guarantees we can eliminate it. When we do correct for ascertainment bias, every correction increases our estimated effect sizes. So we expect that any remaining ascertainment also probably biases our estimates towards zero. We now point this out. (XXX say this somewhere?)

> abstract: "Consistently over time, polygenic scores associated with lower (higher) earnings, education and health are selected for (against)." This is hard to understand. How about: "Consistently over time, polygenic scores predicting higher earnings, education, and health also predict lower fertility."

Done.

> p\. 2: "natural selection may substantially increase the correlation between genetic scores and income, i.e. the "genetic lottery" (Harden, 2021)." Although the meaning eventually becomes clear, this sentence may be very cryptic.

We have made it more explicit: "natural selection may substantially increase the correlation between genetic scores and income, increasing genetic differences between different social groups, and thus making the "genetic lottery" (Harden 2021) more unfair."

> p\. 4, footnote 1: That the covariance between the number of children and the polygenic score for a given phenotype is equal to the evolutionary change in the phenotype between generations is, after some adjustment of terminology, Robertson's (1966, 1968) Secondary Theorem of Natural Selection (see also Walsh & Lynch, 2018, pp. 165-167). I think this should be pointed out.

We now cite Robertson (1966).

> p\. 4: It should be "stabilizing" selection, not "balancing" selection. The latter refers to any kind of selection that favors the coexistence of two or more alleles at a site. The former refers to selection that favors an intermediate value of a quantitative trait, and this is clearly what the authors mean. While I believe the term "diversifying" selection is sometimes uses for selection favoring both extremes of a quantitative trait, "disruptive" selection is probably more common. A similar point applies on p. 20.

We have changed vocabulary accordingly.

> p\. 10: I am interested in the functional form chosen for u(y). This has sensible properties for sigma \< 1 near 1, including being increasing in y for y somewhat greater than zero and being concave. Intuitively, this means that utility increases with income (net of costs of childrearing) but less so as income increases. Is there any particular reason for this functional form as opposed to others that also satisfy these properties? Also, is there any empirical evidence that "utility" obeys this relationship?

The functional form is a standard one in economics. It has the specific property that relative risk aversion is constant. That is, if you have CRRA utility and £100, and you are indifferent to a 50-50 bet between winning £30 and losing £20, then if you had £1000 you would be indifferent between a 50-50 to win £300 or lose £200.

The literature on estimating utility functions is vast (see e.g. Chuang and Schechter 2015). One issue is that of context. CRRA may make sense for risky financial decisions. Here, as Reviewer 2 points out, the "curvature" of the utility function is used not to explain attitudes to risk, but to trade off income against children at different income levels. This author wishes there was a single utility function that worked to explain all human behaviour across contexts... in reality, economists tend to use "what works". We use the CRRA form because it is tractable; we expect that using other forms, intermediate levels of risk aversion would still generate our empirical results, but the maths would be uglier. For a useful discussion of microeconomic theories of fertility, including functional forms, see Jones et al. (2008).

> p\. 30: I feel that more thought might be given to the choice of covariates. For example, if fluid IQ is on the causal path between the PGS and EA, then controlling it will bias downward the estimate of the the PGS effect on EA. But since this mediation analysis seems tentative and not crucial to the paper, perhaps some looseness here can be tolerated.

Yes. Conversely, fluid IQ might affect fertility choices via other pathways than human capital in the labour market, and not controlling for it might leave these to be picked up by EA. We prefer to be "conservative" in controlling for it, in the sense that we'd rather bias the estimate indirect effect of EA towards zero than away from zero. We now mention this issue in the appendix.

That said, the mediation analysis is really the central test of our theory. So we have tried to make it seem less tentative. See our response to Reviewer 2.

> p\. 34: For various reasons I think this section (and the bit in the main text that it supports) ought to be eliminated. The point is to show that natural selection increases the correlation between the EA PGS and income from one generation to the next, correcting for the noise in the PGS. I think a qualitative statement about this point is adequate; the more quantitative attempt by the authors has some problems.
>
> (1) "PSEA" is not defined, although I'm guessing this means "polygenic score for educational attainment."
>
> (2) It happens that there is a more recent meta-analysis of relevant twin studies (Silventoinen et al., 2020).
>
> (3) There appears to be substantial passive gene-environment correlation ("genetic nurture," some now call it), as a result of highly educated parents providing both favorable genes and a social status promoting educational attainment (e.g., Kong et al., 2018). I am sure this complicates the analysis in some way. The authors might respond that they deal with this by using within-family estimates, but see below.
>
> (4) The proof of Equation 5 uses the "standard errors-in-variables formula" plim \\hat\\beta = \\lambda\\beta. While this relationship seems plausible, it may not be standard to behavioral geneticists.
>
> (5) When there is genetic nurture, what is the right PGS effect to use in order to address the question here? The effect in the population, or the effect within families? If it is the latter, a problem is that the estimate of a 142.6% change in R\^2 between generations seems far too high and indicative of low statistical precision.
>
> (6) The assumption that eta is independent of p_star seems very strong. It includes, for example, the assumption that the PGS over rare variants is uncorrelated with the PGS over common variants. I think this is unlikely to be true, as a result of assortative mating (Yengo et al., 2018).

We accept point (6) in particular. On the other hand, we do state that this is a "back of the envelope calculation". We've had pushback from people saying "this is no big deal", so we want to show that at least under some assumptions, yes it might be. Note also that as we say:

"If unmeasured PSEA correlates positively with $\varepsilon$, then the reliability coefficient $\lambda$ will be lower than calculated. However, the proportional effect of natural selection on the PSEA-income relationship will remain the same, so long as $\lambda$ does not differ between the two sets of weighted regressions."

The within-family estimate should remove the worries about exogeneity of p\*, but they are estimated with a smaller sample. We agree that the estimate is high. The bootstrapped 95% confidence interval is from around 0% to 300%, which is not really informative, though it does suggest that genetic nurture is more likely to bias our results downwards than upwards. We've removed the discussion.

> p\. 40 and following: I confess to not following the derivations completely. The authors seem adept at what they do, and I trust that they have made no serious errors.

We hope so too.

> (1) This is my first encounter with the Karush-Kun-Tucker conditions or the Inada condition.

KTK are the standard conditions for constrained optimization (see e.g. Sundaram 1996). We think people who want to follow the proofs will know about them. The Inada condition is a bit more specific to economics. We are now more explicit about what it means. It can be derived in this case by differentiating $u$ to give $u'(y) = y^{-\sigma}$ which for $\sigma > 0$ approaches infinity as $y \rightarrow 0$.

> (2) Below Equation 25, I did not verify the second derivative of N_2 with respect to h. This seems to require simple (if tedious) applications of the product and chain rules. The important conclusion, that N_2 is convex in h for sigma close to 1, is something that I will accept by analogy to the case N_1, N_2 > 0.

This is correct.

> (3) I do not understand in what way Equations 13 and 14 determine when the case N_1 = 0, N_2 > 0 holds. Nevertheless, I accept the intuitive conclusion that it holds for intermediate values of h because too small a value of h (which determines s\*) will make N_1\* in Equation 17 greater than zero and because too high a value will make N_2\* in 18 negative. In a similar way, I accept that no children at all is the optimum for very high values of h; this seems reasonable for this model.

That is the right intuition. Here are the details. Equation 13 requires

$$
-bY_{1}^{-\sigma}+a \le 0
$$

since $\lambda_1$ must be non-negative. The LHS is increasing in $Y_1$. $Y_1=1-s$ since $N_1=0$. So the LHS is decreasing in $s$ and the inequality puts a minimum value on $s$. Lastly, optimal choice of education $s^*$ increases in $h$ for $\sigma<1$ (and decreases for $\sigma>1$). Hence for $\sigma<1$, the inequality puts a minimum on $h$, and for $\sigma>1$ it puts a maximum on $h$.

Equation 14 similarly bounds on $h$, as follows. Equation 22 rearranges equation 14. Equation 25 differentiates equation 22 to show how $N_2$ changes in $h$; for $\sigma<1$, it decreases, for $\sigma>1$ it increases. Therefore, the requirement that $N_2>0$ (for this solution) means that when $\sigma<1$, $h$ cannot be too high, and when $\sigma>1$ it cannot be too low.

We have tweaked the discussion slightly to reflect the above. The best way to gain intuition is probably to try a few numerical examples.

## Reviewer 2

> 1\. The authors have improved the presentation of their results, but it can still be further improved. There are still so many results, plus a model, and as a result the paper still reads like a disconnected set of findings and lacks an overall coherent narrative (although a bit less so than before).

We have tried to work on this... XXX.

> In my opinion, the paper's most important finding is that human capital mediates natural selection. I suspect the authors agree with me on this point, since the title of their manuscript says just that. Still, when reading the paper, one does not get to the topic of human capital and natural selection until p. 9 (and the paper has only 15 pages)! The lengthy discussion of the stratified analysis (Figures 3-5) is interesting, and the authors have edited the text (vs. in the earlier version) to explain that these results are related to predictions of their model, but that discussion is too long and distracts from the main point of the paper. Plus, the stratified analyses are\
> limited and only so much can be learned from them. As the authors acknowledge, "these categories are not necessarily exogenous to the polygenic scores. For example, both in the data and in our theoretical model, age at first live birth is a choice variable, which is endogenous to human-capital-related polygenic." Thus, I recommend that that discussion be shortened and the figures consolidated (the authors should find a way to show the must important results in Figures 3-5 on \~1 page.

We think these analyses are important. First, they motivate our theory. As we say, though the mediating variables are endogenous, we build a theory in which (e.g.) education is an endogenous choice variable.

We also think the results matter in their own right. Any theory of contemporary natural selection needs to explain why correlations with fertility are lower at higher levels of income, education, etc. Non-causal explanations would be fine, but theories which don't generate these correlations must be missing something. Note that Fisher, cited by Reviewer 1, used observed correlations quite like ours to differentiate between theories of the fertility-income relationship. Generally, significant correlations have often led to important theoretical literatures, even when the direction of causality is uncertain (e.g. education/wages in labour economics, or democracy/GDP levels in political economy).

We have shortened the discussion as follows... XXX.

> 2\. The authors test their human-capital-mediation theory in two ways in Section 4, but I think their analysis can and should be improved, since this section contains the core empirical results supporting the authors' main claim about human capital mediation. First, the authors show that a PGS's correlation with two measures of human capital negatively correlates with selection effects for the PGS (Figure 6); second, they run a formal mediation analysis.
>
> \> Can the authors introduce a basic econometric framework and assess whether their results in Figure 6 are consistent with those from the mediation analysis (i.e., could the amount of mediation reported in Table 5 for each phenotype explain the results shown in Figure 6)?

The result in Figure 6 is that the size of a score's selection effect correlates with the score's correlation with education and earnings. Intuitively, the fact that part of the selection effect comes from the mediation indeed helps to explains table 6. That is, if

*PGS effect on fertility = PGS correlation with education X education effect on fertility + direct effect of PGS on fertility*

then indeed PGS effect on fertility, and PGS correlation with education, will be correlated. In that sense, the mediation analysis is just a more precise version of Figure 6 (more precise because it also estimates the education effect on fertility). We're not sure if there is an obvious way to work out "how much" of the Figure 6 correlation is explained by mediation. It may be a bit like asking "how much of a bivariate correlation is explained by the same correlation but with controls": these are really two alternate models of the same process, one less controlled than than the other. We are probably hitting the limits of our econometric expertise, so we are open to suggestions.

> \> In the mediation analysis, why do the authors only use education as the mediator? The mediation framework they use can easily accommodate many mediators (see Section 6 of the Supplementary Information of the "EA2" paper by Okbay et al., Nature Genetics 2016). I would like to see a more full-fledged mediation analysis that also includes other proxies for human capital as mediators, including earnings in a respondent's first job as well as household income (though the later is not an ideal variable for that, especially for females).

We would like to have more mediators, but appropriate variables are hard to find. Earnings in first job reduces our N by a factor of 10, which makes results too noisy to be useful. As the reviewer suggests, current household income is very much endogenous not only to human capital, but also to previous fertility, which makes it inappropriate as a mediator.

> \> Also, in the mediation analysis, some of the ratios of indirect to total effects are negative or larger than 100%. This could indicate that the indirect and direct effects have opposite signs, but could also be due to the imprecision of the estimates. The authors should estimate 95% CIs for the ratio of indirect/total effects (these can be obtained by bootstrapping). They could then focus their discussion on the ratios that are significant and not on those that are just noisy estimates.

Bootstrapping ratios is not so simple, since if the confidence interval for the denominator contains 0, the ratio confidence set may contain infinity and may not be a single interval, and the standard approach of taking quantiles can be misleading (see Franz 2007). Alternatives like the Fieller method have the same problem. Another issue is that to estimate Bonferroni-corrected significance at p = 0.05/33 we'd need 0.075- and 99.925-percentiles of the distribution, which would require many thousands of bootstraps (infeasibly slow, given our sample size).

Following the advice of Franz (2007), we compromise: we bootstrap uncorrected 95% CIs of the ratio, but only when the denominator (i.e. the total effect) is significantly different from 0 at $p$ \< 0.05/33. This is substantively sensible anyway: if we can't reject that a PGS has no effect on fertility, it does not make much sense to ask whether the effect is mediated. We continue to report p values for the indirect effect only, following Zhao et al. (2010).

> \> Also, in the older version, the authors had a nice figure illustrating how education mediates selection (their old Appendix Figure 17, which showed results from re-estimating Equation 1 for each PGS while controlling for education vs. the baseline estimates). The mediation analysis is in a way more thorough version of that figure, but that figure (ideally also showing the effects of controlling for earnings and any other human capital variables) was nice and helpful and could go in the main text

We have introduced a new figure in the main text, showing the results of the mediation analysis with the bootstrapped confidence intervals.

> \> More generally, I think the authors can improve their empirical analysis of human capital as a mediator of natural selection (the above comments should help with this).

We welcome any further thoughts.

> \
> 3. To adjust for ascertainment bias, the authors adjust their estimates using three weighting schemes. Why not combine all the variables from the three schemes and add other observables that could impact ascertainment (including also sex, proximity to a recruitment center, number of individuals in the household, and household income)? If the authors' goal is to adjust their estimates for ascertainment in the UKB, they should use all relevant observables to create UKB weights so that their estimates more closely resemble those they'd obtain in a more representative sample. This is particularly important here since we're interested in selection, and estimates can be severely biased in a selected sample like the UKB.\
> \
> Also, instead of presenting a small subset of the results with various weighting schemes in Figure 2, it'd be preferable for the authors to pick what they think is the best weighting scheme and then present all results with that scheme. Then, they can briefly discuss the robustness of their results to various weighting schemes.

XXX Probably get in touch with those guys with their own weighting.\

> 4\. On p. 4, line 13, the authors write that "Beta is the expected polygenic score among children of the sample", with a footnote explaining the reason. However, since the authors use N=number of children rather than relative number of children (=number of children/mean number of children), I don't think that's correct. See Beauchamp PNAS 2016. More generally, why not use relative lifetime reproductive success (rLRS) instead of number of children as the dependent variable in equation 1?

We now use RLRS as the dependent variable for both respondents and parents. Results are not much changed, except there is now slightly more evidence for larger selection effects over time: 9 polygenic scores show a significant increase in size when we median-split respondents by birth year.

> 5\. How were the 33 PGSs selected? For instance, why didn't the authors include a PGS of risk preferences? (I'm not requesting that the authors include that specific PGS, but only that they explain their selection process.)

We address this in the Materials and Methods section: "Polygenic scores were chosen so as to cover a reasonably broad range of traits, and based on the availability of a large and powerful GWAS which did not include UK Biobank."

> 6\. Did the authors use an age cutoff to include males and females in their selection analyses. To ensure measured fertility is completed fertility, the authors should only use males who were at least 50 years old and females who were at least 45 years old at the time of measurement.

We used 45 years old. We have adjusted this to 50 male/45 female as suggested. Results are little changed, except that EA3 is no longer undergoing significant diversifying selection in the respondents' generation.

> 7\. The authors write that sigma captures risk aversion in income in their model. But that's confusing and not a good way of describing this for their model, because as far as I understand there's no income risk! They should clarify this and say that sigma regulates the curvature of the utility function, and explain what this means in their riskless model.

It is a fair point. We have rewritten accordingly: "$\sigma > 0$ measures the curvature of the utility function, i.e. the decline in marginal utility of income as income increases".\
\
OTHER COMMENTS AND SUGGESTIONS

> 8\. Most figures are hard to read. For example, Figures 3-5 are hard to read. In Figure 3a, the dot color indicate education and the dot border color indicates significance; that's very difficult to read quickly (for instance, a blue dot with a yellow border looks green). Instead, the authors could use different symbols (crosses, squares) for education and colors for significance, or vice-versa.

We have experimented a lot with these figures, and they may not be easy to improve. For example, we tried using symbols, but in our judgment the added visual noise makes it harder to see the pattern of the results across rows. Below we show three alternative versions of one of our figures: our original, a version using symbols, and one with subtle tweaks to colors and stroke sizes. We prefer our tweaked version (and have used it in the revision), but we are open to suggestions.

```{r different-plots, echo = FALSE, fig.cap = "Alternative plot styles", fig.subcap = c("original", "shapes", "tweaked"), fig.ncol = 1, fig.height=2.5, fig.align="center"}

res_edu <- res_edu %>% filter(grepl("^[a-o]", term))
n_regs <- nrow(res_edu)

standard_ggplot(res_edu, n_regs = n_regs, fill_col = `Age left FTE`, fill_direction = -1)

shapes_ggplot(res_edu, n_regs = n_regs, fill_col = `Age left FTE`, fill_direction = -1)

tweaked_ggplot(res_edu, n_regs = n_regs, fill_col = `Age left FTE`, fill_direction = -1)
```

> 9\. The authors should provide more info to help readers understand the details of the analyses whose results are shown in Appendix Tables 3-4.

We now say: 'We report the number of scores showing significant changes over time (i.e. a significant interaction between polygenic score and the "late born" dummy): either a significant change in sign, a significant increase in size, or a significant decrease in size.'

> 10\. The authors do not refer to Figure 10 in the main text.

Fixed.

> 11\. As mentioned above, only so much can be learned from the stratified analyses, and I recommend that less space be devoted to them in the main text. Still, the authors should consider showing the results (probably in the Appendix, given space restriction) stratified by geographical variables, which are more plausibly exogenous (though not perfectly so), such as being born in the North vs. South, in coal mining area vs. not, in rural areas vs. in towns. This may not related directly to their model, though.

See above for our thoughts on exogeneity versus theory. We have added in results for respondents' RLRS, stratified by Townsend deprivation of birthplace. These look very similar to the results for parents' RLRS. Note, though, that since polygenic scores are inherited, and correlate with parents' mobility (Abdellaoui et al. 2019), geographic variables are indeed not exogenous.

## References

Abdellaoui, A., Hugh-Jones, D., Yengo, L., Kemper, K.E., Nivard, M.G., Veul, L., Holtz, Y., Zietsch, B.P., Frayling, T.M., Wray, N.R. and Yang, J., 2019. Genetic correlates of social stratification in Great Britain. *Nature human behaviour*, *3*(12), pp.1332-1342.

Acemoglu, D., Johnson, S., Robinson, J.A. and Yared, P., 2008. Income and democracy. *American economic review*, *98*(3), pp.808-42.

Chuang, Y. and Schechter, L., 2015. Stability of experimental and survey measures of risk, time, and social preferences: A review and some new results. *Journal of development economics*, *117*, pp.151-170.

Franz, V.H., 2007. Ratios: A short guide to confidence limits and proper use. *arXiv preprint arXiv:0710.2024*.

Gauthier, A.H., 2007. The impact of family policies on fertility in industrialized countries: a review of the literature. *Population research and policy review*, *26*(3), pp.323-346.

Jones, L.E., Schoonbroodt, A. and Tertilt, M., 2008. *Fertility theories: can they explain the negative fertility-income relationship?* (No. w14266). National Bureau of Economic Research.

Sundaram, R.K., 1996. *A first course in optimization theory*. Cambridge University Press.

Zhao, X., Lynch Jr, J.G. and Chen, Q., 2010. Reconsidering Baron and Kenny: Myths and truths about mediation analysis. *Journal of consumer research*, *37*(2), pp.197-206.

### 2022-03-14 model-appendix.Rmd
Add proof that dN^*/da gets larger at higher levels of human capital
Response to BG reviewers: - add response document - move to RLRS not absolute N children   - some changes to over-time results - add analysis for respondents by Townsend deprivation - add bootstrap for effect of natural selection on income-EA3 R2, in FE model - add graph of mediation analysis - tweak plot style - other tweaks

@@
 \end{align}
-Note that the Inada condition ($\lim_{x\rightarrow0}u'(x)=\infty$)
+Note that the Inada condition (that marginal utility of income grows without 
+bound as income approaches zero, $\lim_{x\rightarrow0}u'(x)=\infty$)
 for period 1 rules out $s^{*}=1$ and $N_{1}=1/b$, so we need not
@@
 
-Conditions (\ref{eq:kkt1}) and (\ref{eq:kkt2}) give the bounds for
-this case. When $\sigma<1$, (\ref{eq:kkt1}) puts a minimum on $h$
-and (\ref{eq:kkt2}) puts a maximum on $h$. The situation is reversed
-for $\sigma>1$.
+This case holds for intermediate values on $h$. Equation (\ref{eq:dNdh2}) 
+shows that for $\sigma < 1$, $N_2$ decreases in $h$; the requirement that
+$N_2>0$ therefore puts a maximum on $h$. When $\sigma>1$, $N_2$ increases in
+$h$ and this puts a minimum on $h$. The requirement $N_1 = 0$ provides the 
+other bound. Equation (\ref{eq:kkt1}) 
+requires $-bY_{1}^{-\sigma}+a \le 0$ since $\lambda_1$ must be non-negative. 
+The LHS is increasing in $Y_1$, and hence decreasing in $s$ as $Y_1 = 1-s$ since 
+$N_1=0$. Lastly, optimal choice of education $s^*$ increases in $h$ for 
+$\sigma<1$, and decreases for $\sigma > 1$. Hence for $\sigma < 1$, (\ref{eq:kkt1})
+puts a minimum on $h$, and for $\sigma > 1$ it puts a maximum on $h$.
 
@@
 \end{proof}
+
+\subsection*{Effect of $a$}
+
+\begin{lem}
+\label{lemma-a}For $\sigma<1$, $d^2N^*/dadh > 0$, i.e. the effect of $a$
+increases at higher levels of human capital.
+\end{lem}
+\begin{proof}
+Differentiating (\ref{eq:dNdh1}) with respect to $a$ gives
+
+\[
+\frac{d^2N^*_t}{dadh} = -\frac{1}{b}\frac{1}{1-2\sigma}a^{-1}\left(\frac{a}{b}\right)^{1/(1-2\sigma)}\frac{1-\sigma}{2\sigma-1}h^{(1-\sigma)/(2\sigma-1)-1}
+\]
+
+for $t=1,2$ when $N^*_1,N^*_2>0$. For $\sigma > 0.5$ this is positive.
+
+When $N^*_1 = 0,N^*_2>0$, differentiating (\ref{eq:dNdh2}) with respect to $a$ gives
+
+\[
+\frac{d^2N_{2}^{*}}{dadh} = \frac{1}{b}\frac{1}{\sigma}a^{-1}\left(\frac{a}{b}\right)^{-1/\sigma}\frac{1-\sigma}{\sigma}(s^{*}h)^{(1-2\sigma)/\sigma}(s^{*}+h\frac{ds^{*}}{dh})
+\]
+
+which is $-a^{-1}/\sigma$ times (\ref{eq:dNdh2}) and hence is positive.
+
+\end{proof}

### 2022-03-14 why-natural-selection.Rmd
Response to BG reviewers: - add response document - move to RLRS not absolute N children   - some changes to over-time results - add analysis for respondents by Townsend deprivation - add bootstrap for effect of natural selection on income-EA3 R2, in FE model - add graph of mediation analysis - tweak plot style - other tweaks

@@
 mean genetic scores are small, natural selection may substantially increase the
-correlation between genetic scores and income, i.e. the "genetic lottery"
-[@harden2021genetic].
+correlation between genetic scores and income, increasing genetic differences 
+between different social groups, and thus making the "genetic lottery"
+[@harden2021genetic] more unfair.
 
@@
 
-To test more directly for natural selection, we regress respondents' number of
-children ($N_i$) on their polygenic scores (PGS):
+To test more directly for natural selection, we regress respondents' relative 
+lifetime reproductive success ($RLRS_i$) on their polygenic scores (PGS):
 
@@
 \begin{equation}
-N_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
+RLRS_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
 \label{eq:regression}
@@
 
-[^normalization]: The selection effect $\beta$ equals $Cov(N, PGS)/Var(PGS)$
+[^normalization]: XXX The selection effect $\beta$ equals $Cov(N, PGS)/Var(PGS)$
 where $N$ is the number of children. Since PGS are normalized to variance 1 and
@@
 score weighted by the number of children, which is the average polygenic score 
-in the next generation.
+in the next generation [@robertson1966mathematical].
 
@@
 
-We also check for balancing and diversifying selection by estimating
+We also check for stabilizing and disruptive selection by estimating
 \@ref(eq:regression) with a quadratic term.[^balance-diverse-def] In particular,
-we find diversifying selection for educational attainment polygenic scores: at
+we find disruptive selection for educational attainment polygenic scores: at
 higher values of these scores, the negative effect on fertility is smaller
@@
 
-[^balance-diverse-def]: Balancing selection selects for intermediate values, 
-reducing the variance of the trait under selection. Diversifying selection
+[^balance-diverse-def]: Stabilizing selection selects for intermediate values, 
+reducing the variance of the trait under selection. Disruptive selection
 selects for extreme values and increases the variance.
@@
 We also examine selection effects among respondents' parents, using information
-on respondents' number of siblings. Effect sizes of polygenic scores are highly
-correlated across the two generations. Median-splitting respondents by year of
-birth, we find little evidence of change in effect sizes among either the
-parents' or the respondents' generation. We also check whether selection effects
-vary by socio-economic status and AFLB in the parents' generation. We have no
+on respondents' number of siblings to calculate parents' RLRS. Effect sizes of
+polygenic scores are highly correlated across the two generations (Appendix
+Figure \@ref(fig:plot-res-sibs)). Median-splitting respondents by year of birth,
+we find little evidence of change in effect sizes among either the parents' or
+the respondents' generation. We also check whether selection effects vary by
+socio-economic status and AFLB in the parents' generation. We have no
 information about parents' income, so we use the 1971 Townsend deprivation score
@@
 the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
-to 1. Letting $u(\cdot)$ take the constant relative risk aversion form
-$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$, where $\sigma>0$ captures risk aversion
-in income, we examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
+to 1, and let $u(\cdot)$ take the constant relative risk aversion form
+$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$. $\sigma > 0$ measures the curvature
+of the utility function, i.e. the decline in marginal utility of income as income
+increases. We examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
 *fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$ in
@@
 between polygenic scores and fertility.[^one-period-results] The key insight of
-the model is that for middling levels of risk aversion, the substitution effect
+the model is that for middling levels of $\sigma$, the substitution effect
 dominates at low income levels, but is balanced by the income effect as income
@@
 res_mediation %<>% mutate(
-                     prop = estimate_ind/estimate_total,
                      # 0.025 not 0.05 because two-sided
-                     sig  = abs(statistic_ind) > qnorm(1 - 0.025/33)
-                   )
-
-n_pos_sig <- sum(res_mediation$sig & res_mediation$prop > 0)
-
-median_prop <- median(res_mediation$prop[res_mediation$prop > 0 & 
-                                           res_mediation$sig])
-
-```
-
-Second, we run a mediation analysis to directly test whether the correlation
-between each polygenic score and fertility is mediated by educational attainment
-(Appendix Table \@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the
+                     sig     = abs(statistic_ind) > qnorm(1 - 0.025/33),
+                     pos_sig = sig & prop_ind > 0
+                   )
+
+n_pos_sig <- sum(res_mediation$pos_sig)
+median_prop <- median(res_mediation$prop_ind[res_mediation$pos_sig])
+
+```
+
+Second, for the `r nrow(res_mediation)` scores where the bivariate correlation in
+\eqref{eq:regression} is significant at $p < 0.05/33$, we run a mediation
+analysis to directly test whether the correlation between each polygenic score
+and fertility is mediated by educational attainment (Appendix Table
+\@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the
 indirect effect of the score on fertility *via* educational attainment is
-significantly different from zero ($p < 0.05/33$). Among these scores, the
-median proportion of the total effect explained by the indirect effect is 
-`r percent(median_prop)`. The educational attainment variable is a relatively
-crude measure of human capital: more accurate measures would likely
-explain more of the total effect.
+significantly different from zero ($p < 0.05/33$) and takes the same sign as the
+overall effect. Among these scores, the median proportion of the total effect
+explained by the indirect effect is `r percent(median_prop)`. Figure
+\@ref(fig:plot-mediation) shows estimated proportions explained along
+with bootstrap 95% confidence intervals (uncorrected; 100 bootstraps).
+The educational attainment variable is a relatively crude measure of human 
+capital: more accurate measures would likely explain more of the total effect.
+
+```{r plot-mediation, fig.align = "center", fig.cap = "Proportion of selection effect mediated by educational attainment, polygenic scores with significant selection effects"}
+
+res_mediation %>% 
+      mutate(
+        term = pretty_names(term),
+        term = fct_reorder(term, prop_ind, order_abs(1)),
+        `Indirect effect p < 0.05/33` = sig 
+      ) %>% 
+      ggplot(aes(y = term)) + 
+        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/33`), 
+                   width = .5) + 
+        geom_errorbar(aes(
+                          xmin = prop_ind_conf_low, 
+                          xmax = prop_ind_conf_high,
+                          linetype = "Bootstrap 95% c.i.",
+                        ),
+                        width = 0.3,
+                        color = "grey30"
+                      ) +
+        scale_fill_manual(values = pval_palette) +
+        my_vline +
+        scale_x_continuous(labels = scales::percent, n.breaks = 7) +
+        coord_cartesian(xlim = c(-0.5, 1)) +
+        labs(x = "Proportion mediated", y = "") +
+        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i." = 1))
+
+```
 
@@
 
-res_all %<>% filter(dep.var == "N children", reg.type == "controlled")
+res_all %<>% filter(dep.var == "RLRS", reg.type == "controlled")
 res_fe_fertility %<>% filter(Regression == "Raw")
@@
 $a$ in the model [cf. @jones2008fertility]. This theory would not explain why
-selection effects are smaller at higher incomes and education levels. (Indeed,
+selection effects are smaller at higher incomes and education levels. Indeed,
 in the model, the effect of $a$ on fertility becomes stronger at higher levels
-of human capital.) A third alternative is that traits under selection are linked
+of human capital. A third alternative is that traits under selection are linked
 to externalizing behaviour and risk-seeking. This might be partially captured by
-our risk-aversion parameter $\sigma$, which affects fertility by changing the
-marginal utility of income, but a more direct channel is risky sexual behaviour
+our parameter $\sigma$, which can be interpreted as a measure of risk aversion
+over income. However, a more direct channel is risky sexual behaviour
 [@mills2021identification]. The data here provide some support for this: scores
@@
 \clearpage
-## Balancing and diversifying selection
-
-Balancing selection reduces variance in the trait under selection, while
-diversifying selection increases variance. To check for these, we rerun equation
+## Stabillizing and disruptive selection
+
+Stabilizing selection reduces variance in the trait under selection, while
+disruptive selection increases variance. To check for these, we rerun equation
 \@ref(eq:regression), adding a quadratic term in $PGS_i$, and using our
 age/qualification weights. Scores for hip circumference show significant
-balancing selection ($p < 0.05/33$, negative coefficient on quadratic term).
-Scores for educational attainment (EA2 and EA3) show significant diversifying
+stabilizing selection ($p < 0.05/33$, negative coefficient on quadratic term).
+The EA2 score for educational attainment shows significant disruptive
 selection ($p < 0.05/33$, positive coefficient), which reduces the strength of
@@
 
-We also checked for balancing selection in the parents' generation, using
+We also checked for stabilizing selection in the parents' generation, using
 age/qualification weights multiplied by the inverse of number of siblings.
-Scores for EA2 and EA3 again show significant diversifying selection ($p <
-0.05/33$, positive coefficient). Other scores including hip circumference were 
-not significant.
-
-```{r plot-purifying, fig.align = "center", fig.cap = "Purifying/diversifying selection: predicted number of children by polygenic score.", fig.height = 3}
+Scores for EA2 and EA3 show significant disruptive selection ($p <
+0.05/33$, positive coefficient on quadratic). Other scores including 
+hip circumference were not significant.
+
+```{r plot-purifying, fig.align = "center", fig.cap = "Stabilizing/disruptive selection: predicted number of children by polygenic score.", fig.height = 3}
 
@@
 
-op <- par(mfrow = c(1, 3))
+op <- par(mfrow = c(1, 2))
 for (score in sig) {
@@
 
-The UK Biobank data contains information on respondents' number of
-siblings, i.e. their parents' number of children. Since respondents' polygenic 
-scores are equal in expectation
-to the mean scores of their parents, we can use this to look at selection 
-effects in the parents' generation. We estimate equation (\ref{eq:regression})
-using *number of siblings* (including the respondent) as the dependent variable.
-The parents' generation has an additional source of ascertainment bias: sampling
-parents of respondents overweights parents who have many children. For instance,
-parents of three children will have, on average, three times more children
-represented in UK Biobank than parents of one child. Parents of no children will
-by definition not be represented. To compensate, we reweight our preferred
-weightings (Age/Qualification) by the inverse of *number of siblings*.
-
-Figure \@ref(fig:plot-res-sibs) shows regressions of *number of siblings* on
+The UK Biobank data contains information on respondents' number of siblings
+(including them), i.e. their parents' number of children. Since respondents'
+polygenic scores are equal in expectation to the mean scores of their parents,
+we can use this to look at selection effects in the parents' generation. We
+estimate equation (\ref{eq:regression}) using parents' RLRS as the dependent
+variable. The parents' generation has an additional source of ascertainment
+bias: sampling parents of respondents overweights parents who have many
+children. For instance, parents of three children will have, on average, three
+times more children represented in UK Biobank than parents of one child. Parents
+of no children will by definition not be represented. To compensate, we reweight
+our preferred weightings (Age/Qualification) by the inverse of *number of
+siblings*.
+
+Figure \@ref(fig:plot-res-sibs) shows regressions of parents' RLRS on
 polygenic scores. For a clean comparison with the respondents' generation, we
-rerun regressions on *number of children* excluding those with no children, and
+rerun regressions on respondents' RLRS excluding those with no children, and
 show results in the figure. Selection effects are highly correlated across the
@@
 are larger for the parents' generation. We treat this result cautiously, because
-when we split respondents up by year of birth, we find few differences in effect
-sizes between early- and late-born respondents, for either generation. In other
-words, since estimated effect sizes change across "generations", but do not
-change over time within either generation, the change may be due to remaining
-ascertainment bias within the sample or other effects. In particular, effect
-sizes in both generations may depend on polygenic scores' correlation with
+effect sizes in both generations may depend on polygenic scores' correlation with
 childlessness, and we cannot estimate this for the parents' generation.
 
-Excluding childless people in the parents' generation could bias our estimates.
-To learn about this, we compare effect sizes excluding and including childless
+To learn more about this, we compare effect sizes excluding and including childless
 people in the *current* generation. The correlation
@@
 
-Tables \@ref(tab:tbl-siblings-by-period) and
-\@ref(tab:tbl-children-by-period) summarize the results. There is no
-evidence for changes in selection effects within the parents' generation. In the
-respondents' generation, effect sizes were significantly larger in absolute size 
-among the later-born for four polygenic scores: cognitive ability, EA2, EA3 and
-extraversion. These changes are inconsistent with the intergenerational change, 
+Tables \@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)
+summarize the results. We report the number of scores showing significant
+changes over time (i.e. a significant interaction between polygenic score and
+the "late born" dummy): either a significant change in sign, a significant
+increase in size, or a significant decrease in size. There is little evidence
+for changes in selection effects within the parents' generation, with just one
+score showing a significant decrease in size. In the respondents' generation,
+effect sizes were significantly larger in absolute size among the later-born for
+nine polygenic scores: ADHD, age at menopause, caffeine consumption, cognitive
+ability, coronary artery disease, EA2, EA3, extraversion and major depressive
+disorder. These changes are inconsistent with the intergenerational change,
 where estimated effect sizes were larger among the earlier, parents' generation.
-One score, conscientiousness, showed a significant change in sign, from 
+One score, for alzheimer's disease, showed a significant change in sign, from
 negative to positive effects on fertility.
 
-Overall, there is weak evidence for change over time. The clearest results are
-that (a) the direction of selection, and (b) the pattern of relative effect sizes
-across scores, are broadly consistent over time.
+Overall, there is some evidence for an increase in the strength of selection in
+recent history. The clearest result is that the pattern of relative effect sizes
+across scores is broadly consistent over time.
 
@@
       count() %>% 
-      period_table("Change in selection effects between parents of early and late born respondents (regressions on number of siblings). Weights: Age/Qualification times 1/number of siblings.")
+      period_table("Change in selection effects between parents of early and late born respondents (regressions on parents' RLRS). Weights: Age/Qualification times 1/number of siblings.")
 
@@
       count() %>% 
-      period_table("Change in selection effects between early and late born respondents (regressions on number of children). Weights: Age/Qualification.")
+      period_table("Change in selection effects between early and late born respondents (regressions on RLRS). Weights: Age/Qualification.")
 ```
@@
 
-We run a standard mediation analysis in the framework of @baron1986moderator. We estimate 
+We run a standard mediation analysis in the framework of @baron1986moderator.
+For each polygenic score where the bivariate correlation with fertility is 
+significant at $p$ < 0.05/33, we estimate 
 \begin{align}
@@
 \end{align}
-where $N_i$ is fertility, $PGS_i$ is each polygenic score, $EA_i$ is educational
+where $N_i$ is fertility, $PGS_i$ is the polygenic score, $EA_i$ is educational
 attainment (age of leaving fulltime education), and $X_i$ is a vector of
@@
 ```{r plot-res-fe-fertility, fig.cap = "Selection effects controlling for sibling-group fixed effects, with and without a control for education (left education before 16, 16-18, or after 18).
-Each set of 29 results is from a single regression of \\emph{n children} on 29 polygenic scores.
+Each set of 29 results is from a single regression of RLRS on 29 polygenic scores.
 Standard errors clustered by sibling group."}
@@
 
-## Estimating the effect of natural selection on the contribution of PSEA to inequality
+## Estimating the effect of natural selection on the contribution of polygenic scores for educational attainment to inequality
 
@@
 prop_wt_fe <- scales::percent(prop_wt_fe, accuracy = 0.1)
+
+prop_wt_fe_bounds <- scales::percent(res$r2_fe_bounds, accuracy = 0.1)
 
@@
 educational attainment explained by the "true" polygenic score for educational
-attainment $p^{*}$, i.e. the best linear unbiased predictor of educational
+attainment (PSEA) $p^{*}$, i.e. the best linear unbiased predictor of educational
 attainment from genetic data. We write the measured polygenic
@@
 confounding environmental variation such as genetic nurture. Within-siblings
-regression would give a smaller estimate of the effect of EA3. However, this
-would also affect the reliability of the regression [@griliches1986errors],
-making it hard to back out estimates of the true $R^2$. Nevertheless, we
-can calculate the proportional increase in $R^2$ due to natural selection, again 
-so long as the reliability coefficient is unaffected by
-the weighting scheme. Running within-siblings regressions of EA3 on income group
-gives within-$R^2$ estimates of `r res$r2_income_ea3_fe` (weighted by
-age and qualification) and `r res$r2_income_ea3_fe_wt` (multiplying weights by number of children).
-Thus, this procedure estimates the proportional change in $R^2_{income}$ due to natural selection as `r prop_wt_fe`.
+regression avoid this, but the smaller N makes estimates noisier. Running
+within-siblings regressions of EA3 on income group gives within-$R^2$ estimates
+of `r res$r2_income_ea3_fe` (weighted by age and qualification) and `r
+res$r2_income_ea3_fe_wt` (multiplying weights by number of children). Thus, this
+procedure estimates the proportional change in $R^2_{income}$ due to natural
+selection as `r prop_wt_fe`. However, bootstrap 95% confidence intervals
+are large (`r prop_wt_fe_bounds[1]` to `r prop_wt_fe_bounds[2]`). We therefore
+view this result more as suggesting that environmental confounds likely
+bias the estimated change in $R^2_{income}$ towards zero.
 

### 2022-03-14 _drake.R

@@ -171,27 +171,27 @@ plan <- drake_plan(
   
   res_all = {
     res_sibs <- run_regs_basic(
-            dep_var     = "n_sibs", 
+            dep_var     = "RLRS_parents", 
             score_names = score_names,
             famhist     = famhist_pw,
             weights     = famhist_pw$weights
           )
     res_chn <- run_regs_basic(
-            dep_var     = "n_children", 
+            dep_var     = "RLRS", 
             score_names = score_names, 
             famhist     = famhist_kids
           )
     dplyr::bind_rows(
-            "N siblings" = res_sibs, 
-            "N children" = res_chn, 
-            .id = "dep.var"
+            "RLRS_parents" = res_sibs, 
+            "RLRS"         = res_chn, 
+            .id            = "dep.var"
           )
   },
   
   res_quadratic = {
     map_dfr(score_names, 
               ~run_regs_fml(
-                             "n_children ~ {score_name} + I({score_name}^2)", 
+                             "RLRS ~ {score_name} + I({score_name}^2)", 
                              score_name = .x,
                              famhist    = famhist_kids
                            ),
@@ -202,7 +202,7 @@ plan <- drake_plan(
   res_sibs_quadratic = {
     map_dfr(score_names, 
             ~run_regs_fml(
-              "n_sibs ~ {score_name} + I({score_name}^2)", 
+              "RLRS_parents ~ {score_name} + I({score_name}^2)", 
               score_name = .x,
               famhist    = famhist_pw,
               weights    = quote(weights)
@@ -214,22 +214,22 @@ plan <- drake_plan(
   res_pcs = {
     pc_names <- grep("PC", names(pcs), value = TRUE)
     res_sibs_pcs <- run_regs_pcs( 
-      dep_var     = "n_sibs", 
+      dep_var     = "RLRS_parents", 
       famhist     = famhist_pw,
       pcs         = pcs,
       weights     = quote(famhist$weights) # famhist will be famhist_pw in the function
     )
     
     res_chn_pcs <- run_regs_pcs(
-      dep_var     = "n_children", 
+      dep_var     = "RLRS", 
       famhist     = famhist_kids,
       pcs         = pcs,
       weights     = NULL 
     )
     dplyr::bind_rows(
-      "N siblings" = res_sibs_pcs, 
-      "N children" = res_chn_pcs, 
-      .id = "dep.var"
+      "RLRS_parents" = res_sibs_pcs, 
[…]

### 2022-03-14 make-census-weights.R

@@ -61,7 +61,7 @@ make_census_age_qual <- function (DC1108EW_file, DC5202SC_file) {
   sc_age_qual <- readr::read_csv(DC5202SC_file, skip = 12)
   sc_age_qual %<>% 
     select(
-      age_cat = `...1`,            # new names reflect values in the columns
+      age_cat =  X1,            # new names reflect values in the columns
       qual    = `Ethnicity (Flat)`, 
       value   = `White: Total`
     ) %>% 

### 2022-03-14 make-data.R

@@ -304,7 +304,7 @@ add_deprivation_data <- function (famhist, dep_data_dir) {
 
 
 add_fertility_prs <- function (famhist, fertility_data_dir) {
-  fertility_prs <- readr::read_table(
+  fertility_prs <- readr::read_table2(
                      file.path(fertility_data_dir, 
                                "UKB.AMC.NEB_POOLED.HM3.EUR.SBLUP.10k.csv")
   )

### 2022-03-14 regressions.R

@@ -118,7 +118,7 @@ run_regs_period <- function (children, score_name, famhist, weight_data) {
   
   famhist <- inner_join(famhist, weight_data, by = "f.eid")
   
-  dep_var  <- if (children) "n_children" else "n_sibs"
+  dep_var  <- if (children) "RLRS" else "RLRS_parents"
   subset   <- if (children) quote(kids_ss) else NULL
   famhist$year_split <- famhist$YOB >= 1950
   # evaluates to 1950:
@@ -142,7 +142,7 @@ run_regs_period <- function (children, score_name, famhist, weight_data) {
 
 
 run_regs_subset <- function(score_name, subset, famhist) {
-  fml <- as.formula(glue::glue("n_children ~ {score_name}"))
+  fml <- as.formula(glue::glue("RLRS ~ {score_name}"))
   mod <- lm(fml, famhist, subset = eval(subset))
   res <- tidy(mod, conf.int = TRUE)
   res <- filter(res, term == {{score_name}})
@@ -164,7 +164,7 @@ run_regs_fml <- function(fml, ..., subset = NULL, famhist, weights = NULL) {
 
 
 run_regs_age_flb_parents_cross <- function(famhist, score_names, age_birth_var) {
-  fml <- sprintf("n_sibs ~ %s + {score_name}:%s", age_birth_var, age_birth_var)
+  fml <- sprintf("RLRS_parents ~ %s + {score_name}:%s", age_birth_var, age_birth_var)
   res <- map_dfr(score_names, 
                  ~run_regs_fml(
                    fml        = fml,
@@ -196,7 +196,7 @@ run_regs_mnlogit <- function (score_names, fhl_mlogit) {
 
 run_age_anova <- function (score_name, control, famhist) {
   fml <- as.formula(glue::glue(
-    "n_children ~ {score_name}*({control} + 
+    "RLRS ~ {score_name}*({control} + 
           age_at_recruitment + I(age_at_recruitment^2))"))
   res <- lm(fml, data = famhist) %>% 
            car::Anova() %>% 
@@ -231,15 +231,6 @@ run_reg_income_dv <- function (famhist, score_names, ashe_income, age_qual_weigh
   famhist$pay_hat <- pay_hat
 }
 
-run_income_dist <- function (famhist) {
-  # predict individual income from PGS
-  # 3 densities:
-  # - empirically from last_job_pay, weighted by e.g. age_qual
-  # - predicted from reg of last_job_pay on PGS, same weighting
[…]

### 2022-03-14 response-to-reviewers-BE.Rmd
Shorten section 2, move some results to appendix.

@@
 ---
-title: "Response to reviewers"
-author: "David Hugh-Jones and Abdel Abdellaoui"
-date: "`r format(Sys.Date(), '%d %B %Y')`"
 output: 
@@
   - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
+mainfont: Baskerville
 ---
@@
 
+Code to produce this paper is now publicly available at <https://github.com/hughjonesd/negative-selection>. XXX
+
+With best wishes,
+
+David Hugh-Jones\
+Abdel Abdellaoui
+
 ## Editor
@@
 We now mention this work in our conclusion.
-XXX maybe do more.
 
@@
 Gauthier (2007).
-Second, my strong gut feeling is that it is wrong for governments to (bluntly) institute breeding programs for their citizens.
+Second, my strong gut feeling is that it is wrong for governments to (bluntly put) institute breeding programs for their citizens.
 I wrote more informally on this topic [here](https://wyclif.substack.com/p/the-natural-selection-paper-part-822?s=w).
@@
 (Of course the polygenic score for educational attainment does also predict IQ.) The negative fertility-income relationship, in developed countries, is a very well-established stylized fact, and we mention some papers that also show income is causal.
+We now mention two of the cited papers.
 
@@
 When we do correct for ascertainment bias, every correction increases our estimated effect sizes.
-So we expect that any remaining ascertainment also probably biases our estimates towards zero.
-We now point this out.
-(XXX say this somewhere?)
+So we expect that any remaining ascertainment is also probably biasing our estimates towards zero.
+We now say this.
 
@@
 
-We have tried to work on this... XXX.
+We have tried to work on this. Unconnected results and minor points have been pushed to footnotes and/or the appendix.
+
+We don't want to go too far, by pretending that the theory came to us fully-formed in our ivory tower, and then we tested it.
+In fact, the motivating results led us to develop the theory, first informally and then more precisely; only the results in section 4 should be considered a clean test.
 
@@
 
-We have shortened the discussion as follows... XXX.
+As described above, we have shortened Section 2 to focus on the key results of variation by income, education, number of partners, partner presence and AFLB.
+These empirics now take up just 39 lines of text, plus Figures 3-5.
 
@@
 
-Following the advice of Franz (2007), we compromise: we bootstrap uncorrected 95% CIs of the ratio, but only when the denominator (i.e.
+Following the advice of Franz (2007), we now compromise: we bootstrap uncorrected 95% CIs of the ratio, but only when the denominator (i.e.
 the total effect) is significantly different from 0 at $p$ \< 0.05/33.
@@
 
-We welcome any further thoughts.
+We welcome any further suggestions.
 
@@
 
+Oops, yes. We hope the corrected footnote is now right.
+
 We now use RLRS as the dependent variable for both respondents and parents.
-Results are not much changed, except there is now slightly more evidence for larger selection effects over time: 9 polygenic scores show a significant increase in size when we median-split respondents by birth year.
+Results are not much changed, except there is now slightly more evidence for larger selection effects over time: 9 polygenic scores show a significant increase in effect size on respondents' RLRS when we median-split them by birth year.
+In turn, this makes us less sure that we can rule out that the welfare state is a relevant explanation for natural selection, so we have slightly qualified our comments on this topic.
 
@@
 We have adjusted this to 50 male/45 female as suggested.
-Results are little changed, except that EA3 is no longer undergoing significant diversifying selection in the respondents' generation.
+Results are little changed, except that EA3 is no longer undergoing diversifying selection in the respondents' generation at the p = 0.05/33 significance threshold.
 
@@
 
-We have experimented a lot with these figures, and they may not be easy to improve.
+We experimented a lot with these figures, and they may not be easy to improve.
 For example, we tried using symbols, but in our judgment the added visual noise makes it harder to see the pattern of the results across rows.
 Below we show three alternative versions of one of our figures: our original, a version using symbols, and one with subtle tweaks to colors and stroke sizes.
-We prefer our tweaked version (and have used it in the revision), but we are open to suggestions.
+We prefer our tweaked version (and have used it in the revision), but we will bow to the reviewers or editor.
 

### 2022-03-14 why-natural-selection.Rmd
Shorten section 2, move some results to appendix.

@@
           my_theme() +
-          scale_linetype_manual(name = "", values = c("95% c.i." = 1))
+          scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1))
 
@@
 
-```{r NOTES-TODO, eval=FALSE}
-"
-
-# TODO response to PNAS
-
-* Only use education, not first job pay, in "controlled" regressions; check
-  description is correct - done.
-* Add CIs - done.
-* Clarify collider bias/endogeneity - done.
-* Formalize theory - done.
-  - Comment on results where effects are actually reversed. This works
-    nicely if you throw in a money cost - done computational example.
-* Clarify causality? - done...
-  - Helps to deal with the "collider bias" argument of reviewer
-  - Might also be a better test of mediation via human capital
-  - One other possibility. If we use only people with first profession
-    teacher, that (as it happens) shuts down the EA3-income correlation.
-    It also shuts down the EA3-fertility relationship, suggesting that
-    income is indeed the mediator.
-* Clarify "what we learn about genetic links to fertility by using economic theory to interpret relationships between polygenic scores and fertility."
-  - Point is that we can explain how natural selection is occurring.
-* "Discuss empirical predictions for AFLB grounded in dynamic theory"
-  - hmm, tricky, maybe we can include it in discussion of economic results at
-  end of formal section.
-  - think our theory with 2 periods probably covers this
-* Why don't women see stronger effects? But note, they do: not significant
-  after correction for multiple testing, but a 25% difference. 
-  - I think we leave this for reviewers.
-* Quadratic regressions for parents - DONE
-
-# Open questions/TODO
-
-* Limit max no. of children in regressions (use an "in_children" bool)
-  - not necessary. Very few have abnormally many.
-* Consider doing separate regressions for "any children" vs. "n children | 1 or more"
-  - heckit too slow and gives crazy answers, just use plm/ols
-  - some variables do look different
-* If we weight with the GHS, and plot the means of figure 1 using this w
-  weighting, most things stay the same - but the change in EA is reversed! EA
-  increases over time. Could this be because the sample includes many
-  relatively educated older people? If so, then mean effects are *overestimated*
-  by figure 1.
-* Reweight other regressions (e.g. the omnibus regression in the appendix)?
-
-* If you take "year of last job started" and see if it is after/before
-  the first child, you can figure out whether mothers keep working.
-  - better still, look at all jobs started and see if any are within e.g. 5 years of
-  first child's birth
-  - this could then be another way to get at income vs substitution effects. You'd
-  predict negative selection effects among mothers who worked. Of course the
-  choice to work is endogenous.
-* double check which scores are reverse coded!
-
-# WHAT MODEL FITS YOUR DATA
-## ASSUMPTIONS
-* Genetics produce human capital h. 
-* h and education are compliments in producing earnings
-* There are income and substitution effects on fertility:
-  - income effects are stronger at high income levels (Becker and Tomes 1976)
-    and among couples
-* People seek partners when in the labour market. Women match a person, observe
-  their quality, and decide whether to have a first child.
-* Then they (may) have more children, either with the partner or not 
-  (high quality partners stay around more)
-* High quality partners increase household income
-
-## RESULTS
-* People with low h are more likely to leave education early and enter the labour market
-  - because e and h are complements
-  - this is part of the substitution effect, as they expect to spend some time
-    raising children rather than being in work
-* At low values of h, increases in h reduce fertility (substitution effect dominates)
-  at higher values, smaller reductions or increases (income effect stronger)
-  - because h predicts income
-* In 1-parent households, increases in h reduce fertility (sub. effect), in 
-  2-parent households, smaller reductions or increases (income effect stronger)
-  - because h predicts income and because stronger income effect among couples
-* People with low h start having children earlier (b/c leave educ earlier)
-  - so part of the substitution effect comes via this
-  - controlling for this the income effect dominates
-  - hence increases in h reduce fertility less once we control for AFLB
-
-## THOUGHTS
-* Should we distinguish between men and women in couples?
-  - within couples, male h should increase fertility (via earnings)
-  - female h should decrease it (ditto?)
-  - see explore-sex-vs-pay.R
-
-## RIVAL MECHANISMS
-* Lynn 2011 mentions two:
-  - social class gradient on ability to delay gratification
-  - modern society protects the weak (Darwin 1871, Spencer 1874)
-    NB this latter doesn't predict more fertility among "dysgenic" ppl, just
-    equal fertility and then genetic drift does its work
-    - examples include modern sanitation, post-1800
-    - and monogamy!
-"
-```
-
 # Introduction
@@
 # Results
-
-```{r calc-sex-ratio}
-female_ratio <- mean(famhist$sex == 0)
-```
 
@@
 \begin{equation}
-RLRS_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
+\mathrm{RLRS}_i = \alpha + \beta\mathrm{PGS}_i + \varepsilon_i
 \label{eq:regression}
@@
 
-[^normalization]: XXX The selection effect $\beta$ equals $Cov(N, PGS)/Var(PGS)$
+[^normalization]:The selection effect $\beta$ equals $Cov(RLRS, PGS)/Var(PGS)$
 where $N$ is the number of children. Since PGS are normalized to variance 1 and
 mean 0, this reduces to 
-$Cov(N, PGS) \equiv E(N\ PGS) - E(N) E(PGS) = E(N\ PGS)$. This is the polygenic 
-score weighted by the number of children, which is the average polygenic score 
+$Cov(RLRS, PGS) = E(RLRS\times PGS) - E(RLRS) E(PGS) = E(RLRS \times PGS)$. This is the polygenic 
+score weighted by relative lifetime reproductive success, which is the average polygenic score 
 in the next generation [@robertson1966mathematical].
@@
 
-We also check for stabilizing and disruptive selection by estimating
-\@ref(eq:regression) with a quadratic term.[^balance-diverse-def] In particular,
-we find disruptive selection for educational attainment polygenic scores: at
-higher values of these scores, the negative effect on fertility is smaller
-(Appendix Figure \@ref(fig:plot-purifying)).
-
-[^balance-diverse-def]: Stabilizing selection selects for intermediate values, 
-reducing the variance of the trait under selection. Disruptive selection
-selects for extreme values and increases the variance.
-
-```{r plot-res-wt, fig.cap = "Selection effects: weighted and unweighted regressions.  Each point represents a single bivariate regression of number of children on a polygenic score."}
+```{r plot-res-wt, fig.cap = "Selection effects: weighted and unweighted regressions.
+Each point represents a single bivariate regression of number of children on a polygenic score.
+P value threshold is 0.05, Bonferroni-corrected for multiple comparisons.
+Confidence intervals are uncorrected."}
 
@@
 lifetime number of sexual partners.
-We emphasize that these categories are not necessarily exogenous to the polygenic
-scores. For example, both in the data and in our theoretical model, age at first
-live birth is a choice variable, which is endogenous to human-capital-related
-polygenic scores (Appendix Figure \@ref(fig:plot-age-flb-dv)). Nevertheless,
-differences in equation \@ref(eq:regression) across subgroups constrain the set
-of possible explanations for natural selection. As we show below, a simple
-economic model predicts the differences we find.
 
@@
 
-```{r calc-income-edu-controlled}
-
-drake::loadd(res_income_controlled, res_edu_controlled)
-n_regs_ic <- nrow(res_income_controlled)
-n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
-
-n_regs_educ <- nrow(res_edu_controlled)
-n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
-```
-
 Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
@@
 and by household income. Effects are larger and more significant for the lowest
-income category, and for the lowest education category.  These results could be
-explained by age, if older respondents have lower income and are less educated,
-and also show more natural selection on polygenic scores. However, when we rerun
-the regressions, interacting the polygenic score with income category and also
-with a quadratic in age, the interaction with income remains significant at
-$p < 0.05/`r n_regs_ic`$ for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
-if we interact the PGS with age of leaving full time education and a quadratic
-in age, the interaction with age leaving full time education remains significant
-at $p < 0.05/`r n_regs_educ`$ for `r n_sig_educ` out of `r n_regs_educ` regressions.
-
-Selection effects are also different between men and women (Appendix Figure
-\@ref(fig:plot-sexes)). Differences are particularly large for educational
-attainment, height and MDD. Several polygenic scores for mental illness and
-personality traits are more selected for (or less against) among women,
-including major depressive disorder (MDD), schizophrenia and neuroticism, while
-extraversion is more selected for among men. Scores for waist circumference and
-waist-hip ratio are less selected for among women. One
-possible reason for these sex differences is that polygenic scores may affect
-fertility via success in marriage markets, and men and women may value different
-characteristics in these.
+income category, and for the lowest education category. These results are robust 
+to controlling for respondents' age (Appendix section \@ref(sec:age-control)).
 
@@
 Effects are larger among those not living with a spouse or partner
-(Figure \@ref(fig:plot-n-partners)b).
+(Figure \@ref(fig:plot-n-partners)b).[^sex]
+
+[^sex]: We also compared selection effects between men and women (Appendix Figure
+\@ref(fig:plot-sexes)).
 
@@
 
-[^aflb-no-men]: This information is unavailable for men.
-
-```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and household type.", fig.subcap = c("Lifetime number of sexual partners", "Household type"), fig.ncol = 1, fig.align = "center", fig.height = 4}
+[^aflb-no-men]: AFLB is unavailable for men.
+
+```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and household type.", fig.subcap = c("Lifetime number of sexual partners", "Household type"), fig.ncol = 1, fig.align = "center", fig.height = 4, fig.width = 4.5}
 
@@
 `r cor_raw_flb`. Thus, selection effects seem to come through two opposing
-channels: an effect on AFLB, and an opposite-signed effect on number of children
-after AFLB is controlled for.
+channels: a correlation with AFLB, and an opposite-signed correlation with 
+number of children after AFLB is controlled for.
 
@@
 ```
+
+We emphasize that our categories are not necessarily exogenous to the polygenic
+scores. For example, in the data, and also in our theoretical model, age at first
+live birth is a choice variable, which is endogenous to human-capital-related
+polygenic scores (Appendix Figure \@ref(fig:plot-age-flb-dv)). Nevertheless,
+differences in equation \@ref(eq:regression) across subgroups constrain the set
+of possible explanations for natural selection. As we show below, a simple
+economic model predicts the differences we find.
 
@@
 Figure \@ref(fig:plot-res-sibs)). Median-splitting respondents by year of birth,
-we find little evidence of change in effect sizes among either the parents' or
-the respondents' generation. We also check whether selection effects vary by
+we find little evidence of change in effect sizes among the parents' generation.
+There is some evidence that selection effect sizes are increasing in the 
+respondents' generation, with 9 polygenic scores showing a significant increase. 
+We also check whether selection effects vary by
 socio-economic status and AFLB in the parents' generation. We have no
@@
 predictions, along with our corresponding empirical results for the correlation
-between polygenic scores and fertility.[^one-period-results] The key insight of
+between polygenic scores and RLRS.[^one-period-results] The key insight of
 the model is that for middling levels of $\sigma$, the substitution effect
@@
 mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
-@becker1981treatise.
+@becker1981treatise. There is also a literature showing negative correlations
+between IQ and fertility [e.g. @lynn2004new; @reeve2018systematic].
 
 ------------------------------------------------------------------------------------
-    **Theory**                                        **Empirical results**
---- ------------------------------------------------- ------------------------------
-1.  The fertility-human capital                       Figures \@ref(fig:plot-means-over-time) 
-    relationship is negative:                         and \@ref(fig:plot-res-wt).
-    $\frac{dN^{*}}{dh} < 0$.
-
-2.  The fertility-human capital                       Figure \@ref(fig:plot-income-educ-level)a. 
-    relationship is weaker (closer to zero)           Selection effects are also weaker at higher
-    at higher wages and/or levels of                  polygenic scores for 
-    human capital.                                    educational attainment (Appendix Figure
+    **Theory: the fertility-human capital           **Empirical results**
+    relationship is...**                                      
+--- ----------------------------------------------- --------------------------------
+1.  Negative: $\frac{dN^{*}}{dh} < 0$.              Figures \@ref(fig:plot-means-over-time) 
+                                                    and \@ref(fig:plot-res-wt).
+
+2.  Weaker (closer to zero) at higher wages and/or  Figure \@ref(fig:plot-income-educ-level)a.
+    levels of human capital.                        Selection effects are also weaker at 
+                                                    higher polygenic scores for 
+                                                    educational attainment (Appendix Figure
                                                     \@ref(fig:plot-purifying)).
 
-3.  The relationship is more negative when            Stronger effects for single parents
-    the time burden of children, $b$, is              (Figure \@ref(fig:plot-n-partners)).
-    larger.
-
-4.  The relationship is weaker at higher              Figure \@ref(fig:plot-income-educ-level)b.
-    levels of education $s$.
-
-5.  The relationship is weaker among those            Effects weaker among those starting 
-    who start fertility in period 2                   fertility later 
-    ($N_{1}^{*} = 0$) than among those who            (Figure \@ref(fig:plot-age-flb)).
-    start fertility in period 1 
+3.  More negative when the time burden              Stronger effects for single parents
+    of children, $b$, is larger.                    (Figure \@ref(fig:plot-n-partners)).
+
+4.  Weaker at higher levels of education ($s$).     Figure \@ref(fig:plot-income-educ-level)b.
+
+5.  Weaker among those who start fertility          Effects weaker among those starting 
+    in period 2 ($N_{1}^{*} = 0$) than among        fertility later 
+    those who start fertility in period 1           (Figure \@ref(fig:plot-age-flb)).
     ($N_{1}^{*} > 0$).
+
 ----------------------------------------------------------------------------------------
@@
 empirical results.
-
-```{r literature, eval = FALSE}
-"Other models of the literature
-
-## LITERATURE NOTES
-
-* Lundberg and Pollak JEP 2007:
-    - several models where nonmarital fertility is more likely for poorer women
-    - Burdett and Ermisch 2002, Willis 1999, Neal 2004
-    - Unmarried parents may not be able to credibly commit to each other
-
-* Happel et al. 1984
-   - childbirth is chosen to minimize loss of earnings from interrupted work.
-   - early childbirth if your skills decay a lot  while childrearing. (No point
-     gaining them then losing them.) Late childbirth otherwise.
-   - late childbirth if you have acquired skills before marriage.
-   - Capital-market imperfections make it wise to delay childbirth (till man has
-     high income to cope while wife is at home). More so for men's wages which
-     rise faster.
-   - empirically: (among married couples) 
-     - high-skill occupation women have children later. Higher-earning men have 
-     them earlier.
-
-- Caucutt et al. 2001
-  - high waged and high educated women (of recent cohorts) more likely to
-    delay childbirth
-
-- Cohen et al. 2013
-  - analyse effect of Israeli subsidy
-  - along the way, find that income affects fertility negatively at low income,
-    positively at high income, supporting Becker and Tomes 76
-
-- Becker and Tomes 1976
-  - investment in children quality and quantity when quality must be equal
-    across children; and some quality comes from an innate endowment.
-  - "observed" quantity-of-children income elasticity rises as income rises
-  - while "observed" quality-of-children elasticity falls as income rises, 
-    b/c more of total quality must be paid for by parents themselves. 
-  - the need to pay more for quality results in lower quantity, since total cost is 
-    quantity X quality
-  - something fishy about argument somehow...
-
-- Lindo 2008
-  - men who lose a job have children sooner, but fewer children in total
-  - controls for all skills effects
-  - so overall effect of income is positive, but timing changes
-
-- Monstad et al. 2008
-  - using an education reform to vary education
-  - no evidence of a causal effect on fertility, but births start later
-  * maybe the usual correlation reflects underlying skills. Your data could
-    test this.
-
-- Autor et al. 2017
-  - the China shock (exogenous shock to income) 
-  - they use Chinese import penetration by industry (instrumented using
-    other countries) times industry employment shares; and separate it for
-    males and females
-  - shocks to male employment decreased fertility and marriage; 
-    increased unmarried mothers
-  - shocks to female employment *increase* marriage and motherhood
-  * this roughly seems to be the consensus
-  * Whereas, you find that male and female *skills* (as 'instrumented' by PGS)
-    have the same effect on fertility - i.e. less skills = higher fertility
-
-- Baudin et al. 2015
-  * look at childlessness vs. fertility (extensive/intensive margin)
-  * some people are too poor to afford children
-  * 'a husband alleviates the time cost of having children'. But for some groups (see appendix)
-    there is a +ve income/fertility gradient
-  * fertility decreases with female educn due to subst effect
-  * baby boom and bust in US: a decline in childlessness caused by poverty, then an increase
-    caused by substn effect ('opportunity cost')
-  * fertility decreases with educn both for single mothers and married women
-    - slightly (25%) higher decrease per year of educn (Table 2) for single mothers
-
-```
 
@@
 and fertility is mediated by educational attainment (Appendix Table
-\@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the
-indirect effect of the score on fertility *via* educational attainment is
-significantly different from zero ($p < 0.05/33$) and takes the same sign as the
-overall effect. Among these scores, the median proportion of the total effect
-explained by the indirect effect is `r percent(median_prop)`. Figure
-\@ref(fig:plot-mediation) shows estimated proportions explained along
-with bootstrap 95% confidence intervals (uncorrected; 100 bootstraps).
-The educational attainment variable is a relatively crude measure of human 
-capital: more accurate measures would likely explain more of the total effect.
+\@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the indirect effect of
+the score on fertility *via* educational attainment takes the same sign as the
+overall effect, and is significantly different from zero ($p < 0.05/33$). Among
+these scores, the median proportion of the total effect explained by the
+indirect effect is `r percent(median_prop)`. Figure \@ref(fig:plot-mediation)
+shows estimated proportions explained along with bootstrap 95% confidence
+intervals (uncorrected; 100 bootstraps). The educational attainment variable is
+a relatively crude measure of human capital: more accurate measures would likely
+explain more of the total effect.
 
@@
 
-```{r calc-fe-fertility}
-drake::loadd(res_fe_fertility)
-drake::loadd(res_all)
-
-educ_reduction <- with(res_fe_fertility, 
-       estimate[Regression == "Controlled" & term == "EA3_excl_23andMe_UK"] / 
-       estimate[Regression == "Raw" & term == "EA3_excl_23andMe_UK"]
-     )
-educ_reduction <- 1 - educ_reduction
-
-res_all %<>% filter(dep.var == "RLRS", reg.type == "controlled")
-res_fe_fertility %<>% filter(Regression == "Raw")
-res_all %<>% left_join(res_fe_fertility, by = "term")
-
-lm_pool_w <- tidy(lm(estimate.y ~ estimate.x, res_all))
-coef_pool_w <- lm_pool_w[[2, "estimate"]]
-se_pool_w <- lm_pool_w[[2, "std.error"]]
-
-```
-
-These results support our theory that natural selection on polygenic scores is
-driven by their *correlation* with human capital. We also test whether polygenic
-scores *cause* fertility by running within-siblings regressions (Appendix Figure
-\@ref(fig:plot-res-fe-fertility)). With a reduced sample size, all
-within-sibling effects are insignificant after Bonferroni correction. However,
-effect sizes are positively correlated with effect sizes from the pooled model,
-and about 50% smaller (regressing within-sibling on pooled effect sizes, $b$ =
-`r format(coef_pool_w, digits = 3)`). This attenuation is in line with the ~50%
-decrease in heritability in within-sibling GWASs on age at first birth and
-educational attainment [@Howe2021.03.05.433935]. We see these results as
-providing tentative evidence that polygenic scores cause fertility, with effects
-being partly driven by correlations with environmental variation in human
-capital. We also reran within-siblings regressions adding a control for
-education. Most effect sizes barely change, suggesting that our measure of
-education does not in general mediate differences in fertility among siblings.
-However, for the educational attainment polygenic score (EA3), effect sizes are
-reduced by `r percent(educ_reduction)`.
-
 ```{r calc-risk-attitude}
@@
 welfare benefits which incentivize child-bearing might be taken up more among
-low-income individuals. However, most effect sizes appear unchanged over a large
-span of twentieth-century history (Appendix Tables
+low-income individuals. However, the majority of effect sizes appear unchanged
+over a large span of twentieth-century history (Appendix Tables
 \@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)), during
@@
 [@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
-affect fertility [@gauthier2007impact]. A second alternative theory is that
+affect fertility [@gauthier2007impact]. Future
+work could test this theory more explicitly. A second alternative theory is that
 polygenic scores correlate with the motivation to have children, i.e. parameter
@@
 
+Code for this paper is available at 
+https://github.com/hughjonesd/negative-selection.
+
 ## Acknowledgements
@@
 ## Natural selection by sex
+
+Figure \@ref(fig:plot-sexes) plots selection effects by sex. Differences are
+particularly large for educational attainment, height and MDD. Several polygenic
+scores for mental illness and personality traits are more selected for (or less
+against) among women, including major depressive disorder (MDD), schizophrenia
+and neuroticism, while extraversion is more selected for among men. Scores for
+waist circumference and waist-hip ratio are less selected for among women. One
+possible reason for these sex differences is that polygenic scores may affect
+fertility via success in marriage markets, and men and women may value different
+characteristics in these.
 
@@
 \clearpage
+## Controlling for age {#sec:age-control}
+
+```{r calc-income-edu-controlled}
+
+drake::loadd(res_income_controlled, res_edu_controlled)
+n_regs_ic <- nrow(res_income_controlled)
+n_sig_ic <- sum(res_income_controlled$p.value < 0.05/n_regs_ic)
+
+n_regs_educ <- nrow(res_edu_controlled)
+n_sig_educ <- sum(res_edu_controlled$p.value < 0.05/n_regs_educ)
+```
+
+Results in Figure \@ref(fig:plot-income-educ-level) could be
+explained by age, if older respondents have lower income and are less educated,
+and also show more natural selection on polygenic scores. However, when we rerun
+the regressions, interacting the polygenic score with income category and also
+with a quadratic in age, the interaction with income remains significant at
+$p < 0.05/`r n_regs_ic`$ for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
+if we interact the PGS with age of leaving full time education and a quadratic
+in age, the interaction with age leaving full time education remains significant
+at $p < 0.05/`r n_regs_educ`$ for `r n_sig_educ` out of `r n_regs_educ` regressions.
+
+\FloatBarrier
+\clearpage
 ## Parents' generation {#sec-parents}
@@
 
-To learn more about this, we compare effect sizes excluding and including childless
-people in the *current* generation. The correlation
-between the two sets of effect sizes is `r cor_inc_vs_excl_childless`. So, patterns across
+To learn more about this, we compare effect sizes excluding and including
+childless people in the *current* generation. The correlation between the two
+sets of effect sizes is `r cor_inc_vs_excl_childless`. So, patterns across
 different scores are broadly similar whether the childless are counted or not.
-However, absolute effect sizes are smaller when the
-childless are excluded, for `r n_smaller` out of 33 scores; the median
-percentage change is `r percent(median_pct_diff)`.
+However, absolute effect sizes are smaller when the childless are excluded, for
+`r n_smaller` out of 33 scores; the median percentage change is 
+`r percent(median_pct_diff)`.
 
@@
 ## Within-siblings regressions
+
+```{r calc-fe-fertility}
+drake::loadd(res_fe_fertility)
+drake::loadd(res_all)
+
+educ_reduction <- with(res_fe_fertility, 
+       estimate[Regression == "Controlled" & term == "EA3_excl_23andMe_UK"] / 
+       estimate[Regression == "Raw" & term == "EA3_excl_23andMe_UK"]
+     )
+educ_reduction <- 1 - educ_reduction
+
+res_all %<>% filter(dep.var == "RLRS", reg.type == "controlled")
+res_fe_fertility %<>% filter(Regression == "Raw")
+res_all %<>% left_join(res_fe_fertility, by = "term")
+
+lm_pool_w <- tidy(lm(estimate.y ~ estimate.x, res_all))
+coef_pool_w <- lm_pool_w[[2, "estimate"]]
+se_pool_w <- lm_pool_w[[2, "std.error"]]
+
+```
+
+Results in the main text support our theory that natural selection on polygenic scores is
+driven by their *correlation* with human capital. Here, we test whether polygenic
+scores *cause* fertility by running within-siblings regressions (Figure
+\@ref(fig:plot-res-fe-fertility)). With a reduced sample size, all
+within-sibling effects are insignificant after Bonferroni correction. However,
+effect sizes are positively correlated with effect sizes from the pooled model,
+and about 50% smaller (regressing within-sibling on pooled effect sizes, $b$ =
+`r format(coef_pool_w, digits = 3)`). This attenuation is in line with the ~50%
+decrease in heritability in within-sibling GWASs on age at first birth and
+educational attainment [@Howe2021.03.05.433935]. We see these results as
+providing tentative evidence that polygenic scores cause fertility, with effects
+being partly driven by correlations with environmental variation in human
+capital. We also reran within-siblings regressions adding a control for
+education. Most effect sizes barely change, suggesting that our measure of
+education does not in general mediate differences in fertility among siblings.
 

### 2022-03-15 response-to-reviewers-BE.Rmd
Response to BG reviewers:   - tweak colors and graphs   - don't split Fig 3 by sex   - make section 3 briefer   - unify tables of change over time

@@
   - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
+  - \usepackage{tcolorbox}
+  - \newtcolorbox{myquote}{colback=green!5!white, colframe=green!75!black}
+  - \renewenvironment{quote}{\begin{myquote}}{\end{myquote}}
 mainfont: Baskerville
+fontsize: 12pt
 ---
 
-```{r echo=FALSE}
+```{r setup, echo=FALSE}
 
@@
     ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
-               shape = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
+               shape = {{fill_col}}, ...)) +
     my_vline + 
     confint_geom_segment +
-    geom_point(size = 1.8, alpha = 0.8) +
-    scale_color_manual(values = c("FALSE" = "grey45", "TRUE" = "orange"), ...) +
+    geom_point(size = 1.8, aes(color = {{fill_col}})) +
+    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/33)) +
+    scale_color_manual(
+                         values = rev(c("lightblue3", "dodgerblue2", "dodgerblue4"))
+                       ) +
     scale_shape_manual(values = 1:n_cats + 14) +
@@
     my_theme() +
-    scale_linetype_manual(name = "", values = c("95% c.i." = 1))
-
+    scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
+    scale_alpha_discrete(range = c(1, 0),
+                         guide = guide_legend(
+                           override.aes = list(
+                             shape = c(1, 16),
+                             color = "black",
+                             alpha = 1,
+                             size = 1.8
+                           )
+                         ))
 }
@@
 
-Thank you 😊.
+Thank you!
 
@@
 > p\. 40 and following: I confess to not following the derivations completely. The authors seem adept at what they do, and I trust that they have made no serious errors.
-
-We hope so too.
-
+>
 > (1) This is my first encounter with the Karush-Kun-Tucker conditions or the Inada condition.
@@
 
-Equation 14 similarly bounds on $h$, as follows.
+Equation 14 similarly bounds $h$, as follows.
 Equation 22 rearranges equation 14.
@@
 We think these analyses are important.
-First, they motivate our theory.
-As we say, though the mediating variables are endogenous, we build a theory in which (e.g.) education is an endogenous choice variable.
+First, they motivate our theory, in which education, income, AFLB etc.
+are indeed endogenous choice variables which are co-determined with fertility.
+If these mediators were *not* endogenous, we'd need a different theory.
 
@@
 These empirics now take up just 39 lines of text, plus Figures 3-5.
+We no longer split respondents by sex in Figure 4.
+This saves two figures, and should reduce space in the final version.
 
@@
 We would like to have more mediators, but appropriate variables are hard to find.
-Earnings in first job reduces our N by a factor of 10, which makes results too noisy to be useful.
+Earnings in first job reduces our N by a factor of 10, which makes results too noisy to be informative.
 As the reviewer suggests, current household income is very much endogenous not only to human capital, but also to previous fertility, which makes it inappropriate as a mediator.
@@
 Alternatives like the Fieller method have the same problem.
-Another issue is that to estimate Bonferroni-corrected significance at p = 0.05/33 we'd need 0.075- and 99.925-percentiles of the distribution, which would require many thousands of bootstraps (infeasibly slow, given our sample size).
+Another issue is that to estimate Bonferroni-corrected significance at p = 0.05/33 we'd need 0.00075- and 0.99925-quantiles of the distribution, which would require many thousands of bootstraps (unfeasibly slow, given our sample size).
 
@@
 
-We experimented a lot with these figures, and they may not be easy to improve.
-For example, we tried using symbols, but in our judgment the added visual noise makes it harder to see the pattern of the results across rows.
-Below we show three alternative versions of one of our figures: our original, a version using symbols, and one with subtle tweaks to colors and stroke sizes.
+We experimented a lot with our basic figures, and they may not be easy to improve.
+The graphs' key message is "effect sizes in one category are larger than others".
+To see that the eye needs to be able to quickly differentiate between categories.
+We have tried several variants using symbols, but in our judgment the added visual noise makes it harder to follow the categories across rows.
+The figure below shows three alternative versions: our original, a version using symbols, and one with subtle tweaks to colors and stroke sizes, which we hope fix the greenness issue and make p-values clearer.
 We prefer our tweaked version (and have used it in the revision), but we will bow to the reviewers or editor.
 
-```{r different-plots, echo = FALSE, fig.cap = "Alternative plot styles", fig.subcap = c("original", "shapes", "tweaked"), fig.ncol = 1, fig.height=2.5, fig.align="center"}
+We've worked on Figures 1 and 2 to improve readability.
+
+```{r different-plots, echo = FALSE, fig.cap = "Alternative plot styles", fig.subcap = c("Original", "Symbols", "Tweaked"), fig.ncol = 1, fig.height=2.5, fig.align="center"}
 
@@
 
-Acemoglu, D., Johnson, S., Robinson, J.A. and Yared, P., 2008. Income and democracy. *American economic review*, *98*(3), pp.808-42.
-
 Chuang, Y.

### 2022-03-15 why-natural-selection.Rmd
Response to BG reviewers:   - tweak colors and graphs   - don't split Fig 3 by sex   - make section 3 briefer   - unify tables of change over time

@@
 Lines are 95% confidence intervals.
-Green triangles show a significant linear increase over time (p < 0.05/33).
-Red squares show a significant decrease."}
+Circles show a significant linear increase (green) or decrease (red) over time (p < 0.05/33)."}
 
@@
       geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
-        fatten = 1.25) +
-      facet_wrap(~score_name, ncol = 5) +
+        fatten = 1.2) +
+      facet_wrap(~score_name, ncol = 6) +
       my_hline +
       scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
-      scale_shape_manual(values = c("-" = "square", "+" = "triangle", "o" = "circle open")) +
+      scale_shape_manual(values = c("-" = "circle", "+" = "circle", "o" = "square")) +
       theme(
+        line        = element_line(size = 0.5, lineend = "round"),
         strip.text  = element_text(size = 6), 
@@
 
-[^normalization]:The selection effect $\beta$ equals $Cov(RLRS, PGS)/Var(PGS)$
-where $N$ is the number of children. Since PGS are normalized to variance 1 and
-mean 0, this reduces to 
-$Cov(RLRS, PGS) = E(RLRS\times PGS) - E(RLRS) E(PGS) = E(RLRS \times PGS)$. This is the polygenic 
-score weighted by relative lifetime reproductive success, which is the average polygenic score 
-in the next generation [@robertson1966mathematical].
+[^normalization]:The selection effect $\beta$ equals $Cov(RLRS, PGS)/Var(PGS)$.
+Since PGS are normalized to variance 1 and mean 0, this reduces to $Cov(RLRS,
+PGS) = E(RLRS\times PGS) - E(RLRS) E(PGS) = E(RLRS \times PGS)$. This is the
+polygenic score weighted by relative lifetime reproductive success, which is the
+average polygenic score in the next generation [@robertson1966mathematical].
 
@@
 and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
-We examine socio-demographic variables including income, education and sex, and
-family structure variables including age at first live birth, presence of a partner, and
-lifetime number of sexual partners.
+We examine socio-demographic variables including income and education, and
+family structure variables including age at first live birth, presence of a
+partner, and lifetime number of sexual partners. Figure
+\@ref(fig:plot-income-educ-level) plots selection effects for each polygenic
+score, grouping respondents by age of completing full-time education, and by
+household income. Effects are larger and more significant for the lowest income
+category, and for the lowest education category. These results are robust to
+controlling for respondents' age (Appendix section \@ref(sec:age-control)).
 
@@
 
-Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
-polygenic score, grouping respondents by age of completing full-time education,
-and by household income. Effects are larger and more significant for the lowest
-income category, and for the lowest education category. These results are robust 
-to controlling for respondents' age (Appendix section \@ref(sec:age-control)).
-
-```{r calc-age-flb-cross}
-
-drake::loadd(res_age_flb_cross)
-
-cor_age_flb <- res_age_flb_cross %>% 
-      mutate(category = gsub(":.*", "", term)) %>% 
-      pivot_wider(score_name, names_from = category, values_from = estimate) %>%              {cor(.$`age_flb_cat10-22`, .$`age_flb_cat28-52`)}
-
-```
-
-We next focus on variables related to household type and reproductive
-strategy. We split males and females by lifetime number of sexual
+Turning to family structure, we split respondents by lifetime number of sexual
 partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
-For both sexes, selection effects are larger and more significant among
+Selection effects are larger and more significant among
 those with more than 3 lifetime partners. Next we split respondents by whether
@@
 
-[^sex]: We also compared selection effects between men and women (Appendix Figure
-\@ref(fig:plot-sexes)).
+[^sex]: The same pattern holds if we analyse men and women separately (Appendix Figure
+\@ref(fig:plot-n-partners-sex)). We also directly compared selection effects 
+between men and women (Appendix Figure \@ref(fig:plot-sexes)).
+
+```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and presence of a partner.", fig.subcap = c("Lifetime number of sexual partners", "Presence of a partner"), fig.ncol = 1, fig.align = "center", fig.height = 4}
+
+drake::loadd(res_partners_joint)
+res_partners_joint %<>%       
+      filter(grepl(":", term)) %>% 
+      mutate(`N partners` = ifelse(grepl("TRUE:", term), "3 or less", "4 or more"))
+
+n_regs <- as.double(nrow(res_partners_joint))
+res_partners_joint %>% 
+      standard_ggplot(fill_col = `N partners`, n_regs = n_regs, 
+                        score_col = score_name)
+
+drake::loadd(res_with_partner)
+res_with_partner %<>% filter(grepl(":", term))
+
+n_regs <- as.double(nrow(res_with_partner))
+
+res_with_partner %>% 
+      mutate(
+        Household = ifelse(grepl("TRUE", term), "With partner", 
+                      "Without partner")
+      ) %>% 
+      standard_ggplot(fill_col = Household, n_regs = n_regs, 
+                        score_col = score_name)
+```
+
+```{r calc-age-flb-cross}
+
+drake::loadd(res_age_flb_cross)
+
+cor_age_flb <- res_age_flb_cross %>% 
+      mutate(category = gsub(":.*", "", term)) %>% 
+      pivot_wider(score_name, names_from = category, values_from = estimate) %>%              {cor(.$`age_flb_cat10-22`, .$`age_flb_cat28-52`)}
+
+```
 
@@
 children born. Figure \@ref(fig:plot-age-flb) shows effect sizes estimated
-separately for each tercile of AFLB. Several effects are strikingly different
-across terciles. ADHD and MDD are selected for amongst the youngest third of
-mothers, but selected against among the oldest two-thirds. Educational
-attainment is selected for among the oldest two-thirds of mothers, but is not
-significantly selected among the youngest third. Similarly, several polygenic
-scores for body measurements are selected against only among older mothers. The
-correlation between effect sizes for the youngest and oldest terciles is 
-`r cor_age_flb`.
+separately for each tercile of AFLB. Effects are strikingly different
+across terciles. Educational attainment, ADHD and MDD are selected for amongst
+the youngest third of mothers, but selected against among the oldest two-thirds.
+Similarly, several polygenic scores for body measurements are selected against
+only among older mothers. The correlation between effect sizes for the youngest
+and oldest terciles is `r cor_age_flb`.
 
 [^aflb-no-men]: AFLB is unavailable for men.
-
-```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and household type.", fig.subcap = c("Lifetime number of sexual partners", "Household type"), fig.ncol = 1, fig.align = "center", fig.height = 4, fig.width = 4.5}
-
-sexes_theme <-  my_theme(
-                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
-                  legend.position      = "bottom", 
-                  legend.box.spacing   = unit(0, "lines"),
-                  legend.box.margin    = margin(-5, 0, 0, -85),
-                  legend.margin        = margin(0, 0, 0, 0),
-                  legend.box.just      = "left",
-                  legend.justification = "left",
-                  legend.spacing       = unit(1, "lines")
-                )
-sexes_guides <- guides(
-                  # fill  = guide_legend(ncol = 1),
-                  # color = guide_legend(ncol = 1)
-                )
-
-drake::loadd(res_partners)
-res_partners %<>%       
-      filter(grepl(":", term)) %>% 
-      mutate(`N partners` = ifelse(grepl("TRUE:", term), "3 or less", "4 or more"))
-
-n_regs <- as.double(nrow(res_partners))
-res_partners %>% 
-      mutate(
-        score_name = pretty_names(score_name)
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(
-        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        x      = estimate, 
-        fill   = `N partners`, 
-        colour = p < 0.05/{{n_regs}}
-      )) + 
-        my_vline +
-        confint_geom_segment() +
-        pval_geom_point() +
-        facet_wrap(vars(sex), scales = "free_y") +
-        my_fill_scale(n = 2) +
-        pval_scale() +
-        sexes_theme +
-        my_labs(x = "", y = "") + 
-        scale_y_reordered() +
-        scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
-        sexes_guides
-
-drake::loadd(res_with_partner_sex)
-res_with_partner_sex %<>% filter(grepl(":", term))
-
-n_regs <- as.double(nrow(res_with_partner_sex))
-
-res_with_partner_sex %>% 
-      mutate(
-        score_name = pretty_names(score_name),
-        Household = ifelse(grepl("TRUE", term), "With partner", 
-                      "Without partner")
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(
-        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        x      = estimate, 
-        fill   = Household, 
-        colour = p < 0.05/{{n_regs}}
-      )) + 
-        my_vline +
-        confint_geom_segment() +
-        pval_geom_point() +
-        facet_wrap(vars(sex), scales = "free_y") +
-        my_fill_scale(n = 2) +
-        pval_scale() +
-        sexes_theme +
-        my_labs(x = "", y = "") + 
-        scale_y_reordered() +
-        scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
-        sexes_guides
-
-```
 
@@
 
+------------------------------------------------------------------------------------
+    **Theory: the fertility-human capital         **Empirical results**
+    relationship is...**                                      
+--- --------------------------------------------- ----------------------------------
+1.  Negative: $\frac{dN^{*}}{dh} < 0$.            Figures \@ref(fig:plot-means-over-time) 
+                                                  and \@ref(fig:plot-res-wt).
+
+2.  Weaker (closer to zero) at higher wages       Figure 
+    and/or levels of human capital.               \@ref(fig:plot-income-educ-level)a.
+                                                  Selection effects are also weaker at
+                                                  higher polygenic scores for 
+                                                  educational attainment 
+                                                  (Appendix Figure
+                                                  \@ref(fig:plot-purifying)).
+
+3.  More negative when the time burden            Stronger effects for single parents
+    of children $b$ is larger.                    (Figure \@ref(fig:plot-n-partners)).
+
+4.  Weaker at higher levels of education $s$.     Figure \@ref(fig:plot-income-educ-level)b.
+
+5.  Weaker among those who start fertility        Effects weaker among those starting 
+    in period 2 ($N_{1}^{*} = 0$) than among      fertility later 
+    those who start fertility in period 1         (Figure \@ref(fig:plot-age-flb)).
+    ($N_{1}^{*} > 0$).
+
+------------------------------------------------------------------------------------
+
+Table: (\#tab:theory) Predictions from the theoretical model and corresponding 
+empirical results.
+
 Thus, a simple economic model can explain many of our results.
@@
 
-------------------------------------------------------------------------------------
-    **Theory: the fertility-human capital           **Empirical results**
-    relationship is...**                                      
---- ----------------------------------------------- --------------------------------
-1.  Negative: $\frac{dN^{*}}{dh} < 0$.              Figures \@ref(fig:plot-means-over-time) 
-                                                    and \@ref(fig:plot-res-wt).
-
-2.  Weaker (closer to zero) at higher wages and/or  Figure \@ref(fig:plot-income-educ-level)a.
-    levels of human capital.                        Selection effects are also weaker at 
-                                                    higher polygenic scores for 
-                                                    educational attainment (Appendix Figure
-                                                    \@ref(fig:plot-purifying)).
-
-3.  More negative when the time burden              Stronger effects for single parents
-    of children, $b$, is larger.                    (Figure \@ref(fig:plot-n-partners)).
-
-4.  Weaker at higher levels of education ($s$).     Figure \@ref(fig:plot-income-educ-level)b.
-
-5.  Weaker among those who start fertility          Effects weaker among those starting 
-    in period 2 ($N_{1}^{*} = 0$) than among        fertility later 
-    those who start fertility in period 1           (Figure \@ref(fig:plot-age-flb)).
-    ($N_{1}^{*} > 0$).
-
-----------------------------------------------------------------------------------------
-
-Table: (\#tab:theory) Predictions from the theoretical model and corresponding 
-empirical results.
-
 ```{r plot-cors-earnings-educ, fig.cap="Selection effects by correlations with earnings and educational attainment. Each point represents one polygenic score."}
@@
 and fertility is mediated by educational attainment (Appendix Table
-\@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the indirect effect of
+\@ref(tab:tbl-res-mediation)). Figure \@ref(fig:plot-mediation) shows estimated
+proportions explained along with bootstrap 95% confidence intervals
+(uncorrected; 100 bootstraps). For `r n_pos_sig` scores, the indirect effect of
 the score on fertility *via* educational attainment takes the same sign as the
@@
 these scores, the median proportion of the total effect explained by the
-indirect effect is `r percent(median_prop)`. Figure \@ref(fig:plot-mediation)
-shows estimated proportions explained along with bootstrap 95% confidence
-intervals (uncorrected; 100 bootstraps). The educational attainment variable is
-a relatively crude measure of human capital: more accurate measures would likely
-explain more of the total effect.
+indirect effect is `r percent(median_prop)`. The educational attainment variable
+is a relatively crude measure of human capital: more accurate measures would
+likely explain more of the total effect.
 
@@
                         width = 0.3,
-                        color = "grey30"
+                        color = "grey50"
                       ) +
@@
 
-## Natural selection by sex
+## Selection effects by sex
 
@@
 stabilizing selection ($p < 0.05/33$, negative coefficient on quadratic term).
-The EA2 score for educational attainment shows significant disruptive
-selection ($p < 0.05/33$, positive coefficient), which reduces the strength of
-selection against educational attainment at very high levels of the PGS. Figure
-\@ref(fig:plot-purifying) plots predicted number of children against polygenic
-score from these regressions.
+The EA2 score for educational attainment shows significant disruptive selection
+($p < 0.05/33$, positive coefficient), which reduces the strength of selection
+against educational attainment at very high levels of the PGS. (The quadratic on
+the EA3 score has a similar coefficient but is not significant at $p < 0.05/33$.)
+Figure \@ref(fig:plot-purifying) plots predicted number of children
+against polygenic score from these regressions.
 
@@
 age/qualification weights multiplied by the inverse of number of siblings.
-Scores for EA2 and EA3 show significant disruptive selection ($p <
-0.05/33$, positive coefficient on quadratic). Other scores including 
-hip circumference were not significant.
+Scores for EA2 and EA3 show significant disruptive selection ($p < 0.05/33$, 
+positive coefficient on quadratic). Other scores including hip circumference 
+were not significant.
 
@@
 \clearpage
+## Number of partners and presence of partner by sex
+
+Figure \@ref(fig:plot-n-partners-sex) splits up Figure \@ref(fig:plot-n-partners)
+by sex. The pattern of results is the same in both sexes: selection effects are stronger among those with more lifetime sexual partners, and among those not
+currently living with a partner.
+
+```{r plot-n-partners-sex, fig.cap = "Selection effects by number of sexual partners and presence of a partner, for men and women separately.", fig.subcap = c("Lifetime number of sexual partners", "Presence of a partner"), fig.ncol = 1, fig.align = "center", fig.height = 4}
+
+sexes_theme <-  my_theme(
+                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
+                  legend.position      = "bottom", 
+                  #legend.box.spacing   = unit(0, "lines"),
+                  legend.box.margin    = margin(-5, 0, 0, -85),
+                  legend.margin        = margin(0, 0, 0, 0),
+                  legend.box.just      = "left",
+                  legend.justification = "left",
+                  legend.spacing.y     = unit(0, "lines"),
+                  #legend.spacing       = unit(0.5, "lines"),
+                  legend.title         = element_text(size = 10)
+                  #legend.text          = element_text(size = 8)
+                )
+sexes_guides <- guides(
+                  fill  = guide_legend(ncol = 1),
+                  color = guide_legend(ncol = 1)
+                )
+
+drake::loadd(res_partners)
+res_partners %<>%       
+      filter(grepl(":", term)) %>% 
+      mutate(`N partners` = ifelse(grepl("TRUE:", term), "3 or less", "4 or more"))
+
+n_regs <- as.double(nrow(res_partners))
+res_partners %>% 
+      mutate(
+        score_name = pretty_names(score_name)
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(
+        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        x      = estimate, 
+        fill   = `N partners`, 
+        colour = p < 0.05/{{n_regs}}
+      )) + 
+        my_vline +
+        confint_geom_segment() +
+        pval_geom_point() +
+        facet_wrap(vars(sex), scales = "free_y") +
+        my_fill_scale(n = 2) +
+        pval_scale() +
+        sexes_theme +
+        my_labs(x = "", y = "") + 
+        scale_y_reordered() +
+        scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
+        sexes_guides
+
+drake::loadd(res_with_partner_sex)
+res_with_partner_sex %<>% filter(grepl(":", term))
+
+n_regs <- as.double(nrow(res_with_partner_sex))
+
+res_with_partner_sex %>% 
+      mutate(
+        score_name = pretty_names(score_name),
+        Household = ifelse(grepl("TRUE", term), "With partner", 
+                      "Without partner")
+      ) %>% 
+      rename(p = p.value) %>% 
+      ggplot(aes(
+        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
+        x      = estimate, 
+        fill   = Household, 
+        colour = p < 0.05/{{n_regs}}
+      )) + 
+        my_vline +
+        confint_geom_segment() +
+        pval_geom_point() +
+        facet_wrap(vars(sex), scales = "free_y") +
+        my_fill_scale(n = 2) +
+        pval_scale() +
+        sexes_theme +
+        my_labs(x = "", y = "") + 
+        scale_y_reordered() +
+        scale_linetype_manual(
+                                name   = "", 
+                                values = c("95% c.i. uncorrected" = 1)
+                              ) +
+        sexes_guides
+
+```
+
+\FloatBarrier
+\clearpage
 ## Parents' generation {#sec-parents}
-### Selection effects
+### Selection effects and change over time
 
@@
 
-Tables \@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)
-summarize the results. We report the number of scores showing significant
+Table \@ref(tab:tbl-change-by-period) 
+summarizes the results. We report the number of scores showing significant
 changes over time (i.e. a significant interaction between polygenic score and
@@
 nine polygenic scores: ADHD, age at menopause, caffeine consumption, cognitive
-ability, coronary artery disease, EA2, EA3, extraversion and major depressive
-disorder. These changes are inconsistent with the intergenerational change,
+ability, Coronary Artery Disease, EA2, EA3, extraversion and Major Depressive
+Disorder. These changes are inconsistent with the intergenerational change,
 where estimated effect sizes were larger among the earlier, parents' generation.
@@
 
-```{r tbl-siblings-by-period}
-
-period_table <- function (tbl, caption) {
-  tbl %>% 
+```{r tbl-change-by-period}
+
+tbl_resp <- summary_period %>% 
+              group_by(Change_Respondents) %>% 
+              count()
+
+tbl_par <- summary_period %>% 
+             group_by(Change_Parents) %>% 
+             count()
+
+tbl_both <- full_join(tbl_par, tbl_resp, 
+                        by = c("Change_Parents" = "Change_Respondents"))
+
+tbl_both %>% 
+            rename(
+              Change              = Change_Parents, 
+              `Parents' RLRS`     = n.x, 
+              `Respondents' RLRS` = n.y
+            ) %>% 
             as_hux() %>%
@@
             set_left_padding(everywhere, 1, 0) %>% 
-      set_right_padding(everywhere, 2, 0) %>% 
-      set_caption(caption) %>% 
+            set_right_padding(everywhere, 3, 0) %>% 
+            set_caption("Numbers of polygenic scores showing changes in selection effects between early and late born. 
+      Weights for respondents: Age/Qualification. Weights for parents: Age/Qualification times 1/number of siblings.") %>% 
             set_caption_width(1)
-}
-
-summary_period %>% 
-      group_by(Change_Parents) %>% 
-      count() %>% 
-      period_table("Change in selection effects between parents of early and late born respondents (regressions on parents' RLRS). Weights: Age/Qualification times 1/number of siblings.")
-
-```
-
-```{r tbl-children-by-period}
-summary_period %>% 
-      group_by(Change_Respondents) %>% 
-      count() %>% 
-      period_table("Change in selection effects between early and late born respondents (regressions on RLRS). Weights: Age/Qualification.")
+
 ```

### 2022-03-15 _drake.R

@@ -439,6 +439,20 @@ plan <- drake_plan(
           select(-row_number, -subset)
   },
   
+  res_partners_joint = {
+    res <- map_dfr(score_names,
+                     ~ run_regs_fml(
+                       fml        =
+                         "RLRS ~ lo_partners + {score_name}:lo_partners",
+                       score_name = .x,
+                       famhist    = famhist_kids
+                     ),
+                     .id = "score_name"
+                   )
+    res %<>% filter(term != "(Intercept)")
+    res
+  },
+  
   res_with_partner = {
     res <- map_dfr(score_names,
             ~run_regs_fml(

### 2022-03-19 why-natural-selection.Rmd
Work on graphs. Use medians not means for percentage differences.

@@
 The direction of natural selection is reversed among older parents, or after controlling for age at first live birth.
-These patterns are in line with economic theories of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market.
+These patterns are in line with the economic theory of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market.
 Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
-bibliography: "negative-selection.bib"
+bibliography: "why-natural-selection.bib"
 output: 
@@
 
+  if (conf_int) {
+    confint_scale <- scale_linetype_manual(
+                       name   = "", 
+                       values = c("95% c.i. uncorrected" = 1),
+                       guide  = guide_legend(order = 1)
+                     )
+    confint_geom_segment <- confint_geom_segment()
+  } else {
+    confint_scale <- NULL
+    confint_geom_segment <- NULL
+  }
+
   n_regs <- as.double(n_regs)
-
-  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL
 
@@
     ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
-                     fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
+               shape = {{fill_col}}, ...)) +
     my_vline + 
     confint_geom_segment +
-          pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
-          pval_scale() +
-          my_fill_scale(n = n_cats, direction = fill_direction) +
+    geom_point(size = 1.8, alpha = 0.9, aes(color = {{fill_col}})) +
+    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/{{n_regs}})) +
+    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction) +
+    scale_shape_manual(values = 1:n_cats + 14) +
     my_labs() +
     my_theme() +
-          scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1))
-
+    confint_scale +
+    scale_alpha_discrete(range = c(1, 0), 
+                           guide = guide_legend(
+                             order = 0,
+                             override.aes = list(
+                               shape = c(1, 16),
+                               color = "black",
+                               alpha = 1,
+                               size = 1.8
+                             )
+                         )) 
 }
@@
 
-pval_geom_point <- function (size = 1.6, stroke = 0.8, fill = NULL, ...) {
+pval_geom_point <- function (size = 1.6, stroke = 0.9, fill = NULL, ...) {
   if (is.null(fill)) {
@@
   geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i. uncorrected"), 
-                    color = "grey60", alpha = 0.25, size = 1.1, 
+                    color = "grey55", alpha = 0.25, size = 1.1, 
                     lineend = "round")
@@
 
-  if (n <= 2) {
+  scale <- if (n <= 2) {
     dots <- list(...)
@@
   } else {
-    scale_color_brewer(aesthetics = aesthetics, ...)
+    scale <- scale_color_brewer(aesthetics = aesthetics, ...)
+    shades::lightness(scale, shades::scalefac(0.75))
   }
+
+  scale
 }
@@
 
-add_n <- function (var, famhist_var = var, dv = "RLRS", data = famhist) {
+add_n <- function (var, famhist_var = var, dv = "RLRS", data = famhist, 
+                     reverse = FALSE) {
   var <- as.factor(var)
   n <- n_in_regs(famhist_var, dv, data)
+  # this is a horrible hack which just happens to work
+  # check all Ns manually!
+  if (reverse) n <- rev(n)
   levels(var) <- sprintf("%s (N = %s)", levels(var), 
@@
 
-As yet we know little about the social mechanisms behind natural selection. A
-potential explanation comes from the economic theory of fertility [@becker1960economic]. In this theory, higher potential earnings
-have two opposite effects on fertility: a fertility-increasing *income effect*
-(higher income makes children more affordable), and a fertility-lowering
-*substitution effect* (time spent on childrearing has a higher cost in foregone
-earnings). Thus, an individual's *human capital* -- skills and personality
-traits which are valuable in labour markets -- can increase or decrease his or
-her fertility. This can lead to selection on inherited variants which are linked
-to human capital. Furthermore, the economic theory predicts systematic variation
-in the relative strength of the income and substitution effects across different
-social groups.
+As yet we know little about the social mechanisms behind natural selection. The
+economic theory of fertility [@becker1960economic] offers a potential
+explanation. Higher potential earnings have two opposite effects on fertility: a
+fertility-increasing *income effect* (higher income makes children more
+affordable), and a fertility-lowering *substitution effect* (time spent on
+childrearing has a higher cost in foregone earnings). Thus, an individual's
+*human capital* -- skills and personality traits which are valuable in labour
+markets -- can increase or decrease their fertility. Genetic variants  which are 
+linked to human capital will then be selected for or against. Also, the economic
+theory predicts that the relative strength of income and substitution effects
+will vary systematically across different social groups.
 
@@
 We then show that a simple model of human capital, education and fertility
-choices can give rise to these empirical results. To test the theory, we show
-that polygenic scores' correlation with fertility is associated with their
-correlation with education and earnings. Furthermore, a mediation analysis shows
-that part of the correlation with fertility is indeed *via* education.
-Thus, contemporary natural selection on polygenic scores can be explained by scores' correlation with earnings-increasing human capital.
-
-Lastly, we discuss the effects of natural selection. While measured effects on
-mean genetic scores are small, natural selection may substantially increase the
-correlation between genetic scores and income, increasing genetic differences 
-between different social groups, and thus making the "genetic lottery"
-[@harden2021genetic] more unfair.
+choices can give rise to these empirical results. At higher incomes, the income
+and substitution effects are balanced, while among lower-income people, or
+single parents who face a bigger time burden from childcare, the substitution
+effect dominates. The theory predicts that polygenic scores' correlation
+with fertility is associated with their correlation with education and earnings,
+and we confirm this. We then run a mediation analysis, which shows that part of the
+correlation with fertility is indeed mediated by educational attainment. 
+Thus, contemporary natural selection on polygenic scores can be explained 
+by scores' correlation with earnings-increasing human capital.
+
+Lastly, we discuss the effects of natural selection. While our estimated effects 
+on measured polygenic scores are small, natural selection substantially 
+increases the correlation between polygenic scores and income, increasing 
+genetic differences between different social groups, and thus making the 
+"genetic lottery" [@harden2021genetic] more unfair.
 
@@
 changes could reflect natural selection within the UK population, but also
-emigration, or ascertainment bias within the sample [@10.1093/aje/kwx246].
-
-```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank.
-Points are means for 5-year intervals.
-Lines are 95% confidence intervals.
-Circles show a significant linear increase (green) or decrease (red) over time (p < 0.05/33)."}
+emigration, or ascertainment bias in the sample [@10.1093/aje/kwx246].
+
+```{r plot-means-over-time, fig.height = 7, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank.
+Symbols show means for 5-year intervals.
+Bars are 95% confidence intervals.
+Triangles denote a significant linear increase or decrease over time (p < 0.05/33)."}
 
@@
       ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
-      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
-        fatten = 1.2) +
-      facet_wrap(~score_name, ncol = 6) +
+      geom_linerange(aes(ymin = score_lo, ymax = score_hi), color = "grey70", 
+                       size = 3) +
+      geom_point(aes(y = score), size = 1.5) +
+      facet_wrap(~score_name, ncol = 5) +
       my_hline +
-      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
-      scale_shape_manual(values = c("-" = "circle", "+" = "circle", "o" = "square")) +
+      scale_color_manual(values = c("-" = "darkred", "+" = "darkgreen", "o" = "black")) +
+      scale_shape_manual(values = c(
+                                      "-" = "triangle down open", 
+                                      "+" = "triangle open", 
+                                      "o" = "circle open"
+                                    )) +
       theme(
@@
 
-To test more directly for natural selection, we regress respondents' relative 
-lifetime reproductive success ($RLRS_i$) on their polygenic scores (PGS):
+To test for natural selection more directly, we regress respondents' relative 
+lifetime reproductive success (RLRS) on each polygenic score (PGS):
 
@@
 
-RLRS is defined as respondents' number of children, divided by the mean number
+RLRS is defined as respondent $i$'s number of children, divided by the mean number
 of children of people born in the same year. The "selection effect", $\beta$,
@@
 of the causal effect of a polygenic score on fertility. However, natural
-selection is a correlational phenomenon: polygenic scores which correlate with
-high fertility are being selected for, whatever the underlying causal
-relationship.
+selection is a matter of correlation not causation: polygenic scores which 
+correlate with high fertility are being selected for, whatever the underlying 
+causal mechanism.
 
@@
 drake::loadd(res_wt_age_qual_weights)
+drake::loadd(res_wt_van_alten_weights)
 drake::loadd(res_sibs_parent_weights)
-drake::loadd(res_all)
+drake::loadd(res_unweighted)
 drake::loadd(res_sex)
@@
 res_wt_combined <- bind_rows(
-  None                = res_all %>% 
-                          filter(
-                            reg.type == "controlled", 
-                            dep.var == "RLRS"
-                          ) %>% 
-                          select(-reg.type, -dep.var, -score_name),
+  None                = res_unweighted %>% 
+                          select(-score_name) %>% 
+                          filter(term != "(Intercept)"),
+  Main                = res_wt_van_alten_weights,
   Geographical        = res_wt_msoa_weights,
@@
 
-res_significance <- res_wt_combined %>% 
-                      group_by(term) %>% 
-                      summarize(
-                        all = all(p.value < 0.05/33), 
-                        any = any(p.value < 0.05/33)
-                      )
-res_sig_all <- sum(res_significance$all)
-res_sig_any <- sum(res_significance$any)
+n_sig <- sum(res_wt_van_alten_weights$p.value < 0.05/33)
 
@@
                   `Age/Qualification` = `Age/Qualification`/None,
+                  Main                = Main/None,
                   `Age/Qual/AFLB`     = `Age/Qual/AFLB`/estimate_females
@@
 
-To correct for ascertainment bias, we weight participants using population data.
-We try three alternative weighting schemes: (1) "Geographical" weighting, by
-geography, age and presence/absence of a partner; (2) age and highest
-educational qualification; and (3) for women only, age, highest qualification,
-and age at first live birth (AFLB). Figure \@ref(fig:plot-res-wt) plots
-selection effects among the entire sample, estimated with the three weighting
-schemes, and also unweighted. Effects for `r res_sig_all` out of 33 polygenic
-scores are significant at $p < 0.05/33$ under all four weighting schemes, with
-`r res_sig_any` scores reaching significance under at least one scheme.[^balance-diverse-def]
-Weighting makes a large difference to estimates: mean effect sizes across all
-polygenic scores are increased by a factor of `r mean_rel$Geographical`
-(geographical weighting), `r mean_rel[["Age/Qualification"]]`
-(age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
-Estimates might be further affected by weighting on other demographic
-variables.[^weighting]
+Figure \@ref(fig:plot-res-wt) plots selection effects in the whole
+sample.[^balance-diverse-def] To correct for ascertainment bias, we use
+participant weights from @van2021effects, which match the UK Biobank eligible
+population on sex, birth year, location, education, employment, health,
+household size and tenure, number of cars and age at death. Weighting makes a
+large difference: effect sizes go up by a mean of 
+`r scales::percent(mean_rel$Main - 1)`.[^weighting] `r n_sig` out of 33 weighted
+selection effects are significant at $p$ < 0.05/33.
 
@@
 
-[^weighting]: To make interpretation straightforward, we do not use weights in 
-results below unless noted. However, all our qualitative results are robust to
-using weights.
+[^weighting]: We use these weights throughout. All our qualitative results are
+robust if we run unweighted regressions. Appendix Table \@ref(tab:tbl-res-wt)
+shows results from alternative weighting schemes.
 
@@
 res_wt_combined %>% 
-      mutate(
-        Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
-                    "Age/Qual/AFLB", "None"),
+      filter(Weights %in% c("Main", "None")) %>% 
+      mutate(
+        Weights  = recode(Weights, "Main" = "Weighted", "None" = "Unweighted"),
         term     = pretty_names(term),
@@
       ) %>% 
-      rename(p = p.value) %>% 
-      # can't use standard_ggplot because we have non-standard geom_point()
-      ggplot(aes(
-        x      = estimate, 
-        y      = term, 
-        yend   = term,
-        shape  = Weights, 
-        fill   = Weights, 
-        colour = p < 0.05/{{n_regs}}
-      )) +
-      my_vline +
-      geom_point(size = 1.6, stroke = 0.8, alpha = 0.9) +
-      confint_geom_segment() +
-      pval_scale() +
-      scale_fill_manual(values = c(
-        "None"              = "grey15", 
-        "Geographical"      = "dodgerblue4",
-        "Age/Qualification" = "steelblue2",
-        "Age/Qual/AFLB"      = "darkseagreen4"
-      )) +
-      scale_shape_manual(values = c(
-        "None" = "square filled", 
-        "Geographical" = "circle filled",
-        "Age/Qualification" = "circle filled", 
-        "Age/Qual/AFLB" = "circle filled"
-      )) + 
-      labs(x = "Effect size", y = "") +
-      scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
-      my_theme()
-
-```
-
-To motivate our economic model, we split the sample by demographic
-and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
-We examine socio-demographic variables including income and education, and
+      standard_ggplot(fill_col = Weights, n_regs = n_regs)
+
+```
+
+We now show the empirical puzzles which motivate our economic model. Each
+concerns differences in the strength of natural selection across different
+subgroups in the sample. We re-estimate \@ref(eq:regression) splitting the
+sample by demographic and social variables, including income and education, and
 family structure variables including age at first live birth, presence of a
-partner, and lifetime number of sexual partners. Figure
-\@ref(fig:plot-income-educ-level) plots selection effects for each polygenic
-score, grouping respondents by age of completing full-time education, and by
-household income. Effects are larger and more significant for the lowest income
-category, and for the lowest education category. These results are robust to
-controlling for respondents' age (Appendix section \@ref(sec:age-control)).
+partner, and lifetime number of sexual partners.
+
+```{r calc-pct-diffs}
+drake::loadd(res_edu)
+drake::loadd(res_income)
+drake::loadd(res_partners_joint)
+drake::loadd(res_with_partner)
+res_partners_joint %<>% filter(grepl(":", term))
+res_with_partner %<>% filter(grepl(":", term))
+
+res_edu_wide <- full_join(
+                  res_edu %>% filter(age_fte_cat == "< 16"),
+                  res_edu %>% filter(age_fte_cat == "> 18"),
+                  by = "score_name"
+                )
+
+res_income_wide <- full_join(
+                     res_income %>% filter(income_cat == 1),
+                     res_income %>% filter(income_cat == 5),
+                     by = "score_name"
+                   )
+
+res_partners_wide <- full_join(
+                       res_partners_joint %>% filter(grepl("lo_partnersFALSE", term)),
+                       res_partners_joint %>% filter(grepl("lo_partnersTRUE", term)),
+                       by = "score_name"
+                     )
+
+res_with_partner_wide <- full_join(
+                           res_with_partner %>% 
+                             filter(grepl("with_partnerFALSE", term)),
+                           res_with_partner %>% 
+                             filter(grepl("with_partnerTRUE", term)),
+                           by = "score_name"
+                         )
+
+median_pct_diff <- function (dfr) {
+   prop <- dfr %>% filter(
+                    p.value.x < 0.05/33,
+                    sign(estimate.x) == sign(estimate.y)
+                  ) %>% 
+                  mutate(
+                    pct_diff = estimate.x/estimate.y
+                  ) %>% 
+                  pull(pct_diff) %>% 
+                  mean()
+   # percentage difference:
+   prop - 1
+}
+
+pct_diff_edu <- median_pct_diff(res_edu_wide)
+pct_diff_income <- median_pct_diff(res_income_wide)
+pct_diff_n_partners <- median_pct_diff(res_partners_wide)
+pct_diff_with_partner <- median_pct_diff(res_with_partner_wide)
+```
+
+Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
+polygenic score, grouping respondents by age of completing full-time education,
+and by household income. Effects are larger and more significant for the lowest
+education category, and for the lowest income category. The median percentage
+difference between the lowest and highest education categories, among scores
+which are significant for the lowest category and have the same sign across
+categories, is `r scales::percent(pct_diff_edu)`. Between the lowest and
+highest income categories, it is `r scales::percent(pct_diff_income)`. These
+results are robust to controlling for respondents' age (Appendix section
+\@ref(sec:age-control)). Turning to family structure, we split respondents by
+lifetime number of sexual partners, at the median value of 3 (Figure
+\@ref(fig:plot-n-partners)a). Now, selection effects are larger and more
+significant among those with more than 3 lifetime partners, with a median percentage
+difference of `r scales::percent(pct_diff_n_partners)`. Next we split
+respondents by whether they were living with a spouse or partner at the time of
+interview (Figure \@ref(fig:plot-n-partners)b).  Effects are larger among those not living with a spouse or partner. The median percentage difference is 
+`r scales::percent(pct_diff_with_partner)`.[^sex]
+
+[^sex]: The same pattern holds if we analyse men and women separately (Appendix Figure
+\@ref(fig:plot-n-partners-sex)). We also directly compared selection effects 
+between men and women (Appendix Figure \@ref(fig:plot-sexes)).
 
@@
 
-Turning to family structure, we split respondents by lifetime number of sexual
-partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
-Selection effects are larger and more significant among
-those with more than 3 lifetime partners. Next we split respondents by whether
-they were living with a spouse or partner at the time of interview.
-Effects are larger among those not living with a spouse or partner
-(Figure \@ref(fig:plot-n-partners)b).[^sex]
-
-[^sex]: The same pattern holds if we analyse men and women separately (Appendix Figure
-\@ref(fig:plot-n-partners-sex)). We also directly compared selection effects 
-between men and women (Appendix Figure \@ref(fig:plot-sexes)).
-
 ```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and presence of a partner.", fig.subcap = c("Lifetime number of sexual partners", "Presence of a partner"), fig.ncol = 1, fig.align = "center", fig.height = 4}
@@
 res_partners_joint %>% 
+      mutate(
+        `N partners` = add_n(`N partners`, "n_partners_split", reverse = TRUE,
+                               data = famhist %>%
+                               mutate(n_partners_split = n_partners <= 3))
+      ) %>% 
       standard_ggplot(fill_col = `N partners`, n_regs = n_regs, 
@@
         Household = ifelse(grepl("TRUE", term), "With partner", 
-                      "Without partner")
+                      "Without partner"),
+        Household = add_n(Household, "with_partner", reverse = TRUE)
       ) %>% 
@@
       pivot_wider(score_name, names_from = category, values_from = estimate) %>%              {cor(.$`age_flb_cat10-22`, .$`age_flb_cat28-52`)}
+
+```
+
+```{r calc-corr-age-flb}
+
+drake::loadd(res_age_flb)
+drake::loadd(res_sex)
+
+res_sex_comparison <- res_sex %>% 
+      filter(sex == "Female") %>% 
+      mutate(score_name = term) %>% 
+      select(score_name, term, estimate:conf.high)
+
+res_age_flb %<>% filter(term != "age_flb")
+
+res_combined <- bind_rows(
+        raw      = res_sex_comparison, 
+        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
+        .id      = "reg.type"
+      ) %>% 
+      arrange(score_name)
+
+raw_flb_comparison <- res_combined %>% 
+      pivot_wider(
+        id_cols     = score_name, 
+        names_from  = reg.type, 
+        values_from = estimate
+      )
+
+cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
+opp_signed <- sum(
+        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
+      )
 
@@
 only among older mothers. The correlation between effect sizes for the youngest
-and oldest terciles is `r cor_age_flb`.
+and oldest terciles is `r cor_age_flb`. To investigate this further, we estimate
+equation \@ref(eq:regression) among females, *controlling* for AFLB. In 
+`r opp_signed` out of `r length(score_names)` cases, effects change sign when
+controls are added. The correlation between effect sizes controlling for AFLB,
+and raw effect sizes, is `r cor_raw_flb`. Thus, selection effects seem to come
+through two opposing channels: a correlation with AFLB, and an opposite-signed
+correlation with number of children after AFLB is controlled for.
 
 [^aflb-no-men]: AFLB is unavailable for men.
-
-```{r calc-corr-age-flb}
-
-drake::loadd(res_age_flb)
-drake::loadd(res_sex)
-
-res_sex_comparison <- res_sex %>% 
-      filter(sex == "Female") %>% 
-      mutate(score_name = term) %>% 
-      select(score_name, term, estimate:conf.high)
-
-res_age_flb %<>% filter(term != "age_flb")
-
-res_combined <- bind_rows(
-        raw      = res_sex_comparison, 
-        with_age = res_age_flb %>% select(score_name, term, estimate:conf.high), 
-        .id      = "reg.type"
-      ) %>% 
-      arrange(score_name)
-
-raw_flb_comparison <- res_combined %>% 
-      pivot_wider(
-        id_cols     = score_name, 
-        names_from  = reg.type, 
-        values_from = estimate
-      )
-
-cor_raw_flb <- cor(raw_flb_comparison$raw, raw_flb_comparison$with_age)
-opp_signed <- sum(
-        sign(raw_flb_comparison$raw) != sign(raw_flb_comparison$with_age)
-      )
-
-```
-
-To investigate this further, we estimate equation \@ref(eq:regression) among
-females, *controlling* for AFLB. In `r opp_signed` out of 
-`r length(score_names)` cases, effects change sign when controls are added. The
-correlation between effect sizes controlling for AFLB, and raw effect sizes, is
-`r cor_raw_flb`. Thus, selection effects seem to come through two opposing
-channels: a correlation with AFLB, and an opposite-signed correlation with 
-number of children after AFLB is controlled for.
 
@@
 
-We emphasize that our categories are not necessarily exogenous to the polygenic
-scores. For example, in the data, and also in our theoretical model, age at first
-live birth is a choice variable, which is endogenous to human-capital-related
-polygenic scores (Appendix Figure \@ref(fig:plot-age-flb-dv)). Nevertheless,
-differences in equation \@ref(eq:regression) across subgroups constrain the set
-of possible explanations for natural selection. As we show below, a simple
-economic model predicts the differences we find.
+We emphasize that these categories are not exogenous to polygenic scores. For
+example -- both in the data (Appendix Figure \@ref(fig:plot-age-flb-dv)) and in
+our theoretical model -- education and age at first live birth are choice
+variables, which are endogenous to a person's human capital and to relevant
+polygenic scores. Nevertheless, differences in selection effects across
+subgroups constrain the set of possible explanations. A good theory of contemporary
+natural selection needs to show how these differences come about. As we describe
+below, a model based on the economic theory of fertility can do just that.
 
@@
 There is some evidence that selection effect sizes are increasing in the 
-respondents' generation, with 9 polygenic scores showing a significant increase. 
+respondents' generation, with 8 polygenic scores showing a significant increase. 
 We also check whether selection effects vary by
-socio-economic status and AFLB in the parents' generation. We have no
-information about parents' income, so we use the 1971 Townsend deprivation score
-of respondents' birthplace as a proxy [@townsend1987deprivation]. Results show
-the same pattern as for respondents. Effect sizes are larger and more often
-significant in the most deprived areas (Appendix Figure
-\@ref(fig:plot-siblings-townsend)). Effects are larger among younger parents,
-and change sign when controlling for AFLB, among both mothers and fathers
-(Appendix Figures \@ref(fig:plot-age-birth-parents-cross), 
-\@ref(fig:plot-age-birth-parents)). Lastly, we check for a
-"quantity-quality tradeoff" between parents' number of children and number of
-grandchildren. We do not find any: in fact, the correlation between respondents' 
-and parents' RLRS is positive ($\rho$ = `r sib_chn_cor$estimate`, 
-$p < 2 \times 10^{-16}$).
+AFLB and socio-economic status in the parents' generation, using the 1971 
+Townsend deprivation score of respondents' birthplace as a proxy for income
+[@townsend1987deprivation]. Results show the same pattern as for the
+respondents' generation. Effect sizes are larger and more often significant in
+the most deprived areas (Appendix Figure \@ref(fig:plot-siblings-townsend)).
+Effects are larger among younger fathers and mothers, and change sign when
+controlling for AFLB (Appendix Figures \@ref(fig:plot-age-birth-parents-cross),
+\@ref(fig:plot-age-birth-parents)). Lastly, we check for a "quantity-quality
+tradeoff" between parents' number of children and number of grandchildren. We
+don't find any: in fact, the correlation between respondents' and parents' RLRS
+is positive ($\rho$ = `r sib_chn_cor$estimate`, $p < 2 \times 10^{-16}$).
 
@@
 
-To show this, we write down a simple model of fertility choices. $h$ is
+To show this, consider a simple model of fertility choices. $h$ is
 an individual's level of human capital. For now, we simply identify this with
@@
 from the number of children $N$ and from income $Y\equiv(1-bN)W$:
-
 \[
@@
 \]
-
 Here $a$ captures the strength of preference for children. $u(\cdot)$ captures
@@
 in line with the literature: this can be thought of as the expected number of
-children among similar people. The marginal benefit of an extra child is
-$\frac{dU}{dN} = -bWu'(Y)+a$. The effect of an increase in human capital on this
-marginal benefit is
-
+children among people with a given $a$, $b$ and $W$. The marginal
+benefit of an extra child is $\frac{dU}{dN} = -bWu'(Y)+a$. The effect of an
+increase in human capital on this marginal benefit is
 \[
-\frac{d^{2}U}{dNdh}=\underbrace{-bu'(Y)}_{\textrm{Substitution effect}}\underbrace{-bYu''(Y)}_{\textrm{Income effect}}.
+\frac{d^{2}U}{dNdW}=\underbrace{-bu'(Y)}_{\textrm{Substitution effect}}\underbrace{-bYu''(Y)}_{\textrm{Income effect}}.
 \]
-
 The *substitution effect* is negative and reflects that when wages increase,
@@
 that credit markets are imperfect so that agents cannot borrow. Write
-
 \begin{equation}
@@
 \end{equation}
-
-Instead of identifying human capital with wages, we now allow
-individuals to choose a level of education $s \in [0,1]$, which has
-a time cost in period 1. Education is
-complementary to human capital $h > 0$, and increases period 2 wages, which take 
-the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
-to 1, and let $u(\cdot)$ take the constant relative risk aversion form
-$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$. $\sigma > 0$ measures the curvature
-of the utility function, i.e. the decline in marginal utility of income as income
-increases. We examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
-*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$ in
-a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical
-predictions, along with our corresponding empirical results for the correlation
-between polygenic scores and RLRS.[^one-period-results] The key insight of
-the model is that for middling levels of $\sigma$, the substitution effect
-dominates at low income levels, but is balanced by the income effect as income
-increases.
+Instead of identifying human capital with wages, we now allow individuals to
+spend time $s \in [0,1]$ on education in period 1.
+Education is complementary to human capital $h > 0$, and increases period 2
+wages, which take the simple functional form $w(s,h) = sh$. We normalize period
+1 wages to 1, and let $u(\cdot)$ take the constant relative risk aversion form
+$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$. $\sigma > 0$ measures the
+curvature of the utility function, i.e. the decline in marginal utility of
+income as income increases. We examine total fertility 
+$N^{*} = N_{1}^{*} + N_{2}^{*}$ and the *fertility-human capital relationship*,
+$\frac{dN^{*}}{dh}$. For $\sigma < 1$ and close enough to 1, Table
+\@ref(tab:theory) shows five theoretical predictions, along with our
+corresponding empirical results for the correlation between polygenic scores and
+RLRS.[^one-period-results] The key insight of the model is that for middling
+levels of $\sigma$, the substitution effect dominates at low income levels, but
+as income increases, the income and substitution effect balance out.
 
@@
 
-Thus, a simple economic model can explain many of our results.
-Other empirical work in economics also supports our mechanism.
+Thus, a simple economic model can explain many of our results. Other empirical
+work in economics also supports the link from human capital to fertility.
 @caucutt2002women and @monstad2008education show that education and skills
@@
 US fertility decreases faster with education among single mothers than married
-mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
-@becker1981treatise. There is also a literature showing negative correlations
-between IQ and fertility [e.g. @lynn2004new; @reeve2018systematic].
+mothers [@baudin2015fertility], in line with our prediction 3 and as predicted
+by @becker1981treatise. A related literature shows negative correlations between 
+IQ and fertility [e.g. @lynn2004new; @reeve2018systematic].
 
@@
 
-We test this explanation in two ways. First, the economic theory predicts that
+We test the economic theory in two ways. First, it predicts that
 genetic variants will be selected for (or against) in proportion to their
@@
 control for age at first live birth, suggesting that within AFLB categories, the
-income effect dominates. (See also the Supplementary Animations for the
-uncorrected effects per AFLB group, which show the correlation reversing after
-age 22.)
+income effect dominates.
 
@@
                      # 0.025 not 0.05 because two-sided
-                     sig     = abs(statistic_ind) > qnorm(1 - 0.025/33),
+                     sig     = abs(statistic_ind) > qnorm(1 - 0.025/
+                                                            nrow(res_mediation)),
                      pos_sig = sig & prop_ind > 0
@@
 
-Second, for the `r nrow(res_mediation)` scores where the bivariate correlation in
-\eqref{eq:regression} is significant at $p < 0.05/33$, we run a mediation
-analysis to directly test whether the correlation between each polygenic score
-and fertility is mediated by educational attainment (Appendix Table
-\@ref(tab:tbl-res-mediation)). Figure \@ref(fig:plot-mediation) shows estimated
-proportions explained along with bootstrap 95% confidence intervals
-(uncorrected; 100 bootstraps). For `r n_pos_sig` scores, the indirect effect of
-the score on fertility *via* educational attainment takes the same sign as the
-overall effect, and is significantly different from zero ($p < 0.05/33$). Among
-these scores, the median proportion of the total effect explained by the
-indirect effect is `r percent(median_prop)`. The educational attainment variable
-is a relatively crude measure of human capital: more accurate measures would
-likely explain more of the total effect.
-
-```{r plot-mediation, fig.align = "center", fig.cap = "Proportion of selection effect mediated by educational attainment, polygenic scores with significant selection effects"}
+Second, we run a mediation analysis to directly test whether the correlation
+between each polygenic score and fertility is mediated by educational attainment
+(Appendix Table \@ref(tab:tbl-res-mediation)). We use the 
+`r nrow(res_mediation)` scores where the selection effect is significant at $p$ <
+0.05/33. Figure \@ref(fig:plot-mediation) shows estimated
+proportions explained by educational attainment, along with bootstrap 95%
+confidence intervals (uncorrected; 100 bootstraps). For `r n_pos_sig` scores,
+the indirect effect of the score on fertility *via* educational attainment takes
+the same sign as the overall effect, and is significantly different from zero
+($p$ < 0.05/`r nrow(res_mediation)`). Among these scores, the median proportion of
+the total effect explained by the indirect effect is 
+`r scales::percent(median_prop)`. The educational attainment variable is a
+relatively crude measure of human capital: more accurate measures would likely
+explain more of the total effect.
+
+```{r plot-mediation, fig.align = "center", fig.cap = "Proportion of selection effect mediated by educational attainment, among polygenic scores with significant selection effects"}
 
@@
         term = fct_reorder(term, prop_ind, order_abs(1)),
-        `Indirect effect p < 0.05/33` = sig 
+        `Indirect effect p < 0.05/23` = sig 
       ) %>% 
       ggplot(aes(y = term)) + 
-        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/33`), 
+        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/23`), 
                    width = .5) + 
@@
 welfare benefits which incentivize child-bearing might be taken up more among
-low-income individuals. However, the majority of effect sizes appear unchanged
-over a large span of twentieth-century history (Appendix Tables
-\@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)), during
+low-income people. However, the majority of effect sizes appear unchanged
+over a large span of twentieth-century history (Appendix Table
+\@ref(tab:tbl-change-by-period)), during
 which government spending on child-related benefits varied considerably
 [@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
-affect fertility [@gauthier2007impact]. Future
+affect fertility [@gauthier2007impact; see also @bergsvik2021can]. Future
 work could test this theory more explicitly. A second alternative theory is that
@@
 $a$ in the model [cf. @jones2008fertility]. This theory would not explain why
-selection effects are smaller at higher incomes and education levels. Indeed,
-in the model, the effect of $a$ on fertility becomes stronger at higher levels
+selection effects are smaller at higher incomes and education levels. In fact,
+in the model, $a$'s effect on fertility gets stronger at higher levels
 of human capital. A third alternative is that traits under selection are linked
@@
 our parameter $\sigma$, which can be interpreted as a measure of risk aversion
-over income. However, a more direct channel is risky sexual behaviour
-[@mills2021identification]. The data here provide some support for this: scores
+over income; a more direct channel is risky sexual behaviour
+[@mills2021identification]. The data here provide some support for this story: scores
 which might plausibly be linked to externalizing behaviour, like ADHD and
-younger age at smoking initiation, are selected for. However, this theory is
-less good at explaining variation in selection across the full range of scores,
-including physical measures such as waist-hip ratio and BMI. We test this theory
-directly by re-estimating equation \@ref(eq:regression) controlling for a
-measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
-sizes between regressions with and without controls is `r median(ratio)`; all
-scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
-so when controlling for risk attitude. This non-result could simply reflect the
-imprecision of the risk attitude measure, which is a single yes/no question.
-However, this measure *does* predict the overall number of children, highly
-significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of 33 regressions).
-Given that, and the statistical power we get from our sample size, we believe
-that the non-result is real: while risk attitude does predict fertility in the
-sample, it is not an important channel for natural selection.
+younger age at smoking initiation, are selected for. However, risk-seeking 
+seems unlikely to explain variation in fertility across the full range of scores
+under selection, including physical measures like waist-hip ratio and BMI. We
+test this theory directly by re-estimating equation \@ref(eq:regression)
+controlling for a measure of risk attitude (UK Biobank field 2040). The median
+ratio of effect sizes between regressions with and without controls is 
+`r median(ratio)`; all scores which are significant at $p < 0.05/33$ in
+uncontrolled regressions remain so when controlling for risk attitude. This
+non-result could simply reflect the imprecision of the risk attitude measure,
+which is a single yes/no question. But this measure *does* predict the overall
+number of children, highly significantly ($p < 2 \times 10^{-16}$ in 
+`r n_risk_sig` out of 33 regressions). Given that, and the statistical power we
+get from our sample size, we believe that the non-result is real: while risk
+attitude does predict fertility in the sample, it is not an important channel
+for natural selection.
 
@@
 correlates with that score's association with labour market earnings. Strikingly,
-some of these results were predicted by @fisher1930genetical, pp. 253-254.
-
-The economic theory of fertility provides a parsimonious explanation for these
+some of these results were predicted by @fisher1930genetical, pp. 253-254. The
+economic theory of fertility gives a parsimonious explanation for these
 findings. Because of the substitution effect of earnings on fertility, scores
@@
 
-Polygenic scores which correlate with high (low) earnings and more (less)
-education are being selected against (selected for). In addition, many of the
-phenotypes under positive selection are linked to disease risk. Many people
-would probably prefer to have high educational attainment, a low risk of ADHD
-and major depressive disorder, and a low risk of coronary artery disease, but
-natural selection is pushing against genes associated with these traits.
-Potentially, this could increase the health burden on modern populations, but
-that depends on effect sizes. Our results suggest that naïve estimates can be
-affected by sample ascertainment bias. There may be remaining sources of
-ascertainment bias after our weighting; if so, we expect that, like the sources
-of ascertainment we have controlled for, they probably bias our results towards
-zero. This problem may be less serious in surveys which aim to be representative
-(as the UK Biobank does not). However, there is still scope for bias, since not
-all respondents consent to the collection of genetic data. For instance,
-completion rates for genotype data in the US Health and Retirement Study were
-around 80-85% [@hrs2020]. Researchers should be aware of the risks of
+Polygenic scores which correlate with lower earnings and less education are
+being selected for. In addition, many of the phenotypes under positive selection
+are linked to disease risk. Many people would probably prefer to have high
+educational attainment, a low risk of ADHD and major depressive disorder, and a
+low risk of coronary artery disease, but natural selection is pushing against
+genes associated with these traits. Potentially, this could increase the health
+burden on modern populations, but that depends on effect sizes. Our results show
+that naïve estimates can be affected by sample ascertainment bias. There may be
+remaining sources of ascertainment bias after our weighting; if so, we expect
+that, like the sources of ascertainment we have controlled for, they probably
+bias our results towards zero. Researchers should be aware of the risks of
 ascertainment when studying modern natural selection.
@@
 We also do not know how estimated effect sizes of natural selection will change
-as more accurate measures of genetic variation are produced, or whether genetic
+as more accurate polygenic scores are produced, or whether genetic
 variants underlying other phenotypes will show a similar pattern to those
-studied here. In addition, effects of polygenic scores may be inflated in
+studied here. Also, effects of polygenic scores may be inflated in
 population-based samples, because of indirect genetic effects, gene-environment
@@
 @selzam2019comparing; @kong2018nature; @Howe2021.03.05.433935], although we do
-not expect that this should impact their association with number of offspring,
-or the resulting changes in allele frequencies. In short, it is probably too
-early to tell whether modern natural selection has a substantively important
-effect on population means of phenotypes under selection. Nevertheless, we note
-that selection effects on our measured polygenic scores are still relatively
-small, even after reweighting to account for ascertainment bias.
+not expect that this should change their association with number of offspring,
+or the resulting changes in allele frequencies. Although effects on
+our measured polygenic scores are small even after weighting, individually small
+disadvantages can cumulate to create larger effects. Lastly, note that our
+data comes from people born before 1970. Recent evidence suggests that fertility
+patterns may be changing [@doepke2022economics]. Overall, it is probably 
+too early to tell whether modern natural selection has a substantively important 
+effect on population averages of phenotypes under selection.
 
@@
 increase inequality with respect to polygenic scores. For example, Figure
-\@ref(fig:plot-mean-EA3-by-income) graphs mean polygenic scores for educational
+\@ref(fig:plot-mean-EA3-by-income) plots mean polygenic scores for educational
 attainment (EA3) among children from households of different income groups. The
@@
 Overall, natural selection increases the correlation of polygenic scores with
-income for `r n_increases` out of 33 polygenic scores, with a median
-percentage increase of `r median_pct_change`% (Appendix Table
-\@ref(tab:tbl-inequality-effects)).
-
-If inequalities in polygenic scores are important for understanding social
-structure and mobility [@belsky2018genetic;
-@Rimfeld_2018; @harden2021genetic], then these increases are substantive. A
-back-of-the-envelope calculation, using estimated heritability for educational
-attainment of 0.4 [@branigan2013variation], gives the proportion of variance in
-household income attributable to the "true" polygenic score for educational
-attainment as `r r2_income_true_psea` among respondents. Natural selection would
-raise this proportion to `r r2_income_true_psea_wt` among respondents' children,
-counterfactually holding environmental conditions constant. Similarly, since
-many polygenic scores are predictive of disease risk, they could potentially
-increase health inequalities. In general, the evolutionary history of
-anatomically modern humans is related to disease risk [@benton2021influence];
-understanding the role of contemporary natural selection may aid in mapping the
-genetic architecture of current health disparities.
+income for `r n_increases` out of 33 polygenic scores, with a median percentage
+increase of `r median_pct_change`% in the respondents' generation (Appendix
+Table \@ref(tab:tbl-inequality-effects)).  If inequalities in polygenic scores
+are important for understanding social structure and mobility
+[@belsky2018genetic; @Rimfeld_2018; @harden2021genetic], then these increases
+are substantive. Similarly, since many polygenic scores are predictive of
+disease risk, they could potentially increase health inequalities. In general,
+the evolutionary history of anatomically modern humans is related to disease
+risk [@benton2021influence]; understanding the role of contemporary natural
+selection may help researchers to map the genetic architecture of health
+disparities.
 
@@
 Blue is actual.
-Grey is hypothetical in the absence of selection effects.
-Respondents weighted by Age/Qualification.", fig.width = 4}
+Grey is hypothetical in the absence of selection effects.", fig.width = 4}
 income_EA3 <- famhist %>% 
@@
 pessimism" [@sarraf2019modernity]. Others are more sanguine, and argue that
-natural selection's effects are outweighed by environmental improvements, such
-as those underlying the Flynn effect [@flynn1987massive]. The evidence
+natural selection's effects are outweighed by environmental improvements, like
+those underlying the Flynn effect [@flynn1987massive]. The evidence
 here may add some nuance to this debate. Patterns of natural selection have been
 relatively consistent across the past two generations, but they are not the
-outcome of a universal, society-wide phenomenon. Instead they result from 
+outcome of a single, society-wide phenomenon. Instead they result from 
 opposing forces, operating in different parts of society and pulling in
@@
 for example via Mendelian randomization [@davey2003mendelian].
-Conversely, theory and empirical results from economics can shine a light on the
+Conversely, economic theories and empirical results can shine a light on the
 mechanisms behind natural selection, and thereby on the nature of individual
@@
 stratification, we corrected the polygenic scores for 100 principal components
-(PCs).
-
-To compute polygenic scores and PCs, the same procedures were followed as 
+(PCs). To compute polygenic scores and PCs, the same procedures were followed as 
 described in @abdellaoui2019genetic.
@@
 
-Population data for weighting is taken from the 2011 UK Census and the 2006
-General Household Survey (GHS). Weighting for Age/Qualification and 
-Age/Qualification/AFLB weights was done using marginal totals from a linear
-model, using the `calibrate()` function in
-the R "survey" package [@lumley2020survey]. Geographical weighting was done with
-iterative post-stratification using the `rake()` function, on Census Middle Layer
-Super Output Areas, sex and presence/absence of a partner.
+Weighting data was kindly provided by @van2021effects.
 
 Code for this paper is available at 
-https://github.com/hughjonesd/negative-selection.
+https://github.com/hughjonesd/why-natural-selection.
 
@@
         term = pretty_names(term),
-        term = fct_reorder(term, Female)
+        term = fct_reorder(term, Female),
+        p = diff.p.value
       ) 
@@
 res_sex %>% 
-      mutate(
-        term = pretty_names(term),
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(y = term, x = estimate, fill = sex, 
-            colour = p < 0.05/{{n_regs}})) +
-        my_vline +
-        geom_linerange(data = res_sex_wide, 
+      standard_ggplot(fill_col = sex, n_regs = n_regs, score_col = term,
+                        conf_int = FALSE) +
+      geom_linerange(data = res_sex_wide, inherit.aes = FALSE, 
+                     color = "steelblue4", alpha = 1,
               aes(
@@
                 x      = NULL,
-            fill   = NULL,
-            xmin   = min_est, 
-            xmax   = max_est, 
-            colour = diff.p.value < 0.05/{{n_comps}}
-          ), 
-        group = "") +
-        pval_geom_point() +
-        pval_scale(name = "p < 0.05 (corrected)") +
-        my_fill_scale(n = 2) +
-        my_labs() +
-        my_theme()
+                xmin   = min_est + 0.0005, 
+                xmax   = max_est - 0.0005,
+                linetype = p < 0.05/{{n_comps}}
+              )) +
+      scale_linetype_manual(values = c("TRUE" = 1, "FALSE" = 2))
 
@@
 \clearpage
-## Weighted regressions
+## Alternative weighting schemes
+
+We report results for 3 alternative weighting schemes: weighting by
+age/qualification; geographical (weighting by Middle Super Output Area); and for
+women only, age, qualification and age at first live birth. Population data for
+weighting is taken from the 2011 UK Census and the 2006 General Household Survey
+(GHS). Weighting for Age/Qualification and Age/Qualification/AFLB weights was
+done using marginal totals from a linear model, using the `calibrate()` function
+in the R "survey" package [@lumley2020survey]. Geographical weighting was done
+with iterative post-stratification using the `rake()` function, on Census Middle
+Layer Super Output Areas, sex and presence/absence of a partner.
 
@@
 rel_sizes %>% 
-      arrange(desc(Geographical)) %>% 
+      arrange(desc(Main)) %>% 
       bind_rows(rel_sizes_summary) %>% 
@@
       style_headers(bold = TRUE) %>% 
-      set_bottom_border(1, 2:4) %>% 
+      set_bottom_border(1, -1) %>% 
       set_bottom_border(2, everywhere) %>% 
@@
       set_font_size(8) %>% 
-      set_width(0.7) %>% # ensures line wrapping below
-      set_col_width(c(.34, .22, .22, .22)) %>% 
+      set_width(0.8) %>% # ensures line wrapping below
+      set_col_width(c(.28, .18, .18, .18, .18)) %>% 
       add_footnote(paste0(
@@
 disruptive selection increases variance. To check for these, we rerun equation
-\@ref(eq:regression), adding a quadratic term in $PGS_i$, and using our
-age/qualification weights. Scores for hip circumference show significant
-stabilizing selection ($p < 0.05/33$, negative coefficient on quadratic term).
-The EA2 score for educational attainment shows significant disruptive selection
-($p < 0.05/33$, positive coefficient), which reduces the strength of selection
-against educational attainment at very high levels of the PGS. (The quadratic on
-the EA3 score has a similar coefficient but is not significant at $p < 0.05/33$.)
-Figure \@ref(fig:plot-purifying) plots predicted number of children
-against polygenic score from these regressions.
+\@ref(eq:regression), adding a quadratic term in $PGS_i$. Scores for hip
+circumference show significant stabilizing selection ($p < 0.05/33$, negative
+coefficient on quadratic term). The EA2 score for educational attainment shows
+significant disruptive selection ($p < 0.05/33$, positive coefficient), which
+reduces the strength of selection against educational attainment at very high
+levels of the PGS. (The quadratic on the EA3 score has a similar coefficient but
+is not significant at $p < 0.05/33$.) Figure \@ref(fig:plot-purifying) plots
+predicted number of children against polygenic score from these regressions.
 
 We also checked for stabilizing selection in the parents' generation, using
-age/qualification weights multiplied by the inverse of number of siblings.
-Scores for EA2 and EA3 show significant disruptive selection ($p < 0.05/33$, 
-positive coefficient on quadratic). Other scores including hip circumference 
-were not significant.
-
-```{r plot-purifying, fig.align = "center", fig.cap = "Stabilizing/disruptive selection: predicted number of children by polygenic score.", fig.height = 3}
+weights multiplied by the inverse of number of siblings. Scores for EA2 and EA3
+show significant disruptive selection ($p < 0.05/33$, positive coefficient on
+quadratic). Other scores including hip circumference were not significant.
+
+```{r plot-purifying, fig.align = "center", fig.cap = "Stabilizing/disruptive selection: predicted number of children by polygenic score."}
 
@@
 
-stopifnot(sig == c("EA2_noUKB", "hip_combined"))
+stopifnot(sig == c("ADHD_2017", "bmi_combined", "EA2_noUKB", 
+                     "hip_combined", "wc_combined"))
 
@@
 
-op <- par(mfrow = c(1, 2))
+op <- par(mfrow = c(2, 3))
+
 for (score in sig) {
@@
 with a quadratic in age, the interaction with income remains significant at
-$p < 0.05/`r n_regs_ic`$ for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
+$p$ < 0.05/`r n_regs_ic` for `r n_sig_ic` out of `r n_regs_ic` regressions. Similarly
 if we interact the PGS with age of leaving full time education and a quadratic
 in age, the interaction with age leaving full time education remains significant
-at $p < 0.05/`r n_regs_educ`$ for `r n_sig_educ` out of `r n_regs_educ` regressions.
+at $p$ < 0.05/`r n_regs_educ` for `r n_sig_educ` out of `r n_regs_educ` regressions.
 
@@
 res_partners %>% 
-      mutate(
-        score_name = pretty_names(score_name)
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(
-        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        x      = estimate, 
-        fill   = `N partners`, 
-        colour = p < 0.05/{{n_regs}}
-      )) + 
-        my_vline +
-        confint_geom_segment() +
-        pval_geom_point() +
-        facet_wrap(vars(sex), scales = "free_y") +
-        my_fill_scale(n = 2) +
-        pval_scale() +
-        sexes_theme +
-        my_labs(x = "", y = "") + 
-        scale_y_reordered() +
-        scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
+      standard_ggplot(fill_col = `N partners`, score_col = score_name, 
+                        n_regs = n_regs) +
+        facet_grid(cols = vars(sex))
         sexes_guides
@@
       mutate(
-        score_name = pretty_names(score_name),
         Household = ifelse(grepl("TRUE", term), "With partner", 
@@
       ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(
-        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        x      = estimate, 
-        fill   = Household, 
-        colour = p < 0.05/{{n_regs}}
-      )) + 
-        my_vline +
-        confint_geom_segment() +
-        pval_geom_point() +
-        facet_wrap(vars(sex), scales = "free_y") +
-        my_fill_scale(n = 2) +
-        pval_scale() +
-        sexes_theme +
-        my_labs(x = "", y = "") + 
-        scale_y_reordered() +
-        scale_linetype_manual(
-                                name   = "", 
-                                values = c("95% c.i. uncorrected" = 1)
-                              ) +
+      standard_ggplot(fill_col = Household, score_col = score_name, 
+                        n_regs = n_regs) +
+        facet_grid(cols = vars(sex)) + 
         sexes_guides
@@
 ```{r calc-res-sibs}
-drake::loadd(res_wt_age_qual_weights)
+drake::loadd(res_wt_van_alten_weights)
 drake::loadd(res_children_comparison)
 
-cmp <- left_join(res_wt_age_qual_weights, res_children_comparison, by = "term")
+cmp <- left_join(res_wt_van_alten_weights, res_children_comparison, by = "term")
 
@@
 estimate equation (\ref{eq:regression}) using parents' RLRS as the dependent
-variable. The parents' generation has an additional source of ascertainment
+variable.[^parent-RLRS] The parents' generation has an additional source of ascertainment
 bias: sampling parents of respondents overweights parents who have many
@@
 times more children represented in UK Biobank than parents of one child. Parents
-of no children will by definition not be represented. To compensate, we reweight
-our preferred weightings (Age/Qualification) by the inverse of *number of
-siblings*.
+of no children will by definition not be represented. To compensate, we multiply
+our weights by the inverse of *number of siblings*.
+
+[^parent-RLRS]: We don't have data on parents' year of birth for most 
+respondents. To create parents' RLRS, we divide respondents' number of siblings 
+by the average number of siblings of all respondents born in the same year,
+weighting the average by respondents' inverse of number of siblings to 
+compensate for ascertainment bias.
 
@@
 `r n_smaller` out of 33 scores; the median percentage change is 
-`r percent(median_pct_diff)`.
+`r scales::percent(median_pct_diff)`.
 
@@
 respondents.
-Age/qualification weights.
 Parental generation weights multiplied by 1/number of siblings.
@@
 stopifnot(sum(summary_period$Change_Parents == "Insignificant") == 32)
-stopifnot(sum(summary_period$Change_Respondents == "Size increasing") == 9)
-stopifnot(sum(summary_period$Change_Respondents == "Change sign") == 1)
-stopifnot(sum(summary_period$Change_Respondents == "Insignificant") == 23)
+stopifnot(sum(summary_period$Change_Respondents == "Size increasing") == 8)
+stopifnot(sum(summary_period$Change_Respondents == "Insignificant") == 25)
 ```
@@
 birth year, median split at 1950 ("early born" versus "late born"). We use both
-respondents' RLRS and parents' RLRS as a dependent variable. We weight
-using age/qualification cells, and further adjust for selection in the parents'
-generation (see above).
+respondents' RLRS and parents' RLRS as a dependent variable. We use our standard
+weights, and further adjust for selection in the parents' generation (see above).
 
@@
 effect sizes were significantly larger in absolute size among the later-born for
-nine polygenic scores: ADHD, age at menopause, caffeine consumption, cognitive
+eight polygenic scores: ADHD, age at menopause, cognitive
 ability, Coronary Artery Disease, EA2, EA3, extraversion and Major Depressive
@@
 where estimated effect sizes were larger among the earlier, parents' generation.
-One score, for alzheimer's disease, showed a significant change in sign, from
-negative to positive effects on fertility.
-
-Overall, there is some evidence for an increase in the strength of selection in
-recent history. The clearest result is that the pattern of relative effect sizes
-across scores is broadly consistent over time.
+
+Overall, while there is some suggestive evidence for an increase in the strength
+of selection in recent history, the clearest result is that the pattern of
+relative effect sizes across scores is broadly consistent over time.
 
@@
             set_right_padding(everywhere, 3, 0) %>% 
-            set_caption("Numbers of polygenic scores showing changes in selection effects between early and late born. 
-      Weights for respondents: Age/Qualification. Weights for parents: Age/Qualification times 1/number of siblings.") %>% 
+            set_caption("Numbers of polygenic scores showing changes in selection effects between early and late born. Parental generation weights multiplied by 1/number of siblings.") %>% 
             set_caption_width(1)
@@
 Higher = more deprived.
-Estimates weighted by 1/number of siblings.", fig.align = "center"}
+Weights multiplied by 1/number of siblings.", fig.align = "center"}
 
@@
 ```{r plot-age-birth-parents-cross, fig.cap = "Selection effects (parents' RLRS) by age at first live birth terciles.
-Estimates weighted by 1/number of siblings.", fig.subcap = c("Mothers", "Fathers"), fig.ncol = 1}
+Weights multiplied by 1/number of siblings.", fig.subcap = c("Mothers", "Fathers"), fig.ncol = 1}
 
@@
 
-```{r plot-age-birth-parents, fig.cap = "Selection effects (parents' generation) among eldest siblings, controlling for parents' age at birth. Estimates weighted by 1/number of siblings."}
+```{r plot-age-birth-parents, fig.cap = "Selection effects (parents' generation) among eldest siblings, controlling for parents' age at birth. Weights multiplied by 1/number of siblings."}
 
@@
 
-```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at respondent's birth, eldest siblings. Estimates weighted by 1/number of siblings."}
+```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at respondent's birth, eldest siblings. Weights multiplied by 1/number of siblings."}
 
@@
 We run a standard mediation analysis in the framework of @baron1986moderator.
-For each polygenic score where the bivariate correlation with fertility is 
+For each polygenic score where the bivariate correlation with RLRS is 
 significant at $p$ < 0.05/33, we estimate 
 \begin{align}
-N_i & = \alpha + \beta PGS_i + \gamma EA_i + X_i\mu + \varepsilon_i \\
+RLRS_i & = \alpha + \beta PGS_i + \gamma EA_i + X_i\mu + \varepsilon_i \\
 EA_i & = \delta + \zeta PGS_i + X_i\mu + \eta_i
 \end{align}
-where $N_i$ is fertility, $PGS_i$ is the polygenic score, $EA_i$ is educational
-attainment (age of leaving fulltime education), and $X_i$ is a vector of
-controls. The total effect of $PGS$ on number of children is $\beta + \gamma
-\zeta$. The "indirect effect" mediated by $EA$ is $\gamma \zeta$. The standard
-error of the indirect effect can be calculated as
+where $RLRS_i$ is relative lifetime reproductive success, $PGS_i$ is the
+polygenic score, $EA_i$ is educational attainment (age of leaving fulltime
+education), and $X_i$ is a vector of controls. The total effect of $PGS$ on
+number of children is $\beta + \gamma \zeta$. The "indirect effect" mediated by
+$EA$ is $\gamma \zeta$. The standard error of the indirect effect can be
+calculated as
 
@@
 controlling for fluid IQ is conservative, in the sense that IQ might also measure
-human capital; nevertheless, we control for it because fluid IQ might affect
-fertility decisions via other channels than human capital.
+human capital in the labour market; nevertheless, we control for it because 
+fluid IQ might affect fertility decisions via other channels than human capital.
 
@@
 n_pos_sig <- sum(res_mediation$sig & res_mediation$prop_ind > 0)
+
+footnote <- sprintf("* p < 0.05/33. Analysis run on %s PGS which correlated 
+                    significantly with fertility.", nrow(res_mediation))
 
@@
       set_font_size(9) %>% 
-      add_footnote("* p < 0.05/33.", font_size = 8)
+      add_footnote(footnote, font_size = 8)
 ```
@@
 
-Results in the main text support our theory that natural selection on polygenic scores is
-driven by their *correlation* with human capital. Here, we test whether polygenic
-scores *cause* fertility by running within-siblings regressions (Figure
-\@ref(fig:plot-res-fe-fertility)). With a reduced sample size, all
-within-sibling effects are insignificant after Bonferroni correction. However,
-effect sizes are positively correlated with effect sizes from the pooled model,
-and about 50% smaller (regressing within-sibling on pooled effect sizes, $b$ =
-`r format(coef_pool_w, digits = 3)`). This attenuation is in line with the ~50%
-decrease in heritability in within-sibling GWASs on age at first birth and
-educational attainment [@Howe2021.03.05.433935]. We see these results as
-providing tentative evidence that polygenic scores cause fertility, with effects
-being partly driven by correlations with environmental variation in human
-capital. We also reran within-siblings regressions adding a control for
-education. Most effect sizes barely change, suggesting that our measure of
-education does not in general mediate differences in fertility among siblings.
-
 ```{r plot-res-fe-fertility, fig.cap = "Selection effects controlling for sibling-group fixed effects, with and without a control for education (left education before 16, 16-18, or after 18).
@@
 
-standard_ggplot(res_fe_fertility, fill_col = Regression, n_regs = n_regs,
-                  fill_direction = -1)
+res_fe_fertility %>% 
+  mutate(
+    Controls = ifelse(Regression == "Controlled", "FE + education", "FE only")
+  ) %>% 
+  standard_ggplot(fill_col = Controls, n_regs = n_regs,
+                  fill_direction = -1) +
+    guides(alpha = guide_legend(
+                   override.aes = list(
+                             shape = 1,
+                             color = "black",
+                             alpha = 1,
+                             size = 1.8
+                   )
+                 )
+    )
 
@@
 
-We run a single regression on 29 polygenic scores within `r n_groups` sibling
+Results in the main text support our theory that natural selection on polygenic scores is
+driven by their *correlation* with human capital. Here, we test whether polygenic
+scores *cause* fertility by running within-siblings regressions. We run a single regression on 29 polygenic scores within `r n_groups` sibling
 groups (N = `r n`). Thus, we control both for environmental confounds (since
@@
 
+With a reduced sample size, all within-sibling effects are insignificant after
+Bonferroni correction. However, effect sizes are positively correlated with
+effect sizes from the pooled model, and about 70% smaller (regressing
+within-sibling on pooled effect sizes, $b$ = `r format(coef_pool_w, digits = 3)`).
+This attenuation is broadly consistent with the decrease in heritability
+in within-sibling GWASs on age at first birth and educational attainment
+[@Howe2021.03.05.433935]. We see these results as providing tentative evidence
+that polygenic scores cause fertility, with effects being partly driven by
+correlations with environmental variation in human capital. We also reran
+within-siblings regressions adding a control for education. Most effect sizes
+barely change, suggesting that our measure of education does not in general
+mediate differences in fertility among siblings.
+
 \clearpage
@@
 polygenic scores and household income (UKB data field 738). Column "With
-selection" uses respondents' scores weighted by Age/Qualification, times number
-of children. Column "Without selection" uses scores weighted by
-Age/Qualification only, i.e. the counterfactual correlation if all respondents
+selection" uses respondents' scores, multiplying weights by number
+of children. Column "Without selection" uses our standard weights, i.e. 
+it estimates the counterfactual correlation if all respondents
 had the same number of children.
@@
 
-## Estimating the effect of natural selection on the contribution of polygenic scores for educational attainment to inequality
-
-```{r calc-load-ineq-ea3}
-
-res <- drake::readd(res_ineq_ea3)
-
-prop_effect_selection <- res$r2_income_ea3_wt/res$r2_income_ea3 - 1
-prop_effect_selection <- scales::percent(prop_effect_selection, accuracy = 0.1)
-
-prop_wt_fe <- res$r2_income_ea3_fe_wt/res$r2_income_ea3_fe - 1
-prop_wt_fe <- scales::percent(prop_wt_fe, accuracy = 0.1)
-
-prop_wt_fe_bounds <- scales::percent(res$r2_fe_bounds, accuracy = 0.1)
-
-res <- lapply(res, format, digits = 3)
-
-```
-
-From the meta-analysis of twin studies in @branigan2013variation, we take the
-heritability of educational attainment to be around 0.4. This is the variance of
-educational attainment explained by the "true" polygenic score for educational
-attainment (PSEA) $p^{*}$, i.e. the best linear unbiased predictor of educational
-attainment from genetic data. We write the measured polygenic
-score of educational attainment, "EA3" from @lee2018gene, as
-\[
-p = p^* + \eta.
-\]
-
-For realized educational attainment $EA$, write:
-\[
-EA = \alpha+\beta p^* + \varepsilon = \alpha + \beta(p - \eta) + \varepsilon
-\]
-We assume the classical errors-in-variables model, i.e. that $\eta$
-is independent of both $p$ and $\varepsilon$. Under these conditions,
-the ratio between the estimated $R^{2}$ in a regression of $p$ on
-$EA$, and the true $R^{2}$ of $p^{*}$ on $EA$, satisfies: 
-\begin{equation}
-\frac{\textrm{plim}\hat{R}^{2}}{R^{2}} 
-= \frac{\sigma_{p}^{2} - \sigma_{\eta}^{2}}{\sigma_{p}^{2}}
-= \frac{\sigma_{p^*}^{2}}{\sigma_{p}^{2}}
-\equiv \lambda, \label{eq:r2-ratio}
-\end{equation}
-as we show below. The right hand side is the "reliability ratio".
-We regress educational attainment on EA3 in our sample, weighting
-by our Age/Qualification weights, to give 
-$\textrm{plim}\hat{R}^2 = `r res$r2_edu_ea3`$. Plugging this
-into the above, with $R^2 = h^2\approx0.4$, gives $\lambda=`r res$lambda`$.
-
-We now consider the $R^2$ of $p$ on income, that is, the proportion
-of variance in income that is explained by the true polygenic score
-of educational attainment. Applying (\ref{eq:r2-ratio}) again gives
-that
-\begin{equation}
-\frac{\textrm{plim}\hat{R}_{\textrm{income}}^{2}}{R_{\textrm{income}}^{2}}=\lambda.\label{eq:r2-ratio-income}
-\end{equation}
-
-We compute $\hat{R}_{\textrm{income}}^{2}$by regressing income group on EA3,
-first weighting by age/qualification, and then by age/qualification weights
-times respondent's number of children. The estimated $R^2$ are 
-`r res$r2_income_ea3` and `r res$r2_income_ea3_wt` respectively. Applying
-(\ref{eq:r2-ratio-income}) to both estimates gives the proportion of variance in
-income explained by true polygenic score for education attainment (a) among
-respondents as `r res$r2_income_true_psea`; (b) accounting for natural selection
-as `r res$r2_income_true_psea_wt`. The second figure gives the counterfactual
-estimate of $R_{\textrm{income}}^{2}$ after one generation of natural selection,
-holding environmental conditions unchanged.
-
-The errors-in-variables conditions are quite restrictive. In particular, they
-rule out that the unmeasured part of PSEA ($\eta$) correlates with the error 
-term $\varepsilon$, e.g. with environmental causes of educational attainment. 
-If unmeasured PSEA correlates positively with $\varepsilon$, then the
-reliability coefficient $\lambda$ will be lower than calculated. However, the
-proportional effect of natural selection on the PSEA-income relationship will
-remain the same, so long as $\lambda$ does not differ between the two sets of
-weighted regressions.
-
-The estimate above uses a pooled regression of income on EA3, which may include
-confounding environmental variation such as genetic nurture. Within-siblings
-regression avoid this, but the smaller N makes estimates noisier. Running
-within-siblings regressions of EA3 on income group gives within-$R^2$ estimates
-of `r res$r2_income_ea3_fe` (weighted by age and qualification) and `r
-res$r2_income_ea3_fe_wt` (multiplying weights by number of children). Thus, this
-procedure estimates the proportional change in $R^2_{income}$ due to natural
-selection as `r prop_wt_fe`. However, bootstrap 95% confidence intervals
-are large (`r prop_wt_fe_bounds[1]` to `r prop_wt_fe_bounds[2]`). We therefore
-view this result more as suggesting that environmental confounds likely
-bias the estimated change in $R^2_{income}$ towards zero.
-
-### Proof of \@ref(eq:r2-ratio)
-
-Write 
-\begin{align}
-\hat{R}^{2} & =1-\frac{\sum_{i=1}^{N}(\hat{\alpha}+\hat{\beta}p_{i}-EA_{i})^{2}}{\sum_{i=1}^{N}(EA_{i}-\bar{EA})^{2}}\label{eq:r-squared}
-\end{align}
-
-Using the fact that $\hat{\alpha}=\bar{EA}-\hat{\beta}\bar{p}$, where bars signify
-sample averages, we can simplify the numerator in the above:
-
-\[
-\sum_{i=1}^{N}(\hat{\alpha}+\hat{\beta}p_{i}-EA_{i})^{2}=\sum_{i=1}^{N}(\bar{EA}-\hat{\beta}\bar{p}+\hat{\beta}p_{i}-EA_{i})^{2}=\sum_{i=1}^{N}(\hat{\beta}(p_{i}-\bar{p})-(EA_{i}-\bar{EA}))^{2}
-\]
-
-Now, we can compute the following probability limit:
-
-\begin{align}
- & \text{plim}\frac{\sum_{i=1}^{N}(\hat{\beta}(p_{i}-\bar{p})-(EA_{i}-\bar{EA}))^{2}}{N}\nonumber \\
-= & \text{plim}\frac{\sum_{i=1}^{N}\hat{\beta}^{2}(p_{i}-\bar{p})^{2}+(EA_{i}-\bar{EA})^{2}-2\hat{\beta}(p_{i}-\bar{p})(EA_{i}-\bar{EA})}{N}\nonumber \\
-= & \beta^{2}\lambda^{2}\sigma_{p}^{2}+\sigma_{EA}^{2}-2\beta\lambda\sigma_{p,EA},\label{eq:plim}
-\end{align}
-
-using the standard errors-in-variables formula $\textrm{plim}\hat{\beta}=\lambda\beta$,
-and where $\sigma_{p,EA}$ is the covariance between $p$ and $EA$.
-
-Write
-\begin{align*}
-\sigma_{p,EA} & =\textrm{Cov}(p^{*}+\eta,\alpha+\beta p^{*}+\varepsilon)\\
- & =0+\beta\sigma_{p^{*}}^{2}+\beta\textrm{Cov}(\eta,p^{*})+\textrm{Cov}(\eta,\varepsilon)\\
- & =\beta\sigma_{p^{*}}^{2}
-\end{align*}
-
-using the errors-in-variables assumption that $\eta$ is independent
-of $p^{*}$ and $\varepsilon$. Plug this into (\ref{eq:plim}) to
-give
-\begin{align*}
- & \text{plim}\frac{\sum_{i=1}^{N}(\hat{\beta}(p_{i}-\bar{p})-(EA_{i}-\bar{EA}))^{2}}{N}\\
-= & \beta^{2}\lambda^{2}\sigma_{p}^{2}+\sigma_{EA}^{2}-2\beta^{2}\lambda\sigma_{p^{*}}^{2}\\
-= & \beta^{2}\lambda\sigma_{p^{*}}^{2}+\sigma_{EA}^{2}-2\beta^{2}\lambda\sigma_{p^{*}}^{2}\textrm{, using \ensuremath{\sigma_{p^{*}}^{2}=\lambda\sigma_{p}^{2}},}\\
-= & \sigma_{EA}^{2}-\beta^{2}\lambda\sigma_{p^{*}}^{2}
-\end{align*}
-
-Inserting this into (\ref{eq:r-squared}) gives
-\begin{align*}
-\textrm{plim}\hat{R}^{2} & =1-\frac{\sigma_{EA}^{2}}{\sigma_{EA}^{2}}+\frac{\beta^{2}\lambda\sigma_{p^{*}}^{2}}{\sigma_{EA}^{2}}\\
- & =\lambda\frac{\beta^{2}\sigma_{p^{*}}^{2}}{\sigma_{EA}^{2}}\\
- & =\lambda R^{2}
-\end{align*}
-since for bivariate regression $R^{2}$ is the square of the correlation
-between $EA$ and $p^{*}$, which can be written $r=$$\beta\frac{\sigma_{p^{*}}^{2}}{\sigma_{EA}\sigma_{p^{*}}}$.
-We thank an anonymous commenter on https://stats.stackexchange.com/questions/238878/how-do-errors-in-variables-affect-the-r2
-for much of this proof.
-
 \clearpage
@@
 
-```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores on number of siblings/children. Sibling estimates weighted by 1/number of siblings."}
+```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores.
+Parental generation weights multiplied by 1/number of siblings.", fig.subcap = c("Respondents", "Parents"), fig.ncol = 1}
 
@@
 n_regs <- as.double(nrow(res_all))
+res_all %<>% mutate(
+               `PGS` = ifelse(reg.type == "controlled", 
+                                "Residualized for 100 PCs", "Unresidualized")
+             )
+
 res_all %>% 
-      mutate(
-        term     = pretty_names(term),
-        term     = reorder_within(term, estimate, dep.var, order_abs(1)),
-        dep.var  = fct_inorder(dep.var),
-        reg.type = fct_recode(reg.type, 
-                               `Resid. for 100 PCs`  = "controlled", 
-                               Unresidualized        = "raw"
-                             )
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(y = term, x = estimate, fill = reg.type, 
-        colour = p < 0.05/{{n_regs}})) +
-        my_vline + 
-        facet_wrap(vars(dep.var), scales = "free") +
-        pval_geom_point(position = position_dodge(0.4), alpha = 0.8) +
-        my_labs() +
-        pval_scale() +
-        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
-        my_theme() +
-        scale_y_reordered() +
-        theme(
-          axis.text.x  = element_text(size = 7),
-          legend.text  = element_text(size = 8),
-          legend.title = element_text(size = 9)
-        )
+      filter(dep.var == "RLRS") %>% 
+      standard_ggplot(fill_col = PGS, n_regs = n_regs)
+
+res_all %>% 
+      filter(dep.var == "RLRS_parents") %>% 
+      standard_ggplot(fill_col = PGS, n_regs = n_regs)
 
@@
 Each dot represents one bivariate regression.
-Sibling estimates weighted by 1/number of siblings.
+Parental generation weights multiplied by 1/number of siblings.
 Absolute effect sizes are plotted.
@@
         dep.var  = fct_relevel(dep.var, "RLRS"),
+        dep.var  = fct_recode(dep.var,
+          "Parents"     = "RLRS_parents",
+          "Respondents" = "RLRS"
+        ),
         top_pcs  = ifelse(as.numeric(pc_name) <= 5, term, "")
@@
       rename(p = p.value) %>% 
-      ggplot(aes(estimate, dep.var, colour = p < 0.05/{{n_regs}})) + 
-        geom_point(alpha = 0.8, position = pj) +
+      ggplot(aes(estimate, dep.var, shape = p < 0.05/{{n_regs}})) + 
+        geom_point(alpha = 0.8, color = "steelblue4", position = pj) +
         geom_text(aes(estimate, dep.var, label = top_pcs), position = pj, 
@@
         labs(x = "Effect size (log scale)", y = "") + 
-        pval_scale(aesthetics = "colour")
+        scale_shape_manual(values = c("FALSE" = "circle open", "TRUE" = "circle"))
 

### 2022-03-19 why-natural-selection.Rmd
Use alternative plot with shapes

--- 2022-03-15 why-natural-selection.Rmd
@@
   if (is.null(conf_int)) conf_int <- "conf.low" %in% names(dfr)
+  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL
+
+  n_regs <- as.double(n_regs)
+
+  dfr %>% mutate(
+    !! score_col := pretty_names(!! score_col),
+    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
+  ) %>% 
+    rename(p = p.value) %>% 
+    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
+               shape = {{fill_col}}, ...)) +
+    my_vline + 
+    confint_geom_segment +
+    geom_point(size = 1.8, aes(color = {{fill_col}})) +
+    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/33)) +
+    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction) +
+    scale_shape_manual(values = 1:n_cats + 14) +
+    my_labs() +
+    my_theme() +
+    scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
+    scale_alpha_discrete(range = c(1, 0),
+                         guide = guide_legend(
+                           override.aes = list(
+                             shape = c(1, 16),
+                             color = "black",
+                             alpha = 1,
+                             size = 1.8
+                           )
+                         ))
+}
+
+old_ggplot <- function (
+        dfr, 
+        fill_col, 
+        n_regs, 
+        ...,
+        score_col = quo(term), 
+        order_idx = 1, 
+        fill_direction = 1,
+        n_cats    = NULL,
+        conf_int  = NULL
+      ) {
+  if (! missing(score_col)) score_col <- enquo(score_col)
+
+  mfc <- missing(fill_col)
+  fill_col <- if (mfc) quo(NULL) else enquo(fill_col)
+  if (missing(n_cats)) {
+    n_cats <- if (mfc) 1 else length(unique(pull(dfr, {{fill_col}})))
+  }
+
+  if (is.null(conf_int)) conf_int <- "conf.low" %in% names(dfr)
 

### 2022-03-19 why-natural-selection.Rmd
Use alternative plot with shapes

@@
 ```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank.
-Points are means for 5-year intervals.
+Circles are means for 5-year intervals.
 Lines are 95% confidence intervals.
-Circles show a significant linear increase (green) or decrease (red) over time (p < 0.05/33)."}
+Solid circles show a significant linear increase (green) or decrease (red) over time (p < 0.05/33)."}
 
@@
       ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
-      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
-        fatten = 1.2) +
+      geom_linerange(aes(ymin = score_lo, ymax = score_hi), color = "grey60", 
+                       size = 3) +
+      geom_point(aes(y = score), size = 2) +
       facet_wrap(~score_name, ncol = 6) +
       my_hline +
-      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
-      scale_shape_manual(values = c("-" = "circle", "+" = "circle", "o" = "square")) +
+      scale_color_manual(values = c("-" = "darkred", "+" = "darkgreen", "o" = "black")) +
+      scale_shape_manual(values = c("-" = "circle", "+" = "circle", "o" = "circle open")) +
       theme(
@@
 To test more directly for natural selection, we regress respondents' relative 
-lifetime reproductive success ($RLRS_i$) on their polygenic scores (PGS):
+lifetime reproductive success (RLRS) on their polygenic scores (PGS):
 
@@
 
-To motivate our economic model, we split the sample by demographic
-and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
-We examine socio-demographic variables including income and education, and
-family structure variables including age at first live birth, presence of a
-partner, and lifetime number of sexual partners. Figure
-\@ref(fig:plot-income-educ-level) plots selection effects for each polygenic
-score, grouping respondents by age of completing full-time education, and by
-household income. Effects are larger and more significant for the lowest income
-category, and for the lowest education category. These results are robust to
-controlling for respondents' age (Appendix section \@ref(sec:age-control)).
+We now show the empirical puzzles which motivate our economic model. Each
+concerns differences in the strength of natural selection across different
+subgroups in the sample. Splitting the sample by demographic and social
+variables, we re-estimate \@ref(eq:regression) within each subgroup. We use
+socio-economic variables including income and education, and family structure
+variables including age at first live birth, presence of a partner, and lifetime
+number of sexual partners.
+
+Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
+polygenic score, grouping respondents by age of completing full-time education,
+and by household income. Effects are larger and more significant for the lowest
+income category, and for the lowest education category. These results are robust
+to controlling for respondents' age (Appendix section \@ref(sec:age-control)).
 
@@
 
-We emphasize that our categories are not necessarily exogenous to the polygenic
-scores. For example, in the data, and also in our theoretical model, age at first
-live birth is a choice variable, which is endogenous to human-capital-related
-polygenic scores (Appendix Figure \@ref(fig:plot-age-flb-dv)). Nevertheless,
-differences in equation \@ref(eq:regression) across subgroups constrain the set
-of possible explanations for natural selection. As we show below, a simple
-economic model predicts the differences we find.
+We emphasize that our categories are not necessarily exogenous to polygenic
+scores. For example -- both in the data (Appendix Figure
+\@ref(fig:plot-age-flb-dv)) and in our model -- age at first live birth is a
+choice variable, which is endogenous to an individual's human capital and
+related polygenic scores. Nevertheless, observed differences in equation
+\@ref(eq:regression) across subgroups constrain the set of possible
+theories; a good theory of contemporary natural selection needs to explain
+how these results arise. As we show below, a model based on the economic
+theory of fertility can do just that.
 
@@
 
-Instead of identifying human capital with wages, we now allow
-individuals to choose a level of education $s \in [0,1]$, which has
-a time cost in period 1. Education is
-complementary to human capital $h > 0$, and increases period 2 wages, which take 
-the  simple functional form $w(s,h) = sh$. We normalize period 1 wages
-to 1, and let $u(\cdot)$ take the constant relative risk aversion form
-$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$. $\sigma > 0$ measures the curvature
-of the utility function, i.e. the decline in marginal utility of income as income
-increases. We examine total fertility $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the 
-*fertility-human capital relationship*, $\frac{dN^{*}}{dh}$. For $\sigma < 1$ in
-a neighbourhood of $\sigma = 1$, Table \@ref(tab:theory) shows five theoretical
-predictions, along with our corresponding empirical results for the correlation
-between polygenic scores and RLRS.[^one-period-results] The key insight of
-the model is that for middling levels of $\sigma$, the substitution effect
-dominates at low income levels, but is balanced by the income effect as income
-increases.
+Instead of identifying human capital with wages, we now allow individuals to
+choose a level of education $s \in [0,1]$, which has a time cost in period 1.
+Education is complementary to human capital $h > 0$, and increases period 2
+wages, which take the  simple functional form $w(s,h) = sh$. We normalize period
+1 wages to 1, and let $u(\cdot)$ take the constant relative risk aversion form
+$u(y)=\frac{y^{1 - \sigma} - 1}{1 - \sigma}$. $\sigma > 0$ measures the
+curvature of the utility function, i.e. the decline in marginal utility of
+income as income increases. We examine total fertility 
+$N^{*} = N_{1}^{*} + N_{2}^{*}$ and the *fertility-human capital relationship*,
+$\frac{dN^{*}}{dh}$. For $\sigma < 1$ in a neighbourhood of $\sigma = 1$, Table
+\@ref(tab:theory) shows five theoretical predictions, along with our
+corresponding empirical results for the correlation between polygenic scores and
+RLRS.[^one-period-results] The key insight of the model is that for middling
+levels of $\sigma$, the substitution effect dominates at low income levels, but
+is balanced by the income effect as income increases.
 
@@
 mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
-@becker1981treatise. There is also a literature showing negative correlations
+@becker1981treatise. A related literature shows negative correlations
 between IQ and fertility [e.g. @lynn2004new; @reeve2018systematic].
@@
 low-income individuals. However, the majority of effect sizes appear unchanged
-over a large span of twentieth-century history (Appendix Tables
-\@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)), during
+over a large span of twentieth-century history (Appendix Table
+\@ref(tab:tbl-change-by-period)), during
 which government spending on child-related benefits varied considerably
@@
 Code for this paper is available at 
-https://github.com/hughjonesd/negative-selection.
+https://github.com/hughjonesd/why-natural-selection.
 
@@
 siblings*.
+
+[^parent-RLRS]: We don't have data on parents' year of birth for most 
+respondents. To create parents' RLRS, we divide respondents' number of siblings 
+by the average number of siblings of all respondents born in the same year,
+weighting the average by respondents' inverse of number of siblings to 
+compensate for ascertainment bias.
 

### 2022-03-19 why-natural-selection.Rmd
Use shades to darken light colours

@@
 
-old_ggplot <- function (
-        dfr, 
-        fill_col, 
-        n_regs, 
-        ...,
-        score_col = quo(term), 
-        order_idx = 1, 
-        fill_direction = 1,
-        n_cats    = NULL,
-        conf_int  = NULL
-      ) {
-  if (! missing(score_col)) score_col <- enquo(score_col)
-
-  mfc <- missing(fill_col)
-  fill_col <- if (mfc) quo(NULL) else enquo(fill_col)
-  if (missing(n_cats)) {
-    n_cats <- if (mfc) 1 else length(unique(pull(dfr, {{fill_col}})))
-  }
-
-  if (is.null(conf_int)) conf_int <- "conf.low" %in% names(dfr)
-
-  n_regs <- as.double(n_regs)
-
-  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL
-
-  dfr %>% mutate(
-          !! score_col := pretty_names(!! score_col),
-          !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
-        ) %>% 
-        rename(p = p.value) %>% 
-        ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
-                     fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
-          my_vline + 
-          confint_geom_segment +
-          pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
-          pval_scale() +
-          my_fill_scale(n = n_cats, direction = fill_direction) +
-          my_labs() +
-          my_theme() +
-          scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1))
-
-}
-
 my_theme <- function (...) theme(axis.text.y = element_text(size = 7), ...)

### 2022-03-19 why-natural-selection.Rmd
fix one plot

@@
 loadNamespace("scales")
-loadNamespace("shades")
 
@@
   if (is.null(conf_int)) conf_int <- "conf.low" %in% names(dfr)
+
+  n_regs <- as.double(n_regs)
+
   confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL
-
-  n_regs <- as.double(n_regs)
 
@@
         ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
-               shape = {{fill_col}}, ...)) +
+                     fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
           my_vline + 
           confint_geom_segment +
-    geom_point(size = 1.8, aes(color = {{fill_col}})) +
-    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/33)) +
-    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction) +
-    scale_shape_manual(values = 1:n_cats + 14) +
+          pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
+          pval_scale() +
+          my_fill_scale(n = n_cats, direction = fill_direction) +
           my_labs() +
           my_theme() +
-    scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
-    scale_alpha_discrete(range = c(1, 0),
-                         guide = guide_legend(
-                           override.aes = list(
-                             shape = c(1, 16),
-                             color = "black",
-                             alpha = 1,
-                             size = 1.8
-                           )
-                         ))
+          scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1))
+
 }
@@
   } else {
-    shades::lightness(scale_color_brewer(aesthetics = aesthetics, ...), 
-                        shades::scalefac(0.85))
+    scale_color_brewer(aesthetics = aesthetics, ...)
   }
@@
 standard_ggplot(res_fe_fertility, fill_col = Regression, n_regs = n_regs,
-                  fill_direction = -1) +
-  guides(alpha = guide_legend(
-                   override.aes = list(
-                             shape = 1,
-                             color = "black",
-                             alpha = 1,
-                             size = 1.8
-                   )
-                 )
-  )
+                  fill_direction = -1)
 

### 2022-03-23 regressions.R
Rework plot 1; polish flow in section 2

@@ -15,6 +15,7 @@ suppressPackageStartupMessages({
   loadNamespace("psych")
 })
 
+
 calc_pgs_over_time <- function (famhist, score_names) {
   pgs_over_time <- famhist %>% 
         select(YOB, all_of(score_names)) %>% 

### 2022-03-26 response-to-reviewers-BE.Rmd
Tweak
update readme
Rename bibliography

@@
 
-Code to produce this paper is now publicly available at <https://github.com/hughjonesd/negative-selection>. XXX
+Code to produce this paper is now publicly available at <https://github.com/hughjonesd/why-natural-selection>.
 
@@
 This seems to require simple (if tedious) applications of the product and chain rules.
-The important conclusion, that N_2 is convex in h for sigma close to 1, is something that I will accept by analogy to the case N_1, N_2 > 0.
+The important conclusion, that N_2 is convex in h for sigma close to 1, is something that I will accept by analogy to the case N_1, N_2 \> 0.
 
@@
 
-> (3) I do not understand in what way Equations 13 and 14 determine when the case N_1 = 0, N_2 > 0 holds.
+> (3) I do not understand in what way Equations 13 and 14 determine when the case N_1 = 0, N_2 \> 0 holds.
 Nevertheless, I accept the intuitive conclusion that it holds for intermediate values of h because too small a value of h (which determines s\*) will make N_1\* in Equation 17 greater than zero and because too high a value will make N_2\* in 18 negative.
@@
 We prefer our tweaked version (and have used it in the revision), but we will bow to the reviewers or editor.
+For comparison, a version of the whole paper using the symbols plot is available [here](https://raw.githubusercontent.com/hughjonesd/why-natural-selection/alt-plots/why-natural-selection-weighted.pdf).
 

### 2022-03-28 why-natural-selection.Rmd
Talk about RLRS in mediation analysis

@@
 We run a standard mediation analysis in the framework of @baron1986moderator.
-For each polygenic score where the bivariate correlation with fertility is 
+For each polygenic score where the bivariate correlation with RLRS is 
 significant at $p$ < 0.05/33, we estimate 
 \begin{align}
-N_i & = \alpha + \beta PGS_i + \gamma EA_i + X_i\mu + \varepsilon_i \\
+RLRS_i & = \alpha + \beta PGS_i + \gamma EA_i + X_i\mu + \varepsilon_i \\
 EA_i & = \delta + \zeta PGS_i + X_i\mu + \eta_i
 \end{align}
-where $N_i$ is fertility, $PGS_i$ is the polygenic score, $EA_i$ is educational
+where $RLRS_i$ is fertility, $PGS_i$ is the polygenic score, $EA_i$ is educational
 attainment (age of leaving fulltime education), and $X_i$ is a vector of

### 2022-03-28 response-to-reviewers-BE.Rmd
Update response to reviewers; correct paper to talk about RLRS not fertility in appendix

@@
 
-Code to produce this paper is now publicly available at <https://github.com/hughjonesd/why-natural-selection>.
+We've also made some other changes:
+
+-   We moved the results on causality of polygenic scores to the appendix. We thought they were a distraction from the key point of Section 4, which is to test our theory.
+
+-   Code to reproduce this paper is now publicly available at <https://github.com/hughjonesd/why-natural-selection>.
 
@@
 This is related to the income effect in our theory.
-We now mention this work in our conclusion.
+We now mention this work in our conclusion ("Strikingly, some of these results were predicted by Fisher (1930), pp.
+253-254.")
 
@@
 
-Indeed, we don't want to talk about policy in the paper - we think we do not yet know enough.
-For what it's worth, here are my personal thoughts as corresponding author.
-First, government schemes to encourage fertility have a poor track record, see e.g.
-Gauthier (2007).
-Second, my strong gut feeling is that it is wrong for governments to (bluntly put) institute breeding programs for their citizens.
-I wrote more informally on this topic [here](https://wyclif.substack.com/p/the-natural-selection-paper-part-822?s=w).
+Indeed, we don't want to talk about policy in the paper - we think we do not yet know enough about the underlying forces.
+The proposed policy sounds close to some things that already exist e.g.
+child benefit in the UK, which is a non-means-tested per-child benefit.
+For what it's worth, my personal view is that government schemes to encourage fertility have a poor track record, see e.g.
+Gauthier (2007), though see also the recent Bergsvig et al.
+(2021).
+I wrote more informally on the policy issue [here](https://wyclif.substack.com/p/the-natural-selection-paper-part-822?s=w).
 
@@
 We agree this is an important topic, but we are more focused on human capital and earnings than IQ specifically.
-(Of course the polygenic score for educational attainment does also predict IQ.) The negative fertility-income relationship, in developed countries, is a very well-established stylized fact, and we mention some papers that also show income is causal.
-We now mention two of the cited papers.
+(Of course the polygenic score for educational attainment also captures a portion of the polygenic effects on IQ.) The negative fertility-income relationship, in developed countries, is a very well-established stylized fact, and we mention some papers that also show income is causal.
+We now mention two of the cited papers: "A related literature shows negative correlations between IQ and fertility (e.g.
+Lynn and Van Court 2004; Reeve, Heeney, and Menie 2018)".
 
@@
 So we expect that any remaining ascertainment is also probably biasing our estimates towards zero.
-We now say this.
+We now say this: "There may be remaining sources of ascertainment bias after our weighting; if so, we expect that, like the sources of ascertainment we have controlled for, they probably bias our results towards zero."
 
@@
 
-We have changed vocabulary accordingly.
+We have changed vocabulary accordingly, using "stabilizing" and "disruptive" throughout.
 
@@
 We prefer to be "conservative" in controlling for it, in the sense that we'd rather bias the estimate indirect effect of EA towards zero than away from zero.
-We now mention this issue in the appendix.
+We now mention this issue in the appendix: "Note that controlling for fluid IQ is conservative, in the sense that IQ might also measure human capital; nevertheless, we control for it because fluid IQ might affect fertility decisions via other channels than human capital."
 
@@
 We accept point (6) in particular.
-On the other hand, we do state that this is a "back of the envelope calculation".
-We've had pushback from people saying "this is no big deal", so we want to show that at least under some assumptions, yes it might be.
-Note also that as we say:
-
-"If unmeasured PSEA correlates positively with $\varepsilon$, then the reliability coefficient $\lambda$ will be lower than calculated.
-However, the proportional effect of natural selection on the PSEA-income relationship will remain the same, so long as $\lambda$ does not differ between the two sets of weighted regressions."
-
-The within-family estimate should remove the worries about exogeneity of p\*, but they are estimated with a smaller sample.
-We agree that the estimate is high.
-The bootstrapped 95% confidence interval is from around 0% to 300%, which is not really informative, though it does suggest that genetic nurture is more likely to bias our results downwards than upwards.
+We had pushback from people saying "this is no big deal", so we want to show that at least under some assumptions, yes it might be.
+The within-family estimate should remove many of the worries about exogeneity of p\*, but they are estimated with a smaller sample.
+We agree that the estimate seems high, and indeed, the bootstrapped 95% confidence interval is from around 0% to 300%, which is not really informative, though it does suggest that genetic nurture is more likely to bias our results downwards than upwards.
 We've removed the discussion.
@@
 
-KTK are the standard conditions for constrained optimization (see e.g.
+KKT are the standard first order conditions for constrained optimization (see e.g.
 Sundaram 1996).
@@
 The Inada condition is a bit more specific to economics.
-We are now more explicit about what it means.
+We are now more explicit about what it means: "the Inada condition (that marginal utility of income grows without
+
+bound as income approaches zero, $\lim_{x\rightarrow0}u'(x)=\infty$ )".
 It can be derived in this case by differentiating $u$ to give $u'(y) = y^{-\sigma}$ which for $\sigma > 0$ approaches infinity as $y \rightarrow 0$.
+An intuitive interpretation is that if you have extremely low income, your need for income overrides other considerations.
 
@@
 Any theory of contemporary natural selection needs to explain why correlations with fertility are lower at higher levels of income, education, etc.
-Non-causal explanations would be fine, but theories which don't generate these correlations must be missing something.
+Non-causal explanations are fine, but theories which don't generate these correlations must be missing something.
 Note that Fisher, cited by Reviewer 1, used observed correlations quite like ours to differentiate between theories of the fertility-income relationship.
-Generally, significant correlations have often led to important theoretical literatures, even when the direction of causality is uncertain (e.g.
+Generally, significant correlations have often led to important theoretical literatures, even when the direction of causality is initially uncertain (e.g.
 education/wages in labour economics, or democracy/GDP levels in political economy).
@@
 We would like to have more mediators, but appropriate variables are hard to find.
-Earnings in first job reduces our N by a factor of 10, which makes results too noisy to be informative.
+Earnings in first job reduces our N by a factor of 10, and this makes results too noisy to be informative.
 As the reviewer suggests, current household income is very much endogenous not only to human capital, but also to previous fertility, which makes it inappropriate as a mediator.
@@
 
-XXX Probably get in touch with those guys with their own weighting.\
+We agree with this.
+Combining all the variables is hard: age at first live birth is only available for females, and some of the data comes from different sources.
+Our ideal option is to use the weights that are being developed by Sjoerd van Alten and coauthors.
+Their working paper is specifically dedicated to weighting UK Biobank, so it seems good to use the "industry standard" solution rather than rolling our own.
+We now use these weights throughout.
+XXX
 
@@
 We now use RLRS as the dependent variable for both respondents and parents.
-Results are not much changed, except there is now slightly more evidence for larger selection effects over time: 9 polygenic scores show a significant increase in effect size on respondents' RLRS when we median-split them by birth year.
+Beyond the obvious change in absolute effect sizes, results are not much changed, except there is now slightly more evidence for larger selection effects over time: 9 polygenic scores show a significant increase in effect size on respondents' RLRS when we median-split them by birth year.
+Probably this was masked when we used absolute fertility, by the fall in fertility over time.
 In turn, this makes us less sure that we can rule out that the welfare state is a relevant explanation for natural selection, so we have slightly qualified our comments on this topic.
@@
 We have adjusted this to 50 male/45 female as suggested.
-Results are little changed, except that EA3 is no longer undergoing diversifying selection in the respondents' generation at the p = 0.05/33 significance threshold.
+Results are little changed, except that EA3 is no longer undergoing disruptive/diversifying selection in the respondents' generation at the p = 0.05/33 significance threshold.
 
@@
 It is a fair point. We have rewritten accordingly: "$\sigma > 0$ measures the curvature of the utility function, i.e. the decline in marginal utility of income as income increases".\
-\
-OTHER COMMENTS AND SUGGESTIONS
 
+> OTHER COMMENTS AND SUGGESTIONS
+>
 > 8\.
@@
 
-We experimented a lot with our basic figures, and they may not be easy to improve.
+We experimented a lot with our basic figures.
 The graphs' key message is "effect sizes in one category are larger than others".
@@
 We prefer our tweaked version (and have used it in the revision), but we will bow to the reviewers or editor.
-For comparison, a version of the whole paper using the symbols plot is available [here](https://raw.githubusercontent.com/hughjonesd/why-natural-selection/alt-plots/why-natural-selection-weighted.pdf).
+For comparison, a version of the paper using the symbols plot is available [here](https://raw.githubusercontent.com/hughjonesd/why-natural-selection/alt-plots/why-natural-selection-weighted.pdf).
 
@@
 
+Bergsvik, J., Fauske, A.
+and Hart, R.K., 2021.
+Can Policies Stall the Fertility Fall?
+A Systematic Review of the (Quasi‐) Experimental Literature.
+*Population and Development Review*, *47*(4), pp.913-964.
+
 Chuang, Y.

### 2022-03-28 why-natural-selection.Rmd
Update response to reviewers; correct paper to talk about RLRS not fertility in appendix

@@
 \end{align}
-where $RLRS_i$ is fertility, $PGS_i$ is the polygenic score, $EA_i$ is educational
-attainment (age of leaving fulltime education), and $X_i$ is a vector of
-controls. The total effect of $PGS$ on number of children is $\beta + \gamma
-\zeta$. The "indirect effect" mediated by $EA$ is $\gamma \zeta$. The standard
-error of the indirect effect can be calculated as
+where $RLRS_i$ is relative lifetime reproductive success, $PGS_i$ is the
+polygenic score, $EA_i$ is educational attainment (age of leaving fulltime
+education), and $X_i$ is a vector of controls. The total effect of $PGS$ on
+number of children is $\beta + \gamma \zeta$. The "indirect effect" mediated by
+$EA$ is $\gamma \zeta$. The standard error of the indirect effect can be
+calculated as
 

### 2022-05-03 why-natural-selection.Rmd

@@
     ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
-                     fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
+               shape = {{fill_col}}, ...)) +
     my_vline + 
     confint_geom_segment +
-          pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
-          pval_scale() +
-          my_fill_scale(n = n_cats, direction = fill_direction) +
+    geom_point(size = 1.8, aes(color = {{fill_col}})) +
+    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/33)) +
+    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction) +
+    scale_shape_manual(values = 1:n_cats + 14) +
     my_labs() +
     my_theme() +
-          scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1))
-
+    scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
+    scale_alpha_discrete(range = c(1, 0),
+                         guide = guide_legend(
+                           override.aes = list(
+                             shape = c(1, 16),
+                             color = "black",
+                             alpha = 1,
+                             size = 1.8
+                           )
+                         ))
 }

### 2022-05-03 response-to-reviewers-BE.Rmd
trivia
Fix a bug in run_regs_fml(); numerous tweaks to text

@@
 
--   Code to reproduce this paper is now publicly available at <https://github.com/hughjonesd/why-natural-selection>.
+-   Code is now publicly available at <https://github.com/hughjonesd/why-natural-selection>.
 
@@
 For a useful discussion of microeconomic theories of fertility, including functional forms, see Jones et al.
-(2008).
+(2008) and more recently Doepke et al.
+(2022) which uses a very similar form to ours to capture birth timing.
 
@@
 Combining all the variables is hard: age at first live birth is only available for females, and some of the data comes from different sources.
-Our ideal option is to use the weights that are being developed by Sjoerd van Alten and coauthors.
+In the new version, we use weights kindly provided to us by Sjoerd van Alten and coauthors.
 Their working paper is specifically dedicated to weighting UK Biobank, so it seems good to use the "industry standard" solution rather than rolling our own.
-We now use these weights throughout.
-XXX
+We now use these weights throughout, but detail results from alternative weighting schemes in the appendix.
 
@@
 We now use RLRS as the dependent variable for both respondents and parents.
-Beyond the obvious change in absolute effect sizes, results are not much changed, except there is now slightly more evidence for larger selection effects over time: 9 polygenic scores show a significant increase in effect size on respondents' RLRS when we median-split them by birth year.
+Beyond the obvious change in absolute effect sizes, results are not much changed, except there is now slightly more evidence for larger selection effects over time: 8 polygenic scores show a significant increase in effect size on respondents' RLRS when we median-split them by birth year.
 Probably this was masked when we used absolute fertility, by the fall in fertility over time.
@@
 
+Doepke, M., Hannusch, A., Kindermann, F. and Tertilt, M., 2022. *The economics of fertility: A new era* (No. w29948). National Bureau of Economic Research.
+
 Franz, V.H., 2007. Ratios: A short guide to confidence limits and proper use. *arXiv preprint arXiv:0710.2024*.

### 2022-05-03 why-natural-selection.Rmd
Fix a bug in run_regs_fml(); numerous tweaks to text

@@
 
-To correct for ascertainment bias, we weight participants using population data.
-We try three alternative weighting schemes: (1) "Geographical" weighting, by
-geography, age and presence/absence of a partner; (2) age and highest
-educational qualification; and (3) for women only, age, highest qualification,
-and age at first live birth (AFLB). Figure \@ref(fig:plot-res-wt) plots
-selection effects among the entire sample, estimated with the three weighting
-schemes, and also unweighted. Effects for `r res_sig_all` out of 33 polygenic
-scores are significant at $p < 0.05/33$ under all four weighting schemes, with
-`r res_sig_any` scores reaching significance under at least one scheme.[^balance-diverse-def]
-Weighting makes a large difference to estimates: mean effect sizes across all
-polygenic scores are increased by a factor of `r mean_rel$Geographical`
-(geographical weighting), `r mean_rel[["Age/Qualification"]]`
-(age/qualification) or `r mean_rel[["Age/Qual/AFLB"]]` (age/qualification/AFLB).
-Estimates might be further affected by weighting on other demographic
-variables.[^weighting]
+Figure \@ref(fig:plot-res-wt) plots selection effects among the entire
+sample.[^balance-diverse-def] To correct for ascertainment bias, we weight
+participants using the model of @van2021effects. Weighting makes a
+large difference to estimates, increasing effect sizes by a mean of 
+`r percent(mean_rel$Main - 1)`.[^weighting]
 
@@
       scale_fill_manual(values = c(
-        "None"              = "grey15", 
-        "Geographical"      = "dodgerblue4",
-        "Age/Qualification" = "steelblue2",
-        "Age/Qual/AFLB"      = "darkseagreen4"
+        "Unweighted"        = "grey15", 
+        "Weighted"          = "darkseagreen4"
       )) +
       scale_shape_manual(values = c(
-        "None" = "square filled", 
-        "Geographical" = "circle filled",
-        "Age/Qualification" = "circle filled", 
-        "Age/Qual/AFLB" = "circle filled"
+        "Unweighted" = "square filled", 
+        "Weighted"   = "circle filled"
       )) + 
@@
 which might plausibly be linked to externalizing behaviour, like ADHD and
-younger age at smoking initiation, are selected for. However, this theory is
-less good at explaining variation in selection across the full range of scores,
-including physical measures such as waist-hip ratio and BMI. We test this theory
+younger age at smoking initiation, are selected for. This theory is less good,
+though, at explaining variation in selection across the full range of scores,
+including physical measures such as waist-hip ratio and BMI. We test the theory
 directly by re-estimating equation \@ref(eq:regression) controlling for a
@@
 imprecision of the risk attitude measure, which is a single yes/no question.
-However, this measure *does* predict the overall number of children, highly
+But this measure *does* predict the overall number of children, highly
 significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of 33 regressions).
@@
 not expect that this should impact their association with number of offspring,
-or the resulting changes in allele frequencies. In short, it is probably too
-early to tell whether modern natural selection has a substantively important
-effect on population means of phenotypes under selection. Nevertheless, we note
-that selection effects on our measured polygenic scores are still relatively
-small, even after reweighting to account for ascertainment bias.
+or the resulting changes in allele frequencies. Lastly, although effects on
+our measured polygenic scores are small, even after weighting, disadvantages which
+are individually small may cumulate and/or interact to create larger effects. 
+Overall, it is probably too early to tell whether modern natural selection has a 
+substantively important effect on population means.
 
@@
 relatively consistent across the past two generations, but they are not the
-outcome of a universal, society-wide phenomenon. Instead they result from 
+outcome of a single, society-wide phenomenon. Instead they result from 
 opposing forces, operating in different parts of society and pulling in
@@
 
-Population data for weighting is taken from the 2011 UK Census and the 2006
-General Household Survey (GHS). Weighting for Age/Qualification and 
-Age/Qualification/AFLB weights was done using marginal totals from a linear
-model, using the `calibrate()` function in
-the R "survey" package [@lumley2020survey]. Geographical weighting was done with
-iterative post-stratification using the `rake()` function, on Census Middle Layer
-Super Output Areas, sex and presence/absence of a partner.
+Weighting data was kindly provided by @van2021effects.
 
@@
 times more children represented in UK Biobank than parents of one child. Parents
-of no children will by definition not be represented. To compensate, we reweight
-our preferred weightings (Age/Qualification) by the inverse of *number of
-siblings*.
+of no children will by definition not be represented. To compensate, we multiply
+our weights by the inverse of *number of siblings*.
 
@@
 driven by their *correlation* with human capital. Here, we test whether polygenic
-scores *cause* fertility by running within-siblings regressions (Figure
-\@ref(fig:plot-res-fe-fertility)). With a reduced sample size, all
-within-sibling effects are insignificant after Bonferroni correction. However,
-effect sizes are positively correlated with effect sizes from the pooled model,
-and about 50% smaller (regressing within-sibling on pooled effect sizes, $b$ =
-`r format(coef_pool_w, digits = 3)`). This attenuation is in line with the ~50%
-decrease in heritability in within-sibling GWASs on age at first birth and
-educational attainment [@Howe2021.03.05.433935]. We see these results as
-providing tentative evidence that polygenic scores cause fertility, with effects
-being partly driven by correlations with environmental variation in human
-capital. We also reran within-siblings regressions adding a control for
-education. Most effect sizes barely change, suggesting that our measure of
-education does not in general mediate differences in fertility among siblings.
+scores *cause* fertility by running within-siblings regressions. We run a single regression on 29 polygenic scores within `r n_groups` sibling
+groups (N = `r n`). Thus, we control both for environmental confounds (since
+scores are randomly allocated within sib-groups by meiosis), and for genetic
+confounds captured by our polygenic scores. We remove four scores which
+correlate highly with other scores (educational attainment 2, hip circumference,
+waist circumference and waist-hip ratio). Figure \@ref(fig:plot-res-fe-fertility)
+shows the results.
+
+With a reduced sample size, all within-sibling effects are insignificant after
+Bonferroni correction. However, effect sizes are positively correlated with
+effect sizes from the pooled model, and about 50% smaller (regressing
+within-sibling on pooled effect sizes, $b$ = 
+`r format(coef_pool_w, digits = 3)`). This attenuation is in line with the
+~50% decrease in heritability in
+within-sibling GWASs on age at first birth and educational attainment
+[@Howe2021.03.05.433935]. We see these results as providing tentative evidence
+that polygenic scores cause fertility, with effects being partly driven by
+correlations with environmental variation in human capital. We also reran
+within-siblings regressions adding a control for education. Most effect sizes
+barely change, suggesting that our measure of education does not in general
+mediate differences in fertility among siblings.
 
@@
 
-We run a single regression on 29 polygenic scores within `r n_groups` sibling
-groups (N = `r n`). Thus, we control both for environmental confounds (since
-scores are randomly allocated within sib-groups by meiosis), and for genetic
-confounds captured by our polygenic scores. We remove four scores which
-correlate highly with other scores (educational attainment 2, hip circumference,
-waist circumference and waist-hip ratio). Figure \@ref(fig:plot-res-fe-fertility)
-shows the results.
-
 \clearpage
@@
 
-```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores on number of siblings/children. Sibling estimates weighted by 1/number of siblings.", fig.height = 6}
+```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores. Parental generation weights multiplied by 1/number of siblings.", fig.height = 6}
 

### 2022-05-03 _drake.R

@@ -204,15 +204,27 @@ plan <- drake_plan(
           )
   },
   
+  res_unweighted = {
+    map_dfr(score_names, 
+            ~run_regs_fml(
+              "RLRS ~ {score_name}", 
+              score_name  = .x,
+              famhist     = famhist_kids,
+              use_weights = NULL
+            ),
+            .id = "score_name"
+    )
+  },
+  
   res_quadratic = {
     map_dfr(score_names, 
               ~run_regs_fml(
                              "RLRS ~ {score_name} + I({score_name}^2)", 
-                             score_name = .x,
-                             famhist    = famhist_kids,
+                             score_name  = .x,
+                             famhist     = famhist_kids,
                              # weight-everything: added weights
                              # for some reason these need to be made explicit
-                             weights    = quote(weights)
+                             use_weights = quote(weights)
                            ),
               .id = "score_name"
             )
@@ -222,9 +234,9 @@ plan <- drake_plan(
     map_dfr(score_names, 
             ~run_regs_fml(
               "RLRS_parents ~ {score_name} + I({score_name}^2)", 
-              score_name = .x,
-              famhist    = famhist_pw,
-              weights    = quote(weights)
+              score_name  = .x,
+              famhist     = famhist_pw,
+              use_weights = quote(weights)
             ),
             .id = "score_name"
     )
@@ -379,7 +391,7 @@ plan <- drake_plan(
             subset = quote(birth_order == 1),
             famhist = famhist_pw,
             # weight-everything: these are now parent_va_weights
-            weights = weights,
+            use_weights = quote(weights),
             .id = "id"
           )
     
@@ -407,9 +419,9 @@ plan <- drake_plan(
     res <-  map_dfr(score_names,
               ~run_regs_fml(
                 "RLRS_parents ~ Quin71 + {score_name}:Quin71",
-                score_name = .x,
-                famhist    = famhist_townsend_71,
-                weights    = weights
+                score_name  = .x,
+                famhist     = famhist_townsend_71,
+                use_weights = quote(weights)
               ),
               .id = "score_name"
             )
@@ -434,11 +446,11 @@ plan <- drake_plan(
           )
     res <- pmap_dfr(vars,
             run_regs_fml,
-            fml     = "{dep.var} ~ {score_name}",
-            subset  = quote(birth_order == 1),
-            famhist = famhist_pw,
[…]

### 2022-05-03 regressions.R

@@ -156,10 +156,10 @@ run_regs_subset <- function(score_name, subset, famhist) {
 }
 
 # weights added for weight-everything branch
-run_regs_fml <- function(fml, ..., subset = NULL, famhist, weights = quote(weights)) {
+run_regs_fml <- function(fml, ..., subset = NULL, famhist, use_weights = quote(weights)) {
   glue_args <- list(...)
   fml <- as.formula(glue::glue_data(fml, .x = glue_args))
-  mod <- lm(fml, famhist, subset = eval(subset), weights = eval(weights))
+  mod <- lm(fml, famhist, subset = eval(subset), weights = eval(use_weights))
   res <- tidy(mod, conf.int = TRUE)
   attr(res, "call") <- mod$call
   res

### 2022-05-03 why-natural-selection.Rmd
Tweak text

@@
 changes could reflect natural selection within the UK population, but also
-emigration, or ascertainment bias within the sample [@10.1093/aje/kwx246].
+emigration, or ascertainment bias in the sample [@10.1093/aje/kwx246].
 
@@
 
-Figure \@ref(fig:plot-res-wt) plots selection effects among the entire
+Figure \@ref(fig:plot-res-wt) plots selection effects in the entire
 sample.[^balance-diverse-def] To correct for ascertainment bias, we weight
 participants using the model of @van2021effects. Weighting makes a
-large difference to estimates, increasing effect sizes by a mean of 
-`r percent(mean_rel$Main - 1)`.[^weighting]
+large difference: effect sizes go up by a mean of 
+`r percent(mean_rel$Main - 1)`.[^weighting] `n_sig` out of 33 selection effects
+are significant at $p$ < 0.05/33.
 
@@
 to controlling for respondents' age (Appendix section \@ref(sec:age-control)).
+Turning to family structure, we split respondents by lifetime number of sexual
+partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
+Now, selection effects are larger and more significant among
+those with more than 3 lifetime partners. Next we split respondents by whether
+they were living with a spouse or partner at the time of interview.
+Effects are larger among those not living with a spouse or partner
+(Figure \@ref(fig:plot-n-partners)b).[^sex]
+
+[^sex]: The same pattern holds if we analyse men and women separately (Appendix Figure
+\@ref(fig:plot-n-partners-sex)). We also directly compared selection effects 
+between men and women (Appendix Figure \@ref(fig:plot-sexes)).
 
@@
 ```
-
-Turning to family structure, we split respondents by lifetime number of sexual
-partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
-Selection effects are larger and more significant among
-those with more than 3 lifetime partners. Next we split respondents by whether
-they were living with a spouse or partner at the time of interview.
-Effects are larger among those not living with a spouse or partner
-(Figure \@ref(fig:plot-n-partners)b).[^sex]
-
-[^sex]: The same pattern holds if we analyse men and women separately (Appendix Figure
-\@ref(fig:plot-n-partners-sex)). We also directly compared selection effects 
-between men and women (Appendix Figure \@ref(fig:plot-sexes)).
 
@@
 
-We emphasize that our categories are not necessarily exogenous to polygenic
-scores. For example -- both in the data (Appendix Figure
-\@ref(fig:plot-age-flb-dv)) and in our model -- age at first live birth is a
-choice variable, which is endogenous to an individual's human capital and
-related polygenic scores. Nevertheless, observed differences in equation
-\@ref(eq:regression) across subgroups constrain the set of possible
-theories; a good theory of contemporary natural selection needs to explain
-how these results arise. As we show below, a model based on the economic
-theory of fertility can do just that.
+We emphasize that our categories are not exogenous to polygenic scores. For
+example -- both in the data (Appendix Figure \@ref(fig:plot-age-flb-dv)) and in
+our theoretical model -- age at first live birth is a choice variable, which is
+endogenous to a person's human capital and their related polygenic scores.
+Nevertheless, differences in selection effects across subgroups constrain the
+set of possible theories; a good theory of contemporary natural selection needs
+to explain how these results come about. As we show below, a model based on the
+economic theory of fertility can do just that.
 
@@
 "quantity-quality tradeoff" between parents' number of children and number of
-grandchildren. We do not find any: in fact, the correlation between respondents' 
+grandchildren. We don't find any: in fact, the correlation between respondents' 
 and parents' RLRS is positive ($\rho$ = `r sib_chn_cor$estimate`, 

### 2022-05-04 why-natural-selection.Rmd
Many tweaks to text. Weight mediation bootstraps.

@@
 The direction of natural selection is reversed among older parents, or after controlling for age at first live birth.
-These patterns are in line with economic theories of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market.
+These patterns are in line with the economic theory of fertility, in which earnings-increasing human capital may either increase or decrease fertility via income and substitution effects in the labour market.
 Studying natural selection can help us understand the genetic architecture of health outcomes: we find evidence in modern day Great Britain for multiple natural selection pressures that vary between subgroups in the direction and strength of their effects, that are strongly related to the socio-economic system, and that may contribute to health inequalities across income groups."
@@
         ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
-               shape = {{fill_col}}, ...)) +
+                     fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
           my_vline + 
           confint_geom_segment +
-    geom_point(size = 1.8, aes(color = {{fill_col}})) +
-    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/33)) +
-    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction) +
-    scale_shape_manual(values = 1:n_cats + 14) +
+          pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
+          pval_scale() +
+          my_fill_scale(n = n_cats, direction = fill_direction) +
           my_labs() +
           my_theme() +
-    scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
-    scale_alpha_discrete(range = c(1, 0),
-                         guide = guide_legend(
-                           override.aes = list(
-                             shape = c(1, 16),
-                             color = "black",
-                             alpha = 1,
-                             size = 1.8
-                           )
-                         ))
+          scale_linetype_manual(name = "", 
+                                  values = c("95% c.i. uncorrected" = 1)) +
+          guides(
+            fill     = guide_legend(order = 1),
+            stroke   = guide_legend(order = 2),
+            linetype = guide_legend(order = 3)
+          )
 }
@@
 
-```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank.
+```{r plot-means-over-time, fig.height = 7, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank.
 Circles are means for 5-year intervals.
 Lines are 95% confidence intervals.
-Solid circles show a significant linear increase (green) or decrease (red) over time (p < 0.05/33)."}
+Triangles show a significant linear increase or decrease over time (p < 0.05/33)."}
 
@@
                        size = 3) +
-      geom_point(aes(y = score), size = 2) +
-      facet_wrap(~score_name, ncol = 6) +
+      geom_point(aes(y = score), size = 1.5) +
+      facet_wrap(~score_name, ncol = 5) +
       my_hline +
       scale_color_manual(values = c("-" = "darkred", "+" = "darkgreen", "o" = "black")) +
-      scale_shape_manual(values = c("-" = "circle", "+" = "circle", "o" = "circle open")) +
+      scale_shape_manual(values = c(
+                                      "-" = "triangle down open", 
+                                      "+" = "triangle open", 
+                                      "o" = "circle open"
+                                    )) +
       theme(
@@
 
-To test more directly for natural selection, we regress respondents' relative 
-lifetime reproductive success (RLRS) on their polygenic scores (PGS):
+To test for natural selection more directly, we regress respondents' relative 
+lifetime reproductive success (RLRS) on each polygenic score (PGS):
 
@@
 
-Figure \@ref(fig:plot-res-wt) plots selection effects in the entire
-sample.[^balance-diverse-def] To correct for ascertainment bias, we weight
-participants using the model of @van2021effects. Weighting makes a
+Figure \@ref(fig:plot-res-wt) plots selection effects in the whole
+sample.[^balance-diverse-def] To correct for ascertainment bias, we use
+participant weights from @van2021effects, which match the UK Biobank eligible
+population on sex, birth year, location, education, employment, health,
+household size and tenure, number of cars and age at death. Weighting makes a
 large difference: effect sizes go up by a mean of 
-`r percent(mean_rel$Main - 1)`.[^weighting] `n_sig` out of 33 selection effects
-are significant at $p$ < 0.05/33.
+`r scales::percent(mean_rel$Main - 1)`.[^weighting] `r n_sig` out of 33 weighted
+selection effects are significant at $p$ < 0.05/33.
 
@@
       ) %>% 
-      rename(p = p.value) %>% 
-      # can't use standard_ggplot because we have non-standard geom_point()
-      ggplot(aes(
-        x      = estimate, 
-        y      = term, 
-        yend   = term,
-        shape  = Weights, 
-        fill   = Weights, 
-        colour = p < 0.05/{{n_regs}}
-      )) +
-      my_vline +
-      geom_point(size = 1.6, stroke = 0.8, alpha = 0.9) +
-      confint_geom_segment() +
-      pval_scale() +
-      scale_fill_manual(values = c(
-        "Unweighted"        = "grey15", 
-        "Weighted"          = "darkseagreen4"
-      )) +
-      scale_shape_manual(values = c(
-        "Unweighted" = "square filled", 
-        "Weighted"   = "circle filled"
-      )) + 
-      labs(x = "Effect size", y = "") +
-      scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
-      my_theme()
+      standard_ggplot(fill_col = Weights, n_regs = n_regs)
 
@@
 concerns differences in the strength of natural selection across different
-subgroups in the sample. Splitting the sample by demographic and social
-variables, we re-estimate \@ref(eq:regression) within each subgroup. We use
-socio-economic variables including income and education, and family structure
-variables including age at first live birth, presence of a partner, and lifetime
-number of sexual partners.
+subgroups in the sample. We re-estimate \@ref(eq:regression) splitting the
+sample by demographic and social variables, including income and education, and
+family structure variables including age at first live birth, presence of a
+partner, and lifetime number of sexual partners.
+
+```{r calc-pct-diffs}
+drake::loadd(res_edu)
+drake::loadd(res_income)
+drake::loadd(res_partners_joint)
+drake::loadd(res_with_partner)
+res_partners_joint %<>% filter(grepl(":", term))
+res_with_partner %<>% filter(grepl(":", term))
+
+res_edu_wide <- full_join(
+                  res_edu %>% filter(age_fte_cat == "< 16"),
+                  res_edu %>% filter(age_fte_cat == "> 18"),
+                  by = "score_name"
+                )
+
+res_income_wide <- full_join(
+                     res_income %>% filter(income_cat == 1),
+                     res_income %>% filter(income_cat == 5),
+                     by = "score_name"
+                   )
+
+res_partners_wide <- full_join(
+                       res_partners_joint %>% filter(grepl("lo_partnersFALSE", term)),
+                       res_partners_joint %>% filter(grepl("lo_partnersTRUE", term)),
+                       by = "score_name"
+                     )
+
+res_with_partner_wide <- full_join(
+                           res_with_partner %>% 
+                             filter(grepl("with_partnerFALSE", term)),
+                           res_with_partner %>% 
+                             filter(grepl("with_partnerTRUE", term)),
+                           by = "score_name"
+                         )
+
+mean_pct_diff <- function (dfr) {
+   prop <- dfr %>% filter(
+                    p.value.x < 0.05/33,
+                    sign(estimate.x) == sign(estimate.y)
+                  ) %>% 
+                  mutate(
+                    pct_diff = estimate.x/estimate.y
+                  ) %>% 
+                  pull(pct_diff) %>% 
+                  mean()
+   # percentage difference:
+   prop - 1
+}
+
+pct_diff_edu <- mean_pct_diff(res_edu_wide)
+pct_diff_income <- mean_pct_diff(res_income_wide)
+pct_diff_n_partners <- mean_pct_diff(res_partners_wide)
+pct_diff_with_partner <- mean_pct_diff(res_with_partner_wide)
+```
 
@@
 and by household income. Effects are larger and more significant for the lowest
-income category, and for the lowest education category. These results are robust
-to controlling for respondents' age (Appendix section \@ref(sec:age-control)).
-Turning to family structure, we split respondents by lifetime number of sexual
-partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
-Now, selection effects are larger and more significant among
-those with more than 3 lifetime partners. Next we split respondents by whether
-they were living with a spouse or partner at the time of interview.
-Effects are larger among those not living with a spouse or partner
-(Figure \@ref(fig:plot-n-partners)b).[^sex]
+education category, and for the lowest income category. The mean percentage
+difference between the lowest and highest education categories, among scores
+which are significant for the lowest category and have the same sign across
+categories, is `r scales::percent(pct_diff_edu)`. Between the lowest and
+highest income categories, it is `r scales::percent(pct_diff_income)`. These
+results are robust to controlling for respondents' age (Appendix section
+\@ref(sec:age-control)). Turning to family structure, we split respondents by
+lifetime number of sexual partners, at the median value of 3 (Figure
+\@ref(fig:plot-n-partners)a). Now, selection effects are larger and more
+significant among those with more than 3 lifetime partners, with a mean percentage
+difference of `r scales::percent(pct_diff_n_partners)`. Next we split
+respondents by whether they were living with a spouse or partner at the time of
+interview (Figure \@ref(fig:plot-n-partners)b).  Effects are larger among those not living with a spouse or partner. The mean percentage difference is 
+`r scales::percent(pct_diff_with_partner)`.[^sex]
 
@@
 
-We emphasize that our categories are not exogenous to polygenic scores. For
+We emphasize that these categories are not exogenous to polygenic scores. For
 example -- both in the data (Appendix Figure \@ref(fig:plot-age-flb-dv)) and in
-our theoretical model -- age at first live birth is a choice variable, which is
-endogenous to a person's human capital and their related polygenic scores.
-Nevertheless, differences in selection effects across subgroups constrain the
-set of possible theories; a good theory of contemporary natural selection needs
-to explain how these results come about. As we show below, a model based on the
-economic theory of fertility can do just that.
+our theoretical model -- education and age at first live birth are choice
+variables, which are endogenous to a person's human capital and to relevant
+polygenic scores. Nevertheless, differences in selection effects across
+subgroups constrain the set of possible explanations. A good theory of contemporary
+natural selection needs to show how these differences come about. As we describe
+below, a model based on the economic theory of fertility can do just that.
 
@@
 We also check whether selection effects vary by
-socio-economic status and AFLB in the parents' generation. We have no
-information about parents' income, so we use the 1971 Townsend deprivation score
-of respondents' birthplace as a proxy [@townsend1987deprivation]. Results show
-the same pattern as for respondents. Effect sizes are larger and more often
-significant in the most deprived areas (Appendix Figure
-\@ref(fig:plot-siblings-townsend)). Effects are larger among younger parents,
-and change sign when controlling for AFLB, among both mothers and fathers
-(Appendix Figures \@ref(fig:plot-age-birth-parents-cross), 
-\@ref(fig:plot-age-birth-parents)). Lastly, we check for a
-"quantity-quality tradeoff" between parents' number of children and number of
-grandchildren. We don't find any: in fact, the correlation between respondents' 
-and parents' RLRS is positive ($\rho$ = `r sib_chn_cor$estimate`, 
-$p < 2 \times 10^{-16}$).
+AFLB and socio-economic status in the parents' generation, using the 1971 
+Townsend deprivation score of respondents' birthplace as a proxy for income
+[@townsend1987deprivation]. Results show the same pattern as for the
+respondents' generation. Effect sizes are larger and more often significant in
+the most deprived areas (Appendix Figure \@ref(fig:plot-siblings-townsend)).
+Effects are larger among younger fathers and mothers, and change sign when
+controlling for AFLB (Appendix Figures \@ref(fig:plot-age-birth-parents-cross),
+\@ref(fig:plot-age-birth-parents)). Lastly, we check for a "quantity-quality
+tradeoff" between parents' number of children and number of grandchildren. We
+don't find any: in fact, the correlation between respondents' and parents' RLRS
+is positive ($\rho$ = `r sib_chn_cor$estimate`, $p < 2 \times 10^{-16}$).
 
@@
 increase in human capital on this marginal benefit is
-
 \[
-\frac{d^{2}U}{dNdh}=\underbrace{-bu'(Y)}_{\textrm{Substitution effect}}\underbrace{-bYu''(Y)}_{\textrm{Income effect}}.
+\frac{d^{2}U}{dNdW}=\underbrace{-bu'(Y)}_{\textrm{Substitution effect}}\underbrace{-bYu''(Y)}_{\textrm{Income effect}}.
 \]
-
 The *substitution effect* is negative and reflects that when wages increase,
@@
 \end{equation}
-
 Instead of identifying human capital with wages, we now allow individuals to
-choose a level of education $s \in [0,1]$, which has a time cost in period 1.
+spend time $s \in [0,1]$ on education in period 1.
 Education is complementary to human capital $h > 0$, and increases period 2
@@
 $N^{*} = N_{1}^{*} + N_{2}^{*}$ and the *fertility-human capital relationship*,
-$\frac{dN^{*}}{dh}$. For $\sigma < 1$ in a neighbourhood of $\sigma = 1$, Table
+$\frac{dN^{*}}{dh}$. For $\sigma < 1$ and close enough to 1, Table
 \@ref(tab:theory) shows five theoretical predictions, along with our
@@
 levels of $\sigma$, the substitution effect dominates at low income levels, but
-is balanced by the income effect as income increases.
+as income increases, the income and substitution effect balance out.
 
@@
 US fertility decreases faster with education among single mothers than married
-mothers [@baudin2015fertility], in line with our prediction 3 and as predicted by
-@becker1981treatise. A related literature shows negative correlations
-between IQ and fertility [e.g. @lynn2004new; @reeve2018systematic].
+mothers [@baudin2015fertility], in line with our prediction 3 and as predicted
+by @becker1981treatise. A related literature shows negative correlations between 
+IQ and fertility [e.g. @lynn2004new; @reeve2018systematic].
 
@@
 welfare benefits which incentivize child-bearing might be taken up more among
-low-income individuals. However, the majority of effect sizes appear unchanged
+low-income people. However, the majority of effect sizes appear unchanged
 over a large span of twentieth-century history (Appendix Table
@@
 [@hoc1999socsec]. In general, there is only weak evidence that welfare benefits
-affect fertility [@gauthier2007impact]. Future
+affect fertility [@gauthier2007impact; see also @bergsvik2021can]. Future
 work could test this theory more explicitly. A second alternative theory is that
@@
 our parameter $\sigma$, which can be interpreted as a measure of risk aversion
-over income. However, a more direct channel is risky sexual behaviour
-[@mills2021identification]. The data here provide some support for this: scores
+over income; a more direct channel is risky sexual behaviour
+[@mills2021identification]. The data here provide some support for this story: scores
 which might plausibly be linked to externalizing behaviour, like ADHD and
-younger age at smoking initiation, are selected for. This theory is less good,
-though, at explaining variation in selection across the full range of scores,
-including physical measures such as waist-hip ratio and BMI. We test the theory
-directly by re-estimating equation \@ref(eq:regression) controlling for a
-measure of risk attitude (UK Biobank field 2040).  The median ratio of effect
-sizes between regressions with and without controls is `r median(ratio)`; all
-scores which are significant at $p < 0.05/33$ in uncontrolled regressions remain
-so when controlling for risk attitude. This non-result could simply reflect the
-imprecision of the risk attitude measure, which is a single yes/no question.
-But this measure *does* predict the overall number of children, highly
-significantly ($p < 2 \times 10^{-16}$ in `r n_risk_sig` out of 33 regressions).
-Given that, and the statistical power we get from our sample size, we believe
-that the non-result is real: while risk attitude does predict fertility in the
-sample, it is not an important channel for natural selection.
+younger age at smoking initiation, are selected for. However, risk-seeking 
+seems unlikely to explain variation in fertility across the full range of scores
+under selection, including physical measures like waist-hip ratio and BMI. We
+test this theory directly by re-estimating equation \@ref(eq:regression)
+controlling for a measure of risk attitude (UK Biobank field 2040). The median
+ratio of effect sizes between regressions with and without controls is 
+`r median(ratio)`; all scores which are significant at $p < 0.05/33$ in
+uncontrolled regressions remain so when controlling for risk attitude. This
+non-result could simply reflect the imprecision of the risk attitude measure,
+which is a single yes/no question. But this measure *does* predict the overall
+number of children, highly significantly ($p < 2 \times 10^{-16}$ in 
+`r n_risk_sig` out of 33 regressions). Given that, and the statistical power we
+get from our sample size, we believe that the non-result is real: while risk
+attitude does predict fertility in the sample, it is not an important channel
+for natural selection.
 
@@
 Potentially, this could increase the health burden on modern populations, but
-that depends on effect sizes. Our results suggest that naïve estimates can be
+that depends on effect sizes. Our results show that naïve estimates can be
 affected by sample ascertainment bias. There may be remaining sources of
@@
 of ascertainment we have controlled for, they probably bias our results towards
-zero. This problem may be less serious in surveys which aim to be representative
-(as the UK Biobank does not). However, there is still scope for bias, since not
-all respondents consent to the collection of genetic data. For instance,
-completion rates for genotype data in the US Health and Retirement Study were
-around 80-85% [@hrs2020]. Researchers should be aware of the risks of
-ascertainment when studying modern natural selection.
+zero. Researchers should be aware of the risks of ascertainment when studying 
+modern natural selection.
 
 We also do not know how estimated effect sizes of natural selection will change
-as more accurate measures of genetic variation are produced, or whether genetic
+as more accurate polygenic scores are produced, or whether genetic
 variants underlying other phenotypes will show a similar pattern to those
-studied here. In addition, effects of polygenic scores may be inflated in
+studied here. Also, effects of polygenic scores may be inflated in
 population-based samples, because of indirect genetic effects, gene-environment
@@
 @selzam2019comparing; @kong2018nature; @Howe2021.03.05.433935], although we do
-not expect that this should impact their association with number of offspring,
-or the resulting changes in allele frequencies. Lastly, although effects on
-our measured polygenic scores are small, even after weighting, disadvantages which
-are individually small may cumulate and/or interact to create larger effects. 
-Overall, it is probably too early to tell whether modern natural selection has a 
-substantively important effect on population means.
+not expect that this should change their association with number of offspring,
+or the resulting changes in allele frequencies. Although effects on
+our measured polygenic scores are small even after weighting, individually small
+disadvantages can cumulate to create larger effects. Lastly, note that our
+data comes from people born before 1970. Recent evidence suggests that fertility
+patterns may be changing [@doepke2022economics]. Overall, it is probably 
+too early to tell whether modern natural selection has a substantively important 
+effect on population averages of phenotypes under selection.
 
@@
 Overall, natural selection increases the correlation of polygenic scores with
-income for `r n_increases` out of 33 polygenic scores, with a median
-percentage increase of `r median_pct_change`% (Appendix Table
-\@ref(tab:tbl-inequality-effects)).  If inequalities in polygenic scores are
-important for understanding social structure and mobility 
+income for `r n_increases` out of 33 polygenic scores, with a median percentage
+increase of `r median_pct_change`% in the respondents' generation (Appendix
+Table \@ref(tab:tbl-inequality-effects)).  If inequalities in polygenic scores
+are important for understanding social structure and mobility
 [@belsky2018genetic; @Rimfeld_2018; @harden2021genetic], then these increases
@@
 risk [@benton2021influence]; understanding the role of contemporary natural
-selection may aid in mapping the genetic architecture of current health
+selection may help researchers to map the genetic architecture of health
 disparities.
@@
 pessimism" [@sarraf2019modernity]. Others are more sanguine, and argue that
-natural selection's effects are outweighed by environmental improvements, such
-as those underlying the Flynn effect [@flynn1987massive]. The evidence
+natural selection's effects are outweighed by environmental improvements, like
+those underlying the Flynn effect [@flynn1987massive]. The evidence
 here may add some nuance to this debate. Patterns of natural selection have been
@@
 
+```{r plot-res-fe-fertility, fig.cap = "Selection effects controlling for sibling-group fixed effects, with and without a control for education (left education before 16, 16-18, or after 18).
+Each set of 29 results is from a single regression of RLRS on 29 polygenic scores.
+Standard errors clustered by sibling group."}
+
+drake::loadd(res_fe_fertility)
+n_regs <- nrow(res_fe_fertility)/2 # Raw and Controlled treated separately
+
+standard_ggplot(res_fe_fertility, fill_col = Regression, n_regs = n_regs,
+                  fill_direction = -1)
+
+n_groups <- attr(res_fe_fertility, "n_groups")
+n <- attr(res_fe_fertility, "n")
+
+```
+
 Results in the main text support our theory that natural selection on polygenic scores is
@@
 mediate differences in fertility among siblings.
-
-```{r plot-res-fe-fertility, fig.cap = "Selection effects controlling for sibling-group fixed effects, with and without a control for education (left education before 16, 16-18, or after 18).
-Each set of 29 results is from a single regression of RLRS on 29 polygenic scores.
-Standard errors clustered by sibling group."}
-
-drake::loadd(res_fe_fertility)
-n_regs <- nrow(res_fe_fertility)/2 # Raw and Controlled treated separately
-
-standard_ggplot(res_fe_fertility, fill_col = Regression, n_regs = n_regs,
-                  fill_direction = -1) +
-  guides(alpha = guide_legend(
-                   override.aes = list(
-                             shape = 1,
-                             color = "black",
-                             alpha = 1,
-                             size = 1.8
-                   )
-                 )
-  )
-
-n_groups <- attr(res_fe_fertility, "n_groups")
-n <- attr(res_fe_fertility, "n")
-
-```
 

### 2022-05-04 regressions.R

@@ -433,9 +433,15 @@ run_mediation <- function (famhist, res_all) {
   res <- purrr::map_dfr(sig_scores, run_one_mediation, famhist = famhist)
   
   # bootstraps
+  famhist <- famhist %>% filter(!is.na(weights))
   boot_res <- purrr::map_dfr(1:100, function (r) {
                               fh_boot <- famhist %>% 
-                                           slice_sample(prop = 1, replace = TRUE)
+                                           slice_sample(
+                                             prop = 1, 
+                                             replace = TRUE,
+                                             # weight-everything: weights added
+                                             weight_by = famhist$weights
+                                           )
                               br <- purrr::map_dfr(sig_scores, 
                                                    run_one_mediation, famhist = fh_boot)
                               br <- br[c("term", "estimate_total", "estimate_ind")]

### 2022-05-04 response-to-reviewers-BE.Rmd
More text tweaks. No need for picture in response to reviewers.

@@
 
-We experimented a lot with our basic figures.
-The graphs' key message is "effect sizes in one category are larger than others".
-To see that the eye needs to be able to quickly differentiate between categories.
-We have tried several variants using symbols, but in our judgment the added visual noise makes it harder to follow the categories across rows.
-The figure below shows three alternative versions: our original, a version using symbols, and one with subtle tweaks to colors and stroke sizes, which we hope fix the greenness issue and make p-values clearer.
-We prefer our tweaked version (and have used it in the revision), but we will bow to the reviewers or editor.
-For comparison, a version of the paper using the symbols plot is available [here](https://raw.githubusercontent.com/hughjonesd/why-natural-selection/alt-plots/why-natural-selection-weighted.pdf).
-
-We've worked on Figures 1 and 2 to improve readability.
-
-```{r different-plots, echo = FALSE, fig.cap = "Alternative plot styles", fig.subcap = c("Original", "Symbols", "Tweaked"), fig.ncol = 1, fig.height=2.5, fig.align="center"}
-
-res_edu <- res_edu %>% filter(grepl("^[a-o]", term))
-n_regs <- nrow(res_edu)
-
-standard_ggplot(res_edu, n_regs = n_regs, fill_col = `Age left FTE`, fill_direction = -1)
-
-shapes_ggplot(res_edu, n_regs = n_regs, fill_col = `Age left FTE`, fill_direction = -1)
-
-tweaked_ggplot(res_edu, n_regs = n_regs, fill_col = `Age left FTE`, fill_direction = -1)
-```
+We have changed our figures to use both shape and colour for categories. We hope this is an improvement. We've also worked on Figure 1.
 

### 2022-05-04 why-natural-selection.Rmd
More text tweaks. No need for picture in response to reviewers.

@@
 ```{r plot-means-over-time, fig.height = 7, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank.
-Circles are means for 5-year intervals.
-Lines are 95% confidence intervals.
-Triangles show a significant linear increase or decrease over time (p < 0.05/33)."}
+Symbols show means for 5-year intervals.
+Bars are 95% confidence intervals.
+Triangles denote a significant linear increase or decrease over time (p < 0.05/33)."}
 
@@
 There is some evidence that selection effect sizes are increasing in the 
-respondents' generation, with 9 polygenic scores showing a significant increase. 
+respondents' generation, with 8 polygenic scores showing a significant increase. 
 We also check whether selection effects vary by

### 2022-05-04 why-natural-selection.Rmd
Really use medians for percentage differences

@@
     ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
-                     fill = {{fill_col}}, ..., colour = p < 0.05/{{n_regs}})) +
+               shape = {{fill_col}}, ...)) +
     my_vline + 
     confint_geom_segment +
-          pval_geom_point(fill = if(rlang::quo_is_null(fill_col)) "steelblue") +
-          pval_scale() +
-          my_fill_scale(n = n_cats, direction = fill_direction) +
+    geom_point(size = 1.8, alpha = 0.9, aes(color = {{fill_col}})) +
+    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/{{n_regs}})) +
+    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction) +
+    scale_shape_manual(values = 1:n_cats + 14) +
     my_labs() +
     my_theme() +
-          scale_linetype_manual(name = "", 
-                                  values = c("95% c.i. uncorrected" = 1)) +
-          guides(
-            fill     = guide_legend(order = 1),
-            stroke   = guide_legend(order = 2),
-            linetype = guide_legend(order = 3)
-          )
+    confint_scale +
+    scale_alpha_discrete(range = c(1, 0), 
+                           guide = guide_legend(
+                             order = 0,
+                             override.aes = list(
+                               shape = c(1, 16),
+                               color = "black",
+                               alpha = 1,
+                               size = 1.8
+                             )
+                         )) 
 }
@@
   geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i. uncorrected"), 
-                    color = "grey60", alpha = 0.25, size = 1.1, 
+                    color = "grey55", alpha = 0.2, size = 1.1, 
                     lineend = "round")
@@
       ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
-      geom_linerange(aes(ymin = score_lo, ymax = score_hi), color = "grey60", 
+      geom_linerange(aes(ymin = score_lo, ymax = score_hi), color = "grey70", 
                        size = 3) +
@@
 
-mean_pct_diff <- function (dfr) {
+median_pct_diff <- function (dfr) {
    prop <- dfr %>% filter(
@@
                   pull(pct_diff) %>% 
-                  mean()
+                  median()
    # percentage difference:
@@
 
-pct_diff_edu <- mean_pct_diff(res_edu_wide)
-pct_diff_income <- mean_pct_diff(res_income_wide)
-pct_diff_n_partners <- mean_pct_diff(res_partners_wide)
-pct_diff_with_partner <- mean_pct_diff(res_with_partner_wide)
+pct_diff_edu <- median_pct_diff(res_edu_wide)
+pct_diff_income <- median_pct_diff(res_income_wide)
+pct_diff_n_partners <- median_pct_diff(res_partners_wide)
+pct_diff_with_partner <- median_pct_diff(res_with_partner_wide)
 ```
@@
 and by household income. Effects are larger and more significant for the lowest
-education category, and for the lowest income category. The mean percentage
+education category, and for the lowest income category. The median percentage
 difference between the lowest and highest education categories, among scores
@@
 \@ref(fig:plot-n-partners)a). Now, selection effects are larger and more
-significant among those with more than 3 lifetime partners, with a mean percentage
+significant among those with more than 3 lifetime partners, with a median percentage
 difference of `r scales::percent(pct_diff_n_partners)`. Next we split
 respondents by whether they were living with a spouse or partner at the time of
-interview (Figure \@ref(fig:plot-n-partners)b).  Effects are larger among those not living with a spouse or partner. The mean percentage difference is 
+interview (Figure \@ref(fig:plot-n-partners)b).  Effects are larger among those not living with a spouse or partner. The median percentage difference is 
 `r scales::percent(pct_diff_with_partner)`.[^sex]
@@
 
-Polygenic scores which correlate with high (low) earnings and more (less)
-education are being selected against (selected for). In addition, many of the
-phenotypes under positive selection are linked to disease risk. Many people
-would probably prefer to have high educational attainment, a low risk of ADHD
-and major depressive disorder, and a low risk of coronary artery disease, but
-natural selection is pushing against genes associated with these traits.
-Potentially, this could increase the health burden on modern populations, but
-that depends on effect sizes. Our results show that naïve estimates can be
-affected by sample ascertainment bias. There may be remaining sources of
-ascertainment bias after our weighting; if so, we expect that, like the sources
-of ascertainment we have controlled for, they probably bias our results towards
-zero. Researchers should be aware of the risks of ascertainment when studying 
-modern natural selection.
+Polygenic scores which correlate with lower earnings and less education are
+being selected for. In addition, many of the phenotypes under positive selection
+are linked to disease risk. Many people would probably prefer to have high
+educational attainment, a low risk of ADHD and major depressive disorder, and a
+low risk of coronary artery disease, but natural selection is pushing against
+genes associated with these traits. Potentially, this could increase the health
+burden on modern populations, but that depends on effect sizes. Our results show
+that naïve estimates can be affected by sample ascertainment bias. There may be
+remaining sources of ascertainment bias after our weighting; if so, we expect
+that, like the sources of ascertainment we have controlled for, they probably
+bias our results towards zero. Researchers should be aware of the risks of
+ascertainment when studying modern natural selection.
 
@@
 res_partners %>% 
-      mutate(
-        score_name = pretty_names(score_name)
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(
-        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        x      = estimate, 
-        fill   = `N partners`, 
-        colour = p < 0.05/{{n_regs}}
-      )) + 
-        my_vline +
-        confint_geom_segment() +
-        pval_geom_point() +
-        facet_wrap(vars(sex), scales = "free_y") +
-        my_fill_scale(n = 2) +
-        pval_scale() +
-        sexes_theme +
-        my_labs(x = "", y = "") + 
-        scale_y_reordered() +
-        scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
-        sexes_guides
+      standard_ggplot(fill_col = `N partners`, score_col = score_name, 
+                        n_regs = n_regs) +
+        facet_grid(cols = vars(sex))
 
@@
       ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(
-        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
-        x      = estimate, 
-        fill   = Household, 
-        colour = p < 0.05/{{n_regs}}
-      )) + 
-        my_vline +
-        confint_geom_segment() +
-        pval_geom_point() +
-        facet_wrap(vars(sex), scales = "free_y") +
-        my_fill_scale(n = 2) +
-        pval_scale() +
-        sexes_theme +
-        my_labs(x = "", y = "") + 
-        scale_y_reordered() +
-        scale_linetype_manual(
-                                name   = "", 
-                                values = c("95% c.i. uncorrected" = 1)
-                              ) +
-        sexes_guides
+      standard_ggplot(fill_col = Household, score_col = score_name, 
+                        n_regs = n_regs) +
+        facet_grid(cols = vars(sex))
 
@@
 Bonferroni correction. However, effect sizes are positively correlated with
-effect sizes from the pooled model, and about 50% smaller (regressing
-within-sibling on pooled effect sizes, $b$ = 
-`r format(coef_pool_w, digits = 3)`). This attenuation is in line with the
-~50% decrease in heritability in
-within-sibling GWASs on age at first birth and educational attainment
+effect sizes from the pooled model, and about 70% smaller (regressing
+within-sibling on pooled effect sizes, $b$ = `r format(coef_pool_w, digits = 3)`).
+This attenuation is broadly consistent with the decrease in heritability
+in within-sibling GWASs on age at first birth and educational attainment
 [@Howe2021.03.05.433935]. We see these results as providing tentative evidence
@@
 
-```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores. Parental generation weights multiplied by 1/number of siblings.", fig.height = 6}
+```{r plot-regs-controlled-pcs, fig.cap = "Selection effects using unresidualized polygenic scores.
+Parental generation weights multiplied by 1/number of siblings.", fig.subcap = c("Respondents", "Parents"), fig.ncol = 1}
 
@@
 n_regs <- as.double(nrow(res_all))
+res_all %<>% mutate(
+               `PGS` = ifelse(reg.type == "controlled", 
+                                "Residualized for 100 PCs", "Unresidualized")
+             )
+
 res_all %>% 
-      mutate(
-        term     = pretty_names(term),
-        term     = reorder_within(term, estimate, dep.var, order_abs(1)),
-        dep.var  = fct_inorder(dep.var),
-        dep.var  = fct_recode(dep.var, 
-                                `Respondents' RLRS` = "RLRS",
-                                `Parents' RLRS` = "RLRS_parents" 
-                              ),
-        reg.type = fct_recode(reg.type, 
-                               `Resid. for 100 PCs`  = "controlled", 
-                               Unresidualized        = "raw"
-                             )
-      ) %>% 
-      rename(p = p.value) %>% 
-      ggplot(aes(y = term, x = estimate, fill = reg.type, 
-        colour = p < 0.05/{{n_regs}})) +
-        my_vline + 
-        facet_grid(rows = vars(dep.var), scales = "free") +
-        pval_geom_point(position = position_dodge(0.4), alpha = 0.8) +
-        my_labs() +
-        pval_scale() +
-        my_fill_scale(n = 2, guide = guide_legend(title = "")) +
-        my_theme() +
-        scale_y_reordered() +
-        theme(
-          axis.text.x  = element_text(size = 7),
-          legend.text  = element_text(size = 8),
-          legend.title = element_text(size = 9)
-        )
+      filter(dep.var == "RLRS") %>% 
+      standard_ggplot(fill_col = PGS, n_regs = n_regs)
+
+res_all %>% 
+      filter(dep.var == "RLRS_parents") %>% 
+      standard_ggplot(fill_col = PGS, n_regs = n_regs)
 

### 2022-05-04 why-natural-selection.Rmd
Final final tweaks to pictures.

@@
 options(digits = 2)
-options(dplyr.summarise.inform = FALSE) # shut up about summarize
+options(dplyr.summarise.inform = FALSE)
 
@@
 my_vline <- geom_vline(xintercept = 0, colour = "grey20", linetype = "dotted")
-pval_palette <- c("FALSE" = "grey65", "TRUE" = "#f3b372")
 
@@
   n_regs <- as.double(n_regs)
-
+  shape_values <- c(16, 15, 17, 18, 20)[1:n_cats]
+  if (fill_direction == -1) shape_values <- rev(shape_values)
   dfr %>% mutate(
@@
     my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction) +
-    scale_shape_manual(values = 1:n_cats + 14) +
+    scale_shape_manual(values = shape_values) +
     my_labs() +
@@
 
-pval_scale <- function (...) scale_color_manual(values = pval_palette, ...)
-
 pval_geom_point <- function (size = 1.6, stroke = 0.9, fill = NULL, ...) {
@@
 
-my_fill_scale <- function (n, aesthetics = "fill", ...) {
+my_fill_scale <- function (n, aesthetics = "fill", direction = 1, ...) {
 
@@
   } else {
-    scale <- scale_color_brewer(aesthetics = aesthetics, ...)
-    shades::lightness(scale, shades::scalefac(0.75))
+    # scale <- scale_color_viridis_d(aesthetics = aesthetics, option = "D",
+    #                                  direction = direction, end = 0.6, ...)
+    if (direction != -1) fill_colors <- rev(fill_colors)
+    scale_color_manual(aesthetics = aesthetics, 
+                         values = shades::gradient(fill_colors,
+                                                     steps = n), 
+                         ...) 
   }
@@
       ) %>% 
-      standard_ggplot(fill_col = Weights, n_regs = n_regs)
+      standard_ggplot(fill_col = Weights, n_regs = n_regs, fill_direction = -1)
 
@@
       ) %>% 
-      standard_ggplot(n_regs = n_regs, fill_col = `Age left FTE`, 
-                        fill_direction = -1)
+      standard_ggplot(n_regs = n_regs, fill_col = `Age left FTE`)
 
@@
       ) %>% 
-      standard_ggplot(n_regs = n_regs, fill_col = `Income`, 
-                        fill_direction = -1)
+      standard_ggplot(n_regs = n_regs, fill_col = `Income`)
 ```
@@
       standard_ggplot(fill_col = `N partners`, n_regs = n_regs, 
-                        score_col = score_name)
+                        score_col = score_name, fill_direction = -1)
 
@@
       standard_ggplot(fill_col = Household, n_regs = n_regs, 
-                        score_col = score_name)
+                        score_col = score_name, fill_direction = -1)
 ```
@@
       standard_ggplot(score_col = score_name, 
-                      fill_col = `Age at first live birth`, n_regs = n_regs,
-                      fill_direction = -1)
+                      fill_col = `Age at first live birth`, n_regs = n_regs)
 
@@
       mutate(
-        PGS      = ifelse(abs(effect_size) > 0.02, PGS, ""),
+        PGS      = ifelse(abs(effect_size) > 0.027, PGS, ""),
         PGS      = sub("(^\\w*?)_.*", "\\1", PGS)
@@
       ) %>% 
-      ggplot(aes(effect_size, cor, colour = Correlation)) + 
-        geom_point() +
+      ggplot(aes(effect_size, cor)) + 
+        geom_point(color = "steelblue4") +
         geom_text(aes(label = PGS), size = 3, colour = "black", 
@@
         coord_cartesian(clip = "off") +
-        scale_colour_brewer(type = "qual", palette = 2) +
         theme(
@@
                       ) +
-        scale_fill_manual(values = pval_palette) +
+        scale_fill_manual(values = c("FALSE" = "grey65", "TRUE" = "orange2")) +
         my_vline +
@@
                 geom_col(position = "dodge") +
-                scale_fill_manual(values = c("navy", "grey60")) +
+                scale_fill_manual(values = c("darkred", "grey65")) +
                 theme(
@@
 
-```{r plot-sexes, fig.cap = "Selection effects by sex. Highlighted lines are significant differences at p < 0.05/33. Highlighted points are significantly different from 0 at p < 0.05/66."}
+```{r plot-sexes, fig.cap = "Selection effects by sex. Solid lines are significant differences at p < 0.05/33. Solid points are significantly different from 0 at p < 0.05/66."}
 
@@
               )) +
-      scale_linetype_manual(values = c("TRUE" = 1, "FALSE" = 2))
+      scale_linetype_manual(values = c("TRUE" = 1, "FALSE" = 3))
 
@@
       standard_ggplot(fill_col = `N partners`, score_col = score_name, 
-                        n_regs = n_regs) +
+                        n_regs = n_regs, fill_direction = -1) +
         facet_grid(cols = vars(sex))
@@
       standard_ggplot(fill_col = Household, score_col = score_name, 
-                        n_regs = n_regs) +
+                        n_regs = n_regs, fill_direction = -1) +
         facet_grid(cols = vars(sex))
@@
                     score_col = score_name, 
-                    order_idx = 5
+                    order_idx = 5,
+                    fill_direction = -1
                   )
@@
     ) %>% 
-    standard_ggplot(fill_col = Control, n_regs = n_regs, order_idx = 2)
+    standard_ggplot(fill_col = Control, n_regs = n_regs, order_idx = 2,
+                      fill_direction = -1)
 
@@
 
-```{r plot-age-flb-dv, fig.cap = "Effects of polygenic scores on age at first live birth."}
+```{r plot-age-flb-dv, fig.cap = "Effects of polygenic scores on age at first live birth.", fig.align = "center"}
+
+drake::loadd(res_age_flb_dv)
 
@@
 
-```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at respondent's birth, eldest siblings. Weights multiplied by 1/number of siblings."}
+```{r plot-age-birth-parents-dv, fig.cap = "Effects of polygenic scores on parents' age at respondent's birth, eldest siblings. Weights multiplied by 1/number of siblings.", fig.align = "center"}
+
+drake::loadd(res_age_birth_parents_dv)
 
@@
       ) %>% 
-      standard_ggplot(fill_col = `Dependent variable`, n_regs = n_regs)
+      standard_ggplot(fill_col = `Dependent variable`, n_regs = n_regs, 
+                        fill_direction = -1)
 
@@
 
+n_regs <- nrow(res_mediation)
 res_mediation %<>% mutate(
                      # 0.025 not 0.05 because two-sided
-                     sig  = abs(statistic_ind) > qnorm(1 - 0.025/33),
+                     sig  = abs(statistic_ind) > qnorm(1 - 0.025/n_regs),
                      star = ifelse(sig, "*", "")
@@
 
-footnote <- sprintf("* p < 0.05/33. Analysis run on %s PGS which correlated 
-                    significantly with fertility.", nrow(res_mediation))
+footnote <- sprintf("* p < 0.05/%s. Analysis run on %s PGS which correlated significantly with fertility.", n_regs, n_regs)
 
@@
 Table \@ref(tab:tbl-res-mediation) shows results. For `r n_pos_sig` out of 
-`r nrow(res_mediation)` scores, the indirect effect on fertility via human 
-capital is significantly different from 0 at $p = 0.05/33$ and has the
-same sign as the total effect. We also calculate
-the proportion of the total effect that is mediated via the indirect effect,
-along with uncorrected 95% confidence intervals (100 bootstraps).
+`r n_regs` scores, the indirect effect on fertility via human capital is
+significantly different from 0 at $p$ = 0.05/`r n_regs` and has the same sign
+as the total effect. We also calculate the proportion of the total effect that
+is mediated via the indirect effect, along with uncorrected 95% confidence
+intervals (100 bootstraps).
 
@@
   ) %>% 
-  standard_ggplot(fill_col = Controls, n_regs = n_regs,
-                  fill_direction = -1) +
+  standard_ggplot(fill_col = Controls, n_regs = n_regs, fill_direction = -1) +
     guides(alpha = guide_legend(
@@
       filter(dep.var == "RLRS") %>% 
-      standard_ggplot(fill_col = PGS, n_regs = n_regs)
+      standard_ggplot(fill_col = PGS, n_regs = n_regs, fill_direction = -1)
 
@@
       filter(dep.var == "RLRS_parents") %>% 
-      standard_ggplot(fill_col = PGS, n_regs = n_regs)
+      standard_ggplot(fill_col = PGS, n_regs = n_regs, fill_direction = -1)
 

### 2022-05-04 why-natural-selection.Rmd
Final final final tweaks. Never say final.

@@
         Weights  = recode(Weights, "Main" = "Weighted", "None" = "Unweighted"),
-        term     = pretty_names(term),
-        term     = fct_reorder(term, estimate, order_abs(1))
-      ) %>% 
-      standard_ggplot(fill_col = Weights, n_regs = n_regs, fill_direction = -1)
+      ) %>% 
+      standard_ggplot(fill_col = Weights, n_regs = n_regs, fill_direction = -1,
+                        order_idx = 2)
 
@@
 
-```{r plot-cors-earnings-educ, fig.cap="Selection effects by correlations with earnings and educational attainment. Each point represents one polygenic score."}
+```{r plot-cors-earnings-educ, fig.cap="Selection effects by correlations with earnings and educational attainment. Each point represents one polygenic score. Selected scores are annotated."}
 
@@
 attainment (EA3) among children from households of different income groups. The
-blue bars show the actual means, i.e. parents' mean polygenic score weighted by
+red bars show the actual means, i.e. parents' mean polygenic score weighted by
 number of children. The grey bars show the hypothetical means if all households
@@
       standard_ggplot(fill_col = sex, n_regs = n_regs, score_col = term,
-                        conf_int = FALSE) +
+                        conf_int = FALSE, fill_direction = -1) +
       geom_linerange(data = res_sex_wide, inherit.aes = FALSE, 
@@
 res_sibs_kids <-  bind_rows(
-                    Siblings = res_sibs_parent_weights, 
-                    Children = res_children_comparison,
-                    .id = "Dep. var."
-                  )
-
-standard_ggplot(res_sibs_kids, `Dep. var.`, n_regs = length(score_names))
+                    Parents = res_sibs_parent_weights, 
+                    Respondents = res_children_comparison,
+                    .id = "Generation"
+                  )
+
+standard_ggplot(res_sibs_kids, Generation, n_regs = length(score_names),
+                  fill_direction = -1)
 ```

### 2022-05-04 response-to-reviewers-BE.Rmd
Edit response

@@
 
-We have made it more explicit: "natural selection may substantially increase the correlation between genetic scores and income, increasing genetic differences between different social groups, and thus making the "genetic lottery" (Harden 2021) more unfair."
+We have made it more explicit: "natural selection substantially increases the correlation between polygenic scores and income, increasing genetic differences between different social groups, and thus making the "genetic lottery" (Harden 2021) more unfair."
 
@@
 We prefer to be "conservative" in controlling for it, in the sense that we'd rather bias the estimate indirect effect of EA towards zero than away from zero.
-We now mention this issue in the appendix: "Note that controlling for fluid IQ is conservative, in the sense that IQ might also measure human capital; nevertheless, we control for it because fluid IQ might affect fertility decisions via other channels than human capital."
+We now mention this issue in the appendix: "Note that controlling for fluid IQ is conservative, in the sense that IQ might also measure human capital in the labour market; nevertheless, we control for it because fluid IQ might affect fertility decisions via other channels than human capital."
 
@@
 We had pushback from people saying "this is no big deal", so we want to show that at least under some assumptions, yes it might be.
-The within-family estimate should remove many of the worries about exogeneity of p\*, but they are estimated with a smaller sample.
+The within-family estimate should remove many of the worries about exogeneity of p\*, but it is estimated with a smaller sample.
 We agree that the estimate seems high, and indeed, the bootstrapped 95% confidence interval is from around 0% to 300%, which is not really informative, though it does suggest that genetic nurture is more likely to bias our results downwards than upwards.
@@
 Note that Fisher, cited by Reviewer 1, used observed correlations quite like ours to differentiate between theories of the fertility-income relationship.
-Generally, significant correlations have often led to important theoretical literatures, even when the direction of causality is initially uncertain (e.g.
+Generally, significant correlations have often led to important theoretical literatures, even when causality is initially uncertain (e.g.
 education/wages in labour economics, or democracy/GDP levels in political economy).
@@
 
-*PGS effect on fertility = PGS correlation with education X education effect on fertility + direct effect of PGS on fertility*
+*PGS effect on fertility = PGS correlation with education* $\times$ *education effect on fertility + direct effect of PGS on fertility*
 
@@
 As the reviewer suggests, current household income is very much endogenous not only to human capital, but also to previous fertility, which makes it inappropriate as a mediator.
+We welcome any other ideas.
+See also our response to Reviewer 1's comment on fluid IQ.
 
@@
 We now say: 'We report the number of scores showing significant changes over time (i.e.
-a significant interaction between polygenic score and the "late born" dummy): either a significant change in sign, a significant increase in size, or a significant decrease in size.'
+a significant interaction between polygenic score and the "late born" dummy): either a significant change in sign, a significant increase in effect size, or a significant decrease in size.'
 

### 2022-05-04 why-natural-selection.Rmd
typos and fixes for edits

@@
 Figure \@ref(fig:plot-sexes) plots selection effects by sex. Differences are
-particularly large for educational attainment, height and MDD. Several polygenic
-scores for mental illness and personality traits are more selected for (or less
-against) among women, including major depressive disorder (MDD), schizophrenia
-and neuroticism, while extraversion is more selected for among men. Scores for
-waist circumference and waist-hip ratio are less selected for among women. One
-possible reason for these sex differences is that polygenic scores may affect
-fertility via success in marriage markets, and men and women may value different
-characteristics in these.
+particularly large for educational attainment, height, ADHD and MDD. Several
+polygenic scores for mental illness and personality traits are more selected for
+(or less against) among women, including major depressive disorder (MDD),
+schizophrenia and neuroticism, while extraversion is more selected for among
+men.
 
@@
 
-We report results for 3 alternative weighting schemes: weighting by
-age/qualification; geographical (weighting by Middle Super Output Area); and for
-women only, age, qualification and age at first live birth. Population data for
-weighting is taken from the 2011 UK Census and the 2006 General Household Survey
-(GHS). Weighting for Age/Qualification and Age/Qualification/AFLB weights was
-done using marginal totals from a linear model, using the `calibrate()` function
-in the R "survey" package [@lumley2020survey]. Geographical weighting was done
-with iterative post-stratification using the `rake()` function, on Census Middle
-Layer Super Output Areas, sex and presence/absence of a partner.
+We compare results for our main weights to 3 alternative weighting schemes:
+weighting by age/qualification; geographical (weighting by Middle Super Output
+Area); and for women only, age, qualification and age at first live birth.
+Population data for weighting is taken from the 2011 UK Census and the 2006
+General Household Survey (GHS). Weighting for Age/Qualification and
+Age/Qualification/AFLB weights was done using marginal totals from a linear
+model, using the `calibrate()` function in the R "survey" package
+[@lumley2020survey]. Geographical weighting was done with iterative
+post-stratification using the `rake()` function, on Census Middle Layer Super
+Output Areas, sex and presence/absence of a partner.
 
@@
 
-Table \@ref(tab:tbl-change-by-period) 
-summarizes the results. We report the number of scores showing significant
-changes over time (i.e. a significant interaction between polygenic score and
-the "late born" dummy): either a significant change in sign, a significant
-increase in size, or a significant decrease in size. There is little evidence
-for changes in selection effects within the parents' generation, with just one
-score showing a significant decrease in size. In the respondents' generation,
-effect sizes were significantly larger in absolute size among the later-born for
-eight polygenic scores: ADHD, age at menopause, cognitive
-ability, Coronary Artery Disease, EA2, EA3, extraversion and Major Depressive
-Disorder. These changes are inconsistent with the intergenerational change,
-where estimated effect sizes were larger among the earlier, parents' generation.
+Table \@ref(tab:tbl-change-by-period) summarizes the results. We report the
+number of scores showing significant changes over time (i.e. a significant
+interaction between polygenic score and the "late born" dummy): either a
+significant change in sign, a significant increase in effect size, or a
+significant decrease in size. There is little evidence for changes in selection
+effects within the parents' generation, with just one score showing a
+significant decrease in size. In the respondents' generation, effect sizes were
+significantly larger in absolute size among the later-born for eight polygenic
+scores: ADHD, age at menopause, cognitive ability, Coronary Artery Disease, EA2,
+EA3, extraversion and Major Depressive Disorder. These changes are inconsistent
+with the intergenerational change, where estimated effect sizes were larger
+among the earlier, parents' generation.
 
@@
 effect sizes are negatively correlated with the effect sizes from bivariate
-regressions without the age at birth control (father's age at birth: $\rho$ 
+regressions without the control for age at birth (father's age at birth: $\rho$ 
 `r cor_unc_fath`; mother's age at birth: $\rho$ `r cor_unc_moth`).

### 2022-05-14 why-natural-selection.Rmd
Fix abstract error; finally order guides right and justify them; tweak colours

@@
 We test for natural selection through the association between 33 polygenic scores and fertility, across two generations, using data from UK Biobank (N = 409,629 British subjects with European ancestry).
-Consistently over time, polygenic scores that predict lower earnings, education and health also predict lower fertility.
+Consistently over time, polygenic scores that predict higher earnings, education and health also predict lower fertility.
 Selection effects are concentrated among lower SES groups, younger parents, people with more lifetime sexual partners, and people not living with a partner.
@@
                        values = c("95% c.i. uncorrected" = 1),
-                       guide  = guide_legend(order = 1)
+                       guide  = guide_legend(order = 3)
                      )
@@
 
+  my_shape_fill_guide <- guide_legend(order = 1)
   n_regs <- as.double(n_regs)
@@
     geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/{{n_regs}})) +
-    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction) +
-    scale_shape_manual(values = shape_values) +
+    my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction, 
+                    guide = my_shape_fill_guide) +
+    scale_shape_manual(values = shape_values, guide = my_shape_fill_guide) +
     my_labs() +
-    my_theme() +
+    my_theme(legend.justification = "top") +
     confint_scale +
@@
                            guide = guide_legend(
-                             order = 0,
+                             order = 2,
                              override.aes = list(
@@
 
-my_theme <- function (...) theme(axis.text.y = element_text(size = 7), ...)
+my_theme <- function (...) {
+  theme(
+          axis.text.y = element_text(size = 7), 
+          ...
+        )
+}
 
@@
     #                                  direction = direction, end = 0.6, ...)
+    fill_colors <- c(fill_colors[1], "skyblue", fill_colors[2])
     if (direction != -1) fill_colors <- rev(fill_colors)
@@
                          values = shades::gradient(fill_colors,
-                                                     steps = n), 
+                                                     steps = n, space = "Lab"), 
                          ...) 
@@
                           xmax = prop_ind_conf_high,
-                          linetype = "Bootstrap 95% c.i.",
+                          linetype = "Bootstrap 95% c.i. uncorrected",
                         ),
-                        width = 0.3,
+                        width = 0.2,
                         color = "grey50"
                       ) +
-        scale_fill_manual(values = c("FALSE" = "grey65", "TRUE" = "orange2")) +
+        scale_fill_manual(values = c("FALSE" = "grey65", "TRUE" = "blue3")) +
         my_vline +
@@
         labs(x = "Proportion mediated", y = "") +
-        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i." = 1))
+        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i. uncorrected" = 1))
 
@@
 had equal numbers of children. Natural selection against genes associated with
-educational attainment, which lowers the mean, is stronger at the bottom of the
-income distribution, and this increases the differences between groups.
-Overall, natural selection increases the correlation of polygenic scores with
-income for `r n_increases` out of 33 polygenic scores, with a median percentage
-increase of `r median_pct_change`% in the respondents' generation (Appendix
-Table \@ref(tab:tbl-inequality-effects)).  If inequalities in polygenic scores
-are important for understanding social structure and mobility
-[@belsky2018genetic; @Rimfeld_2018; @harden2021genetic], then these increases
-are substantive. Similarly, since many polygenic scores are predictive of
-disease risk, they could potentially increase health inequalities. In general,
-the evolutionary history of anatomically modern humans is related to disease
-risk [@benton2021influence]; understanding the role of contemporary natural
-selection may help researchers to map the genetic architecture of health
-disparities.
+educational attainment is stronger at the bottom of the income distribution, and
+this increases the differences between groups. Overall, natural selection
+increases the correlation of polygenic scores with income for `r n_increases`
+out of 33 polygenic scores, with a median percentage increase of 
+`r median_pct_change`% in the respondents' generation (Appendix Table
+\@ref(tab:tbl-inequality-effects)).  If inequalities in polygenic scores are
+important for understanding social structure and mobility [@belsky2018genetic;
+@Rimfeld_2018; @harden2021genetic], then these increases are substantive.
+Similarly, since many polygenic scores are predictive of disease risk, they
+could potentially increase health inequalities. In general, the evolutionary
+history of anatomically modern humans is related to disease risk
+[@benton2021influence]; understanding the role of contemporary natural selection
+may help researchers to map the genetic architecture of health disparities.
 

### 2022-05-14 why-natural-selection.Rmd
More colour tweaks

@@
   my_shape_fill_guide <- guide_legend(order = 1)
+
   n_regs <- as.double(n_regs)
@@
 
-pval_geom_point <- function (size = 1.6, stroke = 0.9, fill = NULL, ...) {
-  if (is.null(fill)) {
-    geom_point(shape = 21, size = size, stroke = stroke, ...)
-  } else {
-    geom_point(shape = 21, size = size, stroke = stroke, fill = fill, ...)
-  }
-}
-
 confint_geom_segment <- function () {
@@
     #                                  direction = direction, end = 0.6, ...)
-    fill_colors <- c(fill_colors[1], "skyblue", fill_colors[2])
+    fill_colors <- c(fill_colors[1], "#008b8b", fill_colors[2])
     if (direction != -1) fill_colors <- rev(fill_colors)

### 2022-05-15 why-natural-selection.Rmd
Further graphical improvements:   - use color and fill for p values   - fewer grid lines   - many tweaks   - remove cruft

@@
                      )
-    confint_geom_segment <- confint_geom_segment()
+    confint_geom_segment <- geom_segment(
+                              aes(
+                                x        = conf.low, 
+                                xend     = conf.high,
+                                linetype = "95% c.i. uncorrected"
+                              ), 
+                              color   = "grey45", 
+                              alpha   = 0.2, 
+                              size    = 1.1, 
+                              lineend = "round"
+                            )
   } else {
@@
   n_regs <- as.double(n_regs)
-  shape_values <- c(16, 15, 17, 18, 20)[1:n_cats]
+  shape_values <- paste(c("circle", "square", "triangle", "diamond", 
+                            "triangle down"), "filled")
+  shape_values <- shape_values[1:n_cats]
   if (fill_direction == -1) shape_values <- rev(shape_values)
@@
     confint_geom_segment +
-    geom_point(size = 1.8, alpha = 0.9, aes(color = {{fill_col}})) +
-    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/{{n_regs}})) +
+    geom_point(size = 1.8, alpha = 0.8, aes(color = {{fill_col}}, 
+                 fill = stage(p < 0.05/{{n_regs}}, after_scale = ifelse(
+                 fill == "white", fill, color)))) +
     my_fill_scale(aesthetics = "color", n = n_cats, direction = fill_direction, 
                     guide = my_shape_fill_guide) +
+    scale_fill_manual(values = c("TRUE" = "black", "FALSE" = "white"), 
+                        guide = guide_legend(
+                                  title = sprintf("p < 0.05/%s", n_regs),
+                                  override.aes = list(shape = "circle filled"),
+                                  order = 2
+                                )) + 
     scale_shape_manual(values = shape_values, guide = my_shape_fill_guide) +
     my_labs() +
-    my_theme(legend.justification = "top") +
-    confint_scale +
-    scale_alpha_discrete(range = c(1, 0), 
-                           guide = guide_legend(
-                             order = 2,
-                             override.aes = list(
-                               shape = c(1, 16),
-                               color = "black",
-                               alpha = 1,
-                               size = 1.8
-                             )
-                         )) 
+    my_theme(
+      legend.justification = "top", 
+      panel.grid.major.y   = element_blank()
+    ) +
+    confint_scale
 }
@@
           axis.text.y        = element_text(size = 7),
+          panel.grid.major   = element_line(size = rel(0.5)),
           ...
         )
-}
-
-confint_geom_segment <- function () {
-  geom_segment(aes(x = conf.low, xend = conf.high, linetype = "95% c.i. uncorrected"), 
-                    color = "grey55", alpha = 0.2, size = 1.1, 
-                    lineend = "round")
 }
@@
         geom_text(aes(label = PGS), size = 3, colour = "black", 
-                    check_overlap = TRUE, nudge_x = 0.002) +
+                    check_overlap = TRUE, nudge_x = 0.008) +
         facet_wrap( ~ Correlation, scales = "free") + 
@@
                       ) +
-        scale_fill_manual(values = c("FALSE" = "grey65", "TRUE" = "blue3")) +
+        scale_fill_manual(values = c("FALSE" = "grey65", "TRUE" = "darkgreen")) +
         my_vline +
@@
         labs(x = "Proportion mediated", y = "") +
-        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i. uncorrected" = 1))
+        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i. uncorrected" = 1)) +
+        my_theme(
+          panel.grid.major.y = element_blank(),
+          panel.grid.minor.x = element_blank()
+        )
 
@@
 attainment (EA3) among children from households of different income groups. The
-red bars show the actual means, i.e. parents' mean polygenic score weighted by
+green bars show the actual means, i.e. parents' mean polygenic score weighted by
 number of children. The grey bars show the hypothetical means if all households
@@
 ```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean polygenic score for educational attainment (EA3) of children by household income group.
-Blue is actual.
+Green is actual.
 Grey is hypothetical in the absence of selection effects.", fig.width = 4}
@@
                 geom_col(position = "dodge") +
-                scale_fill_manual(values = c("darkred", "grey65")) +
+                scale_fill_manual(values = c("darkgreen", "grey65")) +
                 theme(
-                  legend.background = element_rect(fill = "white", colour = NA),
+                  legend.background = element_rect(
+                                        fill = "white", 
+                                        colour = "black",
+                                        size = rel(0.5)
+                                      ),
+                  legend.margin = margin(t = 0, b = 5.5, l = 5.5, 
+                                           r = 5.5),
                   legend.title = element_blank(), 
@@
 
-```{r plot-age-birth-parents-cross, fig.cap = "Selection effects (parents' RLRS) by age at first live birth terciles.
+```{r plot-age-birth-parents-cross, fig.cap = "Selection effects (parents' RLRS) among eldest siblings, by parents' age at first live birth terciles.
 Weights multiplied by 1/number of siblings.", fig.subcap = c("Mothers", "Fathers"), fig.ncol = 1}
@@
 
-```{r plot-age-birth-parents, fig.cap = "Selection effects (parents' generation) among eldest siblings, controlling for parents' age at birth. Weights multiplied by 1/number of siblings."}
+```{r plot-age-birth-parents, fig.cap = "Selection effects (parents' RLRS) among eldest siblings, controlling for parents' age at birth. Weights multiplied by 1/number of siblings."}
 

### 2022-05-17 why-natural-selection.Rmd
update van alten citation

@@
 sample.[^balance-diverse-def] To correct for ascertainment bias, we use
-participant weights from @van2021effects, which match the UK Biobank eligible
+participant weights from @vanalten2022reweighting, which match the UK Biobank eligible
 population on sex, birth year, location, education, employment, health,
@@
 
-Weighting data was kindly provided by @van2021effects.
+Weighting data was kindly provided by @vanalten2022reweighting.
 

### 2022-05-30 why-natural-selection.Rmd
Use condensed font in graphs

@@
         error      = FALSE,
-        fig.height = 3.5
+        fig.height = 3.5,
+        dev        = "cairo_pdf"
       )
@@
 theme_set(theme_minimal())
+theme_update(
+  text = element_text(family = "Abadi MT Condensed Light")
+)
 
@@
         line        = element_line(size = 0.5, lineend = "round"),
-        strip.text  = element_text(size = 6), 
+        strip.text  = element_text(size = 8), 
         axis.text.x = element_text(size = 7),
@@
         geom_text(aes(label = PGS), size = 3, colour = "black", 
-                    check_overlap = TRUE, nudge_x = 0.008) +
+                    family = "Abadi MT Condensed Light",
+                    check_overlap = TRUE, nudge_x = 0.004) +
         facet_wrap( ~ Correlation, scales = "free") + 
@@
         labs(x = "Proportion mediated", y = "") +
-        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i. uncorrected" = 1)) +
+        scale_linetype_manual(name = "", 
+          values = c("Bootstrap 95% c.i. uncorrected" = 1)) +
         my_theme(
           panel.grid.major.y = element_blank(),
-          panel.grid.minor.x = element_blank()
+          panel.grid.minor.x = element_blank(),
+          legend.justification = "top"
         )
@@
         geom_text(aes(estimate, dep.var, label = top_pcs), position = pj, 
-                    hjust = "right", color = "black", size = 3) +
+                    hjust = "right", color = "black", size = 3,
+                    family = "Abadi MT Condensed Light") +
         scale_x_log10(labels = scales::label_number(accuracy = 0.0001)) +

### 2022-06-09 why-natural-selection.Rmd
Changes for final revision:   - use fewer controls in mediation analysis   - only show confidence intervals in mediation analysis plot when they are correct   - tiny colour tweaks

@@
 ```{r plot-res-wt, fig.cap = "Selection effects: weighted and unweighted regressions.
-Each point represents a single bivariate regression of number of children on a polygenic score.
+Each point represents a single bivariate regression of RLRS on a polygenic score.
 P value threshold is 0.05, Bonferroni-corrected for multiple comparisons.
@@
                                                             nrow(res_mediation)),
-                     pos_sig = sig & prop_ind > 0
+                     pos_sig = sig & prop_ind > 0,
+                     bounded_ci = sign(estimate_total_conf_low) ==
+                                  sign(estimate_total_conf_high)
                    )
@@
 
-```{r plot-mediation, fig.align = "center", fig.cap = "Proportion of selection effect mediated by educational attainment, among polygenic scores with significant selection effects"}
-
-res_mediation %>% 
+(ref:franzcite) [@franz2007ratios]
+
+```{r plot-mediation, fig.align = "center", fig.cap = "Proportion of selection effect mediated by educational attainment, among polygenic scores with significant selection effects.
+Bootstrap confidence intervals for the proportion are shown only where the interval is bounded (ref:franzcite)."}
+
+res_mediation <- res_mediation %>% 
       mutate(
@@
         `Indirect effect p < 0.05/23` = sig 
-      ) %>% 
+      )
+
+res_mediation %>% 
       ggplot(aes(y = term)) + 
@@
                         ),
+                        data = res_mediation %>% filter(bounded_ci),
                         width = 0.2,
-                        color = "grey50"
+                        color = "grey20"
                       ) +
-        scale_fill_manual(values = c("FALSE" = "grey65", "TRUE" = "darkgreen")) +
+        scale_fill_manual(values = c("FALSE" = "grey65", "TRUE" = "steelblue4")) +
         my_vline +
@@
 attainment (EA3) among children from households of different income groups. The
-green bars show the actual means, i.e. parents' mean polygenic score weighted by
+blue bars show the actual means, i.e. parents' mean polygenic score weighted by
 number of children. The grey bars show the hypothetical means if all households
@@
 ```{r plot-mean-EA3-by-income, fig.align = "center", fig.cap = "Mean polygenic score for educational attainment (EA3) of children by household income group.
-Green is actual.
+Blue is actual.
 Grey is hypothetical in the absence of selection effects.", fig.width = 4}
@@
                 geom_col(position = "dodge") +
-                scale_fill_manual(values = c("darkgreen", "grey65")) +
+                scale_fill_manual(values = c("steelblue4", "grey65")) +
                 theme(
@@
 education), and $X_i$ is a vector of controls. The total effect of $PGS$ on
-number of children is $\beta + \gamma \zeta$. The "indirect effect" mediated by
+$RLRS$ is $\beta + \gamma \zeta$. The "indirect effect" mediated by
 $EA$ is $\gamma \zeta$. The standard error of the indirect effect can be
@@
 
-where $\hat\sigma_\zeta$ is the standard error of $\hat\zeta$, etc.
-
-We are not concerned about endogeneity of polygenic scores, since our theory
-requires only that these correlate with human capital. However, endogeneity of
-educational attainment is a concern, i.e., educational attainment might
-correlate with unobserved characteristics that affect one's fertility (and which
-are not related to human capital). To partially address this, we include
-controls for age, sex, fluid IQ, height and risk attitude in $X$. Note that
-controlling for fluid IQ is conservative, in the sense that IQ might also measure
-human capital in the labour market; nevertheless, we control for it because 
-fluid IQ might affect fertility decisions via other channels than human capital.
+where $\hat\sigma_\zeta$ is the standard error of $\hat\zeta$, etc. We include 
+controls for age and sex in $X$.
 

### 2022-06-09 regressions.R

@@ -184,19 +184,6 @@ run_regs_age_flb_parents_cross <- function(famhist, score_names, age_birth_var) 
 }
 
 
-run_regs_mnlogit <- function (score_names, fhl_mlogit) {
-  loadNamespace("mnlogit")
-  run_1_mnlogit <- function(score_name) {
-    fml <- as.formula(glue::glue("n_ch_fac ~ 1 | {score_name}"))
-    res <- try(mod <- mnlogit::mnlogit(fml, fhl_mlogit, ncores = 2))
-    if (inherits(res, "try-error")) return(res)
-    mod$model <- NULL # otherwise it stores a copy of the entire data 8-(
-    return(mod)
-  }
-  lapply(score_names, run_1_mnlogit)
-}
-
-
 run_age_anova <- function (score_name, control, famhist) {
   fml <- as.formula(glue::glue(
     "RLRS ~ {score_name}*({control} + 
@@ -383,7 +370,7 @@ run_mediation <- function (famhist, res_all) {
   
   run_one_mediation <- function (score_name, famhist = famhist) {
     
-    controls <- "age_at_recruitment + sex + fluid_iq + height + f.2040.0.0"
+    controls <- "age_at_recruitment + sex"
     f_mediator <- as.formula(glue::glue("age_fulltime_edu ~ {score_name} +
                                            {controls}"))
 
@@ -445,7 +432,8 @@ run_mediation <- function (famhist, res_all) {
                                              # weight_by = famhist$weights
                                            )
                               br <- purrr::map_dfr(sig_scores, 
-                                                   run_one_mediation, famhist = fh_boot)
+                                                   run_one_mediation, 
+                                                   famhist = fh_boot)
                               br <- br[c("term", "estimate_total", "estimate_ind")]
                               br["rep"] <- r
                               br
@@ -459,8 +447,12 @@ run_mediation <- function (famhist, res_all) {
                prop_ind = estimate_ind/estimate_total
              ) %>%
              summarize(
-               prop_ind_conf_low   = quantile(prop_ind, 0.025),
-               prop_ind_conf_high  = quantile(prop_ind, 0.975)
+               estimate_ind_conf_low    = quantile(estimate_ind, 0.025),
+               estimate_ind_conf_high   = quantile(estimate_ind, 0.975),
+               estimate_total_conf_low  = quantile(estimate_total, 0.025),
+               estimate_total_conf_high = quantile(estimate_total, 0.975),
[…]

### 2022-12-09 why-natural-selection.Rmd
get back to published version of paper

@@
         `Proportion (%)`  = prop_ind * 100,
+        bounded_ci        = sign(estimate_total_conf_low) ==
+                              sign(estimate_total_conf_high),
         `Proportion 95% c.i. (%)` = glue::glue(
-                     "[{prop_ind_conf_low * 100}, {prop_ind_conf_high * 100}]"
-                     )
-      ) %>%
+                     "[{prop_ind_conf_low * 100}, {prop_ind_conf_high * 100}]"),
+        `Proportion 95% c.i. (%)` = ifelse(bounded_ci, 
+                                             `Proportion 95% c.i. (%)`,
+                                             "Unbounded")
+      ) %>% 
+      select(-bounded_ci) %>%
       as_hux() %>%
@@
 is mediated via the indirect effect, along with uncorrected 95% confidence
-intervals (100 bootstraps).
+intervals (100 bootstraps). Note that if the confidence interval for the total
+effect contains zero, the confidence interval for the proportion may be unbounded
+[@franz2007ratios].
 
